### 2011-08-04 test/DESeq/suppl_ii.Rnw
DESeq is used for OMZ175.

\documentclass{article}

\usepackage{Sweave}
\usepackage[a4paper]{geometry}
\usepackage{graphicx}

% Labels, hard-coded to make sure they stay fixed
\input{labels.tex}

\SweaveOpts{keep.source=TRUE,eps=FALSE,width=4,height=4}

% meant to be run with R 2.11, DESeq 1.1.6, edgeR 1.6.0

\begin{document}

\begin{minipage}{\textwidth}
\center
{\large{\textbf{\textsf{Differential expression analysis for sequence count data}}}}
\medskip

Simon Anders, Wolfgang Huber
\bigskip

\LARGE{\textsc{\textsf{\textbf{Supplement II}}}}
\vspace{4ex}
\end{minipage}

\noindent In order to make the analyses presented in our paper transparent and reproducible, 
we document in the present supplement the complete R code used to perform
them and to create all the plots in the paper and in Supplement I. The purpose of this 
document is \emph{not} to explain the use of the \emph{DESeq} package; for this, 
please read the vignette (manual) supplied with the package. 
\medskip

\section{Required packages}

We start with loading all requried packages:
<<results=hide>>=
library( DESeq )
library( edgeR )
library( hexbin )
library( latticeExtra )
library( gplots )
library( geneplotter )
@

This Sweave file has been run with these package versions:

<<>>=
sessionInfo()
@

\section{Reading in the data}

All data files necessary to run the code shown here are supplied in the supplementary data 
tarball available with the paper.

\subsection{Fly RNA-Seq data}

The RNA-Seq data from gene overexpression in fly embryos is unpublished, an excerpt
of which was kindly provided to us by Wilczynski et al.\ (Furlong Lab, EMBL Heidelberg)
in the form of the following table (file \texttt{fly\_RNA\_counts.tsv}, supplied 
with supplementary data):

<<>>=
countsTableFly <- read.delim( "fly_RNA_counts.tsv" )
condsFly <- c( "A", "A", "B", "B" )
# Add dummy names to avoid confusion later
rownames( countsTableFly ) <- paste( "Gene", 1:nrow(countsTableFly), sep="_" )
@

\subsection{Neural stem cell RNA-Seq data}

The RNA-Seq data comparing neural stem cell lines from gliablastomas and non-cancerous
tissue was kindly provided to us by Engstr\"om et al.\ (Bertone Lab, EMBL-EBI Hinxton)
in the form of the following table (file \texttt{NeuralStemCellData.tab}, supplied 
with supplementary data):

<<>>=
countsTableNS <- read.delim( "NeuralStemCellData.tab", row.names=1 )

head(countsTableNS)
@

The conditions are

<<>>=
condsNS <- c( "GNS_dupe", "GNS", "GNS", "GNSL", "NS", "NS" )
@

where the first entry is for the sample that will be eliminated later as it is derived
from the same patient as the second one.

\subsection{Yeast RNA-Seq data}

We have downloaded the deposited reads for the yeast RNA-Seq experiment by Nagalakshmi et al.\
from the NCBI Short Read Archive, remapped them to the yeast genome with Bowtie\footnote{Langmead et al.,
Ultrafast and memory-efficient alignment of short DNA sequences to the human genome. Genome Biol.\ 10, R25} 
and the counted the coverage of each gene with 
htseq-count\footnote{http://www-huber.embl.de/users/anders/HTSeq/doc/count.html}.

We read in this data with

<<>>=
countsYeast <- read.delim( "counts_Nagalakshmi_et_al.tab", row.names=1 ) 
head( countsYeast )
@

\section{ChIP-Seq data}

The supplementary file \texttt{PolII\_counts.tsv} contains one column for each sample
in the Kasowski et al.\ paper, and one row for each polymerase-II binding region
that the authors identified, giving the number of reads counted to map within each binding
region. This file has been derived from the original data file in Kasowski et al.'s 
supplementary data with a simple reformatting.

We read in the data with

<<>>=
countsPolII <- read.delim( "PolII_counts.tsv" )

# make rownames
row.names(countsPolII) <- sprintf( "%s:5d-%d", countsPolII$chrom, 
   countsPolII$start, countsPolII$stop )

# The IDs should start with "GM", not "X":
colnames(countsPolII) <- sub( "X", "GM", colnames(countsPolII) )

# A few individuals in this data file have four or more replicates. We have picked 
# (at random) two of them:
countsPolII <- countsPolII[ , grepl( "GM12878", colnames(countsPolII) ) | 
   grepl( "GM12891", colnames(countsPolII) ) ]

head(countsPolII)
@

\section{Analysis of fly data}

\subsection{Analysis with DESeq}

For the standard DESeq analysis, we instantiate a \emph{CountDataSet}, then estimate
the size factors and the variance functions, and finally perform the binomial test.

<<>>=
cdsFly <- newCountDataSet( countsTableFly, condsFly )
cdsFly <- estimateSizeFactors( cdsFly )
cdsFly <- estimateVarianceFunctions( cdsFly )
resFly <- nbinomTest( cdsFly, "A", "B" )
@

\subsection{Analysis with edgeR}

For the edgeR analysis, we test for different modes: One choice is whether edgeR 
should run in the common (``C'') or the tag-wise (``T'') dispersion mode, the other
is whether edgeR should be given the actual number of sequenced reads (``N''), as
suggested in the edgeR manual, or the same size factors\footnote{As edgeR expects read counts, we 
multiply these by $10^7$ so that they get a plausible magnitude. edgeR only 
cares about the relative ratios of the \texttt{lib.size} values, and hence, this factor 
is unimportant. However, it expect them to be large numbers (it seems to round them 
to integers), and hence, the ``\texttt{1e7*}'' below ensures that no information is lost in 
rounding. Using another large number does not change the results.} as DESeq used (``D'').

We instantiate four \emph{DGEList} objects and perform the analysis four times:

<<>>=
# using DESeq's size factors:

dglFlyD <- DGEList( counts=countsTableFly, group=condsFly, lib.size=1e7*sizeFactors(cdsFly) )
dglFlyD <- estimateCommonDisp( dglFlyD )
dglFlyD <- estimateTagwiseDisp( dglFlyD )

# test with tag-wise dispersion:

edgerResFlyDT <- exactTest( dglFlyD, common.disp=FALSE )
edgerResFlyDTPadj <- p.adjust( edgerResFlyDT$table$p.value, method="BH" )

# common dispersion, DESeq's size factors:

edgerResFlyDC <- exactTest( dglFlyD, common.disp=TRUE )
edgerResFlyDCPadj <- p.adjust( edgerResFlyDC$table$p.value, method="BH" )

# with the original library sizes:

# the total number of sequenced counts were as follows:
libsizesFly <- c( A1=13613446, A2=12106517, B1=12920058, B2=11599459 )

dglFlyN <- DGEList( counts=countsTableFly, group=condsFly, lib.size=libsizesFly )
dglFlyN <- estimateCommonDisp( dglFlyN )
dglFlyN <- estimateTagwiseDisp( dglFlyN )

# test with tag-wise dispersion:

edgerResFlyNT <- exactTest( dglFlyN, common.disp=FALSE )
edgerResFlyNTPadj <- p.adjust( edgerResFlyNT$table$p.value, method="BH" )

# common dispersion, DESeq's size factors:

edgerResFlyNC <- exactTest( dglFlyN, common.disp=TRUE )
edgerResFlyNCPadj <- p.adjust( edgerResFlyNC$table$p.value, method="BH" )
@

\subsection{Plots and comparison}

\subsubsection{Figures on variance estimation}

We start with the plots of Fig.\ \ref{varestFig} in the paper.

The common dispersion values estimated by edgeR are:

<<>>=
dglFlyD$common.dispersion
dglFlyN$common.dispersion
@

The values for the two library size settings hardly differ, we shall use the first one
for the dashed orange line.

<<>>=
alpha <- dglFlyD$common.dispersion
@

The \emph{DESeq} package offers the function \texttt{rawVarFunc}, that
returns the fitted raw variance function for a condition $\rho$, i.e., 
the function $\hat v_\rho(q)$ defined in Eq.\ (9) of the paper. For the solid orange line
in the plot, we define a function \texttt{basevarFunc} that returns the fitted variance
function $w_\rho(q)$ by undoing the subtraction of Eq.\ (9) that \texttt{rawVarFunc}
does internally:

<<>>=
baseVarFunc <- function( cds, cond ) {
   rvf <- rawVarFunc( cds, cond )
   sf <- sizeFactors(cds)[ conditions(cds) == cond ]
   xim <- sum(1/sf) / length(sf)
   function( q ) rvf( q ) + xim * q
}   
@

We also need the gene-wise variance estimates (for the scatter plot part of Fig.\ \ref{varestFig})

<<>>=
diagForA <- varianceFitDiagnostics( cdsFly, "A" )
@

The scatter points in the plots are actual hexagonal bins, produced with the \emph{hexbin}
package. While this looks nice, it unfortunately requires a rather involved call to routines
from the \emph{lattice} to get a proper log-log plot:

<<var_mean_fly,fig=TRUE,  width=4,height=4>>=
# pdf( "var_mean_fly.pdf", width=4, height=4 )

# two functions to make nice exponents for the axis labels

xscale.components.log <- function( ... ) {
   res = xscale.components.default( ... )
   res$bottom$labels$labels = do.call( expression, 
      lapply( res$bottom$labels$at, function(a)
         substitute( 10^b, list(b=a) ) ) )
   res
}

yscale.components.log = function( ... ) {
   res = yscale.components.default( ... )    
   res$left$labels$labels = do.call( expression, 
      lapply( res$left$labels$at, function(a)
         substitute( 10^b, list(b=a) ) ) )
   res
}

xg <- seq( log10(1/3), 
   log10(max( diagForA$baseMean )), length.out=1000 )

print( xyplot(    
   baseVar ~ baseMean, 
   diagForA[ diagForA$baseMean>0,], 
   panel = function( ... ) { 
      panel.hexbinplot( ... )

      # The solid orange line with the fitted base variance
      # (Note that we have to manually switch between natural
      # and logarithmic axis scaling, because hexbin cannot deal
      # with automatic logarithmic scaling):

      llines( xg, log10( baseVarFunc(cdsFly,"A")( 10^xg ) ),
         col="orange", lwd=2 )

      # The dashed orange line with edgeR's variance fit,
      # i.e., variance = mean + alpha * mean^2:

      llines( xg, log10( 10^xg + alpha * (10^xg)^2 ),
         col="orange", lwd=2, lty="dashed" )

      # For each sample with condition "A", plot a purple line
      # with the shot noise, variance = mean, where the mean
      # needs to be scaled by the size factor:

      for( sf in sizeFactors( cdsFly )[ conditions(cdsFly) == "A" ] )
         llines( xg, log10( 10^xg*sf ), col="purple", lwd=1.5 ) },

   trans=function(x) x^(1/6), 
   inv=function(x) x^6,
   xbins=80,
   scales = list( 
      x = list( log=TRUE, axs="i" ),
      y = list( log=TRUE, axs="i", limits=c( 3e-5,3e8 ), tick.number=8 ) ),
   xlab="mean", ylab="variance",
   xscale.components = xscale.components.log,
   yscale.components = yscale.components.log
) )

# dev.off()
@

For Fig.\ \ref{varestFig}b, the code is quite similar:

<<cv2_mean_fly, fig=TRUE,  width=4, height=4>>=
# pdf( "cv2_mean_fly.pdf", width=4, height=4 )

print(xyplot(    
   I( baseVar/baseMean^2 ) ~ baseMean, 
   diagForA[ diagForA$baseMean>0,], 
   panel = function( ... ) { 
      panel.hexbinplot( ... )

      # As we convert from a variance to an SCV, we should adjust for
      # bias, as described in Supplementary Note C. The function
      # adjustScvForBias takes care of this. It requires, as second
      # argument, the information how many samples were used for the
      # variance estimate. The 'attr(...)' expression supplies this,
      # returning in this case simply the value 2:

      llines( xg, 
         adjustScvForBias( 
	    baseVarFunc(cdsFly,"A")( 10^xg ) / (10^xg)^2, 
	    attr( rawVarFunc(cdsFly,"A")(NA), "size" ) ),
         col="orange", lwd=2 )

      # The other lines are as before, only without the log of
      # the y value but with a division by mean^2.

      llines( xg, ( 10^xg + alpha * (10^xg)^2 ) / (10^xg)^2,
         col="orange", lwd=2, lty="dashed" )

      for( sf in sizeFactors( cdsFly )[ conditions(cdsFly) == "A" ] )
         llines( xg, 1 / (10^xg * sf), col="purple", lwd=1.5 ) },

   trans = function(x) x^(1/4), 
   inv = function(x) x^4,
   xbins = 80,
   scales = list( 
      x = list( log=TRUE, axs="i" ),
      y = list( log=FALSE, axs="i", tick.number=8, limits=c(0,.2) ) ),
   xlab = "mean", ylab="squared coefficient of variation",
   xscale.components = xscale.components.log,
))

# dev.off()
@

Just for completeness, the code for Suppl.\ Fig.\ \ref{scvZoomOut}, which is nearly identical to 
the previous code block:

<<cv2_mean_full_fly, fig=TRUE,  width=4, height=4>>=
# pdf( "cv2_mean_full_fly.pdf", width=4, height=4 )

print(xyplot(    
   I( baseVar/baseMean^2 ) ~ baseMean, 
   diagForA[ diagForA$baseMean>0,], 
   panel = function( ... ) { 
      panel.hexbinplot( ... ) 
      llines( xg, 
         adjustScvForBias( 
	    baseVarFunc(cdsFly,"A")( 10^xg ) / (10^xg)^2, 
	    attr( rawVarFunc(cdsFly,"A")(NA), "size" ) ),
         col="orange", lwd=2 )
      llines( xg, ( 10^xg + alpha * (10^xg)^2 ) / (10^xg)^2,
         col="orange", lwd=2, lty="dashed" )
      for( sf in sizeFactors( cdsFly )[ conditions(cdsFly) == "A" ] )
         llines( xg, 1 / (10^xg * sf), col="purple", lwd=1.5 ) },
   trans = function(x) x^(1/4), 
   inv = function(x) x^4,
   xbins = 80,
   scales = list( x = list( log=TRUE, axs="i" ),
      y = list( log=FALSE, axs="i", tick.number=8 ) ),
   xlab = "mean", ylab="squared coefficient of variation",
   xscale.components = xscale.components.log
))

# dev.off()   
@

Finally, for the residuals ECDF plot (Fig.\ \ref{resEcdfFly}), the apropriate function
of DESeq was used:

<<resEcdf_A_fly, fig=TRUE, width=8, height=8>>=
# pdf( "resEcdf_A_fly.pdf", width=8, height=8 )

residualsEcdfPlot( cdsFly, "A" )

# dev.off()
@

\subsection{Figures for DE analysis}

For Fig.\ \ref{DEFig}, we plot one dot for each gene, using semi-transparent colour values
(i.e., specified with alpha channel) in order to make overplotting of dots apparent.
(The \emph{hexbin} package cannot be used as we have two colours, black and red,
to indicate significance.) Extra code, in the form of a panel function, is needed in 
order to draw the small arrows indicating genes with fold changes outside the displayed
value range. We also customise the axis function to get nicer labels.

<<DE_fly, fig=TRUE,  width=4, height=4>>=
# pdf( "DE_fly.pdf", width=4, height=4 )

print(xyplot( 
   log2FoldChange ~ I( baseMean ),
   resFly,  
   pch=16, cex=.3, 
   col=ifelse( resFly$padj < .1, "#FF000040", "#00000040" ),
   panel = function( x, y, col, ... ) {
      above <- (y > 5.8)
      below <- (y < -5.8)
      inside <- !( above | below )
      panel.xyplot( x=x[inside], y=y[inside], col=col[inside], ... ) 
      panel.arrows( x[above], 5.8, x[above], 5.95, col=col[above],
         length=".1", unit="native" )
      panel.arrows( x[below], -5.8, x[below], -5.95, col=col[below],
         length=".1", unit="native" ) },
   axis = function( side, ... ) {
      if( side=="left") {
         panel.axis( side, outside=TRUE, at=seq(-14,14,by=1), labels=FALSE )
         panel.axis( side, outside=TRUE, at=seq(-10,10,by=5), labels=TRUE )
      } 
      if( side=="bottom") {
         panel.axis( side, outside=TRUE, at=seq(-2,10,by=1), rot=0,
            labels = do.call( expression, 
	       lapply( seq(-2,10,by=1), function(a) 
	          substitute( 10^b, list(b=a) ) ) ) )
      } },
   xlab = "mean", ylab = "log2 fold change",
   scales = list( 
      x = list( log=TRUE ), 
      y = list( log=FALSE, limits=c( -6, 6 ) ) ) ))

# dev.off()
@

The volcano plot (Fig.\ \ref{vulcano}) has been produced with:

<<vulcano_fly, fig=TRUE,  width=4, height=4>>=
# pdf( "vulcano_fly.pdf", width=4, height=4 )

print(xyplot( -log10( pval ) ~ log2FoldChange, 
   resFly, 
   pch=20, cex=.2,
   col=ifelse( resFly$padj<.1, "#FF000050", "#00000050" ),
   axis = function( side, ... ) {
      if( side=="bottom") {
         panel.axis( side, outside=TRUE, at=seq(-14,14,by=1), labels=FALSE )
         panel.axis( side, outside=TRUE, at=seq(-10,10,by=5), labels=TRUE )
      }
      if( side=="left") {
         panel.axis( side, outside=TRUE, at=seq(0,25,by=1), labels=FALSE )
         panel.axis( side, outside=TRUE, at=seq(0,25,by=5), 
            labels = do.call( expression, 
	       lapply( seq(0,25,by=5), function(a) 
	          substitute( 10^-b, list(b=a) ) ) ) )
      } },
   xlab = "log2 fold change", ylab = "p value",
   scales = list( 
      x = list( limits=c( -6, 6 ) ),
      y = list( limits=c( 0, 25 ) ) ) ))

# dev.off()
@

\subsection{Hit density}

As edgeR removes rows with zero counts, we should only look at those which are present
in edgeR's result table:

<<>>=
goodRows <- resFly$id %in% rownames(edgerResFlyDT$table)
@

We calculate the densities of hits as follows:

<<>>=
densAll   <- density( log10(resFly$baseMean)[ goodRows ] )
densHits  <- density( log10(resFly$baseMean)[ goodRows & resFly$padj < .1 ] )
densedgDT <- density( log10(resFly$baseMean)[ goodRows ][ edgerResFlyDTPadj < .1 ] )
densedgDC <- density( log10(resFly$baseMean)[ goodRows ][ edgerResFlyDCPadj < .1 ] )
densedgNT <- density( log10(resFly$baseMean)[ goodRows ][ edgerResFlyNTPadj < .1 ] )
densedgNC <- density( log10(resFly$baseMean)[ goodRows ][ edgerResFlyNCPadj < .1 ] )
@

We plot these for densities as follows:

<<hits_dens_fly_all, fig=TRUE,  width=6, height=4>>=
# pdf( "hits_dens_fly_all.pdf", width=6, height=4 )

plot( densAll$x, densAll$y / 7, type="l", ylim=c(0,.05),
   xlab="log10 mean", ylab="density", xaxs="i", yaxs="i", col="gray" )
lines( densHits$x, densHits$y * densHits$n / densAll$n, 
   col="red", lwd=2 )
lines( densedgDT$x, densedgDT$y * densedgDT$n / densAll$n, 
   col="blue", lwd=2, lty="dashed" )
lines( densedgDC$x, densedgDC$y * densedgDC$n / densAll$n, 
   col="lightblue", lwd=2, lty="dashed" )
lines( densedgNT$x, densedgNT$y * densedgNT$n / densAll$n, 
   col="blue", lwd=2, lty="solid" )
lines( densedgNC$x, densedgNC$y * densedgNC$n / densAll$n, 
   col="lightblue", lwd=2, lty="solid" )
text( densAll$x[310]+.2, densAll$y[310]/7, "x7", col="gray" )

# dev.off()
@

For this data set, there is hardly a noticeably difference between the two ways of 
specifying library size. Hence, we omitted the dashed lines from the figure in the paper:

<<hits_dens_fly, fig=TRUE,  width=6, height=4>>=
# pdf( "hits_dens_fly.pdf", width=6, height=4 )

plot( densAll$x, densAll$y / 7, type="l", ylim=c(0,.05),
   xlab="log10 mean", ylab="density", xaxs="i", yaxs="i", col="gray" )
lines( densHits$x, densHits$y * densHits$n / densAll$n, 
   col="red", lwd=2 )
lines( densedgNT$x, densedgNT$y * densedgNT$n / densAll$n, 
   col="blue", lwd=2, lty="solid" )
lines( densedgNC$x, densedgNC$y * densedgNC$n / densAll$n, 
   col="lightblue", lwd=2, lty="solid" )
text( densAll$x[310]+.2, densAll$y[310]/7, "x7", col="gray" )

# dev.off()
@

The effect of the tag-wise dispersion estimate (i.e., of moderated testing) is also hardly
appreciable. This is fully expected because of the low number of replicates.
\medskip

The paper gives numbers of hits, obtained as follows:

Percentage of genes with DESeq $p$ value below 5\%:

<<>>=
length(which( resFly$pval < .05 )) / nrow( resFly )
@

Number of hits at 10\% FDR, found by DESeq:

<<>>=
length( which( resFly$padj < .1 ) )
@

The number of genes judged significant by edgeR is a bit higher and not very sensitive
to the choice of library size specification and dispersion mode:

<<>>=
c( DT = length( which( edgerResFlyDTPadj < .1 ) ),
   DC = length( which( edgerResFlyDCPadj < .1 ) ),
   NT = length( which( edgerResFlyNTPadj < .1 ) ),
   NC = length( which( edgerResFlyNCPadj < .1 ) ) )
@

The numbers of hits shared between DESeq and edgeR are as follows:

<<>>=
list( 
  table( DESeq = resFly$padj[goodRows] < .1, edgeR_DT = edgerResFlyDTPadj < .1 ),
  table( DESeq = resFly$padj[goodRows] < .1, edgeR_DC = edgerResFlyDCPadj < .1 ),
  table( DESeq = resFly$padj[goodRows] < .1, edgeR_NT = edgerResFlyNTPadj < .1 ),
  table( DESeq = resFly$padj[goodRows] < .1, edgeR_NC = edgerResFlyDCPadj < .1 ) )
@

\subsection{Same-vs-same comparison}

For the same-vs-same comparison of the two ``A'' samples, we first remove those
genes which have zero counts in both samples and make a new CountDataSet, then run
the standard analysis:\footnote{With a DESeq version prior to 1.1.6, use 
``\texttt{pool=TRUE}'' rather than ``\texttt{method="blind"}''}

<<>>=
a <- counts(cdsFly)[,conditions(cdsFly)=="A"]
cdsA <- newCountDataSet( a[rowSums(a)>0,], c( "A1", "A2" ) )
cdsA <- estimateSizeFactors( cdsA )
cdsA <- estimateVarianceFunctions( cdsA, method="blind" )
resA <- nbinomTest( cdsA, "A1", "A2" )
@

We do the same with edgeR. (As edgeR does not have an option 
for ``blind'' variance estimation as DESeq does, we tell it first 
that both samples are replicates, then estimate the dispersion,
and then change the group labels before the actual test:

<<>>=
dglA <- DGEList( counts(cdsA), sizeFactors( cdsA ) * 1e7, 
   c( "A", "A" ) )
dglA <- estimateCommonDisp( dglA )
dglA <- estimateTagwiseDisp( dglA )
dglA$samples$group = c( "A1", "A2" )
eResA <- exactTest( dglA, common.disp=FALSE  ) 
@

Finally, we assume a Poisson model and fit a GLM:

<<>>=
pvalPois <- 
   apply( counts(cdsA), 1, function(cl) 
      anova( 
         glm( 
	    counts ~ sample, 
	    data.frame( counts=cl, sample=colnames(counts(cdsA)) ), 
            family=poisson ), 
	 test="Chisq" )$`P(>|Chi|)`[2] )
@

This code generates the ECDF curves (Fig.\ \ref{repl_pval}a):

<<repl_pval_fly_A_ecdf, fig=TRUE,  width=6, height=6>>=
# pdf( file="repl_pval_fly_A_ecdf.pdf", width=6, height=6)

print(ecdfplot( ~ data | which, 
   make.groups( 
      `DESeq, below 100` = resA$pval[ resA$baseMean < 100 ],
      `DESeq, above 100` = resA$pval[ resA$baseMean > 100 ],   
      `DESeq, all` = resA$pval,
      `edgeR, below 100` = eResA$table$p.value[ 
          resA$baseMean < 100 ],
      `edgeR, above 100` = eResA$table$p.value[ 
          resA$baseMean > 100 ],
      `edgeR, all` = eResA$table$p.value,
      `Poisson, below 100` = pvalPois[ resA$baseMean < 100 ],
      `Poisson, above 100` = pvalPois[ resA$baseMean > 100 ],	
      `Poisson, all` = pvalPois ),
   layout=c(3,3), as.table=TRUE, xlab = "p value",
   panel = function( ...) {
      lsegments( 0, 0, 1, 1, col="darkgray" )
      panel.ecdfplot( ... ) },
   scales = list( tick.number = 4, cex=1 ) ))

# dev.off()
@

The zoom-in (Fig.\ \ref{repl_pval}b):

<<repl_pval_fly_A_ecdf_zoom, fig=TRUE,  width=6, height=6>>=
# pdf( file="repl_pval_fly_A_ecdf_zoom.pdf", width=6, height=6)

print(ecdfplot( ~ data | which, 
   make.groups( 
      `DESeq, below 100` = resA$pval[ resA$baseMean < 100 ],
      `DESeq, above 100` = resA$pval[ resA$baseMean > 100 ],   
      `DESeq, all` = resA$pval,
      `edgeR, below 100` = eResA$table$p.value[ 
          resA$baseMean < 100 ],
      `edgeR, above 100` = eResA$table$p.value[ 
          resA$baseMean > 100 ],
      `edgeR, all` = eResA$table$p.value,
      `Poisson, below 100` = pvalPois[ resA$baseMean < 100 ],
      `Poisson, above 100` = pvalPois[ resA$baseMean > 100 ],	
      `Poisson, all` = pvalPois ),
   layout=c(3,3), as.table=TRUE, xlab = "p value",
   panel = function( ...) {
      lsegments( 0, 0, 1, 1, col="darkgray" )
      panel.ecdfplot( ... ) },
   xlim=c( -.005, .1 ), ylim=c( -.005, .1 ),
   scales = list( tick.number = 4, cex=1 ) ))

# dev.off()
@

\section{Analysis of neural stem cell data}

We performed the same analysis as described in the previous section for
the neural stem cell set. For completeness, we give here the full code to
produce all the figures relating to this data set, but with only few comments,
as the code is very similar to the already described code.

\subsection{Analysis with DESeq and edgeR}

As stated above and in the paper, we omit the first sample. Apart from that, 
the analysis is performed as before:

<<>>=
cdsNS <- newCountDataSet( countsTableNS[,-1], condsNS[-1] )
cdsNS <- estimateSizeFactors( cdsNS )
cdsNS <- estimateVarianceFunctions( cdsNS )
resNS <- nbinomTest( cdsNS, "GNS", "NS" )
@

For edgeR, we also have to explicitly remove the ``GNSL'' column. Apart from that, as before:

<<>>=
colGNSL <- which( conditions(cdsNS) == "GNSL" )

dglNS.D <- DGEList( 
   counts = counts(cdsNS)[,-colGNSL], 
   group = factor( conditions(cdsNS)[-colGNSL] ), 
   lib.size = 1e7*sizeFactors(cdsNS)[-colGNSL] )
dglNS.D <- estimateCommonDisp( dglNS.D )
dglNS.D <- estimateTagwiseDisp( dglNS.D )

edgerResNS.DT <- exactTest( dglNS.D, common.disp=FALSE )
edgerResNS.DTPadj <- p.adjust( edgerResNS.DT$table$p.value, method="BH" )

edgerResNS.DC <- exactTest( dglNS.D, common.disp=TRUE )
edgerResNS.DCPadj <- p.adjust( edgerResNS.DC$table$p.value, method="BH" )

libSizesNS <- c( G144=7604834, G166=13625570, G179=12291910,
   CB541=12872125, CB660=10502656 )

dglNS.N <- DGEList( 
   counts = counts(cdsNS)[,-colGNSL], 
   group = factor( conditions(cdsNS)[-colGNSL] ), 
   lib.size = libSizesNS[-colGNSL] )
dglNS.N <- estimateCommonDisp( dglNS.N )
dglNS.N <- estimateTagwiseDisp( dglNS.N )

edgerResNS.NT <- exactTest( dglNS.N, common.disp=FALSE )
edgerResNS.NTPadj <- p.adjust( edgerResNS.NT$table$p.value, method="BH" )

edgerResNS.NC <- exactTest( dglNS.N, common.disp=TRUE )
edgerResNS.NCPadj <- p.adjust( edgerResNS.NC$table$p.value, method="BH" )
@

DESeq finds this number of significant hits at 10\% FDR:

<<>>=
length( which( resNS$padj < .1 ) )
@

For edgeR, the numbers are:

<<>>=
c( DT = length( which( edgerResNS.DTPadj < .1 ) ),
   DC = length( which( edgerResNS.DCPadj < .1 ) ),
   NT = length( which( edgerResNS.NTPadj < .1 ) ),
   NC = length( which( edgerResNS.NCPadj < .1 ) ) )
@

\subsection{Figures}

Figure \ref{cv2_Paul} has been made with this code:

<<cv2_mean, fig=TRUE,  width=4, height=4>>=
# pdf( "cv2_mean.pdf", width=4, height=4 )

diagForGNS <- varianceFitDiagnostics( cdsNS, "GNS" )
alpha <- dglNS.D$common.dispersion

print(xyplot(    
   I( baseVar/baseMean^2 ) ~ baseMean, 
   diagForGNS[ diagForGNS$baseMean>0,], 
   panel = function( ... ) { 
      panel.hexbinplot( ... ) 
      llines( 
         xg, 
         adjustScvForBias( 
           baseVarFunc(cdsNS,"GNS")( 10^xg ) / (10^xg)^2, 
           attr( rawVarFunc(cdsNS,"GNS")(NA), "size" ) ),
         col="orange", lwd=2 )
      llines( xg, ( 10^xg + alpha * (10^xg)^2 ) / (10^xg)^2,
         col="orange", lwd=2, lty="dashed" )
      for( sf in sizeFactors( cdsNS )[ conditions(cdsNS) == "GNS" ] )
         llines( xg, 1 / (10^xg * sf), col="purple", lwd=1.5 ) },
   trans = function(x) x^(1/4), 
   inv = function(x) x^4,
   xbins = 80,
   scales = list( x = list( log=TRUE, axs="i" ),
      y = list( log=FALSE, axs="i", tick.number=8 ) ),
   xlab = "mean", ylab="squared coefficient of variation"
))

# dev.off()
@

For Fig.\ \ref{DE_Paul}, this code was used:

<<DE,fig=TRUE,  width=4, height=4>>=
# pdf( "DE.pdf", width=4, height=4 )

print(xyplot( 
   log2FoldChange ~ I( baseMean ),
   resNS, 
   pch=16, cex=.3, 
   col=ifelse( resNS$padj < .1, "#FF000040", "#00000040" ),
   axis = function( side, ... ) {
      if( side=="left") {
         panel.axis( side, outside=TRUE, at=seq(-14,14,by=1), labels=FALSE )
         panel.axis( side, outside=TRUE, at=seq(-10,10,by=5), labels=TRUE )
      } 
      if( side=="bottom") {
         panel.axis( side, outside=TRUE, at=seq(-2,10,by=1), 
            labels=paste( "10^", seq(-2,10,by=1), sep="" ) )
      } },
   xlab = "mean", ylab = "log2 fold change",
   scales = list( x=list(log=TRUE), y=list(log=FALSE) ) ))

# dev.off()
@

Figure \ref{hits_dens_Paul}, with the density estimates, was done with:

<<hits_dens,fig=TRUE,  width=6, height=4 >>=
# pdf( "hits_dens.pdf", width=6, height=4 )

goodRowsNS <- resNS$id %in% rownames(edgerResNS.DT$table)

densAll.NS   <- density( log10(resNS$baseMean)[ goodRowsNS ] )
densHits.NS  <- density( log10(resNS$baseMean)[ goodRowsNS & resNS$padj < .1 ] )
densedgDT.NS <- density( log10(resNS$baseMean)[ goodRowsNS ][ edgerResNS.DTPadj < .1 ] )
densedgDC.NS <- density( log10(resNS$baseMean)[ goodRowsNS ][ edgerResNS.DCPadj < .1 ] )
densedgNT.NS <- density( log10(resNS$baseMean)[ goodRowsNS ][ edgerResNS.NTPadj < .1 ] )
densedgNC.NS <- density( log10(resNS$baseMean)[ goodRowsNS ][ edgerResNS.NCPadj < .1 ] )

plot( densAll.NS$x, densAll.NS$y / 7, type="l", ylim=c(0,.07),
   xlab="log10 mean", ylab="density", xaxs="i", yaxs="i", 
   col="gray" )
lines( densHits.NS$x, densHits.NS$y * densHits.NS$n / densAll.NS$n, 
   col="red", lwd=2 )
lines( densedgDT.NS$x, densedgDT.NS$y * densedgDT.NS$n / densAll.NS$n, 
   col="blue", lwd=2, lty="dashed" )
lines( densedgDC.NS$x, densedgDC.NS$y * densedgDC.NS$n / densAll.NS$n, 
   col="lightblue", lwd=2, lty="dashed" )
lines( densedgNT.NS$x, densedgNT.NS$y * densedgNT.NS$n / densAll.NS$n, 
   col="blue", lwd=2, lty="solid" )
lines( densedgNC.NS$x, densedgNC.NS$y * densedgNC.NS$n / densAll.NS$n, 
   col="lightblue", lwd=2, lty="solid" )
text( densAll.NS$x[310]+.2, densAll.NS$y[310]/7, "x7", col="gray" )

# dev.off()
@

Finally, Figure \ref{repl_pval_Paul} with the type-I error control histograms

<<repl_pval_A_ecdf, fig=TRUE,  width=8, height=6 >>=
# pdf( file="repl_pval_A_ecdf.pdf", width=6, height=4.5)

cdsGNS <- newCountDataSet( counts(cdsNS)[,1:2], c( "GNS_A", "GNS_B" ) )
cdsGNS <- estimateSizeFactors( cdsGNS )
cdsGNS <- estimateVarianceFunctions( cdsGNS, method="blind" )
resGNS <- nbinomTest( cdsGNS, "GNS_A", "GNS_B" )
medGNS <- median( getBaseMeansAndVariances( counts(cdsGNS), 
             sizeFactors(cdsGNS) )$baseMean )

dglGNS <- DGEList( counts(cdsGNS), sizeFactors( cdsGNS ) * 1e7, 
   c( "GNS", "GNS" ) )
dglGNS <- estimateCommonDisp( dglGNS )
dglGNS <- estimateTagwiseDisp( dglGNS )
dglGNS$samples$group = c( "GNS_1", "GNS_2" )
eResGNS <- exactTest( dglGNS, common.disp=FALSE ) 
eMedGNS <- median( rowMeans(dglGNS$pseudo.alt) )

print(ecdfplot( ~ data | which, 
   make.groups( 
      `DESeq lower half` = resGNS$pval[ resGNS$baseMean < medGNS ],
      `DESeq upper half` = resGNS$pval[ resGNS$baseMean > medGNS ],   
      `DESeq all` = resGNS$pval,
      `edgeR lower half` = eResGNS$table$p.value[ 
          rowMeans(dglGNS$pseudo.alt) < eMedGNS ],
      `edgeR upper half` = eResGNS$table$p.value[ 
          rowMeans(dglGNS$pseudo.alt) > eMedGNS ],
      `edgeR all` = eResGNS$table$p.value ),
   layout=c(3,2), xlab = "p value", as.table=TRUE, 
   panel = function( ...) {
      lsegments( 0, 0, 1, 1, col="darkgray" )
      panel.ecdfplot( ... ) },
   scales = list( tick.number=4, cex=1 ) ))

# dev.off()   
@

\section{Working without replicates}

To see how much hits we lose without replicates, we reduce the fly data set
to only one sample per condition

<<>>=
cdsFly2 <- cdsFly[ ,c( "A1", "B1" ) ]
cdsFly2 <- estimateVarianceFunctions( cdsFly2, method="blind" )
resFly2 <- nbinomTest( cdsFly2, "A", "B" )
@

Here we see a comparison of the hits with the full data set vs the reduced data set:

<<>>=
addmargins( table( full = resFly$padj < .1, reduced = resFly2$padj < .1 ) )
@

The same for the neural stem cell data set:

<<>>=
cdsNS2 <- cdsNS[ ,c( "G144", "CB541" ) ]
cdsNS2 <- estimateVarianceFunctions( cdsNS2, method="blind" )
resNS2 <- nbinomTest( cdsNS2, "GNS", "NS" )
addmargins( table( full = resNS$padj < .1, reduced = resNS2$padj < .1 ) )
@

\section{VST}

To demonstrate the VST, we first make a new CountDataSet from the neural cell data,
this time with all samples. To make clear that the dendrogram we get in the end was made
without using class labels, we do not pass condition labels.

<<heatmap,fig=TRUE,width=6,height=6>>=
# pdf( "heatmap.pdf" )

cdsNS.all <- newCountDataSet( countsTableNS, rep( "dummy", ncol(countsTableNS) ) )
cdsNS.all <- estimateSizeFactors( cdsNS.all )

vsdNS <- getVarianceStabilizedData( cdsNS.all )
distMatrix <- as.matrix(dist(t( vsdNS )))
rownames( distMatrix ) <- c( "GNS (*)", "GNS (*)", "GNS", "GNS (L)", "NS", "NS" )

heatmap.2( distMatrix, density.info="none", trace="none", revC=TRUE,
   col=colorRampPalette(c("darkblue","yellow"))(100) )

# dev.off()
@

\section{Yeast RNA-Seq data}

For the yeast data, we have replicates, but no conditions to compare. DESeq's CountDataSet
class is not designed to deal with this situation, and we fall back to the set of DESeq
functions provided to directly work with matrix data:

<<>>=
sfYeast <- estimateSizeFactorsForMatrix( countsYeast )

# A function to get a base variance function:
get_bvf <- function( cl ) {
   rvf <- estimateVarianceFunctionForMatrix( countsYeast[,cl], sfYeast[cl] )
   xim <- sum(1/sfYeast[cl])/length(cl)
   function( q ) rvf( q ) + xim * q
}

# The variance estimated between the replicates:
bvf_dT_tech <- get_bvf( c("dT_ori","dT_tech") )
bvf_dT_bio  <- get_bvf( c("dT_ori","dT_bio" ) )
bvf_RH_tech <- get_bvf( c("RH_ori","RH_tech") )
bvf_RH_bio  <- get_bvf( c("RH_ori","RH_bio" ) )

# The common-scale means
meansYeast <- getBaseMeansAndVariances( countsYeast, sfYeast )$baseMean
@

The plot of Fig. \ref{noiseNaga} is produced with

<<naga, fig=TRUE,  width=4, height=4>>=
#pdf("naga.pdf", width=4, height=4)

layout( cbind(1:2), heights = c( 1, 3 ) )
par( cex=.75 )

par(mai=c(0.2, 1, 0.5, 0.5))
plot( NULL, main="", xaxs="i", yaxs="i", ylab="density", 
   xlim=c( -.3, 4 ), ylim=c( 0, 1200 ), xaxt="n" )
hist( log10( meansYeast ), breaks=30, add=TRUE )

par(mai=c(1, 1, 0.01, 0.5))
plot( NULL, log="x", ylim=c(0, .65), xaxs="i", yaxs="i",
   xlab="mean", ylab="squared coefficient of variation",
   xlim=c( 10^-.3, 10^4 ) )
xg <- 10^seq( 0, 4, length.out=1000 ) 
rect( xg[-1000], 1/(xg[-1000]*max(sfYeast)), 
   xg[-1], 1/(xg[-1000]*min(sfYeast)), col="#FFD0FF", lty="blank" )
lines( xg, bvf_dT_bio(xg)/xg^2, col="brown" )
lines( xg, bvf_dT_tech(xg)/xg^2, col="darkblue" )
lines( xg, bvf_RH_bio(xg)/xg^2, col="brown", lty="dashed" )
lines( xg, bvf_RH_tech(xg)/xg^2, col="darkblue", lty="dashed" )
lines( xg, 1/xg, col="#FFD0FF" )

# dev.off()
@

\section{ChIP-Seq}

We first look at a heat map for the Pol-II count data:

<<PolII_heatmap, fig=TRUE>>=
cdsPolII.1 <- newCountDataSet( countsPolII, rep( "dummy", 12 ) )
cdsPolII.1 <- estimateSizeFactors( cdsPolII.1 )
heatmap( as.matrix( dist( t( getVarianceStabilizedData( cdsPolII.1 ) ) ) ),
   cexCol=.7, cexRow=.7, symm=TRUE,
   col=colorRampPalette(c("darkblue","yellow"))(100) )
@

The heatmap shows clear evidence of batch effects for GM12878: Replicates 1 to 4
form one batch and 6 and 7 another one. For GM12891, replicates 1 to 3 form
one batch and 4 to 6 another one.

For Fig.\ \ref{figJan}, we have to chose two replicates for each of the groups A1, A2,
B1, and B2. In order to be balanced with respect to batches, we chose the
replicates such that each batch is present in each group, as follows 
(Unused samples get denoted with an ``x''):

<<>>=
condsPolII <- c( 
   GM12878_1 = "A1", GM12878_2 = "A2", GM12878_3 = "x", 
   GM12878_4 = "x",  GM12878_6 = "A1", GM12878_7 = "A2",
   GM12891_1 = "B1", GM12891_2 = "B2", GM12891_3 = "x",
   GM12891_4 = "B1", GM12891_5 = "B2", GM12891_6 = "x" )

cdsPolII.2 <- newCountDataSet( countsPolII, condsPolII )
cdsPolII.2 <- estimateSizeFactors( cdsPolII.2 )
cdsPolII.2 <- estimateVarianceFunctions( cdsPolII.2 )
resPolII.A1vsA2 <- nbinomTest( cdsPolII.2, "A1", "A2" )
resPolII.B1vsB2 <- nbinomTest( cdsPolII.2, "B1", "B2" )
resPolII.A1vsB1 <- nbinomTest( cdsPolII.2, "A1", "B1" )
@

For the comparison of 4 replicates against 4, we use only the
first letter:

<<>>=
cdsPolII.3 <- newCountDataSet( countsPolII, substr( condsPolII, 1, 1 ) )
conditions( cdsPolII.3 )
@

<<>>=
cdsPolII.3 <- estimateSizeFactors( cdsPolII.3 )
cdsPolII.3 <- estimateVarianceFunctions( cdsPolII.3 )
resPolII.AvsB <- nbinomTest( cdsPolII.3, "A", "B" )
@

Number of hits at 10\% FDR:

<<>>=
table( resPolII.A1vsA2$padj < .1 )
table( resPolII.B1vsB2$padj < .1 )
table( resPolII.A1vsB1$padj < .1 )
table( resPolII.AvsB$padj < .1 )

@

The same with a Poisson GLM:

<<>>=
do_glm <- function( cds, cols )
   apply( counts(cds)[, cols], 1, function(r)
      anova( 
         glm( r ~ conditions(cds)[cols], family=poisson ), 
         test="Chisq" )$`P(>|Chi|)`[2] )

pvalPois.A1vsA2 <- do_glm( cdsPolII.2, which( conditions(cdsPolII.2) %in% c("A1","A2") ) )
pvalPois.B1vsB2 <- do_glm( cdsPolII.2, which( conditions(cdsPolII.2) %in% c("B1","B2") ) )
pvalPois.A1vsB1 <- do_glm( cdsPolII.2, which( conditions(cdsPolII.2) %in% c("A1","B1") ) )
pvalPois.AvsB   <- do_glm( cdsPolII.2, which( conditions(cdsPolII.3) %in% c("A","B") ) )
@

The ECDF plot of Fig.\ \ref{figJan} is now made with:

<<ecdf_Jan_pois, fig=TRUE,  width = 6, height = 4.5>>=

# pdf( "ecdf_Jan_pois.pdf", width = 6, height = 4.5 )
print(ecdfplot( ~ data | which, 
   make.groups( 
      `D: A1 vs A2` = resPolII.A1vsA2$pval,
      `D: B1 vs B2` = resPolII.B1vsB2$pval,
      `D: A1 vs B1` = resPolII.A1vsB1$pval,
      `D: A vs B`   = resPolII.AvsB$pval,
      `P: A1 vs A2` = pvalPois.A1vsA2,
      `P: B1 vs B2` = pvalPois.B1vsB2,
      `P: A1 vs B1` = pvalPois.A1vsB1,
      `P: A vs B`   = pvalPois.AvsB ),
   layout=c(4,2), xlab = "p value", as.table=TRUE,  
   panel = function( ...) {
      lsegments( 0, 0, 1, 1, col="darkgray" )
      panel.ecdfplot( ... ) },
   scales = list( tick.number=4, cex=1 ) ))

# dev.off()      
@

\section{Simulation}

The details of simulations in Supplementary Note G are given here.

\subsection{Constant dispersion}

The count data was drawn as follows

<<>>=

# Number of genes to simulate
ngenes <- 20000

# True mean expression value across conditions drawn from an
# exponential distribution
true_q0middle <- rexp( ngenes, 1/250 )

# These values might be typical size factors as could be found
# in a real experiment
sf <- c( .5, 1.7, 1.4, .9 )

# Two replicates for each condition:
conds <- c( "A", "A", "B", "B" )

# Mark around 30% of the genes as differentially expressed:
true_isDE <- runif( ngenes ) < .3

# For the DE genes, draw a log fold change from a normal
true_lfc <- rnorm( ngenes, sd=.7 ) * true_isDE

# Go up and down from the true_q0middle according to these log
# fold changes to get the values q for the conditions
true_q0 <- data.frame(
   A = true_q0middle * 2^(  true_lfc/2 ),
   B = true_q0middle * 2^( -true_lfc/2 ) )

# For the variance-mean relation, let's assume that the shape 
# assumed by edgeR is correct, i.e., variance = mean + alpha * mean^2
# and use this alpha:   
alpha <- .015

# Parametrise the NB to mean and variance:
rnbinomMV <- function( n, mu, v ) 
  rnbinom( n, prob = mu/v, size = mu^2/(v-mu) )

# Now draw counts
countsTable <- 
   sapply( 1:4, function(j)
      sapply( true_q0[[ conds[j] ]], function(q)
         rnbinomMV( 1, sf[j]*q, sf[j]*q + sf[j]^2 * q^2 * alpha ) ) )

# The good rows:
nonzero <- rowSums(countsTable)>0 
@

Now, we analyse this count table:

<<>>=
# A standard analysis with DESeq:
cds <- newCountDataSet( countsTable, conds )
cds <- estimateSizeFactors( cds )
cds <- estimateVarianceFunctions( cds )
res <- nbinomTest( cds, "A", "B" )

# The same with edgeR. (edgeR gets the advantage of knowing the true
# library sizes)
dgl <- DGEList( counts=countsTable, group=conds, lib.size=sf*1e7 )
dgl <- estimateCommonDisp(dgl)
edgerRes <- exactTest( dgl )
edgerResPadj <- p.adjust( edgerRes$table$p.value, method="BH" )
@

Compare the results.

Type-I error control: Which percentage of genes without true DE have a p value below 5\%?

For DESeq:
<<>>=
length(which( res$pval[nonzero]<.05 & !true_isDE[nonzero] )) / length(nonzero)
@

For edgeR:
<<>>=
length(which( edgerRes$table$p.value<.05 & !true_isDE[nonzero] )) / length(nonzero)
@

Detection power: Which percentage of the true positives is detected at 10\% FDR?

For DESeq:
<<>>=
length(which( res$padj[nonzero]<.1 & true_isDE[nonzero] )) / 
   length(which( true_isDE[nonzero] ))
@

For edgeR:
<<>>=
length(which( edgerResPadj<.1 & true_isDE[nonzero] )) / 
   length(which( true_isDE[nonzero] ))
@

FDR control:

For DESeq:
<<>>=
length(which( res$padj[nonzero]<.1 & !true_isDE[nonzero] )) / 
   length(which( res$padj[nonzero]<.1 ))
@

For edgeR:
<<>>=
length(which( edgerResPadj<.1 & !true_isDE[nonzero] )) / 
   length(which( edgerResPadj<.1 ))
@

For Fig.\ \ref{pvalcomp}a, we used this code:

<<pval_nonDE_B, fig=TRUE, width=4, height=3>>=
# pdf( "pval_DE_B.pdf", width=4, height=4 )

print(ecdfplot( ~ data | which, 
   make.groups( 
      `DESeq` = res$pval[nonzero][ !true_isDE[nonzero] ],
      `edgeR` = edgerRes$table$p.value[ !true_isDE[nonzero] ] ),
   layout=c(2,1), as.table=TRUE, xlab = "p value",
   panel = function( ...) {
      lsegments( 0, 0, 1, 1, col="darkgray" )
      panel.ecdfplot( ... ) },
   xlim=c( -.005, .1 ), ylim=c( -.005, .1 ),
   scales = list( tick.number = 4, cex=1 ) ))

# dev.off()   
@

and for Fig.\ \ref{pvalcomp}b

<<pval_DE, fig=TRUE, width=4, height=4>>=
# pdf( "pval_DE.pdf", width=4, height=4 )

print(xyplot(  
   res$pval[true_isDE & nonzero] ~ edgerRes$table$p.value[true_isDE[nonzero]],
   panel = function( ... ) {
      panel.hexbinplot( ... )
      llines( c(-35, 0), c(-35, 0), col="#30FF5090" ) },
   trans = function(x) x^(1/4), 
   inv = function(x) x^4,
   xbins = 150,
   scales = list(
      x = list(log=TRUE),
      y = list(log=TRUE), apect=1 ),
   xlim = c( 1e-15, 10 ),
   ylim = c( 1e-15, 10 ),
   aspect = 1,
   xlab="p value from edgeR", ylab="p value from DESeq", asp=1
))

# dev.off()
@

\subsection{Non-constant dispersion}

We prepare data as before:

<<>>=
ngenes <- 20000
true_q0middle <- rexp( ngenes, 1/250 )
sf <- c( .5, 1.7, 1.4, .9 )
conds <- c( "A", "A", "B", "B" )
true_isDE <- runif( ngenes ) < .3
true_lfc <- rnorm( ngenes, sd=.7 ) * true_isDE
true_q0 <- data.frame(
   A = true_q0middle * 2^(  true_lfc/2 ),
   B = true_q0middle * 2^( -true_lfc/2 ) )
@

Now, we define a mean-dependent raw SCV value and use it in
drawing the count data.

<<>>=
alpha <- function(mu) .01 + 9/(mu+100)

countsTable <- 
   sapply( 1:4, function(j)
      sapply( true_q0[[ conds[j] ]], function(q)
         rnbinomMV( 1, sf[j]*q, sf[j]*q + sf[j]^2 * q^2 * alpha( q ) ) ) )
@

Then, continue as before.
<<>>=
nonzero <- rowSums(countsTable)>0

cds <- newCountDataSet( countsTable, conds )
cds <- estimateSizeFactors( cds )
cds <- estimateVarianceFunctions( cds )
res <- nbinomTest( cds, "A", "B" )

dgl <- DGEList( counts=countsTable, group=conds, lib.size=sf*1e7 )
dgl <- estimateCommonDisp(dgl)
dgl <- estimateTagwiseDisp(dgl)
edgerRes <- exactTest( dgl, common.disp=FALSE )
edgerResPadj <- p.adjust( edgerRes$table$p.value, method="BH" )
@

Here is a plot showing the simulated (purple) and the 
fitted (solid red and green) raw SCV curves:

<<scvSim2,fig=TRUE>>=
scvPlot( cds )
xg <- 10^seq( -2, 4, length.out=1000 )
lines( xg, alpha( xg ), col="purple" )
@

Finally, the ECDF plot
<<pvalcompsim2, fig=TRUE,  width=8, height=6 >>=
# pdf( file="pvalcompsim2", width=8, height=6 )

baseMeans <- t(t(countsTable)/sf)
lowerHalf <- baseMeans < median( baseMeans )

print(ecdfplot( ~ data | which, 
   make.groups( 
      `DESeq, below median` = res$pval[nonzero][ 
         !true_isDE[nonzero] & lowerHalf[nonzero] ],
      `DESeq, above median` = res$pval[nonzero][ 
         !true_isDE[nonzero] & !lowerHalf[nonzero] ],   
      `DESeq, all`          = res$pval[nonzero][ 
         !true_isDE[nonzero] ],
      `edgeR, below median` = edgerRes$table$p.value[ 
         !true_isDE[nonzero] & lowerHalf[nonzero] ],
      `edgeR, above median` = edgerRes$table$p.value[ 
         !true_isDE[nonzero] & !lowerHalf[nonzero] ],
      `edgeR, all` = edgerRes$table$p.value[ 
         !true_isDE[nonzero] ] ),
   layout=c(3,2), as.table=TRUE, xlab = "p value",
   panel = function( ...) {
      lsegments( 0, 0, 1, 1, col="darkgray" )
      panel.ecdfplot( ... ) },
   xlim=c( -.005, .1 ), ylim=c( -.005, .1 ),
   scales = list( tick.number = 4, cex=1 ) ))

# dev.off()   
@
\section{Other figures}

\subsection{Bias-adjustment function}

The code for Fig.\ \ref{biasAdjust} is:
<<biasAdjust, fig=TRUE, width=4, height=4>>=
# pdf("biasAdjust.pdf", width=4, height=4)

plot( NULL, xlim=c(0,2), ylim=c(0,2), 
   xlab = expression( ( hat(sigma)^2 - hat(mu) ) / hat(mu)^2 ), 
   ylab = expression( gamma ) )
xg <- seq( 0, 2, length.out=1000 )
for( i in 2:15 ) 
   lines( xg, adjustScvForBias( xg, i ), 
      col=colorRampPalette(c("red","black"))(15)[i] )

# dev.off()      
@

See the code of the functions \emph{DESeq::adjustScvForBias} and 
\emph{DESeq:::prepareScvBiasCorrectionFits} for details on
how the variance adjustment is performed by means of fitting a
LOESS regression through results from numerical simulations.

\section{Figure of $p(a,b)$}

The code for Fig.\ \ref{figp1}:

<<p1, fig=TRUE,  width=4, height=4>>=
# pdf("p1.pdf", width=4, height=4)

dnbinomMV <- function( x, mu, v ) 
  dnbinom( x, prob = mu/v, size = mu^2/(v-mu) )

plot( 0:10000, 
   sapply( 0:10000, function(k) 
      dnbinomMV( k, 7000, 7000+.1*7000^2 ) * 
         dnbinomMV( 10000-k, 4000, 4000+.1*4000^2 ) ), 
   type='l', xlab="a", ylab="p(a,b)" )
abline( v = 7000 / (7000+4000) * 10000 )

# dev.off()
@

Finally, save all data:

<<>>=
save.image( "suppl_ii.RData" )
@

\end{document}

### 2011-08-04 R/class_and_slots.R

@@ -1,0 +1,196 @@
+setClass( "CountDataSet", 
+   contains = "eSet",
+   representation = representation( 
+      rawVarFuncs = "environment",
+      rawVarFuncTable = "character",
+      multivariateConditions = "logical" ),
+   prototype = prototype( new( "VersionedBiobase",
+      versions = c( classVersion("eSet"), CountDataSet = "1.1.0" ) ) )
+)
+
+newCountDataSet <- function( countData, conditions, sizeFactors=NULL,
+      phenoData = NULL, featureData = NULL )
+{
+   countData <- as.matrix( countData )
+   if( any( round( countData ) != countData ) )
+      stop( "The countData is not integer." )
+   mode( countData ) <- "integer"
+
+   if( is.null( sizeFactors ) )
+      sizeFactors <- rep( NA_real_, ncol(countData) )
+   if( is.null( phenoData ) )
+      phenoData <- annotatedDataFrameFrom( countData, byrow=FALSE )
+   if( is.null( featureData ) ) 
+      featureData <- annotatedDataFrameFrom( countData, byrow=TRUE )
+      
+   phenoData$`sizeFactor` <- sizeFactors
+   varMetadata( phenoData )[ "sizeFactor", "labelDescription" ] <-
+      "size factor (relative estimate of sequencing depth)"
+   
+   if( is( conditions, "matrix" ) )
+      conditions <- as.data.frame( conditions )
+   
+   if( is( conditions, "data.frame" ) || is( conditions, "AnnotatedDataFrame" ) ) {
+      stopifnot( nrow( conditions ) == ncol( countData ) )
+      conditions <- as( conditions, "AnnotatedDataFrame" )
+      dimLabels( conditions ) <- dimLabels( phenoData )
+      rownames( pData(conditions) ) <- rownames( pData(phenoData) )
+         # TODO: What if the rownames were set?
+      phenoData <- combine( phenoData, conditions )
+      multivariateConditions <- TRUE
+      rvft <- c( `_all` = NA_character_ )
+   } else {
+      conditions <- factor( conditions )
+      stopifnot( length( conditions ) == ncol( countData ) )
+      phenoData$`condition` <- factor( conditions )
+      varMetadata( phenoData )[ "condition", "labelDescription" ] <-
+         "experimental condition, treatment or phenotype"
+      multivariateConditions <- FALSE
+      rvft <- rep( NA_character_, length(levels(conditions)) )
+   }
+   
[…]

### 2011-08-04 R/core.R

@@ -1,0 +1,308 @@
+estimateSizeFactorsForMatrix <- function( counts, locfunc = median )
+{
+   loggeomeans <- rowMeans( log(counts) ) 
+   apply( counts, 2, function(cnts) 
+      exp( locfunc( ( log(cnts) - loggeomeans )[ is.finite(loggeomeans) ] ) ) )
+}
+
+
+getBaseMeansAndVariances <- function( counts, sizeFactors ) {
+
+   # Devides the counts by sizeFactors and calculates the estimates for
+   # base means and variances for each gene.
+   
+   data.frame(
+      baseMean = rowMeans( t( t(counts) / sizeFactors ) ),
+      baseVar = rowVars( t( t(counts) / sizeFactors ) ) )
+}   
+
+estimateVarianceFunctionForMatrix <- function( counts, sizeFactors, 
+         locfit_extra_args=list(), lp_extra_args=list() ) {
+
+   stopifnot( ncol( counts ) == length( sizeFactors ) )
+   bmv <- getBaseMeansAndVariances( counts, sizeFactors ) 
+   estimateVarianceFunctionFromBaseMeansAndVariances( bmv$baseMean,
+      bmv$baseVar, sizeFactors, locfit_extra_args, lp_extra_args )
+}      
+   
+modelMatrixToConditionFactor <- function( modelMatrix ) {
+
+   mmconds <- 1:nrow(modelMatrix)
+   for( i in 2:nrow(modelMatrix) )
+      for( j in 1:(i-1) )
+         if( all( modelMatrix[i,] == modelMatrix[j,] ) ) {
+            mmconds[i] = mmconds[j]
+            break }
+   factor( as.integer( factor( mmconds ) ) )
+}
+
+
+getBaseMeansAndPooledVariances <- function( counts, sizeFactors, conditions ) {
+
+   basecounts <- t( t(counts) / sizeFactors )
+   replicated_sample <- conditions %in% names(which(table(conditions)>1))
+   df <- sum(replicated_sample) - length( unique( conditions[ replicated_sample ] ) ) 
+
+   data.frame(
+      baseMean = rowMeans( basecounts ),
+      baseVar =
+	 rowSums( 
+	    sapply( 
+               tapply( 
+        	  ( 1:ncol(counts) )[ replicated_sample ], 
+        	  factor( conditions[ replicated_sample ] ), 
+        	  function(cols) 
[…]

### 2011-08-04 R/deseq.R

@@ -1,0 +1,57 @@
+
+# Read a table
+# newCountDataSet <- function()
+{
+   countData <- countsTable 
+   countData <- as.matrix( countData )
+   if( any( round( countData ) != countData ) )
+      stop( "The countData is not integer." )
+   mode( countData ) <- "integer"
+}
+
+commentfunc2 <- function ()
+{
+library(DESeq)
+countsFile <- "/Users/goshng/Documents/Projects/rnaseq/output/omz/1/bwa/count.txt"
+countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)
+rownames(countsTable) <- countsTable$gene
+countsTable <- countsTable[,-1]
+conds <- c("OMZ175", "OMZHKRR")
+cds <- newCountDataSet(countsTable, conds)
+cds <- estimateSizeFactors(cds)
+cds <- estimateVarianceFunctions(cds,method="blind")
+res <- nbinomTest(cds, "OMZ175", "OMZHKRR")
+}
+
+plotDE <- function( res )
+{
+   plot(res$baseMean, res$log2FoldChange, log="x", pch=20, cex=.5, 
+        col = ifelse( res$padj < .1, "red", "black" ) )
+}
+
+commentfunc <- function ()
+{
+cat ("SCV file: count-de.ps\n")
+postscript ("/Users/goshng/Documents/Projects/rnaseq/output/omz/1/bwa/count-de.ps",  width=10, height=10, horizontal = FALSE, onefile = FALSE,
+paper = "special")
+plotDE(res)
+y <- dev.off()
+
+cat ("SCV file: count-scv.ps\n")
+postscript ("/Users/goshng/Documents/Projects/rnaseq/output/omz/1/bwa/count-scv.ps",  width=10, height=10, horizontal = FALSE, onefile = FALSE,
+paper = "special")
+scvPlot(cds)
+y <- dev.off()
+
+resSig <- res[res$padj < .1,]
+y <- length(countsTable[,1])
+cat ("Total number of genes is ", y, "\n",sep="")
+y <- length(resSig[,1])
+cat ("The number of differentially expressed genes is ", y, "\n",sep="")
+cat ("\nList of differentially expressed genes\n")
+print(resSig[order(resSig$pval),],with=1000)
+cat ("\nThe same table as above but in order of down-regulated first of the differentially expressed genes\n")
+print(resSig[order(resSig$foldChange,-resSig$baseMean),])
[…]

### 2011-08-04 R/methods.R

@@ -1,0 +1,318 @@
+estimateSizeFactors <- function( cds, locfunc=median )
+{
+   stopifnot( is( cds, "CountDataSet" ) )
+   sizeFactors(cds) <- estimateSizeFactorsForMatrix( counts(cds), locfunc )
+   cds
+}
+
+estimateVarianceFunctions <- function( cds, 
+   method = c( "normal", "blind", "pooled" ), pool=NULL, 
+   locfit_extra_args=list(), lp_extra_args=list(), modelFrame = NULL )
+{
+   stopifnot( is( cds, "CountDataSet" ) )   
+   if( any( is.na( sizeFactors(cds) ) ) )
+      stop( "NAs found in size factors. Have you called already 'estimateSizeFactors'?" )
+   
+   if( length(method) != 3 & !is.null( pool ) )
+      stop( "Do not specify both the 'pool' and the 'method' argument." )      
+   if( !is.null( pool ) ) {
+      if( pool == FALSE )
+         method <- "normal"
+      else if( pool == TRUE )
+         method <- "blind"
+      else
+         stop( "Argument 'pool' used incorrectly." )
+      warning( "The 'pool' argument to 'estimatevarianceFunction' is deprecated. Use the 'method' argument instead." )
+   } else
+      method <- match.arg( method )
+   
+   if( cds@multivariateConditions && ! method %in% c( "blind", "pooled" ) )
+      stop( "You have specified multivariate conditions (i.e., passed a data frame with conditions). In this case, you need to specify ethod 'pooled' or 'blind'." )
+   
+   cds@rawVarFuncs <- new.env( hash=TRUE )
+   
+   if( method == "blind" ) {
+      cds@rawVarFuncs[["_blind"]] <-
+         estimateVarianceFunctionForMatrix( counts(cds), 
+	    sizeFactors(cds), locfit_extra_args, lp_extra_args )
+      if( cds@multivariateConditions )
+         rawVarFuncTable(cds) <- c( "_all" = "_blind" )
+      else {
+         a <- rep( "_blind", length( levels( conditions(cds) ) ) )
+	 names(a) <- levels( conditions(cds) )
+	 rawVarFuncTable(cds) <- a } }
+
+   else if( method == "normal" ) {
[…]

### 2011-08-04 data/ua159.pl

@@ -1,0 +1,41 @@
+use strict;
+
+my @genesUnique;
+my $line;
+open IN, "ua159.smu109.info.csv";
+while ($line = <IN>)
+{
+  chomp $line;
+  my $i = 0;
+  while ($i != -1)
+  {
+    $i = index $line, "SMU109", $i;
+    if ($i != -1)
+    {
+      my @e = split (/[\|,]/, substr ($line, $i));
+      my $gene = $e[0];
+      push @genesUnique, $gene;
+      $i += length($gene);
+      print "$gene\n";
+    }
+  }
+}
+close IN;
+
+print "=======================================\n";
+open SIG, "ua159.out.compare";
+while ($line = <SIG>)
+{
+  chomp $line;
+  my @e = split /\s+/, $line;
+  my $g = $e[1];
+  print "For $g\n";
+  foreach my $gUnique (@genesUnique)
+  {
+    if ($g eq $gUnique)
+    {
+      print "         FOUND $g : $line\n";
+    }
+  }
+}
+close SIG;

### 2011-08-04 pl/bwa-sam.pl

@@ -122,6 +122,7 @@ elsif ($cmd eq "parse")
       chomp;
       my @e = split /\t/;
       my $flag = $e[1];
+      my $chr = $e[2];
       my $mapQuality = $e[4];
       unless ($#e > 10)
         {
@@ -137,6 +138,38 @@ elsif ($cmd eq "parse")
       $e[17] =~ /XG:i:(\d+)/; my $xg = $1;
       $e[18] =~ /MD:Z:(.+)/;  my $md = $1;
       print "$flag\t$mapQuality\t$xt\t$nm\t$x0\t$x1\t$xm\t$xo\t$xg\t$md\n";
+      if ($s % 1000000 == 0)
+        {
+          print STDERR "Reads $s\r";
+        }
+    }
+  print STDERR "Skipped reads: $skippedRead\n"; 
+}
+elsif ($cmd eq "pos")
+{
+
+  my $s = 0;
+  my $skippedRead = 0;
+  while (<$infile>)
+    {
+      $s++;
+      chomp;
+      # See this thread of dicussion for finding ending positions.
+      # http://sourceforge.net/mailarchive/forum.php?thread_name=4A6F0237.30806%40broadinstitute.org&forum_name=samtools-help
+      my @e = split /\t/;
+      my $chr = $e[2];
+      my $mapQuality = $e[4];
+      unless ($mapQuality > $mapq and $#e > 10)
+        {
+          $skippedRead++; 
+          next;
+        }
+      my $shortReadIdentifier = $e[0];
+      my $shortReadStart = $e[3];
+      $_ = $e[5];
+      my $shortReadEnd = $e[3]-1;
+      s/(\d+)[NMD]/$shortReadEnd+=$1/eg;
+      print $outfile "$shortReadIdentifier\t$chr\t$shortReadStart\t$shortReadEnd\n"; 
       if ($s % 100000 == 0)
         {
           print STDERR "Reads $s\r";
@@ -174,6 +207,7 @@ Command:
   mapq  - print MAPQ values
   size  - print the number of short reads mapped
   parse - print FLAG, MAPQ, XT, X0, X1, XM, XO, XG
+  pos   - print short read name, chr, start, and end positions
 
   1  QNAME   String
   2  FLAG    Int 
@@ -190,6 +224,10 @@ Command:
 =head1 OPTIONS
 
 =over 8
+
+=item B<-mapq> <number>
+
+Minimum mapping quality to be called short read maps.
 
 =item B<-in> <file>
 

### 2011-08-04 sh/bwa-mpileup.sh

@@ -40,10 +40,11 @@ function bwa-mpileup {
       GENOMEFASTA=$(basename $REFGENOMEFASTA)
       NUMFASTQFILE=$(grep NUMFASTQFILE $SPECIESFILE | cut -d":" -f2)
       REFGENOMELENGTH=$(grep REFGENOMELENGTH $SPECIESFILE | cut -d":" -f2)
+      READDEPTH=$(grep READDEPTH $SPECIESFILE | cut -d":" -f2)
       #for g in $(eval echo {2..$NUMFASTQFILE}); do
-      for g in $(eval echo {2..2}); do
+      for g in $(eval echo {1..1}); do
         FASTQNUM=FASTQ$(printf "%02d" $g)
-        COMMAND1="$SAMTOOLS mpileup -6 -C50 -d 100000 \
+        COMMAND1="$SAMTOOLS mpileup -6 -C50 -d $READDEPTH \
                   -f $DATADIR/$GENOMEFASTA \
                   $BWADIR/$FASTQNUM.sorted.bam \
                   > $BWADIR/$FASTQNUM.pileup"

### 2011-08-04 sh/bwa-summary.sh

@@ -17,7 +17,7 @@
 # along with Mauve Analysis.  If not, see <http://www.gnu.org/licenses/>.
 ###############################################################################
 
-function bwa-sam {
+function bwa-summary {
   PS3="Choose the species for $FUNCNAME: "
   select SPECIES in ${SPECIESS[@]}; do 
     if [ "$SPECIES" == "" ];  then
@@ -29,7 +29,7 @@ function bwa-sam {
       global-variable $SPECIES $REPETITION
       read-species
 
-      echo -n "Do you wish to run a batch? (e.g., y/n)"
+      echo -n "Do you wish to run a batch? (e.g., y/n) "
       read WISH
       if [ "$WISH" == "y" ]; then
         BATCH=YES
@@ -40,23 +40,38 @@ function bwa-sam {
       NUMFASTQFILE=$(grep NUMFASTQFILE $SPECIESFILE | cut -d":" -f2)
       for g in $(eval echo {1..$NUMFASTQFILE}); do
         FASTQNUM=FASTQ$(printf "%02d" $g)
-        COMMAND1="$SAMTOOLS view $BWADIR/$FASTQNUM.sorted.bam \
-          | perl pl/bwa-sam.pl size > $BWADIR/$FASTQNUM-sum.size"
+
         COMMAND2="$SAMTOOLS view $BWADIR/$FASTQNUM.sorted.bam \
-          | perl pl/bwa-sam.pl mapq > $BWADIR/$FASTQNUM-sum.mapq"
+          | perl pl/bwa-sam.pl parse > $BWADIR/$FASTQNUM-sum.parse"
+
+        COMMAND3="$SAMTOOLS view $BWADIR/$FASTQNUM.sorted.bam \
+          | perl pl/bwa-sam.pl pos > $BWADIR/$FASTQNUM-sum.pos"
 
         if [ "$BATCH" == "YES" ]; then
-          echo $COMMAND1 >> $BATCHFILE
           echo $COMMAND2 >> $BATCHFILE
+          #echo $COMMAND3 >> $BATCHFILE
         else
-          echo $COMMAND1 | bash
           echo $COMMAND2 | bash
-          echo "Check $BWADIR/$FASTQNUM.bed"
+          #echo $COMMAND3 | bash
+          echo "Check $BWADIR/$FASTQNUM-sum files"
         fi
+        # Find the total number of reads and mapped reads.
[…]

### 2011-08-04 sh/fastq-summary.sh

@@ -26,6 +26,7 @@ function fastq-summary {
         TOTALLENGTH=$(zcat $GZIPFASTAQFILE|wc -l)
         TOTALLENGTH=$(($TOTALLENGTH / 4))
         echo "The total number of short reads is $TOTALLENGTH"
+
         NUMBERMAPPEDREAD=$(trim $(cat $BWADIR/$FASTQNUM.bed|wc -l))
         PERCENTMAPPEDREAD=$(($NUMBERMAPPEDREAD * 100 / $TOTALLENGTH))
         echo "The number of mapped reads using BWA is $NUMBERMAPPEDREAD ($PERCENTMAPPEDREAD%)"

### 2011-08-04 sh/run.sh

@@ -58,10 +58,12 @@ source sh/fastxtoolkit-fastx_quality_stats.sh
 
 source sh/fastq-sample.sh
 source sh/degseq.sh
+source sh/deseq.sh
 source sh/edgeR.sh
 source sh/transcript-cufflinks.sh
 source sh/maq-align.sh
-source sh/bwa-sam.sh
+source sh/bwa-summary.sh
+source sh/de-count.sh
 source sh/bwa-mpileup.sh
 
 #####################################################################
@@ -93,7 +95,7 @@ CHOICES=( init-file-system \
           # bwa-samtools-bed \
           # bwa-batch-align-to-bed \
           bwa-samtools-wig \
-          bwa-sam \
+          bwa-summary \
           ---TRACK---\
           bwa-R-saveimage \
           bwa-danko-writewiggle \
@@ -105,6 +107,8 @@ CHOICES=( init-file-system \
           ---FASTQ---\
           fastq-sample \
           ---DE---\
+          de-count \
+          deseq \
           bwa-degseq-bed \
           bowtie-refflat \
           degseq \
@@ -155,13 +159,15 @@ select CHOICE in ${CHOICES[@]}; do
   elif [ "$CHOICE" == "bwa-samtools-wig" ]; then $CHOICE; break
   elif [ "$CHOICE" == "fastq-sample" ]; then $CHOICE; break
   elif [ "$CHOICE" == "degseq" ]; then $CHOICE; break
+  elif [ "$CHOICE" == "deseq" ]; then $CHOICE; break
   elif [ "$CHOICE" == "edgeR" ]; then $CHOICE; break
   elif [ "$CHOICE" == "warranty" ]; then $CHOICE; break
   elif [ "$CHOICE" == "copyright" ]; then $CHOICE; break
   elif [ "$CHOICE" == "copyright" ]; then $CHOICE; break
   elif [ "$CHOICE" == "transcript-cufflinks" ]; then $CHOICE; break
   elif [ "$CHOICE" == "maq-align" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "bwa-sam" ]; then $CHOICE; break
+  elif [ "$CHOICE" == "bwa-summary" ]; then $CHOICE; break
+  elif [ "$CHOICE" == "de-count" ]; then $CHOICE; break
   elif [ "$CHOICE" == "quit" ]; then break
   else
     echo -e "You need to enter something\n"

### 2011-08-04 test/DESeq/deseq.R

@@ -1,0 +1,818 @@
+loadlibrary <- function ()
+{
+  library( DESeq )
+  library( edgeR ) 
+  library( hexbin ) 
+  library( latticeExtra ) 
+  library( gplots )
+  library( geneplotter )
+  sessionInfo()
+}
+
+loadfly <- function ()
+{
+  countsTableFly <- read.delim( "fly_RNA_counts.tsv" )
+  condsFly <- c( "A", "A", "B", "B" )
+  # Add dummy names to avoid confusion later
+  rownames( countsTableFly ) <- paste( "Gene", 1:nrow(countsTableFly), sep="_" )
+
+  cdsFly <- newCountDataSet( countsTableFly, condsFly )
+  cdsFly <- estimateSizeFactors( cdsFly )
+  cdsFly <- estimateVarianceFunctions( cdsFly )
+  resFly <- nbinomTest( cdsFly, "A", "B" )
+}
+
+loadneural <- function ()
+{
+countsTableNS <- read.delim( "NeuralStemCellData.tab", row.names=1 )
+condsNS <- c( "GNS_dupe", "GNS", "GNS", "GNSL", "NS", "NS" )
+}
+
+loadyeast <- function ()
+{
+countsYeast <- read.delim( "counts_Nagalakshmi_et_al.tab", row.names=1 ) 
+}
+
+runedgeR <- function ()
+{
+  # using DESeq's size factors:
+
+  dglFlyD <- DGEList( counts=countsTableFly, group=condsFly, lib.size=1e7*sizeFactors(cdsFly) )
+  dglFlyD <- estimateCommonDisp( dglFlyD )
+  dglFlyD <- estimateTagwiseDisp( dglFlyD )
+
+  # test with tag-wise dispersion:
+
+  edgerResFlyDT <- exactTest( dglFlyD, common.disp=FALSE )
+  edgerResFlyDTPadj <- p.adjust( edgerResFlyDT$table$p.value, method="BH" )
+
+  # common dispersion, DESeq's size factors:
+
+  edgerResFlyDC <- exactTest( dglFlyD, common.disp=TRUE )
+  edgerResFlyDCPadj <- p.adjust( edgerResFlyDC$table$p.value, method="BH" )
+
+  # with the original library sizes:
+
+  # the total number of sequenced counts were as follows:
+  libsizesFly <- c( A1=13613446, A2=12106517, B1=12920058, B2=11599459 )
+
[…]

### 2011-08-04 test/DESeq/pasilla.R

@@ -1,0 +1,138 @@
+#library( "DESeq" )
+
+# Could not find this package.
+# library( "pasilla" )
+# data( "pasillaGenes" )
+
+loadfly <- function ()
+{
+  countsTableFly <- read.delim( "fly_RNA_counts.tsv" )
+
+  # Q: Which one is right?
+  condsFly <- c( "A", "A", "B", "B" )
+  # condsFly <- factor(c( "A", "A", "B", "B" ))
+
+  # Add dummy names to avoid confusion later
+  rownames( countsTableFly ) <- paste( "Gene", 1:nrow(countsTableFly), sep="_" )
+}
+
+deseq1_fly <- function ()
+{
+  cdsFly <- newCountDataSet( countsTableFly, condsFly )
+  cdsFly <- estimateSizeFactors( cdsFly )
+  cdsFly <- estimateVarianceFunctions( cdsFly )
+}
+
+deseq2_fly <- function ()
+{
+  resFly <- nbinomTest( cdsFly, "A", "B" )
+}
+
+# Q: Not working
+# ls(cdsFly@fitInfo)
+plotDispEsts <- function( cds, cond )
+{
+   plot( 
+      rowMeans( counts( cds, normalized=TRUE ) ),
+      cds@fitInfo[[cond]]$perGeneDispEsts,
+      pch = '.', log="xy" )
+   xg <- 10^seq( -.5, 5, length.out=300 )
+   lines( xg, cds@fitInfo[[cond]]$dispFun( xg ), col="red" )   
+}
+
+# plotDispEsts ( cdsFly, "A")
+# print(head(fData(cdsFly)))
+
+
+plotDE <- function( res )
+   plot( 
+      res$baseMean, 
+      res$log2FoldChange, 
+      log="x", pch=20, cex=.3, 
+      col = ifelse( res$padj < .1, "red", "black" ) )
+
+# plotDE( resFly )
+
+res <- resFly
+cds <- cdsFly
+# Histogram of p values
+#hist(res$pval, breaks=100, col="skyblue", border="slateblue", main="")
+
+# Filter for significant genes at FDR 10%
+resSig <- res[ res$padj < 0.1, ]
+#print(head( resSig[ order(resSig$pval), ] ))
+
+# Downregulated
+#print(head( resSig[ order( resSig$foldChange, -resSig$baseMean ), ] ))
+# Upregulated
+#print(head( resSig[ order( -resSig$foldChange, -resSig$baseMean ), ] ))
+
[…]

### 2012-01-05 package/smutans/inst/doc/smutans.Rnw
smutans package is being developed.

%\VignetteIndexEntry{Analyzing Streptococcus mutans RNA-Seq data}
%\VignettePackage{smutans}

% To compile this document
% library('weaver'); rm(list=ls()); Sweave('smutans.Rnw', driver=weaver()); system('pdflatex smutans')

\documentclass{article}

\usepackage{Sweave}
\usepackage[letterpaper]{geometry}
\usepackage{hyperref,graphicx}
\usepackage{color}

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=4,height=4.5}
\newcommand{\Robject}[1]{\texttt{#1}}
\newcommand{\Rpackage}[1]{\textit{#1}}
\newcommand{\Rclass}[1]{\textit{#1}}
\newcommand{\Rfunction}[1]{{\small\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

\renewcommand{\floatpagefraction}{0.7}

\author{Sang Chul Choi\\[1em]Cornell University,\\ Ithaca, NY\\[1em]
\texttt{schoi@cornell.edu}}

\title{\textsf{\textbf{Analysing \textit{Streptococcus mutans} RNA-Seq data using R packages}}}

\begin{document}

\maketitle

\begin{abstract}
A number of RNA samples from \textit{Streptococcus mutans} are analyzed using R
packages: \Rpackage{DESeq}, \Rpackage{edgeR}, \Rpackage{goseq}, and \Rpackage{baySeq}.
Tasks include the analysis of count data from RNA-Seq for the detection of
differentially expressed genes.
Tests of enrichment for differentially expressed genes are performed using
\Rpackage{goseq}.
The package \Rpackage{smutans} contains the procedures.
This vignette explains the use of the package. 
\end{abstract}

\tableofcontents

%--------------------------------------------------
\section{Input data and preparations} \label{sec:prep}
%--------------------------------------------------

The \Rpackage{smutans} package expects count data in
the form of \Rclass{CountDataSet}.

To obtain such a count table for your own data, you will need
to create it from your sequence alignments and suitable annotation. Within Bioconductor,
you can use the function \Rfunction{summerizeOverlaps} in the
\Rpackage{GenomicRanges} package. See the vignette, Ref.\ \cite{summerizeOverlaps},
for a worked example. Another possibility (outside of Bioconductor)
is the \emph{htseq-count} script distributed with the HTSeq Python framework
\cite{htseq}. (You do not need to know any Python to use
\texttt{htseq-count}.) A third possibility might be given by the Bioconductor
package \Rpackage{easyrnaseq} (by Nicolas Delhomme; in preparation, available
soon; package name may change).

Another easy way to produce such a table from the output
  of the aligner is to use the \texttt{htseq-count} script distributed
  with the \emph{HTSeq} package. Even though \emph{HTSeq} is a Python
  package, you do not need to know any Python to use
  \texttt{htseq-count}. See
  \url{http://www-huber.embl.de/users/anders/HTSeq/doc/count.html}. (If you use
htseq-count, be sure to remove the extra lines with general counters (``ambiguous'' etc.)
when importing the data.)

The count values must be raw counts of
sequencing reads. This is important for \Rpackage{smutans}'s statistical model to hold,
as only raw reads allow to assess the measurement precision correctly. (Hence,
do not supply rounded values of normalized counts, or counts of covered base pairs.)

Furthermore, it is important that each column stems from an independent biological
replicate.  For purely technical replicates (e.\,g. when the same
library preparation was distributed over multiple lanes of the
sequencer in order to increase coverage), please sum up their counts
to get a single column, corresponding to a unique biological
replicate.  This is needed in order to allow \Rpackage{DESEq} to
estimate variability in the experiment correctly.

As an example dataset, we use the gene level read counts from the
\Rpackage{pasilla} data package. This dataset is from an experiment on
\emph{Drosophila melanogaster} cell cultures and
investigated the effect of RNAi knock-down of the splicing factor
\emph{pasilla}~\cite{Brooks2010}. The data are presented in the object
called \Robject{pasillaGenes}. For a description how this data object
was created from the raw data of Ref.~\cite{Brooks2010}, see the vingette
included with the \Rpackage{pasilla} package.

The \Robject{pasillaGenes} object is of class
\Rclass{CountDataSet}, which is the data container used by
\Rpackage{smutans}. We load the needed packages and the data as
follows.

<<options,results=hide,echo=FALSE>>=
options(digits=3)
@

There are 13 RNA-seq samples in \Robject{smutansGenes} that can be loaded
from the package.
<<smutans2,results=hide>>=
library( "smutans" )
data( "smutansGenes" )
@

\Robject{smutansGenes} contains the counts and also metadata about the samples:

<<look_into_smutansGenes>>=
head( counts(smutansGenes) )
pData( smutansGenes )
@

As you can see, the samples differ by experimental condition (glucose or
galactose) and by library type (ua159 or tw1). First, we only
look at the ua159 data for now. We will look at the tw1 data later.

We create an \Rclass{Smutans} object using \Rclass{CountDataSet} object,
smutansGenes.
<<ua159onlySamples>>=
ua159only <- newSmutans( smutansGenes )
ua159only <- smutans.de2type( ua159only, type="ua159", 
                              condA="glucose", condB="galactose" )
@

\Rfunction{smutans.de2type} uses a two-factor data set to test differential
expression of genes.
Note that the \Rclass{Smutans} contains \Rclass{CountDataSet} objects. 
\Rfunction{smutans.de2type} estimates the effective library size, and the
dispersions. Then, it tests for differential expression.

Because it is recommended to check
the dispersion estimates and to make sure that the data quality is as
expected, we plot the per-gene estimates against the normalized mean
expressions per gene, and then overlay the fitted curve in red.

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFit}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength.}
\label{figFit}
\end{figure}

%Calling the function produces the plot (Fig.\ \ref{figFit}).

<<figFit,fig=TRUE>>=
smutans.plotDispersionEstimates(ua159only)
@

The plot in Figure~\ref{figFit} is doubly logarithmic; this may be
helpful or misleading, and it is worth experimenting with other
plotting styles.

As we estimated the
dispersion from only two samples, we should expect the estimates to
scatter with quite some sampling variance around their true values.
Hence, we \Rpackage{smutans} should not use the per-gene estimates directly in the test,
because using too low dispersion values leads to false positives.
Many of the values below the red line are likely to be underestimates of the
true dispersions, and hence, it is prudent to instead rather use the fitted value.
On the othe hand, not all of he values above the red line are overestimations, and
hence, the conservative choice is to keep them instead of replacing them with
their fitted values. If you do not like this default behaviour, you can change it
with the option \texttt{sharingMode} of \Rfunction{estimateDispersions}. Note that

In any case, the dispersion values which finally should be used by the subsequent testing are stored in
the feature data slot of \texttt{cds}:

%------------------------------------------------------------
\section{Inference: Calling differential expression} \label{sec:DE}
%------------------------------------------------------------

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figDE}
\caption{Plot of normalised mean versus $log_2$ fold change
(this plot is sometimes also called the ``MA-plot'') for the contrast ``untreated''
versus ``treated''.}
\label{figDE}
\end{figure}

\subsection{Standard comparison between two experimental conditions}

Having estimated the dispersion for each gene, it is now straight-forward to look for
differentially expressed genes. To contrast two conditions, e.g., to see whether there is differential
expression between conditions ``untreated'' and ``treated'', we simply call the function \Rfunction{nbinomTest}.
It performs the tests as described in the paper and returns a data frame with the $p$ values and other
useful information.

For each gene, we get its mean expression level (at the base scale) as a joint estimate from both conditions, and
estimated separately for each  condition, the fold change from
the first to the second condition, the logarithm (to basis 2) of the fold change, and the
$p$ value for the statistical significance of this change. The \texttt{padj} column contains the
$p$ values, adjusted for multiple testing with the Benjamini-Hochberg procedure (see the R function
\Rfunction{p.adjust}), which controls false discovery
rate (FDR).

Let us first plot the $\log_2$ fold changes against the base means, colouring in red those genes
that are significant at 10\% FDR.
<<figDE,fig=TRUE>>=
smutans.plotDiffExp(ua159only)
@ %$
See Figure~\ref{figDE} for the plot. As we will use this plot more often, we have
stored its code in a function.

%------------------------------------------------------------
\section{Inference: Calling differential expression} \label{sec:DE}
%------------------------------------------------------------

<<allSamples>>=
allSample <- newSmutans( smutansGenes )
allSample <- smutans.de2( allSample )
@

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFit2}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength.}
\label{figFit}
\end{figure}

Calling the function produces the plot (Fig.\ \ref{figFit2}).

<<figFit2,fig=TRUE>>=
smutans.plotDispersionEstimates(allSample)
@

<<compareUA159onlyAndAllSample>>=
padj.ua159only <- smutans.padj( ua159only )
padj.allSample <- smutans.padj( allSample )
tab = table( "ua159 only" = padj.ua159only < .1, "all samples" = padj.allSample < .1 )
addmargins( tab )
@
Of the \Sexpr{sum(tab)} gene, \Sexpr{tab[2,2]} genes are found in both analyses. The analysis using all
of the sample found \Sexpr{tab[1,2]} genes. The analysis using only ua159 found \Sexpr{tab[2,1]} genes.

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figCompareUA159onlyAndAllSample}
\caption{Comparison of $p$ values (unadjusted) from the test using only the
four paired-end samples and the test using all seven samples. We can see that
the latter tend to be smaller, as expected from the higher power of the test with all samples.}
\label{figCompareUA159onlyAndAllSample}
\end{figure}

<<figCompareUA159onlyAndAllSample,fig=TRUE>>=
pval.ua159only <- smutans.pval( ua159only )
pval.allSample <- smutans.pval( allSample )
bottom = function(x, theta=1e-12) pmax(x, theta)
plot( bottom(pval.ua159only), bottom(pval.allSample), log="xy", pch=20, cex=.3 )
abline(a=0, b=1, col="blue")
@

<<allSampleWithoutType>>=
allSampleWithoutType <- newSmutans( smutansGenes )
allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
                                                condA="glucose", 
                                                condB="galactose" )
@

<<compareUA159onlyAndAllSample>>=
padj.allSampleWithoutType <- smutans.padj( allSampleWithoutType )
tab2 <- table(
   `all samples simple` = padj.allSampleWithoutType < 0.1,
   `all samples GLM`    = padj.allSample < 0.1 )
addmargins( tab2 )
largeNumber <- 10
@
Of the \Sexpr{sum(tab2)} gene, \Sexpr{tab2[2,2]} genes are found in both analyses. The analysis using all
of the sample found \Sexpr{tab2[1,2]} genes. The analysis using all of the same
sample but without considering type found \Sexpr{tab2[2,1]} genes.

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figLfc}
\caption{Scatterplot of direct (\Robject{lfc}) versus \emph{moderated} log-ratios (\Robject{mod\_lfc}).
  The moderation criterion used is variance stabilisation.
  The red points correspond to values that were infinite in \Robject{lfc}
  and were abitrarily set to $\pm\Sexpr{largeNumber}$ for the purpose of plotting.
  These values vary in a finite range (as shown in the plot) in \Robject{mod\_lfc}.}
\label{figLfc}
\end{figure}

<<figLfc,fig=TRUE>>=
select <- 1:40
smutans.de2TypeLfc( ua159only, condA="glucose", condB="galactose" )
@

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figHeatmap1}
\caption{Heatmaps showing the expression data of the top \Sexpr{length(select)}
  differentially expressed genes. Left, the variance stabilisation
  transformed data,
  right, the original count data.
  The right plot is dominated by a small number of data points with large values.}
\label{figHeatmap1}
\end{figure}

<<figHeatmap1,fig=TRUE>>=
smutans.de2TypeHeatmap( ua159only )
@

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figClust}
\caption{Heatmap showing the Euclidean distances between the samples
  as calculated from the variance-stabilising transformation of the
  count data.}
\label{figClust}
\end{figure}

<<figClust,fig=TRUE>>=
smutans.de2Clust( allSample )
@

\section{Session Info}
<<sessi>>=
sessionInfo()
@

\bibliographystyle{unsrt}
\bibliography{smutans}

\end{document}

### 2012-01-05 R/deseq-cornell-ua159-tw1.R

@@ -1,0 +1,137 @@
+# Rscript R/deseq.R count.txt species conditionA conditionB
+.Last <- function() {
+  cat("\n")
+}
+
+plotDispEsts <- function( cds )
+{
+   plot(
+      rowMeans( counts( cds, normalized=TRUE ) ),
+      fitInfo(cds)$perGeneDispEsts,
+      pch = '.', log="xy" )
+   xg <- 10^seq( -.5, 5, length.out=300 )
+   lines( xg, fitInfo(cds)$dispFun( xg ), col="red" )
+}
+
+plotDE <- function( res )
+   plot(
+      res$baseMean,
+      res$log2FoldChange,
+      log="x", pch=20, cex=.3,
+      col = ifelse( res$padj < .1, "red", "black" ) )
+
+args <- commandArgs(trailingOnly = TRUE)
+#if (length(args) != 2)
+#{
+#  cat ("Rscript R/deseq-cornell-ua159-tw1.R output/cornell/1/bwa/count.txt 0.01\n")
+#  quit("yes")
+#}
+#countsFile <- args[1]
+#cutoffpval <- as.numeric(args[2])
+#print(countsFile)
+#print(cutoffpval)
+countsFile <- "output/cornell/1/bwa/count.txt"
+cutoffpval <- 0.01
+
+prepare.input.data <- function ()
+{
+library(DESeq)
+# Count data preparation
+conds <- scan(file=paste(countsFile,"index",sep="."), what="character")
+subconds <- conds %in% c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL")
+countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)
+rownames(countsTable) <- countsTable$gene
+countsTable <- countsTable[,-1]
+countsTable <- countsTable[,subconds]
+conds <- conds[subconds]
+conds.type <- conds %in% c("UA159GLU", "UA159GAL")
+factor.type <- rep("ua159",length(conds))
+factor.type[!conds.type] <- "tw1"
+conds.condition <- conds %in% c("UA159GLU", "TW1GLU")
+factor.condition <- rep("glucose",length(conds))
+factor.condition[!conds.condition] <- "galactose"
+samples <- data.frame(type=factor.type,condition=factor.condition)
+rownames(samples) <- colnames(countsTable) 
+
+design <- samples
[…]

### 2012-01-05 package/smutans/R/class_and_slots.R

@@ -1,0 +1,18 @@
+setClass( "Smutans", 
+  representation = representation( 
+    countDataSet = "CountDataSet", 
+    cds = "ANY", 
+    res = "list", 
+    padj = "numeric",
+    pval = "numeric"
+    )
+)
+
+newSmutans <- function( countData )
+{
+  cds <- new( "Smutans", 
+              countDataSet = countData,
+              cds = countData )
+  cds 
+}
+

### 2012-01-05 package/smutans/R/methods.R

@@ -1,0 +1,219 @@
+setGeneric("smutans.counts", function(object) standardGeneric("smutans.counts"))
+setMethod("smutans.counts",
+          "Smutans",
+  function( object ) {
+    counts( object@countDataSet )
+  }
+)
+
+setGeneric("smutans.counts<-", function(object,value) standardGeneric("smutans.counts<-"))
+setReplaceMethod("smutans.counts", 
+                 "Smutans",
+  function( object, value ) {
+    object@counts <- value
+    object
+  }
+)
+
+setGeneric("smutans.de2type", function(object,...) standardGeneric("smutans.de2type"))
+setMethod("smutans.de2type", 
+          "Smutans", 
+  function(object, type, condA, condB) {
+    ua159Samples <- pData(object@countDataSet)$type == type
+    countsTable <- counts(object@countDataSet)[ , ua159Samples ]
+    conds <- pData(object@countDataSet)$condition[ ua159Samples ]
+    cds <- newCountDataSet( countsTable, conds )
+    cds <- estimateSizeFactors( cds )
+    object@cds <- estimateDispersions( cds )
+    object@res <- nbinomTest( object@cds, condA, condB )
+    object@pval <- object@res$pval
+    object@padj <- object@res$padj
+    object
+  }
+)
+
+setGeneric("smutans.plotDispersionEstimates", function(object,...) standardGeneric("smutans.plotDispersionEstimates"))
+setMethod("smutans.plotDispersionEstimates", 
+          "Smutans", 
+  function(object, file="default", 
+           xlab="Base mean expression", ylab="Dispersion values", qval=0.05 ) 
+  {
+    if (file != "default")
+    {
+      postscript(file, width=6, height=6, horizontal = FALSE, onefile = FALSE, paper = "special")
+      oldpar <- par (mar=c(4.5, 4.5, 0.5, 0.5))
+    }
+    plot( rowMeans( counts( object@cds, normalized=TRUE ) ),
+      fitInfo(object@cds)$perGeneDispEsts,
+      xlab=xlab, ylab=ylab,
+      pch = '.', log="xy" )
+    xg <- 10^seq( -.5, 5, length.out=300 )
+    lines( xg, fitInfo(object@cds)$dispFun( xg ), col="red" )
+    if (file != "default")
+    {
+      par(oldpar)
+      dev.off()
+    }
+  }
+)
+
[…]

### 2012-01-05 package/test.R

@@ -1,0 +1,33 @@
+# library(DESeq)
+# load("smutans/data/smutansGenes.RData")
+source("smutans/R/class_and_slots.R")
+source("smutans/R/methods.R")
+#ua159only <- smutans.de2type( ua159only, type="ua159", 
+#                              condA="glucose", condB="galactose" )
+#allSample <- smutans.de2( allSample )
+#
+#padj.ua159only <- smutans.padj( ua159only )
+#padj.allSample <- smutans.padj( allSample )
+#tab = table( "ua159 only" = padj.ua159only < .1, "all samples" = padj.allSample < .1 )
+#addmargins( tab )
+#
+#pval.ua159only <- smutans.pval( ua159only )
+#pval.allSample <- smutans.pval( allSample )
+#bottom = function(x, theta=1e-12) pmax(x, theta)
+#plot( bottom(pval.ua159only), bottom(pval.allSample), log="xy", pch=20, cex=.3 )
+#abline(a=0, b=1, col="blue")
+#
+#allSampleWithoutType <- newSmutans( smutansGenes )
+#allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
+#                                                condA="glucose", 
+#                                                condB="galactose" )
+padj.allSampleWithoutType <- smutans.padj( allSampleWithoutType )
+tab2 <- table(
+   `all samples simple` = padj.allSampleWithoutType < 0.1,
+   `all samples GLM`    = padj.allSample < 0.1 )
+addmargins( tab2 )
+
+# smutans.de2TypeHeatmap( ua159only )
+# smutans.de2TypeLfc( ua159only, condA="glucose", condB="galactose" )
+smutans.de2Clust( allSample )
+

### 2012-01-05 pl/de-count.pl

@@ -45,6 +45,7 @@ GetOptions( \%params,
             'verbose',
             'version' => sub { print $VERSION."\n"; exit; },
             'first',
+            'singlegenome',
             'shortread=s',
             'genepos=s',
             'out=s',
@@ -189,7 +190,7 @@ elsif ($cmd eq "join")
       # multiple chromosomes.
       # We have to change feature-genome.out-geneonly or
       # feature-genome.pl script.
-      if ($g->{chr} eq $s->{chr}) # Why did I comment this?
+      if (exists $params{singlegenome} || $g->{chr} eq $s->{chr}) # Why did I comment this?
       {
         my $v = $g->{count};
         if ($g->{start} <= $s->{start} and $s->{start} <= $g->{end})

### 2012-01-05 test/DESeq/deseq.R

@@ -811,8 +811,8 @@ abline( v = 7000 / (7000+4000) * 10000 )
 
 ###################################################################
 # From the manual
-#library( "DESeq" )
-#library( "pasilla" )
-#data( "pasillaGenes" )
+library( "DESeq" )
+library( "pasilla" )
+data( "pasillaGenes" )
 
 

### 2012-01-05 test/edgeR/run.R

@@ -1,19 +1,62 @@
-library(edgeR)
-library(limma)
-targets = read.delim(file = "Targets.txt",sep=" ",stringsAsFactors = FALSE)
+# library(edgeR)
+# library(limma)
+
+
+
+ua159.tw1.glu.gal.test = function()
+{
+################################################################################
+# A Test of multifactor analysis: UA159
+# 
+# 0 for glucose, and 1 for galactose
+# w for UA159, and m for TW1
+targets = read.delim(file = "ua159-tw1-glu-gal.txt",sep=" ",stringsAsFactors = FALSE)
+sugar <- factor(c(0,0,1,0,0,1,1,1,0,0,1,1,1))
+# mutant <- factor(c(w,m,m,w,w,w,w,w,m,m,m,m,w))
 d = readDGE(targets,head=F)
-d <- d[rowSums(d$counts) >= 5,] # Filter out those with counts fewer than 5.
+d <- calcNormFactors(d)
+cpm.sugar <- cpm(d)
+print(rownames(d[rowSums(cpm.sugar > 1) < 2,]$counts))
+d <- d[rowSums(cpm.sugar > 1) >= 2,]
+design <- model.matrix(~sugar + d$samples$group)
+rownames(design) <- rownames(d$samples)
+colnames(design)[2] <- "sugar"
+colnames(design)[3] <- "mutant"
+d <- estimateGLMCommonDisp(d,design)
+glmfit.sugar <- glmFit(d,design,dispersion=d$common.dispersion)
+lrt.sugar <- glmLRT(d,glmfit.sugar,coef=3)
+lrt.mutant <- glmLRT(d,glmfit.sugar,coef=2)
+# topTags(lrt.sugar)
+}
+
+################################################################################
+# A Test
+np835.proc <- function ()
+{
+targets = read.delim(file = "np835.txt",sep=" ",stringsAsFactors = FALSE)
+d.835 = readDGE(targets,head=F)
+d.835 <- calcNormFactors(d.835)
+cpm.835 <- cpm(d.835)
+print(rownames(d.835[rowSums(cpm.835 > 1) < 2,]$counts))
+d.835 <- d.835[rowSums(cpm.835 > 1) >= 2,]
+plotMDS(d.835)
+
+# d <- d[rowSums(d$counts) >= 5,] # Filter out those with counts fewer than 5.
 # dim(d) # Show the number of genes and the number of samples
 # d$samples$lib.size # Show the total number of reads for each of the samples
[…]

### 2012-01-05 package/smutans/inst/doc/smutans.Rnw
Just an update.

@@
 
-\title{\textsf{\textbf{Analysing \textit{Streptococcus mutans} RNA-Seq data using R packages}}}
+\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-Seq data using R packages}}}
 
@@
 \begin{abstract}
-A number of RNA samples from \textit{Streptococcus mutans} are analyzed using R
-packages: \Rpackage{DESeq}, \Rpackage{edgeR}, \Rpackage{goseq}, and \Rpackage{baySeq}.
-Tasks include the analysis of count data from RNA-Seq for the detection of
-differentially expressed genes.
-Tests of enrichment for differentially expressed genes are performed using
-\Rpackage{goseq}.
-The package \Rpackage{smutans} contains the procedures.
-This vignette explains the use of the package. 
+We analyze a number of RNA samples from \textit{Streptococcus mutans} using R
+packages: \Rpackage{DESeq}, \Rpackage{edgeR}, and \Rpackage{goseq}.
+Tasks include the analysis of count data from RNA-Seq for
+the detection of differentially expressed genes.  Tests of enrichment for
+differentially expressed genes are performed using \Rpackage{goseq}.  The
+package \Rpackage{smutans} contains the procedures.  This vignette explains the
+use of the package. 
 \end{abstract}
@@
 %--------------------------------------------------
-\section{Input data and preparations} \label{sec:prep}
+\section{Data preparations} \label{sec:prep}
 %--------------------------------------------------
 
-The \Rpackage{smutans} package expects count data in
-the form of \Rclass{CountDataSet}.
+The \Rpackage{smutans} package expects count data in the form of
+\Rclass{CountDataSet}. Using BWA and samtools we aligned short reads from RNA
+sequencing. Using gene annotations in the UA159 genome we counted short reads
+that overlapped the genes.  The count values are raw counts of sequencing reads.
+\Rpackage{DESeq}'s statistical model, or negative binomial model, hold for raw
+reads that allows to assess the measurement precision correctly.  We did not
+have rounded values of normalized counts, or counts of covered base pairs.
 
-To obtain such a count table for your own data, you will need
-to create it from your sequence alignments and suitable annotation. Within Bioconductor,
-you can use the function \Rfunction{summerizeOverlaps} in the
-\Rpackage{GenomicRanges} package. See the vignette, Ref.\ \cite{summerizeOverlaps},
-for a worked example. Another possibility (outside of Bioconductor)
-is the \emph{htseq-count} script distributed with the HTSeq Python framework
-\cite{htseq}. (You do not need to know any Python to use
-\texttt{htseq-count}.) A third possibility might be given by the Bioconductor
-package \Rpackage{easyrnaseq} (by Nicolas Delhomme; in preparation, available
-soon; package name may change).
+Several exeriments of \textit{S.\ mutans} RNA sampling were performed.
+\fixme{Find the papers that are related with the experiments.} One of them was
+presented in the object called \Robject{smutansGenes}.  The
+\Robject{smutansGenes} object is of class \Rclass{CountDataSet}, which is the
+data container used by \Rpackage{DESeq}. We have a covenient class called
+\Rclass{Smutans}. The class contains slots for \Rclass{CountDataSet} objects,
+and results from analyses.  Let's load the needed packages and the data as
+follows.
 
-Another easy way to produce such a table from the output
-  of the aligner is to use the \texttt{htseq-count} script distributed
-  with the \emph{HTSeq} package. Even though \emph{HTSeq} is a Python
-  package, you do not need to know any Python to use
-  \texttt{htseq-count}. See
-  \url{http://www-huber.embl.de/users/anders/HTSeq/doc/count.html}. (If you use
-htseq-count, be sure to remove the extra lines with general counters (``ambiguous'' etc.)
-when importing the data.)
-
-The count values must be raw counts of
-sequencing reads. This is important for \Rpackage{smutans}'s statistical model to hold,
-as only raw reads allow to assess the measurement precision correctly. (Hence,
-do not supply rounded values of normalized counts, or counts of covered base pairs.)
-
-Furthermore, it is important that each column stems from an independent biological
-replicate.  For purely technical replicates (e.\,g. when the same
-library preparation was distributed over multiple lanes of the
-sequencer in order to increase coverage), please sum up their counts
-to get a single column, corresponding to a unique biological
-replicate.  This is needed in order to allow \Rpackage{DESEq} to
-estimate variability in the experiment correctly.
-
-As an example dataset, we use the gene level read counts from the
-\Rpackage{pasilla} data package. This dataset is from an experiment on
-\emph{Drosophila melanogaster} cell cultures and
-investigated the effect of RNAi knock-down of the splicing factor
-\emph{pasilla}~\cite{Brooks2010}. The data are presented in the object
-called \Robject{pasillaGenes}. For a description how this data object
-was created from the raw data of Ref.~\cite{Brooks2010}, see the vingette
-included with the \Rpackage{pasilla} package.
-
-The \Robject{pasillaGenes} object is of class
-\Rclass{CountDataSet}, which is the data container used by
-\Rpackage{smutans}. We load the needed packages and the data as
-follows.
+We can create \Rclass{CountDataSet} objects using functions that are not
+exported. One should create an R script at the directory that contains the
+source code of the \Rpackage{smutans}. Put the following scripts in the R script
+and run the script at the directory. That should create RData files at
+the directory smutans/data: sm835Genes.RData, smutansGenes.RData.
+<<samplePreparation,eval=FALSE>>=
+#not run
+library(DESeq)
+source("smutans/R/core.R")
+smutans.prepareData835NPP( )
+smutans.prepareDataUA159TW1( ) 
+q("no")
+@
 
@@
 
-There are 13 RNA-seq samples in \Robject{smutansGenes} that can be loaded
-from the package.
 <<smutans2,results=hide>>=
@@
 
+There are \Sexpr{length(colnames(counts(smutansGenes)))} RNA-seq samples in
+\Robject{smutansGenes} that can be loaded from the package.
 \Robject{smutansGenes} contains the counts and also metadata about the samples:
@@
 head( counts(smutansGenes) )
+@
+
+<<look_into_smutansGenes>>=
 pData( smutansGenes )
@@
 
-As you can see, the samples differ by experimental condition (glucose or
-galactose) and by library type (ua159 or tw1). First, we only
-look at the ua159 data for now. We will look at the tw1 data later.
+The samples differ by experimental condition (glucose or
+galactose) and by library type (ua159 or tw1). First, we only look at the ua159
+data for now. We will look at the tw1 data later. 
+We create an \Rclass{Smutans} object using the \Rclass{CountDataSet} object,
+\Robject{smutansGenes}.
 
-We create an \Rclass{Smutans} object using \Rclass{CountDataSet} object,
-smutansGenes.
+%--------------------------------------------------
+\section{Differential expression analysis} \label{sec:de}
+%--------------------------------------------------
+
+%------------------------------------------------------------
+\subsection{Inference: Using the UA159 samples for condition effect} 
+\label{subsec:ua159only}
+%------------------------------------------------------------
+
 <<ua159onlySamples>>=
@@
 \Rfunction{smutans.de2type} uses a two-factor data set to test differential
-expression of genes.
+expression of genes. We compare 4 pairwise comparisons using a two-factor data
+set. A pair of comparisons between glucose and galactose given either type ua159
+or tw1. Another pairs of comparisons between types ua159 and tw1 given either
+condition glucose or galactose. We focus on one of the four pairwise comparisons
+first. Later, we will perform analyses for the rest of the three comparisons.
 Note that the \Rclass{Smutans} contains \Rclass{CountDataSet} objects.
 \Rfunction{smutans.de2type} estimates the effective library size, and the
-dispersions. Then, it tests for differential expression.
+dispersions using functions featured by \Rpackage{DESeq}. Then, it tests for
+differential expression.
 
-Because it is recommended to check
+Because \Rpackage{DESeq} recommended to check
 the dispersion estimates and to make sure that the data quality is as
 expected, we plot the per-gene estimates against the normalized mean
-expressions per gene, and then overlay the fitted curve in red.
+expressions per gene, and then overlay the fitted curve in red
+(Fig.\ \ref{figFit}).
 
@@
 \end{figure}
-
-%Calling the function produces the plot (Fig.\ \ref{figFit}).
-
 <<figFit,fig=TRUE>>=
@@
 
-The plot in Figure~\ref{figFit} is doubly logarithmic; this may be
-helpful or misleading, and it is worth experimenting with other
-plotting styles.
+As we estimated the dispersion from only two samples, we should expect the
+estimates to scatter with quite some sampling variance around their true values.
+Hence, \Rpackage{DESeq} should not use the per-gene estimates directly in the
+test, because using too low dispersion values leads to false positives.  Many of
+the values below the red line are likely to be underestimates of the true
+dispersions, and hence, it is prudent to instead rather use the fitted value.
+On the othe hand, not all of he values above the red line are overestimations,
+and hence, the conservative choice is to keep them instead of replacing them
+with their fitted values.
 
-As we estimated the
-dispersion from only two samples, we should expect the estimates to
-scatter with quite some sampling variance around their true values.
-Hence, we \Rpackage{smutans} should not use the per-gene estimates directly in the test,
-because using too low dispersion values leads to false positives.
-Many of the values below the red line are likely to be underestimates of the
-true dispersions, and hence, it is prudent to instead rather use the fitted value.
-On the othe hand, not all of he values above the red line are overestimations, and
-hence, the conservative choice is to keep them instead of replacing them with
-their fitted values. If you do not like this default behaviour, you can change it
-with the option \texttt{sharingMode} of \Rfunction{estimateDispersions}. Note that
+Having estimated the dispersion for each gene, we look for differentially
+expressed genes. To contrast two conditions, e.g., to see whether there is
+differential expression between conditions ``glucose'' and ``galactose'',
+\Rfunction{smutans.de2type} already performed the function
+\Rfunction{nbinomTest} provided by \Rpackage{DESeq}.  Let us first plot the
+$\log_2$ fold changes against the base means, colouring in red those genes that
+are significant at 10\% FDR.
 
-In any case, the dispersion values which finally should be used by the subsequent testing are stored in
-the feature data slot of \texttt{cds}:
-
-%------------------------------------------------------------
-\section{Inference: Calling differential expression} \label{sec:DE}
-%------------------------------------------------------------
+For each gene, we get its mean expression level (at the base scale) as a joint
+estimate from both conditions, and estimated separately for each  condition, the
+fold change from the first to the second condition, the logarithm (to basis 2)
+of the fold change, and the $p$ value for the statistical significance of this
+change. The \texttt{padj} column contains the $p$ values, adjusted for multiple
+testing with the Benjamini-Hochberg procedure (see the R function
+\Rfunction{p.adjust}), which controls false discovery rate (FDR).
 
@@
 \caption{Plot of normalised mean versus $log_2$ fold change
-(this plot is sometimes also called the ``MA-plot'') for the contrast ``untreated''
-versus ``treated''.}
+(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
+versus ``galactose''.}
 \label{figDE}
 \end{figure}
-
-\subsection{Standard comparison between two experimental conditions}
-
-Having estimated the dispersion for each gene, it is now straight-forward to look for
-differentially expressed genes. To contrast two conditions, e.g., to see whether there is differential
-expression between conditions ``untreated'' and ``treated'', we simply call the function \Rfunction{nbinomTest}.
-It performs the tests as described in the paper and returns a data frame with the $p$ values and other
-useful information.
-
-For each gene, we get its mean expression level (at the base scale) as a joint estimate from both conditions, and
-estimated separately for each  condition, the fold change from
-the first to the second condition, the logarithm (to basis 2) of the fold change, and the
-$p$ value for the statistical significance of this change. The \texttt{padj} column contains the
-$p$ values, adjusted for multiple testing with the Benjamini-Hochberg procedure (see the R function
-\Rfunction{p.adjust}), which controls false discovery
-rate (FDR).
-
-Let us first plot the $\log_2$ fold changes against the base means, colouring in red those genes
-that are significant at 10\% FDR.
 <<figDE,fig=TRUE>>=
@@
 @ %$
-See Figure~\ref{figDE} for the plot. As we will use this plot more often, we have
-stored its code in a function.
+See Figure~\ref{figDE} for the plot.
+
+We continue to analyze the data set using type tw1 to compare the two
+conditions. Now, it should be straightfoward to perform the single-factor
+analysis. Figures \ref{figFittw1only} and \ref{figDEtw1only} shows the
+dispersion distribution and MA-plot for the contrast ``glucose'' versus
+``galactose'' using type tw1 only. 
+<<tw1onlySamples>>=
+tw1only <- newSmutans( smutansGenes )
+tw1only <- smutans.de2type( tw1only, type="tw1", 
+                            condA="glucose", condB="galactose" )
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figFittw1only}
+\caption{Empirical (black dots) and fitted (red lines) dispersion values
+plotted against mean expression strength.}
+\label{figFittw1only}
+\end{figure}
+<<figFittw1only,fig=TRUE>>=
+smutans.plotDispersionEstimates(tw1only)
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figDEtw1only}
+\caption{Plot of normalised mean versus $log_2$ fold change
+(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
+versus ``galactose''.}
+\label{figDEtw1only}
+\end{figure}
+<<figDEtw1only,fig=TRUE>>=
+smutans.plotDiffExp(tw1only)
+@
+
+Note that \Rfunction{smutans.de2} already performed the negative bionomial test
+for each of the genes. We fetch adjusted p-values from the single-factor
+analysis and the two-factor one to compare the number of genes that are found to
+be differentially expressed using the same FDR rate of 0.1.
+<<compareUA159onlyAndTW1only>>=
+padj.ua159only <- smutans.padj( ua159only )
+padj.tw1only <- smutans.padj( tw1only )
+tabUA159onlyTW1only = table( "ua159 only" = padj.ua159only < .1, "tw1 only" = padj.tw1only < .1 )
+addmargins( tabUA159onlyTW1only )
+@
+Of the \Sexpr{sum(tabUA159onlyTW1only)} gene, \Sexpr{tabUA159onlyTW1only[2,2]}
+genes are found in both analyses. The analysis using all of tw1 only found
+\Sexpr{tabUA159onlyTW1only[1,2]} genes. The analysis using only ua159 found
+\Sexpr{tabUA159onlyTW1only[2,1]} genes.
+
+Until now we focused on the effect of type not condition. 
+<<glucoseOnlySamples>>=
+glucoseOnly <- newSmutans( smutansGenes, title="glucoseOnly" )
+glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
+                            condA="ua159", condB="tw1" )
+galactoseOnly <- newSmutans( smutansGenes, title="galactoseOnly" )
+galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
+                              condA="ua159", condB="tw1" )
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figFitGlucoseOnly}
+\caption{Empirical (black dots) and fitted (red lines) dispersion values
+plotted against mean expression strength. \textbf{figFitGlucoseOnly}}
+\label{figFitGlucoseOnly}
+\end{figure}
+<<figFitGlucoseOnly,fig=TRUE>>=
+smutans.plotDispersionEstimates(glucoseOnly)
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figDEGlucoseOnly}
+\caption{Plot of normalised mean versus $log_2$ fold change
+(this plot is sometimes also called the ``MA-plot'') for the contrast ``ua159''
+versus ``tw1''. \textbf{figDEGlucoseOnly}}
+\label{figDEGlucoseOnly}
+\end{figure}
+<<figDEGlucoseOnly,fig=TRUE>>=
+smutans.plotDiffExp(glucoseOnly)
+@
+
+Note that \Rfunction{smutans.de2} already performed the negative bionomial test
+for each of the genes. We fetch adjusted p-values from the single-factor
+analysis and the two-factor one to compare the number of genes that are found to
+be differentially expressed using the same FDR rate of 0.1.
+
+\fixme{Two objects and one qvalue cutoff can be handled by a function. Let's
+make a function in core.R.}
+<<compareGlucoseOnlyAndGalactoseOnly>>=
+#padj.glucoseOnly <- smutans.padj( glucoseOnly )
+#padj.galactoseOnly <- smutans.padj( galactoseOnly )
+#tabGlucoseOnlyGalactoseOnly = table( "glucose only" = padj.glucoseOnly < .1, 
+#                                     "galactose only" = padj.galactoseOnly < .1 )
+tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
+                                                      galactoseOnly )
+addmargins( tabGlucoseOnlyGalactoseOnly )
+@
+Of the \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} gene, \Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]}
+genes are found in both analyses. The analysis using all of galactose only found
+\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} genes. The analysis using glucose only found
+\Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} genes.
 
 %------------------------------------------------------------
-\section{Inference: Calling differential expression} \label{sec:DE}
+\subsection{Inference: Using all of the samples for condition effect} 
+\label{subsec:allSample}
 %------------------------------------------------------------
 
+We must have at least one citation \cite{Anders:2010:GB}.
+
+We could compare the two conditions, glucose and galactose, using either type
+ua159 or tw1. This may or may not be desirable depending on the experimental
+design. Using more samples might be more powerful in detecting differentially
+expressed genes. Because we have a two-factor data set, \Robject{smutansGenes},
+we wish to determine which genes are differentially expressed in one of the two
+factors. Here, we first find genes differentially expressed in different
+conditions of glucose and galactose using both types of ua159 and tw1. We will
+compare those genes and those found in the single type analysis that we did
+previously. Let's start the analysis by creating a new \Rclass{Smutans} object,
+called \Robject{allSample}. 
 <<allSamples>>=
@@
 @
+\Rfunction{smutans.de2} is the basic function that uses a two-factor data set to
+perform two-factor analysis. The two factors are type and condition. We are
+interested in condition effect. We first check 
+the dispersion estimates and to make sure that the data quality is as
+expected, we plot the per-gene estimates against the normalized mean
+expressions per gene, and then overlay the fitted curve in red
+(Fig.\ \ref{figFit2}).
 
@@
 plotted against mean expression strength.}
-\label{figFit}
+\label{figFit2}
 \end{figure}
-
-Calling the function produces the plot (Fig.\ \ref{figFit2}).
 
@@
 
+\fixme{Say something about Figure \ref{figFit2}.}
+
+Note that \Rfunction{smutans.de2} already performed the negative bionomial test
+for each of the genes. We fetch adjusted p-values from the single-factor
+analysis and the two-factor one to compare the number of genes that are found to
+be differentially expressed using the same FDR rate of 0.1.
 <<compareUA159onlyAndAllSample>>=
@@
 of the sample found \Sexpr{tab[1,2]} genes. The analysis using only ua159 found \Sexpr{tab[2,1]} genes. 
+Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from the
+two-factor analysis versus those from the single-factor analysis. The p-values
+from the two-factor analysis are overall smaller than those from the
+single-factor analysis. We would expect that this was the case because of higher
+power of the two-factor analysis.
 
@@
 @
+
+\fixme{Check the fitted regression using type and condition. The following are
+from \Rpackage{DESeq} document.} 
+The first three columns show the fitted coefficients, converted to a logarithm base 2 scale.
+The log2 fold change due to the condition is shown in the third column. As indicated
+by the column name, it is the effect of ``untreated'', i.e., the log ratio of
+``untreated'' versus ``treated''. (This is unfortunately the other way round as
+before, due to the peculiarities of contrast coding.) Note that the library type also
+had noticeable influence on the expression, and hence would have increased the dispersion
+estimates (and so reduced power) if we had not fitted an effect for it.
+The column \emph{deviance}
+is the deviance of the fit. (Comparing the deviances with a $\chi^2$ likelihood ratio test is
+how \Rfunction{nbinomGLMTest} calculates the $p$ values.) The last column, \emph{converged},
+indicates whether the calculation of coefficients and deviance has fully converged.
+(If it is false too often, you can try to change the GLM control parameters, as
+explained in the help page to \Rfunction{fitNbinomGLMs}.)
+
+Finally, we show that taking the library type into account was important to have
+good detection power by doing the analysis again using the standard workflow,
+as outlined earlier, and without informing \Rpackage{DESeq} of the library types:
 
@@
 
+We now proceed to compare the two types by considering all of the samples.
+<<allSampleType>>=
+allSampleType <- newSmutans( smutansGenes, title="All samples for type" )
+allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
+smutans.plotDispersionEstimates(allSampleType)
+tabGlucoseOnlyAllSample <- smutans.compareResult( glucoseOnly, 
+                                                  allSampleType )
+addmargins( tabGlucoseOnlyAllSample )
+@
+Of the \Sexpr{sum(tabGlucoseOnlyAllSample)} gene,
+\Sexpr{tabGlucoseOnlyAllSample[2,2]} genes are found in both analyses. The
+analysis using all of the sample found \Sexpr{tabGlucoseOnlyAllSample[1,2]}
+genes. The analysis using all of the same sample but without considering type
+found \Sexpr{tabGlucoseOnlyAllSample[2,1]} genes.
+
+%------------------------------------------------------------
+\subsection{Heatmap analyses} 
+\label{subsec:heatmap}
+%------------------------------------------------------------
+
+Potentially useful figures can be plotted from \Rpackage{DESeq}. Instead of
+browsing differentially expressed genes, they can be plotted in group. The
+original fold changes could be moderated to be amenable to distance calculation. 
+Figure \ref{figLfc} shows moderated log fold changes versus the original log
+fold changes.
+
 \begin{figure}
@@
 
+Now we plot the first \Sexpr{length(select)} genes clustered based on the
+moderated log fold changes.
+
 \begin{figure}
@@
 
+We can cluster samples using variance stabilized data. Figure \ref{figClust}
+shows the 13 sampled clustered. 
+The clustering correctly reflects our experimental design, i.e., samples are
+more similar when they have the same treatment or the same library type.
+(To make this conclusion, it was important to re-estimate e dispersion with
+mode ``blind'' (in the calculation for \Robject{cdsFullBlind} above, as only then,
+the variance stabilizing transformation is not informed about the design and
+hence not biased towards a result supporting it.) Such an analysis is useful
+for quality control, and (even though we finish our vignette with it), it may
+be useful to this first in any analysis.
+
 \begin{figure}
 \centering
-\includegraphics[width=.6\textwidth]{smutans-figClust}
+\includegraphics[width=.9\textwidth]{smutans-figClust}
 \caption{Heatmap showing the Euclidean distances between the samples
@@
 smutans.de2Clust( allSample )
+@
+
+%--------------------------------------------------
+\section{Differential expression analysis using 835NP, 835P, and UA159 no CSP}
+\label{sec:de835}
+%--------------------------------------------------
+
+<<sm835,results=hide>>=
+data( "sm835Genes" )
+@
+
+<<look_into_smutansGenes>>=
+head( counts(sm835Genes) )
+pData( sm835Genes )
+@
+
+<<ua159onlySamples>>=
+npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
+npAndP <- smutans.de2( npAndP, type="ua159", 
+                       condA="835np", condB="835p" )
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figFitNpAndP}
+\caption{Empirical (black dots) and fitted (red lines) dispersion values
+plotted against mean expression strength. \textbf{figFitNpAndP}}
+\label{figFitNpAndP}
+\end{figure}
+<<figFitNpAndP,fig=TRUE>>=
+smutans.plotDispersionEstimates(npAndP)
 @

### 2012-01-05 package/smutans/R/core.R

@@ -1,0 +1,81 @@
+smutans.compareResult <- function (resA, resB, qval=0.1)
+{
+  padj.resA <- smutans.padj( resA )
+  padj.resB <- smutans.padj( resB )
+  tab <- table( padj.resA < qval, 
+                padj.resB < qval,
+                dnn=c(smutans.title(resA), smutans.title(resB)) )
+  tab
+}
+
+smutans.prepareDataUA159TW1 <- function ()
+{
+  # Count data preparation
+  countsFile <- "smutans/inst/extdata/count.txt"
+  conds <- scan(file=paste(countsFile,"index",sep="."), what="character")
+  subconds <- conds %in% c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL")
+  countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)
+  rownames(countsTable) <- countsTable$gene
+  countsTable <- countsTable[,-1]
+  countsTable <- countsTable[,subconds]
+  conds <- conds[subconds]
+  conds.type <- conds %in% c("UA159GLU", "UA159GAL")
+  factor.type <- rep("ua159",length(conds))
+  factor.type[!conds.type] <- "tw1"
+  conds.condition <- conds %in% c("UA159GLU", "TW1GLU")
+  factor.condition <- rep("glucose",length(conds))
+  factor.condition[!conds.condition] <- "galactose"
+  samples <- data.frame(type=factor.type,condition=factor.condition)
+  rownames(samples) <- colnames(countsTable) 
+
+  design <- samples
+  smutansGenes <- newCountDataSet( countsTable, design )
+  expdata = new("MIAME", 
+     name="S. mutans UA159, TW1, Glucose, and Galactose", 
+     lab="University of Florida, and Cornell University", 
+     contact="Drs. Robert Burne, Michael Stanhope, and Adam Siepel", 
+     title="Streptococcus mutans RNA-Seq Studies", 
+     url="http://www.ncbi.nlm.nih.gov/projects/geo/query/acc.cgi?acc=XXX", 
+     abstract="RNA-seq of 13 biological replicates from Streptococcus mutans")
+  pubMedIds(expdata) <- "999999999"
+  experimentData(smutansGenes) <- expdata
+  save(smutansGenes, file=file.path(".", "smutansGenes.RData"))
+}
+
+smutans.prepareData835NPP <- function ()
+{
+  # Count data preparation
[…]

### 2012-01-05 package/smutans/R/methods.R

@@ -1,3 +1,20 @@
+setGeneric("smutans.title", function(object) standardGeneric("smutans.title"))
+setMethod("smutans.title",
+          "Smutans",
+  function( object ) {
+    object@title
+  }
+)
+
+setGeneric("smutans.title<-", function(object,value) standardGeneric("smutans.title<-"))
+setReplaceMethod("smutans.title", 
+                 "Smutans",
+  function( object, value ) {
+    object@title <- value
+    object
+  }
+)
+
 setGeneric("smutans.counts", function(object) standardGeneric("smutans.counts"))
 setMethod("smutans.counts",
           "Smutans",
@@ -11,23 +28,6 @@ setReplaceMethod("smutans.counts",
                  "Smutans",
   function( object, value ) {
     object@counts <- value
-    object
-  }
-)
-
-setGeneric("smutans.de2type", function(object,...) standardGeneric("smutans.de2type"))
-setMethod("smutans.de2type", 
-          "Smutans", 
-  function(object, type, condA, condB) {
-    ua159Samples <- pData(object@countDataSet)$type == type
-    countsTable <- counts(object@countDataSet)[ , ua159Samples ]
-    conds <- pData(object@countDataSet)$condition[ ua159Samples ]
-    cds <- newCountDataSet( countsTable, conds )
-    cds <- estimateSizeFactors( cds )
-    object@cds <- estimateDispersions( cds )
-    object@res <- nbinomTest( object@cds, condA, condB )
-    object@pval <- object@res$pval
-    object@padj <- object@res$padj
     object
   }
 )
@@ -82,19 +82,66 @@ setMethod("smutans.plotDiffExp",
   }
 )
 
-setGeneric("smutans.de2", function(object) standardGeneric("smutans.de2"))
+setGeneric("smutans.de2type", function(object,...) standardGeneric("smutans.de2type"))
+setMethod("smutans.de2type", 
+          "Smutans", 
+  function(object, type="", condition="", condA, condB) {
+    ua159Samples <- pData(object@countDataSet)$type == type
+    countsTable <- counts(object@countDataSet)[ , ua159Samples ]
+    conds <- pData(object@countDataSet)$condition[ ua159Samples ]
+    cds <- newCountDataSet( countsTable, conds )
+    cds <- estimateSizeFactors( cds )
+    object@cds <- estimateDispersions( cds )
[…]

### 2012-01-05 package/test.R

@@ -1,9 +1,23 @@
+#library(DESeq)
+#source("smutans/R/core.R")
+#smutans.prepareData835NPP ()
+#q("no")
 # library(DESeq)
-# load("smutans/data/smutansGenes.RData")
+load("smutans/data/sm835Genes.RData")
 source("smutans/R/class_and_slots.R")
+source("smutans/R/core.R")
 source("smutans/R/methods.R")
+
+head( counts(sm835Genes) )
+pData( sm835Genes )
+
+npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
+npAndP <- smutans.de2( npAndP, type="ua159", 
+                       condA="835np", condB="835p" )
+smutans.plotDispersionEstimates(npAndP)
+
 #ua159only <- smutans.de2type( ua159only, type="ua159", 
-#                              condA="glucose", condB="galactose" )
+                              #condA="glucose", condB="galactose" )
 #allSample <- smutans.de2( allSample )
 #
 #padj.ua159only <- smutans.padj( ua159only )
@@ -21,13 +35,13 @@ source("smutans/R/methods.R")
 #allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
 #                                                condA="glucose", 
 #                                                condB="galactose" )
-padj.allSampleWithoutType <- smutans.padj( allSampleWithoutType )
-tab2 <- table(
-   `all samples simple` = padj.allSampleWithoutType < 0.1,
-   `all samples GLM`    = padj.allSample < 0.1 )
-addmargins( tab2 )
+#padj.allSampleWithoutType <- smutans.padj( allSampleWithoutType )
+#tab2 <- table(
+   #`all samples simple` = padj.allSampleWithoutType < 0.1,
+   #`all samples GLM`    = padj.allSample < 0.1 )
+#addmargins( tab2 )
 
 # smutans.de2TypeHeatmap( ua159only )
 # smutans.de2TypeLfc( ua159only, condA="glucose", condB="galactose" )
-smutans.de2Clust( allSample )
+#smutans.de2Clust( allSample )
 

### 2012-01-06 package/smutans/inst/doc/smutans.Rnw
A smutans R package version of 0.1.2 is used. I still need to work on UA159/TW1 comparisons.

@@
 smutans.prepareDataUA159TW1( ) 
+smutans.prepareGoseq( )
 q("no")
@@
 
-We must have at least one citation \cite{Anders:2010:GB}.
+% We must have at least one citation \cite{Anders:2010:GB}.
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=.9\textwidth]{smutans-figClust}
-\caption{Heatmap showing the Euclidean distances between the samples
-  as calculated from the variance-stabilising transformation of the
-  count data.}
-\label{figClust}
-\end{figure}
+%\begin{figure}
+%\centering
+%\includegraphics[width=.9\textwidth]{smutans-figClust}
+%\caption{Heatmap showing the Euclidean distances between the samples
+%  as calculated from the variance-stabilising transformation of the
+%  count data.}
+%\label{figClust}
+%\end{figure}
 
 <<figClust,fig=TRUE>>=
-smutans.de2Clust( allSample )
+# smutans.de2Clust( allSample )
+@
+
+We find some functional categories enriched with differentially expressed genes.
+<<goseqUA159only>>=
+data( "smutans.feature.genes" )
+data( "smutans.go.genes" )
+data( "smutans.cat.desc" )
+smutans.de2Goseq ( ua159only, qval=1e-2, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
 @
@@
 %--------------------------------------------------
+
+A count data set is a table of numbers of short reads that are assigned to
+predefined genes. We use one of R packages for testing for differential
+expression: \Rpackage{DESeq}.
+
+To find functional categories enriched with differentially expressed genes we
+first find a list of differentially expressed genes. A set of genes are assigned
+to functional gene categories: \Robject{smutans.go.genes}. Features of the genes in
+UA159 include gene lengths: \Robject{smutans.feature.genes}.
+Functional categories with numbers of genes assigned are in
+\Robject{smutans.cat.desc}.
+
+We use RNA-seq data to define genes whose expression changes between
+samples \cite{Oshlack2010}. Short reads are first mapped to a
+reference genome using approaches such as BWA \cite{Li2009a}.
+Reads fitting into one of the annotated intervals are counted.
+These counts are passed to a statistical package which treats differences
+in read counts for statistical significance. For this purpose, we made
+use of DEseq \cite{Anders2010}. This prodecure returns a list of
+differentially expressed genes, which can be passed to downstream
+analysis, such as functionaly category enrichment test (e.g., Young et
+al., 2010).
+
+The abtract briefly describes main steps of analyzing RNA-seq data for
+determining differentially expressed genes \cite{Oshlack2010}. We
+used a short read aligner called BWA \cite{Li2009a} to map the
+sequenced short reads on a reference genome, \textit{Streptococcus mutans}
+UA159 (NCBI accession NC\_004350). Because the aligner allowed a few gaps
+for efficient alignment of millions of sequences of size being 100 base
+pairs long, short reads that consist of RNAs and sequencing adapter
+sequences would be not mapped. We used cutadapt \cite{Martin2011} to remove
+parts of adapter sequences from short reads before mapping on the reference
+genome. Mapped short read alignments were converted into readable formats
+\cite{Li2009b}. We counted short reads aligned to annotated genes in the
+reference genome, forming a table of read counts of the genes. Statistical
+software R \cite{R2011} with its one of R package
+called DEseq \cite{Anders2010} were subsequently employed to infer
+significantly differentially expressed in a pair of biological conditions.
+We also generated new tracks for our recently released Streptococcus Genome
+Browser (http://strep-genome.bscb.cornell.edu) that summarize gene
+expression levels based the short read alignments. This allowed us to
+visualize parts of the reference genome for comparing different RNA
+samples. Further analyses such as functionaly category enrichment test
+\cite{Young2010} were pursued.
+
+We could find no functional categories enriched with differentially expressed
+genes in the pairwise comparisons of 835NP, 835P, and UA159 no CSP.
+
+Let's make a list of differentially expressed genes. It could be interesting to
+see the cluster of the three kinds of samples. Another kind of heatmap plots for
+grouping genes could be interesting. But, basically we have a list of genes with
+some adjusted p-values for detecting differentially expressed genes.
 
@@
 
-<<ua159onlySamples>>=
+<<sm835Samples>>=
 npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
@@
                        condA="835np", condB="835p" )
+npAndUA159 <- newSmutans( sm835Genes, title="835NP vs. UA159" )
+npAndUA159 <- smutans.de2( npAndUA159, type="ua159", 
+                           condA="835np", condB="ua159nocsp" )
+pAndUA159 <- newSmutans( sm835Genes, title="835P vs. UA159" )
+pAndUA159 <- smutans.de2( pAndUA159, type="ua159", 
+                          condA="835p", condB="ua159nocsp" )
 @
@@
 \includegraphics[width=.6\textwidth]{smutans-figFitNpAndP}
+\includegraphics[width=.6\textwidth]{smutans-figFitNpAndUA159}
+\includegraphics[width=.6\textwidth]{smutans-figFitPAndUA159}
 \caption{Empirical (black dots) and fitted (red lines) dispersion values
@@
 smutans.plotDispersionEstimates(npAndP)
+@
+<<figFitNpAndUA159,fig=TRUE>>=
+smutans.plotDispersionEstimates(npAndUA159)
+@
+<<figFitPAndUA159,fig=TRUE>>=
+smutans.plotDispersionEstimates(pAndUA159)
+@
+
+<<sm835Samples>>=
+smutans.de2List( npAndP, "npAndP.csv" )
+smutans.de2List( npAndUA159, "npAndUA159.csv" )
+smutans.de2List( pAndUA159, "pAndUA159.csv" )
+@
+
+%\begin{figure}
+%\centering
+%\includegraphics[width=.9\textwidth]{smutans-figClustNpAndP}
+%\caption{Empirical (black dots) and fitted (red lines) dispersion values
+%plotted against mean expression strength. \textbf{figClustNpAndP}}
+%\label{figClustNpAndP}
+%\end{figure}
+<<figClustNpAndP,fig=TRUE>>=
+# smutans.de2Clust( npAndP )
+@
+%\begin{figure}
+%\centering
+%\includegraphics[width=.9\textwidth]{smutans-figClustNpAndUA159}
+%\caption{Empirical (black dots) and fitted (red lines) dispersion values
+%plotted against mean expression strength. \textbf{figClustNpAndUA159}}
+%\label{figClustNpAndUA159}
+%\end{figure}
+<<figClustNpAndUA159,fig=TRUE>>=
+# smutans.de2Clust( npAndUA159 )
+@
+%\begin{figure}
+%\centering
+%\includegraphics[width=.9\textwidth]{smutans-figClustPAndUA159}
+%\caption{Empirical (black dots) and fitted (red lines) dispersion values
+%plotted against mean expression strength. \textbf{figClustPAndUA159}}
+%\label{figClustPAndUA159}
+%\end{figure}
+<<figClustPAndUA159,fig=TRUE>>=
+# smutans.de2Clust( pAndUA159 )
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{smutans-figHeatmapNpAndP}
+\caption{Empirical (black dots) and fitted (red lines) dispersion values
+plotted against mean expression strength. \textbf{figHeatmapNpAndP}}
+\label{figHeatmapNpAndP}
+\end{figure}
+<<figHeatmapNpAndP,fig=TRUE>>=
+smutans.de2TypeHeatmap( npAndP )
+@
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figHeatmapNpAndUA159}
+\caption{Empirical (black dots) and fitted (red lines) dispersion values
+plotted against mean expression strength. \textbf{figHeatmapNpAndUA159}}
+\label{figHeatmapNpAndUA159}
+\end{figure}
+<<figHeatmapNpAndUA159,fig=TRUE>>=
+smutans.de2TypeHeatmap( npAndUA159 )
+@
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figHeatmapPAndUA159}
+\caption{Empirical (black dots) and fitted (red lines) dispersion values
+plotted against mean expression strength. \textbf{figHeatmapPAndUA159}}
+\label{figHeatmapPAndUA159}
+\end{figure}
+<<figHeatmapPAndUA159,fig=TRUE>>=
+smutans.de2TypeHeatmap( pAndUA159 )
 @
@@
 \bibliographystyle{unsrt}
-\bibliography{smutans}
+\bibliography{siepel-mutans}
 

### 2012-01-06 package/smutans/R/class_and_slots.R

@@ -2,7 +2,7 @@ setClass( "Smutans",
   representation = representation( 
     countDataSet = "CountDataSet", 
     cds = "ANY", 
-    res = "list", 
+    res = "ANY", 
     padj = "numeric",
     pval = "numeric",
     title = "character"

### 2012-01-06 package/smutans/R/core.R

@@ -79,3 +79,15 @@ smutans.prepareData835NPP <- function ()
   save(sm835Genes, file=file.path("smutans", "data", "sm835Genes.RData"))
 }
 
+smutans.prepareGoseq <- function ()
+{
+  f <- "smutans/inst/extdata/feature-genome.out-geneonly"
+  smutans.feature.genes = read.table(file=f,head=F)
+  save(smutans.feature.genes, file=file.path("smutans", "data", "smutans.feature.genes.RData"))
+  f <- "smutans/inst/extdata/smutans.gene2go"
+  smutans.go.genes <- read.table(file=f,head=F)
+  save(smutans.go.genes, file=file.path("smutans", "data", "smutans.go.genes.RData"))
+  f <- "smutans/inst/extdata/smutans.go2ngene"
+  smutans.cat.desc = read.table(file=f,head=F,sep="\t",quote="")
+  save(smutans.cat.desc, file=file.path("smutans", "data", "smutans.cat.desc.RData"))
+}

### 2012-01-06 package/smutans/R/methods.R

@@ -221,18 +221,21 @@ setGeneric("smutans.de2TypeHeatmap",
 )
 setMethod("smutans.de2TypeHeatmap", 
           "Smutans", 
-  function(object, file="default") {
+  function(object, file="default", nselect=40) {
     cdsBlind <- estimateDispersions( object@cds, method="blind" )
     vsd <- getVarianceStabilizedData( cdsBlind )
-    select <- order(object@res$pval)[1:40]
-    colors <- colorRampPalette(c("white","darkblue"))(100)
+    select <- order(object@res$pval)[1:nselect]
+    colors <- colorRampPalette(c("white","black"))(100)
+    # colors <- colorRampPalette(brewer.pal(9,"Blues"))(100)
 
     if (file != "default")
     {
       postscript(file, width=6, height=6, horizontal = FALSE, onefile = FALSE, paper = "special")
       oldpar <- par (mar=c(4.5, 4.5, 0.5, 0.5))
     }
-    heatmap( vsd[select,], col = colors, scale = "none")
+    # heatmap( vsd[select,], col = colors, scale = "none")
+    heatmap.2( vsd[select,], col = colors, scale = "none", trace="none",
+               density.info="none" ) 
     if (file != "default")
     {
       par(oldpar)
@@ -246,7 +249,7 @@ setGeneric("smutans.de2Clust",
 )
 setMethod("smutans.de2Clust", 
           "Smutans", 
-  function(object, file="default") {
+  function(object, file="default", margins=c(5,5)) {
     if (file != "default")
     {
       postscript(file, width=6, height=6, horizontal = FALSE, onefile = FALSE, paper = "special")
@@ -255,10 +258,15 @@ setMethod("smutans.de2Clust",
     cdsFullBlind <- estimateDispersions( object@cds, method = "blind" )
     vsdFull <- getVarianceStabilizedData( cdsFullBlind )
     dists <- dist( t( vsdFull ) )
-    heatmap( as.matrix( dists ),
-       symm=TRUE, scale="none", margins=c(10,10),
-       col = colorRampPalette(c("darkblue","white"))(100),
-       labRow = paste( pData(cdsFullBlind)$condition, pData(cdsFullBlind)$type ) )
+    heatmap.2( as.matrix( dists ), Colv=TRUE,
+       symm=TRUE, scale="none", margins=margins,
+       col = colorRampPalette(c("black","white"))(100),
[…]

### 2012-01-06 package/test.R

@@ -1,20 +1,105 @@
+sm835result <- function () {
+  library(smutans)
+  data( "sm835Genes" )
+  data( "smutans.feature.genes" )
+  data( "smutans.go.genes" )
+  data( "smutans.cat.desc" )
+  npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
+  npAndP <- smutans.de2( npAndP, type="ua159", 
+                         condA="835np", condB="835p" )
+  npAndUA159 <- newSmutans( sm835Genes, title="835NP vs. UA159" )
+  npAndUA159 <- smutans.de2( npAndUA159, type="ua159", 
+                             condA="835np", condB="ua159nocsp" )
+  pAndUA159 <- newSmutans( sm835Genes, title="835P vs. UA159" )
+  pAndUA159 <- smutans.de2( pAndUA159, type="ua159", 
+                            condA="835p", condB="ua159nocsp" )
+  smutans.de2List( npAndP, "npAndP.csv" )
+  smutans.de2List( npAndUA159, "npAndUA159.csv" )
+  smutans.de2List( pAndUA159, "pAndUA159.csv" )
+  smutans.de2Clust( npAndP )
+  # smutans.de2Clust( npAndP, margins=c(7,7) ) this is for v0.1.3
+}
+# smutans.de2Clust( npAndUA159 )
+# smutans.de2Clust( pAndUA159 )
+# smutans.de2TypeHeatmap( npAndP )
+# smutans.de2TypeHeatmap( npAndUA159 )
+# smutans.de2TypeHeatmap( pAndUA159 )
+#smutans.de2Goseq ( npAndP, qval=1e-2, 
+#                   feature.gene=smutans.feature.genes,
+#                   go.genes=smutans.go.genes,
+#                   cat.desc=smutans.cat.desc )
+
+
 #library(DESeq)
 #source("smutans/R/core.R")
+#smutans.prepareGoseq( )
 #smutans.prepareData835NPP ()
 #q("no")
-# library(DESeq)
-load("smutans/data/sm835Genes.RData")
+#library(DESeq)
+#library(goseq)
+#load("smutans/data/smutansGenes.RData")
+#load("smutans/data/sm835Genes.RData")
 source("smutans/R/class_and_slots.R")
 source("smutans/R/core.R")
 source("smutans/R/methods.R")
 
-head( counts(sm835Genes) )
-pData( sm835Genes )
+#head( counts(sm835Genes) )
+#pData( sm835Genes )
 
[…]

### 2012-01-06 package/smutans/inst/doc/smutans.Rnw
Just an update.

@@
 Tasks include the analysis of count data from RNA-Seq for
-the detection of differentially expressed genes.  Tests of enrichment for
-differentially expressed genes are performed using \Rpackage{goseq}.  The
-package \Rpackage{smutans} contains the procedures.  This vignette explains the
-use of the package. 
+the detection of differentially expressed genes, and tests of enrichment for
+differentially expressed genes using \Rpackage{goseq}.  The
+package \Rpackage{smutans} contains procedures that make use of functions
+provided by the above R packages.  
 \end{abstract}
@@
 the directory smutans/data: sm835Genes.RData, smutansGenes.RData.
+
+http://www.oralgen.lanl.gov/oralgen/downloads/smut.csv
+
+http://www.oralgen.lanl.gov/cgi-bin/gene_id_search.cgi?dbname=smut&gene_id=SMu0001
+
+\fixme{Where are flat files for genes of S. mutans UA159?}
+
 <<samplePreparation,eval=FALSE>>=
@@
 source("smutans/R/core.R")
+smutans.prepareData34( )
 smutans.prepareData835NPP( )
@@
 
+\fixme{OMZ175, UA159CSP, SMU86CSP pairwise comparisons should be done fairly
+easily after detailed analyses.}
+
 \section{Session Info}

### 2012-01-06 package/smutans/R/core.R

@@ -8,6 +8,43 @@ smutans.compareResult <- function (resA, resB, qval=0.1)
   tab
 }
 
+smutans.prepareData34 <- function ()
+{
+  # Count data preparation
+  countsFile <- "smutans/inst/extdata/count.txt"
+  conds <- scan(file=paste(countsFile,"index",sep="."), what="character")
+
+  # subconds <- conds %in% c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL")
+  countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)
+  rownames(countsTable) <- countsTable$gene
+  countsTable <- countsTable[,-1]
+  # countsTable <- countsTable[,subconds]
+
+  # conds <- conds[subconds]
+  # conds.type <- conds %in% c("UA159GLU", "UA159GAL")
+  factor.type <- rep("ua159",length(conds))
+  # factor.type[!conds.type] <- "tw1"
+  #conds.condition <- conds %in% c("UA159GLU", "TW1GLU")
+  #factor.condition <- rep("glucose",length(conds))
+  factor.condition <- conds
+
+  samples <- data.frame(type=factor.type,condition=factor.condition)
+  rownames(samples) <- colnames(countsTable) 
+
+  design <- samples
+  sm34Genes <- newCountDataSet( countsTable, design )
+  expdata = new("MIAME", 
+     name="S. mutans UA159, TW1, Glucose, and Galactose", 
+     lab="University of Florida, and Cornell University", 
+     contact="Drs. Robert Burne, Michael Stanhope, and Adam Siepel", 
+     title="Streptococcus mutans RNA-Seq Studies", 
+     url="http://www.ncbi.nlm.nih.gov/projects/geo/query/acc.cgi?acc=XXX", 
+     abstract="RNA-seq of 34 biological replicates from Streptococcus mutans")
+  pubMedIds(expdata) <- "999999999"
+  experimentData(sm34Genes) <- expdata
+  save(sm34Genes, file=file.path("smutans", "data", "sm34Genes.RData"))
+}
+
 smutans.prepareDataUA159TW1 <- function ()
 {
   # Count data preparation
@@ -91,3 +128,23 @@ smutans.prepareGoseq <- function ()
[…]

### 2012-01-06 package/test.R

@@ -34,14 +34,19 @@ sm835result <- function () {
 #source("smutans/R/core.R")
 #smutans.prepareGoseq( )
 #smutans.prepareData835NPP ()
+#smutans.prepareData34 ()
 #q("no")
 #library(DESeq)
 #library(goseq)
 #load("smutans/data/smutansGenes.RData")
 #load("smutans/data/sm835Genes.RData")
+#load("smutans/data/sm34Genes.RData")
 source("smutans/R/class_and_slots.R")
 source("smutans/R/core.R")
 source("smutans/R/methods.R")
+npAndP <- newSmutans( sm34Genes, title="835NP vs. 835P" )
+npAndP <- smutans.de2( npAndP, type="ua159", 
+                       condA="835NP", condB="835P" )
 
 #head( counts(sm835Genes) )
 #pData( sm835Genes )

### 2012-01-10 package/smutans/inst/doc/smutans.Rnw
Just an update.

--- 2012-01-06 package/smutans/inst/doc/smutans.Rnw
@@
-%\VignetteIndexEntry{Analyzing Streptococcus mutans RNA-Seq data}
+%\VignetteIndexEntry{Analyzing Streptococcus mutans RNA-seq data}
 %\VignettePackage{smutans}
@@
 \usepackage{color}
+\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
 
 \SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=4,height=4.5}
-\newcommand{\Robject}[1]{\texttt{#1}}
-\newcommand{\Rpackage}[1]{\textit{#1}}
-\newcommand{\Rclass}[1]{\textit{#1}}
-\newcommand{\Rfunction}[1]{{\small\texttt{#1}}}
+\newcommand{\Robject}[1]{object \texttt{#1}}
+\newcommand{\Rpackage}[1]{R package \textit{#1}}
+\newcommand{\Rclass}[1]{class \textit{#1}}
+\newcommand{\Rfunction}[1]{{function \small\texttt{#1}}}
 \newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}
@@
 
-\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-Seq data using R packages}}}
+\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data using R packages}}}
 
@@
 \begin{abstract}
-We analyze a number of RNA samples from \textit{Streptococcus mutans} using R
-packages: \Rpackage{DESeq}, \Rpackage{edgeR}, and \Rpackage{goseq}.
-Tasks include the analysis of count data from RNA-Seq for
-the detection of differentially expressed genes.  Tests of enrichment for
-differentially expressed genes are performed using \Rpackage{goseq}.  The
-package \Rpackage{smutans} contains the procedures.  This vignette explains the
-use of the package. 
+We analyze a number of RNA samples from \textit{Streptococcus mutans}.
+
+using 
+\Rpackage{DESeq}, \Rpackage{edgeR}, and \Rpackage{goseq}.
+Tasks include the analysis of count data from RNA-seq for
+detection of differentially expressed genes, and tests of enrichment for
+differentially expressed genes using \Rpackage{goseq}.
+
+We use RNA-seq data to define genes whose expression changes between
+samples \cite{Oshlack2010}. Short reads are first mapped to a
+reference genome using approaches such as BWA \cite{Li2009a}.
+Reads fitting into one of the annotated intervals are counted.
+These counts are passed to a statistical package which treats differences
+in read counts for statistical significance. For this purpose, we made
+use of \Rpackage{DEseq} \cite{Anders2010}. This prodecure returns a list of
+differentially expressed genes, which can be passed to downstream
+analysis, such as functional category enrichment test using 
+\Rpackage{goseq} \cite{Young2010}.
+
+This
+\Rpackage{smutans} contains procedures that make use of functions
+provided by the above R packages.  
 \end{abstract}
@@
 %--------------------------------------------------
-
-The \Rpackage{smutans} package expects count data in the form of
-\Rclass{CountDataSet}. Using BWA and samtools we aligned short reads from RNA
-sequencing. Using gene annotations in the UA159 genome we counted short reads
-that overlapped the genes.  The count values are raw counts of sequencing reads.
-\Rpackage{DESeq}'s statistical model, or negative binomial model, hold for raw
-reads that allows to assess the measurement precision correctly.  We did not
-have rounded values of normalized counts, or counts of covered base pairs.
-
-Several exeriments of \textit{S.\ mutans} RNA sampling were performed.
-\fixme{Find the papers that are related with the experiments.} One of them was
-presented in the object called \Robject{smutansGenes}.  The
-\Robject{smutansGenes} object is of class \Rclass{CountDataSet}, which is the
-data container used by \Rpackage{DESeq}. We have a covenient class called
-\Rclass{Smutans}. The class contains slots for \Rclass{CountDataSet} objects,
-and results from analyses.  Let's load the needed packages and the data as
-follows.
-
-We can create \Rclass{CountDataSet} objects using functions that are not
-exported. One should create an R script at the directory that contains the
-source code of the \Rpackage{smutans}. Put the following scripts in the R script
-and run the script at the directory. That should create RData files at
-the directory smutans/data: sm835Genes.RData, smutansGenes.RData.
-<<samplePreparation,eval=FALSE>>=
-#not run
-library(DESeq)
-source("smutans/R/core.R")
-smutans.prepareData835NPP( )
-smutans.prepareDataUA159TW1( ) 
-smutans.prepareGoseq( )
-q("no")
-@
-
-<<options,results=hide,echo=FALSE>>=
-options(digits=3)
-@
-
-<<smutans2,results=hide>>=
-library( "smutans" )
-data( "smutansGenes" )
-@
-
-There are \Sexpr{length(colnames(counts(smutansGenes)))} RNA-seq samples in
-\Robject{smutansGenes} that can be loaded from the package.
-\Robject{smutansGenes} contains the counts and also metadata about the samples:
-
-<<look_into_smutansGenes>>=
-head( counts(smutansGenes) )
-@
-
-<<look_into_smutansGenes>>=
-pData( smutansGenes )
-@
-
-The samples differ by experimental condition (glucose or
-galactose) and by library type (ua159 or tw1). First, we only look at the ua159
-data for now. We will look at the tw1 data later. 
-We create an \Rclass{Smutans} object using the \Rclass{CountDataSet} object,
-\Robject{smutansGenes}.
-
-%--------------------------------------------------
-\section{Differential expression analysis} \label{sec:de}
-%--------------------------------------------------
-
-%------------------------------------------------------------
-\subsection{Inference: Using the UA159 samples for condition effect} 
-\label{subsec:ua159only}
-%------------------------------------------------------------
-
-<<ua159onlySamples>>=
-ua159only <- newSmutans( smutansGenes )
-ua159only <- smutans.de2type( ua159only, type="ua159", 
-                              condA="glucose", condB="galactose" )
-@
-
-\Rfunction{smutans.de2type} uses a two-factor data set to test differential
-expression of genes. We compare 4 pairwise comparisons using a two-factor data
-set. A pair of comparisons between glucose and galactose given either type ua159
-or tw1. Another pairs of comparisons between types ua159 and tw1 given either
-condition glucose or galactose. We focus on one of the four pairwise comparisons
-first. Later, we will perform analyses for the rest of the three comparisons.
-Note that the \Rclass{Smutans} contains \Rclass{CountDataSet} objects.
-\Rfunction{smutans.de2type} estimates the effective library size, and the
-dispersions using functions featured by \Rpackage{DESeq}. Then, it tests for
-differential expression.
-
-Because \Rpackage{DESeq} recommended to check
-the dispersion estimates and to make sure that the data quality is as
-expected, we plot the per-gene estimates against the normalized mean
-expressions per gene, and then overlay the fitted curve in red
-(Fig.\ \ref{figFit}).
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figFit}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength.}
-\label{figFit}
-\end{figure}
-<<figFit,fig=TRUE>>=
-smutans.plotDispersionEstimates(ua159only)
-@
-
-As we estimated the dispersion from only two samples, we should expect the
-estimates to scatter with quite some sampling variance around their true values.
-Hence, \Rpackage{DESeq} should not use the per-gene estimates directly in the
-test, because using too low dispersion values leads to false positives.  Many of
-the values below the red line are likely to be underestimates of the true
-dispersions, and hence, it is prudent to instead rather use the fitted value.
-On the othe hand, not all of he values above the red line are overestimations,
-and hence, the conservative choice is to keep them instead of replacing them
-with their fitted values.
-
-Having estimated the dispersion for each gene, we look for differentially
-expressed genes. To contrast two conditions, e.g., to see whether there is
-differential expression between conditions ``glucose'' and ``galactose'',
-\Rfunction{smutans.de2type} already performed the function
-\Rfunction{nbinomTest} provided by \Rpackage{DESeq}.  Let us first plot the
-$\log_2$ fold changes against the base means, colouring in red those genes that
-are significant at 10\% FDR.
-
-For each gene, we get its mean expression level (at the base scale) as a joint
-estimate from both conditions, and estimated separately for each  condition, the
-fold change from the first to the second condition, the logarithm (to basis 2)
-of the fold change, and the $p$ value for the statistical significance of this
-change. The \texttt{padj} column contains the $p$ values, adjusted for multiple
-testing with the Benjamini-Hochberg procedure (see the R function
-\Rfunction{p.adjust}), which controls false discovery rate (FDR).
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figDE}
-\caption{Plot of normalised mean versus $log_2$ fold change
-(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
-versus ``galactose''.}
-\label{figDE}
-\end{figure}
-<<figDE,fig=TRUE>>=
-smutans.plotDiffExp(ua159only)
-@ %$
-See Figure~\ref{figDE} for the plot.
-
-We continue to analyze the data set using type tw1 to compare the two
-conditions. Now, it should be straightfoward to perform the single-factor
-analysis. Figures \ref{figFittw1only} and \ref{figDEtw1only} shows the
-dispersion distribution and MA-plot for the contrast ``glucose'' versus
-``galactose'' using type tw1 only. 
-<<tw1onlySamples>>=
-tw1only <- newSmutans( smutansGenes )
-tw1only <- smutans.de2type( tw1only, type="tw1", 
-                            condA="glucose", condB="galactose" )
-@
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figFittw1only}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength.}
-\label{figFittw1only}
-\end{figure}
-<<figFittw1only,fig=TRUE>>=
-smutans.plotDispersionEstimates(tw1only)
-@
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figDEtw1only}
-\caption{Plot of normalised mean versus $log_2$ fold change
-(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
-versus ``galactose''.}
-\label{figDEtw1only}
-\end{figure}
-<<figDEtw1only,fig=TRUE>>=
-smutans.plotDiffExp(tw1only)
-@
-
-Note that \Rfunction{smutans.de2} already performed the negative bionomial test
-for each of the genes. We fetch adjusted p-values from the single-factor
-analysis and the two-factor one to compare the number of genes that are found to
-be differentially expressed using the same FDR rate of 0.1.
-<<compareUA159onlyAndTW1only>>=
-padj.ua159only <- smutans.padj( ua159only )
-padj.tw1only <- smutans.padj( tw1only )
-tabUA159onlyTW1only = table( "ua159 only" = padj.ua159only < .1, "tw1 only" = padj.tw1only < .1 )
-addmargins( tabUA159onlyTW1only )
-@
-Of the \Sexpr{sum(tabUA159onlyTW1only)} gene, \Sexpr{tabUA159onlyTW1only[2,2]}
-genes are found in both analyses. The analysis using all of tw1 only found
-\Sexpr{tabUA159onlyTW1only[1,2]} genes. The analysis using only ua159 found
-\Sexpr{tabUA159onlyTW1only[2,1]} genes.
-
-Until now we focused on the effect of type not condition. 
-<<glucoseOnlySamples>>=
-glucoseOnly <- newSmutans( smutansGenes, title="glucoseOnly" )
-glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
-                            condA="ua159", condB="tw1" )
-galactoseOnly <- newSmutans( smutansGenes, title="galactoseOnly" )
-galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
-                              condA="ua159", condB="tw1" )
-@
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figFitGlucoseOnly}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength. \textbf{figFitGlucoseOnly}}
-\label{figFitGlucoseOnly}
-\end{figure}
-<<figFitGlucoseOnly,fig=TRUE>>=
-smutans.plotDispersionEstimates(glucoseOnly)
-@
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figDEGlucoseOnly}
-\caption{Plot of normalised mean versus $log_2$ fold change
-(this plot is sometimes also called the ``MA-plot'') for the contrast ``ua159''
-versus ``tw1''. \textbf{figDEGlucoseOnly}}
-\label{figDEGlucoseOnly}
-\end{figure}
-<<figDEGlucoseOnly,fig=TRUE>>=
-smutans.plotDiffExp(glucoseOnly)
-@
-
-Note that \Rfunction{smutans.de2} already performed the negative bionomial test
-for each of the genes. We fetch adjusted p-values from the single-factor
-analysis and the two-factor one to compare the number of genes that are found to
-be differentially expressed using the same FDR rate of 0.1.
-
-\fixme{Two objects and one qvalue cutoff can be handled by a function. Let's
-make a function in core.R.}
-<<compareGlucoseOnlyAndGalactoseOnly>>=
-#padj.glucoseOnly <- smutans.padj( glucoseOnly )
-#padj.galactoseOnly <- smutans.padj( galactoseOnly )
-#tabGlucoseOnlyGalactoseOnly = table( "glucose only" = padj.glucoseOnly < .1, 
-#                                     "galactose only" = padj.galactoseOnly < .1 )
-tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
-                                                      galactoseOnly )
-addmargins( tabGlucoseOnlyGalactoseOnly )
-@
-Of the \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} gene, \Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]}
-genes are found in both analyses. The analysis using all of galactose only found
-\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} genes. The analysis using glucose only found
-\Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} genes.
-
-%------------------------------------------------------------
-\subsection{Inference: Using all of the samples for condition effect} 
-\label{subsec:allSample}
-%------------------------------------------------------------
-
-% We must have at least one citation \cite{Anders:2010:GB}.
-
-We could compare the two conditions, glucose and galactose, using either type
-ua159 or tw1. This may or may not be desirable depending on the experimental
-design. Using more samples might be more powerful in detecting differentially
-expressed genes. Because we have a two-factor data set, \Robject{smutansGenes},
-we wish to determine which genes are differentially expressed in one of the two
-factors. Here, we first find genes differentially expressed in different
-conditions of glucose and galactose using both types of ua159 and tw1. We will
-compare those genes and those found in the single type analysis that we did
-previously. Let's start the analysis by creating a new \Rclass{Smutans} object,
-called \Robject{allSample}. 
-<<allSamples>>=
-allSample <- newSmutans( smutansGenes )
-allSample <- smutans.de2( allSample )
-@
-\Rfunction{smutans.de2} is the basic function that uses a two-factor data set to
-perform two-factor analysis. The two factors are type and condition. We are
-interested in condition effect. We first check 
-the dispersion estimates and to make sure that the data quality is as
-expected, we plot the per-gene estimates against the normalized mean
-expressions per gene, and then overlay the fitted curve in red
-(Fig.\ \ref{figFit2}).
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figFit2}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength.}
-\label{figFit2}
-\end{figure}
-
-<<figFit2,fig=TRUE>>=
-smutans.plotDispersionEstimates(allSample)
-@
-
-\fixme{Say something about Figure \ref{figFit2}.}
-
-Note that \Rfunction{smutans.de2} already performed the negative bionomial test
-for each of the genes. We fetch adjusted p-values from the single-factor
-analysis and the two-factor one to compare the number of genes that are found to
-be differentially expressed using the same FDR rate of 0.1.
-<<compareUA159onlyAndAllSample>>=
-padj.ua159only <- smutans.padj( ua159only )
-padj.allSample <- smutans.padj( allSample )
-tab = table( "ua159 only" = padj.ua159only < .1, "all samples" = padj.allSample < .1 )
-addmargins( tab )
-@
-Of the \Sexpr{sum(tab)} gene, \Sexpr{tab[2,2]} genes are found in both analyses. The analysis using all
-of the sample found \Sexpr{tab[1,2]} genes. The analysis using only ua159 found \Sexpr{tab[2,1]} genes. 
-Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from the
-two-factor analysis versus those from the single-factor analysis. The p-values
-from the two-factor analysis are overall smaller than those from the
-single-factor analysis. We would expect that this was the case because of higher
-power of the two-factor analysis.
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figCompareUA159onlyAndAllSample}
-\caption{Comparison of $p$ values (unadjusted) from the test using only the
-four paired-end samples and the test using all seven samples. We can see that
-the latter tend to be smaller, as expected from the higher power of the test with all samples.}
-\label{figCompareUA159onlyAndAllSample}
-\end{figure}
-
-<<figCompareUA159onlyAndAllSample,fig=TRUE>>=
-pval.ua159only <- smutans.pval( ua159only )
-pval.allSample <- smutans.pval( allSample )
-bottom = function(x, theta=1e-12) pmax(x, theta)
-plot( bottom(pval.ua159only), bottom(pval.allSample), log="xy", pch=20, cex=.3 )
-abline(a=0, b=1, col="blue")
-@
-
-\fixme{Check the fitted regression using type and condition. The following are
-from \Rpackage{DESeq} document.} 
-The first three columns show the fitted coefficients, converted to a logarithm base 2 scale.
-The log2 fold change due to the condition is shown in the third column. As indicated
-by the column name, it is the effect of ``untreated'', i.e., the log ratio of
-``untreated'' versus ``treated''. (This is unfortunately the other way round as
-before, due to the peculiarities of contrast coding.) Note that the library type also
-had noticeable influence on the expression, and hence would have increased the dispersion
-estimates (and so reduced power) if we had not fitted an effect for it.
-The column \emph{deviance}
-is the deviance of the fit. (Comparing the deviances with a $\chi^2$ likelihood ratio test is
-how \Rfunction{nbinomGLMTest} calculates the $p$ values.) The last column, \emph{converged},
-indicates whether the calculation of coefficients and deviance has fully converged.
-(If it is false too often, you can try to change the GLM control parameters, as
-explained in the help page to \Rfunction{fitNbinomGLMs}.)
-
-Finally, we show that taking the library type into account was important to have
-good detection power by doing the analysis again using the standard workflow,
-as outlined earlier, and without informing \Rpackage{DESeq} of the library types:
-
-<<allSampleWithoutType>>=
-allSampleWithoutType <- newSmutans( smutansGenes )
-allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
-                                                condA="glucose", 
-                                                condB="galactose" )
-@
-
-<<compareUA159onlyAndAllSample>>=
-padj.allSampleWithoutType <- smutans.padj( allSampleWithoutType )
-tab2 <- table(
-   `all samples simple` = padj.allSampleWithoutType < 0.1,
-   `all samples GLM`    = padj.allSample < 0.1 )
-addmargins( tab2 )
-largeNumber <- 10
-@
-Of the \Sexpr{sum(tab2)} gene, \Sexpr{tab2[2,2]} genes are found in both analyses. The analysis using all
-of the sample found \Sexpr{tab2[1,2]} genes. The analysis using all of the same
-sample but without considering type found \Sexpr{tab2[2,1]} genes.
-
-We now proceed to compare the two types by considering all of the samples.
-<<allSampleType>>=
-allSampleType <- newSmutans( smutansGenes, title="All samples for type" )
-allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
-smutans.plotDispersionEstimates(allSampleType)
-tabGlucoseOnlyAllSample <- smutans.compareResult( glucoseOnly, 
-                                                  allSampleType )
-addmargins( tabGlucoseOnlyAllSample )
-@
-Of the \Sexpr{sum(tabGlucoseOnlyAllSample)} gene,
-\Sexpr{tabGlucoseOnlyAllSample[2,2]} genes are found in both analyses. The
-analysis using all of the sample found \Sexpr{tabGlucoseOnlyAllSample[1,2]}
-genes. The analysis using all of the same sample but without considering type
-found \Sexpr{tabGlucoseOnlyAllSample[2,1]} genes.
-
-%------------------------------------------------------------
-\subsection{Heatmap analyses} 
-\label{subsec:heatmap}
-%------------------------------------------------------------
-
-Potentially useful figures can be plotted from \Rpackage{DESeq}. Instead of
-browsing differentially expressed genes, they can be plotted in group. The
-original fold changes could be moderated to be amenable to distance calculation. 
-Figure \ref{figLfc} shows moderated log fold changes versus the original log
-fold changes.
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figLfc}
-\caption{Scatterplot of direct (\Robject{lfc}) versus \emph{moderated} log-ratios (\Robject{mod\_lfc}).
-  The moderation criterion used is variance stabilisation.
-  The red points correspond to values that were infinite in \Robject{lfc}
-  and were abitrarily set to $\pm\Sexpr{largeNumber}$ for the purpose of plotting.
-  These values vary in a finite range (as shown in the plot) in \Robject{mod\_lfc}.}
-\label{figLfc}
-\end{figure}
-
-<<figLfc,fig=TRUE>>=
-select <- 1:40
-smutans.de2TypeLfc( ua159only, condA="glucose", condB="galactose" )
-@
-
-Now we plot the first \Sexpr{length(select)} genes clustered based on the
-moderated log fold changes.
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figHeatmap1}
-\caption{Heatmaps showing the expression data of the top \Sexpr{length(select)}
-  differentially expressed genes. Left, the variance stabilisation
-  transformed data,
-  right, the original count data.
-  The right plot is dominated by a small number of data points with large values.}
-\label{figHeatmap1}
-\end{figure}
-
-<<figHeatmap1,fig=TRUE>>=
-smutans.de2TypeHeatmap( ua159only )
-@
-
-We can cluster samples using variance stabilized data. Figure \ref{figClust}
-shows the 13 sampled clustered. 
-The clustering correctly reflects our experimental design, i.e., samples are
-more similar when they have the same treatment or the same library type.
-(To make this conclusion, it was important to re-estimate e dispersion with
-mode ``blind'' (in the calculation for \Robject{cdsFullBlind} above, as only then,
-the variance stabilizing transformation is not informed about the design and
-hence not biased towards a result supporting it.) Such an analysis is useful
-for quality control, and (even though we finish our vignette with it), it may
-be useful to this first in any analysis.
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=.9\textwidth]{smutans-figClust}
-%\caption{Heatmap showing the Euclidean distances between the samples
-%  as calculated from the variance-stabilising transformation of the
-%  count data.}
-%\label{figClust}
-%\end{figure}
-
-<<figClust,fig=TRUE>>=
-# smutans.de2Clust( allSample )
-@
-
-We find some functional categories enriched with differentially expressed genes.
-<<goseqUA159only>>=
-data( "smutans.feature.genes" )
-data( "smutans.go.genes" )
-data( "smutans.cat.desc" )
-smutans.de2Goseq ( ua159only, qval=1e-2, 
-                   feature.gene=smutans.feature.genes,
-                   go.genes=smutans.go.genes,
-                   cat.desc=smutans.cat.desc )
-@
-
-%--------------------------------------------------
-\section{Differential expression analysis using 835NP, 835P, and UA159 no CSP}
-\label{sec:de835}
-%--------------------------------------------------
-
-A count data set is a table of numbers of short reads that are assigned to
-predefined genes. We use one of R packages for testing for differential
-expression: \Rpackage{DESeq}.
-
-To find functional categories enriched with differentially expressed genes we
-first find a list of differentially expressed genes. A set of genes are assigned
-to functional gene categories: \Robject{smutans.go.genes}. Features of the genes in
-UA159 include gene lengths: \Robject{smutans.feature.genes}.
-Functional categories with numbers of genes assigned are in
-\Robject{smutans.cat.desc}.
-
-We use RNA-seq data to define genes whose expression changes between
-samples \cite{Oshlack2010}. Short reads are first mapped to a
-reference genome using approaches such as BWA \cite{Li2009a}.
-Reads fitting into one of the annotated intervals are counted.
-These counts are passed to a statistical package which treats differences
-in read counts for statistical significance. For this purpose, we made
-use of DEseq \cite{Anders2010}. This prodecure returns a list of
-differentially expressed genes, which can be passed to downstream
-analysis, such as functionaly category enrichment test (e.g., Young et
-al., 2010).
 
@@
 
-We could find no functional categories enriched with differentially expressed
-genes in the pairwise comparisons of 835NP, 835P, and UA159 no CSP.
+The \Rpackage{smutans} expects count data in the form of
+\Rclass{CountDataSet}, which is a type provided by \Rpackage{DESeq}.  Using BWA
+\cite{Li2009a} and samtools \cite{Li2009b} 
+we aligned short reads from RNA sequencing. Using gene annotations
+in the UA159 genome we counted short reads that overlapped the genes.  The count
+values are raw counts of sequencing reads. The \Rpackage{DESeq}'s statistical
+model, or negative binomial model, hold for raw reads that allows to assess the
+measurement precision correctly; we did not have rounded values of normalized
+counts, or counts of covered base pairs.  Several exeriments of \textit{S.\
+mutans} RNA sampling were performed.  \fixme{Find the papers that are related
+with the experiments.} One of them was presented in the 
+\Robject{smutansGenes} as a part of \Rpackage{smutans}. The
+\Robject{smutansGenes} is of \Rclass{CountDataSet}, which is the
+data type used by \Rpackage{DESeq}. We have a convenient 
+\Rclass{Smutans}, which contains slots for \Rclass{CountDataSet} objects, and
+results from analyses thereof.  Before loading necessary R packages and data
+sets, let us describe what functions were used for preparing data objects such as
+\Robject{smutansGenes}. One should create an R script file at the directory that contains the
+source code directory of the \Rpackage{smutans}. Put the following scripts in the R script
+file
+and run the script at that directory. That should create RData files at
+the directory smutans/data. User should not run the R script because the data
+files should be found in the \Rpackage{smutans}.
+% http://www.oralgen.lanl.gov/oralgen/downloads/smut.csv
+% http://www.oralgen.lanl.gov/cgi-bin/gene_id_search.cgi?dbname=smut&gene_id=SMu0001
+% \fixme{Where are flat files for genes of S. mutans UA159?}
+<<samplePreparation,eval=FALSE>>=
+# not run
+# but create a R scirpt file with the following scripts, and
+# run it, i.e., Rscript run.R
+library(DESeq)
+source("smutans/R/core.R")
+smutans.prepareData34( )
+smutans.prepareData835NPP( )
+smutans.prepareDataUA159TW1( ) 
+smutans.prepareGoseq( )
+smutans.prepareDataOMZ175( ) 
+smutans.prepareDataUA159CSP( ) 
+smutans.prepareDataSMU86CSP( )
+q("no")
+@
+<<options,results=hide,echo=FALSE>>=
+options(digits=3)
+@
+Now, see some of the data objects. Load the \Rpackage{smutans}, and a data
+\Robject{smutansGenes}. Note that the object is of type \Rclass{CountDataSet} of
+\Rpackage{DESeq} in which users should find \Rfunction{counts} and
+\Rfunction{pData}. 
+<<smutans2,results=hide>>=
+library( "smutans" )
+data( "smutansGenes" )
+@
+There are 
+\Sexpr{length(colnames(counts(smutansGenes)))}
+RNA-seq samples in \Robject{smutansGenes}, which
+contains the counts and also metadata about the samples:
+<<look_into_smutansGenes>>=
+head( counts(smutansGenes) )
+@
+<<look_into_smutansGenes>>=
+pData( smutansGenes )
+@
+The samples differ by experimental condition (``glucose'' or
+``galactose'') and by library type (``ua159'' or ``tw1'').
 
-Let's make a list of differentially expressed genes. It could be interesting to
-see the cluster of the three kinds of samples. Another kind of heatmap plots for
-grouping genes could be interesting. But, basically we have a list of genes with
-some adjusted p-values for detecting differentially expressed genes.
-
+Samples of 835NP, 835P, and UA159 no CSP should be loaded using the following
+command.
 <<sm835,results=hide>>=
@@
 @
+
+\fixme{Describe other data objects as well.
+sm34Genes.RData,
+sm835Genes.RData,
+smutans.cat.desc.RData,
+smutans.feature.genes.RData, and
+smutans.go.genes.RData.}
+
+%--------------------------------------------------
+\section{DE analysis using UA159 and TW1 samples with
+glucose and galactose} \label{sec:detw1}
+%--------------------------------------------------
+
+We could compare four pairwise comparisons using the two-factor data set of
+\Robject{smutansGenes}: a pair of comparisons between ``glucose'' and
+``galactose'' given either type ``ua159'' or ``tw1'', and another pair of
+comparisons between types ``ua159'' and ``tw1'' given either condition
+``glucose'' or ``galactose.'' We focus on one of the four pairwise comparisons
+first. Later, we will perform analyses for the rest of the three comparisons.
+We look at the ``ua159'' data for now, and will look at the ``tw1'' data later.
+
+We can cluster samples using variance stabilized data. Figure \ref{figClust}
+shows the \Sexpr{length(colnames(counts(smutansGenes)))} sampled clustered. 
+The clustering should reflect our experimental design, i.e., samples are
+more similar when they have the same treatment or the same library type.
+To make this conclusion, we estimate the dispersion with
+mode ``blind'' (see \Rpackage{DESeq} vignette), as only then,
+the variance stabilizing transformation is not informed about the design and
+hence not biased towards a result supporting it. Such an analysis is useful
+for quality control.  This appears to be useful as a first in any analysis.
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{smutans-figClust}
+\caption{Heatmap showing the Euclidean distances between the samples
+  as calculated from the variance-stabilizing transformation of the
+  count data.}
+\label{figClust}
+\end{figure}
+<<figClust,fig=TRUE>>=
+allSample <- newSmutans( smutansGenes )
+smutans.de2Clust( allSample )
+@
+As expected, two experimental conditions of ``glucose'' and ``galactose'' divide the
+samples. Within each condition, two different types of ``ua159'' and ``tw1'' divide
+them again. The pair of ``glucose ua159'' and ``galactose tw1'' is the most
+distinguished.
+
+%------------------------------------------------------------
+\subsection{Inference using single-factor data} 
+\label{subsec:ua159only}
+%------------------------------------------------------------
+
+We create a \Rclass{Smutans} object using the \Rclass{CountDataSet}
+\Robject{smutansGenes}. The title option would be useful in distinguishing
+different objects of \Rclass{Smutans} as we have more R objects with the type.  
+A subsequent call to
+\Rfunction{smutans.de2} uses a two-factor data set to test differential
+expression of genes.  The \Rfunction{smutans.de2} estimates effective library
+sizes, and dispersions using functions provided by \Rpackage{DESeq}. Then, it
+tests genes for differential expression.
+<<ua159onlySamples>>=
+ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
+ua159only <- smutans.de2( ua159only, type="ua159", 
+                          condA="glucose", condB="galactose" )
+@
+Because \Rpackage{DESeq} recommended to check
+the dispersion estimates and to make sure that the data quality is as
+expected, we plot the per-gene estimates against the normalized mean
+expressions per gene, and then overlay the fitted curve in red color
+(Fig.\ \ref{figFit}).
+
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figFit}
+\caption{Empirical (black dots) and fitted (red lines) dispersion values
+plotted against mean expression strength for the data set of UA159-only sample
+to compare glucose and galactose conditions.}
+\label{figFit}
+\end{figure}
+<<figFit,fig=TRUE>>=
+smutans.plotDispersionEstimates(ua159only)
+@
+\fixme{The following paragraph is adopted from \Rpackage{DESeq} vignette.}
+As we estimated the dispersion from only two samples, we should expect the
+estimates to scatter with quite some sampling variance around their true values.
+Hence, \Rpackage{DESeq} should not use the per-gene estimates directly in the
+test, because using too low dispersion values leads to false positives.  Many of
+the values below the red line are likely to be underestimates of the true
+dispersions, and hence, it is prudent to instead rather use the fitted value.
+On the othe hand, not all of the values above the red line are overestimations,
+and hence, the conservative choice is to keep them instead of replacing them
+with their fitted values.
+Having estimated the dispersion for each gene, we look for differentially
+expressed genes. To contrast two conditions, e.g., to see whether there is
+differential expression between conditions ``glucose'' and ``galactose'',
+\Rfunction{smutans.de2} already performed the 
+\Rfunction{nbinomTest} provided by \Rpackage{DESeq}.  
+Let's list genes with resulting information from \Rpackage{DESeq}.
+<<ua159onlyDE2List>>=
+head(smutans.de2List( ua159only ))
+@
+For each gene, we get its mean expression level (at the base scale) as a joint
+estimate from both conditions, and estimated separately for each condition, the
+fold change from the first to the second condition, i.e., 
+$\frac{baseMeanB}{baseMeanA}$, the logarithm (to basis 2)
+of the fold change, and the $p$ value for the statistical significance of this
+change. The \texttt{padj} column contains the $p$ values, adjusted for multiple
+testing with the Benjamini-Hochberg procedure (see the R function
+\Rfunction{p.adjust}), which controls false discovery rate (FDR). Later, we
+print the gene list out in a CSV file using option file of
+\Rfunction{smutans.de2List}.
+%-----------------------------------------------------------------------------
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figDE}
+\caption{Plot of normalised mean versus $log_2$ fold change
+(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
+versus ``galactose'' using UA159-only sample.}
+\label{figDE}
+\end{figure}
+%-----------------------------------------------------------------------------
+<<figDE,fig=TRUE>>=
+smutans.plotDiffExp(ua159only)
+@
+Figure \ref{figDE} shows a plot of the $\log_2$ fold changes against the base
+means, coloring in red those genes that are significant at 10\% of FDR. 
+Let us use ``tw1'' type to compare ``glucose'' and ``galactose'' conditions.
+We continue to analyze the data set using type ``tw1'' to compare the two
+conditions. We perform the single-factor analysis. 
+<<tw1onlySamples>>=
+tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
+tw1only <- smutans.de2( tw1only, type="tw1", 
+                        condA="glucose", condB="galactose" )
+tabUA159onlyTW1only <- smutans.compareResult( ua159only, tw1only )
+addmargins( tabUA159onlyTW1only )
+@
+The table, \Robject{tabUA159onlyTW1only}, shows numbers of genes found to be 
+differentially expressed in the two analyses: ``UA159 Only'' and ``TW1 Only.''
+Of the total \Sexpr{sum(tabUA159onlyTW1only)} genes, \Sexpr{tabUA159onlyTW1only[2,2]}
+genes were found in both analyses. The analysis using TW1-only sample found
+\Sexpr{tabUA159onlyTW1only[1,2]} genes. The analysis using UA159-only sample found
+\Sexpr{tabUA159onlyTW1only[2,1]} genes.
+
+Until now we focused on the effect of condition either using UA159-only or
+TW1-only samples. We now change the gear to see effects of sample types on
+differentially expressed genes. We create two \Rclass{Smutans} objects:
+\Robject{glucoseOnly} and \Robject{galactoseOnly}, and compare them in detecting
+differentially expressed genes:
+<<glucoseOnlySamples>>=
+glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
+glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
+                            condA="ua159", condB="tw1" )
+galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
+galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
+                              condA="ua159", condB="tw1" )
+tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
+                                                      galactoseOnly )
+addmargins( tabGlucoseOnlyGalactoseOnly )
+@
+Of the total \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} genes,
+\Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]} genes were found in both analyses at FDR
+rate of 10\%. The analysis using galactose-only sample found
+\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} genes. The analysis using glucose-only
+sample found \Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} genes. 
+Table \ref{tab:de} summarizes the number of differentially expressed genes that
+were found using single-factor analyses at FDR of 10\%.
+
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Numbers of differentially expressed genes in different analyses}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{ccccl}
+\hline 
+Sample A & Sample B & Number of Genes & Result File \\
+\hline 
+UA159 Glucose & UA159 Galactose & \Sexpr{sum(tabUA159onlyTW1only[2,])} & ua159only.csv \\
+TW1 Glucose & TW1 Galactose & \Sexpr{sum(tabUA159onlyTW1only[,2])} & tw1only.csv \\
+UA159 Glucose & TW1 Glucose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[2,])} & glucoseOnly.csv \\
+UA159 Galactose & TW1 Galactose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[,2])} & galactoseOnly.csv \\
+\hline 
+\end{tabular}
+\par\end{centering}
+\label{tab:de}
+\end{table}
+%-----------------------------------------------------------------------------
+
+%------------------------------------------------------------
+\subsection{Inference using all of the samples} 
+\label{subsec:allSample}
+%------------------------------------------------------------
+
+It is inconvenient to use four different analyses to find differentially
+expressed genes under one of two factors. Besides, it would be desirable to use all of
+the data to find differentially expressed genes because using more samples might
+be more powerful in detecting differentially expressed genes.  Let us find genes
+differentially expressed in different conditions of ``glucose'' and
+``galactose'' using both types of ``ua159'' and ``tw1.'' We will compare these
+genes and those found in the single-factor analysis that we did previously,
+i.e., \Robject{ua159only}.
+Let's start with multi-factor the analysis by creating a new \Rclass{Smutans} object, called
+\Robject{allSample}. Although we already used \Robject{allSample} when
+clustering the samples, we recreate it for instructive purposes.
+<<allSamples,results=hide>>=
+allSample <- newSmutans( smutansGenes, title="All Samples for Comparing Conditions" )
+allSample <- smutans.de2( allSample )
+@
+The \Rfunction{smutans.de2} uses a two-factor data set to
+perform two-factor analysis. The two factors are type and condition. We are here
+interested in condition effect. We could check 
+the dispersion estimates and to make sure that the data quality is as
+expected by plotting the per-gene estimates against the normalized mean
+expressions per gene, and then overlay the fitted curve.
+<<figFitAllSample,eval=FALSE>>=
+# not run
+smutans.plotDispersionEstimates(allSample)
+@
+%-----------------------------------------------------------------------------
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figCompareUA159onlyAndAllSample}
+\caption{Comparison of $p$ values (unadjusted) from the test using only the
+four paired-end samples and the test using all seven samples. We can see that
+the latter tend to be smaller, as expected from the higher power of the test with all samples.}
+\label{figCompareUA159onlyAndAllSample}
+\end{figure}
+%-----------------------------------------------------------------------------
+<<figCompareUA159onlyAndAllSample,fig=TRUE>>=
+pval.ua159only <- smutans.pval( ua159only )
+pval.allSample <- smutans.pval( allSample )
+bottom = function(x, theta=1e-12) pmax(x, theta)
+plot( bottom(pval.ua159only), bottom(pval.allSample), log="xy", pch=20, cex=.3 )
+abline(a=0, b=1, col="blue")
+@
+<<compareUA159onlyAndAllSample>>=
+tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
+                                                   allSample )
+addmargins( tabUA159onlyAndAllSample )
+@
+Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
+\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
+analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
+genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
+genes.  Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from
+the two-factor analysis versus those from the single-factor analysis. The
+p-values from the two-factor analysis are overall smaller than those from the
+single-factor analysis. We would expect that this was the case because of higher
+power of the two-factor analysis.
+
+We can show that taking the library type into account was important to have
+good detection power by doing the analysis again using the standard workflow,
+as outlined earlier, and without informing \Rpackage{DESeq} of the library
+types. This anlaysis is not necessary. However, it clearly shows the advantage of the
+two-factor analysis.
+<<allSampleWithoutType>>=
+allSampleWithoutType <- newSmutans( smutansGenes, "All Sample without Type" )
+allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
+                                                condA="glucose", 
+                                                condB="galactose" )
+tabWithoutWithAllSample <- smutans.compareResult( allSampleWithoutType, 
+                                                  allSample )
+addmargins( tabWithoutWithAllSample )
+@
+Of the total \Sexpr{sum(tabWithoutWithAllSample)} genes,
+\Sexpr{tabWithoutWithAllSample[2,2]} genes were found in both analyses. The
+analysis using all of the sample found \Sexpr{tabWithoutWithAllSample[1,2]}
+genes. The analysis using all of the same sample but without considering type
+found only \Sexpr{tabWithoutWithAllSample[2,1]} genes.
+
+As we mentioned above, we can have two analyses using the two-factor analysis.
+Until now we compared conditions.
+We proceed to compare the two types by considering all of the samples.
+<<allSampleType,results=verbatim>>=
+allSampleType <- newSmutans( smutansGenes, title="All samples for type" )
+allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
+tabGlucoseOnlyAllSample <- smutans.compareResult( glucoseOnly, 
+                                                  allSampleType )
+addmargins( tabGlucoseOnlyAllSample )
+tabGalactoseOnlyAllSample <- smutans.compareResult( galactoseOnly, 
+                                                    allSampleType )
+addmargins( tabGalactoseOnlyAllSample )
+@
+Of the total \Sexpr{sum(tabGlucoseOnlyAllSample)} genes,
+\Sexpr{tabGlucoseOnlyAllSample[2,2]} genes were found in both analyses. The
+analysis using all of the sample found \Sexpr{tabGlucoseOnlyAllSample[1,2]}
+genes. The analysis using glucose-only sample
+found \Sexpr{tabGlucoseOnlyAllSample[2,1]} genes.  
+In the comparison using galactose-only sample,
+of the \Sexpr{sum(tabGalactoseOnlyAllSample)} genes,
+\Sexpr{tabGalactoseOnlyAllSample[2,2]} genes were found in both analyses. The
+analysis using all of the sample found \Sexpr{tabGalactoseOnlyAllSample[1,2]}
+genes. The analysis using galactose-only sample
+found \Sexpr{tabGalactoseOnlyAllSample[2,1]} genes.
+
+We print out the result of differentially expressed genes in CSV files. A
+spreadsheet program such as EXCEL should be able to open CSV files. Users who
+are more comfortable with the EXCEL program than R could post-process CSV
+files for sorting and finding differentially expressed genes using different
+q-value cutoff values.
+<<smutanSamples>>=
+smutans.de2List( allSample, "allSample.csv" )
+smutans.de2List( allSampleType, "allSampleType.csv" )
+smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
+smutans.de2List( galactoseOnly, "galactoseOnly.csv" )
+smutans.de2List( ua159only, "ua159only.csv" )
+smutans.de2List( tw1only, "tw1only.csv" )
+head( smutans.de2List( allSample ) )
+@
+The output from \Robject{allSample} is different from that of
+\Robject{ua159only} or \Robject{tw1only}.
+The first column of ``allSample.csv'' is for genes.
+The next three columns show the fitted coefficients, converted to a logarithm
+base 2 scale: ``X.Intercept.'', ``typeua159'', and ``conditionglucose.''
+The log2 fold change due to the condition is shown in the 4th column. As indicated
+by the column name, it is the effect of ``glucose'', i.e., the log ratio of
+``glucose'' versus ``galactose'' or $\log(glucose/galactose)$. 
+Note that the library type also
+had noticeable influence on the expression, and hence would have increased the dispersion
+estimates (and so reduced power) if we had not fitted an effect for it.
+The column \emph{deviance}
+is the deviance of the fit. (Comparing the deviances with a $\chi^2$ likelihood ratio test is
+how \Rfunction{nbinomGLMTest} calculates the $p$ values.) The column, \emph{converged},
+indicates whether the calculation of coefficients and deviance has fully converged.
+(If it is false too often, you can try to change the GLM control parameters, as
+explained in the help page to \Rfunction{fitNbinomGLMs}.)
+The second to the last column, \emph{pval}, is the raw p-values from the test
+\Rfunction{nbinomGLMTest} provided by \Rpackage{DESeq}. The last column,
+\emph{padj}, is the multiple testing adjusted p-values.
+
+%------------------------------------------------------------
+\subsection{Heatmap analyses} 
+\label{subsec:heatmap}
+%------------------------------------------------------------
+
+Once we find differentially expressed genes, we could cluster them. This could
+be useful to group genes that behave similarly.  These potentially useful
+figures can be plotted from \Rpackage{DESeq}.  We could also find function
+categories that were enriched with differentially expressed genes. For
+clustering differentially expressed genes we need moderate fold changes to make
+them amenable to distance calculation.  Figure \ref{figLfc} shows moderated log
+fold changes versus the original log fold changes.
+
+<<options,results=hide,echo=FALSE>>=
+largeNumber <- 10
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figLfc}
+\caption{Scatterplot of direct (\Robject{lfc}) versus \emph{moderated} log-ratios (\Robject{mod\_lfc}).
+  The moderation criterion used is variance stabilisation.
+  The red points correspond to values that were infinite in \Robject{lfc}
+  and were abitrarily set to $\pm\Sexpr{largeNumber}$ for the purpose of plotting.
+  These values vary in a finite range (as shown in the plot) in \Robject{mod\_lfc}.}
+\label{figLfc}
+\end{figure}
+<<figLfc,fig=TRUE>>=
+select <- 1:40
+smutans.de2TypeLfc( ua159only, condA="glucose", condB="galactose" )
+@
+Now we plot the first \Sexpr{length(select)} genes clustered based on the
+moderated log fold changes.
+
+%-----------------------------------------------------------------------------
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{smutans-figHeatmap1}
+\caption{Heatmaps showing the expression data of the top \Sexpr{length(select)}
+  differentially expressed genes. Left, the variance stabilization
+  transformed data, right, the original count data.
+  The right plot is dominated by a small number of data points with large values.}
+\label{figHeatmap1}
+\end{figure}
+%-----------------------------------------------------------------------------
+<<figHeatmap1,fig=TRUE>>=
+smutans.de2TypeHeatmap( ua159only )
+@
+
+<<goseqUA159only,eval=FALSE>>=
+data( "smutans.feature.genes" )
+data( "smutans.go.genes" )
+data( "smutans.cat.desc" )
+smutans.de2Goseq ( ua159only, qval=1e-2, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+We find some functional categories enriched with differentially expressed genes
+(table \ref{tab:goseq})
+We find functional categories that are enriched with differentially expressed genes.
+To find functional categories enriched with differentially expressed genes we
+first find a list of differentially expressed genes. A set of genes are assigned
+to functional gene categories: \Robject{smutans.go.genes}. Features of the genes in
+UA159 include gene lengths: \Robject{smutans.feature.genes}.
+Functional categories with numbers of genes assigned are in
+\Robject{smutans.cat.desc}.
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+genes}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<goseqUA159only,echo=FALSE,results=tex>>=
+data( "smutans.feature.genes" )
+data( "smutans.go.genes" )
+data( "smutans.cat.desc" )
+smutans.de2Goseq ( ua159only, qval=1e-2, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseq}
+\end{table}
+%-----------------------------------------------------------------------------
+
+%--------------------------------------------------
+\section{DE analysis using 835NP, 835P, and UA159 no CSP}
+\label{sec:de835}
+%--------------------------------------------------
+
+We first cluster the \Sexpr{length(colnames(counts(sm835Genes)))} samples.
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{smutans-figClustNpAndP}
+\caption{Empirical (black dots) and fitted (red lines) dispersion values
+plotted against mean expression strength.}
+\label{figClustNpAndP}
+\end{figure}
+<<figClustNpAndP,fig=TRUE>>=
+data( "sm835Genes" )
+npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
+smutans.de2Clust( npAndP )
+@
+
+We found no functional categories enriched with differentially expressed
+genes in the pairwise comparisons of 835NP, 835P, and UA159 no CSP.
+Let's make a list of differentially expressed genes. 
+Another kind of heatmap plots for
+grouping genes could be interesting. 
+But, basically we have a list of genes with
+some adjusted p-values for detecting differentially expressed genes.
 
@@
                           condA="835p", condB="ua159nocsp" )
-@
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figFitNpAndP}
-\includegraphics[width=.6\textwidth]{smutans-figFitNpAndUA159}
-\includegraphics[width=.6\textwidth]{smutans-figFitPAndUA159}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength. \textbf{figFitNpAndP}}
-\label{figFitNpAndP}
-\end{figure}
-<<figFitNpAndP,fig=TRUE>>=
-smutans.plotDispersionEstimates(npAndP)
-@
-<<figFitNpAndUA159,fig=TRUE>>=
-smutans.plotDispersionEstimates(npAndUA159)
-@
-<<figFitPAndUA159,fig=TRUE>>=
-smutans.plotDispersionEstimates(pAndUA159)
-@
-
-<<sm835Samples>>=
 smutans.de2List( npAndP, "npAndP.csv" )
@@
 smutans.de2List( pAndUA159, "pAndUA159.csv" )
-@
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=.9\textwidth]{smutans-figClustNpAndP}
-%\caption{Empirical (black dots) and fitted (red lines) dispersion values
-%plotted against mean expression strength. \textbf{figClustNpAndP}}
-%\label{figClustNpAndP}
-%\end{figure}
-<<figClustNpAndP,fig=TRUE>>=
-# smutans.de2Clust( npAndP )
-@
-%\begin{figure}
-%\centering
-%\includegraphics[width=.9\textwidth]{smutans-figClustNpAndUA159}
-%\caption{Empirical (black dots) and fitted (red lines) dispersion values
-%plotted against mean expression strength. \textbf{figClustNpAndUA159}}
-%\label{figClustNpAndUA159}
-%\end{figure}
-<<figClustNpAndUA159,fig=TRUE>>=
-# smutans.de2Clust( npAndUA159 )
-@
-%\begin{figure}
-%\centering
-%\includegraphics[width=.9\textwidth]{smutans-figClustPAndUA159}
-%\caption{Empirical (black dots) and fitted (red lines) dispersion values
-%plotted against mean expression strength. \textbf{figClustPAndUA159}}
-%\label{figClustPAndUA159}
-%\end{figure}
-<<figClustPAndUA159,fig=TRUE>>=
-# smutans.de2Clust( pAndUA159 )
 @
@@
 @
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figHeatmapNpAndUA159}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength. \textbf{figHeatmapNpAndUA159}}
-\label{figHeatmapNpAndUA159}
-\end{figure}
-<<figHeatmapNpAndUA159,fig=TRUE>>=
-smutans.de2TypeHeatmap( npAndUA159 )
+
+%--------------------------------------------------
+\section{DE analysis using OMZ175 and OMZ/HKRR}
+\label{sec:deomz}
+%--------------------------------------------------
+
+<<look_into_smomzGenes>>=
+data( "smomzGenes" )
+head( counts(smomzGenes) )
+pData( smomzGenes )
 @
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figHeatmapPAndUA159}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength. \textbf{figHeatmapPAndUA159}}
-\label{figHeatmapPAndUA159}
-\end{figure}
-<<figHeatmapPAndUA159,fig=TRUE>>=
-smutans.de2TypeHeatmap( pAndUA159 )
+
+<<smomz,results=hide>>=
+omz <- newSmutans( smomzGenes, title="OMZ175 vs. OMZ/HKRR" )
+omz <- smutans.de2( omz, type="ua159", 
+                    condA="OMZ175", condB="OMZHKRR" )
+smutans.de2List( omz, "omz.csv" )
+@
+
+%--------------------------------------------------
+\section{DE analysis using UA159 CSP}
+\label{sec:deua159csp}
+%--------------------------------------------------
+
+<<look_into_smua159cspGenes>>=
+data( "smua159cspGenes" )
+head( counts(smua159cspGenes) )
+pData( smua159cspGenes )
+@
+
+<<smua159csp,results=hide>>=
+ua159csp <- newSmutans( smua159cspGenes, title="UA159 no CSP vs. CSP" )
+ua159csp <- smutans.de2( ua159csp, type="ua159", 
+                         condA="UA159noCSP", condB="UA159CSP" )
+smutans.de2List( ua159csp, "ua159csp.csv" )
+@
+
+%--------------------------------------------------
+\section{DE analysis using SMU86 CSP}
+\label{sec:desmu86csp}
+%--------------------------------------------------
+
+<<look_into_smsmu86cspGenes>>=
+data( "smsmu86cspGenes" )
+head( counts(smsmu86cspGenes) )
+pData( smsmu86cspGenes )
+@
+
+<<smsmu86csp,results=hide>>=
+smu86csp <- newSmutans( smsmu86cspGenes, title="UA159 no CSP vs. CSP" )
+smu86csp <- smutans.de2( smu86csp, type="ua159", 
+                         condA="Smu86noCSP", condB="Smu86CSP" )
+smutans.de2List( smu86csp, "smu86csp.csv" )
 @

### 2012-01-10 package/smutans/R/core.R

@@ -1,10 +1,10 @@
-smutans.compareResult <- function (resA, resB, qval=0.1)
+smutans.compareResult <- function ( x, y, qval=0.1 )
 {
-  padj.resA <- smutans.padj( resA )
-  padj.resB <- smutans.padj( resB )
+  padj.resA <- smutans.padj( x )
+  padj.resB <- smutans.padj( y )
   tab <- table( padj.resA < qval, 
                 padj.resB < qval,
-                dnn=c(smutans.title(resA), smutans.title(resB)) )
+                dnn=c(smutans.title(x), smutans.title(y)) )
   tab
 }
 
@@ -116,6 +116,202 @@ smutans.prepareData835NPP <- function ()
   save(sm835Genes, file=file.path("smutans", "data", "sm835Genes.RData"))
 }
 
+smutans.prepareDataOMZ175 <- function ()
+{
+  # Count data preparation
+  countsFile <- "smutans/inst/extdata/count.txt"
+  conds <- scan(file=paste(countsFile,"index",sep="."), what="character")
+  subconds <- conds %in% c("OMZ175", "OMZHKRR")
+  countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)
+  rownames(countsTable) <- countsTable$gene
+  countsTable <- countsTable[,-1]
+  countsTable <- countsTable[,subconds]
+  conds <- conds[subconds]
+  conds.type <- conds %in% c("OMZ175", "OMZHKRR")
+  factor.type <- rep("ua159",length(conds))
+  conds.condition <- conds %in% c("OMZ175")
+  factor.condition <- rep("OMZ175",length(conds))
+  conds.condition <- conds %in% c("OMZHKRR")
+  factor.condition[conds.condition] <- "OMZHKRR"
+  samples <- data.frame(type=factor.type,condition=factor.condition)
+  rownames(samples) <- colnames(countsTable) 
+
+  design <- samples
+  smomzGenes <- newCountDataSet( countsTable, design )
+  expdata = new("MIAME", 
+     name="S. mutans OMZ175, and OMZHKRR", 
+     lab="University of Florida, and Cornell University", 
+     contact="Drs. Robert Burne, Michael Stanhope, and Adam Siepel", 
[…]

### 2012-01-10 package/smutans/R/methods.R

@@ -15,24 +15,11 @@ setReplaceMethod("smutans.title",
   }
 )
 
-setGeneric("smutans.counts", function(object) standardGeneric("smutans.counts"))
-setMethod("smutans.counts",
-          "Smutans",
-  function( object ) {
-    counts( object@countDataSet )
-  }
+setGeneric("smutans.plotDispersionEstimates", 
+  function(object, file="default", 
+           xlab="Base mean expression", ylab="Dispersion values", qval=0.05,...) 
+           standardGeneric("smutans.plotDispersionEstimates")
 )
-
-setGeneric("smutans.counts<-", function(object,value) standardGeneric("smutans.counts<-"))
-setReplaceMethod("smutans.counts", 
-                 "Smutans",
-  function( object, value ) {
-    object@counts <- value
-    object
-  }
-)
-
-setGeneric("smutans.plotDispersionEstimates", function(object,...) standardGeneric("smutans.plotDispersionEstimates"))
 setMethod("smutans.plotDispersionEstimates", 
           "Smutans", 
   function(object, file="default", 
@@ -57,7 +44,11 @@ setMethod("smutans.plotDispersionEstimates",
   }
 )
 
-setGeneric("smutans.plotDiffExp", function(object,...) standardGeneric("smutans.plotDiffExp"))
+setGeneric("smutans.plotDiffExp", 
+  function( object, file="default", 
+            xlab="Base mean expression", ylab="Log base 2 fold changes", qval=0.05 ) 
+    standardGeneric("smutans.plotDiffExp")
+)
 setMethod("smutans.plotDiffExp", 
           "Smutans",
   function(object, file="default", 
@@ -100,16 +91,22 @@ setMethod("smutans.de2type",
   }
 )
 
-setGeneric("smutans.de2", function(object,...) standardGeneric("smutans.de2"))
+setGeneric("smutans.de2", 
+  function(object, compareCondition="yes", 
+           type="", condition="", condA="", condB="",
+           method="pooled") standardGeneric("smutans.de2")
+)
 setMethod("smutans.de2", 
           "Smutans", 
-  function(object, compareCondition="yes", type="", condition="", condA="", condB="") {
+  function(object, compareCondition="yes", 
+           type="", condition="", condA="", condB="",
+           method="pooled" ) {
[…]

### 2012-01-10 package/test.R

@@ -35,18 +35,41 @@ sm835result <- function () {
 #smutans.prepareGoseq( )
 #smutans.prepareData835NPP ()
 #smutans.prepareData34 ()
+#smutans.prepareDataOMZ175( ) 
+#smutans.prepareDataUA159CSP( ) 
+#smutans.prepareDataSMU86CSP( )
 #q("no")
-#library(DESeq)
-#library(goseq)
-#load("smutans/data/smutansGenes.RData")
-#load("smutans/data/sm835Genes.RData")
-#load("smutans/data/sm34Genes.RData")
+library(DESeq)
+library(goseq)
+load("smutans/data/smutansGenes.RData")
+load("smutans/data/sm835Genes.RData")
+load("smutans/data/sm34Genes.RData")
 source("smutans/R/class_and_slots.R")
 source("smutans/R/core.R")
 source("smutans/R/methods.R")
-npAndP <- newSmutans( sm34Genes, title="835NP vs. 835P" )
-npAndP <- smutans.de2( npAndP, type="ua159", 
-                       condA="835NP", condB="835P" )
+# smutans.de2Clust( allSample )
+allSample <- newSmutans( smutansGenes, title="All Samples for Comparsing Conditions" )
+allSample <- smutans.de2( allSample )
+load("smutans/data/smutans.cat.desc.RData")
+load("smutans/data/smutans.feature.genes.RData")
+load("smutans/data/smutans.go.genes.RData")
+smutans.de2Goseq ( allSample, qval=1e-2, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+
+
+#allSample <- newSmutans( smutansGenes, title="All Samples for Comparsing Conditions" )
+#allSample <- smutans.de2( allSample )
+#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
+#ua159only <- smutans.de2( ua159only, type="ua159", 
+#                          condA="glucose", condB="galactose" )
+#tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
+#tw1only <- smutans.de2type( tw1only, type="tw1", 
+#                            condA="glucose", condB="galactose" )
+#npAndP <- newSmutans( sm34Genes, title="835NP vs. 835P" )
+#npAndP <- smutans.de2( npAndP, type="ua159", 
+#                       condA="835NP", condB="835P" )
 
 #head( counts(sm835Genes) )
 #pData( sm835Genes )

### 2012-01-10 package/smutans/inst/doc/smutans.Rnw
Just an update.

@@
 \begin{abstract}
-We analyze a number of RNA samples from \textit{Streptococcus mutans}.
-
-using 
-\Rpackage{DESeq}, \Rpackage{edgeR}, and \Rpackage{goseq}.
-Tasks include the analysis of count data from RNA-seq for
-detection of differentially expressed genes, and tests of enrichment for
-differentially expressed genes using \Rpackage{goseq}.
-
-We use RNA-seq data to define genes whose expression changes between
-samples \cite{Oshlack2010}. Short reads are first mapped to a
-reference genome using approaches such as BWA \cite{Li2009a}.
-Reads fitting into one of the annotated intervals are counted.
-These counts are passed to a statistical package which treats differences
-in read counts for statistical significance. For this purpose, we made
-use of \Rpackage{DEseq} \cite{Anders2010}. This prodecure returns a list of
-differentially expressed genes, which can be passed to downstream
-analysis, such as functional category enrichment test using 
-\Rpackage{goseq} \cite{Young2010}.
-
-This
-\Rpackage{smutans} contains procedures that make use of functions
-provided by the above R packages.  
+We analyze a number of RNA samples from \textit{Streptococcus mutans}.  We use
+RNA-seq to define genes whose expression changes between samples
+\cite{Oshlack2010}. Short reads are first mapped to a reference genome using
+approaches such as BWA \cite{Li2009a}.  Reads fitting into one of the annotated
+intervals are counted.  These counts are passed to a statistical package which
+treats differences in read counts for statistical significance. For this
+purpose, we made use of \Rpackage{DEseq} \cite{Anders2010} or \Rpackage{edgeR}
+\cite{Robinson2010}. This prodecure returns a list of differentially expressed
+genes, which can be passed to downstream analysis, such as functional category
+enrichment test using \Rpackage{goseq} \cite{Young2010}.  This
+\Rpackage{smutans} contains procedures that make use of functions provided by
+the above R packages. 
 \end{abstract}
@@
 
-The abtract briefly describes main steps of analyzing RNA-seq data for
-determining differentially expressed genes \cite{Oshlack2010}. We
-used a short read aligner called BWA \cite{Li2009a} to map the
-sequenced short reads on a reference genome, \textit{Streptococcus mutans}
-UA159 (NCBI accession NC\_004350). Because the aligner allowed a few gaps
-for efficient alignment of millions of sequences of size being 100 base
-pairs long, short reads that consist of RNAs and sequencing adapter
-sequences would be not mapped. We used cutadapt \cite{Martin2011} to remove
-parts of adapter sequences from short reads before mapping on the reference
-genome. Mapped short read alignments were converted into readable formats
-\cite{Li2009b}. We counted short reads aligned to annotated genes in the
+Steps of analyzing RNA-seq data for determining differentially expressed genes
+are well described in \cite{Oshlack2010}. We used a short read aligner called
+BWA \cite{Li2009a} to map the sequenced short reads on a reference genome,
+\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
+aligner allowed a few gaps for efficient alignment of millions of sequences of
+size being 100 base pairs long, short reads that consist of RNAs and sequencing
+adapter sequences would be not mapped. We used cutadapt \cite{Martin2011} to
+remove parts of adapter sequences from short reads before mapping on the
+reference genome. Mapped short read alignments were converted into readable
+formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
 reference genome, forming a table of read counts of the genes. Statistical
-software R \cite{R2011} with its one of R package
-called DEseq \cite{Anders2010} were subsequently employed to infer
-significantly differentially expressed in a pair of biological conditions.
-We also generated new tracks for our recently released Streptococcus Genome
-Browser (http://strep-genome.bscb.cornell.edu) that summarize gene
-expression levels based the short read alignments. This allowed us to
-visualize parts of the reference genome for comparing different RNA
+software R \cite{R2011} with its one of R package called DEseq \cite{Anders2010}
+was subsequently employed to infer differentially expressed in different
+biological conditions.  We also generated new tracks for our recently released
+Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
+summarize gene expression levels based the short read alignments. This allowed
+us to visualize parts of the reference genome for comparing different RNA
 samples. Further analyses such as functionaly category enrichment test
-\cite{Young2010} were pursued.
+\cite{Young2010} were employed.
 
-The \Rpackage{smutans} expects count data in the form of
-\Rclass{CountDataSet}, which is a type provided by \Rpackage{DESeq}.  Using BWA
-\cite{Li2009a} and samtools \cite{Li2009b} 
-we aligned short reads from RNA sequencing. Using gene annotations
-in the UA159 genome we counted short reads that overlapped the genes.  The count
-values are raw counts of sequencing reads. The \Rpackage{DESeq}'s statistical
-model, or negative binomial model, hold for raw reads that allows to assess the
-measurement precision correctly; we did not have rounded values of normalized
-counts, or counts of covered base pairs.  Several exeriments of \textit{S.\
-mutans} RNA sampling were performed.  \fixme{Find the papers that are related
-with the experiments.} One of them was presented in the 
+The \Rpackage{smutans} expects count data in the form of \Rclass{CountDataSet},
+which is a type provided by \Rpackage{DESeq}.  Using BWA \cite{Li2009a} and
+samtools \cite{Li2009b} we aligned short reads from RNA sequencing. Using gene
+annotations in the UA159 genome we counted short reads that overlapped the
+genes.  The count values are raw counts of sequencing reads. The
+\Rpackage{DESeq}'s statistical model, or negative binomial model, hold for raw
+reads that allows to assess the measurement precision correctly; we did not have
+rounded values of normalized counts, or counts of covered base pairs.  Several
+exeriments of \textit{S.\ mutans} RNA sampling were performed.  \fixme{Find the
+papers that are related with the experiments.} One of them was presented in the
 \Robject{smutansGenes} as a part of \Rpackage{smutans}. The
-\Robject{smutansGenes} is of \Rclass{CountDataSet}, which is the
-data type used by \Rpackage{DESeq}. We have a convenient 
-\Rclass{Smutans}, which contains slots for \Rclass{CountDataSet} objects, and
-results from analyses thereof.  Before loading necessary R packages and data
-sets, let us describe what functions were used for preparing data objects such as
-\Robject{smutansGenes}. One should create an R script file at the directory that contains the
-source code directory of the \Rpackage{smutans}. Put the following scripts in the R script
-file
-and run the script at that directory. That should create RData files at
-the directory smutans/data. User should not run the R script because the data
-files should be found in the \Rpackage{smutans}.
+\Robject{smutansGenes} is of \Rclass{CountDataSet}, which is the data type used
+by \Rpackage{DESeq}. We have a convenient \Rclass{Smutans}, which contains slots
+for \Rclass{CountDataSet} objects, and results from analyses thereof.  Before
+loading necessary R packages and data sets, let us describe what functions were
+used for preparing data objects such as \Robject{smutansGenes}. One should
+create an R script file at the directory that contains the source code directory
+of the \Rpackage{smutans}. Put the following scripts in the R script file and
+run the script at that directory. That should create RData files at the
+directory smutans/data. User should not run the R script because the data files
+should be found in the \Rpackage{smutans}.
+
 % http://www.oralgen.lanl.gov/oralgen/downloads/smut.csv
@@
 @
-Now, see some of the data objects. Load the \Rpackage{smutans}, and a data
+Now, see some of the data objects by loading the \Rpackage{smutans}, and a data
 \Robject{smutansGenes}. Note that the object is of type \Rclass{CountDataSet} of
@@
 \Rfunction{pData}. 
-<<smutans2,results=hide>>=
+<<smutans,results=hide>>=
 library( "smutans" )
@@
 <<allSamples,results=hide>>=
-allSample <- newSmutans( smutansGenes, title="All Samples for Comparing Conditions" )
+allSample <- newSmutans( smutansGenes, 
+                         title="All Samples for Comparing Conditions" )
 allSample <- smutans.de2( allSample )
@@
 %------------------------------------------------------------
-\subsection{Heatmap analyses} 
-\label{subsec:heatmap}
+\subsection{Functional category association analyses} 
+\label{subsec:functional}
 %------------------------------------------------------------

### 2012-01-10 package/smutans/R/methods.R

@@ -73,24 +73,6 @@ setMethod("smutans.plotDiffExp",
   }
 )
 
-setGeneric("smutans.de2type", function(object,...) standardGeneric("smutans.de2type"))
-setMethod("smutans.de2type", 
-          "Smutans", 
-  function(object, type="", condition="", condA, condB) {
-    ua159Samples <- pData(object@countDataSet)$type == type
-    countsTable <- counts(object@countDataSet)[ , ua159Samples ]
-    conds <- pData(object@countDataSet)$condition[ ua159Samples ]
-    cds <- newCountDataSet( countsTable, conds )
-    cds <- estimateSizeFactors( cds )
-    object@cds <- estimateDispersions( cds )
-    object@res <- nbinomTest( object@cds, condA, condB )
-    object@pval <- object@res$pval
-    object@padj <- object@res$padj
-  
-    object
-  }
-)
-
 setGeneric("smutans.de2", 
   function(object, compareCondition="yes", 
            type="", condition="", condA="", condB="",

### 2012-01-10 package/smutans/inst/doc/smutans.Rnw
R package smutans_0.1.3

@@
 \newcommand{\Rfunction}[1]{{function \small\texttt{#1}}}
+\newcommand{\program}[1]{{\texttt{#1}}}
 \newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}
@@
 
-\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data using R packages}}}
+\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data\\using R packages}}}
 
@@
 \begin{abstract}
-We analyze a number of RNA samples from \textit{Streptococcus mutans}.  We use
+We analyze RNA samples from \textit{Streptococcus mutans}.  We use
 RNA-seq to define genes whose expression changes between samples
 \cite{Oshlack2010}. Short reads are first mapped to a reference genome using
-approaches such as BWA \cite{Li2009a}.  Reads fitting into one of the annotated
+approaches such as \program{BWA} \cite{Li2009a}.  Reads fitting into one of the annotated
 intervals are counted.  These counts are passed to a statistical package which
@@
 genes, which can be passed to downstream analysis, such as functional category
-enrichment test using \Rpackage{goseq} \cite{Young2010}.  This
-\Rpackage{smutans} contains procedures that make use of functions provided by
+enrichment test using \Rpackage{goseq} \cite{Young2010}.  The current
+\Rpackage{smutans} contains procedures that use functions provided by
 the above R packages. 
@@
 Steps of analyzing RNA-seq data for determining differentially expressed genes
-are well described in \cite{Oshlack2010}. We used a short read aligner called
-BWA \cite{Li2009a} to map the sequenced short reads on a reference genome,
+are described in \cite{Oshlack2010}. We used a short read aligner called
+\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
 \textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
@@
 size being 100 base pairs long, short reads that consist of RNAs and sequencing
-adapter sequences would be not mapped. We used cutadapt \cite{Martin2011} to
+adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
 remove parts of adapter sequences from short reads before mapping on the
@@
 reference genome, forming a table of read counts of the genes. Statistical
-software R \cite{R2011} with its one of R package called DEseq \cite{Anders2010}
+software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
 was subsequently employed to infer differentially expressed in different
-biological conditions.  We also generated new tracks for our recently released
+biological conditions.  We also generated tracks for our recently released
 Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
@@
 us to visualize parts of the reference genome for comparing different RNA
-samples. Further analyses such as functionaly category enrichment test
-\cite{Young2010} were employed.
+samples. Further analyses such as functional category enrichment test
+\cite{Young2010} were also employed.
 
 The \Rpackage{smutans} expects count data in the form of \Rclass{CountDataSet},
-which is a type provided by \Rpackage{DESeq}.  Using BWA \cite{Li2009a} and
-samtools \cite{Li2009b} we aligned short reads from RNA sequencing. Using gene
+which is a type provided by \Rpackage{DESeq}.  Using \program{BWA} \cite{Li2009a} and
+\program{samtools} \cite{Li2009b} we aligned short reads from RNA sequencing. Using gene
 annotations in the UA159 genome we counted short reads that overlapped the

### 2012-01-10 package/smutans/R/methods.R

@@ -143,7 +143,7 @@ setMethod("smutans.pval",
 )
 
 setGeneric("smutans.de2WithoutType", 
-  function(object,...) standardGeneric("smutans.de2WithoutType")
+  function(object, condA, condB) standardGeneric("smutans.de2WithoutType")
 )
 setMethod("smutans.de2WithoutType", 
           "Smutans", 
@@ -161,7 +161,7 @@ setMethod("smutans.de2WithoutType",
 )
 
 setGeneric("smutans.de2TypeLfc", 
-  function(object,...) standardGeneric("smutans.de2TypeLfc")
+  function(object, condA, condB, file="default") standardGeneric("smutans.de2TypeLfc")
 )
 setMethod("smutans.de2TypeLfc", 
           "Smutans", 
@@ -225,11 +225,11 @@ setMethod("smutans.de2TypeHeatmap",
 )
 
 setGeneric("smutans.de2Clust", 
-  function(object,...) standardGeneric("smutans.de2Clust")
+  function(object,file="default", margins=c(10,10)) standardGeneric("smutans.de2Clust")
 )
 setMethod("smutans.de2Clust", 
           "Smutans", 
-  function(object, file="default", margins=c(10,10)) {
+  function(object,file="default", margins=c(10,10)) {
     if (file != "default")
     {
       postscript(file, width=6, height=6, horizontal = FALSE, onefile = FALSE, paper = "special")

### 2012-01-10 package/smutans/inst/doc/smutans.Rnw
Just an update.

@@
 \hline 
-UA159 Glucose & UA159 Galactose & \Sexpr{sum(tabUA159onlyTW1only[2,])} & ua159only.csv \\
-TW1 Glucose & TW1 Galactose & \Sexpr{sum(tabUA159onlyTW1only[,2])} & tw1only.csv \\
-UA159 Glucose & TW1 Glucose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[2,])} & glucoseOnly.csv \\
-UA159 Galactose & TW1 Galactose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[,2])} & galactoseOnly.csv \\
+UA159 Glucose & UA159 Galactose & \Sexpr{sum(ua159only@padj < .1,na.rm=T)} & ua159only.csv \\
+TW1 Glucose & TW1 Galactose & \Sexpr{sum(tw1only@padj < .1,na.rm=T)} & tw1only.csv \\
+UA159 Glucose & TW1 Glucose & \Sexpr{sum(glucoseOnly@padj < .1,na.rm=T)} & glucoseOnly.csv \\
+UA159 Galactose & TW1 Galactose & \Sexpr{sum(galactoseOnly@padj < .1,na.rm=T)} & galactoseOnly.csv \\
+% UA159 Glucose & UA159 Galactose & \Sexpr{sum(tabUA159onlyTW1only[2,])} & ua159only.csv \\
+% TW1 Glucose & TW1 Galactose & \Sexpr{sum(tabUA159onlyTW1only[,2])} & tw1only.csv \\
+% UA159 Glucose & TW1 Glucose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[2,])} & glucoseOnly.csv \\
+% UA159 Galactose & TW1 Galactose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[,2])} & galactoseOnly.csv \\
 \hline 
@@
 
+%--------------------------------------------------
+\section{DE analysis using predicted transcripts}
+\label{sec:detranscript}
+%--------------------------------------------------
+
+%--------------------------------------------------
+\section{DE analysis using predicted small RNAs and target genes}
+\label{sec:desmallrna}
+%--------------------------------------------------
+
 \section{Session Info}

### 2012-01-10 pl/feature-genome.pl

@@ -240,6 +240,26 @@ elsif ($cmd eq "extract")
     print $outfile "$subseq\n";
   }
 }
+elsif ($cmd eq "parsernaseq")
+{
+  my $i = 0;
+  while (<$infile>)
+  {
+    $i++;
+    my $istr = sprintf ("%04d", $i);
+    my @e = split /\t/;
+    print $outfile "$chromosome\t";
+    print $outfile "$e[3]\t$e[4]\ttx$istr\t0\t+\n";
+  }
+}
+elsif ($cmd eq "rnaz")
+{
+  while (<$infile>)
+  {
+    chomp;
+    print $outfile "$_+\n";
+  }
+}
 
 if (exists $params{in})
 {

### 2012-01-10 test/RNAz/mannwhitney.R

@@ -1,0 +1,96 @@
+gocats <- read.table("/usr/projects/strep/SpyMGAS315/SpyMGAS315_go_bacteria.txt", header=FALSE, col.names=c("gene", "go", "pval"))
+gocats <- gocats[gocats$pval <= 1.0e-5,]
+descrips <- read.table("/usr/projects/strep/SpyMGAS315/SpyMGAS315_go_category_names.txt", sep="\t", header=FALSE, col.names=c("go", "count", "descrip"), quote="")
+gocats$go <- as.character(gocats$go)
+descrips$go <- as.character(descrips$go)
+
+run.mw <- function(x, gocats, gocats.col="go", descrips=NULL, noisy=FALSE) {
+  results <- data.frame() 
+  for (gocat in unique(gocats[,gocats.col])) {
+    genes <- gocats[gocats[,gocats.col]==gocat, "gene"]
+    if (length(genes) >= 10) {
+      f <- is.element(x$gene, genes)
+      notf <- !f
+      if (sum(f) >= 10) {
+        wt <- wilcox.test(x[f,"score"], x[!f,"score"], alternative="greater")
+        if (noisy) cat(gocat, sum(f), wt$p.value, sep="\t")
+        if (!is.null(descrips)) {
+          w <- which(descrips[,gocats.col]==gocat)
+          if (length(w) != 1L) stop("couldn't find description for ", gocat)
+          if (noisy) cat(descrips[w,"descrip"])
+          results <- rbind(results, data.frame(gocat=gocat, count=sum(f), p.value=wt$p.value, go.description=descrips[w,"descrip"]))
+        } else results <- rbind(results, data.frame(gocat=gocat, count=sum(f), p.value=wt$p.value))
+        if (noisy) cat("\n")
+      }
+    }
+  }
+  results
+}
+
+get.significant <- function(results, p.val) {
+  adjust <- p.adjust(results[,p.val], method="BH")
+  if (sum(adjust < 0.05) == 0) {
+    cat("no significant results\n")
+    return(NULL)
+  }
+  temp <- results[adjust < 0.05,c(p.val, "count", "go.description")]
+  format.data.frame(temp[order(temp[,p.val]),], digits=4)
+}
+
+
+
+  
+x <- read.table("in.gene", header=FALSE)
+num.unique.mw <- run.mw(data.frame(gene=x[,1], score=x[,12]), gocats, descips=descrips)
[…]

### 2012-01-11 package/smutans/inst/doc/smutans.Rnw
Just an update.

@@
 @
+
+% Resources
+% Google charts - Venn Diagram
+% http://code.google.com/apis/chart/index.html
 
@@
 
+Create a file with intergenic regions sequences of \textit{S.\ mutans} UA159.
+Create a file with bacterial genome sequences except \textit{S.\ mutans} NN2025.
+BLAST the intergenic regions against the bacterial genomes. Extract sequence
+alignments from the BLAST output. Refine the alignment using MUSCLE. Convert the
+FASTA-format alignments to MSA-format ones.  Find small RNA candidates using
+RNAz.  Find targets of the small RNA candidates using RNAplex. Targets are
+scored for each small RNA candidate. The scores were used to do functional
+category enrichment tests of targets for each small RNA candidate.
+
+Table \ref{tab:goseq} 
+shows functional categories enriched with target genes for predicted small
+RNAs. Let's use \Rfunction{smutans.mannwhitney} to the package and other data
+objects.
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+genes}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{lcccl}
+\hline 
+small RNA & $q^a$ & count & GO & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<goseqUA159only,results=tex>>=
+data( "smutans.genes.criteria" )
+data( "smutans.go.genes" )
+data( "smutans.cat.desc" )
+smutans.mannwhitney( qval=0.05,
+                   genes.criteria=smutans.genes.criteria,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseq}
+\end{table}
+%-----------------------------------------------------------------------------
+
+<<smutansRNA>>=
+data( "smutansGenes2" )
+ua159only <- newSmutans( smutansGenes2, title="UA159 Only" )
+ua159only <- smutans.de2( ua159only, type="ua159", 
+                          condA="glucose", condB="galactose" )
+allSample <- newSmutans( smutansGenes2, 
+                         title="All Samples for Comparing Conditions" )
+allSample <- smutans.de2( allSample )
+tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
+                                                   allSample )
+addmargins( tabUA159onlyAndAllSample )
+@
+Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
+\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
+analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
+genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
+genes. We found a few predicted small RNAs that appeared to be differentially
+expressed.
+<<smutansRNA>>=
+smutans.de2List( allSample )[grep("locus",rownames(smutans.de2List(allSample))),]
+@
+We also found a few predicted small RNAs that appeared to be differentially
+expressed in the comparison of the two types: ``ua159'' and ``tw1.''
+<<smutansRNA>>=
+allSampleType <- newSmutans( smutansGenes2, title="All samples for type" )
+allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
+smutans.de2List( allSampleType )[grep("locus",rownames(smutans.de2List(allSampleType))),]
+@
+
 \section{Session Info}

### 2012-01-11 package/smutans/R/core.R

@@ -45,16 +45,25 @@ smutans.prepareData34 <- function ()
   save(sm34Genes, file=file.path("smutans", "data", "sm34Genes.RData"))
 }
 
-smutans.prepareDataUA159TW1 <- function ()
+smutans.prepareDataUA159TW1 <- function (cutoff.count=10)
 {
   # Count data preparation
   countsFile <- "smutans/inst/extdata/count.txt"
+  countsFile2 <- "smutans/inst/extdata/count2.txt"
   conds <- scan(file=paste(countsFile,"index",sep="."), what="character")
   subconds <- conds %in% c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL")
+
   countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)
   rownames(countsTable) <- countsTable$gene
   countsTable <- countsTable[,-1]
   countsTable <- countsTable[,subconds]
+
+  countsTable2 <- read.delim (countsFile2, header=TRUE, stringsAsFactors=TRUE)
+  rownames(countsTable2) <- countsTable2$gene
+  countsTable2 <- countsTable2[,-1]
+  countsTable2 <- countsTable2[,subconds]
+  countsTable2 <- countsTable2[rowSums(countsTable2) >=cutoff.count,]
+
   conds <- conds[subconds]
   conds.type <- conds %in% c("UA159GLU", "UA159GAL")
   factor.type <- rep("ua159",length(conds))
@@ -76,7 +85,19 @@ smutans.prepareDataUA159TW1 <- function ()
      abstract="RNA-seq of 13 biological replicates from Streptococcus mutans")
   pubMedIds(expdata) <- "999999999"
   experimentData(smutansGenes) <- expdata
-  save(smutansGenes, file=file.path(".", "smutansGenes.RData"))
+  save(smutansGenes, file=file.path("smutans", "data", "smutansGenes.RData"))
+
+  smutansGenes2 <- newCountDataSet( countsTable2, design )
+  expdata = new("MIAME", 
+     name="S. mutans UA159, TW1, Glucose, and Galactose based on transcripts and small RNAs", 
+     lab="University of Florida, and Cornell University", 
+     contact="Drs. Robert Burne, Michael Stanhope, and Adam Siepel", 
+     title="Streptococcus mutans RNA-Seq Studies", 
[…]

### 2012-01-11 package/smutans/R/methods.R

@@ -93,7 +93,7 @@ setMethod("smutans.de2",
         fit1 <- fitNbinomGLMs( object@cds, count ~ type + condition )
         fit0 <- fitNbinomGLMs( object@cds, count ~ type  )
       } else {
-        fit1 <- fitNbinomGLMs( object@cds, count ~ type + condition )
+        fit1 <- fitNbinomGLMs( object@cds, count ~ condition + type )
         fit0 <- fitNbinomGLMs( object@cds, count ~ condition )
       }
       object@pval <- nbinomGLMTest( fit1, fit0 )
@@ -265,13 +265,16 @@ setMethod("smutans.de2Clust",
 )
 
 setGeneric("smutans.de2List", 
-  function(object,file="default") standardGeneric("smutans.de2List")
+  function(object,file="default", qval=0.1) standardGeneric("smutans.de2List")
 )
 setMethod("smutans.de2List", 
           "Smutans", 
-  function(object, file="default") {
+  function(object, file="default", qval=0.1) {
     if (file == "default") {
-      return( object@res )
+      resSig <- object@res[object@res$padj < qval,]
+      # options(width = 1000) 
+      return(resSig[order(resSig$pval),])
+      # return( object@res )
     } else {
       write.csv(object@res, file=file)
     }
@@ -322,6 +325,4 @@ setMethod("smutans.de2Goseq",
     }
   }
 )
-
-
 

### 2012-01-11 package/test.R

@@ -38,25 +38,55 @@ sm835result <- function () {
 #smutans.prepareDataOMZ175( ) 
 #smutans.prepareDataUA159CSP( ) 
 #smutans.prepareDataSMU86CSP( )
+#smutans.prepareDataUA159TW1()
 #q("no")
 library(DESeq)
 library(goseq)
 load("smutans/data/smutansGenes.RData")
+load("smutans/data/smutansGenes2.RData")
 load("smutans/data/sm835Genes.RData")
 load("smutans/data/sm34Genes.RData")
+load("smutans/data/smutans.cat.desc.RData")
+load("smutans/data/smutans.genes.criteria.RData")
+load("smutans/data/smutans.go.genes.RData")
 source("smutans/R/class_and_slots.R")
 source("smutans/R/core.R")
 source("smutans/R/methods.R")
+# allSample2 <- newSmutans( smutansGenes2 )
+# smutans.de2Clust( allSample2 )
+#ua159only <- newSmutans( smutansGenes2, title="UA159 Only" )
+#ua159only <- smutans.de2( ua159only, type="ua159", 
+#                          condA="glucose", condB="galactose" )
+#allSample <- newSmutans( smutansGenes2, 
+#                         title="All Samples for Comparing Conditions" )
+#allSample <- smutans.de2( allSample )
+#tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
+#                                                   allSample )
+#addmargins( tabUA159onlyAndAllSample )
+allSampleType <- newSmutans( smutansGenes2, title="All samples for type" )
+allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
+print(
+smutans.de2List( allSampleType )[grep("locus",rownames(smutans.de2List(allSampleType))),]
+)
+#smutans.plotDispersionEstimates(ua159only)
+#smutans.plotDiffExp(ua159only)
+#smutans.de2List( ua159only )
+
+#smutans.mannwhitney( qval=0.05,
+#                   genes.criteria=smutans.genes.criteria,
+#                   go.genes=smutans.go.genes,
+#                   cat.desc=smutans.cat.desc )
+
 # smutans.de2Clust( allSample )
-allSample <- newSmutans( smutansGenes, title="All Samples for Comparsing Conditions" )
-allSample <- smutans.de2( allSample )
-load("smutans/data/smutans.cat.desc.RData")
-load("smutans/data/smutans.feature.genes.RData")
[…]

### 2012-01-13 package/smutans/inst/doc/smutans.Rnw
An BioC R package, GenomicRanges, is being tested.

@@
 
+We need ways of supporting the predicted transcripts. We need a way of using
+expression levels and counts of starts and ends of short reads to predict
+transcripts. For now, we just use an ad hoc method for finding transcript start
+and end sites. A likelihood can be constructed by multiplying geometric
+probabilities. Given a range of possible transcription boundaries we can find
+estimates of positions and geometirc parmeter $p$ that maximize the likelihood
+for both ends of a transcript.
+
+\Rpackage{GenomicRanges} might be useful in 
+<<tx,eval=FALSE>>=
+source("http://www.bioconductor.org/biocLite.R")
+biocLite("GenomicRanges")
+@
+
 %--------------------------------------------------
@@
 %-----------------------------------------------------------------------------
-<<goseqUA159only,results=tex>>=
+<<goseqUA159only,echo=FALSE,results=tex>>=
 data( "smutans.genes.criteria" )

### 2012-01-13 pl/bwa-pos2wig.pl

@@ -584,7 +584,7 @@ Command:
   pos   - print short read name, chr, start, and end positions
   rrna  - check if reads are in the set of rRNA
 
-  end   - counts the either one base pairs are counted.
+  end   - counts the either one base pairs
   first - find the distances to the first opposite signed value. Using the
   output file from command ``end'' for a given position whose value is non-zero
   I compute the distance to the first opposite signed value. If a given position

### 2012-01-13 pl/transcript-parsernaseq.pl

@@ -53,9 +53,11 @@ GetOptions( \%params,
             'wiggle=s',
             'feature=s',
             'end=s',
+            'inpileup=s',
             'peakcutoff=i',
             'sizecutoff=i',
             'windowsize=i',
+            'toplevel',
             'updowncutoff=i',
             'in=s',
             'out=s',
@@ -152,7 +154,7 @@ elsif ($cmd eq "gene")
     &printError("Command $cmd needs -feature options");
   }
 }
-elsif ($cmd eq "bed" or $cmd eq "gff")
+elsif ($cmd eq "bed" or $cmd eq "gff" or $cmd eq "combine")
 {
   unless (exists $params{parsernaseq})
   {
@@ -171,6 +173,16 @@ elsif ($cmd eq "adjust" or $cmd eq "slope")
   unless (exists $params{operon} and exists $params{end})
   {
     &printError("Command $cmd needs -operon and -end options");
+  }
+}
+elsif ($cmd eq "remove")
+{
+  unless (exists $params{parsernaseq} and 
+          exists $params{inpileup} and 
+          exists $params{feature} and
+          exists $params{out})
+  {
+    &printError("Command $cmd needs -parsernaseq, -feature and -inpileup options");
   }
 }
 
@@ -276,6 +288,23 @@ elsif ($cmd eq "bed")
 }
 elsif ($cmd eq "gff")
 {
+  # Find the highest level.
+  my $topLevel = 1;
+  open PARSERNASEQ, $parsernaseq or die "cannot open < $parsernaseq $!";
+  while (<PARSERNASEQ>)
+  {
+    chomp;
+    my @e = split /\t/;
+    my $v2    = $e[8];
+    $v2 =~ /Name=C(\d+)\_/; 
+    my $v3 = $1;
+    if ($topLevel < $v3)
+    {
+      $topLevel = $v3;
+    }
+  }
+  close PARSERNASEQ;
+
   open PARSERNASEQ, $parsernaseq or die "cannot open < $parsernaseq $!";
   while (<PARSERNASEQ>)
   {
@@ -285,11 +314,191 @@ elsif ($cmd eq "gff")
     my $end   = $e[4];
     my $v1    = $e[5];
     my $v2    = $e[8];
-    $v2 =~ /Name=(C\d+)\_/; 
+    $v2 =~ /Name=C(\d+)\_/; 
     my $v3 = $1;
-    print $outfile "chr1\t$start\t$end\t$v3\t$v1\t+\n";
[…]

### 2012-01-13 sh/bwa-pos2wig.sh

@@ -41,7 +41,7 @@ function bwa-pos2wig {
       REFGENOMELENGTH=$(grep REFGENOMELENGTH $SPECIESFILE | cut -d":" -f2)
       #for g in $(eval echo {1..$NUMFASTQFILE}); do
       for g in $(eval echo {1..1}); do
-        FASTQNUM=FASTQ$(printf "%02d" $g)
+        FASTQNUM=FASTQ$(printf "%03d" $g)
 
         # COMMAND1="perl pl/$FUNCNAME.pl perfect -in $BWADIR/$FASTQNUM-sum.pos -out $BWADIR/$FASTQNUM-perfect.wig"
         COMMAND1="perl pl/$FUNCNAME.pl end \
@@ -53,7 +53,7 @@ function bwa-pos2wig {
           echo $COMMAND1 >> $BATCHFILE
         else
           echo $COMMAND1 | bash
-          echo "Check $BWADIR/$FASTQNUM-perfect.wig"
+          echo "Check $BWADIR/$FASTQNUM-end.wig"
         fi
       done
 

### 2012-01-13 sh/transcript-parsernaseq.sh

@@ -40,31 +40,62 @@ function transcript-parsernaseq {
       NUMFASTQFILE=$(grep NUMFASTQFILE $SPECIESFILE | cut -d":" -f2)
       REFGENOMEFASTA=$(grep REFGENOMEFASTA $SPECIESFILE | cut -d":" -f2)
       # for g in $(eval echo {1..$NUMFASTQFILE}); do
-      for g in $(eval echo {1..1}); do
+      # for g in $(eval echo {1..1}); do
+      for g in 1 19 20 35 36; do
         FASTQNUM=FASTQ$(printf "%03d" $g)
 
         COMMAND1="perl pl/$FUNCNAME.pl pileup \
           -wiggle $BWADIR/$FASTQNUM.wig \
-          -out $BWADIR/$FASTQNUM.parsernaseq.pileup"
+          -out $BWADIR/$FASTQNUM.parsernaseq.pileup1"
         COMMAND2="perl pl/$FUNCNAME.pl gene \
           -feature $BWADIR/feature-genome.out-geneonly \
           -out $BWADIR/feature-genome.out-parsernaseq"
-        COMMAND3="ParseRNASeq \
+
+        # scp $BWADIR/$FASTQNUM.parsernaseq.pileup swiftgen:Documents/Projects/rnaseq/src/ParseRNAseq-1.1
+        # scp $BWADIR/feature-genome.out-parsernaseq swiftgen:Documents/Projects/rnaseq/src/ParseRNAseq-1.1
+
+        COMMAND3="ParseRNAseq \
           $BWADIR/$FASTQNUM.parsernaseq.pileup \
           $REFGENOMEFASTA \
-          $BWADIR/feature-genome.out-parsernaseq2 \
+          $BWADIR/feature-genome.out-parsernaseq \
           $BWADIR/$FASTQNUM.parsernaseq \
           -c 10 -b 25 -force_gp -fmt"
-        COMMAND4="perl pl/$FUNCNAME.pl bed \
+        COMMAND3a="perl pl/$FUNCNAME.pl remove \
+          -feature $BWADIR/feature-genome.out-parsernaseq \
+          -inpileup $BWADIR/$FASTQNUM.parsernaseq.pileup \
           -parsernaseq $BWADIR/$FASTQNUM.parsernaseq \
+          -out $BWADIR/$FASTQNUM.parsernaseq.pileup2"
+        COMMAND3b="perl pl/$FUNCNAME.pl remove \
[…]

### 2012-01-13 test/GenomicRanges/tutorial.R

@@ -1,0 +1,2 @@
+source("http://bioconductor.org/biocLite.R")
+biocLite("GenomicRanges")

### 2012-01-16 package/smutans/inst/doc/smutans.Rnw
Just an update.

@@
 
+\section{Prewriting about UA159 versus TW1}
+
+\subsection{Abstract}
+A bacterial transcriptome of the primary etiological agent of human dental caries,
+\textit{Stretptococcus mutans}, is described using RNA-seq data. Putative
+transcripts boundaries are defined. Small RNAs are predicted by computational
+methods, and confirmed by RNA-seq and Southern blot experiments. We refine the
+previously established importance of CcpA regulation as central metabolism and
+virulence gene expression using RNA-seq data.
+CcpA is known to play a direct role in carbon catabolite repression (CCR).
+We describe genes that are differentially expressed in glucose-grown wild-type
+(UA159) and CcpA-deficient (TW1) strains. We found that more
+genes were differentially expressed in the comparison between wild-type and
+CcpA-deficient strains in galactose-grown strains than in
+glucose-grown strains.
+We also found that more genes differentially expressed in wild-type than in
+CcpA-deficient strain when comparing glucose- and galactose-grown cells. These
+two findings were different from those of \cite{Abranches2008}. \fixme{We may
+want to see effects of different p-value cutoff.} We confirmed that CcpA
+modulates the pathogenic potential of \textit{S.\ mutans} through global control
+of gene expression.
+We predicted a few small RNAs based on sequence alignments and RNA secondary
+structures. We confirmed them experimentally. Target genes of predicted small
+RNAs were predicted. The target genes were enriched in some functional
+categories. We listed highly potential small RNAs with their target genes
+functional categories. 
+Transcripts were predicted using RNA-seq short reads mapped on the wild-type
+UA159 genome.
+
+\subsection{Reviews of papers that cited \cite{Abranches2008}}
+
+CCR in bacteria involves regulatory networks that activate or silence genes in
+response to carbon source and availability.
+
+\subsection{Goal: Refinement of \cite{Abranches2008}}
+
+Comparisons of microarray and RNA-seq are not good way to go.
+
+\subsection{Methods and Materials}
+
+\subsubsection{Bacterial strains and growth conditions}
+
+\subsubsection{RNA isolation}
+
+\subsubsection{RNA sequencing}
+
+\subsubsection{Statistical analysis for differential expressions}
+We used an R package, DESeq, \cite{Anders2010} to determine differentially
+expressed genes based on negative binomial model originally developed by \cite{Robinson2007}.
+Steps of analyzing RNA-seq data for determining differentially expressed genes
+are described in \cite{Oshlack2010}. We used a short read aligner called
+\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
+\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
+aligner allowed a few gaps for efficient alignment of millions of sequences of
+size being 100 base pairs long, short reads that consist of RNAs and sequencing
+adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
+remove parts of adapter sequences from short reads before mapping on the
+reference genome. Mapped short read alignments were converted into readable
+formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
+reference genome, forming a table of read counts of the genes. Statistical
+software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
+was subsequently employed to infer differentially expressed in different
+biological conditions.  We also generated tracks for our recently released
+Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
+summarize gene expression levels based the short read alignments. This allowed
+us to visualize parts of the reference genome for comparing different RNA
+samples. Further analyses such as functional category enrichment test
+\cite{Young2010} were also employed.
+
+\subsubsection{small RNAs}
+
+RNAz is used to find small RNA candidates. Sequence alignments were a
+critical step for finding small RNAs in the approach. Phylogenies of streptococcus species could
+be considered to choose species related with Streptococcus mutans.
+Alternatively, I find sequences similar to intergenic regions, and refine the
+alignments using a multiple sequence alignment tool such as MUSCLE
+\cite{Edgar2004a}. Steps for
+achieving this goal include: 1. Exract intergenic regions including up- and
+down-stream, 2. BLAST the intergenic regions against non-redundant DB to find
+similar sequences \cite{Altschul1990}, 3. Refine each alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the
+alignments for scoring regions for small RNAs
+\cite{Washietl2005,Gruber2010}, and 5. Find transcripts with high
+score of small RNAs. The transcripts must be defined in advance.
+
+RNAplex
+\cite{Tafer2008}
+
+A review paper about small RNA predictions and targets
+\cite{Tafer2008a}
+
+RNAplfold
+\cite{Bernhart2006}
+
+\subsubsection{Transcripts}
+
+``-c 10 -b 25 -force_gp''
+
+\subsubsection{Gene category associations and virulence genes}
+
+We assigned genes to Gene Ontology (GO) categories by comparing the {\em
+  Streptococcus} genes to bacterial proteins from the Uniref90 database
+using BLASTP, and then assigning the same GO classification as the target
+gene of the uniProt GOA database if the match had an $E$-value of
+$<1.0\times10^{-5}.$  
+A gene family was assigned a given
+classification if any of its genes was assigned that classification.  To
+test for associations with replacing gene transfers we used the
+recombination intensity estimated for each gene, and for additive gene
+transfers we used the number of additive transfers inferred for each
+family.  To test for significance, we performed a Mann-Whitney $U$ test of
+the values (recombination intensities or numbers of transfers) associated
+with a given category vs.\ the values for the all other genes/families.  We
+used the \cite{Benjamini1995} method to correct for multiple comparisons.
+
+\subsection{Results}
+
+\subsection{Discussion}
+
+\subsection{Abbreviations}
+
+\begin{enumerate}
+\item CCR Carbon catabolite repression
+\item CcpA ?
+\item IPS intracellular polysaccharide
+\end{enumerate}
+
 \section{Session Info}

### 2012-01-17 doc/manuscript.tex
Just an update.

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color} 
\usepackage{setspace} 
\usepackage{rotating}
\usepackage{natbib}
\usepackage{xr}
\usepackage{../../latex/sty/genres}

% Text layout
\oddsidemargin 0in
\evensidemargin 0in
\topmargin -.5in
\textwidth 6.5in
\textheight 9in
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\program}[1]{{\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

\makeatother
\pagestyle{myheadings}
\markright{Transcriptome of Streptococcus mutans}

\begin{document}

\begin{titlepage}

\title{Transcriptome \\
in {\em Streptococcus mutans} }

\author{Sang Chul Choi$^{1}$, Charles D.\ Danko$^{1}$, 
Melissa J.\ Hubisz$^{1}$, \\
Robert A.\ Burne$^{3}$
Michael J.\ Stanhope$^{2}$, and Adam Siepel$^{1}$}

\date{ }
\maketitle

\begin{footnotesize}
\begin{center}
$^1$Department of Biological Statistics and Computational Biology,\\
Cornell University, Ithaca, NY 14853, USA
\\[1ex]
$^2$Department of Population Medicine and Diagnostic Sciences,\\
College of Veterinary Medicine, Cornell University, Ithaca, NY 14853, USA
\\[1ex]
$^3$Department of Oral Biology,\\
College of Dentistry, University of Florida, Gainesville, Florida 32610 USA
\\
\end{center}
\end{footnotesize}

\vspace{1in}

\begin{tabular}{lp{4.5in}}
{\bf Submission type:}& Research Article
\vspace{1ex}\\
{\bf Keywords:}&Carbon catabolite repression, 
{\em Streptococcus mutans}
\vspace{1ex}\\
{\bf Running Head:}&Transcriptome of {\em Streptococcus mutans}
\vspace{1ex}\\ 
{\bf Corresponding Author:}&
\begin{minipage}[t]{4in}
 Adam Siepel\\
 102E Weill Hall, Cornell University\\
 Ithaca, NY 14853\\
 Phone: +1-607-254-1157\\
 Fax: +1-607-255-4698\\
 Email: acs4@cornell.edu
\end{minipage}
\end{tabular}

\thispagestyle{empty}
\end{titlepage}

\doublespacing

\section{Prewriting about UA159 versus TW1}

\subsection{Abstract}
A bacterial transcriptome of the primary etiological agent of human dental caries,
\textit{Stretptococcus mutans}, is described using RNA-seq data. Putative
transcripts boundaries are defined. Small RNAs are predicted by computational
methods, and confirmed by RNA-seq and Southern blot experiments. We refine the
previously established importance of CcpA regulation as central metabolism and
virulence gene expression using RNA-seq data.
CcpA is known to play a direct role in carbon catabolite repression (CCR).
We describe genes that are differentially expressed in glucose-grown wild-type
(UA159) and CcpA-deficient (TW1) strains. We found that more
genes were differentially expressed in the comparison between wild-type and
CcpA-deficient strains in galactose-grown strains than in
glucose-grown strains. How can I explain the reason?
We also found that more genes differentially expressed in wild-type than in
CcpA-deficient strain when comparing glucose- and galactose-grown cells. These
two findings were different from those of \cite{Abranches2008}. \fixme{We may
want to see effects of different p-value cutoff.} We confirmed that CcpA
modulates the pathogenic potential of \textit{S.\ mutans} through global control
of gene expression.
We predicted a few small RNAs based on sequence alignments and RNA secondary
structures. We confirmed them experimentally. Target genes of predicted small
RNAs were predicted. The target genes were enriched in some functional
categories. We listed highly potential small RNAs with their target genes
functional categories. 
Transcripts were predicted using RNA-seq short reads mapped on the wild-type
UA159 genome.

\subsection{Reviews of papers that cited \cite{Abranches2008}}

Bacteria can selectively use substrates from different carbon sources. In the
presence of a
preferred carbon source, catabolic system of secondary substrates are often
regulated or shutdown: called carbon catabolite repression (CCR)
\cite{Gorke2008}. CCR also regulates the expression of virulence factors in many
pathogenic bacteria. A classical example is the glucose-lactose diauxie in
\textit{E.\ coli}. A significant number of genes can be subject to CCR.  
CCR in bacteria involves regulatory networks that activate or silence genes in
response to carbon source and availability.
\textit{S.\ mutans} possesses redundant systems for CCR.
We could measure directly the expression levels of \textit{fruA} in \textit{S.\
mutans}. 
We assess the contribution of CcpA to CCR of predicted operons.
Clearly, deletion of \textit{ccpA} resulted in substantial relief of the CCR of
\textit{fruA}
expression exerted by both glucose and fructose.

\subsection{Goal: Refinement of \cite{Abranches2008}}

Comparisons of microarray and RNA-seq are not good way to go.

Find CCR-sensitive transcripts.

\subsection{Methods and Materials}

\subsubsection{Bacterial strains and growth conditions}
\textit{S.\ mutans} UA159 is a human cariogenic isolate. The whole genome is
availabe at NCBI \cite{Ajdic2002}. Strain TW1 was derived from UA159 by
replacing a part of \textit{ccpA} gene \cite{Wen2002}.

\textit{S.\ mutans} strains were inoculated into TV medium supplemented with
0.5\% galactose or glucose and grown to early exponential phase (OD600, 0.2).
Then fructose was added to the medium at concentrations (0,0.05,3,5,10 mM) for
3.5h to induce \textit{fruA} expression.

\subsubsection{RNA isolation}
RNA was isolated from S. mutans (1) by using 50 ml cultures that were grown
under the desired conditions and harvested by centrifugation at $4^{\circ}$C. Pelleted
cells were resuspended in 400 ul of diethyl pyrocarbonate-treated water, 800 ul
of RNAprotect reagent (Qiagen, Inc., Chatsworth, CA) was added, and the samples
were incubated at room temperature for 5 min with vortexing for 10 s at 1-min
intervals. Cells were then pelleted, resuspended in 250 ul Tris-EDTA (50:10)
buffer, and transferred to 1.5 ml screw-cap tubes containing sterile glass beads
(average diameter, 0.1 mm; Biospec, Bartlesville, OK), 10 ul of 1\% sodium
dodecyl sulfate, and 300 ul of acid phenol-chloroform (5:1). After
centrifugation for 15 min at maximum speed at $4^{\circ}$C, RNA was further purified
using an RNeasy mini kit (Qiagen, Inc., Chatsworth, CA), including on-column
DNase digestion with RNase-free DNase (Qiagen), as recommended by the supplier.

\subsubsection{RNA sequencing}
Sequencing libraries for the Illumina GA platform were constructed by shearing
the enriched cDNA by nebulisation (35psi, 6 min) followed by end-repair with
Klenow polymerase, T4 DNA polymerase and T4 polynucleotide kinase (to blunt-end
the DNA fragments). A single 3' adenosine moiety was added to the cDNA using
Klenow exo- and dATP. The Illumina adapters (containing primer sites for
sequencing and flowcell surface annealing) were ligated onto the repaired ends
on the cDNA and gel-electrophoresis was used to separate library DNA fragments
from unligated adapters by selecting cDNA fragments between 200-250 bps in size.
Fragmentation followed by gel-electrophoresis were used to separate library DNA
fragments and size fragments were recovered following gel extraction at room
temperature to ensure representation of AT rich sequences. Ligated cDNA
fragments were recovered following gel extraction at room temperature to ensure
representation of AT rich sequences. Libraries were amplified by 18 cycles of
PCR with Phusion polymerase. Sequencing libraries were denatured with sodium
hydroxide and diluted to 3.5 pM in hybridisation buffer for loading onto a
single lane of an Illumina GA flowcell. Cluster formation, primer hybridisation
and single-end, 36 cycle sequencing were performed using proprietary reagents
according to manufacturers' recommended protocol (https://icom.illumina.com/).
The efficacy of each stage of library construction was ascertained in a quality
control step that involved measuring the adapter-cDNA on a Agilent DNA 1000
chip. A final dilution of 2 nM of the library was loaded onto the sequencing
machine.

SOLiD sequencing was performed at Agencourt Bioscience (Beverly, MA) and
SeqWright (Houston, TX). Library preparations, fragment library protocols, and
all SOLiD-run parameters followed standard Applied Biosystems protocols.
Illumina/Solexa sequencing followed standard Illumina/Solexa methods and
protocols. Both systems were used because in this study our aim was simply to
collect as much sequence coverage as possible, and in pursuing this goal, we
utilized all available sequencing systems. We note, however, that the relative
performances (overall coverage, data output, etc.) noted in Table 1 for the
Illumina and Applied Biosystems sequencing platforms cannot be compared
directly. This study was not intended or designed to be a direct comparison of
the two systems, and interpreting the differences shown in Table 1 is difficult
given the many differences in cDNA/library preparation and the rapid evolution
of each system over the past year.

\subsubsection{Statistical analysis for differential expressions}
We used an R package, DESeq, \cite{Anders2010} to determine differentially
expressed genes based on negative binomial model originally developed by \cite{Robinson2007}.
Steps of analyzing RNA-seq data for determining differentially expressed genes
are described in \cite{Oshlack2010}. We used a short read aligner called
\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
aligner allowed a few gaps for efficient alignment of millions of sequences of
size being 100 base pairs long, short reads that consist of RNAs and sequencing
adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
remove parts of adapter sequences from short reads before mapping on the
reference genome. Mapped short read alignments were converted into readable
formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
was subsequently employed to infer differentially expressed in different
biological conditions.  We also generated tracks for our recently released
Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
summarize gene expression levels based the short read alignments. This allowed
us to visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functional category enrichment test
\cite{Young2010} were also employed.

\subsubsection{small RNAs}

RNAz is used to find small RNA candidates. Sequence alignments were a
critical step for finding small RNAs in the approach. Phylogenies of streptococcus species could
be considered to choose species related with Streptococcus mutans.
Alternatively, I find sequences similar to intergenic regions, and refine the
alignments using a multiple sequence alignment tool such as MUSCLE
\cite{Edgar2004a}. Steps for
achieving this goal include: 1. Exract intergenic regions including up- and
down-stream, 2. BLAST the intergenic regions against non-redundant DB to find
similar sequences \cite{Altschul1990}, 3. Refine each alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the
alignments for scoring regions for small RNAs
\cite{Washietl2005,Gruber2010}, and 5. Find transcripts with high
score of small RNAs. The transcripts must be defined in advance.

RNAplex
\cite{Tafer2008}

A review paper about small RNA predictions and targets
\cite{Tafer2008a}

RNAplfold
\cite{Bernhart2006}

\subsubsection{Transcripts}

``-c 10 -b 25 -force\_gp''

\subsubsection{Gene category associations and virulence genes}

We assigned genes to Gene Ontology (GO) categories by comparing the {\em
  Streptococcus} genes to bacterial proteins from the Uniref90 database
using BLASTP, and then assigning the same GO classification as the target
gene of the uniProt GOA database if the match had an $E$-value of
$<1.0\times10^{-5}.$  
A gene family was assigned a given
classification if any of its genes was assigned that classification.  To
test for associations with replacing gene transfers we used the
recombination intensity estimated for each gene, and for additive gene
transfers we used the number of additive transfers inferred for each
family.  To test for significance, we performed a Mann-Whitney $U$ test of
the values (recombination intensities or numbers of transfers) associated
with a given category vs.\ the values for the all other genes/families.  We
used the \cite{Benjamini1995} method to correct for multiple comparisons.

\subsection{Results}

CcpA affects CCR of \textit{fruA}. 
\textit{fruA} fructan hydrolase
When Wen2002 deleted CRE sequencs in the \textit{fruA} promoter region, CCR was
almost completely alleviated. 
What are CCR-sensitive genes? 
What are the mechanism of CcpA on CCR of the \textit{fruA} operon? 
Expression levels of \textit{fruA} in \textit{S.\ mutans} strains that were
grown to early exponential phase (OD600, 0.2) in TV medium supplemented with
0.5\% glucose or galactose. \fixme{What does this mean?} 
Deletion of \textit{ccpA} resulted in substantial relief of the CCR of
\textit{fruA} expression exerted by both glucose and fructose. 
CcpA could directly interact with the \textit{fruA} promoter region.

The role of CcpA in gene expression during growth on glucose. 
\textit{S.\ mutans} UA159 and TW1 grown under catabolite-repressing conditions
(glucose), and catabolite-derepressing conditions (galactose). Under
glucose-only condition I found xx number of differentially expressed genes. What
are the functional categories of those differentially expressed genes? What are
the genes that are most profoundly up- or down-regulated (Smu.1421 to Smu.1424,
PDH enzyme complex), ... and more? ... This regulatory control was
short-circuited in the strain growing on glucose and lacking CcpA. I showed that
some of KEGG pathways were affected by TW1 grown under glucose. We found
down-regulated genes, which would have something to do with pathogenicity. 
Upto here, I compared UA159 and TW1 under glucose.

1. UA159 (A) vs. TW1 (B) : fold change is B/A.
When glucose was the sole carbohydrate for growth, 170 genes were differentially
expressed at least twofold as a result of loss of CcpA (Fig. 2; see Table S1 in
the supplemental material). Of the gene products, 22 were predicted to partici-
pate in energy metabolism, 9 were PTS components, and 43 were hypothetical
proteins (Fig. 2). Among the most pro- foundly up-regulated genes in the ccpA
mutant strain were the genes that encoded the components of the PDH enzyme com-
plex (Smu.1421 to Smu.1424), pyruvate-formate lyase, enzymes in the
tricarboxylic acid (TCA) cycle, selected sugar-specific PTS permeases, and
components of the glycogen biosynthetic machinery (see Table S1 in the
supplemental material). 
Con- sistent with the profile data, TW1 cells grown in glucose had 78% higher
PDH activity than the wild-type strain (data not shown). Given the impact on
expression of PDH, TCA en- zymes, and selected glycolytic enzymes, CcpA clearly
legislates gene expression to discriminate between carbohydrate-replete and
carbohydrate-limiting conditions. In particular, when there is excess
carbohydrate, S. mutans moves carbohydrates primar- ily through lactate
dehydrogenase, sacrificing an ATP and gen- erating NAD􏰜. When carbohydrate is
limiting, S. mutans pro- duces mainly formic and acetic acids, shunts
carbohydrate through pyruvate formate lyase, and uses its partial TCA cycle to
regenerate reducing equivalents to maintain an appropriate NAD/NADH balance.
This regulatory control was clearly short-circuited in the strain growing on
glucose and lacking CcpA.
Among the more interesting genes that were down-regu- lated, significant
decreases in expression were noted for the genes encoding components of the
proton-translocating F-ATPase, phosphoglycerate kinase (Pgk), and
fructosyltrans- ferase (ftf), which catalyzes the conversion of sucrose to high-
molecular-weight polymers of fructose, as well as gtfD, which encodes the
glucosyltransferase D enzyme that produces pre- dominantly 􏰐-1,6-linked polymers
of glucose from sucrose. Since acid tolerance (ATPase), acid production from
carbohy- drate (Pgk), and glucan and fructan production (Gtf and Ftf), as well
as the ability to transport carbohydrates and produce intracellular storage
compounds (glycogen), as noted above (PTS and Glg), are central to the
pathogenic potential of S. mutans, CcpA represents a major control point for
expression of critical virulence attributes in cells presented with glucose.

Growth in derepressing conditions induces major changes in the UA159
transcriptome. Here, I compared glucose and galactose conditions for growing
UA159. Energy metabolism was enriched with differentially expressed genes. There
were more up-regulated genes in galactose condition compared with glucose
condition. Some genes are described in detail.

2. I compared glucose (A) and galactose (B) for growing UA159
It is noteworthy that only three genes in the energy metabolism category were
down-regulated, whereas 23 genes in this category were markedly up-regulated. In
addition, the expression of 28 genes encoding hypothetical proteins was affected,
and 19 of them were up-regulated. The genes encod- ing proteins belonging to the
pathways for lactose and galac- tose utilization (2) were the most strongly
induced genes in cells growing on galactose. The galKTE genes, encoding the
enzymes for galac- tose utilization via the Leloir pathway (2), were
up-regulated between 10- and 36-fold, but the genes of the lactose (lac) operon,
encoding the lactose PTS permease and the tagatose- 6-phosphate pathway, were
up-regulated 70- to 300-fold. Since the products of the galKTE operon contribute
to the genera- tion of UDP-galactose, which is required for cell wall biogen-
esis, the lower level of induction of galKTE than of the lac operon may reflect
a need for higher constitutive levels of expression of the Leloir pathway in the
absence of galactose. Components of the multiple-sugar metabolism operon (msm)
were also induced in cells growing on galactose. The Msm pathway is induced by
raffinose and melibiose, both of which are ?-galactosides (5). The enhanced
expression of gene products involved in en- ergy metabolism, including pyruvate
formate lyase and com- ponents of the TCA cycle, in cells growing on galactose
is consistent with the idea that the cells perceive carbohydrate limitation.

Now, I compared glucose (A) and galactose (B) for growing TW1.

Now, I compared UA159 and TW1 under galactose.

What genes are differentially expressed?

Name the transcripts.
I found xx transcripts. Among them xx \% were mono-cistronic operons. xx \% were
poly-cistronic operons. xx transcripts could not defined because RNA-seq short
reads were not precise. Transripts were assigned with functional categories, and
differentially expressed transcripts were found. Transcripts and functional
categories: how can I assess that transcripts consist of genes with similar
functional cateogires.

Name the small RNAs.
RNAz predicted a few small RNAs. Some of them were differentially expressed.
Some functional categories were enriched with target genes of the predicted
genes.

\subsection{Discussion}
We used RNA-seq data to delineate transcript boundaries. Although this map of
transcripts does not precisely tell us where transcripts start and end, it
informs us of which open reading frames can be transcribed together. Potentially
polycistronic operons are defined. We focused on interpretations of operon in
light of CcpA regulations.

\subsection{Abbreviations}

\begin{enumerate}
\item CCR Carbon catabolite repression
\item CcpA ?
\item IPS intracellular polysaccharide
\item CRE catabolite responsive elements
\end{enumerate}

\renewcommand*{\refname}{Literature Cited}
\bibliographystyle{../../latex/bst/mbe}
\bibliography{siepel-mutans}

\end{document}

### 2012-01-18 doc/manuscript.Rnw
Just an update.

--- 2012-01-17 doc/manuscript.tex
@@
 
-\title{Transcriptome \\
-in {\em Streptococcus mutans} }
+\title{Defining Transcriptome using RNA Sequencing\\
+of {\em Streptococcus mutans} }
 
@@
 \vspace{1ex}\\
-{\bf Keywords:}&Carbon catabolite repression, 
-{\em Streptococcus mutans}
+{\bf Keywords:}&Carbon catabolite repression, RNA-seq, non-coding RNAs
 \vspace{1ex}\\
@@
 
-\section{Prewriting about UA159 versus TW1}
+\section{Abstract}
+A bacterial transcriptome from the primary etiological agent of human dental
+caries, \textit{Stretptococcus mutans}, is described using RNA deep sequencing
+data.  We define transcripts boundaries of open reading frames based on sequence
+alignments of short reads on the reference genome of UA159 from the deep
+sequencing.  Based on homologous sequences and RNA secondary structures we also
+report putative small RNAs, several of which are confirmed experimentally using
+both RNA sequencing data, northern blot, and qRT-PCR.  CcpA is known to play a
+direct role in carbon catabolite repression (CCR).  We use the developed
+transcript map, open read frame, and small RNAs to determine differentially
+expressed regions between the wild-type (UA159) and CcpA-deletion (TW1) strains
+in order to refine the previously established importance of CcpA regulation as
+central metabolism and virulence gene expression.  We find that more genes are
+differentially expressed in the comparison between wild-type and CcpA-deletion
+strains that are grown in galactose than in glucose.  We also find that more
+genes differentially expressed in wild-type than in CcpA-deletion strain when
+comparing glucose- and galactose-grown cells.  These two findings are different
+from those from microarry experiment of \cite{Abranches2008}.  We also predicted
+target genes of the small RNAs that were enriched in the following functional
+categories: xxx.  We confirme that CcpA modulates the pathogenic potential of
+\textit{S.\ mutans} through global control of gene expression.
 
-\subsection{Abstract}
-A bacterial transcriptome of the primary etiological agent of human dental caries,
-\textit{Stretptococcus mutans}, is described using RNA-seq data. Putative
-transcripts boundaries are defined. Small RNAs are predicted by computational
-methods, and confirmed by RNA-seq and Southern blot experiments. We refine the
-previously established importance of CcpA regulation as central metabolism and
-virulence gene expression using RNA-seq data.
-CcpA is known to play a direct role in carbon catabolite repression (CCR).
-We describe genes that are differentially expressed in glucose-grown wild-type
-(UA159) and CcpA-deficient (TW1) strains. We found that more
-genes were differentially expressed in the comparison between wild-type and
-CcpA-deficient strains in galactose-grown strains than in
-glucose-grown strains. How can I explain the reason?
-We also found that more genes differentially expressed in wild-type than in
-CcpA-deficient strain when comparing glucose- and galactose-grown cells. These
-two findings were different from those of \cite{Abranches2008}. \fixme{We may
-want to see effects of different p-value cutoff.} We confirmed that CcpA
-modulates the pathogenic potential of \textit{S.\ mutans} through global control
-of gene expression.
-We predicted a few small RNAs based on sequence alignments and RNA secondary
-structures. We confirmed them experimentally. Target genes of predicted small
-RNAs were predicted. The target genes were enriched in some functional
-categories. We listed highly potential small RNAs with their target genes
-functional categories. 
-Transcripts were predicted using RNA-seq short reads mapped on the wild-type
-UA159 genome.
+\section{Abbreviations}
 
-\subsection{Reviews of papers that cited \cite{Abranches2008}}
+\begin{enumerate}
+\item CCR Carbon catabolite repression
+\item CcpA ?
+\item IPS intracellular polysaccharide
+\item CRE catabolite responsive elements
+\end{enumerate}
+
+\section{Introduction}
+Regulation of carbon catabolite repression (CCR) in \textit{Streptococcus
+mutans} can involve a phosphoenolpyruvate:sugar phosphotransferase system (PTS),
+histidin protein (HPr) and a transcriptional regulator, CcpA \citep{Gorke2008}.
+CcpA was shown to regulate metabolism and virulence gene expression in
+\textit{S.\ mutans} by \citet{Abranches2008}. They used microarray experiments
+to measure expression levels of genes. Microarray experiments tend to be
+qualitative. The advent of high throughput sequencing technology motivated us to
+revisit the experiment to refine the result.  RNA deep sequencing data also
+allowed us to define the bacterial transcriptome of \textit{S.\ mutans}. So, we
+could compare putative operon expression levels as well as gene expression
+levels between the wild-type (UA159) and its mutant-type (TW1) whose
+\textit{ccpA} was deleted from the wild-type. This allowed us to profile
+expression of genes and putative operons affected by the regulation of
+\textit{ccpA}.
+
+\citet{Martin2010} developed a hidden Markov model to predict transcripts by
+using site-wise expression levels. I also refined the prediction of transcripts
+using multiple samples. I measured expression levels of known genes and compared
+them from one experiment condition and from another. I can test whether two
+genes are in the same operon or not. I assume that expression levels of two genes 
+vary in similar fashion if the two genes belong to the same operon. What I mean
+by similar fashion is whether they are expressed differentially with similar
+fold changes. I assume that operons maps do not change between the two
+experiments, or multiple experiments. This is not true. I hope that additional
+samples could help to define operons. Eventually, operon maps will be determined
+experimentally. Computational tools could be helpful before those ages come to
+us. \fixme{This idea did not work. We should only use the prediction method.}
+
+Although our RNA-seq data did not sampled specifically non-coding small RNAs, we
+retained small RNAs in the sample by attempting to deplete only ribosomal RNAs
+from the sample. Because non-coding small RNAs and coding messenger RNAs were
+mixed in the sample, we sought a way of predicting small RNAs using
+\program{RNAz} developed by \citet{Gruber2010}.  \program{RNAz} used homologous
+sequences and RNA secondary structures to predict putative non-coding RNAs. We
+confirmed the several predicted small non-coding RNAs experimentally
+\citep{Altuvia2007} including the RNA-seq data set. Because small RNAs could
+be associated with CCR regulation of genes (CITE), we determined differentially expressed small
+RNAs between RNA samples. We found a few small RNAs to be differentially
+expressed.
 
 Bacteria can selectively use substrates from different carbon sources. In the
-presence of a
-preferred carbon source, catabolic system of secondary substrates are often
-regulated or shutdown: called carbon catabolite repression (CCR)
+presence of a preferred carbon source, catabolic system of secondary substrates
+are often regulated or shutdown: called carbon catabolite repression (CCR)
 \cite{Gorke2008}. CCR also regulates the expression of virulence factors in many
 pathogenic bacteria. A classical example is the glucose-lactose diauxie in
-\textit{E.\ coli}. A significant number of genes can be subject to CCR.  
-CCR in bacteria involves regulatory networks that activate or silence genes in
-response to carbon source and availability.
-\textit{S.\ mutans} possesses redundant systems for CCR.
-We could measure directly the expression levels of \textit{fruA} in \textit{S.\
-mutans}. 
-We assess the contribution of CcpA to CCR of predicted operons.
-Clearly, deletion of \textit{ccpA} resulted in substantial relief of the CCR of
-\textit{fruA}
-expression exerted by both glucose and fructose.
+\textit{E.\ coli}. A significant number of genes can be subject to CCR.  CCR in
+bacteria involves regulatory networks that activate or silence genes in response
+to carbon source and availability.  \textit{S.\ mutans} possesses redundant
+systems for CCR.  We could measure directly the expression levels of
+\textit{fruA} in \textit{S.\ mutans}.  We assess the contribution of CcpA to CCR
+of predicted operons.  Clearly, deletion of \textit{ccpA} resulted in
+substantial relief of the CCR of \textit{fruA} expression exerted by both
+glucose and fructose.
 
-\subsection{Goal: Refinement of \cite{Abranches2008}}
+We aims at refining the results of \cite{Abranches2008} in more quantative ways.
+Find CCR-sensitive transcripts. We describe predictions made for transcripts and
+small RNAs based on RNA-seq data. Combining the predictions and known genes we
+determine differentially expressed regions in UA159 genome.
 
-Comparisons of microarray and RNA-seq are not good way to go.
+We need to quantify some uncertainty of transcript predictions.
 
-Find CCR-sensitive transcripts.
+\section{Methods and Materials}
 
-\subsection{Methods and Materials}
-
-\subsubsection{Bacterial strains and growth conditions}
+\subsection{Bacterial strains and growth conditions}
+\fixme{Burne group should be able to describe this part. The following is just
+from somewhere for the purpose of writing template.}
 \textit{S.\ mutans} UA159 is a human cariogenic isolate. The whole genome is
 availabe at NCBI \cite{Ajdic2002}. Strain TW1 was derived from UA159 by
-replacing a part of \textit{ccpA} gene \cite{Wen2002}.
+replacing a part of \textit{ccpA} gene \cite{Wen2002}.  \textit{S.\ mutans}
+strains were inoculated into TV medium supplemented with 0.5\% galactose or
+glucose and grown to early exponential phase (OD600, 0.2).  Then fructose was
+added to the medium at concentrations (0,0.05,3,5,10 mM) for 3.5h to induce
+\textit{fruA} expression.
 
-\textit{S.\ mutans} strains were inoculated into TV medium supplemented with
-0.5\% galactose or glucose and grown to early exponential phase (OD600, 0.2).
-Then fructose was added to the medium at concentrations (0,0.05,3,5,10 mM) for
-3.5h to induce \textit{fruA} expression.
-
-\subsubsection{RNA isolation}
+\subsection{RNA isolation}
+\fixme{Burne group should be able to describe this part. The following is just
+from somewhere for the purpose of writing template.}
 RNA was isolated from S. mutans (1) by using 50 ml cultures that were grown
@@
 
-\subsubsection{RNA sequencing}
+\subsection{RNA sequencing}
+\fixme{Stanhope group should be able to describe this part. The following is just
+from somewhere for the purpose of writing template.}
 Sequencing libraries for the Illumina GA platform were constructed by shearing
@@
 
-\subsubsection{Statistical analysis for differential expressions}
+\subsection{Statistical analysis for differential expressions}
 We used an R package, DESeq, \cite{Anders2010} to determine differentially
@@
 
-\subsubsection{small RNAs}
+\subsection{Transcript predictions}
+\citet{Martin2010} developed a hidden Markov model to predict transcripts by
+using site-wise expression levels. We used the following options for the
+program: ``-c 10 -b 25 -force\_gp''. We binned expression levels into 25 parts.
+We allowed 10 emission states. We also used the genes as the predicted parts of
+transcripts. 
+Because we did not have strand-specific RNA-seq of the 13 samples due to
+limited resources, we could not tell in which directions transcripts based on
+RNA-seq data were oriented.
 
-RNAz is used to find small RNA candidates. Sequence alignments were a
-critical step for finding small RNAs in the approach. Phylogenies of streptococcus species could
-be considered to choose species related with Streptococcus mutans.
+\subsection{small RNAs}
+RNAz is used to find small RNA candidates. Sequence alignments were a critical
+step for finding small RNAs in the approach. Phylogenies of streptococcus
+species could be considered to choose species related with Streptococcus mutans.
 Alternatively, I find sequences similar to intergenic regions, and refine the
 alignments using a multiple sequence alignment tool such as MUSCLE
-\cite{Edgar2004a}. Steps for
-achieving this goal include: 1. Exract intergenic regions including up- and
-down-stream, 2. BLAST the intergenic regions against non-redundant DB to find
-similar sequences \cite{Altschul1990}, 3. Refine each alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the
-alignments for scoring regions for small RNAs
-\cite{Washietl2005,Gruber2010}, and 5. Find transcripts with high
-score of small RNAs. The transcripts must be defined in advance.
+\cite{Edgar2004a}. Steps for achieving this goal include: 1. Exract intergenic
+regions including up- and down-stream, 2. BLAST the intergenic regions against
+non-redundant DB to find similar sequences \cite{Altschul1990}, 3. Refine each
+alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the alignments for
+scoring regions for small RNAs \cite{Washietl2005,Gruber2010}, and 5. Find
+transcripts with high score of small RNAs. The transcripts must be defined in
+advance.  We used RNAplex \citep{Tafer2008} to predict targets of putative small
+RNAs. The method used RNAplfold \citep{Bernhart2006}.  Please, find a review
+paper about small RNA predictions and targets by \citet{Tafer2008a}.
 
-RNAplex
-\cite{Tafer2008}
+\subsection{Gene category associations and virulence genes}
+We used two approaches for finding functional categories
+\citep{Young2010,Suzuki2011}. We briefly descirbe We briefly descirbe the
+prodecure of \citet{Suzuki2011}.  We assigned genes to Gene Ontology (GO)
+categories by comparing the {\em Streptococcus} genes to bacterial proteins from
+the Uniref90 database using BLASTP, and then assigning the same GO
+classification as the target gene of the uniProt GOA database if the match had
+an $E$-value of $<1.0\times10^{-5}.$  A gene family was assigned a given
+classification if any of its genes was assigned that classification.  To test
+for associations with replacing gene transfers we used the recombination
+intensity estimated for each gene, and for additive gene transfers we used the
+number of additive transfers inferred for each family.  To test for
+significance, we performed a Mann-Whitney $U$ test of the values (recombination
+intensities or numbers of transfers) associated with a given category vs.\ the
+values for the all other genes/families.  We used the \cite{Benjamini1995}
+method to correct for multiple comparisons.
 
-A review paper about small RNA predictions and targets
-\cite{Tafer2008a}
-
-RNAplfold
-\cite{Bernhart2006}
-
-\subsubsection{Transcripts}
-
-``-c 10 -b 25 -force\_gp''
-
-\subsubsection{Gene category associations and virulence genes}
-
-We assigned genes to Gene Ontology (GO) categories by comparing the {\em
-  Streptococcus} genes to bacterial proteins from the Uniref90 database
-using BLASTP, and then assigning the same GO classification as the target
-gene of the uniProt GOA database if the match had an $E$-value of
-$<1.0\times10^{-5}.$  
-A gene family was assigned a given
-classification if any of its genes was assigned that classification.  To
-test for associations with replacing gene transfers we used the
-recombination intensity estimated for each gene, and for additive gene
-transfers we used the number of additive transfers inferred for each
-family.  To test for significance, we performed a Mann-Whitney $U$ test of
-the values (recombination intensities or numbers of transfers) associated
-with a given category vs.\ the values for the all other genes/families.  We
-used the \cite{Benjamini1995} method to correct for multiple comparisons.
-
-\subsection{Results}
+\section{Results}
 
@@
 expressed at least twofold as a result of loss of CcpA (Fig. 2; see Table S1 in
-the supplemental material). Of the gene products, 22 were predicted to partici-
+the supplemental material). Of the gene products, 22 were predicted to partici
 pate in energy metabolism, 9 were PTS components, and 43 were hypothetical
-proteins (Fig. 2). Among the most pro- foundly up-regulated genes in the ccpA
+proteins (Fig. 2). Among the most profoundly up-regulated genes in the ccpA
 mutant strain were the genes that encoded the components of the PDH enzyme com-
@@
 supplemental material). 
-Con- sistent with the profile data, TW1 cells grown in glucose had 78% higher
+Consistent with the profile data, TW1 cells grown in glucose had 78\% higher
 PDH activity than the wild-type strain (data not shown). Given the impact on
-expression of PDH, TCA en- zymes, and selected glycolytic enzymes, CcpA clearly
+expression of PDH, TCA enzymes, and selected glycolytic enzymes, CcpA clearly
 legislates gene expression to discriminate between carbohydrate-replete and
@@
 carbohydrate, S. mutans moves carbohydrates primar- ily through lactate
-dehydrogenase, sacrificing an ATP and gen- erating NAD􏰜. When carbohydrate is
+dehydrogenase, sacrificing an ATP and generating NAD When carbohydrate is
 limiting, S. mutans pro- duces mainly formic and acetic acids, shunts
@@
 molecular-weight polymers of fructose, as well as gtfD, which encodes the
-glucosyltransferase D enzyme that produces pre- dominantly 􏰐-1,6-linked polymers
+glucosyltransferase D enzyme that produces predominantly -1,6-linked polymers
 of glucose from sucrose. Since acid tolerance (ATPase), acid production from
@@
 
-\subsection{Discussion}
+\section{Discussion}
 We used RNA-seq data to delineate transcript boundaries. Although this map of
@@
 
-\subsection{Abbreviations}
-
-\begin{enumerate}
-\item CCR Carbon catabolite repression
-\item CcpA ?
-\item IPS intracellular polysaccharide
-\item CRE catabolite responsive elements
-\end{enumerate}
-
 \renewcommand*{\refname}{Literature Cited}

### 2012-01-18 doc/manuscript.Rnw
Just an update.

@@
 
+\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=6,height=6}
 \newcommand{\Rpackage}[1]{R package \textit{#1}}
@@
 A bacterial transcriptome from the primary etiological agent of human dental
-caries, \textit{Stretptococcus mutans}, is described using RNA deep sequencing
+caries, \textit{Stretptococcus mutans}, is described using deep RNA sequencing
 data.  We define transcripts boundaries of open reading frames based on sequence
@@
 qualitative. The advent of high throughput sequencing technology motivated us to
-revisit the experiment to refine the result.  RNA deep sequencing data also
+revisit the experiment to refine the result.  The deep RNA sequencing data also
 allowed us to define the bacterial transcriptome of \textit{S.\ mutans}. So, we
@@
 
-\section{Methods and Materials}
+\section{Materials and Methods}
 
@@
 
+Findings include:
+\begin{enumerate}
+\item New genes, open reading frames that are not annotated in NCBI's genome database.
+\end{enumerate}
+
+We sequenced a whole prokaryotic transcriptome of \textit{S\. mutans}, and
+mapped short reads to the genomic sequence of the reference genome UA159
+\citep{Ajdic2002}. 
+\citet{Martin2010} applied a hidden Markov model to use short reads mapped to a
+genomic sequence in order to parase \textit{B.\ anthracis} genome into a
+sequence of transcripts. Most of annotated genes belong to the predicted
+transcripts. We found discrepancies of annotated genes and their transcriptions:
+1. A gene contains two transcripts, 
+2. Non-coding regions contain transcripts.
+\fixme{We could find more discrepancies. Try to find them}. 
+Expression levels of adjacent genes within a transcript should be more
+correlated to each other than those of adjacent genes belonging to separate transcripts
+\citep{Martin2010}.
+
+We identified a set of transcripts, albeit not a full, including small RNAs, and novel
+transcripts. 
+We hope that our study could serve as a foundation for a comprehensive study of
+the \textit{S.\ mutans} transcriptome.
+Our current knowledge of the transcriptome is limited to annotated or predicted genes.
+
+Whole transcriptome sequencing using high throughput sequencing (HTS)
+technologies, or deep RNA sequencing (RNA-seq) allowed to reveal the
+transcriptome topography of \textit{S.\ mutans}.
+
+\subsection{From Reviews}
+We reconstructed the full-length transcripts by transcriptome assembly using
+xxx.
+
+We tried in vain to use short-read assemblers such as Velvet \citep{Zerbino2008} that had been developed to
+assembly transcriptome.
+Trinity
+
+gene-dense transcriptomes
+
+We ensured a high-quality transcriptome assembly.
+
+In the data generation phase (FIG. 1a), total RNAs or mRNAs are fragmented and
+converted into a library of cDNAs containing sequencing adaptors.  
+In the data analysis phase (FIG. 1b), these short reads are pre-processed to
+remove sequencing errors and other artefacts. The reads are subsequently
+assembled to reconstruct the original RNAs and to assess their abun- dance
+(‘expression counting’).
+
+To increase the number of assem- bled transcripts, especially the less abundant
+ones, ribosomal RNA (rRNA) and abundant transcripts are removed during the
+first steps of library construction.
+In order to retain RNAs without a poly(A) tail in the assem- bled transcriptome,
+rRNA contamination can instead be removed by hybridization-based depletion
+methods. \fixme{What depletion method did we use?} 
+Last, the use of strand-specific RNA-seq protocols aids in the assembly and
+quantification of overlapping transcripts that are derived from opposite strands
+of the genome. This consideration is especially important for gene-dense
+genomes, such as those of bacteria, archaea and lower eukaryotes.
+
+Sequencing. The major factors to consider before sequencing a sample are: the
+choice of sequencing plat- form, the sequencing read length and whether to use a
+paired-end protocol.
+
+Removing artefacts from RNA- seq data sets before assembly improves the read
+quality, which, in turn, improves the accuracy and computa- tional efficiency of
+the assembly. This step is straight- forward and can be executed using several
+tools.  In general, three types of artefacts should be removed from raw RNA-seq
+data: sequencing adaptors, which origi- nate from failed or short DNA
+insertions during library preparation; low-complexity reads; and
+near-identical reads that are derived from PCR amplification. 
+Sequencing errors in NGS reads can be removed or corrected by analysing the
+quality score and/or the k-mer frequency.
+
+Reference-based transcriptome assembly is easier to perform for the simple
+transcriptomes of bac- terial, archaeal and lower eukaryotic organisms, as these
+organisms have few introns and little alternative splicing. Transcription
+boundaries can be inferred from regions of contiguous read coverage in the
+genome even with- out graph construction and traversal37,51,52. Alternative
+transcription start and stop sites can also be inferred based on the 5′ cap or
+poly(A) signals (if cap- or end- specific experimental protocols are used)51,53.
+However, complications arise owing to the gene-dense nature of
+these genomes. Many genes overlap, resulting in adja- cent genes being assembled
+into one transcript, even though they are not from a polycistronic RNA. Strand-
+specific RNA-seq has successfully been used to separate adjacent overlapping
+genes from opposite strands in the genome51,52. Overlapping genes that are
+transcribed from the same strand and that also have comparable expres- sion
+levels cannot easily be separated without using cap- or end-specific RNA-seq.
+
+In summary, reference-based assembly is generally preferable for cases in which
+a high-quality reference genome already exists.
+
+Meanwhile, experimental RNA-seq and sequencing protocols are continually
+improving and should greatly reduce the informatics challenges.  RNA-seq reads
+from third-generation sequencers, such as PacBio70, are longer (up to several
+kilobases). PacBio sequencers are capable of sequencing a single transcript to
+its full length in a single read. If this technology reaches a throughput that
+is comparable to the second-generation technologies, then the need for
+transcriptome assembly will probably be eliminated. Hopefully, the future of
+transcriptome assembly will be ``no assembly required''.
+
+What is the sequencing depth?
+
+\fixme{I might have to try different filters of short reads to see how those
+affect the transcript prediction.}
+
+\subsection{My Results}
+
+The input RNA-seq data is a list of site-wise non-negative integers that
+represents a number of reads mapped to a particular genomic position
+\citet{Martin2010}; a RNA-seq ``coverage'' data. We designated a gene as an
+expressed if the gene belongs to a transcript with at least C2 level.
+
+A gene
+
+We use the term `coverage' to designate 
+We verified the genes using RNA-seq data.
+
 CcpA affects CCR of \textit{fruA}. 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{manuscript-figClust}
+\caption{Heatmap showing the Euclidean distances between the samples
+  as calculated from the variance-stabilizing transformation of the
+  count data.}
+\label{figClust}
+\end{figure}
+<<figClust,fig=TRUE,echo=FALSE>>=
+x <- rnorm(100)
+hist(x)
+@
+
 \end{document}

### 2012-01-18 doc/manuscript.tex
Just an update.

@@
 
+\usepackage{Sweave}
 \begin{document}
@@
 
-\section{Prewriting about UA159 versus TW1}
+\section{Abstract}
+A bacterial transcriptome from the primary etiological agent of human dental
+caries, \textit{Stretptococcus mutans}, is described using deep RNA sequencing
+data.  We define transcripts boundaries of open reading frames based on sequence
+alignments of short reads on the reference genome of UA159 from the deep
+sequencing.  Based on homologous sequences and RNA secondary structures we also
+report putative small RNAs, several of which are confirmed experimentally using
+both RNA sequencing data, northern blot, and qRT-PCR.  CcpA is known to play a
+direct role in carbon catabolite repression (CCR).  We use the developed
+transcript map, open read frame, and small RNAs to determine differentially
+expressed regions between the wild-type (UA159) and CcpA-deletion (TW1) strains
+in order to refine the previously established importance of CcpA regulation as
+central metabolism and virulence gene expression.  We find that more genes are
+differentially expressed in the comparison between wild-type and CcpA-deletion
+strains that are grown in galactose than in glucose.  We also find that more
+genes differentially expressed in wild-type than in CcpA-deletion strain when
+comparing glucose- and galactose-grown cells.  These two findings are different
+from those from microarry experiment of \cite{Abranches2008}.  We also predicted
+target genes of the small RNAs that were enriched in the following functional
+categories: xxx.  We confirme that CcpA modulates the pathogenic potential of
+\textit{S.\ mutans} through global control of gene expression.
 
-\subsection{Abstract}
-A bacterial transcriptome of the primary etiological agent of human dental caries,
-\textit{Stretptococcus mutans}, is described using RNA-seq data. Putative
-transcripts boundaries are defined. Small RNAs are predicted by computational
-methods, and confirmed by RNA-seq and Southern blot experiments. We refine the
-previously established importance of CcpA regulation as central metabolism and
-virulence gene expression using RNA-seq data.
-CcpA is known to play a direct role in carbon catabolite repression (CCR).
-We describe genes that are differentially expressed in glucose-grown wild-type
-(UA159) and CcpA-deficient (TW1) strains. We found that more
-genes were differentially expressed in the comparison between wild-type and
-CcpA-deficient strains in galactose-grown strains than in
-glucose-grown strains. How can I explain the reason?
-We also found that more genes differentially expressed in wild-type than in
-CcpA-deficient strain when comparing glucose- and galactose-grown cells. These
-two findings were different from those of \cite{Abranches2008}. \fixme{We may
-want to see effects of different p-value cutoff.} We confirmed that CcpA
-modulates the pathogenic potential of \textit{S.\ mutans} through global control
-of gene expression.
-We predicted a few small RNAs based on sequence alignments and RNA secondary
-structures. We confirmed them experimentally. Target genes of predicted small
-RNAs were predicted. The target genes were enriched in some functional
-categories. We listed highly potential small RNAs with their target genes
-functional categories. 
-Transcripts were predicted using RNA-seq short reads mapped on the wild-type
-UA159 genome.
+\section{Abbreviations}
 
-\subsection{Reviews of papers that cited \cite{Abranches2008}}
+\begin{enumerate}
+\item CCR Carbon catabolite repression
+\item CcpA ?
+\item IPS intracellular polysaccharide
+\item CRE catabolite responsive elements
+\end{enumerate}
+
+\section{Introduction}
+Regulation of carbon catabolite repression (CCR) in \textit{Streptococcus
+mutans} can involve a phosphoenolpyruvate:sugar phosphotransferase system (PTS),
+histidin protein (HPr) and a transcriptional regulator, CcpA \citep{Gorke2008}.
+CcpA was shown to regulate metabolism and virulence gene expression in
+\textit{S.\ mutans} by \citet{Abranches2008}. They used microarray experiments
+to measure expression levels of genes. Microarray experiments tend to be
+qualitative. The advent of high throughput sequencing technology motivated us to
+revisit the experiment to refine the result.  The deep RNA sequencing data also
+allowed us to define the bacterial transcriptome of \textit{S.\ mutans}. So, we
+could compare putative operon expression levels as well as gene expression
+levels between the wild-type (UA159) and its mutant-type (TW1) whose
+\textit{ccpA} was deleted from the wild-type. This allowed us to profile
+expression of genes and putative operons affected by the regulation of
+\textit{ccpA}.
+
+\citet{Martin2010} developed a hidden Markov model to predict transcripts by
+using site-wise expression levels. I also refined the prediction of transcripts
+using multiple samples. I measured expression levels of known genes and compared
+them from one experiment condition and from another. I can test whether two
+genes are in the same operon or not. I assume that expression levels of two genes 
+vary in similar fashion if the two genes belong to the same operon. What I mean
+by similar fashion is whether they are expressed differentially with similar
+fold changes. I assume that operons maps do not change between the two
+experiments, or multiple experiments. This is not true. I hope that additional
+samples could help to define operons. Eventually, operon maps will be determined
+experimentally. Computational tools could be helpful before those ages come to
+us. \fixme{This idea did not work. We should only use the prediction method.}
+
+Although our RNA-seq data did not sampled specifically non-coding small RNAs, we
+retained small RNAs in the sample by attempting to deplete only ribosomal RNAs
+from the sample. Because non-coding small RNAs and coding messenger RNAs were
+mixed in the sample, we sought a way of predicting small RNAs using
+\program{RNAz} developed by \citet{Gruber2010}.  \program{RNAz} used homologous
+sequences and RNA secondary structures to predict putative non-coding RNAs. We
+confirmed the several predicted small non-coding RNAs experimentally
+\citep{Altuvia2007} including the RNA-seq data set. Because small RNAs could
+be associated with CCR regulation of genes (CITE), we determined differentially expressed small
+RNAs between RNA samples. We found a few small RNAs to be differentially
+expressed.
 
 Bacteria can selectively use substrates from different carbon sources. In the
-presence of a
-preferred carbon source, catabolic system of secondary substrates are often
-regulated or shutdown: called carbon catabolite repression (CCR)
+presence of a preferred carbon source, catabolic system of secondary substrates
+are often regulated or shutdown: called carbon catabolite repression (CCR)
 \cite{Gorke2008}. CCR also regulates the expression of virulence factors in many
 pathogenic bacteria. A classical example is the glucose-lactose diauxie in
-\textit{E.\ coli}. A significant number of genes can be subject to CCR.  
-CCR in bacteria involves regulatory networks that activate or silence genes in
-response to carbon source and availability.
-\textit{S.\ mutans} possesses redundant systems for CCR.
-We could measure directly the expression levels of \textit{fruA} in \textit{S.\
-mutans}. 
-We assess the contribution of CcpA to CCR of predicted operons.
-Clearly, deletion of \textit{ccpA} resulted in substantial relief of the CCR of
-\textit{fruA}
-expression exerted by both glucose and fructose.
+\textit{E.\ coli}. A significant number of genes can be subject to CCR.  CCR in
+bacteria involves regulatory networks that activate or silence genes in response
+to carbon source and availability.  \textit{S.\ mutans} possesses redundant
+systems for CCR.  We could measure directly the expression levels of
+\textit{fruA} in \textit{S.\ mutans}.  We assess the contribution of CcpA to CCR
+of predicted operons.  Clearly, deletion of \textit{ccpA} resulted in
+substantial relief of the CCR of \textit{fruA} expression exerted by both
+glucose and fructose.
 
-\subsection{Goal: Refinement of \cite{Abranches2008}}
+We aims at refining the results of \cite{Abranches2008} in more quantative ways.
+Find CCR-sensitive transcripts. We describe predictions made for transcripts and
+small RNAs based on RNA-seq data. Combining the predictions and known genes we
+determine differentially expressed regions in UA159 genome.
 
-Comparisons of microarray and RNA-seq are not good way to go.
+We need to quantify some uncertainty of transcript predictions.
 
-Find CCR-sensitive transcripts.
+\section{Methods and Materials}
 
-\subsection{Methods and Materials}
-
-\subsubsection{Bacterial strains and growth conditions}
+\subsection{Bacterial strains and growth conditions}
+\fixme{Burne group should be able to describe this part. The following is just
+from somewhere for the purpose of writing template.}
 \textit{S.\ mutans} UA159 is a human cariogenic isolate. The whole genome is
 availabe at NCBI \cite{Ajdic2002}. Strain TW1 was derived from UA159 by
-replacing a part of \textit{ccpA} gene \cite{Wen2002}.
+replacing a part of \textit{ccpA} gene \cite{Wen2002}.  \textit{S.\ mutans}
+strains were inoculated into TV medium supplemented with 0.5\% galactose or
+glucose and grown to early exponential phase (OD600, 0.2).  Then fructose was
+added to the medium at concentrations (0,0.05,3,5,10 mM) for 3.5h to induce
+\textit{fruA} expression.
 
-\textit{S.\ mutans} strains were inoculated into TV medium supplemented with
-0.5\% galactose or glucose and grown to early exponential phase (OD600, 0.2).
-Then fructose was added to the medium at concentrations (0,0.05,3,5,10 mM) for
-3.5h to induce \textit{fruA} expression.
-
-\subsubsection{RNA isolation}
+\subsection{RNA isolation}
+\fixme{Burne group should be able to describe this part. The following is just
+from somewhere for the purpose of writing template.}
 RNA was isolated from S. mutans (1) by using 50 ml cultures that were grown
@@
 
-\subsubsection{small RNAs}
+\subsection{Transcript predictions}
+\citet{Martin2010} developed a hidden Markov model to predict transcripts by
+using site-wise expression levels. We used the following options for the
+program: ``-c 10 -b 25 -force\_gp''. We binned expression levels into 25 parts.
+We allowed 10 emission states. We also used the genes as the predicted parts of
+transcripts. 
+Because we did not have strand-specific RNA-seq of the 13 samples due to
+limited resources, we could not tell in which directions transcripts based on
+RNA-seq data were oriented.
 
-RNAz is used to find small RNA candidates. Sequence alignments were a
-critical step for finding small RNAs in the approach. Phylogenies of streptococcus species could
-be considered to choose species related with Streptococcus mutans.
+\subsection{small RNAs}
+RNAz is used to find small RNA candidates. Sequence alignments were a critical
+step for finding small RNAs in the approach. Phylogenies of streptococcus
+species could be considered to choose species related with Streptococcus mutans.
 Alternatively, I find sequences similar to intergenic regions, and refine the
 alignments using a multiple sequence alignment tool such as MUSCLE
-\cite{Edgar2004a}. Steps for
-achieving this goal include: 1. Exract intergenic regions including up- and
-down-stream, 2. BLAST the intergenic regions against non-redundant DB to find
-similar sequences \cite{Altschul1990}, 3. Refine each alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the
-alignments for scoring regions for small RNAs
-\cite{Washietl2005,Gruber2010}, and 5. Find transcripts with high
-score of small RNAs. The transcripts must be defined in advance.
+\cite{Edgar2004a}. Steps for achieving this goal include: 1. Exract intergenic
+regions including up- and down-stream, 2. BLAST the intergenic regions against
+non-redundant DB to find similar sequences \cite{Altschul1990}, 3. Refine each
+alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the alignments for
+scoring regions for small RNAs \cite{Washietl2005,Gruber2010}, and 5. Find
+transcripts with high score of small RNAs. The transcripts must be defined in
+advance.  We used RNAplex \citep{Tafer2008} to predict targets of putative small
+RNAs. The method used RNAplfold \citep{Bernhart2006}.  Please, find a review
+paper about small RNA predictions and targets by \citet{Tafer2008a}.
 
-RNAplex
-\cite{Tafer2008}
+\subsection{Gene category associations and virulence genes}
+We used two approaches for finding functional categories
+\citep{Young2010,Suzuki2011}. We briefly descirbe We briefly descirbe the
+prodecure of \citet{Suzuki2011}.  We assigned genes to Gene Ontology (GO)
+categories by comparing the {\em Streptococcus} genes to bacterial proteins from
+the Uniref90 database using BLASTP, and then assigning the same GO
+classification as the target gene of the uniProt GOA database if the match had
+an $E$-value of $<1.0\times10^{-5}.$  A gene family was assigned a given
+classification if any of its genes was assigned that classification.  To test
+for associations with replacing gene transfers we used the recombination
+intensity estimated for each gene, and for additive gene transfers we used the
+number of additive transfers inferred for each family.  To test for
+significance, we performed a Mann-Whitney $U$ test of the values (recombination
+intensities or numbers of transfers) associated with a given category vs.\ the
+values for the all other genes/families.  We used the \cite{Benjamini1995}
+method to correct for multiple comparisons.
 
-A review paper about small RNA predictions and targets
-\cite{Tafer2008a}
+\section{Results}
 
-RNAplfold
-\cite{Bernhart2006}
+Findings include:
+\begin{enumerate}
+\item New genes, open reading frames that are not annotated in NCBI's genome database.
+\end{enumerate}
 
-\subsubsection{Transcripts}
+We sequenced a whole prokaryotic transcriptome of \textit{S\. mutans}, and
+mapped short reads to the genomic sequence of the reference genome UA159
+\citep{Ajdic2002}. 
+\citet{Martin2010} applied a hidden Markov model to use short reads mapped to a
+genomic sequence in order to parase \textit{B.\ anthracis} genome into a
+sequence of transcripts. Most of annotated genes belong to the predicted
+transcripts. We found discrepancies of annotated genes and their transcriptions:
+1. A gene contains two transcripts, 
+2. Non-coding regions contain transcripts.
+\fixme{We could find more discrepancies. Try to find them}. 
+Expression levels of adjacent genes within a transcript should be more
+correlated to each other than those of adjacent genes belonging to separate transcripts
+\citep{Martin2010}.
 
-``-c 10 -b 25 -force\_gp''
+We identified a set of transcripts, albeit not a full, including small RNAs, and novel
+transcripts. 
+We hope that our study could serve as a foundation for a comprehensive study of
+the \textit{S.\ mutans} transcriptome.
+Our current knowledge of the transcriptome is limited to annotated or predicted genes.
 
-\subsubsection{Gene category associations and virulence genes}
+Whole transcriptome sequencing using high throughput sequencing (HTS)
+technologies, or deep RNA sequencing (RNA-seq) allowed to reveal the
+transcriptome topography of \textit{S.\ mutans}.
 
-We assigned genes to Gene Ontology (GO) categories by comparing the {\em
-  Streptococcus} genes to bacterial proteins from the Uniref90 database
-using BLASTP, and then assigning the same GO classification as the target
-gene of the uniProt GOA database if the match had an $E$-value of
-$<1.0\times10^{-5}.$  
-A gene family was assigned a given
-classification if any of its genes was assigned that classification.  To
-test for associations with replacing gene transfers we used the
-recombination intensity estimated for each gene, and for additive gene
-transfers we used the number of additive transfers inferred for each
-family.  To test for significance, we performed a Mann-Whitney $U$ test of
-the values (recombination intensities or numbers of transfers) associated
-with a given category vs.\ the values for the all other genes/families.  We
-used the \cite{Benjamini1995} method to correct for multiple comparisons.
+\subsection{From Reviews}
+We reconstructed the full-length transcripts by transcriptome assembly using
+xxx.
 
-\subsection{Results}
+We tried in vain to use short-read assemblers such as Velvet \citep{Zerbino2008} that had been developed to
+assembly transcriptome.
+Trinity
+
+gene-dense transcriptomes
+
+We ensured a high-quality transcriptome assembly.
+
+In the data generation phase (FIG. 1a), total RNAs or mRNAs are fragmented and
+converted into a library of cDNAs containing sequencing adaptors.  
+In the data analysis phase (FIG. 1b), these short reads are pre-processed to
+remove sequencing errors and other artefacts. The reads are subsequently
+assembled to reconstruct the original RNAs and to assess their abun- dance
+(‘expression counting’).
+
+To increase the number of assem- bled transcripts, especially the less abundant
+ones, ribosomal RNA (rRNA) and abundant transcripts are removed during the
+first steps of library construction.
+In order to retain RNAs without a poly(A) tail in the assem- bled transcriptome,
+rRNA contamination can instead be removed by hybridization-based depletion
+methods. \fixme{What depletion method did we use?} 
+Last, the use of strand-specific RNA-seq protocols aids in the assembly and
+quantification of overlapping transcripts that are derived from opposite strands
+of the genome. This consideration is especially important for gene-dense
+genomes, such as those of bacteria, archaea and lower eukaryotes.
+
+Sequencing. The major factors to consider before sequencing a sample are: the
+choice of sequencing plat- form, the sequencing read length and whether to use a
+paired-end protocol.
+
+Removing artefacts from RNA- seq data sets before assembly improves the read
+quality, which, in turn, improves the accuracy and computa- tional efficiency of
+the assembly. This step is straight- forward and can be executed using several
+tools.  In general, three types of artefacts should be removed from raw RNA-seq
+data: sequencing adaptors, which origi- nate from failed or short DNA
+insertions during library preparation; low-complexity reads; and
+near-identical reads that are derived from PCR amplification. 
+Sequencing errors in NGS reads can be removed or corrected by analysing the
+quality score and/or the k-mer frequency.
+
+Reference-based transcriptome assembly is easier to perform for the simple
+transcriptomes of bac- terial, archaeal and lower eukaryotic organisms, as these
+organisms have few introns and little alternative splicing. Transcription
+boundaries can be inferred from regions of contiguous read coverage in the
+genome even with- out graph construction and traversal37,51,52. Alternative
+transcription start and stop sites can also be inferred based on the 5′ cap or
+poly(A) signals (if cap- or end- specific experimental protocols are used)51,53.
+However, complications arise owing to the gene-dense nature of
+these genomes. Many genes overlap, resulting in adja- cent genes being assembled
+into one transcript, even though they are not from a polycistronic RNA. Strand-
+specific RNA-seq has successfully been used to separate adjacent overlapping
+genes from opposite strands in the genome51,52. Overlapping genes that are
+transcribed from the same strand and that also have comparable expres- sion
+levels cannot easily be separated without using cap- or end-specific RNA-seq.
+
+In summary, reference-based assembly is generally preferable for cases in which
+a high-quality reference genome already exists.
+
+Meanwhile, experimental RNA-seq and sequencing protocols are continually
+improving and should greatly reduce the informatics challenges.  RNA-seq reads
+from third-generation sequencers, such as PacBio70, are longer (up to several
+kilobases). PacBio sequencers are capable of sequencing a single transcript to
+its full length in a single read. If this technology reaches a throughput that
+is comparable to the second-generation technologies, then the need for
+transcriptome assembly will probably be eliminated. Hopefully, the future of
+transcriptome assembly will be ``no assembly required''.
+
+What is the sequencing depth?
+
+\subsection{My Results}
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{manuscript-figClust}
+\caption{Heatmap showing the Euclidean distances between the samples
+  as calculated from the variance-stabilizing transformation of the
+  count data.}
+\label{figClust}
+\end{figure}
+
 \end{document}

### 2012-01-18 package/smutans/inst/doc/smutans.Rnw
Just an update.

--- 2012-01-13 package/smutans/inst/doc/smutans.Rnw
@@
 smutans.prepareDataSMU86CSP( )
+smutans.prepareTranscript( )
 q("no")
@@
 
-\Rpackage{GenomicRanges} might be useful in 
+\Rpackage{GenomicRanges} is useful in summarizing predicted transcripts.
+feature-genome.out-geneonly is stored in \Robject{smutans.feature.genes}.  
+File FASTQ001.parsernaseq1 is ParseRNAseq output file for transcript
+predictions. File FASTQ001.parsernaseq.pileup is ParseRNAseq input file for
+site-wise expression levels.
+
+\Rfunction{smutans.prepareTranscript} creates two objects:
+\Robject{smutans.tx} and \Robject{smutans.txPileup}.
 <<tx,eval=FALSE>>=

### 2012-01-18 package/smutans/R/core.R

@@ -171,50 +171,6 @@ smutans.prepareDataOMZ175 <- function ()
   save(smomzGenes, file=file.path("smutans", "data", "smomzGenes.RData"))
 }
 
-smutans.prepareGoseq <- function ()
-{
-  f <- "smutans/inst/extdata/feature-genome.out-geneonly"
-  smutans.feature.genes = read.table(file=f,head=F)
-  save(smutans.feature.genes, file=file.path("smutans", "data", "smutans.feature.genes.RData"))
-  f <- "smutans/inst/extdata/smutans.gene2go"
-  smutans.go.genes <- read.table(file=f,head=F)
-  save(smutans.go.genes, file=file.path("smutans", "data", "smutans.go.genes.RData"))
-  f <- "smutans/inst/extdata/smutans.go2ngene"
-  smutans.cat.desc = read.table(file=f,head=F,sep="\t",quote="")
-  save(smutans.cat.desc, file=file.path("smutans", "data", "smutans.cat.desc.RData"))
-}
-
-smutans.makeData <- function( ngenes=500 ) 
-{
-  q0 <- rexp( ngenes, rate=1/250 )
-  is_DE <- runif( ngenes ) < .3
-  lfc <- rnorm( ngenes, sd=2 )
-  q0A <- ifelse( is_DE, q0 * 2^(  lfc/2 ), q0 )
-  q0B <- ifelse( is_DE, q0 * 2^( -lfc/2 ), q0 )
-  true_sf <- c( 1., 1.3, .7, .9, 1.6 )   
-  conds <- c( "A", "A", "B", "B", "B" )
-  m <- t( sapply( 1:ngenes, function(i) 
-     sapply( 1:5, function( j )
-        rnbinom( 1, mu = true_sf[j] * ifelse( conds[j]=="A", q0A[i], q0B[i] ), 
-           size = 1/.2 ) ) ) )
-  colnames(m) <- c( "A1", "A2", "B1", "B2", "B3" )
-  rownames(m) <- paste( "gene", 1:ngenes, 
-    ifelse( is_DE, "T", "F" ), sep="_" )
-  newCountDataSet( m, conds )
-}
-smutans.prepareGoseq <- function ()
-{
-  f <- "smutans/inst/extdata/feature-genome.out-geneonly"
-  smutans.feature.genes = read.table(file=f,head=F)
-  save(smutans.feature.genes, file=file.path("smutans", "data", "smutans.feature.genes.RData"))
[…]

### 2012-01-18 package/test.R

@@ -1,24 +1,3 @@
-sm835result <- function () {
-  library(smutans)
-  data( "sm835Genes" )
-  data( "smutans.feature.genes" )
-  data( "smutans.go.genes" )
-  data( "smutans.cat.desc" )
-  npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
-  npAndP <- smutans.de2( npAndP, type="ua159", 
-                         condA="835np", condB="835p" )
-  npAndUA159 <- newSmutans( sm835Genes, title="835NP vs. UA159" )
-  npAndUA159 <- smutans.de2( npAndUA159, type="ua159", 
-                             condA="835np", condB="ua159nocsp" )
-  pAndUA159 <- newSmutans( sm835Genes, title="835P vs. UA159" )
-  pAndUA159 <- smutans.de2( pAndUA159, type="ua159", 
-                            condA="835p", condB="ua159nocsp" )
-  smutans.de2List( npAndP, "npAndP.csv" )
-  smutans.de2List( npAndUA159, "npAndUA159.csv" )
-  smutans.de2List( pAndUA159, "pAndUA159.csv" )
-  smutans.de2Clust( npAndP )
-  # smutans.de2Clust( npAndP, margins=c(7,7) ) this is for v0.1.3
-}
 # smutans.de2Clust( npAndUA159 )
 # smutans.de2Clust( pAndUA159 )
 # smutans.de2TypeHeatmap( npAndP )
@@ -30,8 +9,9 @@ sm835result <- function () {
 #                   cat.desc=smutans.cat.desc )
 
 
+library(GenomicRanges)
 #library(DESeq)
-#source("smutans/R/core.R")
+source("smutans/R/core.R")
 #smutans.prepareGoseq( )
 #smutans.prepareData835NPP ()
 #smutans.prepareData34 ()
@@ -39,7 +19,8 @@ sm835result <- function () {
 #smutans.prepareDataUA159CSP( ) 
 #smutans.prepareDataSMU86CSP( )
 #smutans.prepareDataUA159TW1()
-#q("no")
+smutans.prepareTranscript( )
+q("no")
 library(DESeq)
 library(goseq)
 load("smutans/data/smutansGenes.RData")
@@ -187,4 +168,24 @@ smutans.de2List( allSampleType )[grep("locus",rownames(smutans.de2List(allSample
 # smutans.de2TypeHeatmap( ua159only )
[…]

### 2012-01-18 pl/feature-genome.pl

@@ -49,6 +49,7 @@ GetOptions( \%params,
             'intergenicregiononly',
             'startcodon',
             'geneonly',
+            'withproduct',
             'windowsize=i',
             'genomelength=i',
             'out=s',
@@ -197,7 +198,16 @@ elsif ($cmd eq "ptt2")
       {
         $end = $genomelength;
       }
-      print $outfile "$chromosome\t$start\t$end\t$g->{Synonym}\t0\t$g->{Strand}\n";
+
+      if (exists $params{withproduct})
+      {
+        print $outfile "$chromosome\t$start\t$end\t$g->{Synonym}\t0\t$g->{Strand}\t";
+        print $outfile "$g->{Product}\n";
+      }
+      else
+      {
+        print $outfile "$chromosome\t$start\t$end\t$g->{Synonym}\t0\t$g->{Strand}\n";
+      }
     }
   }
   elsif (exists $params{geneonly})
@@ -208,7 +218,16 @@ elsif ($cmd eq "ptt2")
       $g->{Location} =~ /(\d+)\.\.(\d+)/;
       my $start = $1 - 1; # PTT is 1-base, BED is 0-base.
       my $end = $2;
-      print $outfile "$chromosome\t$start\t$end\t$g->{Synonym}\t0\t$g->{Strand}\n";
+
+      if (exists $params{withproduct})
+      {
+        print $outfile "$chromosome\t$start\t$end\t$g->{Synonym}\t0\t$g->{Strand}\t";
+        print $outfile "$g->{Product}\n";
+      }
+      else
+      {
+        print $outfile "$chromosome\t$start\t$end\t$g->{Synonym}\t0\t$g->{Strand}\n";
+      }
     }
   }
 }

### 2012-01-18 sh/feature-genome.sh

@@ -42,6 +42,11 @@ function feature-genome {
         -geneonly \
         -in $REFGENOMEPTT \
         -out $DATADIR/$FUNCNAME.out-geneonly"
+      COMMAND1a="perl pl/$FUNCNAME.pl ptt2 \
+        -withproduct \
+        -geneonly \
+        -in $REFGENOMEPTT \
+        -out $DATADIR/$FUNCNAME.out-geneonly.txt"
       COMMAND2="perl pl/$FUNCNAME.pl ptt \
         -in $REFGENOMEPTT \
         -intergenicregion \
@@ -57,11 +62,13 @@ function feature-genome {
 
       if [ "$BATCH" == "YES" ]; then
         echo $COMMAND1 >> $BATCHFILE
+        echo $COMMAND1a >> $BATCHFILE
         #echo $COMMAND2 >> $BATCHFILE
         echo $COMMAND3 >> $BATCHFILE
         echo $COMMAND4 >> $BATCHFILE
       else
         echo $COMMAND1 | bash
+        echo $COMMAND1a | bash
         #echo $COMMAND2 | bash
         echo $COMMAND3 | bash
         echo $COMMAND4 | bash

### 2012-01-21 doc/manuscript.tex
Just and update.

@@
 
+\fixme{I might have to try different filters of short reads to see how those
+affect the transcript prediction.}
+
 \subsection{My Results}
+
+The input RNA-seq data is a list of site-wise non-negative integers that
+represents a number of reads mapped to a particular genomic position
+\citet{Martin2010}; a RNA-seq ``coverage'' data. We designated a gene as an
+expressed if the gene belongs to a transcript with at least C2 level.
+
+The number of transcripts predicted was
+
+A gene
+
+We use the term `coverage' to designate 
+We verified the genes using RNA-seq data.
 

### 2012-01-21 package/manuscript.Rnw
Just and update.

--- 2012-01-18 doc/manuscript.Rnw
@@
 \usepackage{../../latex/sty/genres}
+\externaldocument{supplementary}
 
@@
 \newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}
+\newcommand{\comment}[1]{}
 
@@
 Melissa J.\ Hubisz$^{1}$, \\
+Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
 Robert A.\ Burne$^{3}$
@@
 target genes of the small RNAs that were enriched in the following functional
-categories: xxx.  We confirme that CcpA modulates the pathogenic potential of
+categories: xxx.  We confirmed that CcpA modulates the pathogenic potential of
 \textit{S.\ mutans} through global control of gene expression.
 
+\comment{
 \section{Abbreviations}
-
 \begin{enumerate}
@@
 \end{enumerate}
+}
 
@@
 
-Although our RNA-seq data did not sampled specifically non-coding small RNAs, we
+Small RNAs and their targets can be predicted \citep{Tafer2008a}.
+Although our RNA-seq data did not sample specifically non-coding small RNAs, we
 retained small RNAs in the sample by attempting to deplete only ribosomal RNAs
@@
 presence of a preferred carbon source, catabolic system of secondary substrates
-are often regulated or shutdown: called carbon catabolite repression (CCR)
-\cite{Gorke2008}. CCR also regulates the expression of virulence factors in many
+are often regulated or shutdown: called carbon catabolite repression
+\citep{Gorke2008}. CCR also regulates the expression of virulence factors in many
 pathogenic bacteria. A classical example is the glucose-lactose diauxie in
@@
 
-We need to quantify some uncertainty of transcript predictions.
+\fixme{We need to quantify some uncertainty of transcript predictions.}
 
@@
 \fixme{Burne group should be able to describe this part. The following is just
-from somewhere for the purpose of writing template.}
+from somewhere for the purpose of writing template. Please, rewrite it.}
 \textit{S.\ mutans} UA159 is a human cariogenic isolate. The whole genome is
-availabe at NCBI \cite{Ajdic2002}. Strain TW1 was derived from UA159 by
-replacing a part of \textit{ccpA} gene \cite{Wen2002}.  \textit{S.\ mutans}
+availabe at NCBI \citep{Ajdic2002}. Strain TW1 was derived from UA159 by
+replacing a part of \textit{ccpA} gene \citep{Wen2002}.  \textit{S.\ mutans}
 strains were inoculated into TV medium supplemented with 0.5\% galactose or
@@
 \fixme{Burne group should be able to describe this part. The following is just
-from somewhere for the purpose of writing template.}
+from somewhere for the purpose of writing template. Please, rewrite it.}
 RNA was isolated from S. mutans (1) by using 50 ml cultures that were grown
@@
 \subsection{RNA sequencing}
-\fixme{Stanhope group should be able to describe this part. The following is just
-from somewhere for the purpose of writing template.}
+\fixme{SC wrote the second paragraph based on email correspondences between PDPB and CDD.
+Please rewrite it if necessary. Stanhope group should be able to describe this
+part. The following first paragraph is from a paper. Do not use them or rephrase
+them in any way. They are for informational purposes.}
 Sequencing libraries for the Illumina GA platform were constructed by shearing
@@
 
-SOLiD sequencing was performed at Agencourt Bioscience (Beverly, MA) and
-SeqWright (Houston, TX). Library preparations, fragment library protocols, and
-all SOLiD-run parameters followed standard Applied Biosystems protocols.
-Illumina/Solexa sequencing followed standard Illumina/Solexa methods and
-protocols. Both systems were used because in this study our aim was simply to
-collect as much sequence coverage as possible, and in pursuing this goal, we
-utilized all available sequencing systems. We note, however, that the relative
-performances (overall coverage, data output, etc.) noted in Table 1 for the
-Illumina and Applied Biosystems sequencing platforms cannot be compared
-directly. This study was not intended or designed to be a direct comparison of
-the two systems, and interpreting the differences shown in Table 1 is difficult
-given the many differences in cDNA/library preparation and the rapid evolution
-of each system over the past year.
+Deep sequencing of the RNA samples were performed at the Cornell University Life
+Sciences Core Laboratories Center (Ithaca, NY).  RNA extraction was performed
+using a modified Qiagen RNAease protocol (Qiagen, Inc., Chatsworth, CA): cells
+were pelleted, resuspended in RNA protect reagent, frozen, then resuspended in
+TE, mechanically disrupted in a bead beater in the presence of glass beads and
+acidic phenol. After these steps, we used the Qiagen kit to obtain total RNAs.
+The RNA sample was then passed through the Ambion MicrobExpress kit twice, which
+removed ribosomal RNAs only using magnetic beads.  This procedure left small
+RNAs and transfer RNAs intact.  We followed the TruSeq Illumina kit to prepare
+the cDNA libraries.  We did not intend to enrich short reads in either small
+RNAs or transcript boundaries.
 
 \subsection{Statistical analysis for differential expressions}
-We used an R package, DESeq, \cite{Anders2010} to determine differentially
-expressed genes based on negative binomial model originally developed by \cite{Robinson2007}.
-Steps of analyzing RNA-seq data for determining differentially expressed genes
-are described in \cite{Oshlack2010}. We used a short read aligner called
-\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
-\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
-aligner allowed a few gaps for efficient alignment of millions of sequences of
-size being 100 base pairs long, short reads that consist of RNAs and sequencing
-adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
-remove parts of adapter sequences from short reads before mapping on the
-reference genome. Mapped short read alignments were converted into readable
-formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
+We used an R package, DESeq, \citep{Anders2010} to determine differentially
+expressed genes based on negative binomial model originally developed by
+\citep{Robinson2007}.  Steps of analyzing RNA-seq data for determining
+differentially expressed genes are described in \cite{Oshlack2010}. We used a
+short read aligner called \program{BWA} \citep{Li2009a} to map the sequenced
+short reads on a reference genome, \textit{Streptococcus mutans} UA159 (NCBI
+accession NC\_004350). Because the aligner allowed a few gaps for efficient
+alignment of millions of sequences of size being 100 base pairs long, short
+reads that consist of RNAs and sequencing adapter sequences would be not mapped.
+We used \program{cutadapt} \citep{Martin2011} to remove parts of adapter
+sequences from short reads before mapping on the reference genome. Mapped short
+read alignments were converted into readable formats using \program{samtools}
+\citep{Li2009b}. We counted short reads aligned to annotated genes in the
 reference genome, forming a table of read counts of the genes. Statistical
-software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
-was subsequently employed to infer differentially expressed in different
-biological conditions.  We also generated tracks for our recently released
-Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
-summarize gene expression levels based the short read alignments. This allowed
-us to visualize parts of the reference genome for comparing different RNA
-samples. Further analyses such as functional category enrichment test
-\cite{Young2010} were also employed.
+software R \citep{R2011} with \Rpackage{DEseq} \citep{Anders2010} was
+subsequently employed to infer differentially expressed in different biological
+conditions.  We also generated tracks for our recently released Streptococcus
+Genome Browser (http://strep-genome.bscb.cornell.edu) that summarize gene
+expression levels based the short read alignments. This allowed us to visualize
+parts of the reference genome for comparing different RNA samples. Further
+analyses such as functional category enrichment test \cite{Young2010} were also
+employed.
 
@@
 using site-wise expression levels. We used the following options for the
-program: ``-c 10 -b 25 -force\_gp''. We binned expression levels into 25 parts.
-We allowed 10 emission states. We also used the genes as the predicted parts of
-transcripts. 
-Because we did not have strand-specific RNA-seq of the 13 samples due to
-limited resources, we could not tell in which directions transcripts based on
-RNA-seq data were oriented.
+program: ``-c 10 -b 25 -force\_gp'' to bin expression levels into 25 parts to
+allow 10 emission states. We also used the genes as the predicted parts of
+transcripts.  Because we did not have strand-specific RNA-seq of the 13 samples
+due to limited resources, we could not tell in which directions transcripts
+based on RNA-seq data were oriented.
 
 \subsection{small RNAs}
-RNAz is used to find small RNA candidates. Sequence alignments were a critical
-step for finding small RNAs in the approach. Phylogenies of streptococcus
-species could be considered to choose species related with Streptococcus mutans.
-Alternatively, I find sequences similar to intergenic regions, and refine the
-alignments using a multiple sequence alignment tool such as MUSCLE
-\cite{Edgar2004a}. Steps for achieving this goal include: 1. Exract intergenic
-regions including up- and down-stream, 2. BLAST the intergenic regions against
-non-redundant DB to find similar sequences \cite{Altschul1990}, 3. Refine each
-alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the alignments for
-scoring regions for small RNAs \cite{Washietl2005,Gruber2010}, and 5. Find
-transcripts with high score of small RNAs. The transcripts must be defined in
-advance.  We used RNAplex \citep{Tafer2008} to predict targets of putative small
-RNAs. The method used RNAplfold \citep{Bernhart2006}.  Please, find a review
-paper about small RNA predictions and targets by \citet{Tafer2008a}.
+We used \program{RNAz} to predict small RNA candidates. Sequence alignments were
+a critical step for finding small RNAs in the approach. Phylogenies of
+streptococcus species could be considered to choose species related with
+Streptococcus mutans.  Alternatively, we found sequences similar to intergenic
+regions, and refined the alignments using a multiple sequence alignment tool
+such as MUSCLE \citep{Edgar2004a}. First, we exracted intergenic regions
+including up- and down-stream. We searched non-redundant database of bacterial
+genomes using BLAST \citep{Altschul1990} for similar intergenic sequences,
+subsequently, refining the sequence alignments using MUSCLE \citep{Edgar2004a}.
+Then, we applied \program{RNAz} to the alignments for scoring regions for small
+RNAs \citep{Washietl2005,Gruber2010}.  Targets of transcripts with high scores
+for small RNAs were also predicted using \program{RNAplex} \citep{Tafer2008},
+and \program{RNAplfold} \citep{Bernhart2006}.  Target genes were scored for each
+small RNA candidate, which was used to perform functional category enrichment
+tests of targets for each small RNA candidate. We could find functional
+categories that a few small RNAs might be associated with.
 
@@
 We used two approaches for finding functional categories
-\citep{Young2010,Suzuki2011}. We briefly descirbe We briefly descirbe the
-prodecure of \citet{Suzuki2011}.  We assigned genes to Gene Ontology (GO)
-categories by comparing the {\em Streptococcus} genes to bacterial proteins from
-the Uniref90 database using BLASTP, and then assigning the same GO
-classification as the target gene of the uniProt GOA database if the match had
-an $E$-value of $<1.0\times10^{-5}.$  A gene family was assigned a given
-classification if any of its genes was assigned that classification.  To test
-for associations with replacing gene transfers we used the recombination
-intensity estimated for each gene, and for additive gene transfers we used the
-number of additive transfers inferred for each family.  To test for
-significance, we performed a Mann-Whitney $U$ test of the values (recombination
-intensities or numbers of transfers) associated with a given category vs.\ the
-values for the all other genes/families.  We used the \cite{Benjamini1995}
-method to correct for multiple comparisons.
+\citep{Young2010,Suzuki2011}. We briefly descirbe the prodecure of
+\citet{Suzuki2011}.  We assigned genes to Gene Ontology (GO) categories by
+comparing the \textit{S.\ mutans} genes to bacterial proteins from the Uniref90
+database using \program{BLASTP}, and then assigning the same GO classification
+as the target gene of the uniProt GOA database if the match had an $E$-value of
+$<1.0\times10^{-5}.$  A gene family was assigned a given classification if any
+of its genes was assigned that classification.  To test for associations with
+replacing gene transfers we used the recombination intensity estimated for each
+gene, and for additive gene transfers we used the number of additive transfers
+inferred for each family.  To test for significance, we performed a Mann-Whitney
+$U$ test of the values (recombination intensities or numbers of transfers)
+associated with a given category vs.\ the values for the all other
+genes/families.  We used the \cite{Benjamini1995} method to correct for multiple
+comparisons.
+\fixme{A short description about \citep{Young2010}.}
 
 \section{Results}
-
+\comment{
 Findings include:
@@
 \end{enumerate}
-
-We sequenced a whole prokaryotic transcriptome of \textit{S\. mutans}, and
+}
+We sequenced a whole prokaryotic transcriptome of \textit{S.\ mutans}, and
 mapped short reads to the genomic sequence of the reference genome UA159
@@
 assembled to reconstruct the original RNAs and to assess their abun- dance
-(‘expression counting’).
+(`expression counting').
 
@@
 
+<<manuscript,echo=FALSE,results=hide>>=
+# not run when smutans R package is installed
+# Use this when smutans R package is being tested.
+library(DESeq)
+library(GenomicRanges)
+library(plotrix)
+source("smutans/R/class_and_slots.R")
+source("smutans/R/core.R")
+source("smutans/R/methods.R")
+load("smutans/data/smutansData.tx.RData")
+load("smutans/data/smutansData.txGenes.RData")
+load("smutans/data/smutansData.txPileup.RData")
+load("smutans/data/smutans.feature.genes.RData")
+load("smutans/data/smutans.genes.criteria.RData")
+load("smutans/data/smutans.go.genes.RData")
+load("smutans/data/smutans.cat.desc.RData")
+load("smutans/data/smutansGenes.RData")
+@
+<<manuscript,eval=FALSE,echo=FALSE,results=hide>>=
+# not run without smutans R package installed
+# Remove eval=FALSE when the R package is installed
+library(smutans)
+load(smutansData.tx)
+load(smutansData.txGenes)
+load(smutansData.txPileup)
+load(smutans.feature.genes)
+data(smutans.genes.criteria)
+data(smutans.go.genes)
+data(smutans.cat.desc)
+data(smutansGenes)
+@
+<<echo=FALSE,result=hide>>=
+grTx <- smutansData.tx[elementMetadata(smutansData.tx)$state > 1]
+mtch2 <- findOverlaps(smutansData.txGenes,grTx)
+grNoncoding <- gaps(smutansData.txGenes)
+# A bug in GenomicRanges: the last segment is the while genome. I removed it
+# from the noncoding ranges.
+grNoncoding <- grNoncoding[-length(grNoncoding)] 
+mtchNoncoding <- findOverlaps(grNoncoding,grTx)
+# Expressed transcripts without annotated genes
+grTxWGenes <- grTx[unique(matchMatrix(mtch2)[,2])]
+grTxWOGenes <- grTx[-unique(matchMatrix(mtch2)[,2])]
+# Find transcripts with genes with conflicted strands
+grGenesPlus <- smutansData.txGenes[strand(smutansData.txGenes)=='+']
+grGenesMnus <- smutansData.txGenes[strand(smutansData.txGenes)=='-']
+mtchPlus <- findOverlaps(grGenesPlus,grTxWGenes)
+mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
+txWithConflictedGenes <- intersect(matchMatrix(mtchPlus)[,2],matchMatrix(mtchMnus)[,2])
+grTx2 <- grTxWGenes[-txWithConflictedGenes] 
+# Lengths of 5' and 3' UTRs
+mtchPlus <- findOverlaps(grGenesPlus,grTx2)
+mtchMnus <- findOverlaps(grGenesMnus,grTx2)
+length5UTR <- c()
+length3UTR <- c()
+for (i in unique(matchMatrix(mtchPlus)[,"subject"])) {
+  x <- matchMatrix(mtchPlus)[matchMatrix(mtchPlus)[,"subject"]==i,"query"]
+  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
+  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
+  length5UTR <- c(length5UTR,y)
+  length3UTR <- c(length3UTR,z)
+}
+for (i in unique(matchMatrix(mtchMnus)[,"subject"])) {
+  x <- matchMatrix(mtchMnus)[matchMatrix(mtchMnus)[,"subject"]==i,"query"]
+  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
+  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
+  length5UTR <- c(length5UTR,z)
+  length3UTR <- c(length3UTR,y)
+}
+# Correlation of expression levels of adjacent genes.
+pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
+pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
+elementMetadata(smutansData.txGenes)["score"] <- pileup.y
+mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
+x <- rle(matchMatrix(mtch3)[,"subject"])
+comp.exp.two.genes <- c()
+for (i in x$values[x$lengths > 1]) {
+  y <- combn(matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==i],2)
+  for (j in 1:dim(y)[2]) {
+    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
+  }
+}
+# Correlation of expression levels of adjacent genes between transcripts.
+comp.exp.two.genes.between.transcript <- c()
+for (i in 1:length(x$values)) {
+  if (i == length(x$values)) {
+    break
+  }
+  y1 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i]]
+  i2 <- i + 1
+  y2 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i2]]
+  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
+  for (j in 1:dim(l)[1]) {
+    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
+  }
+}
+# plot(log10(comp.exp.two.genes.between.transcript[1,]), log10(comp.exp.two.genes.between.transcript[2,]))
+# cor(log10(comp.exp.two.genes.between.transcript[1,]), log10(comp.exp.two.genes.between.transcript[2,]))
+@
+
 \subsection{My Results}
+% Statistics of transcripts, and genes.
+The input RNA-seq data is a list of site-wise non-negative integers that
+represents a number of short reads mapped to a particular genomic position
+\citet{Martin2010}, which is called a RNA-seq ``coverage'' data. We designated a
+gene as an expressed if the gene belongs to a transcript with at least C2
+level.
+The number of predicted expressed transcripts was
+\Sexpr{sum(elementMetadata(smutansData.tx)$state > 1)}
+(\Sexpr{as.integer(sum(elementMetadata(smutansData.tx)$state > 1)/length(smutansData.tx)*100)}\%)
+out of 
+\Sexpr{length(smutansData.tx)} 
+fragments.
+Of the total \Sexpr{length(smutansData.txGenes)} of genes,
+\Sexpr{length(matchMatrix(mtch2)[,1])}
+(\Sexpr{as.integer(length(matchMatrix(mtch2)[,1])/length(smutansData.txGenes)*100)}\%) 
+were designated as expressed. 
+Genes that were not designated as expressed could be due to false gene
+annotations or possibility of true negative expression under the experimental
+conditions. 
+Figure \ref{figLength} shows the length distribution of expressed and
+not-expressed genes. 
+Of the \Sexpr{length(grTx)} expressed transcripts, 
+\Sexpr{length(grTxWOGenes)} contained no annotated genes.
+Figure \ref{figTranscriptLength} shows the length distribution of expressed
+transcripts with annotated coding genes and unannotated regions.
+Unannotated regions can include potential non-coding regions and new
+coding genes.  Transciprts with unannotated regions appeared to be shorter than those
+with coding genes. Most of the unannotated regions with length being near 360 bp
+could be non-coding regions, and the rest of the unannotated regions with length
+being near 1000 bp could be new coding genes. Some of the unannotated genes
+could be pseudogene regions.
+Because we used a RNA-seq technique without considering strand-specific, we
+could not tell the strandness of the predicted transcripts. Of
+the \Sexpr{length(grTx)} expressed transcripts, 
+\Sexpr{length(txWithConflictedGenes)}
+included annotated genes of both strands
+(\Sexpr{as.integer(length(txWithConflictedGenes)/length(grTx)*100)}\%).
+We filter out the transcripts with genes with conflicted both strands leaving
+\Sexpr{length(grTx2)} expressed transcripts. 
+Figure \ref{figUTRLength} shows length distributions of 5' and 3' UTR of the
+expressed transcripts without genes with conflicted strands.
+Many transcripts appeared to start and end near the boundaries of annotated genes.
+Gene expression levels for adjacent genes within expressed transcripts were more
+correlated than those between adjacent transcripts 
+(Figure \ref{figExpressionTwoGene}).
+% Question: What are the promoter scores?
 
-The input RNA-seq data is a list of site-wise non-negative integers that
-represents a number of reads mapped to a particular genomic position
-\citet{Martin2010}; a RNA-seq ``coverage'' data. We designated a gene as an
-expressed if the gene belongs to a transcript with at least C2 level.
+% Small RNAs predicted
+Although we aimed at sequencing messenger RNAs, we expected to find regulatory
+small RNAs in the RNA-seq data because we attempted to deplete structrual
+ribosomal RNAs. \program{RNAz} is a software package for detection of structural
+noncoding RNAs using comparative genomic data \citep{Gruber2010}. The method has
+been applied to different genomic data (CITE). We applied the method to predict
+potential non-coding small RNAs in \textit{S.\ mutans} as a supplement to
+evidence for small RNAs provided by the RNA-seq data. With increased evidence of
+putative small RNAs we could proceed to further experimentally verify
+potentially important small RNAs.  We first predicted
+\Sexpr{length(colnames(smutans.genes.criteria))-1} putative small RNAs. Using
+\program{RNAplex} \citep{Tafer2008} and \program{RNAplfold} \citep{Bernhart2006}
+we predicted target genes of the putative small RNAs.  Table \ref{tab:goseq}
+shows functional categories enriched with target genes for predicted small RNAs.
+We found that a few of the putative small RNAs were structured with stable stems
+(SHOW FIGURES).
 
-A gene
+% UCSC genome browser resources
+We created new tracks for our recently released {\em Streptococcus} Genome
+Browser (http://strep-genome.bscb.cornell.edu) that summarized the results of
+this \textit{S.\ mutans} transcriptome analysis alongside known genes, predicted
+small RNAs, predicted target genes and predicted transcripts. These tracks can
+be used to inspect loci of interest and to compare the results of different
+RNA-seq data sets. They can also be queried and intersected with other tracks
+using the UCSC Table Browser.
 
-We use the term `coverage' to designate 
+<<echo=FALSE,result=hide>>=
+glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
+glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
+                            condA="ua159", condB="tw1" )
+smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
+@
+% Differential expression study
+We sequenced 
+\Sexpr{length(colnames(counts(smutansGenes)))}
+RNA samples for studying differential gene expression to better understand
+biological processes. Two independent factors of the biological processes
+includes the effect of CcpA and the treatment of monosaccharides (glucose versus
+galactose). The two factors results four pair-wise comparisons. We compared two
+conditions of one factor with the other factor being fixed at one condition. For
+example, we could compare UA159 wild-type and TW1 mutant-type strains under the
+condition of glucose-grown cells to find differentially expressed associatively
+with deactivation of \textit{ccpA} genes. We performed a similar comparison with
+galactose-grown cells. While we replicated four biological experiments for the
+condition combination of UA159 wilde-type grown in galactose, we had three for
+the other three conditions. Because of limited resources of RNA deep sequencing,
+we considered three replicates as a maximum number to but still adequately
+detect differentially expressed genes.
+
+We list differentially expressed genes in the comparison of UA159 wild-type and
+TW1 mutant-type in Table \ref{tabGlucoseOnly}.
+
+Table \ref{tabGalactoseOnly}.  
+Table \ref{tabUA159Only}.  
+Table \ref{tabTW1Only}.
+
+\fixme{Let's compare the microarray result and that from RNA-seq.}
+\fixme{Use KEGG for functional category study.}
+\fixme{Display results in UCSC genome browser.}
+\fixme{Let's find strand-specific RNA-seq.}
+\fixme{Add gene descriptions and functional category in the result of DE study.}
+
 We verified the genes using RNA-seq data.
@@
 
+\section{Acknowledgments}
+This study was funded by the National Institute of Allergy and Infectious
+Disease, US National Institutes of Health, under grant number AI073368-01A2
+(to M.J.S.\ and A.S).  Additional support was provided by National Science
+Foundation CAREER Award DBI-0644111 and a David and Lucile Packard
+Fellowship for Science and Engineering (to A.S.).  
+We thank Christopher Browngardt for assistance with labelling functional classes
+of \textit{S.\ mutans} genes.
+
 \renewcommand*{\refname}{Literature Cited}
@@
 \bibliography{siepel-mutans}
+\clearpage{}
 
+\section{Tables}
+
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology enrichments for target genes of putative small RNAs.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{lcccl}
+\hline 
+small RNA & $q^a$ & count & GO & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.mannwhitney( qval=0.05,
+                     genes.criteria=smutans.genes.criteria,
+                     go.genes=smutans.go.genes,
+                     cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseq}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+
+\section{Figures}
+
+<<figLength,fig=TRUE,echo=FALSE>>=
+l <- list(expressed=width(smutansData.txGenes[matchMatrix(mtch2)[,1]]), 
+          notexpressed=width(smutansData.txGenes[-matchMatrix(mtch2)[,1]]))
+bin <- c(as.integer(seq.int(0, 3000, length.out=15)), 8500)
+multhist(l,breaks=bin)
+@
 \begin{figure}
 \centering
-\includegraphics[width=.9\textwidth]{manuscript-figClust}
-\caption{Heatmap showing the Euclidean distances between the samples
-  as calculated from the variance-stabilizing transformation of the
-  count data.}
-\label{figClust}
+\includegraphics[width=.9\textwidth]{manuscript-figLength}
+\caption{{\bf Length distribution for expressed and not-expressed genes.}
+Darker gray
+bars represent frequency of expressed genes, and lighter gray ones that of
+not-expressed genes. The last bin sums from 3000 bp to 8500 bp.
+}
+\label{figLength}
 \end{figure}
-<<figClust,fig=TRUE,echo=FALSE>>=
-x <- rnorm(100)
-hist(x)
+
+<<figTranscriptLength,fig=TRUE,echo=FALSE>>=
+l <- list(coding=width(grTxWGenes), 
+          noncoding=width(grTxWOGenes))
+bin <- c(as.integer(seq.int(0, 10000, length.out=15)), 30000)
+multhist(l,breaks=bin)
 @
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{manuscript-figTranscriptLength}
+\caption{{\bf Length distribution for expressed transcripts with annotated genes
+and unannotated genes.}
+Darker gray bars represent frequency of expressed transcripts with annotated
+genes, and lighter gray ones that of without annotated genes that can be
+potential non-coding regions. The last bin sums from 10000 bp to 30000 bp.
+}
+\label{figTranscriptLength}
+\end{figure}
+
+<<figUTRLength,fig=TRUE,echo=FALSE>>=
+l <- list(five=length5UTR, three=length3UTR)
+bin <- c(as.integer(seq.int(0, 300, length.out=19)), 6500)
+multhist(l,breaks=bin)
+@
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{manuscript-figUTRLength}
+\caption{{\bf Length distribution for 5' and 3' UTR.}
+Darker gray bars represent frequency of 5' UTR lengths, and lighter gray 3' UTR
+lengths. The last bin sums from 300 bp to 6500 bp.
+\fixme{Check the too long UTRs because they do not make sense.}
+}
+\label{figUTRLength}
+\end{figure}
+
+<<figExpressionTwoGene,fig=TRUE,echo=FALSE>>=
+plot( log10(comp.exp.two.genes[1,]), log10(comp.exp.two.genes[2,]),
+      xlab="Expression of first Gene in log10",
+      ylab="Expression of second Gene in log10",
+      pch=19,cex=0.5)
+abline(b=1,lty=2)
+@
+<<figExpressionTwoGeneBetweenTranscript,fig=TRUE,echo=FALSE>>=
+plot( log10(comp.exp.two.genes.between.transcript[1,]),
+      log10(comp.exp.two.genes.between.transcript[2,]),
+      xlab="Expression of first Gene in log10",
+      ylab="Expression of second Gene in log10",
+      pch=19,cex=0.5)
+abline(b=1,lty=2)
+@
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{manuscript-figExpressionTwoGene}
+\includegraphics[width=.6\textwidth]{manuscript-figExpressionTwoGeneBetweenTranscript}
+\caption{{\bf Scatter plot of expression levels of pairs of adjacent genes
+located in the same transcripts.}
+(Top) Two genes in the same tranascripts are plotted using their expression levels in
+log10 scale.
+(Bottom) Two genes in different tranascripts are plotted using their expression levels in
+log10 scale.
+}
+\label{figExpressionTwoGene}
+\end{figure}
 

### 2012-01-21 package/supplementary.Rnw
Just and update.

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color} 
\usepackage{setspace} 
\usepackage{rotating}
\usepackage{natbib}
\usepackage{xr}
\usepackage{../../latex/sty/genres}
\usepackage{url}
\usepackage{longtable}

\externaldocument{manuscript}

% Text layout
\oddsidemargin 0in
\evensidemargin 0in
\topmargin -.5in
\textwidth 6.5in
\textheight 9in
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=6,height=6}
\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\program}[1]{{\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}
\newcommand{\comment}[1]{}

\makeatother
\pagestyle{myheadings}
\markright{Transcriptome of Streptococcus mutans}

%

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}

\makeatother

\renewcommand{\thefootnote}{\textit{\alph{footnote}}}

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **
\markright{Transcriptome of Streptococcus mutans}

\begin{document}

\setcounter{secnumdepth}{-1}

\noindent \begin{LARGE}{\bf Defining Transcriptome using RNA Sequencing
 \\[0.7ex] 
of {\em Streptococcus mutans}}\end{LARGE}\\[2ex]
{\bf 
Sang Chul Choi, Charles D.\ Danko, Melissa J.\ Hubisz, \\
Paulina Dmitrievna Pavinski Bitar, \\
Robert A.\ Burne, Michael J.\ Stanhope, and Adam Siepel}%\\[0.5ex]

\tableofcontents

\listoftables

\listoffigures

\doublespacing

\clearpage

<<echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(GenomicRanges)
library(plotrix)
library(xtable)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutansGenes.RData")
@

\section{Supplementary Text}
A
\subsection{Summary statistics for recombinant trees}
\citep{Gorke2008}

\enlargethispage{\baselineskip}

\renewcommand*{\refname}{Literature Cited}
\bibliographystyle{../../latex/bst/mbe}
\bibliography{siepel-mutans}

\clearpage{}\setcounter{figure}{0}
\setcounter{table}{0}

\makeatletter 
\renewcommand{\thefigure}{S\@arabic\c@figure} 
\renewcommand{\thetable}{S\@arabic\c@table} 
\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\tablename}{Supplementary Table}
\makeatother

%\section{Supplementary Tables}

<<echo=FALSE,results=hide>>=
geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
geneDescription <- geneDescription[,c(4,7)]
colnames(geneDescription) <- c("Gene ID","Gene Description")
@

<<echo=FALSE,results=tex>>=
#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
#                            condA="ua159", condB="tw1" )
#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
load("glucoseOnly.RData")
x <- smutans.de2List( glucoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=0,
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabGlucoseOnly',
                 caption='List of genes for the comparison UA159 and TW1 under glocuse-grown cells',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
@
\clearpage{}

<<echo=FALSE,results=tex>>=
#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
#                              condA="ua159", condB="tw1" )
#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
load("galactoseOnly.RData")
x <- smutans.de2List( galactoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=0,
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabGalactoseOnly',
                 caption='List of genes for the comparison UA159 and TW1 under galactose-grown cells',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
@
\clearpage{}

<<echo=FALSE,results=tex>>=
#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
#ua159only <- smutans.de2( ua159only, type="ua159", 
#                          condA="glucose", condB="galactose" )
#save(ua159only, file=file.path("ua159only.RData"))
load("ua159only.RData")
x <- smutans.de2List( ua159only, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=0,
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabUA159Only',
                 caption='List of genes for the comparison of UA159 wild-type grown with glucose and galactose ',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
@
\clearpage{}

<<echo=FALSE,results=tex>>=
#tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
#tw1only <- smutans.de2( tw1only, type="tw1", 
#                          condA="glucose", condB="galactose" )
#save(tw1only, file=file.path("tw1only.RData"))
load("tw1only.RData")
x <- smutans.de2List( tw1only, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=0,
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabTW1Only',
                 caption='List of genes for the comparison of TW1 mutant-type grown with glucose and galactose ',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
@
\clearpage{}

%\section{Supplementary Figures}

%=============================================================================
%\begin{figure}[!ht]
%\includegraphics[scale=0.42]{figures/clonalorigin}
%\caption[Illustration of recombinant tree]{
%{\bf Illustration of a recombinant tree in the ClonalOrigin model.}
%(A) A recombinant tree for our five-sample clonal frame with four
%recombinant edges (dashed lines with arrows running from donors to
%recipients).  (B) Genomic 
%regions associated with recombinant edges.  Recombination 
%intensities (number of edge types present) and gene tree labels are shown at
%bottom. (C)~Induced local gene trees.
%Bold lines indicate branches of the clonal frame, while dashed lines indicate
%recombinant edges. See section \textbf{Summary statistics of recombinant trees} in
%\textbf{Supplementary Text} for additional details.}
%\label{fig:clonalorigin}
%\end{figure}
%\clearpage{}%
%=============================================================================

\end{document}

### 2012-01-21 package/smutans/R/core.R

@@ -363,6 +363,8 @@ smutans.prepareTranscript <- function ()
   parseRNAseqOutFile <- "smutans/inst/extdata/FASTQ001.parsernaseq1"
   x <- read.table(parseRNAseqOutFile, head=FALSE)
   number.tx <- length(rownames(x))
+  tx.state <- sub("_[[:digit:]]*", "", substring(x$V9,6))
+  tx.state <- sub("C", "", tx.state)
   smutansData.tx <- GRanges( seqnames = Rle("chr1", number.tx),
                               ranges =
                                 IRanges(start=x$V4,
@@ -370,7 +372,7 @@ smutans.prepareTranscript <- function ()
                                         names=sprintf("tx%04d",seq(number.tx))),
                               strand = Rle(strand("*"),number.tx),
                               score=x$V6,
-                              state=substring(x$V9,6)
+                              state=as.integer(tx.state)
                             )
   # ParseRNAseq input pileup or coverage file
   parseRNAseqInFile <- "smutans/inst/extdata/FASTQ001.parsernaseq.pileup"
@@ -391,9 +393,11 @@ smutans.prepareTranscript <- function ()
                                  )
   seqlengths(smutansData.txGenes) <- length(pileup.v)
 
+  smutansData.txPileup <- pileup.v
   save(smutans.feature.genes, file=file.path("smutans", "data", "smutans.feature.genes.RData"))
   save(smutansData.tx, file=file.path("smutans", "data", "smutansData.tx.RData"))
   save(smutansData.txGenes, file=file.path("smutans", "data", "smutansData.txGenes.RData"))
+  save(smutansData.txPileup, file=file.path("smutans", "data", "smutansData.txPileup.RData"))
 }
 
 

### 2012-01-21 package/smutans/R/methods.R

@@ -275,6 +275,9 @@ setMethod("smutans.de2List",
       # options(width = 1000) 
       return(resSig[order(resSig$pval),])
       # return( object@res )
+    } else if (file == "tex") {
+      # Write the result in tex format.
+      print ("AAA")
     } else {
       write.csv(object@res, file=file)
     }

### 2012-01-21 package/test.R

@@ -30,6 +30,7 @@ load("smutans/data/sm34Genes.RData")
 load("smutans/data/smutans.cat.desc.RData")
 load("smutans/data/smutans.genes.criteria.RData")
 load("smutans/data/smutans.go.genes.RData")
+load("smutans/data/smutansData.txPileup.RData")
 source("smutans/R/class_and_slots.R")
 source("smutans/R/core.R")
 source("smutans/R/methods.R")

### 2012-01-21 test/GenomicRanges/run.R

@@ -3,6 +3,8 @@ library(GenomicRanges)
 # We load a ParseRNAseq output file first. We could load a BED-format file later.
 x <- read.table("FASTQ001.parsernaseq1", head=FALSE)
 number.tx <- length(rownames(x))
+tx.state <- sub("_[[:digit:]]*", "", substring(x$V9,6))
+tx.state <- sub("C", "", tx.state)
 gr <- GRanges( seqnames = Rle("chr1", number.tx),
                ranges =
                  IRanges(start=x$V4,
@@ -10,7 +12,7 @@ gr <- GRanges( seqnames = Rle("chr1", number.tx),
                          names=sprintf("tx%04d",seq(number.tx))),
                strand = Rle(strand("*"),number.tx),
                score=x$V6,
-               state=substring(x$V9,6)
+               state=as.integer(tx.state)
              )
 
 # Find the number of transcripts
@@ -53,16 +55,74 @@ mtch <- findOverlaps(grGenes,gr)
 # Plot or summarize number of genes in the transcripts
 stem(table(matchMatrix(mtch)[,2]))
 
+# Find the expressed genes
+mtch2 <- findOverlaps(grGenes,gr[elementMetadata(gr)$state > 1])
+length(matchMatrix(mtch2)[,1])
+
 # Find transcripts without annotated genes as putative non-coding elements
-grNoncoding <- gaps(grGenes)
-mtchNoncoding <- findOverlaps(grNoncoding,gr)
+smutansData.txGenes <- grGenes
+smutansData.tx <- gr
+smutansData.txPileup <- pileup.v
+grTx <- smutansData.tx[elementMetadata(smutansData.tx)$state > 1]
+
+grNoncoding <- gaps(smutansData.txGenes)
+mtchNoncoding <- findOverlaps(grNoncoding,grTx)
+                              
+matchMatrix(mtchNoncoding)[,2]
 stem(table(matchMatrix(mtchNoncoding)[,2]))
+
 
 # Find transcripts with genes with conflicted strands
 # Get plus strand genes.
-grGenesPlus <- grGenes[strand(grGenes)=='+']
-grGenesMnus <- grGenes[strand(grGenes)=='-']
-mtchPlus <- findOverlaps(grGenesPlus,gr)
-mtchMnus <- findOverlaps(grGenesMnus,gr)
[…]

### 2012-01-21 package/manuscript.Rnw
Just an update.

@@
 <<echo=FALSE,result=hide>>=
-glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
-glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
-                            condA="ua159", condB="tw1" )
+#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
+#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
+#                            condA="ua159", condB="tw1" )
+#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
+load("glucoseOnly.RData")
 smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
+tableGlucoseOnly <- smutans.de2List( glucoseOnly, qval=1e-3 )
+
+# Gene descriptions
+geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
+geneDescription <- geneDescription[,c(4,7)]
+colnames(geneDescription) <- c("Gene ID","Gene Description")
+# Functional classes
+functionalDescription <- read.csv("feature-genome.out-geneonly-oralgen-functional.txt",head=FALSE)
+functionalDescription <- functionalDescription[,c(2,12)]
+colnames(functionalDescription) <- c("Gene ID","Functional Class")
+# Find potential differentially expressed genes, and add gene descriptions, and
+# functional classes
+x <- smutans.de2List( glucoseOnly, qval=1e-3 )
+x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
+colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
+x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
+tableGlucoseOnly <- x
 @
+
 % Differential expression study
@@
 TW1 mutant-type in Table \ref{tabGlucoseOnly}.  
+Of \Sexpr{nrow(counts(smutansGenes))} genes, 
+\Sexpr{nrow(tableGlucoseOnly)} genes were found to be differentially expressed
+at FDR of 0.1\%.
 

### 2012-01-21 package/supplementary.Rnw
Just an update.

@@
 library(xtable)
+library(RColorBrewer)
 source("smutans/R/class_and_slots.R")
@@
 colnames(geneDescription) <- c("Gene ID","Gene Description")
+
+functionalDescription <- read.csv("feature-genome.out-geneonly-oralgen-functional.txt",head=FALSE)
+functionalDescription <- functionalDescription[,c(2,12)]
+colnames(functionalDescription) <- c("Gene ID","Functional Class")
 @
@@
                  display=c("s","s","d","d","f","e","s"),
-                 digits=0,
+                 digits=c(0,0,0,0,1,0,0),
                  align=c('l','l','r','r','r','r','p{3.5in}'),
@@
        floating=FALSE) 
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
 @
@@
 
+We confirmed that the most upregulated genes in the TW1 mutant-type encoded the
+components of the PDH enzyme complex (SMU.1421, SMU.1422, SMU.1423, and
+SMU.1424).  
+\Sexpr{length(grep("Energy metabolism",x[,"Functional Class"]))}
+\Sexpr{length(grep("PTS",x[,"Functional Class"]))}
+\Sexpr{length(grep("hypothetical",x[,"Gene Description"]))}
+
+<<figPieGlucoseOnly,fig=TRUE,echo=FALSE,width=6,height=8>>=
+par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
+#par(mfrow=c(2,1))
+trim <- function (x) gsub("^\\s+|\\s+$", "", x)
+y <- x[,"Functional Class"]
+z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+z <- table(as.factor(sort(unlist(z))))
+
+slices <- as.vector(z)
+lbls <- names(z)
+pct <- round(slices/sum(slices)*100)
+c1 <- brewer.pal(n=9,name="Set1")
+pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
+plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
+legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
+@
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figPieGlucoseOnly}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing UA159 wild-type and TW1 mutant-type grown in glucose.}
+Labels for functional categories are colored with nine different colors. The
+first label with red color, 
+``Biosynthesis of cofactors, prosthetic groups, and carriers; Thiamine'', starts
+at the top, and next labels read to the clockwise. The numbers circling the pie
+chart is in per cent.
+}
+\label{figPieGlucoseOnly}
+\end{figure}
+
+\clearpage{}
 <<echo=FALSE,results=tex>>=
@@
 
+<<figPieGalactoseOnly,fig=TRUE,echo=FALSE,width=6,height=8>>=
+par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
+#par(mfrow=c(2,1))
+trim <- function (x) gsub("^\\s+|\\s+$", "", x)
+y <- x[,"Functional Class"]
+z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+z <- table(as.factor(sort(unlist(z))))
+
+slices <- as.vector(z)
+lbls <- names(z)
+pct <- round(slices/sum(slices)*100)
+c1 <- brewer.pal(n=9,name="Set1")
+pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
+plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
+legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
+@
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figPieGalactoseOnly}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing UA159 wild-type and TW1 mutant-type grown in galactose.}
+Labels for functional categories are colored with nine different colors. The
+first label with red color, 
+``Amino acid biosynthesis; Glutamate family'', starts
+at the top, and next labels read to the clockwise. The numbers circling the pie
+chart is in per cent.
+}
+\label{figPieGalactoseOnly}
+\end{figure}
+
+\clearpage{}
 <<echo=FALSE,results=tex>>=
@@
 load("ua159only.RData")
-x <- smutans.de2List( ua159only, qval=1e-3 )
+x <- smutans.de2List( ua159only, qval=1e-10 )
 x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
@@
                  display=c("s","s","d","d","f","e","s"),
-                 digits=0,
+                 digits=c(0,0,0,0,1,0,0),
                  align=c('l','l','r','r','r','r','p{3.5in}'),
                  label='tabUA159Only',
-                 caption='List of genes for the comparison of UA159 wild-type grown with glucose and galactose ',
+                 caption='List of genes for the comparison of UA159 wild-type grown with glucose and galactose. Note that we used 1e-10 as the p-value cutoff.',
                )
@@
 
+<<figPieUA159Only,fig=TRUE,echo=FALSE,width=6,height=8>>=
+par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
+trim <- function (x) gsub("^\\s+|\\s+$", "", x)
+y <- x[,"Functional Class"]
+z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+z <- table(as.factor(sort(unlist(z))))
+
+slices <- as.vector(z)
+lbls <- names(z)
+pct <- round(slices/sum(slices)*100)
+c1 <- brewer.pal(n=9,name="Set1")
+pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
+plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
+
+legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
+#legend(-1.1,1.1,names(z[1:27]),bty='n',cex=0.7,fill=c1)
+#legend(0,0.4,names(z[-c(1:27)]),bty='n',cex=0.7,fill=c1)
+
+@
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figPieUA159Only}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing UA159 wild-type grown in glucose and galactose.}
+Labels for functional categories are colored with nine different colors. The
+first label with red color, 
+``Amino acid biosynthesis; Glutamate family'', starts
+at the top, and next labels read to the clockwise. The numbers circling the pie
+chart is in per cent.
+}
+\label{figPieUA159Only}
+\end{figure}
+
+\clearpage{}
 <<echo=FALSE,results=tex>>=
@@
        floating=FALSE) 
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
 @
 \clearpage{}
+
+<<figPieTW1Only,fig=TRUE,echo=FALSE,width=6,height=8>>=
+par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
+trim <- function (x) gsub("^\\s+|\\s+$", "", x)
+y <- x[,"Functional Class"]
+z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+z <- table(as.factor(sort(unlist(z))))
+
+slices <- as.vector(z)
+lbls <- names(z)
+pct <- round(slices/sum(slices)*100)
+c1 <- brewer.pal(n=9,name="Set1")
+pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
+plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
+legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
+@
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figPieTW1Only}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing TW1 mutant-type grown in glucose and galactose.}
+Labels for functional categories are colored with nine different colors. The
+first label with red color, 
+``Amino acid biosynthesis; Histidine family'', starts
+at the top, and next labels read to the clockwise. The numbers circling the pie
+chart is in per cent.
+}
+\label{figPieTW1Only}
+\end{figure}
 

### 2012-01-22 package/manuscript.Rnw
Just an upate.

@@
 
+Expression levels of \textit{fruA} in \textit{S.\ mutans} strains that were
+grown to early exponential phase (OD600, 0.2) in TV medium supplemented with
+0.5\% glucose or galactose.
+
 \subsection{RNA isolation}
@@
 
-\section{Results}
 \comment{
@@
 }
+
+\section{Results}
 We sequenced a whole prokaryotic transcriptome of \textit{S.\ mutans}, and
@@
 \citep{Martin2010}. 
-
 We identified a set of transcripts, albeit not a full, including small RNAs, and novel
@@
 Our current knowledge of the transcriptome is limited to annotated or predicted genes.
-
 Whole transcriptome sequencing using high throughput sequencing (HTS)
 technologies, or deep RNA sequencing (RNA-seq) allowed to reveal the
-transcriptome topography of \textit{S.\ mutans}.
+transcriptome topography of \textit{S.\ mutans}. We did not aim at precisely
+pin-pointing transcript start or end sites.
+
+We admit that strand-specific RNA-seq would be important in gene-dense
+transcriptomes of species like bacteria or archaea.
 
 \subsection{From Reviews}
-We reconstructed the full-length transcripts by transcriptome assembly using
-xxx.
 
-We tried in vain to use short-read assemblers such as Velvet \citep{Zerbino2008} that had been developed to
-assembly transcriptome.
-Trinity
-
-gene-dense transcriptomes
-
+We could not use short-read assemblers such as Velvet \citep{Zerbino2008} 
+and Trinity that had been developed to assembly transcriptome. Because genes in
+bacteria are so dense, RNA-seq data without strands did not allow us to apply
+any state-of-the-art software of transcriptome assembly.
 We ensured a high-quality transcriptome assembly.
@@
 
+Name the transcripts.
+I found xx transcripts. Among them xx \% were mono-cistronic operons. xx \% were
+poly-cistronic operons. xx transcripts could not defined because RNA-seq short
+reads were not precise. Transripts were assigned with functional categories, and
+differentially expressed transcripts were found. Transcripts and functional
+categories: how can I assess that transcripts consist of genes with similar
+functional cateogires.
+
 % Small RNAs predicted
@@
 (SHOW FIGURES).
+
+Name the small RNAs.
+RNAz predicted a few small RNAs. Some of them were differentially expressed.
+Some functional categories were enriched with target genes of the predicted
+genes.
 
@@
 at FDR of 0.1\%. 
+We confirmed that the most upregulated genes in the TW1 mutant-type encoded the
+components of the PDH enzyme complex (SMU.1421, SMU.1422, SMU.1423, and
+SMU.1424).  
+Figure \ref{figPieGlucoseOnly} shows the proportions of functional categories 
+of the differentially expressed genes. Notably, we found
+\Sexpr{length(grep("Energy metabolism",tableGlucoseOnly[,"Functional Class"]))}
+genes that were involved in energy metabolism. Sugars and fermentation
+subclasses were profound in energy metabolism. We also found 
+\Sexpr{length(grep("PTS",tableGlucoseOnly[,"Functional Class"]))}
+genes in PTS class of signal transduction. Another subclass, 
+``carbohydrates, organic alcohols, and acids,'' in
+``transport and binding proteins'' was following next in the number of
+differentially expressed genes. We also found  
+\Sexpr{length(grep("hypothetical",tableGlucoseOnly[,"Gene Description"]))}
+genes annotated as hypothetical, which might constitute the unknown functional
+class. These results were consistent with those of \citet{Abranches2008}.
+Profoundly upregulated genes in response to the lack of CcpA were
+putative Clp proteinase (SMU.1425), 
+branched-chain alpha-keto acid dehydrogenase subunit E2 (SMU.1421), and
+putative transcriptional regulator (SMU.1599).
+Down-regulated genes include cytoplasmic alpha-amylase (SMU.1590), 
+beta-D-fructosyltransferase (SMU.2028), and
+putative glycerol uptake facilitator protein (SMU.396).
+Fructosyltransferase converts sucrose to fructose.
 
-Table \ref{tabGalactoseOnly}.  
-Table \ref{tabUA159Only}.  
-Table \ref{tabTW1Only}.
+<<echo=FALSE,result=hide>>=
+#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
+#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
+#                              condA="ua159", condB="tw1" )
+#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
+load("galactoseOnly.RData")
+x <- smutans.de2List( galactoseOnly, qval=1e-3 )
+x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
+colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
+x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
+tableGalactoseOnly <- x
+@
+We found that differentially expressed genes in galactose-grown cells were
+smaller than those in glucose-grown cells, which was different from the result
+of \cite{Abranches2008}. 
+Energy metabolism was the most
+profoundly differentially expressed (fig.\ \ref{figPieGalactoseOnly}).
+Subclasses pyruvate family in amino acid biosynthesis, 
+and degradation of proteins, peptides, and glycopeptides in protein fate
+were following the next. Almost quarter of the genes were unknown.
+Down-regulated genes included cytoplasmic alpha-amylase (SMU.1590), and
+beta-D-fructosyltransferase (SMU.2028)
+(table \ref{tabGalactoseOnly}).
 
-\fixme{Let's compare the microarray result and that from RNA-seq.}
-\fixme{Use KEGG for functional category study.}
-\fixme{Display results in UCSC genome browser.}
-\fixme{Let's find strand-specific RNA-seq.}
-\fixme{Add gene descriptions and functional category in the result of DE study.}
+<<echo=FALSE,result=hide>>=
+#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
+#ua159only <- smutans.de2( ua159only, type="ua159", 
+#                          condA="glucose", condB="galactose" )
+#save(ua159only, file=file.path("ua159only.RData"))
+load("ua159only.RData")
+x <- smutans.de2List( ua159only, qval=1e-10 )
+x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
+colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
+x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
+tableUA159Only <- x
+@
+Figure \ref{figPieUA159Only} shows proportons of functional categories that
+included differentially expressed genes between glucose-grown and
+galactose-grown UA159 wild-type cells. Qualitatively, energy metabolism and PTS
+classes were profound. This was silmilar to the comparison between UA159
+wild-type and TW1 mutant-type grown in glucose. More than quarter of the
+differentially expressed genes were unknown, unassigned, or hypothetical.
+Consecutive genes (SMU.1488 to SMU.1496) were upregulated in 
+catabolite-derepressing conditions or in galactose-grown cells
+(\fixme{Are these the genes of the lactose (lac) operon?})
+(table \ref{tabUA159Only}).  Another set of genes (SMU.886 to SMU.889) were also
+upregulated (\fixme{Are these called galKTE?}).  
+Genes from SMU.1535 to SMU.1538 were upregulated. 
+Of the \Sexpr{nrow(tableUA159Only)} differentially expressed genes at
+$1 \times 10^{-10}$ cutoff of multiple testing adjusted p-value, 
+\Sexpr{length(grep("hypothetical",tableUA159Only[,"Gene Description"]))} were
+hypothetical, and most of them were down-regulated. 
+A few of the down-regulated genes were annotated.
 
-We verified the genes using RNA-seq data.
+We also found sugar metabolism and PTS signal transduction to be profound in
+differential expression in the comparison of glucose-grown and galactose-grown
+TW1 mutant-type cells figure \ref{figPieTW1Only}. Many of the genes were also
+hypothetical. We found similar genes differentially expressed in TW1 mutant-type
+cells to those in UA159 wild-type (table \ref{tabTW1Only}).
 
-CcpA affects CCR of \textit{fruA}. 
-\textit{fruA} fructan hydrolase
-When Wen2002 deleted CRE sequencs in the \textit{fruA} promoter region, CCR was
-almost completely alleviated. 
-What are CCR-sensitive genes? 
-What are the mechanism of CcpA on CCR of the \textit{fruA} operon? 
-Expression levels of \textit{fruA} in \textit{S.\ mutans} strains that were
-grown to early exponential phase (OD600, 0.2) in TV medium supplemented with
-0.5\% glucose or galactose. \fixme{What does this mean?} 
+\fixme{We may need dot plots where x-axis represents fold changes, and y-axis
+functional classes. Dots may be colored by p-values. This would show most of
+information in one plot. We could label some of dots to emphasixe. Done?}
+\fixme{Use KEGG for functional category study. Done?}
+\fixme{Display results in UCSC genome browser. Done?}
+\fixme{Use two factor analysis. Done?}
+
+%
+
 Deletion of \textit{ccpA} resulted in substantial relief of the CCR of
 \textit{fruA} expression exerted by both glucose and fructose. 
-CcpA could directly interact with the \textit{fruA} promoter region.
+When \citet{Wen2002} deleted CRE sequencs in the \textit{fruA} promoter region, CCR was
+almost completely alleviated.  They also showed that 
+CcpA directly interacted with the \textit{fruA} promoter region.
 
 The role of CcpA in gene expression during growth on glucose.
+
 \textit{S.\ mutans} UA159 and TW1 grown under catabolite-repressing conditions
 (glucose), and catabolite-derepressing conditions (galactose). Under
-glucose-only condition I found xx number of differentially expressed genes. What
-are the functional categories of those differentially expressed genes? What are
-the genes that are most profoundly up- or down-regulated (Smu.1421 to Smu.1424,
-PDH enzyme complex), ... and more? ... This regulatory control was
-short-circuited in the strain growing on glucose and lacking CcpA. I showed that
-some of KEGG pathways were affected by TW1 grown under glucose. We found
-down-regulated genes, which would have something to do with pathogenicity. 
-Upto here, I compared UA159 and TW1 under glucose.
-
-1. UA159 (A) vs. TW1 (B) : fold change is B/A.
-When glucose was the sole carbohydrate for growth, 170 genes were differentially
-expressed at least twofold as a result of loss of CcpA (Fig. 2; see Table S1 in
-the supplemental material). Of the gene products, 22 were predicted to partici
-pate in energy metabolism, 9 were PTS components, and 43 were hypothetical
-proteins (Fig. 2). Among the most profoundly up-regulated genes in the ccpA
-mutant strain were the genes that encoded the components of the PDH enzyme com-
-plex (Smu.1421 to Smu.1424), pyruvate-formate lyase, enzymes in the
-tricarboxylic acid (TCA) cycle, selected sugar-specific PTS permeases, and
-components of the glycogen biosynthetic machinery (see Table S1 in the
-supplemental material). 
-Consistent with the profile data, TW1 cells grown in glucose had 78\% higher
-PDH activity than the wild-type strain (data not shown). Given the impact on
-expression of PDH, TCA enzymes, and selected glycolytic enzymes, CcpA clearly
-legislates gene expression to discriminate between carbohydrate-replete and
-carbohydrate-limiting conditions. In particular, when there is excess
-carbohydrate, S. mutans moves carbohydrates primar- ily through lactate
-dehydrogenase, sacrificing an ATP and generating NAD When carbohydrate is
-limiting, S. mutans pro- duces mainly formic and acetic acids, shunts
-carbohydrate through pyruvate formate lyase, and uses its partial TCA cycle to
-regenerate reducing equivalents to maintain an appropriate NAD/NADH balance.
-This regulatory control was clearly short-circuited in the strain growing on
-glucose and lacking CcpA.
-Among the more interesting genes that were down-regu- lated, significant
-decreases in expression were noted for the genes encoding components of the
-proton-translocating F-ATPase, phosphoglycerate kinase (Pgk), and
-fructosyltrans- ferase (ftf), which catalyzes the conversion of sucrose to high-
-molecular-weight polymers of fructose, as well as gtfD, which encodes the
-glucosyltransferase D enzyme that produces predominantly -1,6-linked polymers
-of glucose from sucrose. Since acid tolerance (ATPase), acid production from
-carbohy- drate (Pgk), and glucan and fructan production (Gtf and Ftf), as well
-as the ability to transport carbohydrates and produce intracellular storage
-compounds (glycogen), as noted above (PTS and Glg), are central to the
-pathogenic potential of S. mutans, CcpA represents a major control point for
-expression of critical virulence attributes in cells presented with glucose.
-
-Growth in derepressing conditions induces major changes in the UA159
-transcriptome. Here, I compared glucose and galactose conditions for growing
-UA159. Energy metabolism was enriched with differentially expressed genes. There
-were more up-regulated genes in galactose condition compared with glucose
-condition. Some genes are described in detail.
-
-2. I compared glucose (A) and galactose (B) for growing UA159
-It is noteworthy that only three genes in the energy metabolism category were
-down-regulated, whereas 23 genes in this category were markedly up-regulated. In
-addition, the expression of 28 genes encoding hypothetical proteins was affected,
-and 19 of them were up-regulated. The genes encod- ing proteins belonging to the
-pathways for lactose and galac- tose utilization (2) were the most strongly
-induced genes in cells growing on galactose. The galKTE genes, encoding the
-enzymes for galac- tose utilization via the Leloir pathway (2), were
-up-regulated between 10- and 36-fold, but the genes of the lactose (lac) operon,
-encoding the lactose PTS permease and the tagatose- 6-phosphate pathway, were
-up-regulated 70- to 300-fold. Since the products of the galKTE operon contribute
-to the genera- tion of UDP-galactose, which is required for cell wall biogen-
-esis, the lower level of induction of galKTE than of the lac operon may reflect
-a need for higher constitutive levels of expression of the Leloir pathway in the
-absence of galactose. Components of the multiple-sugar metabolism operon (msm)
-were also induced in cells growing on galactose. The Msm pathway is induced by
-raffinose and melibiose, both of which are ?-galactosides (5). The enhanced
-expression of gene products involved in en- ergy metabolism, including pyruvate
-formate lyase and com- ponents of the TCA cycle, in cells growing on galactose
-is consistent with the idea that the cells perceive carbohydrate limitation.
-
-Now, I compared glucose (A) and galactose (B) for growing TW1.
-
-Now, I compared UA159 and TW1 under galactose.
-
-What genes are differentially expressed?
-
-Name the transcripts.
-I found xx transcripts. Among them xx \% were mono-cistronic operons. xx \% were
-poly-cistronic operons. xx transcripts could not defined because RNA-seq short
-reads were not precise. Transripts were assigned with functional categories, and
-differentially expressed transcripts were found. Transcripts and functional
-categories: how can I assess that transcripts consist of genes with similar
-functional cateogires.
-
-Name the small RNAs.
-RNAz predicted a few small RNAs. Some of them were differentially expressed.
-Some functional categories were enriched with target genes of the predicted
-genes.
+glucose-only condition we found xx number of differentially expressed genes.
 

### 2012-01-23 package/manuscript.Rnw
Just an update.

@@
 Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
+\fixme{Add more coauthors and change the order of authors if needed}, \\
 Robert A.\ Burne$^{3}$
@@
 \vspace{1in}
+
+\fixme{Add corresponding authors if needed.}
 
@@
 
-<<manuscript,echo=FALSE,results=hide>>=
+<<echo=FALSE,results=hide>>=
 # not run when smutans R package is installed
@@
 @
-<<manuscript,eval=FALSE,echo=FALSE,results=hide>>=
+<<eval=FALSE,echo=FALSE,results=hide>>=
 # not run without smutans R package installed
@@
 functional classes. Dots may be colored by p-values. This would show most of
-information in one plot. We could label some of dots to emphasixe. Done?}
+information in one plot. We could label some of dots to emphasize. Done?}
+
+% tableGlucoseOnly's functional classes can be multiple. We still need to have
+% different colors using p-values.
+% dotplot(tableGlucoseOnly[,"Functional Class"]~tableGlucoseOnly[,"fold"])
+
+\fixme{Display results in UCSC genome browser. Done?}
+
+\fixme{Young's GO analysis with the comparisons}
+We used a q-value of $1\times 10^{-3}$ to determine differentially expressed
+genes. Note that we had another q-value cutoff to list gene ontology terms for
+functional category enrichment tests; 0.05 was the cutoff value for this.
+We found gene ontology terms enriched with differentially expressed genes in the
+comparison of UA159 and TW1 glucose-grown cells:
+phosphotransferase system, 
+glycogen biosynthetic process, 
+sugar:hydrogen symporter activity, and
+carbohydrate metabolic process (table \ref{tab:goseqGlucoseOnly}).
+Because the number of differentially expressed genes in galactose-grown cells is
+was smaller than that of glucose-grown cells, we found only one gene ontology
+term, branched chain family amino acid biosynthetic process
+(table \ref{tab:goseqGalactoseOnly}).
+The comparisons between glucose-grown and galactose-grown UA159
+(table \ref{tab:goseqUA159Only}), and that from TW1 mutant-type
+(table \ref{tab:goseqTW1Only}) showed very similar functional categories. Ranks
+of gene ontology terms were different. 
+Lactose metabolic process was top in TW1, and
+phosphoenolpyruvate-dependent sugar phosphotransferase system was top in UA159.
+
 \fixme{Use KEGG for functional category study. Done?}
-\fixme{Display results in UCSC genome browser. Done?}
+(table \ref{tab:kegseqGlucoseOnly})
+(table \ref{tab:kegseqGalactoseOnly})
+(table \ref{tab:kegseqUA159Only})
+(table \ref{tab:kegseqTW1Only})
+
 \fixme{Use two factor analysis. Done?}
+Figure \ref{figClust}
+Because the effects of nutrient changes appeared to be similar whether UA159
+wild-type or TW1 mutant-type was used, we used all of the 
+\Sexpr{length(colnames(counts(smutansGenes)))} samples to compare glucose-grown
+and galactose grown cells.
+
+<<echo=FALSE,results=hide>>=
+allSample <- newSmutans( smutansGenes, 
+                         title="All Samples for Comparing Conditions" )
+allSample <- smutans.de2( allSample )
+save(allSample, file=file.path("allSample.RData"))
+#load("allSample.RData")
+tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
+                                                   allSample )
+#addmargins( tabUA159onlyAndAllSample )
+@
+Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
+\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
+analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
+genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
+genes.  Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from
+the two-factor analysis versus those from the single-factor analysis. The
+p-values from the two-factor analysis are overall smaller than those from the
+single-factor analysis. We would expect that this was the case because of higher
+power of the two-factor analysis.
 

### 2012-01-23 package/supplementary.Rnw
Just an upate.
Just an update.

@@
 library(DESeq)
+library(goseq)
 library(GenomicRanges)
@@
 load("smutans/data/smutans.cat.desc.RData")
+load("smutans/data/smutans.keg.genes.RData")
+load("smutans/data/smutans.keg.desc.RData")
 load("smutans/data/smutansGenes.RData")
+@
+<<eval=FALSE,echo=FALSE,results=hide>>=
+# not run without smutans R package installed
+# Remove eval=FALSE when the R package is installed
+library(smutans)
+load(smutansData.tx)
+load(smutansData.txGenes)
+load(smutansData.txPileup)
+load(smutans.feature.genes)
+data(smutans.genes.criteria)
+data(smutans.go.genes)
+data(smutans.cat.desc)
+data(smutans.keg.genes)
+data(smutans.keg.desc)
+data(smutansGenes)
 @
@@
                  label='tabGlucoseOnly',
-                 caption='List of genes for the comparison UA159 and TW1 under glocuse-grown cells',
+                 caption='List of genes for the comparison UA159 and TW1 under glucose-grown cells',
                )
@@
 \clearpage{}
-
-We confirmed that the most upregulated genes in the TW1 mutant-type encoded the
-components of the PDH enzyme complex (SMU.1421, SMU.1422, SMU.1423, and
-SMU.1424).  
-\Sexpr{length(grep("Energy metabolism",x[,"Functional Class"]))}
-\Sexpr{length(grep("PTS",x[,"Functional Class"]))}
-\Sexpr{length(grep("hypothetical",x[,"Gene Description"]))}
 
@@
 
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+genes in the comparison UA159 and TW1 grown in glucose.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.de2Goseq ( glucoseOnly, qval=1e-3, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseqGlucoseOnly}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+genes in the comparison UA159 and TW1 grown in galactose.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.de2Goseq ( galactoseOnly, qval=1e-3, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseqGalactoseOnly}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+genes in the comparison of glucose-grown and galactose-grown UA159 wild-type
+cells.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.de2Goseq ( ua159only, qval=1e-3, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseqUA159Only}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+genes in the comparison of glucose-grown and galactose-grown TW1 mutant-type
+cells.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.de2Goseq ( tw1only, qval=1e-3, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseqTW1Only}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf KEGG enrichments for differentially expressed
+genes in the comparison UA159 and TW1 grown in glucose.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.de2Goseq ( glucoseOnly, qval=1e-3, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.keg.genes[c(4,3)],
+                   cat.desc=smutans.keg.desc)
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:kegseqGlucoseOnly}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf KEGG enrichments for differentially expressed
+genes in the comparison UA159 and TW1 grown in galactose.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.de2Goseq ( galactoseOnly, qval=1e-3, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.keg.genes[c(4,3)],
+                   cat.desc=smutans.keg.desc)
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:kegseqGalactoseOnly}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf KEGG enrichments for differentially expressed
+genes in the comparison of glucose-grown and galactose-grown UA159 wild-type
+cells.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.de2Goseq ( ua159only, qval=1e-3, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.keg.genes[c(4,3)],
+                   cat.desc=smutans.keg.desc)
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:kegseqUA159Only}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf KEGG enrichments for differentially expressed
+genes in the comparison of glucose-grown and galactose-grown TW1 mutant-type
+cells.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.de2Goseq ( tw1only, qval=1e-3, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.keg.genes[c(4,3)],
+                   cat.desc=smutans.keg.desc)
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:kegseqTW1Only}
+\end{table}
+%-----------------------------------------------------------------------------
+
+% END OF TABLE
+
 %\section{Supplementary Figures}
 
-%=============================================================================
-%\begin{figure}[!ht]
-%\includegraphics[scale=0.42]{figures/clonalorigin}
-%\caption[Illustration of recombinant tree]{
-%{\bf Illustration of a recombinant tree in the ClonalOrigin model.}
-%(A) A recombinant tree for our five-sample clonal frame with four
-%recombinant edges (dashed lines with arrows running from donors to
-%recipients).  (B) Genomic 
-%regions associated with recombinant edges.  Recombination 
-%intensities (number of edge types present) and gene tree labels are shown at
-%bottom. (C)~Induced local gene trees.
-%Bold lines indicate branches of the clonal frame, while dashed lines indicate
-%recombinant edges. See section \textbf{Summary statistics of recombinant trees} in
-%\textbf{Supplementary Text} for additional details.}
-%\label{fig:clonalorigin}
-%\end{figure}
-%\clearpage{}%
-%=============================================================================
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{supplementary-figClust}
+\caption{Heatmap showing the Euclidean distances between the samples
+  as calculated from the variance-stabilizing transformation of the
+  count data.}
+\label{figClust}
+\end{figure}
+<<figClust,echo=FALSE,result=hide,fig=TRUE>>=
+allSample <- newSmutans( smutansGenes )
+smutans.de2Clust( allSample )
+@
 

### 2012-01-23 package/smutans/R/core.R

@@ -272,6 +272,14 @@ smutans.prepareGoseq <- function ()
   f <- "smutans/inst/extdata/target.rnaplex.acc.out"
   smutans.genes.criteria = read.table(file=f,head=T,sep="\t",quote="")
   save(smutans.genes.criteria, file=file.path("smutans", "data", "smutans.genes.criteria.RData"))
+
+  f <- "smutans/inst/extdata/smu00001.keg.gene2cat"
+  smutans.keg.genes <- read.delim(file=f,head=F)
+  save(smutans.keg.genes, file=file.path("smutans", "data", "smutans.keg.genes.RData"))
+  f <- "smutans/inst/extdata/smu00001.keg.catdesc"
+  smutans.keg.desc <- read.delim(file=f,head=F)
+  save(smutans.keg.desc, file=file.path("smutans", "data", "smutans.keg.desc.RData"))
+  # smutans.keg.genes <- x[c(4,3)]
 }
 
 smutans.run.mw <- function(x, gocats, gocats.col="go", descrips=NULL, noisy=FALSE) {

### 2012-01-23 package/test.R

@@ -12,14 +12,14 @@
 library(GenomicRanges)
 #library(DESeq)
 source("smutans/R/core.R")
-#smutans.prepareGoseq( )
+smutans.prepareGoseq( )
 #smutans.prepareData835NPP ()
 #smutans.prepareData34 ()
 #smutans.prepareDataOMZ175( ) 
 #smutans.prepareDataUA159CSP( ) 
 #smutans.prepareDataSMU86CSP( )
 #smutans.prepareDataUA159TW1()
-smutans.prepareTranscript( )
+#smutans.prepareTranscript( )
 q("no")
 library(DESeq)
 library(goseq)

### 2012-01-23 pl/feature-genome.pl

@@ -279,6 +279,88 @@ elsif ($cmd eq "rnaz")
     print $outfile "$_+\n";
   }
 }
+elsif ($cmd eq "keg")
+{
+  my %c1;
+  my $l;
+  my $count = 0;
+  my $count2 = 0;
+  my $level1 = "";
+  my $level2 = "";
+  my $level3 = "";
+  my $level4 = "";
+  my @genes = ();
+  my %categoryDescription;
+  my %categoryCountDescription;
+  while ($l = <$infile>)
+  {
+    chomp $l;
+    unless ($l =~ /^D/)
+    {
+      if (scalar (@genes) > 0)
+      {
+        $count2++;
+        foreach my $g (@genes) 
+        {
+          $g =~ s/\_/\./g;
+          print $outfile "$level1\t$level2\t$level3\t$g\n"
+        }
+        my $rec = {};
+        $categoryCountDescription{$level3} = $rec;
+        $rec->{description} = $categoryDescription{$level3};
+        $rec->{count} = scalar (@genes);
+        #print "$count2 (1) $level1 : (2) $level2 : (3) $level3\t";
+        #print (join(", ", @genes));
+        #print "\n";
+        @genes = ();
+      }
+    }
+
+    if ($l =~ /^A\<b\>(.+)\<\/b\>/)
+    {
+      $level1 = $1;
+      # print $level1,"\n";
+    }
+    elsif ($l =~ /^B\s+\<b\>(.+)\<\/b\>/)
+    {
+      $level2 = $1;
+      # print "\t",$level2,"\n";
+    }
+    # elsif ($l =~ /^C\s+(\d+)\s+(.+)\s+\[PATH:smu\d+\]/)
+    elsif ($l =~ /^C\s+(\d+)\s+(.+)/)
+    {
+      $level3 = "KEGG:$1";
+      $count++;
+      # print "\t\t",$level3,"\n";
+      $categoryDescription{$level3} = $2;
+    }
+    elsif ($l =~ /^D\s+(\w+)/)
+    {
+      $level4 = $1;
+      # print "\t\t\t",$level4,"\n";
+      push @genes, $level4;
+    }
+#        00010 Glycolysis / Gluconeogenesis [PATH:smu00010]
+
+#      $rec = {};
+#      $c1{$1} = $rec;
+#      for $field ( split ) {
+#        ($key, $value) = split /=/, $field;
+#        $rec->{$key} = $value;
+#      }
+
+  }
+
[…]

### 2012-02-02 package/manuscript.Rnw
Just an update.

--- 2012-01-22 package/manuscript.Rnw
@@
 \usepackage{xr}
+\usepackage{url}
 \usepackage{../../latex/sty/genres}
@@
 Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
+\fixme{Add more coauthors, or change the order of authors if needed} \\
 Robert A.\ Burne$^{3}$
@@
 \vspace{1in}
+
+\fixme{Add more corresponding authors.}
 
@@
 
-\section{Abstract}
-A bacterial transcriptome from the primary etiological agent of human dental
-caries, \textit{Stretptococcus mutans}, is described using deep RNA sequencing
-data.  We define transcripts boundaries of open reading frames based on sequence
-alignments of short reads on the reference genome of UA159 from the deep
-sequencing.  Based on homologous sequences and RNA secondary structures we also
-report putative small RNAs, several of which are confirmed experimentally using
-both RNA sequencing data, northern blot, and qRT-PCR.  CcpA is known to play a
-direct role in carbon catabolite repression (CCR).  We use the developed
-transcript map, open read frame, and small RNAs to determine differentially
-expressed regions between the wild-type (UA159) and CcpA-deletion (TW1) strains
-in order to refine the previously established importance of CcpA regulation as
-central metabolism and virulence gene expression.  We find that more genes are
-differentially expressed in the comparison between wild-type and CcpA-deletion
-strains that are grown in galactose than in glucose.  We also find that more
-genes differentially expressed in wild-type than in CcpA-deletion strain when
-comparing glucose- and galactose-grown cells.  These two findings are different
-from those from microarry experiment of \cite{Abranches2008}.  We also predicted
-target genes of the small RNAs that were enriched in the following functional
-categories: xxx.  We confirmed that CcpA modulates the pathogenic potential of
-\textit{S.\ mutans} through global control of gene expression.
+\fixme{Refinements: More filtering steps in the short read alignments might be necessary. Done?}
 
-\comment{
-\section{Abbreviations}
-\begin{enumerate}
-\item CCR Carbon catabolite repression
-\item CcpA ?
-\item IPS intracellular polysaccharide
-\item CRE catabolite responsive elements
-\end{enumerate}
+\section*{Abstract}
+A bacterial transcriptome of the primary etiological agent of human dental
+caries, \textit{Streptococcus mutans}, is described using deep RNA sequencing.
+RNA transcripts inferred from the RNA-seq data revealed novel messenger RNAs,
+and small RNAs. Differential expression studies on the role of CcpA in 
+carbon catabolite repression (CCR) suggested that some of the novel mRNAs and
+small RNAs found in our study could be involved in the regulation of CcpA. Our
+study provides more precise landscape of transcriptome changes in response to
+different carbon sources.
+
+\section*{Introduction}
+% S. mutans, CCR and CcpA
+About one fifth of children of ages ranging from 2 to 19 were reported to have
+untreated cavities in the United States \citep{health2010}. Oral bacteria digest
+foods containing carbohydrates left on the teeth to convert them to acids, which
+dissolve the surface of the teeth and create cavities.  The primary etiological
+agent of human dental caries, \textit{Streptococcus mutans}, is one of hundreds
+bacterial species that constitute the oral microbial community
+\citep{Jenkinson2005}.  Bacteria can selectively use substrates from different
+carbon sources via a regulation called carbon catabolite repression (CCR)
+\citep{Gorke2008}.  \textit{S.\ mutans} also possesses biological systems for
+CCR.  CCR were known to regulates the expression of virulence factors in many
+pathogenic bacteria, thereby subjecting a number of genes to CCR.  CcpA was
+shown to regulate metabolism and virulence gene expression in 
+\textit{S.\ mutans} through CCR in transcriptom study using 
+the microarray technique based on DNA hybridization \citep{Abranches2008}.
+
+% RNA-seq, and microarray, bacterial transcriptomes
+It has been a decade since the first genome of the oral bacteria was finished
+\citep{Ajdic2002}.  Studies of the transcriptome in the bacteria were performed
+based on DNA hybridization
+\citep[e.g.,][]{Klein2010,Lemme2011,Redanz2011,Wen2011} whereas those in other
+bacterial species took advantage of high-throughput sequencing technologies:
+e.g.,
+\textit{Mycoplasma pneumoniae} \citep{Guell2009},
+\textit{Methanosarcina mazei} \citep{Jaeger2009},
+\textit{Bacillus subtilis}  \citep{Irnov2010},
+\textit{Helicobacter pylori} \citep{Sharma2010}, and
+\textit{Sulfolobus solfataricus} \citep{Wurtzel2010}, to name a few.
+The advent of the high-throughput RNA deep sequencing technologies allowed
+annotation-free detection of transcripts, sensitive and dynamic ranges of
+measurements in RNA expression levels, and studies of non-coding RNAs.
+
+% Aims
+We used the standard RNA deep sequencing technique to sequence RNAs in
+\textit{S.\ mutans}, focusing on messenger RNAs expression levels in response to
+different nutrients with UA159 wild-type and \textit{ccpA}-deleted TW1
+mutant-type.  We attempted to define transcripts of mRNAs using the RNA-seq
+data, and to find several small non-coding RNAs. Using the map of transcripts
+and expression levels of the transcripts from different biological conditions
+allowed us to determine which transcripts and non-coding small RNAs were
+differentially expressed. This provided more refined views of dynamic changes in
+the transcriptome landscape of the etiological agent of human dental caries,
+\textit{S.\ mutans}, than previously reported by \citet{Abranches2008}.
+
+\section*{Materials and Methods}
+
+\subsection*{Bacterial strains and growth conditions}
+\fixme{Burne group describes bacterial strains and growth conditions and other
+necessary works. Please, add texts and SCC will copy-edit them.}
+
+\subsection*{RNA isolation}
+\fixme{Burne group describes RNA isolations and other works. Please, add texts
+and SCC will copy-edit them.}
+
+\subsection*{RNA sequencing}
+\fixme{SCC wrote the paragraph based on email correspondences between PDPB and CDD.
+Stanhope group describes RNA sequencing. Please add texts and SCC will copy-edit
+them.}
+Deep sequencing of the RNA samples were performed at the Cornell University Life
+Sciences Core Laboratories Center (Ithaca, NY).  RNA extraction was performed
+using a modified Qiagen RNAease protocol (Qiagen, Inc., Chatsworth, CA): cells
+were pelleted, resuspended in RNA protect reagent, frozen, and then resuspended
+in TE, mechanically disrupted in a bead beater in the presence of glass beads
+and acidic phenol. We used the Qiagen kit to obtain total RNAs.  The RNA sample
+was passed through the Ambion MicrobExpress kit twice, which removed ribosomal
+RNAs only using magnetic beads.  This procedure left small RNAs and transfer
+RNAs intact.  Then, we followed the TruSeq Illumina kit to prepare the cDNA
+libraries.
+
+\subsection*{Short reads alignments}
+About 20 millions of short reads were sequenced for each sample.  We used a
+short read aligner, \program{BWA} \citep{Li2009a}, to map the sequenced short
+reads on the reference genome, \textit{S.\ mutans} UA159 (NCBI accession
+NC\_004350). Because the aligner allowed a few gaps for efficient alignments of
+millions of sequences of size being 100 base pairs long, short reads that
+consisted of RNAs with sequencing adapter sequences would not be mapped.  We
+used \program{cutadapt} \citep{Martin2011} to remove parts of adapter sequences
+from short reads before mapping on the reference genome.  Mapped short read
+alignments were converted into readable formats using \program{samtools}
+\citep{Li2009b}.
+
+\subsection*{Transcript predictions}
+RNA transcripts were inferred by applying a hidden Markov model to site-wise
+expression levels \citep{Martin2010}.  We used the pileup command of
+\program{BWA} to convert the short reads alignments to pilup values, which were
+taken as the site-wise expression levels along the genome.  Site-wise expression
+levels are a list of site-wise non-negative integers that represents numbers of
+short reads mapped to a particular genomic position.  We used the following
+options for the transcript inference program, \program{ParseRNAseq}: ``-c 10 -b
+25 -force\_gp'' to bin expression levels into 25 parts, and to allow 10 emission
+states for relative expression intensity. We also used the genes as the
+predicted parts of transcripts.    Because we enriched mRNAs in the cDNA
+preparation, we intented to group annotated genes to segments, which could be
+transcripts. We did not aim at finding precise transcription start or end sites.
+Because we did not have strand-specific RNA-seq of the samples due to limited
+resources, we could not tell in which directions transcripts were oriented
+solely based on RNA-seq data. Instead, we used strands of annotated genes to
+determine strands of predicted transcripts.
+
+\subsection*{Prediction of small RNAs and targets}
+Although our RNA-seq data did not enriched non-coding small RNAs in the cDNA
+preparation, we retained small RNAs in the sample by attempting to deplete only
+ribosomal RNAs from the sample. Because non-coding small RNAs and coding
+mRNAs were mixed in the sample, it was difficult to discriminate small
+non-coding RNAs from mRNAs unless non-coding RNAs expressions were masked by
+neighboring background RNA expressions. 
+Therefore, we sought a way of predicting small RNAs
+using \program{RNAz} \citep{Gruber2010}, which  used homologous sequences and
+RNA secondary structures to predict putative non-coding RNAs. Because sequence
+alignments were a critical step for finding small RNAs in the approach, we
+collated sequences similar to intergenic regions, and refined the alignments using
+a multiple sequence alignment tool. Exracting
+intergenic regions including up- and down-stream, we searched a database of
+bacterial genomes using BLAST \citep{Altschul1990} for similar intergenic
+sequences, subsequently, refining the sequence alignments using MUSCLE
+\citep{Edgar2004a}.  Then, we applied \program{RNAz} to the alignments for
+scoring regions for small RNAs \citep{Washietl2005,Gruber2010}.  Targets of
+transcripts with high scores for small RNAs were also predicted using
+\program{RNAplex} \citep{Tafer2008}, and \program{RNAplfold}
+\citep{Bernhart2006}.  Target genes were scored for each small RNA candidate,
+which was used to perform functional category enrichment tests of targets for
+each small RNA candidate.
+
+\subsection*{Experimental validation of small RNAs}
+\fixme{Burne group might have done or be able to do some experimental
+validations of putative non-coding RNAs shown at table \ref{tabUnannotatedRNA}
+\citep[e.g.,][]{Altuvia2007}.}
+
+\subsection*{Statistical analysis for differential expressions}
+We used \Rpackage{DESeq} \citep{Anders2010} to determine differentially
+expressed genes based on negative binomial model originally developed by
+\citet{Robinson2007}.  Steps of analyzing RNA-seq data for determining
+differentially expressed genes were described in \cite{Oshlack2010}.  We counted
+short reads aligned to annotated genes in the reference genome, forming a table
+of read counts of the genes. Statistical software R \citep{R2011} with
+\Rpackage{DEseq} was subsequently employed to infer differentially expressed
+genes in the different biological conditions.
+
+\subsection*{Gene category associations}
+We compiled three sets of gene categories for testing functional associations.
+Firstly, we assigned genes to Gene Ontology (GO) categories by comparing the
+\textit{S.\ mutans} genes to bacterial proteins from the Uniref90 database using
+\program{BLASTP}, and then assigning the same GO classification as the target
+gene of the uniProt GOA database if the match had an $E$-value of
+$<1.0\times10^{-5}.$  A gene family was assigned a given classification if any
+of its genes was assigned that classification.  Secondly, we collated functional
+classes of genes in the Oral Pathogen Sequence Databases available at
+\url{http://oralgen.lanl.gov}. Thirdly, we used a set of metabolite pathways for
+\textit{S.\ mutans} available at Kyoto Encyclopedia of Genes and Genomes (KEGG)
+\citep{Kanehisa2012}.  Using the sets of gene categories for \textit{S.\ mutans}
+we performed a Mann-Whitney $U$ test of the values of target genes of small RNAs
+associated with a given category versus those for the other categories. We
+used a variation of Fisher's exact test accounting for gene lengths to test
+enrichments of differentially expressed genes \citep{Young2010}.  Both of the
+tests were corrected for multiple testing hypotheses \citep{Benjamini1995}.
+
+\subsection*{UCSC genome browser tracks and data availability}
+We created tracks for our recently released Streptococcus Genome Browser
+(\url{http://strep-genome.bscb.cornell.edu}) that summarized the results of our
+\textit{S.\ mutans} transcriptome analysis with known genes, gene expression
+levels based on the short read alignments, predicted putative transcript
+segments, and predictd small non-coding RNAs.  They should be accessed by clade
+\textit{Streptococcus}, and genome \textit{S. mutans UA159}.  These tracks can
+be used to inspect loci of interest and to compare the results of different
+RNA-seq data sets.  They can also be queried and intersected with other tracks
+using the UCSC Table Browser.
+
+\section*{Results}
+
+<<echo=FALSE,results=hide>>=
+# not run when smutans R package is installed
+# Use this when smutans R package is being tested.
+library(DESeq)
+library(GenomicRanges)
+library(plotrix)
+source("smutans/R/class_and_slots.R")
+source("smutans/R/core.R")
+source("smutans/R/methods.R")
+load("smutans/data/smutansData.tx.RData")
+load("smutans/data/smutansData.txGenes.RData")
+load("smutans/data/smutansData.txGrCombined.RData")
+load("smutans/data/smutansData.txPileup.RData")
+load("smutans/data/smutans.feature.genes.RData")
+load("smutans/data/smutans.genes.criteria.RData")
+load("smutans/data/smutans.go.genes.RData")
+load("smutans/data/smutans.cat.desc.RData")
+load("smutans/data/smutansGenes.RData")
+load("smutans/data/smutansGenes2.RData")
+
+# Copied from supplementary.Rnw. Compare the following with ones above to remove
+# redundant lines.
+load("smutans/data/smutansData.tx.RData")
+load("smutans/data/smutansData.txGenes.RData")
+load("smutans/data/smutansData.txGrCombined.RData")
+load("smutans/data/smutansData.txAnnotation.RData")
+load("smutans/data/smutansData.txRNAz.RData")
+load("smutans/data/smutansData.txPileup.RData")
+load("smutans/data/smutans.feature.genes.RData")
+load("smutans/data/smutans.genes.criteria.RData")
+load("smutans/data/smutans.go.genes.RData")
+load("smutans/data/smutans.cat.desc.RData")
+load("smutans/data/smutans.keg.genes.RData")
+load("smutans/data/smutans.keg.desc.RData")
+load("smutans/data/smutansGenes.RData")
+@
+<<eval=FALSE,echo=FALSE,results=hide>>=
+# not run without smutans R package installed
+# Remove eval=FALSE when the R package is installed
+library(smutans)
+load(smutansData.tx)
+load(smutansData.txGenes)
+load(smutansData.txGrCombined)
+load(smutansData.txPileup)
+load(smutans.feature.genes)
+data(smutans.genes.criteria)
+data(smutans.go.genes)
+data(smutans.cat.desc)
+data(smutansGenes)
+data(smutansGenes2)
+@
+
+<<echo=FALSE,result=hide>>=
+notExpressedTx <- smutansData.txGrCombined[elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5]
+grTx <- smutansData.txGrCombined[!(elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5)]
+# We query genes and annotations (genes and RNAs) againt tx's.
+matchGenesTx <- findOverlaps(smutansData.txGenes,grTx)
+matchAnnotationTx <- findOverlaps(smutansData.txAnnotation,grTx)
+# I use gaps with IRanges because gaps with GenomicRanges does not work as expected.
+grNoAnnotation <- gaps(ranges(smutansData.txAnnotation))
+grNoncoding <- gaps(ranges(smutansData.txGenes))
+grRNAz <- ranges(smutansData.txRNAz)
+# 
+matchUnannotatedRNA <- findOverlaps(grRNAz,grNoAnnotation,type="within")
+txRNAz.unannotated <- grRNAz[matchMatrix(matchUnannotatedRNA)[,1]]
+
+# A bug in GenomicRanges: the last segment is the while genome. I removed it
+# from the noncoding ranges.
+# matchNoncoding <- findOverlaps(grNoncoding,grTx)
+
+# Expressed transcripts without annotated genes
+grTxWGenes <- grTx[unique(matchMatrix(matchGenesTx)[,2])]
+grTxWOGenes <- grTx[-unique(matchMatrix(matchGenesTx)[,2])]
+grTxWOAnnotation <- grTx[-unique(matchMatrix(matchAnnotationTx)[,2])]
+
+# Find transcripts with genes with conflicted strands
+grGenesPlus <- smutansData.txGenes[strand(smutansData.txGenes)=='+']
+grGenesMnus <- smutansData.txGenes[strand(smutansData.txGenes)=='-']
+#mtchPlus <- findOverlaps(grGenesPlus,grTxWGenes)
+#mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
+#txWithConflictedGenes <- intersect(matchMatrix(mtchPlus)[,2],matchMatrix(mtchMnus)[,2])
+
+# Tx with genes strands
+grTx2 <- grTxWGenes[!strand(grTxWGenes)=='*']
+
+# Find the number of genes in segments.
+matchGenesTx2 <- findOverlaps(smutansData.txGenes,grTx2,type="within")
+number.genes.tx <- runLength(Rle(matchMatrix(matchGenesTx2)[,"subject"]))
+
+# Length distribution of UTR regions
+# grTx2 <- grTxWGenes[-txWithConflictedGenes] 
+# Lengths of 5' and 3' UTRs
+mtchPlus <- findOverlaps(grGenesPlus,grTx2,type="within")
+mtchMnus <- findOverlaps(grGenesMnus,grTx2,type="within")
+length5UTR <- c()
+length3UTR <- c()
+for (i in unique(matchMatrix(mtchPlus)[,"subject"])) {
+  x <- matchMatrix(mtchPlus)[matchMatrix(mtchPlus)[,"subject"]==i,"query"]
+  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
+  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
+  length5UTR <- c(length5UTR,y)
+  length3UTR <- c(length3UTR,z)
+}
+for (i in unique(matchMatrix(mtchMnus)[,"subject"])) {
+  x <- matchMatrix(mtchMnus)[matchMatrix(mtchMnus)[,"subject"]==i,"query"]
+  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
+  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
+  length5UTR <- c(length5UTR,z)
+  length3UTR <- c(length3UTR,y)
 }
 
-\section{Introduction}
+# Correlation of expression levels of adjacent genes.
+pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
+pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
+elementMetadata(smutansData.txGenes)["score"] <- pileup.y
+mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
+x <- rle(matchMatrix(mtch3)[,"subject"])
+comp.exp.two.genes <- c()
+for (i in x$values[x$lengths > 1]) {
+  y <- combn(matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==i],2)
+  for (j in 1:dim(y)[2]) {
+    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
+  }
+}
+# Correlation of expression levels of adjacent genes between transcripts.
+comp.exp.two.genes.between.transcript <- c()
+for (i in 1:length(x$values)) {
+  if (i == length(x$values)) {
+    break
+  }
+  y1 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i]]
+  i2 <- i + 1
+  y2 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i2]]
+  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
+  for (j in 1:dim(l)[1]) {
+    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
+  }
+}
+@
+
+\subsection*{Experimental results from Burne group}
+\fixme{Add some if needed.}
+
+\subsection*{Results from Stanhop group when RNA sequencing preparation}
+\fixme{Add some if needed.}
+
+\subsection*{Statistics of short reads}
+
+\subsection*{Cluster of RNA samples}
+First using the table of read counts of annotated genes we formed a cluster of
+the 13 RNA-seq samples (fig.\ \ref{figClust}). The effect of nutrients appeared
+to be stronger than that of the deletion of \textit{ccpA} genes from the UA159
+genome.
+
+\subsection*{Predicted transcripts}
+
+We obtained a set of segments of RNA-seq expressions for each of the 13 RNA-seq
+samples.  All of the segments for each set of RNA-seq expression segments were
+scored by the average proportions of segments shared by the segments in the rest
+of sets.  We placed more highly ranked segments first on the reference genome
+based on the scores, and purged more lowly ranked segments from the
+transcriptome map. In doing so, the
+final set of segments was formed by placing non-overlapping highly ranked
+segments.  Average site-wise expression levels of the segments 
+are shown at figure \ref{figGrCombinedExpression}. 
+We designated segments as expressed that had site-wise pileup values
+greater than 5 and proportion of sites of zero site-wise pileup values less than
+10\%.  A gene was also designated as an expressed if the gene belonged to an
+expressed segment.  Some of the predicted segments were not expressed or very
+lowly expressed (\Sexpr{length(notExpressedTx)} 
+out of 
+\Sexpr{length(smutansData.txGrCombined)}).
+Of the total \Sexpr{length(smutansData.txGenes)} of genes,
+\Sexpr{length(matchMatrix(matchGenesTx)[,1])}
+(\Sexpr{as.integer(length(matchMatrix(matchGenesTx)[,1])/length(smutansData.txGenes)*100)}\%) 
+were designated as expressed. 
+Genes that were not designated as expressed could be due to false gene
+annotations or possibility of true negative expression under the experimental
+conditions. 
+Figure \ref{figLength} shows the length distribution of expressed and
+not-expressed genes. 
+Of the \Sexpr{length(grTx)} expressed segments, 
+\Sexpr{length(grTxWOAnnotation)} contained no annotated genes or RNAs.
+Many of the segments might cover gaps between real operons.
+\fixme{Examine these unannotated regions in the UCSC genome tracks.}
+
+Figure \ref{figTranscriptLength} shows the length distribution of expressed
+segments with annotated coding genes and unannotated regions.  Unannotated
+regions could include potential non-coding regions and novel coding genes.
+Segments of unannotated regions appeared to be shorter than those with
+coding genes. Most of the unannotated regions with length being near 360 bp
+could be non-coding regions, and the rest of the unannotated regions with length
+being near 1000 bp could be new coding genes. Some of the unannotated genes
+could be pseudogene regions.  We used strandnesses of known genes to find
+strands of segments although we could not tell strands of the predicted
+transcripts solely based on our RNA-seq data because of our strand-nonspecific
+RNA-seq technique.  Of
+the \Sexpr{length(grTx)} expressed transcripts, 
+\Sexpr{sum(strand(grTx)=='*')}
+included annotated genes of both strands
+(\Sexpr{as.integer(sum(strand(grTx)=='*')/length(grTx)*100)}\%).
+We filtered out the transcripts with genes with conflicted both strands leaving
+\Sexpr{length(grTx2)} expressed transcripts. 
+Figure \ref{figUTRLength} shows length distributions of 5' and 3' UTR of the
+expressed transcripts without genes with conflicted strands.  Many transcripts
+appeared to start and end near the boundaries of annotated genes.  Gene
+expression levels for adjacent genes within expressed transcripts were more
+correlated than those between adjacent transcripts 
+(fig.\ \ref{figExpressionTwoGene}).
+Among the \Sexpr{length(unique(matchMatrix(matchGenesTx2)[,"subject"]))} segments
+with genes of all of the same strands, 
+\Sexpr{sum(number.genes.tx==1)} segments contained a single gene while 
+\Sexpr{sum(number.genes.tx > 1)} contained more than one gene.
+
+\subsection*{Predictions of small RNAs}
+
+We predicted
+\Sexpr{length(colnames(smutans.genes.criteria))-1} putative non-coding RNAs, most of
+which were already annotated.  
+Of the \Sexpr{length(colnames(smutans.genes.criteria))-1} predicted non-coding RNAs
+\Sexpr{length(txRNAz.unannotated)} were in  
+unannotated non-coding regions (table \ref{tabUnannotatedRNA}).  Using
+\program{RNAplex} \citep{Tafer2008} and \program{RNAplfold} \citep{Bernhart2006}
+we predicted target genes of the putative non-coding RNAs.  Table \ref{tab:goseq}
+shows functional categories enriched with target genes for predicted small RNAs.
+We found that a few of the putative small RNAs were structured with stable stems
+(figs.\ \ref{figRNALocus1}, \ref{figRNALocus2}, and \ref{figRNALocus3}).  
+\fixme{We might be able to 
+further experimentally verify potentially important small RNAs.}
+
+\subsection*{Differential expression of RNA transcripts}
+
+<<echo=FALSE,result=hide>>=
+#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
+#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
+#                            condA="ua159", condB="tw1" )
+#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
+load("glucoseOnly.RData")
+smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
+tableGlucoseOnly <- smutans.de2List( glucoseOnly, qval=1e-3 )
+
+# Gene descriptions
+geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
+geneDescription <- geneDescription[,c(4,7)]
+colnames(geneDescription) <- c("Gene ID","Gene Description")
+# Functional classes
+functionalDescription <- read.csv("feature-genome.out-geneonly-oralgen-functional.txt",head=FALSE)
+functionalDescription <- functionalDescription[,c(2,12)]
+colnames(functionalDescription) <- c("Gene ID","Functional Class")
+# Find potential differentially expressed genes, and add gene descriptions, and
+# functional classes
+x <- smutans.de2List( glucoseOnly, qval=1e-3 )
+x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
+colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
+x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
+tableGlucoseOnly <- x
+@
+
+Two independent factors of the biological processes included the effect of CcpA
+and the treatment of mono-saccharides (glucose versus galactose), resulting in
+four pair-wise comparisons. We compared two conditions of one factor with the
+other factor being fixed at one condition. For example, we compared UA159
+wild-type and TW1 mutant-type strains under the condition of glucose-grown cells
+to determine differentially expressed transcripts or genes potentially due to
+presence and absence of deactivation of \textit{ccpA} genes.  We performed a
+similar comparison with galactose-grown cells. We replicated four
+biological experiments for the condition combination of UA159 wild-type grown
+in galactose, and three replicates for the other three conditions.
+
+\subsection*{Comparison between wild-type and mutant-type in glucose-grown cells}
+
+Table \ref{tabGlucoseOnly} lists differentially expressed genes in the
+comparison of UA159 wild-type and TW1 mutant-type grown with glucose.
+Of \Sexpr{nrow(counts(smutansGenes))} genes, 
+\Sexpr{nrow(tableGlucoseOnly)} genes were found to be differentially expressed
+at FDR of 0.1\%.  We confirmed that the most up-regulated genes in the TW1
+mutant-type encoded the components of the PDH enzyme complex (SMU.1421,
+SMU.1422, SMU.1423, and SMU.1424) as shown previously.  
+These genes appeared to be transcribed in
+a single operon (fig.\ \ref{figUCSC}). 
+Figure \ref{figPieGlucoseOnly} shows the
+proportions of \program{oralgen} databases functional classes by the number of
+differentially expressed genes.  Notably, we found
+\Sexpr{length(grep("Energy metabolism",tableGlucoseOnly[,"Functional Class"]))}
+genes that were involved in energy metabolism where sugars and fermentation
+subclasses were profound. We found 
+\Sexpr{length(grep("PTS",tableGlucoseOnly[,"Functional Class"]))}
+genes in PTS class of signal transduction. Another subclass, 
+``carbohydrates, organic alcohols, and acids,'' in
+``transport and binding proteins'' was following next in the number of
+differentially expressed genes. We also found  
+\Sexpr{length(grep("hypothetical",tableGlucoseOnly[,"Gene Description"]))}
+genes annotated as hypothetical, which might constitute the unknown functional
+class. These results appeared to be consistent with those of \citet{Abranches2008}.
+
+We further tested functional category enrichments using gene ontology terms and KEGG
+pathways. Table \ref{tab:goseqGlucoseOnly} shows that differentially expressed
+genes were associated with the following gene ontology terms:
+phosphotransferase system, 
+glycogen biosynthetic process, and
+sugar:hydrogen symporter activity.
+Among the KEGG pathways, starch and sucrose metabolism and phosphotransferase
+system (PTS) appeared to be highly associated with the differentially expressed
+genes (table \ref{tab:kegseqGlucoseOnly}).
+Figure \ref{figDotplotGlucoseOnly} shows that differentially expressed genes
+tended to increase in expression levels in response to the deletion of
+\textit{ccpA} genes.  Profoundly upregulated genes were putative Clp proteinase
+(SMU.1425), branched-chain alpha-keto acid dehydrogenase subunit E2 (SMU.1421),
+and putative transcriptional regulator (SMU.1599).  Much less proportion of the
+differentially expressed genes were found to be decreased in expression levels
+in response to the lack of CcpA (fig.\ \ref{figDotplotGlucoseOnly}).
+Down-regulated genes include cytoplasmic alpha-amylase (SMU.1590),
+beta-D-fructosyltransferase (SMU.2028), and putative glycerol uptake facilitator
+protein (SMU.396).  Fructosyltransferase converts sucrose to fructose.
+
+Figure \ref{figFoldPValueGlucoseOnly} shows fold changes and $p$-vallues are
+negatively correlated. \fixme{Why did I want to note this observation?}
+
+\subsection*{Comparison between wild-type and mutant-type in galctose-grown cells}
+
+<<echo=FALSE,result=hide>>=
+#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
+#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
+#                              condA="ua159", condB="tw1" )
+#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
+load("galactoseOnly.RData")
+x <- smutans.de2List( galactoseOnly, qval=1e-3 )
+x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
+colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
+x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
+tableGalactoseOnly <- x
+@
+We found that differentially expressed genes in galactose-grown cells were
+smaller than those in glucose-grown cells, which was different from the result
+of \cite{Abranches2008}. 
+Energy metabolism was the most
+profoundly differentially expressed (fig.\ \ref{figPieGalactoseOnly}).
+Subclasses pyruvate family in amino acid biosynthesis, 
+and degradation of proteins, peptides, and glycopeptides in protein fate
+were following the next. Almost a quarter of the genes were unknown.
+Down-regulated genes included cytoplasmic alpha-amylase (SMU.1590), and
+beta-D-fructosyltransferase (SMU.2028)
+(table \ref{tabGalactoseOnly}). We could not find gene ontology terms that were
+significantly associated with differentially expressed genes. It might be
+because many genes were unknown in their functions 
+(fig.\ \ref{figDotplotGalactoseOnly}).
+However, a few KEGG pathways including valine, leucine and isoleucine
+biosynthesis were found to be significant (table \ref{tab:kegseqGalactoseOnly}).
+
+\subsection*{Comparison between glucose- and galactose-grown wild-type cells}
+
+<<echo=FALSE,result=hide>>=
+#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
+#ua159only <- smutans.de2( ua159only, type="ua159", 
+#                          condA="glucose", condB="galactose" )
+#save(ua159only, file=file.path("ua159only.RData"))
+load("ua159only.RData")
+x <- smutans.de2List( ua159only, qval=1e-10 )
+x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
+colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
+x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
+tableUA159Only <- x
+@
+Among the four pairwise comparisons this showed the largest number of
+differentially expressed genes (table \ref{tabUA159Only}). We used a smaller
+cutoff value for listing them. 
+Figure \ref{figPieUA159Only} shows proportons of functional categories that
+included differentially expressed genes between glucose-grown and
+galactose-grown UA159 wild-type cells. Qualitatively, energy metabolism and PTS
+classes were profound. 
+``Energy metabolism; Fermentation'' had highly differentially expressed genes
+(fig.\ \ref{figDotplotUA159Only})
+This was silmilar to the comparison between UA159
+wild-type and TW1 mutant-type grown in glucose. 
+More than a quarter of the
+differentially expressed genes were unknown, unassigned, or hypothetical.
+Consecutive genes (SMU.1488c to SMU.1496) were upregulated in 
+catabolite-derepressing conditions or in galactose-grown cells
+(\fixme{Are these the genes of the lactose (lac) operon?})
+(table \ref{tabUA159Only}).  Another set of genes (SMU.886 to SMU.889) were also
+upregulated (\fixme{Are these called galKTE?}). Whereas the
+four genes appeared to be transcribed together under galactose condition,
+two segments seemed to be transcribed by dividing the four genes into two pairs
+of genes (fig.\ \ref{figUCSC2}).  Genes from SMU.1535 to SMU.1539, which
+appeared to be transcribed together, were upregulated.
+Of the \Sexpr{nrow(tableUA159Only)} differentially expressed genes at
+$1 \times 10^{-10}$ cutoff of multiple testing adjusted p-value, 
+\Sexpr{length(grep("hypothetical",tableUA159Only[,"Gene Description"]))} were
+hypothetical, and most of them were down-regulated
+(fig.\ \ref{figDotplotUA159Only}).
+We also tested functional category enrichments using gene ontology terms 
+(table \ref{tab:goseqUA159Only}) and KEGG pathways 
+(table \ref{tab:kegseqUA159Only})
+where carbohydrate metabolic process and Galactose metabolism were top listed,
+respectively.
+
+\subsection*{Comparison between glucose- and galactose-grown mutant-type cells}
+
+We found sugar metabolism and PTS signal transduction to be profound in
+differential expression in the comparison of glucose-grown and galactose-grown
+TW1 mutant-type cells (fig.\ \ref{figPieTW1Only}). Many of the genes were also
+hypothetical. We found similar genes differentially expressed in TW1 mutant-type
+cells to those in UA159 wild-type 
+(table \ref{tabTW1Only} and fig.\ \ref{figDotplotTW1Only}).
+The comparisons between glucose-grown and galactose-grown UA159, 
+and that from TW1 mutant-type showed very similar functional categories
+although ranks of gene ontology terms were different
+(tables \ref{tab:goseqUA159Only} and \ref{tab:goseqTW1Only}).
+Lactose metabolic process was top listed in TW1, and
+carbohydrate metabolic process was top listed in UA159.
+KEGG pathways in TW1 were also similar to those in UA159 
+(table \ref{tab:kegseqTW1Only}) .
+
+\subsection*{Comparison between glucose- and galactose-grown cells using all samples}
+As previously figure \ref{figClust} shows the nutrient effect was stronger, and
+the effects of nutrient changes appeared to be similar whether UA159
+wild-type or TW1 mutant-type was used, we used all of the 
+\Sexpr{length(colnames(counts(smutansGenes)))} samples to compare glucose-grown
+and galactose grown cells.
+<<echo=FALSE,results=hide>>=
+allSample <- newSmutans( smutansGenes, 
+                         title="All Samples for Comparing Conditions" )
+allSample <- smutans.de2( allSample )
+save(allSample, file=file.path("allSample.RData"))
+#load("allSample.RData")
+tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
+                                                   allSample )
+#addmargins( tabUA159onlyAndAllSample )
+@
+Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
+\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found to be differentially
+expressed in both the analysis of using only UA159-grown cells and the analysis
+of using all of the 13 samples. The
+analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
+genes. The analysis using the partial samples found only
+\Sexpr{tabUA159onlyAndAllSample[2,1]} genes.  We would expect that it was 
+because of higher power of the two-factor analysis with more data.
+
+\subsection*{Differentially expressed small RNAs and transcripts}
+
+<<echo=FALSE,results=hide>>=
+ua159only <- newSmutans( smutansGenes2, title="UA159 Only" )
+ua159only <- smutans.de2( ua159only, type="ua159", 
+                          condA="glucose", condB="galactose" )
+allSample <- newSmutans( smutansGenes2, 
+                         title="All Samples for Comparing Conditions" )
+allSample <- smutans.de2( allSample )
+tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
+                                                   allSample )
+# smutans.de2List( allSample )[grep("locus",rownames(smutans.de2List(allSample))),]
+@
+We found a few predicted small RNAs that appeared to be differentially
+expressed: locus69, locus101, locus7, and locus1. 
+Two non-coding regions of locus69 and locus101 seemed to be too weak.
+However, locus1 and locus7 seemed to have some noticeable differences between
+glucose and galactose conditions.
+We also found a few predicted small RNAs that appeared to be differentially
+expressed in the comparison of the two types UA159 and TW1: locus68,
+locus8, and locus10.  The locus68 was overlapped with T-box region.
+The two regions of locus8 and locus10 showed very weak and subtle differences
+between UA159 and TW1. 
+<<echo=FALSE,results=hide>>=
+allSampleType <- newSmutans( smutansGenes2, title="All samples for type" )
+allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
+# smutans.de2List( allSampleType )[grep("locus",rownames(smutans.de2List(allSampleType))),]
+@
+Figures \ref{figRNALocus1}, \ref{figRNALocus2}, and \ref{figRNALocus3} shows MFE
+structures with encoding positional entropy of differentially expressed putative
+structural non-coding RNAs.
+
+\section*{Discussion}
+
+Our current knowledge of the transcriptome is limited to annotated or predicted
+genes.  Whole transcriptome sequencing using high throughput sequencing
+technologies, or deep RNA sequencing (RNA-seq) allowed to reveal the
+transcriptome topography of \textit{S.\ mutans}.  Instead of aiming at
+pin-pointing transcript start or end sites we focused on differential expression
+of genes in response to nutirent availability and deletion of \textit{ccpA} from
+UA159.  Although we did not use strand-specific RNA-seq that could be important
+in gene-dense transcriptomes of species like bacteria or archaea, our
+application of a hidden Markov model for predicting transcripts to RNA-seq data
+in conjunction with known gene annotations allowed to infer segments of
+putatively co-transcribed regions. In doing so, we attempted to ensure a
+high-quality transcriptome assembly.  Although this map of transcripts did not
+precisely tell us where transcripts start and end, it informed us of which open
+reading frames were transcribed together.  We also found discrepancies of
+annotated genes and their transcriptions: a gene contains two transcripts, and
+non-coding regions contain transcripts.  We hope that our study could serve as a
+foundation for a comprehensive study of the \textit{S.\ mutans} transcriptome.
+
+\section*{Acknowledgments}
+\fixme{Add more peoples or fundings if needed.}
+This study was funded by the National Institute of Allergy and Infectious
+Disease, US National Institutes of Health, under grant number AI073368-01A2
+(to M.J.S.\ and A.S).  Additional support was provided by National Science
+Foundation CAREER Award DBI-0644111 and a David and Lucile Packard
+Fellowship for Science and Engineering (to A.S.).  
+We thank Christopher Browngardt for assistance with labelling functional classes
+of \textit{S.\ mutans} genes.
+
+\renewcommand*{\refname}{Literature Cited}
+\bibliographystyle{../../latex/bst/mbe}
+\bibliography{siepel-mutans}
+\clearpage{}
+
+\section*{Tables}
+
+\clearpage{}
+
+\section*{Figures}
+
+\end{document}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% REMOVED TEXT
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Regulation of carbon catabolite repression (CCR) in \textit{Streptococcus
@@
 
-Small RNAs and their targets can be predicted \citep{Tafer2008a}.
-Although our RNA-seq data did not sample specifically non-coding small RNAs, we
-retained small RNAs in the sample by attempting to deplete only ribosomal RNAs
-from the sample. Because non-coding small RNAs and coding messenger RNAs were
-mixed in the sample, we sought a way of predicting small RNAs using
-\program{RNAz} developed by \citet{Gruber2010}.  \program{RNAz} used homologous
-sequences and RNA secondary structures to predict putative non-coding RNAs. We
-confirmed the several predicted small non-coding RNAs experimentally
-\citep{Altuvia2007} including the RNA-seq data set. Because small RNAs could
-be associated with CCR regulation of genes (CITE), we determined differentially expressed small
-RNAs between RNA samples. We found a few small RNAs to be differentially
-expressed.
-
-Bacteria can selectively use substrates from different carbon sources. In the
-presence of a preferred carbon source, catabolic system of secondary substrates
-are often regulated or shutdown: called carbon catabolite repression
-\citep{Gorke2008}. CCR also regulates the expression of virulence factors in many
-pathogenic bacteria. A classical example is the glucose-lactose diauxie in
-\textit{E.\ coli}. A significant number of genes can be subject to CCR.  CCR in
-bacteria involves regulatory networks that activate or silence genes in response
-to carbon source and availability.  \textit{S.\ mutans} possesses redundant
-systems for CCR.  We could measure directly the expression levels of
-\textit{fruA} in \textit{S.\ mutans}.  We assess the contribution of CcpA to CCR
-of predicted operons.  Clearly, deletion of \textit{ccpA} resulted in
-substantial relief of the CCR of \textit{fruA} expression exerted by both
-glucose and fructose.
-
 We aims at refining the results of \cite{Abranches2008} in more quantative ways.
@@
 determine differentially expressed regions in UA159 genome.
-
-\fixme{We need to quantify some uncertainty of transcript predictions.}
-
-\section{Materials and Methods}
-
-\subsection{Bacterial strains and growth conditions}
-\fixme{Burne group should be able to describe this part. The following is just
-from somewhere for the purpose of writing template. Please, rewrite it.}
-\textit{S.\ mutans} UA159 is a human cariogenic isolate. The whole genome is
-availabe at NCBI \citep{Ajdic2002}. Strain TW1 was derived from UA159 by
-replacing a part of \textit{ccpA} gene \citep{Wen2002}.  \textit{S.\ mutans}
-strains were inoculated into TV medium supplemented with 0.5\% galactose or
-glucose and grown to early exponential phase (OD600, 0.2).  Then fructose was
-added to the medium at concentrations (0,0.05,3,5,10 mM) for 3.5h to induce
-\textit{fruA} expression.
-
-Expression levels of \textit{fruA} in \textit{S.\ mutans} strains that were
-grown to early exponential phase (OD600, 0.2) in TV medium supplemented with
-0.5\% glucose or galactose.
-
-\subsection{RNA isolation}
-\fixme{Burne group should be able to describe this part. The following is just
-from somewhere for the purpose of writing template. Please, rewrite it.}
-RNA was isolated from S. mutans (1) by using 50 ml cultures that were grown
-under the desired conditions and harvested by centrifugation at $4^{\circ}$C. Pelleted
-cells were resuspended in 400 ul of diethyl pyrocarbonate-treated water, 800 ul
-of RNAprotect reagent (Qiagen, Inc., Chatsworth, CA) was added, and the samples
-were incubated at room temperature for 5 min with vortexing for 10 s at 1-min
-intervals. Cells were then pelleted, resuspended in 250 ul Tris-EDTA (50:10)
-buffer, and transferred to 1.5 ml screw-cap tubes containing sterile glass beads
-(average diameter, 0.1 mm; Biospec, Bartlesville, OK), 10 ul of 1\% sodium
-dodecyl sulfate, and 300 ul of acid phenol-chloroform (5:1). After
-centrifugation for 15 min at maximum speed at $4^{\circ}$C, RNA was further purified
-using an RNeasy mini kit (Qiagen, Inc., Chatsworth, CA), including on-column
-DNase digestion with RNase-free DNase (Qiagen), as recommended by the supplier.
-
-\subsection{RNA sequencing}
-\fixme{SC wrote the second paragraph based on email correspondences between PDPB and CDD.
-Please rewrite it if necessary. Stanhope group should be able to describe this
-part. The following first paragraph is from a paper. Do not use them or rephrase
-them in any way. They are for informational purposes.}
-Sequencing libraries for the Illumina GA platform were constructed by shearing
-the enriched cDNA by nebulisation (35psi, 6 min) followed by end-repair with
-Klenow polymerase, T4 DNA polymerase and T4 polynucleotide kinase (to blunt-end
-the DNA fragments). A single 3' adenosine moiety was added to the cDNA using
-Klenow exo- and dATP. The Illumina adapters (containing primer sites for
-sequencing and flowcell surface annealing) were ligated onto the repaired ends
-on the cDNA and gel-electrophoresis was used to separate library DNA fragments
-from unligated adapters by selecting cDNA fragments between 200-250 bps in size.
-Fragmentation followed by gel-electrophoresis were used to separate library DNA
-fragments and size fragments were recovered following gel extraction at room
-temperature to ensure representation of AT rich sequences. Ligated cDNA
-fragments were recovered following gel extraction at room temperature to ensure
-representation of AT rich sequences. Libraries were amplified by 18 cycles of
-PCR with Phusion polymerase. Sequencing libraries were denatured with sodium
-hydroxide and diluted to 3.5 pM in hybridisation buffer for loading onto a
-single lane of an Illumina GA flowcell. Cluster formation, primer hybridisation
-and single-end, 36 cycle sequencing were performed using proprietary reagents
-according to manufacturers' recommended protocol (https://icom.illumina.com/).
-The efficacy of each stage of library construction was ascertained in a quality
-control step that involved measuring the adapter-cDNA on a Agilent DNA 1000
-chip. A final dilution of 2 nM of the library was loaded onto the sequencing
-machine.
-
-Deep sequencing of the RNA samples were performed at the Cornell University Life
-Sciences Core Laboratories Center (Ithaca, NY).  RNA extraction was performed
-using a modified Qiagen RNAease protocol (Qiagen, Inc., Chatsworth, CA): cells
-were pelleted, resuspended in RNA protect reagent, frozen, then resuspended in
-TE, mechanically disrupted in a bead beater in the presence of glass beads and
-acidic phenol. After these steps, we used the Qiagen kit to obtain total RNAs.
-The RNA sample was then passed through the Ambion MicrobExpress kit twice, which
-removed ribosomal RNAs only using magnetic beads.  This procedure left small
-RNAs and transfer RNAs intact.  We followed the TruSeq Illumina kit to prepare
-the cDNA libraries.  We did not intend to enrich short reads in either small
-RNAs or transcript boundaries.
-
-\subsection{Statistical analysis for differential expressions}
-We used an R package, DESeq, \citep{Anders2010} to determine differentially
-expressed genes based on negative binomial model originally developed by
-\citep{Robinson2007}.  Steps of analyzing RNA-seq data for determining
-differentially expressed genes are described in \cite{Oshlack2010}. We used a
-short read aligner called \program{BWA} \citep{Li2009a} to map the sequenced
-short reads on a reference genome, \textit{Streptococcus mutans} UA159 (NCBI
-accession NC\_004350). Because the aligner allowed a few gaps for efficient
-alignment of millions of sequences of size being 100 base pairs long, short
-reads that consist of RNAs and sequencing adapter sequences would be not mapped.
-We used \program{cutadapt} \citep{Martin2011} to remove parts of adapter
-sequences from short reads before mapping on the reference genome. Mapped short
-read alignments were converted into readable formats using \program{samtools}
-\citep{Li2009b}. We counted short reads aligned to annotated genes in the
-reference genome, forming a table of read counts of the genes. Statistical
-software R \citep{R2011} with \Rpackage{DEseq} \citep{Anders2010} was
-subsequently employed to infer differentially expressed in different biological
-conditions.  We also generated tracks for our recently released Streptococcus
-Genome Browser (http://strep-genome.bscb.cornell.edu) that summarize gene
-expression levels based the short read alignments. This allowed us to visualize
-parts of the reference genome for comparing different RNA samples. Further
-analyses such as functional category enrichment test \cite{Young2010} were also
-employed.
-
-\subsection{Transcript predictions}
-\citet{Martin2010} developed a hidden Markov model to predict transcripts by
-using site-wise expression levels. We used the following options for the
-program: ``-c 10 -b 25 -force\_gp'' to bin expression levels into 25 parts to
-allow 10 emission states. We also used the genes as the predicted parts of
-transcripts.  Because we did not have strand-specific RNA-seq of the 13 samples
-due to limited resources, we could not tell in which directions transcripts
-based on RNA-seq data were oriented.
-
-\subsection{small RNAs}
-We used \program{RNAz} to predict small RNA candidates. Sequence alignments were
-a critical step for finding small RNAs in the approach. Phylogenies of
-streptococcus species could be considered to choose species related with
-Streptococcus mutans.  Alternatively, we found sequences similar to intergenic
-regions, and refined the alignments using a multiple sequence alignment tool
-such as MUSCLE \citep{Edgar2004a}. First, we exracted intergenic regions
-including up- and down-stream. We searched non-redundant database of bacterial
-genomes using BLAST \citep{Altschul1990} for similar intergenic sequences,
-subsequently, refining the sequence alignments using MUSCLE \citep{Edgar2004a}.
-Then, we applied \program{RNAz} to the alignments for scoring regions for small
-RNAs \citep{Washietl2005,Gruber2010}.  Targets of transcripts with high scores
-for small RNAs were also predicted using \program{RNAplex} \citep{Tafer2008},
-and \program{RNAplfold} \citep{Bernhart2006}.  Target genes were scored for each
-small RNA candidate, which was used to perform functional category enrichment
-tests of targets for each small RNA candidate. We could find functional
-categories that a few small RNAs might be associated with.
-
-\subsection{Gene category associations and virulence genes}
-We used two approaches for finding functional categories
-\citep{Young2010,Suzuki2011}. We briefly descirbe the prodecure of
-\citet{Suzuki2011}.  We assigned genes to Gene Ontology (GO) categories by
-comparing the \textit{S.\ mutans} genes to bacterial proteins from the Uniref90
-database using \program{BLASTP}, and then assigning the same GO classification
-as the target gene of the uniProt GOA database if the match had an $E$-value of
-$<1.0\times10^{-5}.$  A gene family was assigned a given classification if any
-of its genes was assigned that classification.  To test for associations with
-replacing gene transfers we used the recombination intensity estimated for each
-gene, and for additive gene transfers we used the number of additive transfers
-inferred for each family.  To test for significance, we performed a Mann-Whitney
-$U$ test of the values (recombination intensities or numbers of transfers)
-associated with a given category vs.\ the values for the all other
-genes/families.  We used the \cite{Benjamini1995} method to correct for multiple
-comparisons.
-\fixme{A short description about \citep{Young2010}.}
-
-\comment{
-Findings include:
-\begin{enumerate}
-\item New genes, open reading frames that are not annotated in NCBI's genome database.
-\end{enumerate}
-}
-
-\section{Results}
-We sequenced a whole prokaryotic transcriptome of \textit{S.\ mutans}, and
-mapped short reads to the genomic sequence of the reference genome UA159
-\citep{Ajdic2002}. 
-\citet{Martin2010} applied a hidden Markov model to use short reads mapped to a
-genomic sequence in order to parase \textit{B.\ anthracis} genome into a
-sequence of transcripts. Most of annotated genes belong to the predicted
-transcripts. We found discrepancies of annotated genes and their transcriptions:
-1. A gene contains two transcripts, 
-2. Non-coding regions contain transcripts.
-\fixme{We could find more discrepancies. Try to find them}. 
-Expression levels of adjacent genes within a transcript should be more
-correlated to each other than those of adjacent genes belonging to separate transcripts
-\citep{Martin2010}. 
-We identified a set of transcripts, albeit not a full, including small RNAs, and novel
-transcripts. 
-We hope that our study could serve as a foundation for a comprehensive study of
-the \textit{S.\ mutans} transcriptome.
-Our current knowledge of the transcriptome is limited to annotated or predicted genes.
-Whole transcriptome sequencing using high throughput sequencing (HTS)
-technologies, or deep RNA sequencing (RNA-seq) allowed to reveal the
-transcriptome topography of \textit{S.\ mutans}. We did not aim at precisely
-pin-pointing transcript start or end sites.
-
-We admit that strand-specific RNA-seq would be important in gene-dense
-transcriptomes of species like bacteria or archaea.
-
-\subsection{From Reviews}
-
-We could not use short-read assemblers such as Velvet \citep{Zerbino2008} 
-and Trinity that had been developed to assembly transcriptome. Because genes in
-bacteria are so dense, RNA-seq data without strands did not allow us to apply
-any state-of-the-art software of transcriptome assembly.
-We ensured a high-quality transcriptome assembly.
 
@@
 
-<<manuscript,echo=FALSE,results=hide>>=
-# not run when smutans R package is installed
-# Use this when smutans R package is being tested.
-library(DESeq)
-library(GenomicRanges)
-library(plotrix)
-source("smutans/R/class_and_slots.R")
-source("smutans/R/core.R")
-source("smutans/R/methods.R")
-load("smutans/data/smutansData.tx.RData")
-load("smutans/data/smutansData.txGenes.RData")
-load("smutans/data/smutansData.txPileup.RData")
-load("smutans/data/smutans.feature.genes.RData")
-load("smutans/data/smutans.genes.criteria.RData")
-load("smutans/data/smutans.go.genes.RData")
-load("smutans/data/smutans.cat.desc.RData")
-load("smutans/data/smutansGenes.RData")
-@
-<<manuscript,eval=FALSE,echo=FALSE,results=hide>>=
-# not run without smutans R package installed
-# Remove eval=FALSE when the R package is installed
-library(smutans)
-load(smutansData.tx)
-load(smutansData.txGenes)
-load(smutansData.txPileup)
-load(smutans.feature.genes)
-data(smutans.genes.criteria)
-data(smutans.go.genes)
-data(smutans.cat.desc)
-data(smutansGenes)
-@
-<<echo=FALSE,result=hide>>=
-grTx <- smutansData.tx[elementMetadata(smutansData.tx)$state > 1]
-mtch2 <- findOverlaps(smutansData.txGenes,grTx)
-grNoncoding <- gaps(smutansData.txGenes)
-# A bug in GenomicRanges: the last segment is the while genome. I removed it
-# from the noncoding ranges.
-grNoncoding <- grNoncoding[-length(grNoncoding)] 
-mtchNoncoding <- findOverlaps(grNoncoding,grTx)
-# Expressed transcripts without annotated genes
-grTxWGenes <- grTx[unique(matchMatrix(mtch2)[,2])]
-grTxWOGenes <- grTx[-unique(matchMatrix(mtch2)[,2])]
-# Find transcripts with genes with conflicted strands
-grGenesPlus <- smutansData.txGenes[strand(smutansData.txGenes)=='+']
-grGenesMnus <- smutansData.txGenes[strand(smutansData.txGenes)=='-']
-mtchPlus <- findOverlaps(grGenesPlus,grTxWGenes)
-mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
-txWithConflictedGenes <- intersect(matchMatrix(mtchPlus)[,2],matchMatrix(mtchMnus)[,2])
-grTx2 <- grTxWGenes[-txWithConflictedGenes] 
-# Lengths of 5' and 3' UTRs
-mtchPlus <- findOverlaps(grGenesPlus,grTx2)
-mtchMnus <- findOverlaps(grGenesMnus,grTx2)
-length5UTR <- c()
-length3UTR <- c()
-for (i in unique(matchMatrix(mtchPlus)[,"subject"])) {
-  x <- matchMatrix(mtchPlus)[matchMatrix(mtchPlus)[,"subject"]==i,"query"]
-  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
-  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
-  length5UTR <- c(length5UTR,y)
-  length3UTR <- c(length3UTR,z)
-}
-for (i in unique(matchMatrix(mtchMnus)[,"subject"])) {
-  x <- matchMatrix(mtchMnus)[matchMatrix(mtchMnus)[,"subject"]==i,"query"]
-  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
-  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
-  length5UTR <- c(length5UTR,z)
-  length3UTR <- c(length3UTR,y)
-}
-# Correlation of expression levels of adjacent genes.
-pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
-pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
-elementMetadata(smutansData.txGenes)["score"] <- pileup.y
-mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
-x <- rle(matchMatrix(mtch3)[,"subject"])
-comp.exp.two.genes <- c()
-for (i in x$values[x$lengths > 1]) {
-  y <- combn(matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==i],2)
-  for (j in 1:dim(y)[2]) {
-    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
-  }
-}
-# Correlation of expression levels of adjacent genes between transcripts.
-comp.exp.two.genes.between.transcript <- c()
-for (i in 1:length(x$values)) {
-  if (i == length(x$values)) {
-    break
-  }
-  y1 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i]]
-  i2 <- i + 1
-  y2 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i2]]
-  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
-  for (j in 1:dim(l)[1]) {
-    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
-  }
-}
-# plot(log10(comp.exp.two.genes.between.transcript[1,]), log10(comp.exp.two.genes.between.transcript[2,]))
-# cor(log10(comp.exp.two.genes.between.transcript[1,]), log10(comp.exp.two.genes.between.transcript[2,]))
-@
-
-\subsection{My Results}
-% Statistics of transcripts, and genes.
-The input RNA-seq data is a list of site-wise non-negative integers that
-represents a number of short reads mapped to a particular genomic position
-\citet{Martin2010}, which is called a RNA-seq ``coverage'' data. We designated a
-gene as an expressed if the gene belongs to a transcript with at least C2
-level.
-The number of predicted expressed transcripts was
-\Sexpr{sum(elementMetadata(smutansData.tx)$state > 1)}
-(\Sexpr{as.integer(sum(elementMetadata(smutansData.tx)$state > 1)/length(smutansData.tx)*100)}\%)
-out of 
-\Sexpr{length(smutansData.tx)} 
-fragments.
-Of the total \Sexpr{length(smutansData.txGenes)} of genes,
-\Sexpr{length(matchMatrix(mtch2)[,1])}
-(\Sexpr{as.integer(length(matchMatrix(mtch2)[,1])/length(smutansData.txGenes)*100)}\%) 
-were designated as expressed. 
-Genes that were not designated as expressed could be due to false gene
-annotations or possibility of true negative expression under the experimental
-conditions. 
-Figure \ref{figLength} shows the length distribution of expressed and
-not-expressed genes. 
-Of the \Sexpr{length(grTx)} expressed transcripts, 
-\Sexpr{length(grTxWOGenes)} contained no annotated genes.
-Figure \ref{figTranscriptLength} shows the length distribution of expressed
-transcripts with annotated coding genes and unannotated regions.
-Unannotated regions can include potential non-coding regions and new
-coding genes.  Transciprts with unannotated regions appeared to be shorter than those
-with coding genes. Most of the unannotated regions with length being near 360 bp
-could be non-coding regions, and the rest of the unannotated regions with length
-being near 1000 bp could be new coding genes. Some of the unannotated genes
-could be pseudogene regions.
-Because we used a RNA-seq technique without considering strand-specific, we
-could not tell the strandness of the predicted transcripts. Of
-the \Sexpr{length(grTx)} expressed transcripts, 
-\Sexpr{length(txWithConflictedGenes)}
-included annotated genes of both strands
-(\Sexpr{as.integer(length(txWithConflictedGenes)/length(grTx)*100)}\%).
-We filter out the transcripts with genes with conflicted both strands leaving
-\Sexpr{length(grTx2)} expressed transcripts. 
-Figure \ref{figUTRLength} shows length distributions of 5' and 3' UTR of the
-expressed transcripts without genes with conflicted strands.
-Many transcripts appeared to start and end near the boundaries of annotated genes.
-Gene expression levels for adjacent genes within expressed transcripts were more
-correlated than those between adjacent transcripts 
-(Figure \ref{figExpressionTwoGene}).
-% Question: What are the promoter scores?
-
-Name the transcripts.
-I found xx transcripts. Among them xx \% were mono-cistronic operons. xx \% were
-poly-cistronic operons. xx transcripts could not defined because RNA-seq short
-reads were not precise. Transripts were assigned with functional categories, and
-differentially expressed transcripts were found. Transcripts and functional
-categories: how can I assess that transcripts consist of genes with similar
-functional cateogires.
-
-% Small RNAs predicted
-Although we aimed at sequencing messenger RNAs, we expected to find regulatory
-small RNAs in the RNA-seq data because we attempted to deplete structrual
-ribosomal RNAs. \program{RNAz} is a software package for detection of structural
-noncoding RNAs using comparative genomic data \citep{Gruber2010}. The method has
-been applied to different genomic data (CITE). We applied the method to predict
-potential non-coding small RNAs in \textit{S.\ mutans} as a supplement to
-evidence for small RNAs provided by the RNA-seq data. With increased evidence of
-putative small RNAs we could proceed to further experimentally verify
-potentially important small RNAs.  We first predicted
-\Sexpr{length(colnames(smutans.genes.criteria))-1} putative small RNAs. Using
-\program{RNAplex} \citep{Tafer2008} and \program{RNAplfold} \citep{Bernhart2006}
-we predicted target genes of the putative small RNAs.  Table \ref{tab:goseq}
-shows functional categories enriched with target genes for predicted small RNAs.
-We found that a few of the putative small RNAs were structured with stable stems
-(SHOW FIGURES).
-
-Name the small RNAs.
-RNAz predicted a few small RNAs. Some of them were differentially expressed.
-Some functional categories were enriched with target genes of the predicted
-genes.
-
-% UCSC genome browser resources
-We created new tracks for our recently released {\em Streptococcus} Genome
-Browser (http://strep-genome.bscb.cornell.edu) that summarized the results of
-this \textit{S.\ mutans} transcriptome analysis alongside known genes, predicted
-small RNAs, predicted target genes and predicted transcripts. These tracks can
-be used to inspect loci of interest and to compare the results of different
-RNA-seq data sets. They can also be queried and intersected with other tracks
-using the UCSC Table Browser.
-
-<<echo=FALSE,result=hide>>=
-#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
-#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
-#                            condA="ua159", condB="tw1" )
-#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
-load("glucoseOnly.RData")
-smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
-tableGlucoseOnly <- smutans.de2List( glucoseOnly, qval=1e-3 )
-
-# Gene descriptions
-geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
-geneDescription <- geneDescription[,c(4,7)]
-colnames(geneDescription) <- c("Gene ID","Gene Description")
-# Functional classes
-functionalDescription <- read.csv("feature-genome.out-geneonly-oralgen-functional.txt",head=FALSE)
-functionalDescription <- functionalDescription[,c(2,12)]
-colnames(functionalDescription) <- c("Gene ID","Functional Class")
-# Find potential differentially expressed genes, and add gene descriptions, and
-# functional classes
-x <- smutans.de2List( glucoseOnly, qval=1e-3 )
-x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
-colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
-x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
-x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
-tableGlucoseOnly <- x
-@
-
-% Differential expression study
-We sequenced 
-\Sexpr{length(colnames(counts(smutansGenes)))}
-RNA samples for studying differential gene expression to better understand
-biological processes. Two independent factors of the biological processes
-includes the effect of CcpA and the treatment of monosaccharides (glucose versus
-galactose). The two factors results four pair-wise comparisons. We compared two
-conditions of one factor with the other factor being fixed at one condition. For
-example, we could compare UA159 wild-type and TW1 mutant-type strains under the
-condition of glucose-grown cells to find differentially expressed associatively
-with deactivation of \textit{ccpA} genes. We performed a similar comparison with
-galactose-grown cells. While we replicated four biological experiments for the
-condition combination of UA159 wilde-type grown in galactose, we had three for
-the other three conditions. Because of limited resources of RNA deep sequencing,
-we considered three replicates as a maximum number to but still adequately
-detect differentially expressed genes.
-
-We list differentially expressed genes in the comparison of UA159 wild-type and
-TW1 mutant-type in Table \ref{tabGlucoseOnly}.  
-Of \Sexpr{nrow(counts(smutansGenes))} genes, 
-\Sexpr{nrow(tableGlucoseOnly)} genes were found to be differentially expressed
-at FDR of 0.1\%. 
-We confirmed that the most upregulated genes in the TW1 mutant-type encoded the
-components of the PDH enzyme complex (SMU.1421, SMU.1422, SMU.1423, and
-SMU.1424).  
-Figure \ref{figPieGlucoseOnly} shows the proportions of functional categories 
-of the differentially expressed genes. Notably, we found
-\Sexpr{length(grep("Energy metabolism",tableGlucoseOnly[,"Functional Class"]))}
-genes that were involved in energy metabolism. Sugars and fermentation
-subclasses were profound in energy metabolism. We also found 
-\Sexpr{length(grep("PTS",tableGlucoseOnly[,"Functional Class"]))}
-genes in PTS class of signal transduction. Another subclass, 
-``carbohydrates, organic alcohols, and acids,'' in
-``transport and binding proteins'' was following next in the number of
-differentially expressed genes. We also found  
-\Sexpr{length(grep("hypothetical",tableGlucoseOnly[,"Gene Description"]))}
-genes annotated as hypothetical, which might constitute the unknown functional
-class. These results were consistent with those of \citet{Abranches2008}.
-Profoundly upregulated genes in response to the lack of CcpA were
-putative Clp proteinase (SMU.1425), 
-branched-chain alpha-keto acid dehydrogenase subunit E2 (SMU.1421), and
-putative transcriptional regulator (SMU.1599).
-Down-regulated genes include cytoplasmic alpha-amylase (SMU.1590), 
-beta-D-fructosyltransferase (SMU.2028), and
-putative glycerol uptake facilitator protein (SMU.396).
-Fructosyltransferase converts sucrose to fructose.
-
-<<echo=FALSE,result=hide>>=
-#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
-#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
-#                              condA="ua159", condB="tw1" )
-#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
-load("galactoseOnly.RData")
-x <- smutans.de2List( galactoseOnly, qval=1e-3 )
-x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
-colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
-x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
-x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
-tableGalactoseOnly <- x
-@
-We found that differentially expressed genes in galactose-grown cells were
-smaller than those in glucose-grown cells, which was different from the result
-of \cite{Abranches2008}. 
-Energy metabolism was the most
-profoundly differentially expressed (fig.\ \ref{figPieGalactoseOnly}).
-Subclasses pyruvate family in amino acid biosynthesis, 
-and degradation of proteins, peptides, and glycopeptides in protein fate
-were following the next. Almost quarter of the genes were unknown.
-Down-regulated genes included cytoplasmic alpha-amylase (SMU.1590), and
-beta-D-fructosyltransferase (SMU.2028)
-(table \ref{tabGalactoseOnly}).
-
-<<echo=FALSE,result=hide>>=
-#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
-#ua159only <- smutans.de2( ua159only, type="ua159", 
-#                          condA="glucose", condB="galactose" )
-#save(ua159only, file=file.path("ua159only.RData"))
-load("ua159only.RData")
-x <- smutans.de2List( ua159only, qval=1e-10 )
-x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
-colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
-x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
-x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
-tableUA159Only <- x
-@
-Figure \ref{figPieUA159Only} shows proportons of functional categories that
-included differentially expressed genes between glucose-grown and
-galactose-grown UA159 wild-type cells. Qualitatively, energy metabolism and PTS
-classes were profound. This was silmilar to the comparison between UA159
-wild-type and TW1 mutant-type grown in glucose. More than quarter of the
-differentially expressed genes were unknown, unassigned, or hypothetical.
-Consecutive genes (SMU.1488 to SMU.1496) were upregulated in 
-catabolite-derepressing conditions or in galactose-grown cells
-(\fixme{Are these the genes of the lactose (lac) operon?})
-(table \ref{tabUA159Only}).  Another set of genes (SMU.886 to SMU.889) were also
-upregulated (\fixme{Are these called galKTE?}).  
-Genes from SMU.1535 to SMU.1538 were upregulated. 
-Of the \Sexpr{nrow(tableUA159Only)} differentially expressed genes at
-$1 \times 10^{-10}$ cutoff of multiple testing adjusted p-value, 
-\Sexpr{length(grep("hypothetical",tableUA159Only[,"Gene Description"]))} were
-hypothetical, and most of them were down-regulated. 
-A few of the down-regulated genes were annotated.
-
-We also found sugar metabolism and PTS signal transduction to be profound in
-differential expression in the comparison of glucose-grown and galactose-grown
-TW1 mutant-type cells figure \ref{figPieTW1Only}. Many of the genes were also
-hypothetical. We found similar genes differentially expressed in TW1 mutant-type
-cells to those in UA159 wild-type (table \ref{tabTW1Only}).
-
-\fixme{We may need dot plots where x-axis represents fold changes, and y-axis
-functional classes. Dots may be colored by p-values. This would show most of
-information in one plot. We could label some of dots to emphasixe. Done?}
-\fixme{Use KEGG for functional category study. Done?}
-\fixme{Display results in UCSC genome browser. Done?}
-\fixme{Use two factor analysis. Done?}
-
-%
-
 Deletion of \textit{ccpA} resulted in substantial relief of the CCR of
@@
 glucose-only condition we found xx number of differentially expressed genes.
-
-\section{Discussion}
-We used RNA-seq data to delineate transcript boundaries. Although this map of
-transcripts does not precisely tell us where transcripts start and end, it
-informs us of which open reading frames can be transcribed together. Potentially
-polycistronic operons are defined. We focused on interpretations of operon in
-light of CcpA regulations.
-
-\section{Acknowledgments}
-This study was funded by the National Institute of Allergy and Infectious
-Disease, US National Institutes of Health, under grant number AI073368-01A2
-(to M.J.S.\ and A.S).  Additional support was provided by National Science
-Foundation CAREER Award DBI-0644111 and a David and Lucile Packard
-Fellowship for Science and Engineering (to A.S.).  
-We thank Christopher Browngardt for assistance with labelling functional classes
-of \textit{S.\ mutans} genes.
-
-\renewcommand*{\refname}{Literature Cited}
-\bibliographystyle{../../latex/bst/mbe}
-\bibliography{siepel-mutans}
-\clearpage{}
-
-\section{Tables}
-
-%-----------------------------------------------------------------------------
-\begin{table}[!h]
-\caption{{\bf Gene Ontology enrichments for target genes of putative small RNAs.}}
-\vspace{1ex}
-\noindent \begin{centering}
-\begin{tabular}{lcccl}
-\hline 
-small RNA & $q^a$ & count & GO & Description \\
-\hline 
-%-----------------------------------------------------------------------------
-<<echo=FALSE,results=tex>>=
-smutans.mannwhitney( qval=0.05,
-                     genes.criteria=smutans.genes.criteria,
-                     go.genes=smutans.go.genes,
-                     cat.desc=smutans.cat.desc )
-@
-%-----------------------------------------------------------------------------
-\hline 
-\end{tabular}
-\par\end{centering}
-\begin{flushleft}
-$^a$False discovery rate estimated by the
-Benjamini-Hochberg method.  All categories having at least ten genes and
-$q\leq 0.05$ are displayed.\\ 
-\end{flushleft}
-\label{tab:goseq}
-\end{table}
-%-----------------------------------------------------------------------------
-
-\clearpage{}
-
-\section{Figures}
-
-<<figLength,fig=TRUE,echo=FALSE>>=
-l <- list(expressed=width(smutansData.txGenes[matchMatrix(mtch2)[,1]]), 
-          notexpressed=width(smutansData.txGenes[-matchMatrix(mtch2)[,1]]))
-bin <- c(as.integer(seq.int(0, 3000, length.out=15)), 8500)
-multhist(l,breaks=bin)
-@
-\begin{figure}
-\centering
-\includegraphics[width=.9\textwidth]{manuscript-figLength}
-\caption{{\bf Length distribution for expressed and not-expressed genes.}
-Darker gray
-bars represent frequency of expressed genes, and lighter gray ones that of
-not-expressed genes. The last bin sums from 3000 bp to 8500 bp.
-}
-\label{figLength}
-\end{figure}
-
-<<figTranscriptLength,fig=TRUE,echo=FALSE>>=
-l <- list(coding=width(grTxWGenes), 
-          noncoding=width(grTxWOGenes))
-bin <- c(as.integer(seq.int(0, 10000, length.out=15)), 30000)
-multhist(l,breaks=bin)
-@
-\begin{figure}
-\centering
-\includegraphics[width=.9\textwidth]{manuscript-figTranscriptLength}
-\caption{{\bf Length distribution for expressed transcripts with annotated genes
-and unannotated genes.}
-Darker gray bars represent frequency of expressed transcripts with annotated
-genes, and lighter gray ones that of without annotated genes that can be
-potential non-coding regions. The last bin sums from 10000 bp to 30000 bp.
-}
-\label{figTranscriptLength}
-\end{figure}
-
-<<figUTRLength,fig=TRUE,echo=FALSE>>=
-l <- list(five=length5UTR, three=length3UTR)
-bin <- c(as.integer(seq.int(0, 300, length.out=19)), 6500)
-multhist(l,breaks=bin)
-@
-\begin{figure}
-\centering
-\includegraphics[width=.9\textwidth]{manuscript-figUTRLength}
-\caption{{\bf Length distribution for 5' and 3' UTR.}
-Darker gray bars represent frequency of 5' UTR lengths, and lighter gray 3' UTR
-lengths. The last bin sums from 300 bp to 6500 bp.
-\fixme{Check the too long UTRs because they do not make sense.}
-}
-\label{figUTRLength}
-\end{figure}
-
-<<figExpressionTwoGene,fig=TRUE,echo=FALSE>>=
-plot( log10(comp.exp.two.genes[1,]), log10(comp.exp.two.genes[2,]),
-      xlab="Expression of first Gene in log10",
-      ylab="Expression of second Gene in log10",
-      pch=19,cex=0.5)
-abline(b=1,lty=2)
-@
-<<figExpressionTwoGeneBetweenTranscript,fig=TRUE,echo=FALSE>>=
-plot( log10(comp.exp.two.genes.between.transcript[1,]),
-      log10(comp.exp.two.genes.between.transcript[2,]),
-      xlab="Expression of first Gene in log10",
-      ylab="Expression of second Gene in log10",
-      pch=19,cex=0.5)
-abline(b=1,lty=2)
-@
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{manuscript-figExpressionTwoGene}
-\includegraphics[width=.6\textwidth]{manuscript-figExpressionTwoGeneBetweenTranscript}
-\caption{{\bf Scatter plot of expression levels of pairs of adjacent genes
-located in the same transcripts.}
-(Top) Two genes in the same tranascripts are plotted using their expression levels in
-log10 scale.
-(Bottom) Two genes in different tranascripts are plotted using their expression levels in
-log10 scale.
-}
-\label{figExpressionTwoGene}
-\end{figure}
-
-\end{document}

### 2012-02-02 package/supplementary.Rnw
Just an update.

@@
 library(GenomicRanges)
+library(IRanges)
 library(plotrix)
@@
 load("smutans/data/smutansData.txGenes.RData")
+load("smutans/data/smutansData.txGrCombined.RData")
+load("smutans/data/smutansData.txAnnotation.RData")
+load("smutans/data/smutansData.txRNAz.RData")
 load("smutans/data/smutansData.txPileup.RData")
@@
 load(smutansData.txGenes)
+load(smutansData.txCombined)
+load(smutansData.txAnnotation)
+load(smutansData.txRNAz)
 load(smutansData.txPileup)
@@
 \section{Supplementary Text}
-A
-\subsection{Summary statistics for recombinant trees}
-\citep{Gorke2008}
+Add texts if needed.
+\subsection{Subsection name}
+Add texts if needed and citations such as \citep{Gorke2008}.
 
@@
 
-%\section{Supplementary Tables}
+\clearpage{}
+\section{Supplementary Tables}
+
+%\clearpage{}
+<<echo=FALSE,results=tex>>=
+ir1 <- gaps(ranges(smutansData.txAnnotation))
+ir2 <- ranges(smutansData.txRNAz)
+mtchUnannotatedRNA <- findOverlaps(ir2,ir1,type="within")
+txRNAz.unannotated <- ir2[matchMatrix(mtchUnannotatedRNA)[,1]]
+t1 <- data.frame(names=names(txRNAz.unannotated),start=start(txRNAz.unannotated),end=end(txRNAz.unannotated))
+x.big <- xtable( t1,
+                 display=c("s","s","d","d"),
+                 align=c('l','l','r','r'),
+                 label='tabUnannotatedRNA',
+                 caption='{\\bf List of unannotated RNAs predicted using RNAz.}'
+               )
+print( x.big,
+       caption.placement="top",
+       include.rownames=FALSE)
+@
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology enrichments for target genes of putative small RNAs.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{lcccl}
+\hline 
+small RNA & $q^a$ & count & GO & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.mannwhitney( qval=0.05,
+                     genes.criteria=smutans.genes.criteria,
+                     go.genes=smutans.go.genes,
+                     cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseq}
+\end{table}
+%-----------------------------------------------------------------------------
 
 <<echo=FALSE,results=hide>>=
+trim <- function (x) gsub("^\\s+|\\s+$", "", x)
+
 geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
@@
 
+\clearpage{}
 <<echo=FALSE,results=tex>>=
@@
                  label='tabGlucoseOnly',
-                 caption='List of genes for the comparison UA159 and TW1 under glucose-grown cells',
+                 caption='{\\bf List of genes for the comparison UA159 and TW1 under glucose-grown cells.}',
                )
@@
 x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
+tableGlucoseOnly <- x
 @
-\clearpage{}
-
-<<figPieGlucoseOnly,fig=TRUE,echo=FALSE,width=6,height=8>>=
-par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
-#par(mfrow=c(2,1))
-trim <- function (x) gsub("^\\s+|\\s+$", "", x)
-y <- x[,"Functional Class"]
-z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
-z <- table(as.factor(sort(unlist(z))))
-
-slices <- as.vector(z)
-lbls <- names(z)
-pct <- round(slices/sum(slices)*100)
-c1 <- brewer.pal(n=9,name="Set1")
-pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
-plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
-legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
-@
-\begin{figure}
-\centering
-\includegraphics[width=0.9\textwidth]{supplementary-figPieGlucoseOnly}
-\caption{{\bf Distribution of functional classes of differntially expressed genes
-in comparing UA159 wild-type and TW1 mutant-type grown in glucose.}
-Labels for functional categories are colored with nine different colors. The
-first label with red color, 
-``Biosynthesis of cofactors, prosthetic groups, and carriers; Thiamine'', starts
-at the top, and next labels read to the clockwise. The numbers circling the pie
-chart is in per cent.
-}
-\label{figPieGlucoseOnly}
-\end{figure}
-
-\clearpage{}
-<<echo=FALSE,results=tex>>=
-#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
-#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
-#                              condA="ua159", condB="tw1" )
-#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
-load("galactoseOnly.RData")
-x <- smutans.de2List( galactoseOnly, qval=1e-3 )
-x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
-colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
-x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
-x.big <- xtable( x[order(x[,"p-value"]),],
-                 display=c("s","s","d","d","f","e","s"),
-                 digits=c(0,0,0,0,1,0,0),
-                 align=c('l','l','r','r','r','r','p{3.5in}'),
-                 label='tabGalactoseOnly',
-                 caption='List of genes for the comparison UA159 and TW1 under galactose-grown cells',
-               )
-print( x.big,
-       tabular.environment='longtable',
-       caption.placement="top",
-       include.rownames=FALSE,
-       floating=FALSE) 
-x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
-@
-\clearpage{}
-
-<<figPieGalactoseOnly,fig=TRUE,echo=FALSE,width=6,height=8>>=
-par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
-#par(mfrow=c(2,1))
-trim <- function (x) gsub("^\\s+|\\s+$", "", x)
-y <- x[,"Functional Class"]
-z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
-z <- table(as.factor(sort(unlist(z))))
-
-slices <- as.vector(z)
-lbls <- names(z)
-pct <- round(slices/sum(slices)*100)
-c1 <- brewer.pal(n=9,name="Set1")
-pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
-plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
-legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
-@
-\begin{figure}
-\centering
-\includegraphics[width=0.9\textwidth]{supplementary-figPieGalactoseOnly}
-\caption{{\bf Distribution of functional classes of differntially expressed genes
-in comparing UA159 wild-type and TW1 mutant-type grown in galactose.}
-Labels for functional categories are colored with nine different colors. The
-first label with red color, 
-``Amino acid biosynthesis; Glutamate family'', starts
-at the top, and next labels read to the clockwise. The numbers circling the pie
-chart is in per cent.
-}
-\label{figPieGalactoseOnly}
-\end{figure}
-
-\clearpage{}
-<<echo=FALSE,results=tex>>=
-#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
-#ua159only <- smutans.de2( ua159only, type="ua159", 
-#                          condA="glucose", condB="galactose" )
-#save(ua159only, file=file.path("ua159only.RData"))
-load("ua159only.RData")
-x <- smutans.de2List( ua159only, qval=1e-10 )
-x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
-colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
-x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
-x.big <- xtable( x[order(x[,"p-value"]),],
-                 display=c("s","s","d","d","f","e","s"),
-                 digits=c(0,0,0,0,1,0,0),
-                 align=c('l','l','r','r','r','r','p{3.5in}'),
-                 label='tabUA159Only',
-                 caption='List of genes for the comparison of UA159 wild-type grown with glucose and galactose. Note that we used 1e-10 as the p-value cutoff.',
-               )
-print( x.big,
-       tabular.environment='longtable',
-       caption.placement="top",
-       include.rownames=FALSE,
-       floating=FALSE) 
-x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
-@
-\clearpage{}
-
-<<figPieUA159Only,fig=TRUE,echo=FALSE,width=6,height=8>>=
-par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
-trim <- function (x) gsub("^\\s+|\\s+$", "", x)
-y <- x[,"Functional Class"]
-z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
-z <- table(as.factor(sort(unlist(z))))
-
-slices <- as.vector(z)
-lbls <- names(z)
-pct <- round(slices/sum(slices)*100)
-c1 <- brewer.pal(n=9,name="Set1")
-pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
-plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
-
-legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
-#legend(-1.1,1.1,names(z[1:27]),bty='n',cex=0.7,fill=c1)
-#legend(0,0.4,names(z[-c(1:27)]),bty='n',cex=0.7,fill=c1)
-
-@
-\begin{figure}
-\centering
-\includegraphics[width=0.9\textwidth]{supplementary-figPieUA159Only}
-\caption{{\bf Distribution of functional classes of differntially expressed genes
-in comparing UA159 wild-type grown in glucose and galactose.}
-Labels for functional categories are colored with nine different colors. The
-first label with red color, 
-``Amino acid biosynthesis; Glutamate family'', starts
-at the top, and next labels read to the clockwise. The numbers circling the pie
-chart is in per cent.
-}
-\label{figPieUA159Only}
-\end{figure}
-
-\clearpage{}
-<<echo=FALSE,results=tex>>=
-#tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
-#tw1only <- smutans.de2( tw1only, type="tw1", 
-#                          condA="glucose", condB="galactose" )
-#save(tw1only, file=file.path("tw1only.RData"))
-load("tw1only.RData")
-x <- smutans.de2List( tw1only, qval=1e-3 )
-x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
-colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
-x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
-x.big <- xtable( x[order(x[,"p-value"]),],
-                 display=c("s","s","d","d","f","e","s"),
-                 digits=c(0,0,0,0,1,0,0),
-                 align=c('l','l','r','r','r','r','p{3.5in}'),
-                 label='tabTW1Only',
-                 caption='List of genes for the comparison of TW1 mutant-type grown with glucose and galactose ',
-               )
-print( x.big,
-       tabular.environment='longtable',
-       caption.placement="top",
-       include.rownames=FALSE,
-       floating=FALSE) 
-x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
-@
-\clearpage{}
-
-<<figPieTW1Only,fig=TRUE,echo=FALSE,width=6,height=8>>=
-par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
-trim <- function (x) gsub("^\\s+|\\s+$", "", x)
-y <- x[,"Functional Class"]
-z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
-z <- table(as.factor(sort(unlist(z))))
-
-slices <- as.vector(z)
-lbls <- names(z)
-pct <- round(slices/sum(slices)*100)
-c1 <- brewer.pal(n=9,name="Set1")
-pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
-plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
-legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
-@
-\begin{figure}
-\centering
-\includegraphics[width=0.9\textwidth]{supplementary-figPieTW1Only}
-\caption{{\bf Distribution of functional classes of differntially expressed genes
-in comparing TW1 mutant-type grown in glucose and galactose.}
-Labels for functional categories are colored with nine different colors. The
-first label with red color, 
-``Amino acid biosynthesis; Histidine family'', starts
-at the top, and next labels read to the clockwise. The numbers circling the pie
-chart is in per cent.
-}
-\label{figPieTW1Only}
-\end{figure}
 
@@
 \begin{flushleft}
-$^a$False discovery rate estimated by the
-Benjamini-Hochberg method.  All categories having at least ten genes and
-$q\leq 0.05$ are displayed.\\ 
+$^a$ We used a multiple-testing adjusted $p$-value of $1\times 10^{-3}$ to
+determine differentially expressed genes.  For each category having at least ten
+genes a variation of Fisher's exact test was performed, and another
+multiple-testing adjusted $p$-value, or $q$-value was obtained. The listed
+categories were chosen at the cutoff value of 0.05.
+\\ 
 \end{flushleft}
@@
 \begin{table}[!h]
-\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+\caption{{\bf KEGG enrichments for differentially expressed
+genes in the comparison UA159 and TW1 grown in glucose.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.de2Goseq ( glucoseOnly, qval=1e-3, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.keg.genes[c(4,3)],
+                   cat.desc=smutans.keg.desc)
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:kegseqGlucoseOnly}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+<<echo=FALSE,results=tex>>=
+#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
+#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
+#                              condA="ua159", condB="tw1" )
+#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
+load("galactoseOnly.RData")
+x <- smutans.de2List( galactoseOnly, qval=1e-3 )
+x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
+colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
+x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
+x.big <- xtable( x[order(x[,"p-value"]),],
+                 display=c("s","s","d","d","f","e","s"),
+                 digits=c(0,0,0,0,1,0,0),
+                 align=c('l','l','r','r','r','r','p{3.5in}'),
+                 label='tabGalactoseOnly',
+                 caption='{\\bf List of genes for the comparison UA159 and TW1 under galactose-grown cells}',
+               )
+print( x.big,
+       tabular.environment='longtable',
+       caption.placement="top",
+       include.rownames=FALSE,
+       floating=FALSE) 
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
+tableGalactoseOnly <- x
+@
+
+%\clearpage{}
+%%-----------------------------------------------------------------------------
+%\begin{table}[!h]
+%\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+%genes in the comparison UA159 and TW1 grown in galactose.}}
+%\vspace{1ex}
+%\noindent \begin{centering}
+%\begin{tabular}{cl}
+%\hline 
+%$q^a$ & Description \\
+%\hline 
+%%-----------------------------------------------------------------------------
+%<<echo=FALSE,results=tex>>=
+%smutans.de2Goseq ( galactoseOnly, qval=1e-3, 
+%                   feature.gene=smutans.feature.genes,
+%                   go.genes=smutans.go.genes,
+%                   cat.desc=smutans.cat.desc )
+%@
+%%-----------------------------------------------------------------------------
+%\hline 
+%\end{tabular}
+%\par\end{centering}
+%\begin{flushleft}
+%$^a$False discovery rate estimated by the
+%Benjamini-Hochberg method.  All categories having at least ten genes and
+%$q\leq 0.05$ are displayed.\\ 
+%\end{flushleft}
+%\label{tab:goseqGalactoseOnly}
+%\end{table}
+%%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf KEGG enrichments for differentially expressed
 genes in the comparison UA159 and TW1 grown in galactose.}}
@@
                    feature.gene=smutans.feature.genes,
-                   go.genes=smutans.go.genes,
-                   cat.desc=smutans.cat.desc )
+                   go.genes=smutans.keg.genes[c(4,3)],
+                   cat.desc=smutans.keg.desc)
 @
@@
 \end{flushleft}
-\label{tab:goseqGalactoseOnly}
+\label{tab:kegseqGalactoseOnly}
 \end{table}
 %-----------------------------------------------------------------------------
+
+\clearpage{}
+<<echo=FALSE,results=tex>>=
+#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
+#ua159only <- smutans.de2( ua159only, type="ua159", 
+#                          condA="glucose", condB="galactose" )
+#save(ua159only, file=file.path("ua159only.RData"))
+load("ua159only.RData")
+x <- smutans.de2List( ua159only, qval=1e-10 )
+x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
+colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
+x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
+x.big <- xtable( x[order(x[,"p-value"]),],
+                 display=c("s","s","d","d","f","e","s"),
+                 digits=c(0,0,0,0,1,0,0),
+                 align=c('l','l','r','r','r','r','p{3.5in}'),
+                 label='tabUA159Only',
+                 caption='{\\bf List of genes for the comparison of UA159 wild-type grown with glucose and galactose.} Note that we used 1e-10 as the p-value cutoff.',
+               )
+print( x.big,
+       tabular.environment='longtable',
+       caption.placement="top",
+       include.rownames=FALSE,
+       floating=FALSE) 
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
+tableUA159Only <- x
+@
 
@@
 \begin{table}[!h]
-\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
-genes in the comparison of glucose-grown and galactose-grown TW1 mutant-type
-cells.}}
-\vspace{1ex}
-\noindent \begin{centering}
-\begin{tabular}{cl}
-\hline 
-$q^a$ & Description \\
-\hline 
-%-----------------------------------------------------------------------------
-<<echo=FALSE,results=tex>>=
-smutans.de2Goseq ( tw1only, qval=1e-3, 
-                   feature.gene=smutans.feature.genes,
-                   go.genes=smutans.go.genes,
-                   cat.desc=smutans.cat.desc )
-@
-%-----------------------------------------------------------------------------
-\hline 
-\end{tabular}
-\par\end{centering}
-\begin{flushleft}
-$^a$False discovery rate estimated by the
-Benjamini-Hochberg method.  All categories having at least ten genes and
-$q\leq 0.05$ are displayed.\\ 
-\end{flushleft}
-\label{tab:goseqTW1Only}
-\end{table}
-%-----------------------------------------------------------------------------
-
-\clearpage{}
-%-----------------------------------------------------------------------------
-\begin{table}[!h]
-\caption{{\bf KEGG enrichments for differentially expressed
-genes in the comparison UA159 and TW1 grown in glucose.}}
-\vspace{1ex}
-\noindent \begin{centering}
-\begin{tabular}{cl}
-\hline 
-$q^a$ & Description \\
-\hline 
-%-----------------------------------------------------------------------------
-<<echo=FALSE,results=tex>>=
-smutans.de2Goseq ( glucoseOnly, qval=1e-3, 
-                   feature.gene=smutans.feature.genes,
-                   go.genes=smutans.keg.genes[c(4,3)],
-                   cat.desc=smutans.keg.desc)
-@
-%-----------------------------------------------------------------------------
-\hline 
-\end{tabular}
-\par\end{centering}
-\begin{flushleft}
-$^a$False discovery rate estimated by the
-Benjamini-Hochberg method.  All categories having at least ten genes and
-$q\leq 0.05$ are displayed.\\ 
-\end{flushleft}
-\label{tab:kegseqGlucoseOnly}
-\end{table}
-%-----------------------------------------------------------------------------
-
-\clearpage{}
-%-----------------------------------------------------------------------------
-\begin{table}[!h]
-\caption{{\bf KEGG enrichments for differentially expressed
-genes in the comparison UA159 and TW1 grown in galactose.}}
-\vspace{1ex}
-\noindent \begin{centering}
-\begin{tabular}{cl}
-\hline 
-$q^a$ & Description \\
-\hline 
-%-----------------------------------------------------------------------------
-<<echo=FALSE,results=tex>>=
-smutans.de2Goseq ( galactoseOnly, qval=1e-3, 
-                   feature.gene=smutans.feature.genes,
-                   go.genes=smutans.keg.genes[c(4,3)],
-                   cat.desc=smutans.keg.desc)
-@
-%-----------------------------------------------------------------------------
-\hline 
-\end{tabular}
-\par\end{centering}
-\begin{flushleft}
-$^a$False discovery rate estimated by the
-Benjamini-Hochberg method.  All categories having at least ten genes and
-$q\leq 0.05$ are displayed.\\ 
-\end{flushleft}
-\label{tab:kegseqGalactoseOnly}
-\end{table}
-%-----------------------------------------------------------------------------
-
-\clearpage{}
-%-----------------------------------------------------------------------------
-\begin{table}[!h]
 \caption{{\bf KEGG enrichments for differentially expressed
@@
 
+\clearpage{}
+<<echo=FALSE,results=tex>>=
+#tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
+#tw1only <- smutans.de2( tw1only, type="tw1", 
+#                          condA="glucose", condB="galactose" )
+#save(tw1only, file=file.path("tw1only.RData"))
+load("tw1only.RData")
+x <- smutans.de2List( tw1only, qval=1e-3 )
+x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
+colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
+x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
+x.big <- xtable( x[order(x[,"p-value"]),],
+                 display=c("s","s","d","d","f","e","s"),
+                 digits=c(0,0,0,0,1,0,0),
+                 align=c('l','l','r','r','r','r','p{3.5in}'),
+                 label='tabTW1Only',
+                 caption='{\\bf List of genes for the comparison of TW1 mutant-type grown with glucose and galactose.}',
+               )
+print( x.big,
+       tabular.environment='longtable',
+       caption.placement="top",
+       include.rownames=FALSE,
+       floating=FALSE) 
+x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
+tableTW1Only <- x
+@
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+genes in the comparison of glucose-grown and galactose-grown TW1 mutant-type
+cells.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<echo=FALSE,results=tex>>=
+smutans.de2Goseq ( tw1only, qval=1e-3, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseqTW1Only}
+\end{table}
+%-----------------------------------------------------------------------------
 \clearpage{}
@@
 % END OF TABLE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%\section{Supplementary Figures}
+\clearpage{}
+\section{Supplementary Figures}
 
+<<figClust,echo=FALSE,result=hide,fig=TRUE>>=
+allSample <- newSmutans( smutansGenes )
+smutans.de2Clust( allSample )
+@
 \begin{figure}
@@
 \includegraphics[width=.9\textwidth]{supplementary-figClust}
-\caption{Heatmap showing the Euclidean distances between the samples
+\caption{{\bf Cluster of 13 RNA-seq samples.} 
+Heatmap shows the Euclidean distances between the samples
 as calculated from the variance-stabilizing transformation of the
@@
 \end{figure}
-<<figClust,echo=FALSE,result=hide,fig=TRUE>>=
-allSample <- newSmutans( smutansGenes )
-smutans.de2Clust( allSample )
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Transcript analyses R scripts
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+<<figGrCombinedExpression,fig=TRUE,echo=FALSE>>=
+l <- list(elementMetadata(smutansData.txGrCombined)$pileup)
+bin <- c(as.integer(seq.int(0, 1000, length.out=15)), 50000)
+multhist(l,breaks=bin)
+@
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{supplementary-figGrCombinedExpression}
+\caption{{\bf Average pileup values for segments.}
+The last bin sums from 1000 bp to 5000 bp.
+}
+\label{figGrCombinedExpression}
+\end{figure}
+
+<<echo=FALSE,result=hide>>=
+# library(GenomicRanges)
+# library(IRanges)
+
+notExpressedTx <- smutansData.txGrCombined[elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5]
+grTx <- smutansData.txGrCombined[!(elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5)]
+
+matchGenesTx <- findOverlaps(smutansData.txGenes,grTx)
+matchAnnotationTx <- findOverlaps(smutansData.txAnnotation,grTx)
+
+grNoAnnotation <- gaps(ranges(smutansData.txAnnotation))
+grNoncoding <- gaps(ranges(smutansData.txGenes))
+grRNAz <- ranges(smutansData.txRNAz)
+
+matchUnannotatedRNA <- findOverlaps(grRNAz,grNoAnnotation,type="within")
+txRNAz.unannotated <- grRNAz[matchMatrix(matchUnannotatedRNA)[,1]]
+
+# A bug in GenomicRanges: the last segment is the while genome. I removed it
+# from the noncoding ranges.
+# matchNoncoding <- findOverlaps(grNoncoding,grTx)
+
+# Expressed transcripts without annotated genes
+grTxWGenes <- grTx[unique(matchMatrix(matchGenesTx)[,2])]
+grTxWOGenes <- grTx[-unique(matchMatrix(matchGenesTx)[,2])]
+grTxWOAnnotation <- grTx[-unique(matchMatrix(matchAnnotationTx)[,2])]
+
+# Find transcripts with genes with conflicted strands
+grGenesPlus <- smutansData.txGenes[strand(smutansData.txGenes)=='+']
+grGenesMnus <- smutansData.txGenes[strand(smutansData.txGenes)=='-']
+#mtchPlus <- findOverlaps(grGenesPlus,grTxWGenes)
+#mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
+#txWithConflictedGenes <- intersect(matchMatrix(mtchPlus)[,2],matchMatrix(mtchMnus)[,2])
+
+# Tx with genes strands
+grTx2 <- grTxWGenes[!strand(grTxWGenes)=='*']
+
+# Find the number of genes in segments.
+matchGenesTx2 <- findOverlaps(smutansData.txGenes,grTx2,type="within")
+number.genes.tx <- runLength(Rle(matchMatrix(matchGenesTx2)[,"subject"]))
+
+# Length distribution of UTR regions
+# grTx2 <- grTxWGenes[-txWithConflictedGenes] 
+# Lengths of 5' and 3' UTRs
+mtchPlus <- findOverlaps(grGenesPlus,grTx2,type="within")
+mtchMnus <- findOverlaps(grGenesMnus,grTx2,type="within")
+length5UTR <- c()
+length3UTR <- c()
+for (i in unique(matchMatrix(mtchPlus)[,"subject"])) {
+  x <- matchMatrix(mtchPlus)[matchMatrix(mtchPlus)[,"subject"]==i,"query"]
+  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
+  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
+  length5UTR <- c(length5UTR,y)
+  length3UTR <- c(length3UTR,z)
+}
+for (i in unique(matchMatrix(mtchMnus)[,"subject"])) {
+  x <- matchMatrix(mtchMnus)[matchMatrix(mtchMnus)[,"subject"]==i,"query"]
+  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
+  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
+  length5UTR <- c(length5UTR,z)
+  length3UTR <- c(length3UTR,y)
+}
+
+# Correlation of expression levels of adjacent genes.
+pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
+pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
+elementMetadata(smutansData.txGenes)["score"] <- pileup.y
+mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
+x <- rle(matchMatrix(mtch3)[,"subject"])
+comp.exp.two.genes <- c()
+for (i in x$values[x$lengths > 1]) {
+  y <- combn(matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==i],2)
+  for (j in 1:dim(y)[2]) {
+    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
+  }
+}
+# Correlation of expression levels of adjacent genes between transcripts.
+comp.exp.two.genes.between.transcript <- c()
+for (i in 1:length(x$values)) {
+  if (i == length(x$values)) {
+    break
+  }
+  y1 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i]]
+  i2 <- i + 1
+  y2 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i2]]
+  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
+  for (j in 1:dim(l)[1]) {
+    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
+  }
+}
 @
 
+<<figLength,fig=TRUE,echo=FALSE>>=
+l <- list(expressed=width(smutansData.txGenes[matchMatrix(matchGenesTx)[,1]]), 
+          notexpressed=width(smutansData.txGenes[-matchMatrix(matchGenesTx)[,1]]))
+bin <- c(as.integer(seq.int(0, 3000, length.out=15)), 8500)
+multhist(l,breaks=bin)
+@
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{supplementary-figLength}
+\caption{{\bf Length distribution for expressed and not-expressed genes.}
+Darker gray
+bars represent frequency of expressed genes, and lighter gray ones that of
+not-expressed genes. The last bin sums from 3000 bp to 8500 bp.
+}
+\label{figLength}
+\end{figure}
+
+<<figTranscriptLength,fig=TRUE,echo=FALSE>>=
+l <- list(coding=width(grTxWGenes), 
+          noncoding=width(grTxWOGenes))
+bin <- c(as.integer(seq.int(0, 10000, length.out=15)), 30000)
+multhist(l,breaks=bin)
+@
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{supplementary-figTranscriptLength}
+\caption{{\bf Length distribution for expressed segments with annotated genes
+and unannotated genes.}
+Darker gray bars represent frequency of expressed segments with annotated
+genes, and lighter gray ones that of without annotated genes that can be
+potential non-coding regions. The last bin sums from 10000 bp to 30000 bp.
+}
+\label{figTranscriptLength}
+\end{figure}
+
+<<figUTRLength,fig=TRUE,echo=FALSE>>=
+l <- list(five=length5UTR, three=length3UTR)
+bin <- c(as.integer(seq.int(0, 300, length.out=19)), 6500)
+multhist(l,breaks=bin)
+@
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{supplementary-figUTRLength}
+\caption{{\bf Length distribution for 5' and 3' UTR.}
+Darker gray bars represent frequency of 5' UTR lengths, and lighter gray 3' UTR
+lengths. The last bin sums from 300 bp to 6500 bp.
+\fixme{Check the too long UTRs because they do not make sense.}
+}
+\label{figUTRLength}
+\end{figure}
+
+<<figExpressionTwoGene,fig=TRUE,echo=FALSE>>=
+plot( log10(comp.exp.two.genes[1,]), log10(comp.exp.two.genes[2,]),
+      xlab="Expression of first Gene in log10",
+      ylab="Expression of second Gene in log10",
+      pch=19,cex=0.5)
+abline(b=1,lty=2)
+@
+<<figExpressionTwoGeneBetweenTranscript,fig=TRUE,echo=FALSE>>=
+plot( log10(comp.exp.two.genes.between.transcript[1,]),
+      log10(comp.exp.two.genes.between.transcript[2,]),
+      xlab="Expression of first Gene in log10",
+      ylab="Expression of second Gene in log10",
+      pch=19,cex=0.5)
+abline(b=1,lty=2)
+@
+\begin{figure}
+\centering
+\includegraphics[width=.6\textwidth]{supplementary-figExpressionTwoGene}
+\includegraphics[width=.6\textwidth]{supplementary-figExpressionTwoGeneBetweenTranscript}
+\caption{{\bf Scatter plot of expression levels of pairs of adjacent genes
+located in the same transcripts.}
+(Top) Two genes in the same tranascripts are plotted using their expression levels in
+log10 scale.
+(Bottom) Two genes in different tranascripts are plotted using their expression levels in
+log10 scale.
+}
+\label{figExpressionTwoGene}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus101}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus7}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus68}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus8}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus10}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus69}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus1}
+\caption{{\bf MFE structures drawing encoding positional entropy for
+differentially expressed regions.}
+Predicted non-coding RNAs are shown from top-left to bottom-right: locus101,
+locus7, locus68, locus8, locus10, locus69, and locus1.
+}
+\label{figRNALocus1}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus99}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus103}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus105}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus106}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus107}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus108}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus110}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus2}
+\caption{{\bf MFE structures drawing encoding positional entropy with expressed
+regions.}
+Predicted non-coding RNAs are shown from top-left to bottom-right: 
+locus99, locus103, locus105, locus106, locus107, locus108, locus110, and locus2.
+}
+\label{figRNALocus2}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus5}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus6}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus7}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus8}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus52}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus68}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus69}
+\includegraphics[width=0.3\textwidth]{figures/supplementary-figRNALocus85}
+\caption{{\bf MFE structures drawing encoding positional entropy with expressed
+regions.}
+Predicted non-coding RNAs are shown from top-left to bottom-right: 
+locus5, locus6, locus7, locus8, locus52, locus68, locus69, and locus85.
+}
+\label{figRNALocus3}
+\end{figure}
+
+\clearpage{}
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/ucsc}
+\caption{{\bf UCSC genome browser tracks of RNA-seq data.}
+Tracks are shown from the top: known genes, pileup value tracks for UA159 grown
+in glucose, those for TW1 in glucose, and the five segments predicted based on
+RNA-seq data and known genes. Known genes with forward strand are colored red,
+and those of reverse strand blue. The most top of the five predicted segments
+are also colored in the same way. 
+}
+\label{figUCSC}
+\end{figure}
+
+\clearpage{}
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/ucsc2}
+\caption{{\bf UCSC genome browser tracks of RNA-seq data for \textit{galKTE} genes.}
+Four genes from SMU.886 to SMU.889 were also upregulated in respose to
+galactose. RNA-seq1 is UA159 with glucose, RNA-seq2 TW1 with glucose,
+RNA-seq21 UA159 with galactose, and RNA-seq3 TW1 with galactose.
+}
+\label{figUCSC2}
+\end{figure}
+
+\clearpage{}
+<<figPieGlucoseOnly,fig=TRUE,echo=FALSE,width=6,height=8>>=
+par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
+y <- tableGlucoseOnly[,"Functional Class"]
+z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+z <- table(as.factor(sort(unlist(z))))
+
+slices <- as.vector(z)
+lbls <- names(z)
+pct <- round(slices/sum(slices)*100)
+c1 <- brewer.pal(n=9,name="Set1")
+pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
+plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
+legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
+@
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figPieGlucoseOnly}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing UA159 wild-type and TW1 mutant-type grown in glucose.}
+Labels for functional categories are colored with nine different colors. The
+first label with red color, 
+``Biosynthesis of cofactors, prosthetic groups, and carriers; Thiamine'', starts
+at the top, and next labels read to the clockwise. The numbers circling the pie
+chart is in per cent.
+}
+\label{figPieGlucoseOnly}
+\end{figure}
+
+<<figDotplotGlucoseOnly,fig=TRUE,echo=FALSE,width=12,height=6>>=
+#par(mar=c(4.1, 3.1, 0.1, 0.1))
+par(mai=c(0, 0, 0, 0))
+y <- tableGlucoseOnly[,"Functional Class"]
+# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+# z <- table(as.factor(sort(unlist(z))))
+z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
+xm <- tableGlucoseOnly[rep(seq(length(z)),sapply(z,length)),]
+xm[,"Functional Class"] <- unlist(z)
+# Use xm to plot a dotplot.
+dotplot(xm[,"Functional Class"]~xm[,"fold"],
+        main="",xlab="Log2 fold change",col="black",pch=88)
+@
+\begin{sidewaysfigure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figDotplotGlucoseOnly}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing UA159 wild-type and TW1 mutant-type grown in glucose.}
+}
+\label{figDotplotGlucoseOnly}
+\end{sidewaysfigure}
+
+<<figFoldPValueGlucoseOnly,fig=TRUE,echo=FALSE>>=
+par(mar=c(4.1, 4.1, 0.2, 0.2))
+plot(abs(xm[,"fold"])~log10(xm[,"p-value"]),
+     xlab="log10 p-value",ylab="Absolute log2 fold change",
+     pch=19,cex=0.7)
+@
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figFoldPValueGlucoseOnly}
+\caption{{\bf Scatter plot of fold versus p-vaule in log scale.}
+}
+\label{figFoldPValueGlucoseOnly}
+\end{figure}
+
+\clearpage{}
+<<figPieGalactoseOnly,fig=TRUE,echo=FALSE,width=6,height=8>>=
+par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
+y <- tableGalactoseOnly[,"Functional Class"]
+z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+z <- table(as.factor(sort(unlist(z))))
+
+slices <- as.vector(z)
+lbls <- names(z)
+pct <- round(slices/sum(slices)*100)
+c1 <- brewer.pal(n=9,name="Set1")
+pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
+plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
+legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
+@
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figPieGalactoseOnly}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing UA159 wild-type and TW1 mutant-type grown in galactose.}
+Labels for functional categories are colored with nine different colors. The
+first label with red color, 
+``Amino acid biosynthesis; Glutamate family'', starts
+at the top, and next labels read to the clockwise. The numbers circling the pie
+chart is in per cent.
+}
+\label{figPieGalactoseOnly}
+\end{figure}
+
+<<figDotplotGalactoseOnly,fig=TRUE,echo=FALSE,width=12,height=6>>=
+par(mai=c(0, 0, 0, 0))
+y <- tableGalactoseOnly[,"Functional Class"]
+# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+# z <- table(as.factor(sort(unlist(z))))
+z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
+xm <- tableGalactoseOnly[rep(seq(length(z)),sapply(z,length)),]
+xm[,"Functional Class"] <- unlist(z)
+# Use xm to plot a dotplot.
+dotplot(xm[,"Functional Class"]~xm[,"fold"],
+        main="",xlab="Log2 fold change",col="black",pch=88)
+@
+\begin{sidewaysfigure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figDotplotGalactoseOnly}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing UA159 wild-type and TW1 mutant-type grown in galactose.}
+}
+\label{figDotplotGalactoseOnly}
+\end{sidewaysfigure}
+
+\clearpage{}
+<<figPieUA159Only,fig=TRUE,echo=FALSE,width=6,height=8>>=
+par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
+trim <- function (x) gsub("^\\s+|\\s+$", "", x)
+y <- tableUA159Only[,"Functional Class"]
+z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+z <- table(as.factor(sort(unlist(z))))
+
+slices <- as.vector(z)
+lbls <- names(z)
+pct <- round(slices/sum(slices)*100)
+c1 <- brewer.pal(n=9,name="Set1")
+pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
+plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
+
+legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
+#legend(-1.1,1.1,names(z[1:27]),bty='n',cex=0.7,fill=c1)
+#legend(0,0.4,names(z[-c(1:27)]),bty='n',cex=0.7,fill=c1)
+
+@
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figPieUA159Only}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing UA159 wild-type grown in glucose and galactose.}
+Labels for functional categories are colored with nine different colors. The
+first label with red color, 
+``Amino acid biosynthesis; Glutamate family'', starts
+at the top, and next labels read to the clockwise. The numbers circling the pie
+chart is in per cent.
+}
+\label{figPieUA159Only}
+\end{figure}
+
+<<figDotplotUA159Only,fig=TRUE,echo=FALSE,width=12,height=6>>=
+#par(mar=c(4.1, 3.1, 0.1, 0.1))
+par(mai=c(0, 0, 0, 0))
+y <- tableUA159Only[,"Functional Class"]
+# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+# z <- table(as.factor(sort(unlist(z))))
+z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
+xm <- tableUA159Only[rep(seq(length(z)),sapply(z,length)),]
+xm[,"Functional Class"] <- unlist(z)
+# Use xm to plot a dotplot.
+dotplot(xm[,"Functional Class"]~xm[,"fold"],
+        main="",xlab="Log2 fold change",col="black",pch=88)
+@
+\begin{sidewaysfigure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figDotplotUA159Only}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing UA159 wild-type grown in glucose and galactose.}
+}
+\label{figDotplotUA159Only}
+\end{sidewaysfigure}
+
+\clearpage{}
+<<figPieTW1Only,fig=TRUE,echo=FALSE,width=6,height=8>>=
+par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
+trim <- function (x) gsub("^\\s+|\\s+$", "", x)
+y <- tableTW1Only[,"Functional Class"]
+z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+z <- table(as.factor(sort(unlist(z))))
+
+slices <- as.vector(z)
+lbls <- names(z)
+pct <- round(slices/sum(slices)*100)
+c1 <- brewer.pal(n=9,name="Set1")
+pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
+plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
+legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
+@
+\begin{figure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figPieTW1Only}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing TW1 mutant-type grown in glucose and galactose.}
+Labels for functional categories are colored with nine different colors. The
+first label with red color, 
+``Amino acid biosynthesis; Histidine family'', starts
+at the top, and next labels read to the clockwise. The numbers circling the pie
+chart is in per cent.
+}
+\label{figPieTW1Only}
+\end{figure}
+
+<<figDotplotTW1Only,fig=TRUE,echo=FALSE,width=12,height=6>>=
+#par(mar=c(4.1, 3.1, 0.1, 0.1))
+par(mai=c(0, 0, 0, 0))
+y <- tableTW1Only[,"Functional Class"]
+# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
+# z <- table(as.factor(sort(unlist(z))))
+z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
+xm <- tableTW1Only[rep(seq(length(z)),sapply(z,length)),]
+xm[,"Functional Class"] <- unlist(z)
+# Use xm to plot a dotplot.
+dotplot(xm[,"Functional Class"]~xm[,"fold"],
+        main="",xlab="Log2 fold change",col="black",pch=88)
+@
+\begin{sidewaysfigure}
+\centering
+\includegraphics[width=0.9\textwidth]{supplementary-figDotplotTW1Only}
+\caption{{\bf Distribution of functional classes of differntially expressed genes
+in comparing TW1 wild-type grown in glucose and galactose.}
+}
+\label{figDotplotTW1Only}
+\end{sidewaysfigure}
+
 \end{document}

### 2012-02-02 package/smutans/R/core.R

@@ -367,26 +367,38 @@ smutans.mannwhitney <- function( qval=0.05,genes.criteria,go.genes,cat.desc ) {
 
 smutans.prepareTranscript <- function ()
 {
-  # Predicted transcripts
-  parseRNAseqOutFile <- "smutans/inst/extdata/FASTQ001.parsernaseq1"
-  x <- read.table(parseRNAseqOutFile, head=FALSE)
-  number.tx <- length(rownames(x))
-  tx.state <- sub("_[[:digit:]]*", "", substring(x$V9,6))
-  tx.state <- sub("C", "", tx.state)
-  smutansData.tx <- GRanges( seqnames = Rle("chr1", number.tx),
-                              ranges =
-                                IRanges(start=x$V4,
-                                        end=x$V5,
-                                        names=sprintf("tx%04d",seq(number.tx))),
-                              strand = Rle(strand("*"),number.tx),
-                              score=x$V6,
-                              state=as.integer(tx.state)
-                            )
   # ParseRNAseq input pileup or coverage file
   parseRNAseqInFile <- "smutans/inst/extdata/FASTQ001.parsernaseq.pileup"
   x <- read.table(parseRNAseqInFile, head=FALSE)
   pileup.v <- Rle(x$V2)
-  seqlengths(smutansData.tx) <- length(pileup.v)
+
+  # Predicted transcripts
+  smutansData.txAll <- NULL
+  for (i in c(1,2,3,19,20,21,22,23,24,25,26,27,28)) {
+    parseRNAseqOutFile <- paste("smutans/inst/extdata/FASTQ", 
+                                sprintf("%03d",i),
+                                ".parsernaseq1",sep="")
+    x <- read.table(parseRNAseqOutFile, head=FALSE)
+    number.tx <- length(rownames(x))
+    tx.state <- sub("_[[:digit:]]*", "", substring(x$V9,6))
+    tx.state <- sub("C", "", tx.state)
+    smutansData.tx <- GRanges( seqnames = Rle("chr1", number.tx),
+                                ranges =
+                                  IRanges(start=x$V4,
+                                          end=x$V5,
+                                          names=sprintf("tx%04d",seq(number.tx))),
+                                strand = Rle(strand("*"),number.tx),
+                                score=x$V6,
+                                state=as.integer(tx.state)
+                              )
+    seqlengths(smutansData.tx) <- length(pileup.v)
+    if (i == 1) {
[…]

### 2012-02-02 package/test.R

@@ -12,15 +12,16 @@
 library(GenomicRanges)
 #library(DESeq)
 source("smutans/R/core.R")
-smutans.prepareGoseq( )
+#smutans.prepareGoseq( )
 #smutans.prepareData835NPP ()
 #smutans.prepareData34 ()
 #smutans.prepareDataOMZ175( ) 
 #smutans.prepareDataUA159CSP( ) 
 #smutans.prepareDataSMU86CSP( )
 #smutans.prepareDataUA159TW1()
-#smutans.prepareTranscript( )
-q("no")
+smutans.prepareTranscript( )
+#q("no")
+
 library(DESeq)
 library(goseq)
 load("smutans/data/smutansGenes.RData")
@@ -31,9 +32,46 @@ load("smutans/data/smutans.cat.desc.RData")
 load("smutans/data/smutans.genes.criteria.RData")
 load("smutans/data/smutans.go.genes.RData")
 load("smutans/data/smutansData.txPileup.RData")
+load("smutans/data/smutansData.txGenes.RData")
+load("smutans/data/smutansData.txRNAz.RData")
+load("smutans/data/smutansData.txAll.RData")
+load("smutans/data/smutansData.txAnnotation.RData")
 source("smutans/R/class_and_slots.R")
 source("smutans/R/core.R")
 source("smutans/R/methods.R")
+
+
+
+###################################################################################
+# ???
+###################################################################################
+#geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
+#geneDescription <- geneDescription[,c(4,7)]
+#colnames(geneDescription) <- c("Gene ID","Gene Description")
+#functionalDescription <- read.csv("feature-genome.out-geneonly-oralgen-functional.txt",head=FALSE)
+#functionalDescription <- functionalDescription[,c(2,12)]
+#colnames(functionalDescription) <- c("Gene ID","Functional Class")
+#
+#load("glucoseOnly.RData")
+#x <- smutans.de2List( glucoseOnly, qval=1e-3 )
+#x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
+#colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
+#x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
+#x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
[…]

### 2012-02-02 sh/transcript-parsernaseq.sh

@@ -41,7 +41,9 @@ function transcript-parsernaseq {
       REFGENOMEFASTA=$(grep REFGENOMEFASTA $SPECIESFILE | cut -d":" -f2)
       # for g in $(eval echo {1..$NUMFASTQFILE}); do
       # for g in $(eval echo {1..1}); do
-      for g in 1 19 20 35 36; do
+# TW1-Glu/Gal Comparison   : 2,24,25     vs. 3,26,27
+# UA159-Glu/Gal Comparison : 1,19,20     vs. 21,22,23,28
+      for g in 1 2 3 19 20 21 22 23 24 25 26 27 28; do
         FASTQNUM=FASTQ$(printf "%03d" $g)
 
         COMMAND1="perl pl/$FUNCNAME.pl pileup \
@@ -121,13 +123,13 @@ function transcript-parsernaseq {
 
           # scp $BATCHFILE swiftgen:Documents/Projects/rnaseq/src/ParseRNAseq-1.1
         else
-          #echo $COMMAND1 | bash
+          echo $COMMAND1 | bash
           #echo $COMMAND1
           #echo $COMMAND2
           #echo $COMMAND3
           #echo $COMMAND3a
           #echo $COMMAND4
-          echo $COMMAND5 | bash
+          #echo $COMMAND5 | bash
           #echo $COMMAND5a | bash
           #echo $COMMAND6
           #echo $COMMAND7

### 2012-02-03 package/manuscript.Rnw
Just an update.

@@
 
-<<echo=FALSE,results=hide>>=
+<<eval=FALSE,echo=FALSE,results=hide>>=
 # not run when smutans R package is installed
@@
 source("smutans/R/methods.R")
-load("smutans/data/smutansData.tx.RData")
-load("smutans/data/smutansData.txGenes.RData")
-load("smutans/data/smutansData.txGrCombined.RData")
-load("smutans/data/smutansData.txPileup.RData")
-load("smutans/data/smutans.feature.genes.RData")
-load("smutans/data/smutans.genes.criteria.RData")
-load("smutans/data/smutans.go.genes.RData")
-load("smutans/data/smutans.cat.desc.RData")
-load("smutans/data/smutansGenes.RData")
-load("smutans/data/smutansGenes2.RData")
-
-# Copied from supplementary.Rnw. Compare the following with ones above to remove
-# redundant lines.
 load("smutans/data/smutansData.tx.RData")
@@
 load("smutans/data/smutansGenes.RData")
+load("smutans/data/smutansGenes2.RData")
 @
-<<eval=FALSE,echo=FALSE,results=hide>>=
+<<echo=FALSE,results=hide>>=
 # not run without smutans R package installed
@@
 library(smutans)
-load(smutansData.tx)
-load(smutansData.txGenes)
-load(smutansData.txGrCombined)
-load(smutansData.txPileup)
-load(smutans.feature.genes)
+data(smutansData.tx)
+data(smutansData.txGenes)
+data(smutansData.txGrCombined)
+data(smutansData.txAnnotation)
+data(smutansData.txRNAz)
+data(smutansData.txPileup)
+data(smutans.feature.genes)
 data(smutans.genes.criteria)
@@
 data(smutans.cat.desc)
+data(smutans.keg.genes)
+data(smutans.keg.desc)
 data(smutansGenes)
@@
 <<echo=FALSE,result=hide>>=
-#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
-#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
-#                            condA="ua159", condB="tw1" )
-#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
-load("glucoseOnly.RData")
+glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
+glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
+                            condA="ua159", condB="tw1" )
+save(glucoseOnly, file=file.path("glucoseOnly.RData"))
+#load("glucoseOnly.RData")
+
 smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
@@
 <<echo=FALSE,result=hide>>=
-#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
-#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
-#                              condA="ua159", condB="tw1" )
-#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
-load("galactoseOnly.RData")
+galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
+galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
+                              condA="ua159", condB="tw1" )
+save(galactoseOnly, file=file.path("galactoseOnly.RData"))
+#load("galactoseOnly.RData")
 x <- smutans.de2List( galactoseOnly, qval=1e-3 )
@@
 <<echo=FALSE,result=hide>>=
-#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
-#ua159only <- smutans.de2( ua159only, type="ua159", 
-#                          condA="glucose", condB="galactose" )
-#save(ua159only, file=file.path("ua159only.RData"))
-load("ua159only.RData")
+ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
+ua159only <- smutans.de2( ua159only, type="ua159", 
+                          condA="glucose", condB="galactose" )
+save(ua159only, file=file.path("ua159only.RData"))
+#load("ua159only.RData")
 x <- smutans.de2List( ua159only, qval=1e-10 )

### 2012-02-03 package/supplementary.Rnw
Just an update.

@@
 @
-<<eval=FALSE,echo=FALSE,results=hide>>=
+<<echo=FALSE,results=hide>>=
 # not run without smutans R package installed
@@
 library(smutans)
-load(smutansData.tx)
-load(smutansData.txGenes)
-load(smutansData.txCombined)
-load(smutansData.txAnnotation)
-load(smutansData.txRNAz)
-load(smutansData.txPileup)
-load(smutans.feature.genes)
+data(smutansData.tx)
+data(smutansData.txGenes)
+data(smutansData.txGrCombined)
+data(smutansData.txAnnotation)
+data(smutansData.txRNAz)
+data(smutansData.txPileup)
+data(smutans.feature.genes)
 data(smutans.genes.criteria)
@@
 data(smutansGenes)
+data(smutansGenes2)
 @
@@
 <<echo=FALSE,results=tex>>=
-#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
-#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
-#                            condA="ua159", condB="tw1" )
-#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
-load("glucoseOnly.RData")
+glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
+glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
+                            condA="ua159", condB="tw1" )
+save(glucoseOnly, file=file.path("glucoseOnly.RData"))
+#load("glucoseOnly.RData")
 x <- smutans.de2List( glucoseOnly, qval=1e-3 )
@@
 <<echo=FALSE,results=tex>>=
-#galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
-#galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
-#                              condA="ua159", condB="tw1" )
-#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
-load("galactoseOnly.RData")
+galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
+galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
+                              condA="ua159", condB="tw1" )
+save(galactoseOnly, file=file.path("galactoseOnly.RData"))
+#load("galactoseOnly.RData")
 x <- smutans.de2List( galactoseOnly, qval=1e-3 )
@@
 <<echo=FALSE,results=tex>>=
-#ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
-#ua159only <- smutans.de2( ua159only, type="ua159", 
-#                          condA="glucose", condB="galactose" )
-#save(ua159only, file=file.path("ua159only.RData"))
-load("ua159only.RData")
+ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
+ua159only <- smutans.de2( ua159only, type="ua159", 
+                          condA="glucose", condB="galactose" )
+save(ua159only, file=file.path("ua159only.RData"))
+#load("ua159only.RData")
 x <- smutans.de2List( ua159only, qval=1e-10 )
@@
 <<echo=FALSE,results=tex>>=
-#tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
-#tw1only <- smutans.de2( tw1only, type="tw1", 
-#                          condA="glucose", condB="galactose" )
-#save(tw1only, file=file.path("tw1only.RData"))
-load("tw1only.RData")
+tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
+tw1only <- smutans.de2( tw1only, type="tw1", 
+                          condA="glucose", condB="galactose" )
+save(tw1only, file=file.path("tw1only.RData"))
+#load("tw1only.RData")
 x <- smutans.de2List( tw1only, qval=1e-3 )

### 2012-02-03 package/smutans/R/core.R

@@ -171,6 +171,74 @@ smutans.prepareDataOMZ175 <- function ()
   save(smomzGenes, file=file.path("smutans", "data", "smomzGenes.RData"))
 }
 
+smutans.prepareDataUA159PH75 <- function ()
+{
+  # Count data preparation
+  countsFile <- "smutans/inst/extdata/count.txt"
+  conds <- scan(file=paste(countsFile,"index",sep="."), what="character")
+  subconds <- conds %in% c("UA159PH7", "UA159PH5")
+  countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)
+  rownames(countsTable) <- countsTable$gene
+  countsTable <- countsTable[,-1]
+  countsTable <- countsTable[,subconds]
+  conds <- conds[subconds]
+  conds.type <- conds %in% c("UA159PH7", "UA159PH5")
+  factor.type <- rep("ua159",length(conds))
+  conds.condition <- conds %in% c("UA159PH7")
+  factor.condition <- rep("UA159PH7",length(conds))
+  conds.condition <- conds %in% c("UA159PH5")
+  factor.condition[conds.condition] <- "UA159PH5"
+  samples <- data.frame(type=factor.type,condition=factor.condition)
+  rownames(samples) <- colnames(countsTable) 
+
+  design <- samples
+  smUA159pHGenes <- newCountDataSet( countsTable, design )
+  expdata = new("MIAME", 
+     name="S. mutans UA159 pH7 and pH5", 
+     lab="University of Florida, and Cornell University", 
+     contact="Drs. Robert Burne, Michael Stanhope, and Adam Siepel", 
+     title="Streptococcus mutans RNA-Seq Studies", 
+     url="http://www.ncbi.nlm.nih.gov/projects/geo/query/acc.cgi?acc=XXX", 
+     abstract="RNA-seq of UA159 pH 7 and pH 5 replicates from Streptococcus mutans")
+  pubMedIds(expdata) <- "999999999"
+  experimentData(smUA159pHGenes) <- expdata
+  save(smUA159pHGenes, file=file.path("smutans", "data", "smUA159pHGenes.RData"))
+}
+
+smutans.prepareDataSMU21PH75 <- function ()
+{
+  # Count data preparation
+  countsFile <- "smutans/inst/extdata/count.txt"
+  conds <- scan(file=paste(countsFile,"index",sep="."), what="character")
[…]

### 2012-02-03 package/test.R

@@ -10,7 +10,7 @@
 
 
 library(GenomicRanges)
-#library(DESeq)
+library(DESeq)
 source("smutans/R/core.R")
 #smutans.prepareGoseq( )
 #smutans.prepareData835NPP ()
@@ -19,7 +19,9 @@ source("smutans/R/core.R")
 #smutans.prepareDataUA159CSP( ) 
 #smutans.prepareDataSMU86CSP( )
 #smutans.prepareDataUA159TW1()
-smutans.prepareTranscript( )
+#smutans.prepareTranscript( )
+smutans.prepareDataSMU21PH75() 
+smutans.prepareDataUA159PH75()
 #q("no")
 
 library(DESeq)
@@ -39,6 +41,22 @@ load("smutans/data/smutansData.txAnnotation.RData")
 source("smutans/R/class_and_slots.R")
 source("smutans/R/core.R")
 source("smutans/R/methods.R")
+
+load("smutans/data/smUA159pHGenes.RData")
+load("smutans/data/smSMU21pHGenes.RData")
+
+save.to.dir <- "/Users/goshng/Documents/Projects/RNASeq-Analysis/output/email/to/robert-burne/020212"
+ua159ph75 <- newSmutans( smUA159pHGenes, title="UA159 pH 7 vs. pH 5" )
+ua159ph75 <- smutans.de2( ua159ph75, type="ua159", 
+                          condA="UA159PH7", condB="UA159PH5" )
+f <- paste(save.to.dir, "ua159ph75.csv", sep="/")
+smutans.de2List( ua159ph75, f )
+
+smu21ph75 <- newSmutans( smSMU21pHGenes, title="Smu21 pH 7 vs. pH 5" )
+smu21ph75 <- smutans.de2( smu21ph75, type="ua159", 
+                          condA="Smu21PH7", condB="Smu21PH5" )
+f <- paste(save.to.dir, "smu21ph75.csv", sep="/")
+smutans.de2List( smu21ph75, f )
 
 
 

### 2012-02-03 sh/batch.sh

@@ -88,6 +88,7 @@ function batch-output {
 
 function batch-rmessage {
   echo "bash $BASEDIR/copy-data.sh"
+  echo "work at cac:$CACWORKDIR"
 }
 
 function batch-speciesfile {
@@ -130,6 +131,7 @@ while [ 1 ]; do
       bash run-\$STATUS.sh
     fi  
   fi  
+  sleep 30
 
   if [ "\$STATUS" == "bwa-align" ]; then
     NJOBS=\$(qstat | grep $PROJECTNAME-BWA | wc -l) 
@@ -138,14 +140,17 @@ while [ 1 ]; do
       bash run-\$STATUS.sh
     fi  
   fi  
+  sleep 30
 
   if [ "\$STATUS" == "de-all" ]; then
     NJOBS=\$(qstat | grep $PROJECTNAME-DE | wc -l) 
     if [ \$NJOBS -eq 0 ]; then
-      STATUS=parsernaseq
+      STATUS=sum-de-all
+#      STATUS=parsernaseq
       bash \$STATUS.sh
     fi  
   fi  
+  sleep 30
 
   if [ "\$STATUS" == "parsernaseq" ]; then
     NJOBS=\$(qstat | grep $PROJECTNAME-PARSE | wc -l) 
@@ -154,6 +159,7 @@ while [ 1 ]; do
       bash \$STATUS.sh
     fi  
   fi  
+  sleep 30
 
   if [ "\$STATUS" == "rnaz" ]; then
     NJOBS=\$(qstat | grep $PROJECTNAME-RNAZ | wc -l) 
@@ -162,10 +168,12 @@ while [ 1 ]; do
       bash \$STATUS.sh
     fi  
   fi  
+  sleep 30
 
   if [ "\$STATUS" == "sum-de-all" ]; then
     STATUS=exit
   fi  
+  sleep 30
 
   if [ "\$STATUS" == "exit" ]; then
     break

### 2012-03-01 package/smutans/inst/doc/smutans.Rnw
Just an update.

@@
 
-\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data\\using R packages}}}
+\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data}}}
 
@@
 \begin{abstract}
-We analyze RNA samples from \textit{Streptococcus mutans}.  We use
+We analyze RNA samples from \textit{Streptococcus mutans} using
 RNA-seq to define genes whose expression changes between samples
 \cite{Oshlack2010}. Short reads are first mapped to a reference genome using
-approaches such as \program{BWA} \cite{Li2009a}.  Reads fitting into one of the annotated
+approaches such as \program{BWA} \cite{Li2009a} or \program{subread}.
+Reads fitting into one of the annotated
 intervals are counted.  These counts are passed to a statistical package which
-treats differences in read counts for statistical significance. For this
+detects differences in read counts for statistical significance. For this
 purpose, we made use of \Rpackage{DEseq} \cite{Anders2010} or \Rpackage{edgeR}
@@
 size being 100 base pairs long, short reads that consist of RNAs and sequencing
-adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
+adapter sequences would not be mapped. We used \program{cutadapt} \cite{Martin2011} to
 remove parts of adapter sequences from short reads before mapping on the
 reference genome. Mapped short read alignments were converted into readable
-formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
+formats \cite{Li2009b}. We trimmed sequence by quality score from the 3' end
+with threshold score of 20 \cite{Schmieder2011}.
+We counted short reads aligned to annotated genes in the
 reference genome, forming a table of read counts of the genes. Statistical
@@
 samples. Further analyses such as functional category enrichment test
-\cite{Young2010} were also employed.
+\cite{Young2010} were also employed.  To obtain count tables of short reads we
+used scripts available at 
+\textit{\http://github.com/goshng/RNASeq-Analysis}. In the following we will use
+the output of short reads
 
 The \Rpackage{smutans} expects count data in the form of \Rclass{CountDataSet},
-which is a type provided by \Rpackage{DESeq}.  Using \program{BWA} \cite{Li2009a} and
+which is a type supported by \Rpackage{DESeq}.  Using \program{BWA} \cite{Li2009a} and
 \program{samtools} \cite{Li2009b} we aligned short reads from RNA sequencing. Using gene
@@
 <<smsmu86csp,results=hide>>=
-smu86csp <- newSmutans( smsmu86cspGenes, title="UA159 no CSP vs. CSP" )
+smu86csp <- newSmutans( smsmu86cspGenes, title="SMU86 no CSP vs. CSP" )
 smu86csp <- smutans.de2( smu86csp, type="ua159", 

### 2012-03-01 sh/batch2.sh

@@ -38,12 +38,13 @@ function batch2 {
     batch2-run-cram
     batch2-run-fastq-qc
     batch2-run-bwa-align
-    batch2-run-de
 
     # This would replace fastq-qc, bwa-align, and de.
     # Because this creates similar job scripts to those created by fastq-qc,
     # bwa-align, or de, place qcalignde after them.
     batch2-run-qcalignde
+
+    create-index
 
     # batch2-run-blast
 
@@ -117,6 +118,7 @@ function batch2-output {
 
 
 function batch2-speciesfile {
+  FASTQLABEL=$(grep ^FASTQLABEL\: $SPECIESFILE | cut -d":" -f2)
   TESTFASTQ=$(grep ^TESTFASTQ\: $SPECIESFILE | cut -d":" -f2)
   TESTFASTQNUM=FASTQ$(printf "%03d" $TESTFASTQ)
   CRAMDIR=$(grep ^CRAMDIR\: $SPECIESFILE | cut -d":" -f2)
@@ -572,139 +574,8 @@ EOF
 
 }
 
-function batch2-run-de {
-  STATUS=de
-  GENOMEFASTA=$(basename $REFGENOMEFASTA)
-  GENOMEGFF=$(basename $REFGENOMEGFF)
-
-cat>$BASEDIR/run-$STATUS.sh<<EOF
-#!/bin/bash
-FASTQFILES=( $FASTQFILES )
-sed s/PBSARRAYSIZE/\${#FASTQFILES[@]}/g < batch-$STATUS.sh > tbatch.sh
-nsub tbatch.sh 
-rm tbatch.sh
-EOF
-
-cat>$BASEDIR/batch-$STATUS.sh<<EOF
-#!/bin/bash
-#PBS -l walltime=${DEWALLTIME}:00:00,nodes=1
-#PBS -A ${BATCHACCESS}
-#PBS -j oe
-#PBS -N $PROJECTNAME-DE
-#PBS -q ${QUEUENAME}
-#PBS -m e
-# #PBS -M ${BATCHEMAIL}
-#PBS -t 1-PBSARRAYSIZE
-
-ID=FASTQIDENTIFIER
-function copy-data {
-  cd \$TMPDIR
-
-  # Programs and scripts.
-  cp -r \$PBS_O_WORKDIR/pl .
-  cp \$PBS_O_WORKDIR/job-de* .
-  cp \$PBS_O_WORKDIR/job-de*.R .
-  cp \$HOME/bin/samtools .
[…]

### 2012-03-05 package/manuscript.Rnw
Just an update.

@@
 library(smutans)
+inDir = system.file("extdata", package="smutans", mustWork=TRUE)
+smutansGenes <- 
+  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
+               indexFile =file.path(inDir, "count-ua159.txt.index"),
+               condition = c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL"),
+               firstFactor = c("UA159GLU", "UA159GAL"),
+               firstFactorLabel = c("ua159", "tw1"),
+               secondFactor = c("UA159GLU", "TW1GLU"),
+               secondFactorLabel = c("glucose","galactose"),
+               name="S. mutans UA159, TW1, Glucose, and Galactose")
 data(smutansData.tx)
@@
 data(smutans.keg.desc)
-data(smutansGenes)
 data(smutansGenes2)
@@
                          title="All Samples for Comparing Conditions" )
-allSample <- smutans.de2( allSample )
+allSample <- smutans.de2( allSample, 
+                          condA="glucose", condB="galactose" )
 save(allSample, file=file.path("allSample.RData"))
@@
                          title="All Samples for Comparing Conditions" )
-allSample <- smutans.de2( allSample )
+allSample <- smutans.de2( allSample, 
+                          condA="glucose", condB="galactose" )
 tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
@@
 allSampleType <- newSmutans( smutansGenes2, title="All samples for type" )
-allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
+allSampleType <- smutans.de2( allSampleType, compareCondition="no", 
+                              condA="ua159", condB="tw1" )
 # smutans.de2List( allSampleType )[grep("locus",rownames(smutans.de2List(allSampleType))),]
@@
 \end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% REMOVED TEXT
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Regulation of carbon catabolite repression (CCR) in \textit{Streptococcus
-mutans} can involve a phosphoenolpyruvate:sugar phosphotransferase system (PTS),
-histidin protein (HPr) and a transcriptional regulator, CcpA \citep{Gorke2008}.
-CcpA was shown to regulate metabolism and virulence gene expression in
-\textit{S.\ mutans} by \citet{Abranches2008}. They used microarray experiments
-to measure expression levels of genes. Microarray experiments tend to be
-qualitative. The advent of high throughput sequencing technology motivated us to
-revisit the experiment to refine the result.  The deep RNA sequencing data also
-allowed us to define the bacterial transcriptome of \textit{S.\ mutans}. So, we
-could compare putative operon expression levels as well as gene expression
-levels between the wild-type (UA159) and its mutant-type (TW1) whose
-\textit{ccpA} was deleted from the wild-type. This allowed us to profile
-expression of genes and putative operons affected by the regulation of
-\textit{ccpA}.
-
-\citet{Martin2010} developed a hidden Markov model to predict transcripts by
-using site-wise expression levels. I also refined the prediction of transcripts
-using multiple samples. I measured expression levels of known genes and compared
-them from one experiment condition and from another. I can test whether two
-genes are in the same operon or not. I assume that expression levels of two genes 
-vary in similar fashion if the two genes belong to the same operon. What I mean
-by similar fashion is whether they are expressed differentially with similar
-fold changes. I assume that operons maps do not change between the two
-experiments, or multiple experiments. This is not true. I hope that additional
-samples could help to define operons. Eventually, operon maps will be determined
-experimentally. Computational tools could be helpful before those ages come to
-us. \fixme{This idea did not work. We should only use the prediction method.}
-
-We aims at refining the results of \cite{Abranches2008} in more quantative ways.
-Find CCR-sensitive transcripts. We describe predictions made for transcripts and
-small RNAs based on RNA-seq data. Combining the predictions and known genes we
-determine differentially expressed regions in UA159 genome.
-
-In the data generation phase (FIG. 1a), total RNAs or mRNAs are fragmented and
-converted into a library of cDNAs containing sequencing adaptors.  
-In the data analysis phase (FIG. 1b), these short reads are pre-processed to
-remove sequencing errors and other artefacts. The reads are subsequently
-assembled to reconstruct the original RNAs and to assess their abun- dance
-(`expression counting').
-
-To increase the number of assem- bled transcripts, especially the less abundant
-ones, ribosomal RNA (rRNA) and abundant transcripts are removed during the
-first steps of library construction.
-In order to retain RNAs without a poly(A) tail in the assem- bled transcriptome,
-rRNA contamination can instead be removed by hybridization-based depletion
-methods. \fixme{What depletion method did we use?} 
-Last, the use of strand-specific RNA-seq protocols aids in the assembly and
-quantification of overlapping transcripts that are derived from opposite strands
-of the genome. This consideration is especially important for gene-dense
-genomes, such as those of bacteria, archaea and lower eukaryotes.
-
-Sequencing. The major factors to consider before sequencing a sample are: the
-choice of sequencing plat- form, the sequencing read length and whether to use a
-paired-end protocol.
-
-Removing artefacts from RNA- seq data sets before assembly improves the read
-quality, which, in turn, improves the accuracy and computa- tional efficiency of
-the assembly. This step is straight- forward and can be executed using several
-tools.  In general, three types of artefacts should be removed from raw RNA-seq
-data: sequencing adaptors, which origi- nate from failed or short DNA
-insertions during library preparation; low-complexity reads; and
-near-identical reads that are derived from PCR amplification. 
-Sequencing errors in NGS reads can be removed or corrected by analysing the
-quality score and/or the k-mer frequency.
-
-Reference-based transcriptome assembly is easier to perform for the simple
-transcriptomes of bac- terial, archaeal and lower eukaryotic organisms, as these
-organisms have few introns and little alternative splicing. Transcription
-boundaries can be inferred from regions of contiguous read coverage in the
-genome even with- out graph construction and traversal37,51,52. Alternative
-transcription start and stop sites can also be inferred based on the 5′ cap or
-poly(A) signals (if cap- or end- specific experimental protocols are used)51,53.
-However, complications arise owing to the gene-dense nature of
-these genomes. Many genes overlap, resulting in adja- cent genes being assembled
-into one transcript, even though they are not from a polycistronic RNA. Strand-
-specific RNA-seq has successfully been used to separate adjacent overlapping
-genes from opposite strands in the genome51,52. Overlapping genes that are
-transcribed from the same strand and that also have comparable expres- sion
-levels cannot easily be separated without using cap- or end-specific RNA-seq.
-
-In summary, reference-based assembly is generally preferable for cases in which
-a high-quality reference genome already exists.
-
-Meanwhile, experimental RNA-seq and sequencing protocols are continually
-improving and should greatly reduce the informatics challenges.  RNA-seq reads
-from third-generation sequencers, such as PacBio70, are longer (up to several
-kilobases). PacBio sequencers are capable of sequencing a single transcript to
-its full length in a single read. If this technology reaches a throughput that
-is comparable to the second-generation technologies, then the need for
-transcriptome assembly will probably be eliminated. Hopefully, the future of
-transcriptome assembly will be ``no assembly required''.
-
-What is the sequencing depth?
-
-\fixme{I might have to try different filters of short reads to see how those
-affect the transcript prediction.}
-
-Deletion of \textit{ccpA} resulted in substantial relief of the CCR of
-\textit{fruA} expression exerted by both glucose and fructose. 
-When \citet{Wen2002} deleted CRE sequencs in the \textit{fruA} promoter region, CCR was
-almost completely alleviated.  They also showed that 
-CcpA directly interacted with the \textit{fruA} promoter region.
-
-The role of CcpA in gene expression during growth on glucose.
-
-\textit{S.\ mutans} UA159 and TW1 grown under catabolite-repressing conditions
-(glucose), and catabolite-derepressing conditions (galactose). Under
-glucose-only condition we found xx number of differentially expressed genes.

### 2012-03-05 package/smutans/inst/doc/smutans.Rnw
Just an update.

@@
 detects differences in read counts for statistical significance. For this
-purpose, we made use of \Rpackage{DEseq} \cite{Anders2010} or \Rpackage{edgeR}
-\cite{Robinson2010}. This prodecure returns a list of differentially expressed
+purpose, we make use of \Rpackage{DEseq} \cite{Anders2010}.
+This prodecure returns a list of differentially expressed
 genes, which can be passed to downstream analysis, such as functional category
@@
 
-Steps of analyzing RNA-seq data for determining differentially expressed genes
-are described in \cite{Oshlack2010}. We used a short read aligner called
-\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
-\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
-aligner allowed a few gaps for efficient alignment of millions of sequences of
-size being 100 base pairs long, short reads that consist of RNAs and sequencing
-adapter sequences would not be mapped. We used \program{cutadapt} \cite{Martin2011} to
-remove parts of adapter sequences from short reads before mapping on the
-reference genome. Mapped short read alignments were converted into readable
-formats \cite{Li2009b}. We trimmed sequence by quality score from the 3' end
-with threshold score of 20 \cite{Schmieder2011}.
-We counted short reads aligned to annotated genes in the
-reference genome, forming a table of read counts of the genes. Statistical
-software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
-was subsequently employed to infer differentially expressed in different
-biological conditions.  We also generated tracks for our recently released
-Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
-summarize gene expression levels based the short read alignments. This allowed
-us to visualize parts of the reference genome for comparing different RNA
-samples. Further analyses such as functional category enrichment test
-\cite{Young2010} were also employed.  To obtain count tables of short reads we
-used scripts available at 
-\textit{\http://github.com/goshng/RNASeq-Analysis}. In the following we will use
-the output of short reads
+Analyzing RNA-seq data for determining differentially expressed genes
+are well described \cite{Oshlack2010}. We used a short read aligner called
+\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference
+genome, \textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because
+the aligner allowed a few gaps for efficient alignment of millions of sequences
+of size being 100 base pairs long, short reads that consist of RNAs and
+sequencing adapter sequences would not be mapped. By using \program{cutadapt}
+\cite{Martin2011} we removed parts of adapter sequences from short reads before
+mapping on the reference genome.  We also trimmed short read sequences by quality score from
+the 3' end with threshold score of 20 \cite{Schmieder2011}. Mapped short read
+alignments were converted into readable formats \cite{Li2009b}. We counted short
+reads aligned to annotated genes in the reference genome, forming a table of
+read counts of the genes. Statistical software R \cite{R2011} with
+\Rpackage{DESeq} \cite{Anders2010} was subsequently employed to infer
+differentially expressed in different biological conditions. Most of functions
+in comparing different experiments were derived from \Rpackage{DESeq}. Readers
+should refer to the package for details.  We also generated
+tracks for a Streptococcus Genome Browser
+(http://strep-genome.bscb.cornell.edu) that summarize gene expression levels
+based on the short read alignments. This allowed us to visualize parts of the
+reference genome for comparing different RNA samples. Further analyses such as
+functional category enrichment test \cite{Young2010} were also employed.  To
+obtain count tables of short reads we used scripts available at a website
+http://github.com/goshng/RNASeq-Analysis. In the following we will use
+the count table to determine differentially expressed genes.
 
 The \Rpackage{smutans} expects count data in the form of \Rclass{CountDataSet},
-which is a type supported by \Rpackage{DESeq}.  Using \program{BWA} \cite{Li2009a} and
-\program{samtools} \cite{Li2009b} we aligned short reads from RNA sequencing. Using gene
-annotations in the UA159 genome we counted short reads that overlapped the
-genes.  The count values are raw counts of sequencing reads. The
-\Rpackage{DESeq}'s statistical model, or negative binomial model, hold for raw
-reads that allows to assess the measurement precision correctly; we did not have
-rounded values of normalized counts, or counts of covered base pairs.  Several
-exeriments of \textit{S.\ mutans} RNA sampling were performed.  \fixme{Find the
-papers that are related with the experiments.} One of them was presented in the
-\Robject{smutansGenes} as a part of \Rpackage{smutans}. The
-\Robject{smutansGenes} is of \Rclass{CountDataSet}, which is the data type used
-by \Rpackage{DESeq}. We have a convenient \Rclass{Smutans}, which contains slots
-for \Rclass{CountDataSet} objects, and results from analyses thereof.  Before
-loading necessary R packages and data sets, let us describe what functions were
-used for preparing data objects such as \Robject{smutansGenes}. One should
-create an R script file at the directory that contains the source code directory
-of the \Rpackage{smutans}. Put the following scripts in the R script file and
-run the script at that directory. That should create RData files at the
-directory smutans/data. User should not run the R script because the data files
-should be found in the \Rpackage{smutans}.
-
-% http://www.oralgen.lanl.gov/oralgen/downloads/smut.csv
-% http://www.oralgen.lanl.gov/cgi-bin/gene_id_search.cgi?dbname=smut&gene_id=SMu0001
-% \fixme{Where are flat files for genes of S. mutans UA159?}
+which is a type supported by \Rpackage{DESeq}.  Note that the count values
+are raw counts of sequencing reads.  The \Rpackage{DESeq}'s statistical model,
+or negative binomial model, holds for raw reads counts that allows to assess the
+measurement precision correctly; we did not have rounded values of normalized
+counts, or counts of covered base pairs to form count tables.  Several
+exeriments of \textit{S.\ mutans} 
+RNA sampling were performed to examine effects of glucose and galactose,
+835NP and 835P, OMZ175 and OMZ/HKRR, CSP, and acid tolerance.  The comparison of
+glucose and galactose is presented in the \Robject{smutansGenes} as a part of
+\Rpackage{smutans}.
+
+<<options,results=hide,echo=FALSE>>=
+options(digits=3)
+@
+In the following chapter, we use \Robject{smutansGenes}, which is of type
+\Rclass{CountDataSet} of \Rpackage{DESeq}.  Load the \Rpackage{smutans}, and the
+data \Robject{smutansGenes}. 
+<<smutans,results=hide>>=
+library( "smutans" )
+@
+
+Browse the installed directory of \Rpackage{smutans} for several count
+table and index files. As we go along the chapters of this vignette, we will use
+them.
+<<readSmutans>>=
+inDir <- system.file("extdata", package="smutans", mustWork=TRUE)
+dir(inDir,pattern="count*")
+@
+
+Look at two kinds of files: count.txt and count.txt.index. We use
+\Rfunction{readSmutans} to read a count table and its index files. You can check
+if two factor levels are correctly assigned using \Rfunction{pData}. Refer to
+the help documentation of the function in R.
+<<eval=FALSE>>=
+help(readSmutans)
+@
+<<readSmutans>>=
+smutansGenes <- 
+  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
+               indexFile =file.path(inDir, "count-ua159.txt.index"),
+               condition = c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL"),
+               firstFactor = c("UA159GLU", "UA159GAL"),
+               firstFactorLabel = c("ua159", "tw1"),
+               secondFactor = c("UA159GLU", "TW1GLU"),
+               secondFactorLabel = c("glucose","galactose"),
+               name="S. mutans UA159, TW1, Glucose, and Galactose")
+@
+
+The following functions create R data objects. Users are not recommended to use
+them because they have been already created and saved together in the package.
+They will be replaced by functions of \Rpackage{smutans}. They are documented
+for the sake of reminders.
 <<samplePreparation,eval=FALSE>>=
@@
 smutans.prepareData34( )
-smutans.prepareData835NPP( )
 smutans.prepareDataUA159TW1( ) 
 smutans.prepareGoseq( )
-smutans.prepareDataOMZ175( ) 
-smutans.prepareDataUA159CSP( ) 
-smutans.prepareDataSMU86CSP( )
 smutans.prepareTranscript( )
@@
 @
-<<options,results=hide,echo=FALSE>>=
-options(digits=3)
-@
-Now, see some of the data objects by loading the \Rpackage{smutans}, and a data
-\Robject{smutansGenes}. Note that the object is of type \Rclass{CountDataSet} of
-\Rpackage{DESeq} in which users should find \Rfunction{counts} and
-\Rfunction{pData}. 
-<<smutans,results=hide>>=
-library( "smutans" )
-data( "smutansGenes" )
-@
-There are 
+Now we look at the detail of the loaded count table. There are 
 \Sexpr{length(colnames(counts(smutansGenes)))}
@@
 @
-The samples differ by experimental condition (``glucose'' or
-``galactose'') and by library type (``ua159'' or ``tw1'').
-
-Samples of 835NP, 835P, and UA159 no CSP should be loaded using the following
-command.
-<<sm835,results=hide>>=
-data( "sm835Genes" )
-@
-
-% Resources
-% Google charts - Venn Diagram
-% http://code.google.com/apis/chart/index.html
-
-\fixme{Describe other data objects as well.
-sm34Genes.RData,
-sm835Genes.RData,
-smutans.cat.desc.RData,
-smutans.feature.genes.RData, and
-smutans.go.genes.RData.}
-
-%--------------------------------------------------
-\section{DE analysis using UA159 and TW1 samples with
-glucose and galactose} \label{sec:detw1}
-%--------------------------------------------------
-
-We could compare four pairwise comparisons using the two-factor data set of
-\Robject{smutansGenes}: a pair of comparisons between ``glucose'' and
-``galactose'' given either type ``ua159'' or ``tw1'', and another pair of
-comparisons between types ``ua159'' and ``tw1'' given either condition
-``glucose'' or ``galactose.'' We focus on one of the four pairwise comparisons
-first. Later, we will perform analyses for the rest of the three comparisons.
-We look at the ``ua159'' data for now, and will look at the ``tw1'' data later.
-
-We can cluster samples using variance stabilized data. Figure \ref{figClust}
-shows the \Sexpr{length(colnames(counts(smutansGenes)))} sampled clustered. 
-The clustering should reflect our experimental design, i.e., samples are
-more similar when they have the same treatment or the same library type.
-To make this conclusion, we estimate the dispersion with
-mode ``blind'' (see \Rpackage{DESeq} vignette), as only then,
-the variance stabilizing transformation is not informed about the design and
-hence not biased towards a result supporting it. Such an analysis is useful
-for quality control.  This appears to be useful as a first in any analysis.
+The samples differ by experimental condition (``glucose'' or ``galactose'') and
+by library type (``ua159'' or ``tw1'').  Use the following functions to get
+helps such as opening the vignette and manual of \Rpackage{smutans}. 
+<<eval=FALSE>>=
+help.start()
+vignette(package="smutans")
+openVignette("smutans")
+@
+
+%--------------------------------------------------
+\section{DE analysis using UA159 and TW1 samples with glucose and galactose}
+%--------------------------------------------------
+
+We compared four pairs of experiments using the \Robject{smutansGenes}: two comparisons
+between ``glucose'' and ``galactose'' given either type ``ua159'' or ``tw1'',
+and another two comparisons between types ``ua159'' and ``tw1'' given either
+condition ``glucose'' or ``galactose.'' Before comparing these we clustered
+samples using read counts. Figure \ref{figClust} shows the cluster of the
+\Sexpr{length(colnames(counts(smutansGenes)))} samples. Refer to
+\Rpackage{DESeq} vignette as to how the clustering is conducted.  The clustering
+was useful in ascertaining that the nutrient effect was stronger than the mutant
+effect: two experimental conditions of ``glucose'' and ``galactose''
+distinguished the samples. Within each condition, two different types of
+``ua159'' and ``tw1'' separated them.
 
@@
 @
-As expected, two experimental conditions of ``glucose'' and ``galactose'' divide the
-samples. Within each condition, two different types of ``ua159'' and ``tw1'' divide
-them again. The pair of ``glucose ua159'' and ``galactose tw1'' is the most
-distinguished.
 
@@
 \subsection{Inference using single-factor data} 
-\label{subsec:ua159only}
 %------------------------------------------------------------
 
-We create a \Rclass{Smutans} object using the \Rclass{CountDataSet}
-\Robject{smutansGenes}. The title option would be useful in distinguishing
-different objects of \Rclass{Smutans} as we have more R objects with the type.  
-A subsequent call to
-\Rfunction{smutans.de2} uses a two-factor data set to test differential
-expression of genes.  The \Rfunction{smutans.de2} estimates effective library
-sizes, and dispersions using functions provided by \Rpackage{DESeq}. Then, it
-tests genes for differential expression.
-<<ua159onlySamples>>=
+Let us compare ``glucose'' and ``galactose'' conditions given ``ua159'' type. 
+\Robject{smutansGenes}
+contains short reads counts based on bwa alignment on UA159 genome.  The title
+argument would be useful in distinguishing different objects of \Rclass{Smutans}
+as we have more R objects of the same type. A subsequent call to
+\Rfunction{smutans.de2} uses the two-factor data set to test differential
+expression for the annotated genes by estimating
+effective library sizes, and dispersions using functions provided by
+\Rpackage{DESeq}.
+<<ua159only>>=
 ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
@@
                           condA="glucose", condB="galactose" )
-@
+head(smutans.de2List( ua159only ))
+@
+Descriptions of the columns are in order.
+Each gene has its mean expression level (at the base scale) as a joint estimate
+from both conditions, and estimated separately for each condition, the fold
+change from the first to the second condition, i.e.,
+$\frac{baseMeanB}{baseMeanA}$, the logarithm (to basis 2) of the fold change,
+and the $p$ value for the statistical significance of this change. The
+\texttt{padj} column contains the $p$ values, adjusted for multiple testing with
+the Benjamini-Hochberg procedure (see the \Rfunction{p.adjust}),
+which controls false discovery rate (FDR). Later, we print the gene list in a
+CSV file using option file of \Rfunction{smutans.de2List}. Windows users should
+be able to open CSV files with a spreadsheet program such as EXCEL. Users who
+are more comfortable with the EXCEL program than R could post-process CSV files
+for sorting and finding differentially expressed genes using different q-value
+cutoff values.
+<<eval=TRUE>>=
+smutans.de2List( ua159only, "ua159only.csv" )
+@
+
+By fixing ``tw1'' type similarly we compared ``glucose'' and ``galactose''
+conditions.
+<<tw1onlySamples>>=
+tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
+tw1only <- smutans.de2( tw1only, type="tw1", 
+                        condA="glucose", condB="galactose" )
+@
+We can print the result in a csv file.
+<<eval=TRUE>>=
+# not run
+smutans.de2List( tw1only, "tw1only.csv" )
+@
+We compared ``glucose'' and ``galactose'' with either ``ua159'' or ``tw1''
+fixed. We computed numbers of differentially expressed genes that were
+determined by both or either one of them.
+<<>>=
+tabUA159onlyTW1only <- smutans.compareResult( ua159only, tw1only )
+addmargins( tabUA159onlyTW1only )
+@
+The table, \Robject{tabUA159onlyTW1only}, shows numbers of genes found to be 
+differentially expressed in the two analyses: ``UA159 Only'' and ``TW1 Only.''
+Of the total \Sexpr{sum(tabUA159onlyTW1only)} genes, \Sexpr{tabUA159onlyTW1only[2,2]}
+genes were found in both analyses. The analysis using TW1-only sample found
+\Sexpr{tabUA159onlyTW1only[1,2]} among genes UA159-only sample could not
+find. The analysis using UA159-only sample found
+\Sexpr{tabUA159onlyTW1only[2,1]} genes among ones TW1-only sample could not
+find. UA159-only sample detected more differentially expressed genes.
+
+We now compare sample types on differentially expressed genes. Create two
+\Rclass{Smutans} objects: \Robject{glucoseOnly} and \Robject{galactoseOnly}, and
+compare them in detecting differentially expressed genes:
+<<glucoseOnlySamples>>=
+glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
+glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
+                            condA="ua159", condB="tw1" )
+galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
+galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
+                              condA="ua159", condB="tw1" )
+tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
+                                                      galactoseOnly )
+addmargins( tabGlucoseOnlyGalactoseOnly )
+@
+Of the total \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} genes,
+\Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]} genes were found in both analyses at FDR
+rate of 10\%. The analysis using galactose-only sample found
+\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} of the genes that were negative in
+glucose-only sample.  The analysis using glucose-only
+sample found \Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} of the genes that were
+negative in galactose-only sample. 
+Table \ref{tab:de} summarizes the number of differentially expressed genes that
+were found using single-factor analyses at FDR of 10\%. We can print them in csv
+files using \Rfunction{smutans.de2List}.
+<<eval=TRUE>>=
+# not run
+smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
+smutans.de2List( galactoseOnly, "galactoseOnly.csv" )
+@
+
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Numbers of differentially expressed genes in different analyses}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{ccccl}
+\hline 
+Sample A & Sample B & Number of Genes & Result File \\
+\hline 
+UA159 Glucose & UA159 Galactose & \Sexpr{sum(ua159only@padj < .1,na.rm=T)} & ua159only.csv \\
+TW1 Glucose & TW1 Galactose & \Sexpr{sum(tw1only@padj < .1,na.rm=T)} & tw1only.csv \\
+UA159 Glucose & TW1 Glucose & \Sexpr{sum(glucoseOnly@padj < .1,na.rm=T)} & glucoseOnly.csv \\
+UA159 Galactose & TW1 Galactose & \Sexpr{sum(galactoseOnly@padj < .1,na.rm=T)} & galactoseOnly.csv \\
+\hline 
+\end{tabular}
+\par\end{centering}
+\label{tab:de}
+\end{table}
+%-----------------------------------------------------------------------------
+
+%------------------------------------------------------------
+\subsection{Inference using all of the samples} 
+%------------------------------------------------------------
+
+Let us determine genes differentially expressed in different conditions of
+``glucose'' and ``galactose'' using both types of ``ua159'' and ``tw1'': a
+two-factor analysis. Start
+with the two-factor analysis by creating a new \Rclass{Smutans} object, called
+\Robject{allSample}. Although we already used \Robject{allSample} when
+clustering the samples, we recreate it for instructive purposes.  The
+\Rfunction{smutans.de2} can use a two-factor data set to perform two-factor
+analysis: we are here interested in condition effect. 
+<<allSamples,results=hide>>=
+allSample <- newSmutans( smutansGenes, 
+                         title="All Samples for Comparing Conditions" )
+allSample <- smutans.de2( allSample,
+                          condA="glucose", condB="galactose" )
+tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
+                                                   allSample )
+@
+<<>>=
+addmargins( tabUA159onlyAndAllSample )
+@
+Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
+\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. 
+Using all of the samples detected more differentially expressed genes than did
+ua159-only sample.
+The output from \Robject{allSample} is almost the same except that there are
+convergence column.
+<<>>=
+head( smutans.de2List( allSample ) )
+@
+
+We can show that taking the library type into account was important to have
+good detection power by doing the analysis againt using the standard workflow,
+as outlined earlier, and without informing \Rpackage{DESeq} of the library
+types. This anlaysis is not necessary. However, it clearly shows the advantage of the
+two-factor analysis.
+<<allSampleWithoutType>>=
+allSampleWithoutType <- newSmutans( smutansGenes, "All Sample without Type" )
+allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
+                                                condA="glucose", 
+                                                condB="galactose" )
+tabWithoutWithAllSample <- smutans.compareResult( allSampleWithoutType, 
+                                                  allSample )
+addmargins( tabWithoutWithAllSample )
+@
+Using all of the sample that considers types found more differentially expressed
+genes than did one ignoring the types.
+
+We now proceed to compare the two types by considering all of the samples, and
+compare the result with the single-factor analysis.
+<<allSamplesType,results=hide>>=
+allSampleType <- newSmutans( smutansGenes, title="All samples for type" )
+allSampleType <- smutans.de2( allSampleType, compareCondition="no",
+                              condA="ua159", condB="tw1" )
+tabGlucoseOnlyAllSample <- smutans.compareResult( glucoseOnly, 
+                                                  allSampleType )
+tabGalactoseOnlyAllSample <- smutans.compareResult( galactoseOnly, 
+                                                    allSampleType )
+@                                                    
+<<>>=
+addmargins( tabGlucoseOnlyAllSample )
+addmargins( tabGalactoseOnlyAllSample )
+@
+Of the total \Sexpr{sum(tabGlucoseOnlyAllSample)} genes,
+\Sexpr{tabGlucoseOnlyAllSample[2,2]} genes were found in both analyses. The
+analysis using all of the sample found \Sexpr{tabGlucoseOnlyAllSample[1,2]}
+of the genes that glucose-only sample did not find. 
+The analysis using glucose-only sample
+found \Sexpr{tabGlucoseOnlyAllSample[2,1]} the genes that all samples did not
+find.  In the comparison using galactose-only sample,
+of the \Sexpr{sum(tabGalactoseOnlyAllSample)} genes,
+\Sexpr{tabGalactoseOnlyAllSample[2,2]} genes were found in both analyses. The
+analysis using all of the sample found \Sexpr{tabGalactoseOnlyAllSample[1,2]}
+the genes galactose-only sample did not find. 
+The analysis using galactose-only sample
+found \Sexpr{tabGalactoseOnlyAllSample[2,1]} the genes that all samples did not
+find.
+
+One can print the sults in CSV files.
+<<eval=TRUE>>=
+smutans.de2List( allSample, "allSample.csv" )
+smutans.de2List( allSampleType, "allSampleType.csv" )
+@
+
+%------------------------------------------------------------
+\subsection{Functional category association analyses} 
+%------------------------------------------------------------
+
+We find some functional categories enriched with differentially expressed genes
+using \Rpackage{goseq} \cite{Young2010} (table \ref{tab:goseq}).
+To find functional categories enriched with differentially expressed genes we
+first find a list of differentially expressed genes. A set of genes are assigned
+to functional gene categories: \Robject{smutans.go.genes}. Features of the genes in
+UA159 include gene lengths: \Robject{smutans.feature.genes}.
+Functional categories with numbers of genes assigned are in
+\Robject{smutans.cat.desc}.
+<<goseqUA159only,eval=FALSE>>=
+data( "smutans.feature.genes" )
+data( "smutans.go.genes" )
+data( "smutans.cat.desc" )
+smutans.de2Goseq ( ua159only, qval=1e-2, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+genes}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+$q^a$ & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<goseqUA159only,echo=FALSE,results=tex>>=
+data( "smutans.feature.genes" )
+data( "smutans.go.genes" )
+data( "smutans.cat.desc" )
+smutans.de2Goseq ( ua159only, qval=1e-2, 
+                   feature.gene=smutans.feature.genes,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseq}
+\end{table}
+%-----------------------------------------------------------------------------
+
+%--------------------------------------------------
+\section{DE analysis using 835NP, 835P, and UA159 no CSP}
+%--------------------------------------------------
+
+Sang-Joon Ahn is leading this experiment. Let us create the input data variable.
+We compare three comparisons using three conditions. We do the cluster,
+determine differentially expressed genes, and print those genes.  One of the
+three conditions, 835P, is distinguished from the others.  We found no
+functional categories enriched with differentially expressed genes in the
+pairwise comparisons of 835NP, 835P, and UA159 no CSP.  Let's make a list of
+differentially expressed genes.  
+<<readSmutans>>=
+sm835Genes <- 
+  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
+                indexFile =file.path(inDir, "count-ua159.txt.index"),
+                condition = c("835NP", "835P", "UA159noCSP"),
+                firstFactor = c("835NP", "835P", "UA159noCSP"),
+                firstFactorLabel = c("ua159"),
+                secondFactor = c(),
+                secondFactorLabel = c("835np","835p", "ua159nocsp"),
+                name="S. mutans 835NP, 835P, UA159 no CSP")
+@
+
+<<>>=
+head( counts(sm835Genes), n=1 )
+pData( sm835Genes )
+@
+
+<<figClustNpAndP,fig=TRUE>>=
+npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
+smutans.de2Clust( npAndP )
+@
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{smutans-figClustNpAndP}
+\caption{Empirical (black dots) and fitted (red lines) dispersion values
+plotted against mean expression strength.}
+\label{figClustNpAndP}
+\end{figure}
+We first cluster the \Sexpr{length(colnames(counts(sm835Genes)))} samples
+(fig.\ \ref{figClustNpAndP}).
+
+<<sm835Samples>>=
+npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
+npAndP <- smutans.de2( npAndP, type="ua159", 
+                       condA="835np", condB="835p" )
+npAndUA159 <- newSmutans( sm835Genes, title="835NP vs. UA159" )
+npAndUA159 <- smutans.de2( npAndUA159, type="ua159", 
+                           condA="835np", condB="ua159nocsp" )
+pAndUA159 <- newSmutans( sm835Genes, title="835P vs. UA159" )
+pAndUA159 <- smutans.de2( pAndUA159, type="ua159", 
+                          condA="835p", condB="ua159nocsp" )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( npAndP, "npAndP.csv" )
+smutans.de2List( npAndUA159, "npAndUA159.csv" )
+smutans.de2List( pAndUA159, "pAndUA159.csv" )
+@
+
+%--------------------------------------------------
+\section{DE analysis using OMZ175 and OMZ/HKRR}
+%--------------------------------------------------
+
+We can use either UA159 or OMZ175 genome as a reference genome.  We first
+cluster the samples (fig.\ \ref{figClustomz}).  We did not find many genes in
+this comparison (see below).
+
+We also mapped short reads on OMZ175 contigs. We wished to find genes in OMZ175 but not in
+UA159 to see whether they are enriched with differentially expressed genes. We
+also wished to find genes in UA159 but not in OMZ175 to check in the same
+fashion. What are the distribution of the genes that unique to either UA159 or
+OMZ175 in differentially expressed genes? We could test the independence of
+differntially expressed genes and unique genes.
+Use 0.1 as the q-value cutoff to have two sets of differentially expressed genes
+and not ones. And also divide genes unique to SMU109 and not one. Test the
+independence of differential expression and SMU109 unique genes using Fisher's
+exact test. This was not useful because of small number of differentially
+expressed genes.
+<<readSmutans>>=
+smomzGenes <- 
+  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
+               indexFile =file.path(inDir, "count-ua159.txt.index"),
+               condition = c("OMZ175", "OMZHKRR"),
+               firstFactorLabel = c("ua159"),
+               secondFactorLabel = c("OMZ175", "OMZHKRR"),
+               name="S. mutans OMZ175, and OMZHKRR based on UA159")
+@
+
+<<>>=
+head( counts(smomzGenes), n=1 )
+pData( smomzGenes )
+@
+
+<<figClustomz,fig=TRUE>>=
+omz <- newSmutans( smomzGenes, title="OMZ175 vs. OMZ/HKRR" )
+smutans.de2Clust( omz )
+@
+
+<<smomz,results=hide>>=
+omz <- smutans.de2( omz, type="ua159", 
+                    condA="OMZ175", condB="OMZHKRR" )
+@
+
+<<>>=
+smutans.de2List( omz )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( omz, "omz.csv" )
+@
+
+<<readSmutans>>=
+smomzGenesSMU109 <- 
+  readSmutans (countsFile=file.path(inDir, "count-omz-bwa.txt"),
+               indexFile =file.path(inDir, "count-omz.txt.index"),
+               condition = c("OMZ175", "OMZHKRR"),
+               firstFactorLabel = c("smu109"),
+               secondFactorLabel = c("OMZ175", "OMZHKRR"),
+               name="S. mutans OMZ175, and OMZHKRR on SMU109 contigs")
+@
+
+<<>>=
+head( counts(smomzGenesSMU109), n=1 )
+pData( smomzGenesSMU109 )
+@
+
+<<figClustomzSMU109,fig=TRUE>>=
+omzSMU109 <- newSmutans( smomzGenesSMU109, title="OMZ175 vs. OMZ/HKRR" )
+smutans.de2Clust( omzSMU109 )
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=.4\textwidth]{smutans-figClustomz}
+\includegraphics[width=.4\textwidth]{smutans-figClustomzSMU109}
+\caption{Empirical (black dots) and fitted (red lines) dispersion values
+plotted against mean expression strength. Left-hand side is for UA159-based
+result, and right-hand side for SMU109 or OMZ175 contigs-based.}
+\label{figClustomz}
+\end{figure}
+
+<<smomz,results=hide>>=
+omzSMU109 <- smutans.de2( omzSMU109, type="smu109", 
+                          condA="OMZ175", condB="OMZHKRR" )
+@
+
+<<>>=
+smutans.de2List( omzSMU109 )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( omzSMU109, "omzSMU109.csv" )
+@
+
+%--------------------------------------------------
+\section{DE analysis using UA159 CSP and SMU86 CSP}
+%--------------------------------------------------
+
+We follow the procedure in the previous chapter on the comparison of glucose and
+galactose. The cluster of the samples shows that the effect of types ua159 and
+smu86 is stronger than that of conditions CSP (fig.\ \ref{figClustCSP}).
+Differences due to CSP was more profound in ua159 type than was in smu86
+(table \ref{tab:decsp}).
+
+We used 0.01 as the q-value cutoff to have two sets of differentially expressed genes
+and not ones. And also divide genes unique to SMU86 and not one. Test the
+independence of differential expression and SMU86 unique genes using Fisher's
+exact test. We found evidence that the differences were due to genes
+unique to SMU86. However, we did not find evidence that differences were due to genes
+unique to UA159. The proportion of genes unique to SMU86 was larger than that
+of common genes. This shows that genes unique to SMU86 have more profound
+effects than common genes under CSP, but genes unique to UA159 does not have
+those effects.
+
+<<options,results=hide>>=
+csp.qval.cutoff <- 10^-1
+@
+
+<<readSmutans>>=
+<<readSmutans>>=
+cspGenes <-
+  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
+                indexFile =file.path(inDir, "count-ua159.txt.index"),
+                condition = c("UA159noCSP", "UA159CSP", "Smu86noCSP", "Smu86CSP"),
+                firstFactor = c("UA159noCSP", "UA159CSP"),
+                firstFactorLabel = c("ua159", "smu86"),
+                secondFactor = c("UA159noCSP", "Smu86noCSP"),
+                secondFactorLabel = c("nocsp","csp"),
+                name="S. mutans no CSP and with CSP")
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{smutans-figClustCSP}
+\caption{Heatmap showing the Euclidean distances between the samples
+as calculated from the variance-stabilizing transformation of the
+count data.}
+\label{figClustCSP}
+\end{figure}
+<<figClustCSP,fig=TRUE>>=
+allSampleCSP <- newSmutans( cspGenes )
+smutans.de2Clust( allSampleCSP )
+@
+
+<<>>=
+ua159csp <- newSmutans( cspGenes, title="UA159" )
+ua159csp <- smutans.de2( ua159csp, type="ua159", 
+                         condA="nocsp", condB="csp" )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( ua159csp, "ua159csp.csv" )
+@
+
+<<smu86csp>>=
+smu86csp <- newSmutans( cspGenes, title="SMU86" )
+smu86csp <- smutans.de2( smu86csp, type="smu86", 
+                             condA="nocsp", condB="csp" )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( smu86csp, "smu86csp.csv" )
+@
+
+<<>>=
+tabUA159SMU86 <- smutans.compareResult( ua159csp, smu86csp,
+                                        qval=csp.qval.cutoff )
+addmargins( tabUA159SMU86 )
+@
+
+<<>>=
+nocsp <- newSmutans( cspGenes, title="NO CSP" )
+nocsp <- smutans.de2( nocsp, condition="nocsp", 
+                        condA="ua159", condB="smu86" )
+csp <- newSmutans( cspGenes, title="CSP" )
+csp <- smutans.de2( csp, condition="csp", 
+                        condA="ua159", condB="smu86" )
+tabnocspcsp <- smutans.compareResult( nocsp, csp, 
+                                      qval=csp.qval.cutoff )
+addmargins( tabnocspcsp )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( nocsp, "nocsp.csv" )
+smutans.de2List( csp, "csp.csv" )
+@
+
+<<allSamplesSmu86,results=hide>>=
+allSampleSmu86 <- newSmutans( cspGenes, 
+                              title="All Samples for Comparing CSP and no CSP" )
+allSampleSmu86 <- smutans.de2( allSampleSmu86,
+                               condA="nocsp", condB="csp" )
+tabUA159cspAndAllSampleSmu86 <- smutans.compareResult( ua159csp, 
+                                                       allSampleSmu86,
+                                                       qval=csp.qval.cutoff )
+tabSMU86cspAndAllSampleSmu86 <- smutans.compareResult( smu86csp, 
+                                                       allSampleSmu86, 
+                                                       qval=csp.qval.cutoff )
+@
+<<>>=
+addmargins( tabUA159cspAndAllSampleSmu86 )
+addmargins( tabSMU86cspAndAllSampleSmu86 )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( allSampleSmu86, "allSampleSmu86.csv" )
+@
+
+<<allSamplesCsp,results=hide>>=
+allSampleCsp <- newSmutans( cspGenes, 
+                              title="All Samples for Comparing SMU86" )
+allSampleCsp <- smutans.de2( allSampleCsp, compareCondition="no",
+                               condA="ua159", condB="smu86" )
+tabnocspAndAllSampleCsp <- smutans.compareResult( nocsp, 
+                                                  allSampleCsp,
+                                                  qval=csp.qval.cutoff )
+tabcspAndAllSampleCsp <- smutans.compareResult( csp, 
+                                                allSampleCsp,
+                                                qval=csp.qval.cutoff )
+@
+<<>>=
+addmargins( tabnocspAndAllSampleCsp )
+addmargins( tabcspAndAllSampleCsp )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( allSampleCsp, "allSampleCsp.csv" )
+@
+
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Numbers of differentially expressed genes in CSP
+analyses.} We use multiple testing adjusted p-value of $\Sexpr{csp.qval.cutoff}$.}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{ccccl}
+\hline 
+Sample A & Sample B & Number of Genes & Result File \\
+\hline 
+UA159 no CSP & UA159 CSP    & \Sexpr{sum(ua159csp@padj < csp.qval.cutoff,na.rm=T)} & ua159ph.csv \\
+SMU86 no CSP & SMU86 CSP    & \Sexpr{sum(smu86csp@padj < csp.qval.cutoff,na.rm=T)} & smu21ph.csv \\
+UA159 no CSP & SMU86 no CSP & \Sexpr{sum(nocsp@padj < csp.qval.cutoff,na.rm=T)} & ph7only.csv \\
+UA159 CSP & SMU86 CSP       & \Sexpr{sum(csp@padj < csp.qval.cutoff,na.rm=T)} & ph5only.csv \\
+\hline 
+\end{tabular}
+\par\end{centering}
+\label{tab:decsp}
+\end{table}
+%-----------------------------------------------------------------------------
+
+Use SMU86 contigs as a reference genome.
+
+<<readSmutans>>=
+smu86cspGenes <-
+  readSmutans (countsFile=file.path(inDir, "count-smu86-bwa.txt"),
+                indexFile =file.path(inDir, "count-smu86.txt.index"),
+                condition = c("Smu86noCSP", "Smu86CSP"),
+                firstFactorLabel = c("smu86"),
+                secondFactorLabel = c("nocsp","csp"),
+                name="S. mutans SMU86 nocsp and csp")
+@
+
+<<>>=
+smu86cspSMU86 <- newSmutans( smu86cspGenes, title="SMU86" )
+smu86cspSMU86 <- smutans.de2( smu86cspSMU86, type="smu86", 
+                              condA="nocsp", condB="csp" )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( smu86cspSMU86, "smu86cspSMU86.csv" )
+@
+
+<<eval=TRUE>>=
+x <- scan(file.path(inDir, "SMU86.diff.lst"), what="character")
+y <- smutans.de2List( smu86cspSMU86, qval=10^-2)$id
+z <- smu86cspSMU86@res$id
+x1 <- z[z %in% x]
+x2 <- z[!z %in% x]
+y1 <- z[z %in% y]
+y2 <- z[!z %in% y]
+sum(x1 %in% y1)
+sum(x1 %in% y2)
+sum(x2 %in% y1)
+sum(x2 %in% y2)
+unique.diff <-
+matrix(c(4, 80, 12, 1710),
+       nrow = 2,
+       dimnames =
+       list(c("Diff", "Not diff"),
+            c("Unique", "Common")))
+fisher.test(unique.diff)
+@
+
+<<>>=
+x <- scan(file.path(inDir, "UA159-SMU86.diff.lst"), what="character")
+y <- smutans.de2List( smu86csp, qval=10^-2)$id
+z <- smu86csp@res$id
+x1 <- z[z %in% x]
+x2 <- z[!z %in% x]
+y1 <- z[z %in% y]
+y2 <- z[!z %in% y]
+sum(x1 %in% y1)
+sum(x1 %in% y2)
+sum(x2 %in% y1)
+sum(x2 %in% y2)
+unique.diff <-
+matrix(c(1, 216, 12, 1731),
+       nrow = 2,
+       dimnames =
+       list(c("Diff", "Not diff"),
+            c("Unique", "Common")))
+fisher.test(unique.diff)
+@
+
+%--------------------------------------------------
+\section{DE analysis using acid tolerance}
+%--------------------------------------------------
+
+Sara is leading the experiment. 
+The clustering shows that the effect of type, ua159 vs\. smu21, is stronger
+than pH effect (fig.\ \ref{figClustpH}).
+There seem to be too many differentially expressed genes in the comparison of
+different pH (table \ref{tab:deph}). We therefore used a much smaller cutoff value of
+multiple testing adjusted $p$-value. 
+When comparing two conditions, pH 7 and pH 5, we found more differentially
+expressed genes in ua159 than we did in smu21.
+When comparing two types, ua159 and smu21, we found more differentially
+expressed genes under pH 5 than we found under pH 7.
+
+We used 0.01 as the q-value cutoff to have two sets of differentially expressed genes
+and not ones. And also divide genes unique to SMU21 and not one. Test the
+independence of differential expression and SMU21 unique genes using Fisher's
+exact test. We did not find evidence that the differences were due to genes
+unique to SMU21. However, we found evidence that differences were due to genes
+unique to UA159. The proportion of genes unique to UA159 was smaller than that
+of common genes. This shows that genes unique to SMU21 have similar effects to
+common genes under pH 5, but genes unique to UA159 does not have those effects.
+
+<<options,results=hide>>=
+ph.qval.cutoff <- 10^-10
+@
+<<readSmutans>>=
+phGenes <-
+  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
+               indexFile =file.path(inDir, "count-ua159.txt.index"),
+               condition = c("UA159PH7", "UA159PH5", "Smu21PH7", "Smu21PH5"),
+               firstFactor = c("UA159PH7", "UA159PH5"),
+               firstFactorLabel = c("ua159", "smu21"),
+               secondFactor = c("UA159PH7", "Smu21PH7"),
+               secondFactorLabel = c("ph7","ph5"),
+               name="S. mutans UA159 pH7 and pH5")
+@
+
+\begin{figure}
+\centering
+\includegraphics[width=.9\textwidth]{smutans-figClustpH}
+\caption{Heatmap showing the Euclidean distances between the samples
+as calculated from the variance-stabilizing transformation of the
+count data.}
+\label{figClustpH}
+\end{figure}
+<<figClustpH,fig=TRUE>>=
+allSamplepH <- newSmutans( phGenes )
+smutans.de2Clust( allSamplepH )
+@
+
+<<ua159ph>>=
+ua159ph <- newSmutans( phGenes, title="UA159" )
+ua159ph <- smutans.de2( ua159ph, type="ua159", 
+                        condA="ph7", condB="ph5" )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( ua159ph, "ua159ph.csv" )
+@
+
+<<smu21ph>>=
+smu21ph <- newSmutans( phGenes, title="SMU21" )
+smu21ph <- smutans.de2( smu21ph, type="smu21", 
+                        condA="ph7", condB="ph5" )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( smu21ph, "smu21ph.csv" )
+@
+
+<<>>=
+tabUA159SMU21 <- smutans.compareResult( ua159ph, smu21ph, qval=ph.qval.cutoff )
+addmargins( tabUA159SMU21 )
+@
+
+<<ph7only>>=
+ph7only <- newSmutans( phGenes, title="pH 7" )
+ph7only <- smutans.de2( ph7only, condition="ph7", 
+                        condA="ua159", condB="smu21" )
+ph5only <- newSmutans( phGenes, title="pH 5" )
+ph5only <- smutans.de2( ph5only, condition="ph5", 
+                        condA="ua159", condB="smu21" )
+tabpH7pH5 <- smutans.compareResult( ph7only, ph5only, qval=ph.qval.cutoff )
+
+addmargins( tabpH7pH5 )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( ph7only, "ph7only.csv" )
+smutans.de2List( ph5only, "ph5only.csv" )
+@
+
+Using all of the samples.
+
+<<allSamplesSmu21,results=hide>>=
+allSampleSmu21 <- newSmutans( phGenes, 
+                              title="All Samples for Comparing pH" )
+allSampleSmu21 <- smutans.de2( allSampleSmu21,
+                               condA="ph7", condB="ph5" )
+tabUA159pHAndAllSampleSmu21 <- smutans.compareResult( ua159ph, 
+                                                      allSampleSmu21,
+                                                      qval=ph.qval.cutoff )
+tabSMU21pHAndAllSampleSmu21 <- smutans.compareResult( smu21ph, 
+                                                      allSampleSmu21,
+                                                      qval=ph.qval.cutoff )
+
+<<>>=
+addmargins( tabUA159pHAndAllSampleSmu21 )
+addmargins( tabSMU21pHAndAllSampleSmu21 )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( allSampleSmu21, "allSampleSmu21.csv" )
+@
+
+<<allSamplesph,results=hide>>=
+allSampleph <- newSmutans( phGenes, 
+                           title="All Samples for Comparing UA159 and SMU21" )
+allSampleph <- smutans.de2( allSampleph, compareCondition="no",
+                            condA="ua159", condB="smu21" )
+tabpH7AndAllSampleph <- smutans.compareResult( ph7only, 
+                                               allSampleph, 
+                                               qval=ph.qval.cutoff )
+tabpH5AndAllSampleph <- smutans.compareResult( ph5only, 
+                                               allSampleph, 
+                                               qval=ph.qval.cutoff )
+@
+
+<<>>=
+addmargins( tabpH7AndAllSampleph )
+addmargins( tabpH5AndAllSampleph )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( allSampleph, "allSampleph.csv" )
+@
+
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Numbers of differentially expressed genes in acid tolerance
+analyses.} We use multiple testing adjusted p-value of $\Sexpr{ph.qval.cutoff}$.}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{ccccl}
+\hline 
+Sample A & Sample B & Number of Genes & Result File \\
+\hline 
+UA159 pH 7 & UA159 pH 5 & \Sexpr{sum(ua159ph@padj < ph.qval.cutoff,na.rm=T)} & ua159ph.csv \\
+SMU21 pH 7 & SMU21 pH 5 & \Sexpr{sum(smu21ph@padj < ph.qval.cutoff,na.rm=T)} & smu21ph.csv \\
+UA159 pH 7 & SMU21 pH 7 & \Sexpr{sum(ph7only@padj < ph.qval.cutoff,na.rm=T)} & ph7only.csv \\
+UA159 pH 5 & SMU21 pH 5 & \Sexpr{sum(ph5only@padj < ph.qval.cutoff,na.rm=T)} & ph5only.csv \\
+\hline 
+\end{tabular}
+\par\end{centering}
+\label{tab:deph}
+\end{table}
+%-----------------------------------------------------------------------------
+
+Use SMU21 contigs as a reference genome.
+
+<<readSmutans>>=
+smu21phGenes <-
+  readSmutans (countsFile=file.path(inDir, "count-smu21-bwa.txt"),
+                indexFile =file.path(inDir, "count-smu21.txt.index"),
+                condition = c("Smu21PH7", "Smu21PH5"),
+                firstFactor = c(),
+                firstFactorLabel = c("smu21"),
+                secondFactor = c(),
+                secondFactorLabel = c("ph7", "ph5"),
+                name="S. mutans UA159 pH7 and pH5 on SMU21")
+@
+
+<<smu21phSMU21>>=
+smu21phSMU21 <- newSmutans( smu21phGenes, title="SMU21" )
+smu21phSMU21 <- smutans.de2( smu21phSMU21, type="smu21", 
+                            condA="ph7", condB="ph5" )
+@
+
+<<eval=TRUE>>=
+# not run
+smutans.de2List( smu21phSMU21, "smu21phSMU21.csv" )
+@
+
+<<eval=TRUE>>=
+x <- scan(file.path(inDir, "SMU21.diff.lst"), what="character")
+y <- smutans.de2List( smu21phSMU21, qval=10^-2)$id
+z <- smu21phSMU21@res$id
+x1 <- z[z %in% x]
+x2 <- z[!z %in% x]
+y1 <- z[z %in% y]
+y2 <- z[!z %in% y]
+sum(x1 %in% y1)
+sum(x1 %in% y2)
+sum(x2 %in% y1)
+sum(x2 %in% y2)
+unique.diff <-
+matrix(c(7, 67, 106, 1562),
+       nrow = 2,
+       dimnames =
+       list(c("Diff", "Not diff"),
+            c("Unique", "Common")))
+fisher.test(unique.diff)
+@
+
+<<>>=
+x <- scan(file.path(inDir, "UA159-SMU21.diff.lst"), what="character")
+y <- smutans.de2List( smu21ph, qval=10^-2)$id
+z <- smu21ph@res$id
+x1 <- z[z %in% x]
+x2 <- z[!z %in% x]
+y1 <- z[z %in% y]
+y2 <- z[!z %in% y]
+sum(x1 %in% y1)
+sum(x1 %in% y2)
+sum(x2 %in% y1)
+sum(x2 %in% y2)
+unique.diff <-
+matrix(c(2, 267, 107, 1575),
+       nrow = 2,
+       dimnames =
+       list(c("Diff", "Not diff"),
+            c("Unique", "Common")))
+fisher.test(unique.diff)
+@
+
+%--------------------------------------------------
+\section{DE analysis using predicted transcripts}
+\label{sec:detranscript}
+%--------------------------------------------------
+
+We need ways of supporting the predicted transcripts. We need a way of using
+expression levels and counts of starts and ends of short reads to predict
+transcripts. For now, we just use an ad hoc method for finding transcript start
+and end sites. A likelihood can be constructed by multiplying geometric
+probabilities. Given a range of possible transcription boundaries we can find
+estimates of positions and geometirc parmeter $p$ that maximize the likelihood
+for both ends of a transcript.
+
+\Rpackage{GenomicRanges} is useful in summarizing predicted transcripts.
+feature-genome.out-geneonly is stored in \Robject{smutans.feature.genes}.  
+File FASTQ001.parsernaseq1 is ParseRNAseq output file for transcript
+predictions. File FASTQ001.parsernaseq.pileup is ParseRNAseq input file for
+site-wise expression levels.
+
+\Rfunction{smutans.prepareTranscript} creates two objects:
+\Robject{smutans.tx} and \Robject{smutans.txPileup}.
+<<tx,eval=FALSE>>=
+source("http://www.bioconductor.org/biocLite.R")
+biocLite("GenomicRanges")
+@
+
+%--------------------------------------------------
+\section{DE analysis using predicted small RNAs and target genes}
+\label{sec:desmallrna}
+%--------------------------------------------------
+
+Create a file with intergenic regions sequences of \textit{S.\ mutans} UA159.
+Create a file with bacterial genome sequences except \textit{S.\ mutans} NN2025.
+BLAST the intergenic regions against the bacterial genomes. Extract sequence
+alignments from the BLAST output. Refine the alignment using MUSCLE. Convert the
+FASTA-format alignments to MSA-format ones.  Find small RNA candidates using
+RNAz.  Find targets of the small RNA candidates using RNAplex. Targets are
+scored for each small RNA candidate. The scores were used to do functional
+category enrichment tests of targets for each small RNA candidate.
+
+Table \ref{tab:goseq} 
+shows functional categories enriched with target genes for predicted small
+RNAs. Let's use \Rfunction{smutans.mannwhitney} to the package and other data
+objects.
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
+genes}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{lcccl}
+\hline 
+small RNA & $q^a$ & count & GO & Description \\
+\hline 
+%-----------------------------------------------------------------------------
+<<goseqUA159only,echo=FALSE,results=tex>>=
+data( "smutans.genes.criteria" )
+data( "smutans.go.genes" )
+data( "smutans.cat.desc" )
+smutans.mannwhitney( qval=0.05,
+                   genes.criteria=smutans.genes.criteria,
+                   go.genes=smutans.go.genes,
+                   cat.desc=smutans.cat.desc )
+@
+%-----------------------------------------------------------------------------
+\hline 
+\end{tabular}
+\par\end{centering}
+\begin{flushleft}
+$^a$False discovery rate estimated by the
+Benjamini-Hochberg method.  All categories having at least ten genes and
+$q\leq 0.05$ are displayed.\\ 
+\end{flushleft}
+\label{tab:goseq}
+\end{table}
+%-----------------------------------------------------------------------------
+
+<<smutansRNA>>=
+data( "smutansGenes2" )
+ua159only <- newSmutans( smutansGenes2, title="UA159 Only" )
+ua159only <- smutans.de2( ua159only, type="ua159", 
+                          condA="glucose", condB="galactose" )
+allSample <- newSmutans( smutansGenes2, 
+                         title="All Samples for Comparing Conditions" )
+allSample <- smutans.de2( allSample, 
+                          condA="glucose", condB="galactose" )
+tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
+                                                   allSample )
+addmargins( tabUA159onlyAndAllSample )
+@
+Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
+\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
+analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
+genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
+genes. We found a few predicted small RNAs that appeared to be differentially
+expressed.
+<<smutansRNA>>=
+smutans.de2List( allSample )[grep("locus",rownames(smutans.de2List(allSample))),]
+@
+We also found a few predicted small RNAs that appeared to be differentially
+expressed in the comparison of the two types: ``ua159'' and ``tw1.''
+<<smutansRNA>>=
+allSampleType <- newSmutans( smutansGenes2, title="All samples for type" )
+allSampleType <- smutans.de2( allSampleType, compareCondition="no",
+                              condA="ua159", condB="tw1" )
+smutans.de2List( allSampleType )[grep("locus",rownames(smutans.de2List(allSampleType))),]
+@
+
+%--------------------------------------------------
+\section{Miscellaneous R scripts}
+%--------------------------------------------------
+
 Because \Rpackage{DESeq} recommended to check
@@
 Let's list genes with resulting information from \Rpackage{DESeq}.
-<<ua159onlyDE2List>>=
-head(smutans.de2List( ua159only ))
-@
-For each gene, we get its mean expression level (at the base scale) as a joint
-estimate from both conditions, and estimated separately for each condition, the
-fold change from the first to the second condition, i.e., 
-$\frac{baseMeanB}{baseMeanA}$, the logarithm (to basis 2)
-of the fold change, and the $p$ value for the statistical significance of this
-change. The \texttt{padj} column contains the $p$ values, adjusted for multiple
-testing with the Benjamini-Hochberg procedure (see the R function
-\Rfunction{p.adjust}), which controls false discovery rate (FDR). Later, we
-print the gene list out in a CSV file using option file of
-\Rfunction{smutans.de2List}.
+
 %-----------------------------------------------------------------------------
@@
 means, coloring in red those genes that are significant at 10\% of FDR.
-Let us use ``tw1'' type to compare ``glucose'' and ``galactose'' conditions.
-We continue to analyze the data set using type ``tw1'' to compare the two
-conditions. We perform the single-factor analysis. 
-<<tw1onlySamples>>=
-tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
-tw1only <- smutans.de2( tw1only, type="tw1", 
-                        condA="glucose", condB="galactose" )
-tabUA159onlyTW1only <- smutans.compareResult( ua159only, tw1only )
-addmargins( tabUA159onlyTW1only )
-@
-The table, \Robject{tabUA159onlyTW1only}, shows numbers of genes found to be 
-differentially expressed in the two analyses: ``UA159 Only'' and ``TW1 Only.''
-Of the total \Sexpr{sum(tabUA159onlyTW1only)} genes, \Sexpr{tabUA159onlyTW1only[2,2]}
-genes were found in both analyses. The analysis using TW1-only sample found
-\Sexpr{tabUA159onlyTW1only[1,2]} genes. The analysis using UA159-only sample found
-\Sexpr{tabUA159onlyTW1only[2,1]} genes.
-
-Until now we focused on the effect of condition either using UA159-only or
-TW1-only samples. We now change the gear to see effects of sample types on
-differentially expressed genes. We create two \Rclass{Smutans} objects:
-\Robject{glucoseOnly} and \Robject{galactoseOnly}, and compare them in detecting
-differentially expressed genes:
-<<glucoseOnlySamples>>=
-glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
-glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
-                            condA="ua159", condB="tw1" )
-galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
-galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
-                              condA="ua159", condB="tw1" )
-tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
-                                                      galactoseOnly )
-addmargins( tabGlucoseOnlyGalactoseOnly )
-@
-Of the total \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} genes,
-\Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]} genes were found in both analyses at FDR
-rate of 10\%. The analysis using galactose-only sample found
-\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} genes. The analysis using glucose-only
-sample found \Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} genes. 
-Table \ref{tab:de} summarizes the number of differentially expressed genes that
-were found using single-factor analyses at FDR of 10\%.
-
-%-----------------------------------------------------------------------------
-\begin{table}[!h]
-\caption{{\bf Numbers of differentially expressed genes in different analyses}}
-\vspace{1ex}
-\noindent \begin{centering}
-\begin{tabular}{ccccl}
-\hline 
-Sample A & Sample B & Number of Genes & Result File \\
-\hline 
-UA159 Glucose & UA159 Galactose & \Sexpr{sum(ua159only@padj < .1,na.rm=T)} & ua159only.csv \\
-TW1 Glucose & TW1 Galactose & \Sexpr{sum(tw1only@padj < .1,na.rm=T)} & tw1only.csv \\
-UA159 Glucose & TW1 Glucose & \Sexpr{sum(glucoseOnly@padj < .1,na.rm=T)} & glucoseOnly.csv \\
-UA159 Galactose & TW1 Galactose & \Sexpr{sum(galactoseOnly@padj < .1,na.rm=T)} & galactoseOnly.csv \\
-% UA159 Glucose & UA159 Galactose & \Sexpr{sum(tabUA159onlyTW1only[2,])} & ua159only.csv \\
-% TW1 Glucose & TW1 Galactose & \Sexpr{sum(tabUA159onlyTW1only[,2])} & tw1only.csv \\
-% UA159 Glucose & TW1 Glucose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[2,])} & glucoseOnly.csv \\
-% UA159 Galactose & TW1 Galactose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[,2])} & galactoseOnly.csv \\
-\hline 
-\end{tabular}
-\par\end{centering}
-\label{tab:de}
-\end{table}
-%-----------------------------------------------------------------------------
-
-%------------------------------------------------------------
-\subsection{Inference using all of the samples} 
-\label{subsec:allSample}
-%------------------------------------------------------------
-
-It is inconvenient to use four different analyses to find differentially
-expressed genes under one of two factors. Besides, it would be desirable to use all of
-the data to find differentially expressed genes because using more samples might
-be more powerful in detecting differentially expressed genes.  Let us find genes
-differentially expressed in different conditions of ``glucose'' and
-``galactose'' using both types of ``ua159'' and ``tw1.'' We will compare these
-genes and those found in the single-factor analysis that we did previously,
-i.e., \Robject{ua159only}.
-Let's start with multi-factor the analysis by creating a new \Rclass{Smutans} object, called
-\Robject{allSample}. Although we already used \Robject{allSample} when
-clustering the samples, we recreate it for instructive purposes.
-<<allSamples,results=hide>>=
-allSample <- newSmutans( smutansGenes, 
-                         title="All Samples for Comparing Conditions" )
-allSample <- smutans.de2( allSample )
-@
-The \Rfunction{smutans.de2} uses a two-factor data set to
-perform two-factor analysis. The two factors are type and condition. We are here
-interested in condition effect. We could check 
+
+We could check 
 the dispersion estimates and to make sure that the data quality is as
@@
 @
+<<figCompareUA159onlyAndAllSample,fig=TRUE>>=
+pval.ua159only <- smutans.pval( ua159only )
+pval.allSample <- smutans.pval( allSample )
+bottom = function(x, theta=1e-12) pmax(x, theta)
+plot( bottom(pval.ua159only), bottom(pval.allSample), log="xy", pch=20, cex=.3 )
+abline(a=0, b=1, col="blue")
+@
+Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from
+the two-factor analysis versus those from the single-factor analysis. The
+p-values from the two-factor analysis are overall smaller than those from the
+single-factor analysis. We would expect that this was the case because of higher
+power of the two-factor analysis.
 %-----------------------------------------------------------------------------
@@
 %-----------------------------------------------------------------------------
-<<figCompareUA159onlyAndAllSample,fig=TRUE>>=
-pval.ua159only <- smutans.pval( ua159only )
-pval.allSample <- smutans.pval( allSample )
-bottom = function(x, theta=1e-12) pmax(x, theta)
-plot( bottom(pval.ua159only), bottom(pval.allSample), log="xy", pch=20, cex=.3 )
-abline(a=0, b=1, col="blue")
-@
-<<compareUA159onlyAndAllSample>>=
-tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
-                                                   allSample )
-addmargins( tabUA159onlyAndAllSample )
-@
-Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
-\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
-analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
-genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
-genes.  Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from
-the two-factor analysis versus those from the single-factor analysis. The
-p-values from the two-factor analysis are overall smaller than those from the
-single-factor analysis. We would expect that this was the case because of higher
-power of the two-factor analysis.
-
-We can show that taking the library type into account was important to have
-good detection power by doing the analysis again using the standard workflow,
-as outlined earlier, and without informing \Rpackage{DESeq} of the library
-types. This anlaysis is not necessary. However, it clearly shows the advantage of the
-two-factor analysis.
-<<allSampleWithoutType>>=
-allSampleWithoutType <- newSmutans( smutansGenes, "All Sample without Type" )
-allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
-                                                condA="glucose", 
-                                                condB="galactose" )
-tabWithoutWithAllSample <- smutans.compareResult( allSampleWithoutType, 
-                                                  allSample )
-addmargins( tabWithoutWithAllSample )
-@
-Of the total \Sexpr{sum(tabWithoutWithAllSample)} genes,
-\Sexpr{tabWithoutWithAllSample[2,2]} genes were found in both analyses. The
-analysis using all of the sample found \Sexpr{tabWithoutWithAllSample[1,2]}
-genes. The analysis using all of the same sample but without considering type
-found only \Sexpr{tabWithoutWithAllSample[2,1]} genes.
-
-As we mentioned above, we can have two analyses using the two-factor analysis.
-Until now we compared conditions.
-We proceed to compare the two types by considering all of the samples.
-<<allSampleType,results=verbatim>>=
-allSampleType <- newSmutans( smutansGenes, title="All samples for type" )
-allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
-tabGlucoseOnlyAllSample <- smutans.compareResult( glucoseOnly, 
-                                                  allSampleType )
-addmargins( tabGlucoseOnlyAllSample )
-tabGalactoseOnlyAllSample <- smutans.compareResult( galactoseOnly, 
-                                                    allSampleType )
-addmargins( tabGalactoseOnlyAllSample )
-@
-Of the total \Sexpr{sum(tabGlucoseOnlyAllSample)} genes,
-\Sexpr{tabGlucoseOnlyAllSample[2,2]} genes were found in both analyses. The
-analysis using all of the sample found \Sexpr{tabGlucoseOnlyAllSample[1,2]}
-genes. The analysis using glucose-only sample
-found \Sexpr{tabGlucoseOnlyAllSample[2,1]} genes.  
-In the comparison using galactose-only sample,
-of the \Sexpr{sum(tabGalactoseOnlyAllSample)} genes,
-\Sexpr{tabGalactoseOnlyAllSample[2,2]} genes were found in both analyses. The
-analysis using all of the sample found \Sexpr{tabGalactoseOnlyAllSample[1,2]}
-genes. The analysis using galactose-only sample
-found \Sexpr{tabGalactoseOnlyAllSample[2,1]} genes.
-
-We print out the result of differentially expressed genes in CSV files. A
-spreadsheet program such as EXCEL should be able to open CSV files. Users who
-are more comfortable with the EXCEL program than R could post-process CSV
-files for sorting and finding differentially expressed genes using different
-q-value cutoff values.
-<<smutanSamples>>=
-smutans.de2List( allSample, "allSample.csv" )
-smutans.de2List( allSampleType, "allSampleType.csv" )
-smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
-smutans.de2List( galactoseOnly, "galactoseOnly.csv" )
-smutans.de2List( ua159only, "ua159only.csv" )
-smutans.de2List( tw1only, "tw1only.csv" )
-head( smutans.de2List( allSample ) )
-@
-The output from \Robject{allSample} is different from that of
-\Robject{ua159only} or \Robject{tw1only}.
-The first column of ``allSample.csv'' is for genes.
-The next three columns show the fitted coefficients, converted to a logarithm
-base 2 scale: ``X.Intercept.'', ``typeua159'', and ``conditionglucose.''
-The log2 fold change due to the condition is shown in the 4th column. As indicated
-by the column name, it is the effect of ``glucose'', i.e., the log ratio of
-``glucose'' versus ``galactose'' or $\log(glucose/galactose)$. 
-Note that the library type also
-had noticeable influence on the expression, and hence would have increased the dispersion
-estimates (and so reduced power) if we had not fitted an effect for it.
-The column \emph{deviance}
-is the deviance of the fit. (Comparing the deviances with a $\chi^2$ likelihood ratio test is
-how \Rfunction{nbinomGLMTest} calculates the $p$ values.) The column, \emph{converged},
-indicates whether the calculation of coefficients and deviance has fully converged.
-(If it is false too often, you can try to change the GLM control parameters, as
-explained in the help page to \Rfunction{fitNbinomGLMs}.)
-The second to the last column, \emph{pval}, is the raw p-values from the test
-\Rfunction{nbinomGLMTest} provided by \Rpackage{DESeq}. The last column,
-\emph{padj}, is the multiple testing adjusted p-values.
-
-%------------------------------------------------------------
-\subsection{Functional category association analyses} 
-\label{subsec:functional}
-%------------------------------------------------------------
 
@@
 
-<<goseqUA159only,eval=FALSE>>=
-data( "smutans.feature.genes" )
-data( "smutans.go.genes" )
-data( "smutans.cat.desc" )
-smutans.de2Goseq ( ua159only, qval=1e-2, 
-                   feature.gene=smutans.feature.genes,
-                   go.genes=smutans.go.genes,
-                   cat.desc=smutans.cat.desc )
-@
-We find some functional categories enriched with differentially expressed genes
-(table \ref{tab:goseq})
-We find functional categories that are enriched with differentially expressed genes.
-To find functional categories enriched with differentially expressed genes we
-first find a list of differentially expressed genes. A set of genes are assigned
-to functional gene categories: \Robject{smutans.go.genes}. Features of the genes in
-UA159 include gene lengths: \Robject{smutans.feature.genes}.
-Functional categories with numbers of genes assigned are in
-\Robject{smutans.cat.desc}.
-%-----------------------------------------------------------------------------
-\begin{table}[!h]
-\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
-genes}}
-\vspace{1ex}
-\noindent \begin{centering}
-\begin{tabular}{cl}
-\hline 
-$q^a$ & Description \\
-\hline 
-%-----------------------------------------------------------------------------
-<<goseqUA159only,echo=FALSE,results=tex>>=
-data( "smutans.feature.genes" )
-data( "smutans.go.genes" )
-data( "smutans.cat.desc" )
-smutans.de2Goseq ( ua159only, qval=1e-2, 
-                   feature.gene=smutans.feature.genes,
-                   go.genes=smutans.go.genes,
-                   cat.desc=smutans.cat.desc )
-@
-%-----------------------------------------------------------------------------
-\hline 
-\end{tabular}
-\par\end{centering}
-\begin{flushleft}
-$^a$False discovery rate estimated by the
-Benjamini-Hochberg method.  All categories having at least ten genes and
-$q\leq 0.05$ are displayed.\\ 
-\end{flushleft}
-\label{tab:goseq}
-\end{table}
-%-----------------------------------------------------------------------------
-
-%--------------------------------------------------
-\section{DE analysis using 835NP, 835P, and UA159 no CSP}
-\label{sec:de835}
-%--------------------------------------------------
-
-We first cluster the \Sexpr{length(colnames(counts(sm835Genes)))} samples.
-\begin{figure}
-\centering
-\includegraphics[width=.9\textwidth]{smutans-figClustNpAndP}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength.}
-\label{figClustNpAndP}
-\end{figure}
-<<figClustNpAndP,fig=TRUE>>=
-data( "sm835Genes" )
-npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
-smutans.de2Clust( npAndP )
-@
-
-We found no functional categories enriched with differentially expressed
-genes in the pairwise comparisons of 835NP, 835P, and UA159 no CSP.
-Let's make a list of differentially expressed genes. 
-Another kind of heatmap plots for
-grouping genes could be interesting. 
-But, basically we have a list of genes with
-some adjusted p-values for detecting differentially expressed genes.
-
-<<look_into_smutansGenes>>=
-head( counts(sm835Genes) )
-pData( sm835Genes )
-@
-
-<<sm835Samples>>=
-npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
-npAndP <- smutans.de2( npAndP, type="ua159", 
-                       condA="835np", condB="835p" )
-npAndUA159 <- newSmutans( sm835Genes, title="835NP vs. UA159" )
-npAndUA159 <- smutans.de2( npAndUA159, type="ua159", 
-                           condA="835np", condB="ua159nocsp" )
-pAndUA159 <- newSmutans( sm835Genes, title="835P vs. UA159" )
-pAndUA159 <- smutans.de2( pAndUA159, type="ua159", 
-                          condA="835p", condB="ua159nocsp" )
-smutans.de2List( npAndP, "npAndP.csv" )
-smutans.de2List( npAndUA159, "npAndUA159.csv" )
-smutans.de2List( pAndUA159, "pAndUA159.csv" )
-@
-
 \begin{figure}
@@
 
-%--------------------------------------------------
-\section{DE analysis using OMZ175 and OMZ/HKRR}
-\label{sec:deomz}
-%--------------------------------------------------
-
-<<look_into_smomzGenes>>=
-data( "smomzGenes" )
-head( counts(smomzGenes) )
-pData( smomzGenes )
-@
-
-<<smomz,results=hide>>=
-omz <- newSmutans( smomzGenes, title="OMZ175 vs. OMZ/HKRR" )
-omz <- smutans.de2( omz, type="ua159", 
-                    condA="OMZ175", condB="OMZHKRR" )
-smutans.de2List( omz, "omz.csv" )
-@
-
-%--------------------------------------------------
-\section{DE analysis using UA159 CSP}
-\label{sec:deua159csp}
-%--------------------------------------------------
-
-<<look_into_smua159cspGenes>>=
-data( "smua159cspGenes" )
-head( counts(smua159cspGenes) )
-pData( smua159cspGenes )
-@
-
-<<smua159csp,results=hide>>=
-ua159csp <- newSmutans( smua159cspGenes, title="UA159 no CSP vs. CSP" )
-ua159csp <- smutans.de2( ua159csp, type="ua159", 
-                         condA="UA159noCSP", condB="UA159CSP" )
-smutans.de2List( ua159csp, "ua159csp.csv" )
-@
-
-%--------------------------------------------------
-\section{DE analysis using SMU86 CSP}
-\label{sec:desmu86csp}
-%--------------------------------------------------
-
-<<look_into_smsmu86cspGenes>>=
-data( "smsmu86cspGenes" )
-head( counts(smsmu86cspGenes) )
-pData( smsmu86cspGenes )
-@
-
-<<smsmu86csp,results=hide>>=
-smu86csp <- newSmutans( smsmu86cspGenes, title="SMU86 no CSP vs. CSP" )
-smu86csp <- smutans.de2( smu86csp, type="ua159", 
-                         condA="Smu86noCSP", condB="Smu86CSP" )
-smutans.de2List( smu86csp, "smu86csp.csv" )
-@
-
-%--------------------------------------------------
-\section{DE analysis using predicted transcripts}
-\label{sec:detranscript}
-%--------------------------------------------------
-
-We need ways of supporting the predicted transcripts. We need a way of using
-expression levels and counts of starts and ends of short reads to predict
-transcripts. For now, we just use an ad hoc method for finding transcript start
-and end sites. A likelihood can be constructed by multiplying geometric
-probabilities. Given a range of possible transcription boundaries we can find
-estimates of positions and geometirc parmeter $p$ that maximize the likelihood
-for both ends of a transcript.
-
-\Rpackage{GenomicRanges} is useful in summarizing predicted transcripts.
-feature-genome.out-geneonly is stored in \Robject{smutans.feature.genes}.  
-File FASTQ001.parsernaseq1 is ParseRNAseq output file for transcript
-predictions. File FASTQ001.parsernaseq.pileup is ParseRNAseq input file for
-site-wise expression levels.
-
-\Rfunction{smutans.prepareTranscript} creates two objects:
-\Robject{smutans.tx} and \Robject{smutans.txPileup}.
-<<tx,eval=FALSE>>=
-source("http://www.bioconductor.org/biocLite.R")
-biocLite("GenomicRanges")
-@
-
-%--------------------------------------------------
-\section{DE analysis using predicted small RNAs and target genes}
-\label{sec:desmallrna}
-%--------------------------------------------------
-
-Create a file with intergenic regions sequences of \textit{S.\ mutans} UA159.
-Create a file with bacterial genome sequences except \textit{S.\ mutans} NN2025.
-BLAST the intergenic regions against the bacterial genomes. Extract sequence
-alignments from the BLAST output. Refine the alignment using MUSCLE. Convert the
-FASTA-format alignments to MSA-format ones.  Find small RNA candidates using
-RNAz.  Find targets of the small RNA candidates using RNAplex. Targets are
-scored for each small RNA candidate. The scores were used to do functional
-category enrichment tests of targets for each small RNA candidate.
-
-Table \ref{tab:goseq} 
-shows functional categories enriched with target genes for predicted small
-RNAs. Let's use \Rfunction{smutans.mannwhitney} to the package and other data
-objects.
-%-----------------------------------------------------------------------------
-\begin{table}[!h]
-\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
-genes}}
-\vspace{1ex}
-\noindent \begin{centering}
-\begin{tabular}{lcccl}
-\hline 
-small RNA & $q^a$ & count & GO & Description \\
-\hline 
-%-----------------------------------------------------------------------------
-<<goseqUA159only,echo=FALSE,results=tex>>=
-data( "smutans.genes.criteria" )
-data( "smutans.go.genes" )
-data( "smutans.cat.desc" )
-smutans.mannwhitney( qval=0.05,
-                   genes.criteria=smutans.genes.criteria,
-                   go.genes=smutans.go.genes,
-                   cat.desc=smutans.cat.desc )
-@
-%-----------------------------------------------------------------------------
-\hline 
-\end{tabular}
-\par\end{centering}
-\begin{flushleft}
-$^a$False discovery rate estimated by the
-Benjamini-Hochberg method.  All categories having at least ten genes and
-$q\leq 0.05$ are displayed.\\ 
-\end{flushleft}
-\label{tab:goseq}
-\end{table}
-%-----------------------------------------------------------------------------
-
-<<smutansRNA>>=
-data( "smutansGenes2" )
-ua159only <- newSmutans( smutansGenes2, title="UA159 Only" )
-ua159only <- smutans.de2( ua159only, type="ua159", 
-                          condA="glucose", condB="galactose" )
-allSample <- newSmutans( smutansGenes2, 
-                         title="All Samples for Comparing Conditions" )
-allSample <- smutans.de2( allSample )
-tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
-                                                   allSample )
-addmargins( tabUA159onlyAndAllSample )
-@
-Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
-\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
-analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
-genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
-genes. We found a few predicted small RNAs that appeared to be differentially
-expressed.
-<<smutansRNA>>=
-smutans.de2List( allSample )[grep("locus",rownames(smutans.de2List(allSample))),]
-@
-We also found a few predicted small RNAs that appeared to be differentially
-expressed in the comparison of the two types: ``ua159'' and ``tw1.''
-<<smutansRNA>>=
-allSampleType <- newSmutans( smutansGenes2, title="All samples for type" )
-allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
-smutans.de2List( allSampleType )[grep("locus",rownames(smutans.de2List(allSampleType))),]
-@
-
 \section{Session Info}

### 2012-03-05 package/smutans/R/core.R

@@ -100,145 +100,6 @@ smutans.prepareDataUA159TW1 <- function (cutoff.count=10)
   save(smutansGenes2, file=file.path("smutans", "data", "smutansGenes2.RData"))
 }
 
-smutans.prepareData835NPP <- function ()
-{
-  # Count data preparation
-  countsFile <- "smutans/inst/extdata/count.txt"
-  conds <- scan(file=paste(countsFile,"index",sep="."), what="character")
-  subconds <- conds %in% c("835NP", "835P", "UA159noCSP")
-  countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)
-  rownames(countsTable) <- countsTable$gene
-  countsTable <- countsTable[,-1]
-  countsTable <- countsTable[,subconds]
-  conds <- conds[subconds]
-  conds.type <- conds %in% c("835NP", "835P", "UA159noCSP")
-  factor.type <- rep("ua159",length(conds))
-  # All of the same factor: factor.type[!conds.type] <- "tw1"
-  conds.condition <- conds %in% c("835NP")
-  factor.condition <- rep("835np",length(conds))
-  conds.condition <- conds %in% c("835P")
-  factor.condition[conds.condition] <- "835p"
-  conds.condition <- conds %in% c("UA159noCSP")
-  factor.condition[conds.condition] <- "ua159nocsp"
-  samples <- data.frame(type=factor.type,condition=factor.condition)
-  rownames(samples) <- colnames(countsTable) 
-
-  design <- samples
-  sm835Genes <- newCountDataSet( countsTable, design )
-  expdata = new("MIAME", 
-     name="S. mutans 835NP, 835P, UA159 no CSP", 
-     lab="University of Florida, and Cornell University", 
-     contact="Drs. Robert Burne, Michael Stanhope, and Adam Siepel", 
-     title="Streptococcus mutans RNA-Seq Studies", 
-     url="http://www.ncbi.nlm.nih.gov/projects/geo/query/acc.cgi?acc=XXX", 
-     abstract="RNA-seq of 8 biological replicates from Streptococcus mutans")
-  pubMedIds(expdata) <- "999999999"
-  experimentData(sm835Genes) <- expdata
-  save(sm835Genes, file=file.path("smutans", "data", "sm835Genes.RData"))
-}
-
-smutans.prepareDataOMZ175 <- function ()
-{
-  # Count data preparation
[…]

### 2012-03-05 package/smutans/R/methods.R

@@ -88,17 +88,35 @@ setMethod("smutans.de2",
       fullCountsTable <- counts( object@countDataSet )
       cdsFull <- newCountDataSet( fullCountsTable, design )
       cdsFull <- estimateSizeFactors( cdsFull )
+
       object@cds <- estimateDispersions( cdsFull, method=method )
+      # condAB <- unique(sort(pData(cdsFull)$condition))
+      bmv <- getBaseMeansAndVariances( counts(object@cds), sizeFactors(object@cds) )
       if (compareCondition == "yes") {
         fit1 <- fitNbinomGLMs( object@cds, count ~ type + condition )
         fit0 <- fitNbinomGLMs( object@cds, count ~ type  )
+        colA <- pData(object@cds)$condition == condA
+        colB <- pData(object@cds)$condition == condB
       } else {
         fit1 <- fitNbinomGLMs( object@cds, count ~ condition + type )
         fit0 <- fitNbinomGLMs( object@cds, count ~ condition )
+        colA <- pData(object@cds)$type == condA
+        colB <- pData(object@cds)$type == condB
       }
+      bmvA <- getBaseMeansAndVariances( counts(object@cds)[,colA], sizeFactors(object@cds)[colA] )
+      bmvB <- getBaseMeansAndVariances( counts(object@cds)[,colB], sizeFactors(object@cds)[colB] )
+      bmvBoverA <- bmvB$baseMean/bmvA$baseMean
+
       object@pval <- nbinomGLMTest( fit1, fit0 )
       object@padj <- p.adjust( object@pval, method="BH" )
-      object@res <- data.frame(fit1,pval=object@pval,padj=object@padj)
+      object@res <- data.frame(id=rownames(bmv),
+                               baseMean=bmv$baseMean,
+                               baseMeanA=bmvA$baseMean,
+                               baseMeanB=bmvB$baseMean,
+                               foldChange=bmvBoverA,
+                               log2FoldChange=log2(bmvBoverA),
+                               pval=object@pval,
+                               padj=object@padj, converged=fit1$converged)
     } else if (type!="" && condition=="") {
       ua159Samples <- pData(object@countDataSet)$type == type
       countsTable <- counts(object@countDataSet)[ , ua159Samples ]
@@ -265,19 +283,14 @@ setMethod("smutans.de2Clust",
 )
 
 setGeneric("smutans.de2List", 
-  function(object,file="default", qval=0.1) standardGeneric("smutans.de2List")
[…]

### 2012-03-05 package/test.R

@@ -12,6 +12,8 @@
 library(GenomicRanges)
 library(DESeq)
 source("smutans/R/core.R")
+source("smutans/R/methods.R")
+source("smutans/R/class_and_slots.R")
 #smutans.prepareGoseq( )
 #smutans.prepareData835NPP ()
 #smutans.prepareData34 ()
@@ -20,43 +22,137 @@ source("smutans/R/core.R")
 #smutans.prepareDataSMU86CSP( )
 #smutans.prepareDataUA159TW1()
 #smutans.prepareTranscript( )
-smutans.prepareDataSMU21PH75() 
-smutans.prepareDataUA159PH75()
+#smutans.prepareDataSMU21PH75() 
+#smutans.prepareDataUA159PH75()
+
+smutansGenes <- readSmutans (countsFile="smutans/inst/extdata/count-ua159-bwa.txt",
+             indexFile ="smutans/inst/extdata/count-ua159.txt.index",
+             condition = c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL"),
+             firstFactor = c("UA159GLU", "UA159GAL"),
+             firstFactorLabel = c("ua159", "tw1"),
+             secondFactor = c("UA159GLU", "TW1GLU"),
+             secondFactorLabel = c("glucose","galactose"),
+             name="S. mutans UA159, TW1, Glucose, and Galactose", 
+             lab="University of Florida, and Cornell University", 
+             contact="Drs. Robert Burne, Michael Stanhope, and Adam Siepel", 
+             title="Streptococcus mutans RNA-Seq Studies", 
+             url="http://www.ncbi.nlm.nih.gov/projects/geo/query/acc.cgi?acc=XXX", 
+             abstract="RNA-seq of 13 biological replicates from Streptococcus mutans",
+             pubmedid <- "999999999")
+
+sm835Genes <- 
+  readSmutans (countsFile="smutans/inst/extdata/count-ua159-bwa.txt",
+               indexFile ="smutans/inst/extdata/count-ua159.txt.index",
+                condition = c("835NP", "835P", "UA159noCSP"),
+                firstFactor = c("835NP", "835P", "UA159noCSP"),
+                firstFactorLabel = c("ua159"),
+                secondFactor = c(),
+                secondFactorLabel = c("835np","835p", "ua159nocsp"),
+                name="S. mutans 835NP, 835P, UA159 no CSP", 
[…]

### 2012-03-05 sh/batch2.sh

@@ -456,11 +456,18 @@ function copy-data {
 
   # Programs and scripts.
   cp -r \$PBS_O_WORKDIR/pl .
+  cp \$HOME/$SAMTOOLS samtools
+  cp \$HOME/$BWA bwa
+  cp \$HOME/$SUBREADBUILDINDEX subread-buildindex
+  cp \$HOME/$SUBREADALIGN subread-align
+
+  # All of the batchjob scripts.
+  cp \$PBS_O_WORKDIR/job-cram .
+  cp \$PBS_O_WORKDIR/job-cram2fastq .
+  cp \$PBS_O_WORKDIR/job-fastq-qc .
   cp \$PBS_O_WORKDIR/job-bwa-align .
-  cp ~/bin/samtools .
-  cp ~/bin/bwa .
-  cp ~/bin/subread-buildindex .
-  cp ~/bin/subread-align .
+  cp \$PBS_O_WORKDIR/job-de* .
+  cp \$PBS_O_WORKDIR/job-de*.R .
 
   # Create output directories at the compute node.
   mkdir -p $CDATADIR
@@ -470,22 +477,9 @@ function copy-data {
   # Copy common data
   cp $RDATADIR/$GENOMEFASTA $CDATADIR
   cp $RDATADIR/$GENOMEGFF $CDATADIR
-
-  ./bwa index -p $CBWADIR/$GENOMEFASTA-bwa -a is \\
-    $CDATADIR/$GENOMEFASTA
-  ./subread-buildindex -o \\
-    $CSUBREADDIR/$GENOMEFASTA-subread \\
-    $CDATADIR/$GENOMEFASTA
 }
 
 function retrieve-data {
-  # cp $CBWADIR/*.bam $RBWADIR
-  # cp $CBWADIR/*.pileup $RBWADIR
-  # cp $CBWADIR/*.wig $RBWADIR
-  # cp $CBWADIR/*-sum.pos $RBWADIR
-  # cp $CBWADIR/*-sum.rrna $RBWADIR
-  # cp $CBWADIR/*.sorted.bam $RBWADIR
-  # cp $CSUBREADDIR/*.sorted.bam $RSUBREADDIR
   echo No Copy!
 }
 
@@ -494,84 +488,27 @@ function process-data {
   CORESPERNODE=1
   FASTQFILES=( $FASTQFILES )
[…]

### 2012-04-05 package/manuscript.Rnw
Just an update.

@@
 ``Energy metabolism; Fermentation'' had highly differentially expressed genes
-(fig.\ \ref{figDotplotUA159Only})
-This was silmilar to the comparison between UA159
+(fig.\ \ref{figDotplotUA159Only}).
+This was similar to the comparison between UA159
 wild-type and TW1 mutant-type grown in glucose. 

### 2012-04-05 package/smutans/R/methods.R

@@ -124,6 +124,10 @@ setMethod("smutans.de2",
 
       cds <- newCountDataSet( countsTable, conds )
       cds <- estimateSizeFactors( cds )
+
+      # For no replicate case:
+      # object@cds <- estimateDispersions( cds,method="blind",sharingMode="fit-only")
+      
       object@cds <- estimateDispersions( cds )
       object@res <- nbinomTest( object@cds, condA, condB )
       object@pval <- object@res$pval
@@ -263,15 +267,15 @@ setMethod("smutans.de2Clust",
     heatmap( as.matrix( dists ),
        symm=TRUE, scale="none", margins=margins,
        col = colorRampPalette(c("black","white"))(100),
-       labCol = paste( pData(cdsFullBlind)$condition, pData(cdsFullBlind)$type),
-       labRow = paste( pData(cdsFullBlind)$condition, pData(cdsFullBlind)$type)
+       labCol = paste( rownames(pData(cdsFullBlind)), pData(cdsFullBlind)$condition, pData(cdsFullBlind)$type),
+       labRow = paste( rownames(pData(cdsFullBlind)), pData(cdsFullBlind)$condition, pData(cdsFullBlind)$type)
        )
 #    heatmap( as.matrix( dists ), Colv=TRUE,
 #       symm=TRUE, scale="none", margins=margins,
 #       col = colorRampPalette(c("black","white"))(100),
 #       trace="none", density.info="none",
-#       labRow = paste( pData(cdsFullBlind)$condition, pData(cdsFullBlind)$type),
-#       labCol = paste( pData(cdsFullBlind)$condition, pData(cdsFullBlind)$type),
+#       labRow = paste( rownames(pData(cdsFullBlind)), pData(cdsFullBlind)$condition, pData(cdsFullBlind)$type),
+#       labCol = paste( rownames(pData(cdsFullBlind)), pData(cdsFullBlind)$condition, pData(cdsFullBlind)$type),
 #       dendrogram = "both", keysize = 1.2
 #       )
     if (file != "default")

### 2012-04-05 package/test.R

@@ -9,11 +9,12 @@
 #                   cat.desc=smutans.cat.desc )
 
 
-library(GenomicRanges)
-library(DESeq)
+#library(GenomicRanges)
+#library(DESeq)
 source("smutans/R/core.R")
 source("smutans/R/methods.R")
 source("smutans/R/class_and_slots.R")
+
 #smutans.prepareGoseq( )
 #smutans.prepareData835NPP ()
 #smutans.prepareData34 ()
@@ -25,84 +26,18 @@ source("smutans/R/class_and_slots.R")
 #smutans.prepareDataSMU21PH75() 
 #smutans.prepareDataUA159PH75()
 
-smutansGenes <- readSmutans (countsFile="smutans/inst/extdata/count-ua159-bwa.txt",
-             indexFile ="smutans/inst/extdata/count-ua159.txt.index",
-             condition = c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL"),
-             firstFactor = c("UA159GLU", "UA159GAL"),
-             firstFactorLabel = c("ua159", "tw1"),
-             secondFactor = c("UA159GLU", "TW1GLU"),
-             secondFactorLabel = c("glucose","galactose"),
-             name="S. mutans UA159, TW1, Glucose, and Galactose", 
-             lab="University of Florida, and Cornell University", 
-             contact="Drs. Robert Burne, Michael Stanhope, and Adam Siepel", 
-             title="Streptococcus mutans RNA-Seq Studies", 
-             url="http://www.ncbi.nlm.nih.gov/projects/geo/query/acc.cgi?acc=XXX", 
-             abstract="RNA-seq of 13 biological replicates from Streptococcus mutans",
-             pubmedid <- "999999999")
+agGenes <- readSmutans(countsFile="../output/agalactiae/1/bwa/count-NC_004368.txt", 
+    indexFile="../output/agalactiae/1/run-analysis/count-hcmb.index",
+    condition = c("hm","cm","hb","cb"),
+    firstFactor = c("hm","hb"), 
+    firstFactorLabel = c("human", "cow"), 
+    secondFactor = c("hm", "cm"), 
+    secondFactorLabel = c("milk","broth"), 
+    name="S. agalctiae human, cow, milk, and broth")
 
-sm835Genes <- 
-  readSmutans (countsFile="smutans/inst/extdata/count-ua159-bwa.txt",
-               indexFile ="smutans/inst/extdata/count-ua159.txt.index",
[…]

### 2012-04-05 test/tomato/job-de-store.R

@@ -1,2 +1,108 @@
+save.image(file = "tomato.RData")
+load("tomato.RData")
+
 library(rtracklayer)
+library(GenomicRanges)
+library(VariantAnnotation)
+library("GenomicFeatures")
+library(TxDb.Hsapiens.UCSC.hg19.knownGene)
 gene.range3 <- import.gff3("/v4scratch/sc2265/rnaseq/output/tomato/1/data/ITAG2.3_gene_models.gff3.txt.no-negative")
+#load("gene.range3.Rd")
+vcffile <- "FASTQ001.var.flt.vcf.gz"
+tomatoFaFile <- FaFile( "/v4scratch/sc2265/rnaseq/output/tomato/1/data/S_lycopersicum_chromosomes.2.40.fna" )
+# tomatoFaFile <- FaFile( "S_lycopersicum_chromosomes.2.40.fna" )
+
+# makeTranscriptDb
+gr.mRNA <- gene.range3[gene.range3$type=="mRNA",]
+gr.exon <- gene.range3[gene.range3$type=="exon",]
+gr.cds <- gene.range3[gene.range3$type=="CDS",]
+
+# Find exons with proper CDS.
+grExon <- GRanges( seqnames = gr.exon$space, 
+               ranges = IRanges(start(gr.exon), end = end(gr.exon), names = gr.exon$ID), 
+               strand = strand(gr.exon), 
+               ID = gr.exon$ID, Parent = unlist(gr.exon$Parent) )
+grCDS <- GRanges( seqnames = gr.cds$space, 
+               ranges = IRanges(start(gr.cds), end = end(gr.cds), names = gr.cds$ID), 
+               strand = strand(gr.cds) )
+ol <- findOverlaps(grCDS,grExon,type="within")
+# Unique exons
+# Choose unique exons and CDS would be unique.
+head(unique(as.matrix(ol)[,2])) 
+x <- c()
+print(length(unique(as.matrix(ol)[,2])))
+j <- 0
+for (i in unique(as.matrix(ol)[,2])) {
+  j <- j + 1
+  x <- c(x, which(as.matrix(ol)[,2]==i)[1])
+  if (j %% 1000 == 0) {
+    print(j)
+  }
+}
+ol2 <- as.matrix(ol)[x,]
+rm (j,i,x)
+grExon.unique <- grExon[ol2[,2],]
+grCDS.unique <- grCDS[ol2[,1],]
+
+# tx_id of splicings
[…]

### 2012-04-05 package/manuscript.Rnw
just an update.

@@
 
-\fixme{Refinements: More filtering steps in the short read alignments might be necessary. Done?}
-
 \section*{Abstract}
@@
 \section*{Introduction}
-% S. mutans, CCR and CcpA
+% Biological motivations: S. mutans, CCR and CcpA
 About one fifth of children of ages ranging from 2 to 19 were reported to have
@@
 
-% RNA-seq, and microarray, bacterial transcriptomes
+% Technological motivations: RNA-seq, and microarray, bacterial transcriptomes
 It has been a decade since the first genome of the oral bacteria was finished
@@
 
-% Aims
+% Aims of the study
 We used the standard RNA deep sequencing technique to sequence RNAs in
@@
 different nutrients with UA159 wild-type and \textit{ccpA}-deleted TW1
-mutant-type.  We attempted to define transcripts of mRNAs using the RNA-seq
-data, and to find several small non-coding RNAs. Using the map of transcripts
+mutant-type.  We describe transcripts of the mRNAs using the RNA-seq
+data, and present several small non-coding RNAs in \textit{S.\ mutans}. 
+Using the map of transcripts
 and expression levels of the transcripts from different biological conditions
@@
 \subsection*{RNA sequencing}
-\fixme{SCC wrote the paragraph based on email correspondences between PDPB and CDD.
-Stanhope group describes RNA sequencing. Please add texts and SCC will copy-edit
-them.}
 Deep sequencing of the RNA samples were performed at the Cornell University Life

### 2012-04-05 R/gff2txdb.R

@@ -1,0 +1,91 @@
+library(rtracklayer)
+library(GenomicRanges)
+library(VariantAnnotation)
+library(GenomicFeatures)
+gffFile <- "/Volumes/Elements/Documents/Projects/mauve/bacteria/Streptococcus_mutans_UA159_uid57947/NC_004350.gff"
+sm.gff <- import.gff3(gffFile)
+sm.source <- sm.gff[sm.gff$type=="source",]
+sm.gene <- sm.gff[sm.gff$type=="gene",]
+sm.exon <- sm.gff[sm.gff$type=="exon",]
+sm.CDS <- sm.gff[sm.gff$type=="CDS",]
+
+transcripts <- 
+  data.frame( tx_id=seq(length(sm.gene$locus_tag)),
+              tx_name=sm.gene$locus_tag,
+              tx_chrom=sm.gene$space, 
+              tx_strand=sm.gene$strand,
+              tx_start=start(sm.gene),
+              tx_end=end(sm.gene) )
+
+# Compare CDS and gene locus tags
+x <- c()
+y <- c()
+for (i in sm.gene$locus_tag) {
+  if (sum(sm.CDS$locus_tag == i) > 0) {
+    x <- c(x,start(sm.CDS)[sm.CDS$locus_tag == i])
+    y <- c(y,end(sm.CDS)[sm.CDS$locus_tag == i])
+  } else {
+    x <- c(x,NA)
+    y <- c(y,NA)
+  }
+}
+grCDS.start <- x
+grCDS.end <- y
+rm(i,x,y)
+
+# 
+splicings <- 
+  data.frame( tx_id=seq(length(sm.gene$locus_tag)),
+              exon_rank=seq(length(sm.gene$locus_tag)),
+              exon_start=start(sm.gene),
+              exon_end=end(sm.gene),
+              exon_name=sm.gene$locus_tag,
+              cds_start=grCDS.start,
+              cds_end=grCDS.end )
+
+chrominfo <-
+  data.frame( chrom=names(sm.gene),
+              length=end(sm.source),
+              is_circular=TRUE )
+
+TxDb.Smutans.UA159.uid57947.knownGene <- makeTranscriptDb (transcripts, splicings, chrominfo=chrominfo)
+txdb <- TxDb.Smutans.UA159.uid57947.knownGene 
+
+cds(txdb, columns="tx_name")
+cds(txdb, columns="exon_name")
+
+# Read short reads
+library(ShortRead)
+aln <- readAligned(bam.file,type="BAM")
[…]

### 2012-04-12 package/manuscript.Rnw
Just an update.

@@
 
+\section*{Main findings}
+The resolution of RNA-Seq has advanced many aspects of the existing gene
+annotation, including the determination of gene boundaries for known genes and
+the identification of novel transcribed regions in many of the bacteria that
+have previously been sequenced.
+For example, for L. monocytogenes, 517 primary operons were identified that
+encode 1719 genes, which represent 60% of the annotated genome
+The knowledge gained regarding untranslated regulatory regions, operons and
+sRNAs is triggering re-evaluation and re-annotation of many bacterial genomes.
+
 \section*{Introduction}
@@
 \subsection*{Short reads alignments}
-About 20 millions of short reads were sequenced for each sample.  We used a
+About 20 millions of short reads were sequenced for each sample 
+(table \ref{count}).  We used a
 short read aligner, \program{BWA} \citep{Li2009a}, to map the sequenced short
@@
 matchUnannotatedRNA <- findOverlaps(grRNAz,grNoAnnotation,type="within")
-txRNAz.unannotated <- grRNAz[matchMatrix(matchUnannotatedRNA)[,1]]
+txRNAz.unannotated <- grRNAz[as.matrix(matchUnannotatedRNA)[,1]]
 
@@
 # Expressed transcripts without annotated genes
-grTxWGenes <- grTx[unique(matchMatrix(matchGenesTx)[,2])]
-grTxWOGenes <- grTx[-unique(matchMatrix(matchGenesTx)[,2])]
-grTxWOAnnotation <- grTx[-unique(matchMatrix(matchAnnotationTx)[,2])]
+grTxWGenes <- grTx[unique(as.matrix(matchGenesTx)[,2])]
+grTxWOGenes <- grTx[-unique(as.matrix(matchGenesTx)[,2])]
+grTxWOAnnotation <- grTx[-unique(as.matrix(matchAnnotationTx)[,2])]
 
@@
 #mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
-#txWithConflictedGenes <- intersect(matchMatrix(mtchPlus)[,2],matchMatrix(mtchMnus)[,2])
+#txWithConflictedGenes <- intersect(as.matrix(mtchPlus)[,2],as.matrix(mtchMnus)[,2])
 
@@
 matchGenesTx2 <- findOverlaps(smutansData.txGenes,grTx2,type="within")
-number.genes.tx <- runLength(Rle(matchMatrix(matchGenesTx2)[,"subject"]))
+number.genes.tx <- runLength(Rle(as.matrix(matchGenesTx2)[,"subjectHits"]))
 
@@
 length3UTR <- c()
-for (i in unique(matchMatrix(mtchPlus)[,"subject"])) {
-  x <- matchMatrix(mtchPlus)[matchMatrix(mtchPlus)[,"subject"]==i,"query"]
+for (i in unique(as.matrix(mtchPlus)[,"subjectHits"])) {
+  x <- as.matrix(mtchPlus)[as.matrix(mtchPlus)[,"subjectHits"]==i,"queryHits"]
   y <- min(start(grGenesPlus[x])) - start(grTx2[i])
@@
 }
-for (i in unique(matchMatrix(mtchMnus)[,"subject"])) {
-  x <- matchMatrix(mtchMnus)[matchMatrix(mtchMnus)[,"subject"]==i,"query"]
+for (i in unique(as.matrix(mtchMnus)[,"subjectHits"])) {
+  x <- as.matrix(mtchMnus)[as.matrix(mtchMnus)[,"subjectHits"]==i,"queryHits"]
   y <- min(start(grGenesMnus[x])) - start(grTx2[i])
@@
 mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
-x <- rle(matchMatrix(mtch3)[,"subject"])
+x <- rle(as.matrix(mtch3)[,"subjectHits"])
 comp.exp.two.genes <- c()
 for (i in x$values[x$lengths > 1]) {
-  y <- combn(matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==i],2)
+  y <- combn(as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==i],2)
   for (j in 1:dim(y)[2]) {
@@
   }
-  y1 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i]]
+  y1 <- as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==x$values[i]]
   i2 <- i + 1
-  y2 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i2]]
+  y2 <- as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==x$values[i2]]
   l <- as.matrix(do.call(expand.grid, list(y1,y2)))
@@
 Of the total \Sexpr{length(smutansData.txGenes)} of genes,
-\Sexpr{length(matchMatrix(matchGenesTx)[,1])}
-(\Sexpr{as.integer(length(matchMatrix(matchGenesTx)[,1])/length(smutansData.txGenes)*100)}\%) 
+\Sexpr{length(as.matrix(matchGenesTx)[,1])}
+(\Sexpr{as.integer(length(as.matrix(matchGenesTx)[,1])/length(smutansData.txGenes)*100)}\%) 
 were designated as expressed. 
@@
 (fig.\ \ref{figExpressionTwoGene}).
-Among the \Sexpr{length(unique(matchMatrix(matchGenesTx2)[,"subject"]))} segments
+Among the \Sexpr{length(unique(as.matrix(matchGenesTx2)[,"subjectHits"]))} segments
 with genes of all of the same strands, 

### 2012-04-12 package/supplementary.Rnw
Just an update.

--- 2012-02-03 package/supplementary.Rnw
@@
 load("smutans/data/smutansGenes.RData")
+load("smutans/data/tw1.summary.stat.xtable.RData")
 @
@@
 data(smutans.keg.desc)
-data(smutansGenes)
 data(smutansGenes2)
+data(tw1.summary.stat.xtable)
 @
@@
 %\clearpage{}
+
+<<echo=FALSE,results=tex>>=
+# lab:count
+print( tw1.summary.stat.xtable,
+       caption.placement="top",
+       include.rownames=FALSE )
+@
+
 <<echo=FALSE,results=tex>>=
@@
 mtchUnannotatedRNA <- findOverlaps(ir2,ir1,type="within")
-txRNAz.unannotated <- ir2[matchMatrix(mtchUnannotatedRNA)[,1]]
+txRNAz.unannotated <- ir2[as.matrix(mtchUnannotatedRNA)[,1]]
 t1 <- data.frame(names=names(txRNAz.unannotated),start=start(txRNAz.unannotated),end=end(txRNAz.unannotated))
@@
 <<figLength,fig=TRUE,echo=FALSE>>=
-l <- list(expressed=width(smutansData.txGenes[matchMatrix(matchGenesTx)[,1]]), 
-          notexpressed=width(smutansData.txGenes[-matchMatrix(matchGenesTx)[,1]]))
+l <- list(expressed=width(smutansData.txGenes[as.matrix(matchGenesTx)[,1]]), 
+          notexpressed=width(smutansData.txGenes[-as.matrix(matchGenesTx)[,1]]))
 bin <- c(as.integer(seq.int(0, 3000, length.out=15)), 8500)
@@
 xm[,"Functional Class"] <- unlist(z)
-# Use xm to plot a dotplot.
-dotplot(xm[,"Functional Class"]~xm[,"fold"],
+# Use xm to plot a dotchart.
+dotchart(xm[,"fold"],labels=xm[,"Functional Class"],
         main="",xlab="Log2 fold change",col="black",pch=88)

### 2012-04-12 R/deseq-cornell.R

@@ -1,0 +1,116 @@
+# Rscript R/deseq.R count.txt species conditionA conditionB
+.Last <- function() {
+  cat("\n")
+}
+args <- commandArgs(trailingOnly = TRUE)
+if (length(args) != 5)
+{
+  cat ("Rscript R/deseq.R count.txt species conditionA conditionB\n")
+  quit("yes")
+}
+countsFile <- args[1]
+speciesName <- args[2]
+subcondA <- args[3]
+subcondB <- args[4]
+cutoffpval <- as.numeric(args[5])
+
+library(DESeq)
+conds <- scan(file=paste(countsFile,"index",sep="."), what="character")
+# countsFile <- "output/cornell/1/bwa/count.txt"
+# conds <- c("UA159GLU", "TW1GLU", "TW1GAL", "OMZ175", "OMZHKRR", "OMZ175", "OMZ175", "OMZHKRR", "OMZHKRR", "UA159GLU", "UA159GLU", "UA159GAL", "UA159GAL", "TW1GLU", "TW1GLU", "TW1GAL", "TW1GAL", "UA159GAL", "UA159GAL", "835NP", "835NP", "835NP", "835P", "835P", "835P", "UA159noCSP", "UA159noCSP", "UA159CSP", "UA159CSP", "UA159CSP", "Smu86noCSP", "Smu86noCSP", "Smu86CSP", "Smu86CSP")
+# subcondA <- "UA159GLU"
+# subcondB <- "UA159GAL"
+# speciesName <- "cornell"
+subconds <- conds %in% c(subcondA, subcondB) 
+comparisonName <- paste(speciesName, subcondA, subcondB, sep="-")
+dePSFile <- paste("output/", speciesName, "/1/bwa/", comparisonName, "-de-", cutoffpval, ".ps", sep="")
+deSCVFile <- paste("output/", speciesName, "/1/bwa/", comparisonName, "-scv-", cutoffpval, ".ps", sep="")
+
+countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)
+rownames(countsTable) <- countsTable$gene
+countsTable <- countsTable[,-1]
+y <- length(countsTable[,1])
+cat ("Total number of genes is ", y, "\n",sep="")
+removedCountsTable <- countsTable[rowSums(countsTable[,subconds]) <= sum(subconds)*2.5,]
+y <- length(removedCountsTable[,1])
[…]

### 2012-04-12 R/deseq-omz-omz.R

@@ -1,0 +1,64 @@
+library(DESeq)
+cutoffpval <- 0.05
+countsFile <- "output/omz/1/bwa/count.txt"
+countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)
+rownames(countsTable) <- countsTable$gene
+countsTable <- countsTable[,-1]
+y <- length(countsTable[,1])
+cat ("Total number of genes is ", y, "\n",sep="")
+
+removedCountsTable <- countsTable[countsTable[,1]+countsTable[,2]+countsTable[,3]+countsTable[,4]+countsTable[,5]+countsTable[,6]<=15,]
+y <- length(removedCountsTable[,1])
+cat ("Number of genes removed is ", y, "\n",sep="")
+countsTable <- countsTable[countsTable[,1]+countsTable[,2]+countsTable[,3]+countsTable[,4]+countsTable[,5]+countsTable[,6]>15,]
+
+conds <- c("OMZ175", "OMZHKRR", "OMZ175", "OMZ175", "OMZHKRR", "OMZHKRR")
+
+cds <- newCountDataSet(countsTable, conds)
+cds <- estimateSizeFactors(cds)
+
+cds <- estimateVarianceFunctions(cds)
+res <- nbinomTest(cds, "OMZ175", "OMZHKRR")
+# 1 3 4 vs 2 5 6
+# 1  "OMZ175"
+# 2  "OMZHKRR"
+# 3  "OMZ175"
+# 4  "OMZ175"
+# 5  "OMZHKRR"
+# 6  "OMZHKRR"
+      
+plotDE <- function( res )
+{
+   plot(res$baseMean, res$log2FoldChange, log="x", pch=20, cex=.5, 
+        col = ifelse( res$padj < cutoffpval, "red", "black" ) )
+}
+
+cat ("SCV file: omz-de.ps\n")
+postscript ("output/omz/1/bwa/omz-de.ps",  width=10, height=10, horizontal = FALSE, onefile = FALSE,
+paper = "special")
+plotDE(res)
+y <- dev.off()
+
+cat ("SCV file: omz-scv.ps\n")
+postscript ("output/omz/1/bwa/omz-scv.ps",  width=10, height=10, horizontal = FALSE, onefile = FALSE,
+paper = "special")
+scvPlot(cds)
+y <- dev.off()
+
+resSig <- res[res$padj < cutoffpval,]
+y <- length(countsTable[,1])
[…]

### 2012-04-12 R/job-extract-gene.R

@@ -1,0 +1,28 @@
+############################################################################
+# Create a test fastq files by sampling 10 100-bp short reads for each gene.
+library(easyRNASeq)
+library(rtracklayer)
+gff.file <- "/v4scratch/sc2265/rnaseq/output/omz/1/data/SMU109.gff"
+genome.file <- "/v4scratch/sc2265/rnaseq/output/omz/1/data/SMU109-OMZ175.fa"
+fasta.file <- "SMU109-gene.fa"
+file.create(fasta.file)
+
+gene.range3 <- import.gff3(gff.file)
+gene.range1 <- gene.range3[gene.range3$type=="gene",]
+gene.range2 <- gene.range1[grep("SMU[rt]", gene.range1$locus_tag, invert=TRUE),]
+
+s.mutans.sequence <- read.DNAStringSet(genome.file)
+chrom.list <- names(s.mutans.sequence)
+
+for (i in seq(nrow(gene.range2))) {
+  j <- as.character(gene.range2[i,]$space)
+  x <- start(gene.range2[i,]$ranges)
+  y <- end(gene.range2[i,]$ranges)
+  s.mutans <- s.mutans.sequence[[j]]
+  gene.seq <- subseq(s.mutans,start=x,end=y)
+  if (gene.range2[i,]$strand == '-') {
+    gene.seq <- reverseComplement(gene.seq)
+  }
+  oneRead <- sprintf(">%s\n%s\n",gene.range2[i,]$locus_tag,toString(gene.seq))
+  cat(oneRead, file=fasta.file, append=TRUE)
+}

### 2012-04-12 R/run.sh

@@ -1,0 +1,137 @@
+# Three global variables for a reference genome
+SPECIES=cornell
+COUNTFILE=output/$SPECIES/1/bwa/count.txt
+
+# for PVALUE in 0.01 0.001 0.0001 0.00001 0.000001; do
+for PVALUE in 0.000001; do
+
+CONDDITIONA=835NP
+CONDDITIONB=835P
+Rscript R/deseq-cornell.R \
+  $COUNTFILE $SPECIES $CONDDITIONA $CONDDITIONB $PVALUE \
+  > output/$SPECIES/1/bwa/$SPECIES-$CONDDITIONA-$CONDDITIONB-$PVALUE.deseq 
+
+done
+exit
+
+############################
+# UA159, TW1, GLU, and GAL
+function run-ua159-tw1-glu-gal {
+  SPECIES=cornell
+  COUNTFILE=output/$SPECIES/1/bwa/count.txt
+  CONDDITIONA=UA159GLU
+  CONDDITIONB=UA159GAL
+  Rscript R/deseq-cornell.R \
+    $COUNTFILE $SPECIES $CONDDITIONA $CONDDITIONB $PVALUE \
+    > output/$SPECIES/1/bwa/$SPECIES-$CONDDITIONA-$CONDDITIONB.deseq 
+
+  CONDDITIONA=TW1GLU
+  CONDDITIONB=TW1GAL
+  Rscript R/deseq-cornell.R \
+    $COUNTFILE $SPECIES $CONDDITIONA $CONDDITIONB $PVALUE \
+    > output/$SPECIES/1/bwa/$SPECIES-$CONDDITIONA-$CONDDITIONB.deseq 
+
+  CONDDITIONA=UA159GLU
+  CONDDITIONB=TW1GLU
+  Rscript R/deseq-cornell.R \
+    $COUNTFILE $SPECIES $CONDDITIONA $CONDDITIONB $PVALUE \
+    > output/$SPECIES/1/bwa/$SPECIES-$CONDDITIONA-$CONDDITIONB.deseq 
+
+  CONDDITIONA=UA159GAL
+  CONDDITIONB=TW1GAL
+  Rscript R/deseq-cornell.R \
+    $COUNTFILE $SPECIES $CONDDITIONA $CONDDITIONB $PVALUE \
[…]

### 2012-04-12 pl/orthoparahomlist.pl

@@ -1,0 +1,181 @@
+#!/usr/bin/perl
+# Compare gene sets of two species with BLAST and compile lists with
+# orthologs, paralogs and other homologs
+# Mario Stanke, 6.6.2011
+
+use strict;
+use warnings;
+
+use Getopt::Long; # for parameter specification on the command line
+use Bio::Tools::Run::StandAloneBlast;
+
+my $usage = <<'ENDUSAGE';
+orthoparahomlist.pl       compare gene sets of two species with BLAST and
+                          compile lists with orthologs, paralogs and other homologs
+
+SYNOPSIS
+
+orthoparahomlist.pl --protA=proteinsA.fa --protB=proteinsB.fa 
+
+    protA and protB specify a set of genes from two different species A and B, respectively.
+    Every protein sequence is compared to every other protein sequence with BLASTP. This 
+    script requires an installation of NCBI blastall and that databases have been created with
+    fomatdb.
+    
+OPTIONS
+
+    --help             output this help message
+    --ortho=ortho.lst  Create an output file ortho.lst that contains for each gene a in A the
+                       putative ortholog b in B as inferred by reciprocal best BLAST hit. List
+                       only genes which have a reciprocal best hit. Format:
+                       geneA.id <TAB> geneB.id
+    --para=para.lst    Create an output file para.lst that contains for each gene a in A all other
+                       genes in A that meet the E-value threshold. Include the E-value of the hit 
+                       and the percentage of identical amino acids of the first HSP. Format:
+                       geneA.id1 <TAB> geneB.id1 <TAB> E-value <TAB> percent identical
+                       geneA.id1 <TAB> geneB.id2 <TAB> E-value <TAB> percent identical
+                       geneA.id1 <TAB> geneB.id3 <TAB> E-value <TAB> percent identical
+                       ...
+                       geneA.id1 <TAB> geneB.idM <TAB> E-value <TAB> percent identical
+                       geneA.id2 <TAB> geneB.id1 <TAB> E-value <TAB> percent identical
+                       geneA.id2 <TAB> geneB.id2 <TAB> E-value <TAB> percent identical
+                       ...
[…]

### 2012-04-12 sh/batch2.sh

@@ -865,8 +865,8 @@ txdb <- loadFeatures(txdb.file)
 
 # 1. Total number of short reads in fastq: zcat FASTQ051.fq.gz | wc -l
 indexBam(bam.file)
-bam.what <- c("qname","flag","pos","mapq","cigar")
-bvAll <- scanBam(bam.file, param=ScanBamParam(what=bam.what))
+#bam.what <- c("qname","flag","pos","mapq","cigar")
+#bvAll <- scanBam(bam.file, param=ScanBamParam(what=bam.what))
 bv <- readBamGappedAlignments(bam.file,use.names=TRUE,param=ScanBamParam(what=c("mapq")))
 # 2. Number of short reads mapped 
 # > length(bv)
@@ -952,8 +952,13 @@ fastQIndex <- scan(args[2])
 for (i in fastQIndex) {
   cl.file <- sprintf("%s/FASTQ%03d.sorted.bam.cl", args[1], i)
   load(cl.file)
+  num.total.read <- as.integer(num.total.read)
   table.read.statistics <- 
-    rbind(table.read.statistics, c(i, num.total.read, num.mapped.read, num.unique.read, sum(cl), sum(cl.nc)))
+    rbind(table.read.statistics, 
+    c(i, num.total.read, sprintf("%d (%d)",num.mapped.read, round(num.mapped.read/num.total.read*100)), 
+                         sprintf("%d (%d)",num.unique.read, round(num.unique.read/num.total.read*100)),
+			 sprintf("%d (%d)",sum(cl), round(sum(cl)/num.total.read*100)), 
+			 sprintf("%d (%d)",sum(cl.nc), round(sum(cl.nc)/num.total.read*100))))
   count.table <- data.frame(count.table,cl)
   colnames(count.table)[ncol(count.table)] <- paste("X",i,sep="")
 }
@@ -961,16 +966,16 @@ colnames(count.table) <- sub("X","",colnames(count.table))
 count.table.file <- sprintf("%s/count.txt", args[1])
 write.table(count.table,file=count.table.file,quote=FALSE,sep="\\t",row.names=FALSE)
 
-colnames(table.read.statistics) <- c("Sample ID", "Total number of reads", "Mapped reads", "Unique reads", "Reads on CDS regions", "Reads on noncoding regions")
+colnames(table.read.statistics) <- c("Sample ID", "Number of reads", "Mapped reads", "Unique reads", "Reads on CDS", "Reads on noncoding regions")
 x.big <- xtable( as.data.frame(table.read.statistics),
[…]

### 2012-04-12 sh/cram-smutans.sh

@@ -1,0 +1,54 @@
+#!/bin/bash -x
+
+# FASTQ=FASTQ001.fq
+FASTQ=test.fq
+REFFASTA=NC_004350.fa
+
+BWA_BIN=$HOME/bin/bwa
+SAMTOOLS_BIN=$HOME/bin/samtools
+CRAM_BIN=$HOME/usr/bin/cramtools.jar
+PICARD_SAMTOFASTAQ=$HOME/usr/bin/picard-tools-1.62/SamToFastq.jar 
+PICARD_VALIDATESAMFILE=$HOME/usr/bin/picard-tools-1.62/ValidateSamFile.jar
+PICARD_READGROUP=$HOME/usr/bin/picard-tools-1.62/AddOrReplaceReadGroups.jar
+
+CPU=8
+
+java -Xmx10g -jar \
+  $PICARD_SAMTOFASTAQ \
+  INPUT=$FASTQ.readgroup.bam \
+  FASTQ=$FASTQ.recovered.fq
+exit
+
+# $BWA_BIN index $REFFASTA
+$BWA_BIN aln -I -t $CPU $REFFASTA $FASTQ > $FASTQ.sai
+$BWA_BIN samse $REFFASTA $FASTQ.sai $FASTQ | $SAMTOOLS_BIN view -Sb - > $FASTQ.bam
+# $SAMTOOLS_BIN faidx $REFFASTA
+
+$SAMTOOLS_BIN sort $FASTQ.bam $FASTQ.sorted
+$SAMTOOLS_BIN index $FASTQ.sorted.bam
+
+java -jar $CRAM_BIN cram \
+  --input-bam-file $FASTQ.sorted.bam \
+  --reference-fasta-file $REFFASTA \
+  --output-cram-file $FASTQ.cram
+
+$SAMTOOLS_BIN view -h $FASTQ.sorted.bam -o $FASTQ.sam
+
+java -Xmx2g -jar \
+  $PICARD_READGROUP \
+  RGLB=testrglb \
+  RGPL=illumina \
+  RGPU=testrgpu \
+  RGSM=testrgsm \
+  INPUT=$FASTQ.sam \
+  OUTPUT=$FASTQ.readgroup.bam
+
+java -Xmx2g -jar \
+  $PICARD_VALIDATESAMFILE \
+  INPUT=$FASTQ.readgroup.bam
+
+java -Xmx10g -jar \
[…]

### 2012-04-13 package/manuscript.Rnw
QC first and then CRAM. We use CRAM without quality scores for subsequent analysis.

@@
 \maketitle
+\thispagestyle{empty}
 
@@
 
-\thispagestyle{empty}
 \end{titlepage}
+
+\setcounter{page}{2}
 
@@
 
-\section*{Abstract}
+\section*{ABSTRACT}
+
+\fixme{Add more specific findings while summarizing the results. Connect two
+ideas of the differential expression study and transcriptome study.}
 A bacterial transcriptome of the primary etiological agent of human dental
 caries, \textit{Streptococcus mutans}, is described using deep RNA sequencing.
-RNA transcripts inferred from the RNA-seq data revealed novel messenger RNAs,
-and small RNAs. Differential expression studies on the role of CcpA in 
+The technique of RNA-Seq has allowed 
+the determination of transcript boundaries for annotated genes and
+the identification of previously unknown transcribed regions 
+including mRNAs and small noncoding RNAs in the genome.
+Differential expression studies on the role of CcpA in 
 carbon catabolite repression (CCR) suggested that some of the novel mRNAs and
@@
 different carbon sources.
+The knowledge that we gained from the current study of \textit{S. mutans}
+transcriptome will lead to re-evaluation of the bacterial genome.
 
-\section*{Main findings}
-The resolution of RNA-Seq has advanced many aspects of the existing gene
-annotation, including the determination of gene boundaries for known genes and
-the identification of novel transcribed regions in many of the bacteria that
-have previously been sequenced.
-For example, for L. monocytogenes, 517 primary operons were identified that
-encode 1719 genes, which represent 60% of the annotated genome
-The knowledge gained regarding untranslated regulatory regions, operons and
-sRNAs is triggering re-evaluation and re-annotation of many bacterial genomes.
+\section*{INTRODUCTION}
 
-\section*{Introduction}
 % Biological motivations: S. mutans, CCR and CcpA
@@
 
-\section*{Materials and Methods}
-
-\subsection*{Bacterial strains and growth conditions}
-\fixme{Burne group describes bacterial strains and growth conditions and other
-necessary works. Please, add texts and SCC will copy-edit them.}
-
-\subsection*{RNA isolation}
-\fixme{Burne group describes RNA isolations and other works. Please, add texts
-and SCC will copy-edit them.}
-
-\subsection*{RNA sequencing}
-Deep sequencing of the RNA samples were performed at the Cornell University Life
-Sciences Core Laboratories Center (Ithaca, NY).  RNA extraction was performed
-using a modified Qiagen RNAease protocol (Qiagen, Inc., Chatsworth, CA): cells
-were pelleted, resuspended in RNA protect reagent, frozen, and then resuspended
-in TE, mechanically disrupted in a bead beater in the presence of glass beads
-and acidic phenol. We used the Qiagen kit to obtain total RNAs.  The RNA sample
-was passed through the Ambion MicrobExpress kit twice, which removed ribosomal
-RNAs only using magnetic beads.  This procedure left small RNAs and transfer
-RNAs intact.  Then, we followed the TruSeq Illumina kit to prepare the cDNA
-libraries.
-
-\subsection*{Short reads alignments}
-About 20 millions of short reads were sequenced for each sample 
-(table \ref{count}).  We used a
-short read aligner, \program{BWA} \citep{Li2009a}, to map the sequenced short
-reads on the reference genome, \textit{S.\ mutans} UA159 (NCBI accession
-NC\_004350). Because the aligner allowed a few gaps for efficient alignments of
-millions of sequences of size being 100 base pairs long, short reads that
-consisted of RNAs with sequencing adapter sequences would not be mapped.  We
-used \program{cutadapt} \citep{Martin2011} to remove parts of adapter sequences
-from short reads before mapping on the reference genome.  Mapped short read
-alignments were converted into readable formats using \program{samtools}
-\citep{Li2009b}.
-
-\subsection*{Transcript predictions}
-RNA transcripts were inferred by applying a hidden Markov model to site-wise
-expression levels \citep{Martin2010}.  We used the pileup command of
-\program{BWA} to convert the short reads alignments to pilup values, which were
-taken as the site-wise expression levels along the genome.  Site-wise expression
-levels are a list of site-wise non-negative integers that represents numbers of
-short reads mapped to a particular genomic position.  We used the following
-options for the transcript inference program, \program{ParseRNAseq}: ``-c 10 -b
-25 -force\_gp'' to bin expression levels into 25 parts, and to allow 10 emission
-states for relative expression intensity. We also used the genes as the
-predicted parts of transcripts.    Because we enriched mRNAs in the cDNA
-preparation, we intented to group annotated genes to segments, which could be
-transcripts. We did not aim at finding precise transcription start or end sites.
-Because we did not have strand-specific RNA-seq of the samples due to limited
-resources, we could not tell in which directions transcripts were oriented
-solely based on RNA-seq data. Instead, we used strands of annotated genes to
-determine strands of predicted transcripts.
-
-\subsection*{Prediction of small RNAs and targets}
-Although our RNA-seq data did not enriched non-coding small RNAs in the cDNA
-preparation, we retained small RNAs in the sample by attempting to deplete only
-ribosomal RNAs from the sample. Because non-coding small RNAs and coding
-mRNAs were mixed in the sample, it was difficult to discriminate small
-non-coding RNAs from mRNAs unless non-coding RNAs expressions were masked by
-neighboring background RNA expressions. 
-Therefore, we sought a way of predicting small RNAs
-using \program{RNAz} \citep{Gruber2010}, which  used homologous sequences and
-RNA secondary structures to predict putative non-coding RNAs. Because sequence
-alignments were a critical step for finding small RNAs in the approach, we
-collated sequences similar to intergenic regions, and refined the alignments using
-a multiple sequence alignment tool. Exracting
-intergenic regions including up- and down-stream, we searched a database of
-bacterial genomes using BLAST \citep{Altschul1990} for similar intergenic
-sequences, subsequently, refining the sequence alignments using MUSCLE
-\citep{Edgar2004a}.  Then, we applied \program{RNAz} to the alignments for
-scoring regions for small RNAs \citep{Washietl2005,Gruber2010}.  Targets of
-transcripts with high scores for small RNAs were also predicted using
-\program{RNAplex} \citep{Tafer2008}, and \program{RNAplfold}
-\citep{Bernhart2006}.  Target genes were scored for each small RNA candidate,
-which was used to perform functional category enrichment tests of targets for
-each small RNA candidate.
-
-\subsection*{Experimental validation of small RNAs}
-\fixme{Burne group might have done or be able to do some experimental
-validations of putative non-coding RNAs shown at table \ref{tabUnannotatedRNA}
-\citep[e.g.,][]{Altuvia2007}.}
-
-\subsection*{Statistical analysis for differential expressions}
-We used \Rpackage{DESeq} \citep{Anders2010} to determine differentially
-expressed genes based on negative binomial model originally developed by
-\citet{Robinson2007}.  Steps of analyzing RNA-seq data for determining
-differentially expressed genes were described in \cite{Oshlack2010}.  We counted
-short reads aligned to annotated genes in the reference genome, forming a table
-of read counts of the genes. Statistical software R \citep{R2011} with
-\Rpackage{DEseq} was subsequently employed to infer differentially expressed
-genes in the different biological conditions.
-
-\subsection*{Gene category associations}
-We compiled three sets of gene categories for testing functional associations.
-Firstly, we assigned genes to Gene Ontology (GO) categories by comparing the
-\textit{S.\ mutans} genes to bacterial proteins from the Uniref90 database using
-\program{BLASTP}, and then assigning the same GO classification as the target
-gene of the uniProt GOA database if the match had an $E$-value of
-$<1.0\times10^{-5}.$  A gene family was assigned a given classification if any
-of its genes was assigned that classification.  Secondly, we collated functional
-classes of genes in the Oral Pathogen Sequence Databases available at
-\url{http://oralgen.lanl.gov}. Thirdly, we used a set of metabolite pathways for
-\textit{S.\ mutans} available at Kyoto Encyclopedia of Genes and Genomes (KEGG)
-\citep{Kanehisa2012}.  Using the sets of gene categories for \textit{S.\ mutans}
-we performed a Mann-Whitney $U$ test of the values of target genes of small RNAs
-associated with a given category versus those for the other categories. We
-used a variation of Fisher's exact test accounting for gene lengths to test
-enrichments of differentially expressed genes \citep{Young2010}.  Both of the
-tests were corrected for multiple testing hypotheses \citep{Benjamini1995}.
-
-\subsection*{UCSC genome browser tracks and data availability}
-We created tracks for our recently released Streptococcus Genome Browser
-(\url{http://strep-genome.bscb.cornell.edu}) that summarized the results of our
-\textit{S.\ mutans} transcriptome analysis with known genes, gene expression
-levels based on the short read alignments, predicted putative transcript
-segments, and predictd small non-coding RNAs.  They should be accessed by clade
-\textit{Streptococcus}, and genome \textit{S. mutans UA159}.  These tracks can
-be used to inspect loci of interest and to compare the results of different
-RNA-seq data sets.  They can also be queried and intersected with other tracks
-using the UCSC Table Browser.
-
-\section*{Results}
+\section*{RESULTS}
 
@@
 
-\subsection*{Results from Stanhop group when RNA sequencing preparation}
+\subsection*{Results from Stanhope group when RNA sequencing preparation}
 \fixme{Add some if needed.}
 
-\subsection*{Statistics of short reads}
+\subsection*{Cluster of RNA samples}
 
-\subsection*{Cluster of RNA samples}
 First using the table of read counts of annotated genes we formed a cluster of
@@
 
-\section*{Discussion}
+\section*{DISCUSSION}
 
@@
 
-\section*{Acknowledgments}
+\section*{METHODS}
+
+\subsection*{Bacterial strains and growth conditions}
+\fixme{Burne group describes bacterial strains and growth conditions and other
+necessary works. Please, add texts and SCC will copy-edit them.}
+
+\subsection*{RNA isolation}
+\fixme{Burne group describes RNA isolations and other works. Please, add texts
+and SCC will copy-edit them.}
+
+\subsection*{RNA sequencing}
+Deep sequencing of the RNA samples were performed at the Cornell University Life
+Sciences Core Laboratories Center (Ithaca, NY).  RNA extraction was performed
+using a modified Qiagen RNAease protocol (Qiagen, Inc., Chatsworth, CA): cells
+were pelleted, resuspended in RNA protect reagent, frozen, and then resuspended
+in TE, mechanically disrupted in a bead beater in the presence of glass beads
+and acidic phenol. We used the Qiagen kit to obtain total RNAs.  The RNA sample
+was passed through the Ambion MicrobExpress kit twice, which removed ribosomal
+RNAs only using magnetic beads.  This procedure left small RNAs and transfer
+RNAs intact.  Then, we followed the TruSeq Illumina kit to prepare the cDNA
+libraries.
+
+\subsection*{Short reads alignments}
+About 20 millions of short reads were sequenced for each sample 
+(table \ref{count}).  We used a
+short read aligner, \program{BWA} \citep{Li2009a}, to map the sequenced short
+reads on the reference genome, \textit{S.\ mutans} UA159 (NCBI accession
+NC\_004350). Because the aligner allowed a few gaps for efficient alignments of
+millions of sequences of size being 100 base pairs long, short reads that
+consisted of RNAs with sequencing adapter sequences would not be mapped.  We
+used \program{cutadapt} \citep{Martin2011} to remove parts of adapter sequences
+from short reads before mapping on the reference genome.  Mapped short read
+alignments were converted into readable formats using \program{samtools}
+\citep{Li2009b}.
+
+\subsection*{Transcript predictions}
+RNA transcripts were inferred by applying a hidden Markov model to site-wise
+expression levels \citep{Martin2010}.  We used the pileup command of
+\program{BWA} to convert the short reads alignments to pilup values, which were
+taken as the site-wise expression levels along the genome.  Site-wise expression
+levels are a list of site-wise non-negative integers that represents numbers of
+short reads mapped to a particular genomic position.  We used the following
+options for the transcript inference program, \program{ParseRNAseq}: ``-c 10 -b
+25 -force\_gp'' to bin expression levels into 25 parts, and to allow 10 emission
+states for relative expression intensity. We also used the genes as the
+predicted parts of transcripts.    Because we enriched mRNAs in the cDNA
+preparation, we intented to group annotated genes to segments, which could be
+transcripts. We did not aim at finding precise transcription start or end sites.
+Because we did not have strand-specific RNA-seq of the samples due to limited
+resources, we could not tell in which directions transcripts were oriented
+solely based on RNA-seq data. Instead, we used strands of annotated genes to
+determine strands of predicted transcripts.
+
+\subsection*{Prediction of small RNAs and targets}
+Although our RNA-seq data did not enriched non-coding small RNAs in the cDNA
+preparation, we retained small RNAs in the sample by attempting to deplete only
+ribosomal RNAs from the sample. Because non-coding small RNAs and coding
+mRNAs were mixed in the sample, it was difficult to discriminate small
+non-coding RNAs from mRNAs unless non-coding RNAs expressions were masked by
+neighboring background RNA expressions. 
+Therefore, we sought a way of predicting small RNAs
+using \program{RNAz} \citep{Gruber2010}, which  used homologous sequences and
+RNA secondary structures to predict putative non-coding RNAs. Because sequence
+alignments were a critical step for finding small RNAs in the approach, we
+collated sequences similar to intergenic regions, and refined the alignments using
+a multiple sequence alignment tool. Exracting
+intergenic regions including up- and down-stream, we searched a database of
+bacterial genomes using BLAST \citep{Altschul1990} for similar intergenic
+sequences, subsequently, refining the sequence alignments using MUSCLE
+\citep{Edgar2004a}.  Then, we applied \program{RNAz} to the alignments for
+scoring regions for small RNAs \citep{Washietl2005,Gruber2010}.  Targets of
+transcripts with high scores for small RNAs were also predicted using
+\program{RNAplex} \citep{Tafer2008}, and \program{RNAplfold}
+\citep{Bernhart2006}.  Target genes were scored for each small RNA candidate,
+which was used to perform functional category enrichment tests of targets for
+each small RNA candidate.
+
+\subsection*{Experimental validation of small RNAs}
+\fixme{Burne group might have done or be able to do some experimental
+validations of putative non-coding RNAs shown at table \ref{tabUnannotatedRNA}
+\citep[e.g.,][]{Altuvia2007}.}
+
+\subsection*{Statistical analysis for differential expressions}
+We used \Rpackage{DESeq} \citep{Anders2010} to determine differentially
+expressed genes based on negative binomial model originally developed by
+\citet{Robinson2007}.  Steps of analyzing RNA-seq data for determining
+differentially expressed genes were described in \cite{Oshlack2010}.  We counted
+short reads aligned to annotated genes in the reference genome, forming a table
+of read counts of the genes. Statistical software R \citep{R2011} with
+\Rpackage{DEseq} was subsequently employed to infer differentially expressed
+genes in the different biological conditions.
+
+\subsection*{Gene category associations}
+We compiled three sets of gene categories for testing functional associations.
+Firstly, we assigned genes to Gene Ontology (GO) categories by comparing the
+\textit{S.\ mutans} genes to bacterial proteins from the Uniref90 database using
+\program{BLASTP}, and then assigning the same GO classification as the target
+gene of the uniProt GOA database if the match had an $E$-value of
+$<1.0\times10^{-5}.$  A gene family was assigned a given classification if any
+of its genes was assigned that classification.  Secondly, we collated functional
+classes of genes in the Oral Pathogen Sequence Databases available at
+\url{http://oralgen.lanl.gov}. Thirdly, we used a set of metabolite pathways for
+\textit{S.\ mutans} available at Kyoto Encyclopedia of Genes and Genomes (KEGG)
+\citep{Kanehisa2012}.  Using the sets of gene categories for \textit{S.\ mutans}
+we performed a Mann-Whitney $U$ test of the values of target genes of small RNAs
+associated with a given category versus those for the other categories. We
+used a variation of Fisher's exact test accounting for gene lengths to test
+enrichments of differentially expressed genes \citep{Young2010}.  Both of the
+tests were corrected for multiple testing hypotheses \citep{Benjamini1995}.
+
+\section*{DATA ACCESS}
+
+\subsection*{UCSC genome browser tracks and data availability}
+We created tracks for our recently released Streptococcus Genome Browser
+(\url{http://strep-genome.bscb.cornell.edu}) that summarized the results of our
+\textit{S.\ mutans} transcriptome analysis with known genes, gene expression
+levels based on the short read alignments, predicted putative transcript
+segments, and predictd small non-coding RNAs.  They should be accessed by clade
+\textit{Streptococcus}, and genome \textit{S. mutans UA159}.  These tracks can
+be used to inspect loci of interest and to compare the results of different
+RNA-seq data sets.  They can also be queried and intersected with other tracks
+using the UCSC Table Browser.
+
+\section*{ACKNOWLEDGMENTS}
+
 \fixme{Add more peoples or fundings if needed.}
@@
 
-\renewcommand*{\refname}{Literature Cited}
-\bibliographystyle{../../latex/bst/mbe}
+\section*{FIGURE LEGENDS}
+
+\section*{FIGURES}
+
+\section*{TABLES}
+
+\clearpage{}
+
+\renewcommand*{\refname}{REFERENCES}
+\bibliographystyle{../../latex/bst/genres}
 \bibliography{siepel-mutans}
@@
 
-\section*{Tables}
-
-\clearpage{}
-
-\section*{Figures}
-
 \end{document}

### 2012-04-13 package/supplementary.Rnw
Just an update.
QC first and then CRAM. We use CRAM without quality scores for subsequent analysis.

@@
 \caption{{\bf Average pileup values for segments.}
-The last bin sums from 1000 bp to 5000 bp.
+The last bin sums from 1000 to 50000. 
 }
@@
 # Use xm to plot a dotchart.
-dotchart(xm[,"fold"],labels=xm[,"Functional Class"],
+dotchart(xm[,"fold"],labels=xm[,"Functional Class"],cex=0.5,
         main="",xlab="Log2 fold change",col="black",pch=88)

### 2012-04-13 sh/batch2.sh

@@ -28,16 +28,16 @@ function batch2 {
     batch2-output
     batch2-speciesfile 
     batch2-push-data
-
     batch2-get-data
+
 #  ucsc-data
+    batch2-run-fastqc
     batch2-run-cram
     batch2-run-bwa-align
-
     batch2-run-samtools-pileup
 
-    # This would replace fastq-qc, bwa-align, and de.
-    # Because this creates similar job scripts to those created by fastq-qc,
+    # This would replace fastqc, bwa-align, and de.
+    # Because this creates similar job scripts to those created by fastqc,
     # bwa-align, or de, place qcalignde after them.
     batch2-run-qcalignde
 
@@ -146,8 +146,6 @@ function batch2-speciesfile {
   MAXLINEDE=$(grep ^MAXLINEDE\: $SPECIESFILE | cut -d":" -f2)
   # . minimum alignment mapping quality in _MINMAPQ_
   MINMAPQ=$(grep ^MINMAPQ\: $SPECIESFILE | cut -d":" -f2)
-  # . number CPUs in a compute node in _BWAALIGNNCPU_
-  BWAALIGNNCPU=$(grep ^BWAALIGNNCPU\: $SPECIESFILE | cut -d":" -f2)
   # . set bwa options in _BWAOPTION_
   BWAOPTION=$(grep ^BWAOPTION\: $SPECIESFILE | cut -d":" -f2)
   # . Rscript path in _CACRSCRIPT_
@@ -202,6 +200,69 @@ done
 EOF
 }
 
+function batch2-run-fastqc {
+  STATUS=fastqc
+  GENOMEFASTA=$(basename $REFGENOMEFASTA)
+  
+cat>$BASEDIR/run-$STATUS.sh<<EOF
+#!/bin/bash
+FASTQFILES=( $FASTQFILES )
+sed s/PBSARRAYSIZE/\${#FASTQFILES[@]}/g < batch-$STATUS.sh > tbatch.sh
+nsub tbatch.sh 
+rm tbatch.sh
+EOF
+
[…]

### 2012-04-24 package/tutorial.Rnw
Just an update.

%\VignetteIndexEntry{Analyzing Streptococcus mutans RNA-seq data}
%\VignettePackage{smutans}

% To compile this document
% library('weaver'); rm(list=ls()); Sweave('smutans.Rnw', driver=weaver()); system('pdflatex smutans')

\documentclass{article}

\usepackage{Sweave}
\usepackage[letterpaper]{geometry}
\usepackage{hyperref,graphicx}
\usepackage{color}
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=4,height=4.5}
\newcommand{\Robject}[1]{object \texttt{#1}}
\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\Rclass}[1]{class \textit{#1}}
\newcommand{\Rfunction}[1]{{function \small\texttt{#1}}}
\newcommand{\program}[1]{{\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

\renewcommand{\floatpagefraction}{0.7}

\author{Sang Chul Choi\\[1em]Cornell University,\\ Ithaca, NY\\[1em]
\texttt{schoi@cornell.edu}}

\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data\\using R packages}}}

\begin{document}

\maketitle

\begin{abstract}
We analyze RNA samples from \textit{Streptococcus mutans}.  We use
RNA-seq to define genes whose expression changes between samples
\cite{Oshlack2010}. Short reads are first mapped to a reference genome using
approaches such as \program{BWA} \cite{Li2009a}.  Reads fitting into one of the annotated
intervals are counted.  These counts are passed to a statistical package which
treats differences in read counts for statistical significance. For this
purpose, we made use of \Rpackage{DEseq} \cite{Anders2010} or \Rpackage{edgeR}
\cite{Robinson2010}. This prodecure returns a list of differentially expressed
genes, which can be passed to downstream analysis, such as functional category
enrichment test using \Rpackage{goseq} \cite{Young2010}.  The current
\Rpackage{smutans} contains procedures that use functions provided by
the above R packages. 
\end{abstract}

\tableofcontents

%--------------------------------------------------
\section{Data preparations} \label{sec:prep}
%--------------------------------------------------

Steps of analyzing RNA-seq data for determining differentially expressed genes
are described in \cite{Oshlack2010}. We used a short read aligner called
\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
aligner allowed a few gaps for efficient alignment of millions of sequences of
size being 100 base pairs long, short reads that consist of RNAs and sequencing
adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
remove parts of adapter sequences from short reads before mapping on the
reference genome. Mapped short read alignments were converted into readable
formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
was subsequently employed to infer differentially expressed in different
biological conditions.  We also generated tracks for our recently released
Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
summarize gene expression levels based the short read alignments. This allowed
us to visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functional category enrichment test
\cite{Young2010} were also employed.

The \Rpackage{smutans} expects count data in the form of \Rclass{CountDataSet},
which is a type provided by \Rpackage{DESeq}.  Using \program{BWA} \cite{Li2009a} and
\program{samtools} \cite{Li2009b} we aligned short reads from RNA sequencing. Using gene
annotations in the UA159 genome we counted short reads that overlapped the
genes.  The count values are raw counts of sequencing reads. The
\Rpackage{DESeq}'s statistical model, or negative binomial model, hold for raw
reads that allows to assess the measurement precision correctly; we did not have
rounded values of normalized counts, or counts of covered base pairs.  Several
exeriments of \textit{S.\ mutans} RNA sampling were performed.  \fixme{Find the
papers that are related with the experiments.} One of them was presented in the
\Robject{smutansGenes} as a part of \Rpackage{smutans}. The
\Robject{smutansGenes} is of \Rclass{CountDataSet}, which is the data type used
by \Rpackage{DESeq}. We have a convenient \Rclass{Smutans}, which contains slots
for \Rclass{CountDataSet} objects, and results from analyses thereof.  Before
loading necessary R packages and data sets, let us describe what functions were
used for preparing data objects such as \Robject{smutansGenes}. One should
create an R script file at the directory that contains the source code directory
of the \Rpackage{smutans}. Put the following scripts in the R script file and
run the script at that directory. That should create RData files at the
directory smutans/data. User should not run the R script because the data files
should be found in the \Rpackage{smutans}.

% http://www.oralgen.lanl.gov/oralgen/downloads/smut.csv
% http://www.oralgen.lanl.gov/cgi-bin/gene_id_search.cgi?dbname=smut&gene_id=SMu0001
% \fixme{Where are flat files for genes of S. mutans UA159?}
<<samplePreparation,eval=FALSE>>=
# not run
# but create a R scirpt file with the following scripts, and
# run it, i.e., Rscript run.R
library(DESeq)
source("smutans/R/core.R")
smutans.prepareData34( )
smutans.prepareData835NPP( )
smutans.prepareDataUA159TW1( ) 
smutans.prepareGoseq( )
smutans.prepareDataOMZ175( ) 
smutans.prepareDataUA159CSP( ) 
smutans.prepareDataSMU86CSP( )
q("no")
@
<<options,results=hide,echo=FALSE>>=
options(digits=3)
@
Now, see some of the data objects by loading the \Rpackage{smutans}, and a data
\Robject{smutansGenes}. Note that the object is of type \Rclass{CountDataSet} of
\Rpackage{DESeq} in which users should find \Rfunction{counts} and
\Rfunction{pData}. 
<<smutans,results=hide>>=
library( "smutans" )
data( "smutansGenes" )
@
There are 
\Sexpr{length(colnames(counts(smutansGenes)))}
RNA-seq samples in \Robject{smutansGenes}, which
contains the counts and also metadata about the samples:
<<look_into_smutansGenes>>=
head( counts(smutansGenes) )
@
<<look_into_smutansGenes>>=
pData( smutansGenes )
@
The samples differ by experimental condition (``glucose'' or
``galactose'') and by library type (``ua159'' or ``tw1'').

Samples of 835NP, 835P, and UA159 no CSP should be loaded using the following
command.
<<sm835,results=hide>>=
data( "sm835Genes" )
@

% Resources
% Google charts - Venn Diagram
% http://code.google.com/apis/chart/index.html

\fixme{Describe other data objects as well.
sm34Genes.RData,
sm835Genes.RData,
smutans.cat.desc.RData,
smutans.feature.genes.RData, and
smutans.go.genes.RData.}

%--------------------------------------------------
\section{DE analysis using UA159 and TW1 samples with
glucose and galactose} \label{sec:detw1}
%--------------------------------------------------

We could compare four pairwise comparisons using the two-factor data set of
\Robject{smutansGenes}: a pair of comparisons between ``glucose'' and
``galactose'' given either type ``ua159'' or ``tw1'', and another pair of
comparisons between types ``ua159'' and ``tw1'' given either condition
``glucose'' or ``galactose.'' We focus on one of the four pairwise comparisons
first. Later, we will perform analyses for the rest of the three comparisons.
We look at the ``ua159'' data for now, and will look at the ``tw1'' data later.

We can cluster samples using variance stabilized data. Figure \ref{figClust}
shows the \Sexpr{length(colnames(counts(smutansGenes)))} sampled clustered. 
The clustering should reflect our experimental design, i.e., samples are
more similar when they have the same treatment or the same library type.
To make this conclusion, we estimate the dispersion with
mode ``blind'' (see \Rpackage{DESeq} vignette), as only then,
the variance stabilizing transformation is not informed about the design and
hence not biased towards a result supporting it. Such an analysis is useful
for quality control.  This appears to be useful as a first in any analysis.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{smutans-figClust}
\caption{Heatmap showing the Euclidean distances between the samples
  as calculated from the variance-stabilizing transformation of the
  count data.}
\label{figClust}
\end{figure}
<<figClust,fig=TRUE>>=
allSample <- newSmutans( smutansGenes )
smutans.de2Clust( allSample )
@
As expected, two experimental conditions of ``glucose'' and ``galactose'' divide the
samples. Within each condition, two different types of ``ua159'' and ``tw1'' divide
them again. The pair of ``glucose ua159'' and ``galactose tw1'' is the most
distinguished.

%------------------------------------------------------------
\subsection{Inference using single-factor data} 
\label{subsec:ua159only}
%------------------------------------------------------------

We create a \Rclass{Smutans} object using the \Rclass{CountDataSet}
\Robject{smutansGenes}. The title option would be useful in distinguishing
different objects of \Rclass{Smutans} as we have more R objects with the type.  
A subsequent call to
\Rfunction{smutans.de2} uses a two-factor data set to test differential
expression of genes.  The \Rfunction{smutans.de2} estimates effective library
sizes, and dispersions using functions provided by \Rpackage{DESeq}. Then, it
tests genes for differential expression.
<<ua159onlySamples>>=
ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
ua159only <- smutans.de2( ua159only, type="ua159", 
                          condA="glucose", condB="galactose" )
@
Because \Rpackage{DESeq} recommended to check
the dispersion estimates and to make sure that the data quality is as
expected, we plot the per-gene estimates against the normalized mean
expressions per gene, and then overlay the fitted curve in red color
(Fig.\ \ref{figFit}).

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figFit}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength for the data set of UA159-only sample
to compare glucose and galactose conditions.}
\label{figFit}
\end{figure}
<<figFit,fig=TRUE>>=
smutans.plotDispersionEstimates(ua159only)
@
\fixme{The following paragraph is adopted from \Rpackage{DESeq} vignette.}
As we estimated the dispersion from only two samples, we should expect the
estimates to scatter with quite some sampling variance around their true values.
Hence, \Rpackage{DESeq} should not use the per-gene estimates directly in the
test, because using too low dispersion values leads to false positives.  Many of
the values below the red line are likely to be underestimates of the true
dispersions, and hence, it is prudent to instead rather use the fitted value.
On the othe hand, not all of the values above the red line are overestimations,
and hence, the conservative choice is to keep them instead of replacing them
with their fitted values.
Having estimated the dispersion for each gene, we look for differentially
expressed genes. To contrast two conditions, e.g., to see whether there is
differential expression between conditions ``glucose'' and ``galactose'',
\Rfunction{smutans.de2} already performed the 
\Rfunction{nbinomTest} provided by \Rpackage{DESeq}.  
Let's list genes with resulting information from \Rpackage{DESeq}.
<<ua159onlyDE2List>>=
head(smutans.de2List( ua159only ))
@
For each gene, we get its mean expression level (at the base scale) as a joint
estimate from both conditions, and estimated separately for each condition, the
fold change from the first to the second condition, i.e., 
$\frac{baseMeanB}{baseMeanA}$, the logarithm (to basis 2)
of the fold change, and the $p$ value for the statistical significance of this
change. The \texttt{padj} column contains the $p$ values, adjusted for multiple
testing with the Benjamini-Hochberg procedure (see the R function
\Rfunction{p.adjust}), which controls false discovery rate (FDR). Later, we
print the gene list out in a CSV file using option file of
\Rfunction{smutans.de2List}.
%-----------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figDE}
\caption{Plot of normalised mean versus $log_2$ fold change
(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
versus ``galactose'' using UA159-only sample.}
\label{figDE}
\end{figure}
%-----------------------------------------------------------------------------
<<figDE,fig=TRUE>>=
smutans.plotDiffExp(ua159only)
@
Figure \ref{figDE} shows a plot of the $\log_2$ fold changes against the base
means, coloring in red those genes that are significant at 10\% of FDR. 
Let us use ``tw1'' type to compare ``glucose'' and ``galactose'' conditions.
We continue to analyze the data set using type ``tw1'' to compare the two
conditions. We perform the single-factor analysis. 
<<tw1onlySamples>>=
tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
tw1only <- smutans.de2( tw1only, type="tw1", 
                        condA="glucose", condB="galactose" )
tabUA159onlyTW1only <- smutans.compareResult( ua159only, tw1only )
addmargins( tabUA159onlyTW1only )
@
The table, \Robject{tabUA159onlyTW1only}, shows numbers of genes found to be 
differentially expressed in the two analyses: ``UA159 Only'' and ``TW1 Only.''
Of the total \Sexpr{sum(tabUA159onlyTW1only)} genes, \Sexpr{tabUA159onlyTW1only[2,2]}
genes were found in both analyses. The analysis using TW1-only sample found
\Sexpr{tabUA159onlyTW1only[1,2]} genes. The analysis using UA159-only sample found
\Sexpr{tabUA159onlyTW1only[2,1]} genes.

Until now we focused on the effect of condition either using UA159-only or
TW1-only samples. We now change the gear to see effects of sample types on
differentially expressed genes. We create two \Rclass{Smutans} objects:
\Robject{glucoseOnly} and \Robject{galactoseOnly}, and compare them in detecting
differentially expressed genes:
<<glucoseOnlySamples>>=
glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
                            condA="ua159", condB="tw1" )
galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
                              condA="ua159", condB="tw1" )
tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
                                                      galactoseOnly )
addmargins( tabGlucoseOnlyGalactoseOnly )
@
Of the total \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} genes,
\Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]} genes were found in both analyses at FDR
rate of 10\%. The analysis using galactose-only sample found
\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} genes. The analysis using glucose-only
sample found \Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} genes. 
Table \ref{tab:de} summarizes the number of differentially expressed genes that
were found using single-factor analyses at FDR of 10\%.

%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Numbers of differentially expressed genes in different analyses}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{ccccl}
\hline 
Sample A & Sample B & Number of Genes & Result File \\
\hline 
UA159 Glucose & UA159 Galactose & \Sexpr{sum(ua159only@padj < .1,na.rm=T)} & ua159only.csv \\
TW1 Glucose & TW1 Galactose & \Sexpr{sum(tw1only@padj < .1,na.rm=T)} & tw1only.csv \\
UA159 Glucose & TW1 Glucose & \Sexpr{sum(glucoseOnly@padj < .1,na.rm=T)} & glucoseOnly.csv \\
UA159 Galactose & TW1 Galactose & \Sexpr{sum(galactoseOnly@padj < .1,na.rm=T)} & galactoseOnly.csv \\
% UA159 Glucose & UA159 Galactose & \Sexpr{sum(tabUA159onlyTW1only[2,])} & ua159only.csv \\
% TW1 Glucose & TW1 Galactose & \Sexpr{sum(tabUA159onlyTW1only[,2])} & tw1only.csv \\
% UA159 Glucose & TW1 Glucose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[2,])} & glucoseOnly.csv \\
% UA159 Galactose & TW1 Galactose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[,2])} & galactoseOnly.csv \\
\hline 
\end{tabular}
\par\end{centering}
\label{tab:de}
\end{table}
%-----------------------------------------------------------------------------

%------------------------------------------------------------
\subsection{Inference using all of the samples} 
\label{subsec:allSample}
%------------------------------------------------------------

It is inconvenient to use four different analyses to find differentially
expressed genes under one of two factors. Besides, it would be desirable to use all of
the data to find differentially expressed genes because using more samples might
be more powerful in detecting differentially expressed genes.  Let us find genes
differentially expressed in different conditions of ``glucose'' and
``galactose'' using both types of ``ua159'' and ``tw1.'' We will compare these
genes and those found in the single-factor analysis that we did previously,
i.e., \Robject{ua159only}.
Let's start with multi-factor the analysis by creating a new \Rclass{Smutans} object, called
\Robject{allSample}. Although we already used \Robject{allSample} when
clustering the samples, we recreate it for instructive purposes.
<<allSamples,results=hide>>=
allSample <- newSmutans( smutansGenes, 
                         title="All Samples for Comparing Conditions" )
allSample <- smutans.de2( allSample )
@
The \Rfunction{smutans.de2} uses a two-factor data set to
perform two-factor analysis. The two factors are type and condition. We are here
interested in condition effect. We could check 
the dispersion estimates and to make sure that the data quality is as
expected by plotting the per-gene estimates against the normalized mean
expressions per gene, and then overlay the fitted curve.
<<figFitAllSample,eval=FALSE>>=
# not run
smutans.plotDispersionEstimates(allSample)
@
%-----------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figCompareUA159onlyAndAllSample}
\caption{Comparison of $p$ values (unadjusted) from the test using only the
four paired-end samples and the test using all seven samples. We can see that
the latter tend to be smaller, as expected from the higher power of the test with all samples.}
\label{figCompareUA159onlyAndAllSample}
\end{figure}
%-----------------------------------------------------------------------------
<<figCompareUA159onlyAndAllSample,fig=TRUE>>=
pval.ua159only <- smutans.pval( ua159only )
pval.allSample <- smutans.pval( allSample )
bottom = function(x, theta=1e-12) pmax(x, theta)
plot( bottom(pval.ua159only), bottom(pval.allSample), log="xy", pch=20, cex=.3 )
abline(a=0, b=1, col="blue")
@
<<compareUA159onlyAndAllSample>>=
tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
                                                   allSample )
addmargins( tabUA159onlyAndAllSample )
@
Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
genes.  Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from
the two-factor analysis versus those from the single-factor analysis. The
p-values from the two-factor analysis are overall smaller than those from the
single-factor analysis. We would expect that this was the case because of higher
power of the two-factor analysis.

We can show that taking the library type into account was important to have
good detection power by doing the analysis again using the standard workflow,
as outlined earlier, and without informing \Rpackage{DESeq} of the library
types. This anlaysis is not necessary. However, it clearly shows the advantage of the
two-factor analysis.
<<allSampleWithoutType>>=
allSampleWithoutType <- newSmutans( smutansGenes, "All Sample without Type" )
allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
                                                condA="glucose", 
                                                condB="galactose" )
tabWithoutWithAllSample <- smutans.compareResult( allSampleWithoutType, 
                                                  allSample )
addmargins( tabWithoutWithAllSample )
@
Of the total \Sexpr{sum(tabWithoutWithAllSample)} genes,
\Sexpr{tabWithoutWithAllSample[2,2]} genes were found in both analyses. The
analysis using all of the sample found \Sexpr{tabWithoutWithAllSample[1,2]}
genes. The analysis using all of the same sample but without considering type
found only \Sexpr{tabWithoutWithAllSample[2,1]} genes.

As we mentioned above, we can have two analyses using the two-factor analysis.
Until now we compared conditions.
We proceed to compare the two types by considering all of the samples.
<<allSampleType,results=verbatim>>=
allSampleType <- newSmutans( smutansGenes, title="All samples for type" )
allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
tabGlucoseOnlyAllSample <- smutans.compareResult( glucoseOnly, 
                                                  allSampleType )
addmargins( tabGlucoseOnlyAllSample )
tabGalactoseOnlyAllSample <- smutans.compareResult( galactoseOnly, 
                                                    allSampleType )
addmargins( tabGalactoseOnlyAllSample )
@
Of the total \Sexpr{sum(tabGlucoseOnlyAllSample)} genes,
\Sexpr{tabGlucoseOnlyAllSample[2,2]} genes were found in both analyses. The
analysis using all of the sample found \Sexpr{tabGlucoseOnlyAllSample[1,2]}
genes. The analysis using glucose-only sample
found \Sexpr{tabGlucoseOnlyAllSample[2,1]} genes.  
In the comparison using galactose-only sample,
of the \Sexpr{sum(tabGalactoseOnlyAllSample)} genes,
\Sexpr{tabGalactoseOnlyAllSample[2,2]} genes were found in both analyses. The
analysis using all of the sample found \Sexpr{tabGalactoseOnlyAllSample[1,2]}
genes. The analysis using galactose-only sample
found \Sexpr{tabGalactoseOnlyAllSample[2,1]} genes.

We print out the result of differentially expressed genes in CSV files. A
spreadsheet program such as EXCEL should be able to open CSV files. Users who
are more comfortable with the EXCEL program than R could post-process CSV
files for sorting and finding differentially expressed genes using different
q-value cutoff values.
<<smutanSamples>>=
smutans.de2List( allSample, "allSample.csv" )
smutans.de2List( allSampleType, "allSampleType.csv" )
smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
smutans.de2List( galactoseOnly, "galactoseOnly.csv" )
smutans.de2List( ua159only, "ua159only.csv" )
smutans.de2List( tw1only, "tw1only.csv" )
head( smutans.de2List( allSample ) )
@
The output from \Robject{allSample} is different from that of
\Robject{ua159only} or \Robject{tw1only}.
The first column of ``allSample.csv'' is for genes.
The next three columns show the fitted coefficients, converted to a logarithm
base 2 scale: ``X.Intercept.'', ``typeua159'', and ``conditionglucose.''
The log2 fold change due to the condition is shown in the 4th column. As indicated
by the column name, it is the effect of ``glucose'', i.e., the log ratio of
``glucose'' versus ``galactose'' or $\log(glucose/galactose)$. 
Note that the library type also
had noticeable influence on the expression, and hence would have increased the dispersion
estimates (and so reduced power) if we had not fitted an effect for it.
The column \emph{deviance}
is the deviance of the fit. (Comparing the deviances with a $\chi^2$ likelihood ratio test is
how \Rfunction{nbinomGLMTest} calculates the $p$ values.) The column, \emph{converged},
indicates whether the calculation of coefficients and deviance has fully converged.
(If it is false too often, you can try to change the GLM control parameters, as
explained in the help page to \Rfunction{fitNbinomGLMs}.)
The second to the last column, \emph{pval}, is the raw p-values from the test
\Rfunction{nbinomGLMTest} provided by \Rpackage{DESeq}. The last column,
\emph{padj}, is the multiple testing adjusted p-values.

%------------------------------------------------------------
\subsection{Functional category association analyses} 
\label{subsec:functional}
%------------------------------------------------------------

Once we find differentially expressed genes, we could cluster them. This could
be useful to group genes that behave similarly.  These potentially useful
figures can be plotted from \Rpackage{DESeq}.  We could also find function
categories that were enriched with differentially expressed genes. For
clustering differentially expressed genes we need moderate fold changes to make
them amenable to distance calculation.  Figure \ref{figLfc} shows moderated log
fold changes versus the original log fold changes.

<<options,results=hide,echo=FALSE>>=
largeNumber <- 10
@

\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figLfc}
\caption{Scatterplot of direct (\Robject{lfc}) versus \emph{moderated} log-ratios (\Robject{mod\_lfc}).
  The moderation criterion used is variance stabilisation.
  The red points correspond to values that were infinite in \Robject{lfc}
  and were abitrarily set to $\pm\Sexpr{largeNumber}$ for the purpose of plotting.
  These values vary in a finite range (as shown in the plot) in \Robject{mod\_lfc}.}
\label{figLfc}
\end{figure}
<<figLfc,fig=TRUE>>=
select <- 1:40
smutans.de2TypeLfc( ua159only, condA="glucose", condB="galactose" )
@
Now we plot the first \Sexpr{length(select)} genes clustered based on the
moderated log fold changes.

%-----------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{smutans-figHeatmap1}
\caption{Heatmaps showing the expression data of the top \Sexpr{length(select)}
  differentially expressed genes. Left, the variance stabilization
  transformed data, right, the original count data.
  The right plot is dominated by a small number of data points with large values.}
\label{figHeatmap1}
\end{figure}
%-----------------------------------------------------------------------------
<<figHeatmap1,fig=TRUE>>=
smutans.de2TypeHeatmap( ua159only )
@

<<goseqUA159only,eval=FALSE>>=
data( "smutans.feature.genes" )
data( "smutans.go.genes" )
data( "smutans.cat.desc" )
smutans.de2Goseq ( ua159only, qval=1e-2, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc )
@
We find some functional categories enriched with differentially expressed genes
(table \ref{tab:goseq})
We find functional categories that are enriched with differentially expressed genes.
To find functional categories enriched with differentially expressed genes we
first find a list of differentially expressed genes. A set of genes are assigned
to functional gene categories: \Robject{smutans.go.genes}. Features of the genes in
UA159 include gene lengths: \Robject{smutans.feature.genes}.
Functional categories with numbers of genes assigned are in
\Robject{smutans.cat.desc}.
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
genes}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<goseqUA159only,echo=FALSE,results=tex>>=
data( "smutans.feature.genes" )
data( "smutans.go.genes" )
data( "smutans.cat.desc" )
smutans.de2Goseq ( ua159only, qval=1e-2, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc )
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:goseq}
\end{table}
%-----------------------------------------------------------------------------

%--------------------------------------------------
\section{DE analysis using 835NP, 835P, and UA159 no CSP}
\label{sec:de835}
%--------------------------------------------------

We first cluster the \Sexpr{length(colnames(counts(sm835Genes)))} samples.
\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{smutans-figClustNpAndP}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength.}
\label{figClustNpAndP}
\end{figure}
<<figClustNpAndP,fig=TRUE>>=
data( "sm835Genes" )
npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
smutans.de2Clust( npAndP )
@

We found no functional categories enriched with differentially expressed
genes in the pairwise comparisons of 835NP, 835P, and UA159 no CSP.
Let's make a list of differentially expressed genes. 
Another kind of heatmap plots for
grouping genes could be interesting. 
But, basically we have a list of genes with
some adjusted p-values for detecting differentially expressed genes.

<<look_into_smutansGenes>>=
head( counts(sm835Genes) )
pData( sm835Genes )
@

<<sm835Samples>>=
npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
npAndP <- smutans.de2( npAndP, type="ua159", 
                       condA="835np", condB="835p" )
npAndUA159 <- newSmutans( sm835Genes, title="835NP vs. UA159" )
npAndUA159 <- smutans.de2( npAndUA159, type="ua159", 
                           condA="835np", condB="ua159nocsp" )
pAndUA159 <- newSmutans( sm835Genes, title="835P vs. UA159" )
pAndUA159 <- smutans.de2( pAndUA159, type="ua159", 
                          condA="835p", condB="ua159nocsp" )
smutans.de2List( npAndP, "npAndP.csv" )
smutans.de2List( npAndUA159, "npAndUA159.csv" )
smutans.de2List( pAndUA159, "pAndUA159.csv" )
@

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{smutans-figHeatmapNpAndP}
\caption{Empirical (black dots) and fitted (red lines) dispersion values
plotted against mean expression strength. \textbf{figHeatmapNpAndP}}
\label{figHeatmapNpAndP}
\end{figure}
<<figHeatmapNpAndP,fig=TRUE>>=
smutans.de2TypeHeatmap( npAndP )
@

%--------------------------------------------------
\section{DE analysis using OMZ175 and OMZ/HKRR}
\label{sec:deomz}
%--------------------------------------------------

<<look_into_smomzGenes>>=
data( "smomzGenes" )
head( counts(smomzGenes) )
pData( smomzGenes )
@

<<smomz,results=hide>>=
omz <- newSmutans( smomzGenes, title="OMZ175 vs. OMZ/HKRR" )
omz <- smutans.de2( omz, type="ua159", 
                    condA="OMZ175", condB="OMZHKRR" )
smutans.de2List( omz, "omz.csv" )
@

%--------------------------------------------------
\section{DE analysis using UA159 CSP}
\label{sec:deua159csp}
%--------------------------------------------------

<<look_into_smua159cspGenes>>=
data( "smua159cspGenes" )
head( counts(smua159cspGenes) )
pData( smua159cspGenes )
@

<<smua159csp,results=hide>>=
ua159csp <- newSmutans( smua159cspGenes, title="UA159 no CSP vs. CSP" )
ua159csp <- smutans.de2( ua159csp, type="ua159", 
                         condA="UA159noCSP", condB="UA159CSP" )
smutans.de2List( ua159csp, "ua159csp.csv" )
@

%--------------------------------------------------
\section{DE analysis using SMU86 CSP}
\label{sec:desmu86csp}
%--------------------------------------------------

<<look_into_smsmu86cspGenes>>=
data( "smsmu86cspGenes" )
head( counts(smsmu86cspGenes) )
pData( smsmu86cspGenes )
@

<<smsmu86csp,results=hide>>=
smu86csp <- newSmutans( smsmu86cspGenes, title="UA159 no CSP vs. CSP" )
smu86csp <- smutans.de2( smu86csp, type="ua159", 
                         condA="Smu86noCSP", condB="Smu86CSP" )
smutans.de2List( smu86csp, "smu86csp.csv" )
@

%--------------------------------------------------
\section{DE analysis using predicted transcripts}
\label{sec:detranscript}
%--------------------------------------------------

We need ways of supporting the predicted transcripts. We need a way of using
expression levels and counts of starts and ends of short reads to predict
transcripts. For now, we just use an ad hoc method for finding transcript start
and end sites. A likelihood can be constructed by multiplying geometric
probabilities. Given a range of possible transcription boundaries we can find
estimates of positions and geometirc parmeter $p$ that maximize the likelihood
for both ends of a transcript.

\Rpackage{GenomicRanges} might be useful in 
<<tx,eval=FALSE>>=
source("http://www.bioconductor.org/biocLite.R")
biocLite("GenomicRanges")
@

%--------------------------------------------------
\section{DE analysis using predicted small RNAs and target genes}
\label{sec:desmallrna}
%--------------------------------------------------

Create a file with intergenic regions sequences of \textit{S.\ mutans} UA159.
Create a file with bacterial genome sequences except \textit{S.\ mutans} NN2025.
BLAST the intergenic regions against the bacterial genomes. Extract sequence
alignments from the BLAST output. Refine the alignment using MUSCLE. Convert the
FASTA-format alignments to MSA-format ones.  Find small RNA candidates using
RNAz.  Find targets of the small RNA candidates using RNAplex. Targets are
scored for each small RNA candidate. The scores were used to do functional
category enrichment tests of targets for each small RNA candidate.

Table \ref{tab:goseq} 
shows functional categories enriched with target genes for predicted small
RNAs. Let's use \Rfunction{smutans.mannwhitney} to the package and other data
objects.
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
genes}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{lcccl}
\hline 
small RNA & $q^a$ & count & GO & Description \\
\hline 
%-----------------------------------------------------------------------------
<<goseqUA159only,echo=FALSE,results=tex>>=
data( "smutans.genes.criteria" )
data( "smutans.go.genes" )
data( "smutans.cat.desc" )
smutans.mannwhitney( qval=0.05,
                   genes.criteria=smutans.genes.criteria,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc )
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:goseq}
\end{table}
%-----------------------------------------------------------------------------

<<smutansRNA>>=
data( "smutansGenes2" )
ua159only <- newSmutans( smutansGenes2, title="UA159 Only" )
ua159only <- smutans.de2( ua159only, type="ua159", 
                          condA="glucose", condB="galactose" )
allSample <- newSmutans( smutansGenes2, 
                         title="All Samples for Comparing Conditions" )
allSample <- smutans.de2( allSample )
tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
                                                   allSample )
addmargins( tabUA159onlyAndAllSample )
@
Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
genes. We found a few predicted small RNAs that appeared to be differentially
expressed.
<<smutansRNA>>=
smutans.de2List( allSample )[grep("locus",rownames(smutans.de2List(allSample))),]
@
We also found a few predicted small RNAs that appeared to be differentially
expressed in the comparison of the two types: ``ua159'' and ``tw1.''
<<smutansRNA>>=
allSampleType <- newSmutans( smutansGenes2, title="All samples for type" )
allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
smutans.de2List( allSampleType )[grep("locus",rownames(smutans.de2List(allSampleType))),]
@

\section{Prewriting about UA159 versus TW1}

\subsection{Abstract}
A bacterial transcriptome of the primary etiological agent of human dental caries,
\textit{Stretptococcus mutans}, is described using RNA-seq data. Putative
transcripts boundaries are defined. Small RNAs are predicted by computational
methods, and confirmed by RNA-seq and Southern blot experiments. We refine the
previously established importance of CcpA regulation as central metabolism and
virulence gene expression using RNA-seq data.
CcpA is known to play a direct role in carbon catabolite repression (CCR).
We describe genes that are differentially expressed in glucose-grown wild-type
(UA159) and CcpA-deficient (TW1) strains. We found that more
genes were differentially expressed in the comparison between wild-type and
CcpA-deficient strains in galactose-grown strains than in
glucose-grown strains.
We also found that more genes differentially expressed in wild-type than in
CcpA-deficient strain when comparing glucose- and galactose-grown cells. These
two findings were different from those of \cite{Abranches2008}. \fixme{We may
want to see effects of different p-value cutoff.} We confirmed that CcpA
modulates the pathogenic potential of \textit{S.\ mutans} through global control
of gene expression.
We predicted a few small RNAs based on sequence alignments and RNA secondary
structures. We confirmed them experimentally. Target genes of predicted small
RNAs were predicted. The target genes were enriched in some functional
categories. We listed highly potential small RNAs with their target genes
functional categories. 
Transcripts were predicted using RNA-seq short reads mapped on the wild-type
UA159 genome.

\subsection{Reviews of papers that cited \cite{Abranches2008}}

CCR in bacteria involves regulatory networks that activate or silence genes in
response to carbon source and availability.

\subsection{Goal: Refinement of \cite{Abranches2008}}

Comparisons of microarray and RNA-seq are not good way to go.

\subsection{Methods and Materials}

\subsubsection{Bacterial strains and growth conditions}

\subsubsection{RNA isolation}

\subsubsection{RNA sequencing}

\subsubsection{Statistical analysis for differential expressions}
We used an R package, DESeq, \cite{Anders2010} to determine differentially
expressed genes based on negative binomial model originally developed by \cite{Robinson2007}.
Steps of analyzing RNA-seq data for determining differentially expressed genes
are described in \cite{Oshlack2010}. We used a short read aligner called
\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
aligner allowed a few gaps for efficient alignment of millions of sequences of
size being 100 base pairs long, short reads that consist of RNAs and sequencing
adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
remove parts of adapter sequences from short reads before mapping on the
reference genome. Mapped short read alignments were converted into readable
formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
reference genome, forming a table of read counts of the genes. Statistical
software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
was subsequently employed to infer differentially expressed in different
biological conditions.  We also generated tracks for our recently released
Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
summarize gene expression levels based the short read alignments. This allowed
us to visualize parts of the reference genome for comparing different RNA
samples. Further analyses such as functional category enrichment test
\cite{Young2010} were also employed.

\subsubsection{small RNAs}

RNAz is used to find small RNA candidates. Sequence alignments were a
critical step for finding small RNAs in the approach. Phylogenies of streptococcus species could
be considered to choose species related with Streptococcus mutans.
Alternatively, I find sequences similar to intergenic regions, and refine the
alignments using a multiple sequence alignment tool such as MUSCLE
\cite{Edgar2004a}. Steps for
achieving this goal include: 1. Exract intergenic regions including up- and
down-stream, 2. BLAST the intergenic regions against non-redundant DB to find
similar sequences \cite{Altschul1990}, 3. Refine each alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the
alignments for scoring regions for small RNAs
\cite{Washietl2005,Gruber2010}, and 5. Find transcripts with high
score of small RNAs. The transcripts must be defined in advance.

RNAplex
\cite{Tafer2008}

A review paper about small RNA predictions and targets
\cite{Tafer2008a}

RNAplfold
\cite{Bernhart2006}

\subsubsection{Transcripts}

``-c 10 -b 25 -force_gp''

\subsubsection{Gene category associations and virulence genes}

We assigned genes to Gene Ontology (GO) categories by comparing the {\em
  Streptococcus} genes to bacterial proteins from the Uniref90 database
using BLASTP, and then assigning the same GO classification as the target
gene of the uniProt GOA database if the match had an $E$-value of
$<1.0\times10^{-5}.$  
A gene family was assigned a given
classification if any of its genes was assigned that classification.  To
test for associations with replacing gene transfers we used the
recombination intensity estimated for each gene, and for additive gene
transfers we used the number of additive transfers inferred for each
family.  To test for significance, we performed a Mann-Whitney $U$ test of
the values (recombination intensities or numbers of transfers) associated
with a given category vs.\ the values for the all other genes/families.  We
used the \cite{Benjamini1995} method to correct for multiple comparisons.

\subsection{Results}

CcpA affects CCR of \textit{fruA}. 
\textit{fruA} fructan hydrolase
When Wen2002 deleted CRE sequencs in the \textit{fruA} promoter region, CCR was
almost completely alleviated. 
What are CCR-sensitive genes? 
What are the mechanism of CcpA on CCR of the \textit{fruA} operon? 
Expression levels of \textif{fruA} in \textit{S.\ mutans} strains that were
grown to early exponential phase (OD_600, 0.2) in TV medium supplemented with
0.5\% glucose or galactose. \fixme{What does this mean?} 
Deletion of \textit{ccpA} resulted in substantial relief of the CCR of
\textit{fruA} expression exerted by both glucose and fructose. 
CcpA could directly interact with the \textit{fruA} promoter region.

The role of CcpA in gene expression during growth on glucose. 
\textit{S.\ mutans} UA159 and TW1 grown under catabolite-repressing conditions
(glucose), and catabolite-derepressing conditions (galactose). Under
glucose-only condition I found xx number of differentially expressed genes. What
are the functional categories of those differentially expressed genes? What are
the genes that are most profoundly up- or down-regulated (Smu.1421 to Smu.1424,
PDH enzyme complex), ... and more? ... This regulatory control was
short-circuited in the strain growing on glucose and lacking CcpA. I showed that
some of KEGG pathways were affected by TW1 grown under glucose. We found
down-regulated genes, which would have something to do with pathogenicity. 
Upto here, I compared UA159 and TW1 under glucose.

Growth in derepressing conditions induces major changes in the UA159
transcriptome. Here, I compared glucose and galactose conditions for growing
UA159. Energy metabolism was enriched with differentially expressed genes. There
were more up-regulated genes in galactose condition compared with glucose
condition. Some genes are described in detail.

Now, I compared glucose and galactose for growing TW1.

Now, I compared UA159 and TW1 under galactose.

What genes are differentially expressed?

Name the transcripts.
I found xx transcripts. Among them xx \% were mono-cistronic operons. xx \% were
poly-cistronic operons. xx transcripts could not defined because RNA-seq short
reads were not precise. Transripts were assigned with functional categories, and
differentially expressed transcripts were found. Transcripts and functional
categories: how can I assess that transcripts consist of genes with similar
functional cateogires.

Name the small RNAs.
RNAz predicted a few small RNAs. Some of them were differentially expressed.
Some functional categories were enriched with target genes of the predicted
genes.

\subsection{Discussion}
We used RNA-seq data to delineate transcript boundaries. Although this map of
transcripts does not precisely tell us where transcripts start and end, it
informs us of which open reading frames can be transcribed together. Potentially
polycistronic operons are defined. We focused on interpretations of operon in
light of CcpA regulations.

\subsection{Abbreviations}

\begin{enumerate}
\item CCR Carbon catabolite repression
\item CcpA ?
\item IPS intracellular polysaccharide
\end{enumerate}

\section{Session Info}
<<sessi>>=
sessionInfo()
@

\bibliographystyle{unsrt}
\bibliography{siepel-mutans}

\end{document}

### 2012-04-24 package/manuscript.Rnw
Just an update.

@@
 Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
-\fixme{Add more coauthors, or change the order of authors if needed} \\
 Robert A.\ Burne$^{3}$
@@
 \vspace{1in}
-
-\fixme{Add more corresponding authors.}
 
@@
 ideas of the differential expression study and transcriptome study.}
+
 A bacterial transcriptome of the primary etiological agent of human dental
@@
 including mRNAs and small noncoding RNAs in the genome.
+
 Differential expression studies on the role of CcpA in 
 carbon catabolite repression (CCR) suggested that some of the novel mRNAs and
-small RNAs found in our study could be involved in the regulation of CcpA. Our
-study provides more precise landscape of transcriptome changes in response to
-different carbon sources.
+small RNAs found in our study could be involved in the regulation of CcpA.
+
+Our study provides more precise landscape of transcriptome changes in response to
+different carbon sources in \textit{S. mutans}.
+
 The knowledge that we gained from the current study of \textit{S. mutans}
@@
 shown to regulate metabolism and virulence gene expression in 
-\textit{S.\ mutans} through CCR in transcriptom study using 
+\textit{S.\ mutans} through CCR in transcriptome study using 
 the microarray technique based on DNA hybridization \citep{Abranches2008}.
@@
 
-\subsection*{Experimental results from Burne group}
-\fixme{Add some if needed.}
+<<eval=FALSE,echo=FALSE,result=hide>>=
+# To find Infernal small RNAs expressed in RNA-seq data
+grTx <- smutansData.txGrCombined[!(elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5)]
+smutansData.txInfernal <- smutansData.txAnnotation[1:274]
+ol.Infernal <- findOverlaps(smutansData.txInfernal,grTx,type="within")
+r <- ranges(grTx[unique(ol.Infernal@subjectHits)])
+for (i in seq(length(r))) {
+  cat(paste("chr1:",start(r)[i],"-",end(r)[i]),"\n")
+}
+#chr1: 22518 - 23227 tRNA
+#chr1: 23229 - 25062 tRNA
+#chr1: 97029 - 99661 mir
+#chr1: 99934 - 100544 tRNA
+#chr1: 111689 - 118470 mir
+#chr1: 152804 - 153769 * STnc490k different position
+#chr1: 182583 - 185313 mir
+#chr1: 190870 - 191839 tRNA
+#chr1: 293153 - 293660 * SRP_bact SRP_euk_arch different position
+#chr1: 322211 - 328546 tRNA
+#chr1: 413043 - 415691 * T-box
+#chr1: 552584 - 553935 mir
+#chr1: 626450 - 630612 * P15 (ambiguous) 
+#chr1: 660501 - 661809 * L20_leader (strong) LR-PK1 (ambiguous)
+#chr1: 666062 - 667070 * TPP (ambiguous but strong)
+#chr1: 710297 - 712890 mir
+#chr1: 796924 - 798060 * L21_leader (ambiguous)
+#chr1: 800164 - 814296 * PyrR (ambiguous)
+#chr1: 938383 - 938619 mir
+#chr1: 995160 - 998898 snoR72 (ambiguous)
+#chr1: 1115782 - 1117618 * Glycin (ambiguous)
+#chr1: 1140867 - 1142427 tRNA
+#chr1: 1142434 - 1149502 tRNA
+#chr1: 1163587 - 1166184 PyrR (ambiguous)
+#chr1: 1209827 - 1214087 tRNA
+#chr1: 1214631 - 1215039 * L19_leader (ambiguous)
+#chr1: 1326825 - 1328325 * CRISPER (ambiguous)
+#chr1: 1410520 - 1420619 * His_leader (ambiguous)
+#chr1: 1439314 - 1447074 * T-box (ambigous)
+#chr1: 1447225 - 1451227 Thr_leader
+#chr1: 1487846 - 1488056 mir
+#chr1: 1605925 - 1606244 mir
+#chr1: 1615311 - 1618362 23S-methyl
+#chr1: 1654702 - 1655118 tRNA
+#chr1: 1677566 - 1678101 * T-box (ambiguous)
+#chr1: 1678104 - 1678469 tRNA
+#chr1: 1758499 - 1759936 mir
+#chr1: 1783980 - 1787743 mir
+#chr1: 1838690 - 1844933 * mir
+#chr1: 2023401 - 2026533 tRNA mir
 
-\subsection*{Results from Stanhope group when RNA sequencing preparation}
-\fixme{Add some if needed.}
+# To find RNAz predicted small RNAs expressed in RNA-seq data
+ol <- findOverlaps(smutansData.txRNAz,grTx,type="within")
+r <- ranges(grTx[unique(ol@subjectHits)])
+for (i in seq(length(r))) {
+  cat(paste("chr1:",start(r)[i],"-",end(r)[i]),"\n")
+}
+#chr1: 301108 - 303712 no
+#chr1: 303714 - 304038 * locus100 His_leader (weak)
+#chr1: 340223 - 341140 * locus101 
+#chr1: 378392 - 382328 
+#chr1: 413043 - 415691 * locus103 T-box (strong)
+#chr1: 480819 - 482162 * locus104 (ambiguous)
+#chr1: 681862 - 684666 * locus107 yybP-ykoY (ambiguous)
+#chr1: 742326 - 747734 * locus108 (ambiguous)
+#chr1: 908974 - 911426 * locus110 (weak)
+#chr1: 1036181 - 1038689 locus2
+#chr1: 1209827 - 1214087 locus4 tRNA (false?)
+#chr1: 1304269 - 1305793 locus7 (ambiguous)
+#chr1: 1678104 - 1678469 locus53 tRNA
+#chr1: 1783980 - 1787743 locus69
 
-\subsection*{Cluster of RNA samples}
+# Expressed regions without annotations
+smutansData.txAllAnnotation <- c(smutansData.txRNAz,smutansData.txAnnotation)
+strand(smutansData.txAllAnnotation) <- '*'
+smutansData.txAllAnnotation <- gaps(reduce(smutansData.txAllAnnotation))
+smutansData.txAllAnnotation <- smutansData.txAllAnnotation[-2:-1]
+# Check sum(width(smutansData.txAllAnnotation)) is small or < 10% of 2Mbp.
+x <- c()
+for (i in seq(length(smutansData.txAllAnnotation))) {
+  s <- start(ranges(smutansData.txAllAnnotation))[i]
+  e <- end(ranges(smutansData.txAllAnnotation))[i]
+  x <- c(x,sum(smutansData.txPileup[s:e])/(e-s+1))
+}
+elementMetadata(smutansData.txAllAnnotation)[,"score"]<-x
+smutansData.txNoAnnotation <- smutansData.txAllAnnotation[order(elementMetadata(smutansData.txAllAnnotation)[,"score"])]
+for (i in seq(length(smutansData.txNoAnnotation))) {
+  if (abs(start(smutansData.txNoAnnotation)[i] - end(smutansData.txNoAnnotation)[i]) > 50) {
+    cat(paste("chr1:",start(smutansData.txNoAnnotation)[i],"-",end(smutansData.txNoAnnotation)[i]),"\n")
+  }
+}
+# This does not work. I think that I have to manually check.
+strand(smutansData.txNoAnnotation) <- '+'
+x <- elementMetadata(smutansData.txNoAnnotation)[,"score"]
+x <- as.integer(x*1000/max(x))
+elementMetadata(smutansData.txNoAnnotation)[,"score"] <- x
+smutansData.txNoAnnotation <- sort(smutansData.txNoAnnotation)
+elementMetadata(smutansData.txNoAnnotation)[,"name"] <- seq(length(smutansData.txNoAnnotation))
+export.bed(smutansData.txNoAnnotation,con="b.bed")
+x <- c(46, 69, 73, 74, 75, 76, 80, 97, 103, 212, 261, 280, 288, 317, 440, 450, 478,
+484 , 628 , 753 , 1011, 1019 , 1037 , 1084, 1097, 1176, 1291, 1444, 1508,
+1518, 1600, 1605, 1638, 1645, 1690)
+for (i in x) {
+  cat(paste("chr1:",start(smutansData.txNoAnnotation)[i],"-",end(smutansData.txNoAnnotation)[i]),"\n")
+}
 
-First using the table of read counts of annotated genes we formed a cluster of
-the 13 RNA-seq samples (fig.\ \ref{figClust}). The effect of nutrients appeared
-to be stronger than that of the deletion of \textit{ccpA} genes from the UA159
-genome.
+# 46
+# 69 * strong
+# 73 ambiguous
+# 74 very weak
+# 75 ambiguous
+# 76 ambiguous
+# 80 * weak
+# 97 *
+# 103 *
+# 212 * strong
+# 261 *
+# 280 *
+# 288 * one gene?
+# 317 * in between tRNAs
+# 440 weak
+# 450 * strong
+# 478 * strong
+# 484 * strong
+# 628 *
+# 753 *
+# 1011 * strong
+# 1019 * tail of a gene?
+# 1037 * strong?
+# 1084 * strong
+# 1097 * left? PCR amplification error?
+# 1176 * start or end of genes?
+# 1291 * end of the gene?
+# 1444 * 
+# 1508 * strong
+# 1518 * strong
+# 1600 * start of the gene?
+# 1605 * new gene in between genes?
+# 1638 * start of the gene?
+# 1645 * start of the gene?
+# 1690 * in between genes?
+
+for (i in seq(length(txRNAz.unannotated))) {
+  cat(paste("# ",names(txRNAz.unannotated)[i],"  chr1:",start(txRNAz.unannotated)[i],"-",end(txRNAz.unannotated)[i]),"\n")
+}
+#  locus99   chr1: 302288 - 302403  no
+#  locus100   chr1: 303745 - 303859 did
+#  locus101   chr1: 340263 - 340343 did
+#  locus102   chr1: 378452 - 378505 good one
+#  locus104   chr1: 481458 - 481539 not a good candidate
+#  locus106   chr1: 674067 - 674135 maybe one
+#  locus108   chr1: 747483 - 747567 not a good one
+#  locus110   chr1: 911157 - 911258 not a good one
+#  locus2   chr1: 1036231 - 1036283 not a good one
+#  locus5   chr1: 1244007 - 1244086 not a good one
+#  locus6   chr1: 1297150 - 1297227 not a good one
+#  locus7   chr1: 1304792 - 1304926 not a good one
+#  locus8   chr1: 1313525 - 1313674 not a good one
+#  locus10   chr1: 1564032 - 1564149 not a good one
+#  locus69   chr1: 1787575 - 1787687 not a good one
+#  locus85   chr1: 1864317 - 1864461 not a good one
+@
+
+\subsection*{Small RNAs expressed in RNA-seq}
+
+We compiled 274 predicted RNAs by using cmsearch program from Infernal (v1.0)
+with default options for models in RFAM 9.1 \citep{Gardner2009,Nawrocki2009}.
+By intersecting the predicted RNAs locations and RNA-seq expressed regions,
+we found more evidence for some of the predicted RNAs (table \ref{tab:infernal}).
+Several RNAs were shown to be clearly expressed by RNA-seq experiments:
+T-box strong, RNaseP\_bact\_a, RNaseP\_bact\_b, L20\_leader, tmRNA, FMN, and 6S.
+Other RNAs, TPP and Glycin, were located close to their neighboring genes,
+making it hard to discriminate them from the genes.
+
+Additionally, we predicted 
+\Sexpr{length(smutansData.txRNAz)}
+structural RNAs using RNAz, most of which were overlapped with those from
+the Infernal prediction. We found a few of them to be expressed based on RNA-seq
+experiments (table \ref{tab:RNAz}).
+Region locus100 had tens of short reads mapped on it, which was near
+His\_leader predicted by Infernal. 
+Although region locus101 was ambiguous in its location with respect to its neighboring
+genes, its predicted secondary structure hinted at its potential as a functional
+RNA. Region locus103 was expressed clearly compared with neighboring regions; the
+region was predicted as T-box by Infernal.
+Although region locus107 was ambiguous in its location with respect to its neighboring
+genes expression, its good secondary structure and yybP-ykoY predicted by
+Infernal hinted at the RNA functional potential of the region.
+
+\subsection*{Unannotated regions expressed in RNA-seq}
+
+We focused 16 regions that had not been annotated either by previously known
+genes or by the potentially functional RNA prediction of our current study
+(table \ref{tab:unannotated}). Four regions inlcuding novel6,
+novel7, novel10, and novel11 showed that their expression levels were not
+clearly dicriminated from their flanking genes. 
+Located in between SMU.305 and SMU.307 was, for example, novel6, which was
+connected in expression level by large number of short reads with SMU.307.
+The other 12 regions appeared to show distinct expression levels compared with
+flank regions. Region novel1 was situated in between SMU.61 and SMU.63c, which
+had different strands.
+Although seemingly separated from either of the two genes, it was connected by
+links of average about 20 depths of short reads with the either of them.
+Regions novel2, novel13, and novel15 appeared to be expressed very weakly.
+A few regions were neighboring known RNAs: 
+novel3 located in between SMU.97 and SMU.t19,
+novel4 located in between SMU.t20 and SMU.107, and
+novel6 situated to the right of
+SRP\_bact and SRP\_euk\_arch predicted by Infernal.
+We also examined regions in between known CDS or tRNAs, which could be
+unannotated regions but likely to be in operon with its neighboring genes
+(table \ref{tab:inbewteencds}).
+
+% Of the \Sexpr{length(colnames(smutans.genes.criteria))-1} predicted non-coding RNAs
+% \Sexpr{length(txRNAz.unannotated)} were in  
+% unannotated non-coding regions (table \ref{tabUnannotatedRNA}).
+
+Using
+\program{RNAplex} \citep{Tafer2008} and \program{RNAplfold} \citep{Bernhart2006}
+we predicted target genes of the putative non-coding RNAs.  Table \ref{tab:goseq}
+shows functional categories enriched with target genes for predicted small RNAs.
+
+We found that a few of the putative small RNAs were structured with stable stems
+(figs.\ \ref{figRNALocus1}, \ref{figRNALocus2}, and \ref{figRNALocus3}).
 
@@
 
-\subsection*{Predictions of small RNAs}
+The automatically defined transcripts using site-wise expression values by the
+hidden Markov model and mapped short reads displayed using UCSC genome browser
+tracks guided us to define transcripts.
 
-We predicted
-\Sexpr{length(colnames(smutans.genes.criteria))-1} putative non-coding RNAs, most of
-which were already annotated.  
-Of the \Sexpr{length(colnames(smutans.genes.criteria))-1} predicted non-coding RNAs
-\Sexpr{length(txRNAz.unannotated)} were in  
-unannotated non-coding regions (table \ref{tabUnannotatedRNA}).  Using
-\program{RNAplex} \citep{Tafer2008} and \program{RNAplfold} \citep{Bernhart2006}
-we predicted target genes of the putative non-coding RNAs.  Table \ref{tab:goseq}
-shows functional categories enriched with target genes for predicted small RNAs.
-We found that a few of the putative small RNAs were structured with stable stems
-(figs.\ \ref{figRNALocus1}, \ref{figRNALocus2}, and \ref{figRNALocus3}).  
-\fixme{We might be able to 
-further experimentally verify potentially important small RNAs.}
+tx0391
+tx0392
 
 \subsection*{Differential expression of RNA transcripts}
+
+First using the table of read counts of annotated genes we formed a cluster of
+the 13 RNA-seq samples (fig.\ \ref{figClust}). The effect of nutrients appeared
+to be stronger than that of the deletion of \textit{ccpA} genes from the UA159
+genome.
 

### 2012-04-24 package/supplementary.Rnw
Just an update.

@@
 @
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Infernal predicted small RNAs that appeared to be expressed in UA159 Glucose
+RNA-seq.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cl}
+\hline 
+Position & Description \\
+\hline 
+chr1: 413043 - 415691 & T-box \\
+chr1: 439797 - 440210 & RNaseP\_bact\_a, RNaseP\_bact\_b \\
+chr1: 659901 - 661209 & L20\_leader \\
+chr1: 666062 - 667070 & TPP (ambiguous) \\
+chr1: 1115846 - 1116774 & Glycin (ambiguous) \\
+chr1: 1139350 - 1139673 & tmRNA \\
+chr1: 1202719 - 1202902 & T-box \\
+chr1: 1615122 - 1615307 & FMN \\
+chr1: 1927964 - 1928120 & 6S \\
+\hline 
+\end{tabular}
+\par\end{centering}
+\label{tab:infernal}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf RNAz predicted small RNAs that appeared to be expressed in UA159 Glucose
+RNA-seq.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cll}
+\hline 
+ID & Position & Infernal prediction \\
+\hline 
+locus100 & chr1: 303714 - 304038 & His\_leader \\
+locus101 & chr1: 340223 - 341140 & \\
+locus103 & chr1: 413043 - 415691 & T-box \\
+% locus104 & chr1: 480819 - 482162 & \\
+locus107 & chr1: 681862 - 684666 & yybP-ykoY \\
+% locus85 & chr1: 1864317 -1864461 & \\
+\hline 
+\end{tabular}
+\par\end{centering}
+\label{tab:RNAz}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Expressed regions of unannotated regions in UA159 Glucose
+RNA-seq.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cll}
+\hline 
+ID & Position & Description \\
+\hline 
+novel1 & chr1: 62546 - 63078 & clear, strong \\
+novel2 & chr1: 69293 - 69630 & weak \\
+novel3 & chr1: 99553 - 100331 & clear, strong \\
+novel4 & chr1: 108631 - 109107 & clear, strong \\
+novel5 & chr1: 211452 - 211769 & clear, strong \\
+novel6 & chr1: 293238 - 293666 & vague, strong \\
+novel7 & chr1: 484053 - 484635 & clear, strong \\
+novel8 & chr1: 523427 - 524198 & clear, weak \\
+novel9 & chr1: 530356 - 530591 & clear, weak \\
+novel10 & chr1: 702752 - 703471 & vague, strong \\
+novel11 & chr1: 1202920 - 1203816 & vague \\
+novel12 & chr1: 1254303 - 1255083 & clear, strong \\
+novel13 & chr1: 1370933 - 1371532 & weak \\
+novel14 & chr1: 1757571 - 1758606 & clear, strong \\
+novel15 & chr1: 1769455 - 1770330 & clear, weak \\
+novel16 & chr1: 76962 - 77575 & SMU.76 and SMU.78 \\
+% & chr1: 1291334 - 1291685 & PCR amplification error \\
+% & chr1: 1689061 - 1689404 & PCR amplification error \\
+\hline 
+\end{tabular}
+\par\end{centering}
+\label{tab:unannotated}
+\end{table}
+%-----------------------------------------------------------------------------
+
+\clearpage{}
+%-----------------------------------------------------------------------------
+\begin{table}[!h]
+\caption{{\bf Regions expressed in between CDS regions in UA159 Glucose
+RNA-seq.}}
+\vspace{1ex}
+\noindent \begin{centering}
+\begin{tabular}{cll}
+\hline 
+ID & Position & Description \\
+\hline 
+bt1 & chr1: 68950 - 69214 & between SMU.67 and SMU.68 \\
+bt2 & chr1: 268249 - 269322 & between SMU.281 and SMU.283 \\
+bt3 & chr1: 301821 - 302287 & between SMU.317 and SMU.318 \\
+bt4 & chr1: 328198 - 328359 & between SMU.t34 and SMU.t35 \\
+bt5 & chr1: 473362 - 473629 & start of SMU.503c \\
+bt6 & chr1: 869137 - 869771 & start of SMU.911c \\
+bt7 & chr1: 1176217 - 1176736 & start of SMU.1237c \\
+bt8 & chr1: 1185233 - 1185628 & end of SMU.1249c \\
+bt9 & chr1: 1534290 - 1534622 & end of SMU.1604.c \\
+bt10 & chr1: 1870392 - 1871025 & between SMU.1995c and SMU.1996 \\
+bt11 & chr1: 1893131 - 1893647 & start of SMU.2026c \\
+\hline 
+\end{tabular}
+\par\end{centering}
+\label{tab:inbewteencds}
+\end{table}
+%-----------------------------------------------------------------------------
 

### 2012-04-24 package/tutorial.Rnw
Just an update.

@@
 
-\title{\textsf{\textbf{Analyzing \textit{Streptococcus mutans} RNA-seq data\\using R packages}}}
+\title{\textsf{\textbf{A Toturial of BioConductor R packages}}}
 
@@
 \begin{abstract}
-We analyze RNA samples from \textit{Streptococcus mutans}.  We use
-RNA-seq to define genes whose expression changes between samples
-\cite{Oshlack2010}. Short reads are first mapped to a reference genome using
-approaches such as \program{BWA} \cite{Li2009a}.  Reads fitting into one of the annotated
-intervals are counted.  These counts are passed to a statistical package which
-treats differences in read counts for statistical significance. For this
-purpose, we made use of \Rpackage{DEseq} \cite{Anders2010} or \Rpackage{edgeR}
-\cite{Robinson2010}. This prodecure returns a list of differentially expressed
-genes, which can be passed to downstream analysis, such as functional category
-enrichment test using \Rpackage{goseq} \cite{Young2010}.  The current
-\Rpackage{smutans} contains procedures that use functions provided by
-the above R packages. 
+BioConductor R packages have been useful in analyzing RNA-seq data sets. I wish
+to record for the instructional purposes.
 \end{abstract}
-
 \tableofcontents
 
-%--------------------------------------------------
-\section{Data preparations} \label{sec:prep}
-%--------------------------------------------------
+\section{Genomes}
+
+\subsection{Sample sequences from a genome}
+
+\section{Short reads}
 
 Steps of analyzing RNA-seq data for determining differentially expressed genes
-are described in \cite{Oshlack2010}. We used a short read aligner called
-\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
-\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
-aligner allowed a few gaps for efficient alignment of millions of sequences of
-size being 100 base pairs long, short reads that consist of RNAs and sequencing
-adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
-remove parts of adapter sequences from short reads before mapping on the
-reference genome. Mapped short read alignments were converted into readable
-formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
-reference genome, forming a table of read counts of the genes. Statistical
-software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
-was subsequently employed to infer differentially expressed in different
-biological conditions.  We also generated tracks for our recently released
-Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
-summarize gene expression levels based the short read alignments. This allowed
-us to visualize parts of the reference genome for comparing different RNA
-samples. Further analyses such as functional category enrichment test
-\cite{Young2010} were also employed.
+are described in \cite{Oshlack2010}.
 
-The \Rpackage{smutans} expects count data in the form of \Rclass{CountDataSet},
-which is a type provided by \Rpackage{DESeq}.  Using \program{BWA} \cite{Li2009a} and
-\program{samtools} \cite{Li2009b} we aligned short reads from RNA sequencing. Using gene
-annotations in the UA159 genome we counted short reads that overlapped the
-genes.  The count values are raw counts of sequencing reads. The
-\Rpackage{DESeq}'s statistical model, or negative binomial model, hold for raw
-reads that allows to assess the measurement precision correctly; we did not have
-rounded values of normalized counts, or counts of covered base pairs.  Several
-exeriments of \textit{S.\ mutans} RNA sampling were performed.  \fixme{Find the
-papers that are related with the experiments.} One of them was presented in the
-\Robject{smutansGenes} as a part of \Rpackage{smutans}. The
-\Robject{smutansGenes} is of \Rclass{CountDataSet}, which is the data type used
-by \Rpackage{DESeq}. We have a convenient \Rclass{Smutans}, which contains slots
-for \Rclass{CountDataSet} objects, and results from analyses thereof.  Before
-loading necessary R packages and data sets, let us describe what functions were
-used for preparing data objects such as \Robject{smutansGenes}. One should
-create an R script file at the directory that contains the source code directory
-of the \Rpackage{smutans}. Put the following scripts in the R script file and
-run the script at that directory. That should create RData files at the
-directory smutans/data. User should not run the R script because the data files
-should be found in the \Rpackage{smutans}.
+\subsection{Create short reads}
 
-% http://www.oralgen.lanl.gov/oralgen/downloads/smut.csv
-% http://www.oralgen.lanl.gov/cgi-bin/gene_id_search.cgi?dbname=smut&gene_id=SMu0001
-% \fixme{Where are flat files for genes of S. mutans UA159?}
-<<samplePreparation,eval=FALSE>>=
-# not run
-# but create a R scirpt file with the following scripts, and
-# run it, i.e., Rscript run.R
-library(DESeq)
-source("smutans/R/core.R")
-smutans.prepareData34( )
-smutans.prepareData835NPP( )
-smutans.prepareDataUA159TW1( ) 
-smutans.prepareGoseq( )
-smutans.prepareDataOMZ175( ) 
-smutans.prepareDataUA159CSP( ) 
-smutans.prepareDataSMU86CSP( )
-q("no")
-@
-<<options,results=hide,echo=FALSE>>=
-options(digits=3)
-@
-Now, see some of the data objects by loading the \Rpackage{smutans}, and a data
-\Robject{smutansGenes}. Note that the object is of type \Rclass{CountDataSet} of
-\Rpackage{DESeq} in which users should find \Rfunction{counts} and
-\Rfunction{pData}. 
-<<smutans,results=hide>>=
-library( "smutans" )
-data( "smutansGenes" )
-@
-There are 
-\Sexpr{length(colnames(counts(smutansGenes)))}
-RNA-seq samples in \Robject{smutansGenes}, which
-contains the counts and also metadata about the samples:
-<<look_into_smutansGenes>>=
-head( counts(smutansGenes) )
-@
-<<look_into_smutansGenes>>=
-pData( smutansGenes )
-@
-The samples differ by experimental condition (``glucose'' or
-``galactose'') and by library type (``ua159'' or ``tw1'').
+\subsection{Read short reads}
 
-Samples of 835NP, 835P, and UA159 no CSP should be loaded using the following
-command.
-<<sm835,results=hide>>=
-data( "sm835Genes" )
-@
+\Rpackage{ShortRead}
 
-% Resources
-% Google charts - Venn Diagram
-% http://code.google.com/apis/chart/index.html
+library(ShortRead)
+fastqFile <- "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.recovered.fq.gz"
+fq <- readFastq(fastqFile, withIds=TRUE)
+head(sread(fq), 3)
+head(quality(fq), 3)
+sampler <- FastqSampler(fastqFile, 100000)
+fqSample <- yield(sampler)
+report(qa(fqSample),dest="/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/report001")
 
-\fixme{Describe other data objects as well.
-sm34Genes.RData,
-sm835Genes.RData,
-smutans.cat.desc.RData,
-smutans.feature.genes.RData, and
-smutans.go.genes.RData.}
+qa2 <- qa(fq, basename(fastqFile)) 
+qa3 <- qa(fqSample, basename(fastqFile)) 
+report(qa2,dest="/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/report001")
+X11 error in report
 
-%--------------------------------------------------
-\section{DE analysis using UA159 and TW1 samples with
-glucose and galactose} \label{sec:detw1}
-%--------------------------------------------------
+indexBam(reads_fn)
 
-We could compare four pairwise comparisons using the two-factor data set of
-\Robject{smutansGenes}: a pair of comparisons between ``glucose'' and
-``galactose'' given either type ``ua159'' or ``tw1'', and another pair of
-comparisons between types ``ua159'' and ``tw1'' given either condition
-``glucose'' or ``galactose.'' We focus on one of the four pairwise comparisons
-first. Later, we will perform analyses for the rest of the three comparisons.
-We look at the ``ua159'' data for now, and will look at the ``tw1'' data later.
+Reweight each read.
 
-We can cluster samples using variance stabilized data. Figure \ref{figClust}
-shows the \Sexpr{length(colnames(counts(smutansGenes)))} sampled clustered. 
-The clustering should reflect our experimental design, i.e., samples are
-more similar when they have the same treatment or the same library type.
-To make this conclusion, we estimate the dispersion with
-mode ``blind'' (see \Rpackage{DESeq} vignette), as only then,
-the variance stabilizing transformation is not informed about the design and
-hence not biased towards a result supporting it. Such an analysis is useful
-for quality control.  This appears to be useful as a first in any analysis.
+\subsection{Sample short reads}
 
-\begin{figure}
-\centering
-\includegraphics[width=.9\textwidth]{smutans-figClust}
-\caption{Heatmap showing the Euclidean distances between the samples
-  as calculated from the variance-stabilizing transformation of the
-  count data.}
-\label{figClust}
-\end{figure}
-<<figClust,fig=TRUE>>=
-allSample <- newSmutans( smutansGenes )
-smutans.de2Clust( allSample )
-@
-As expected, two experimental conditions of ``glucose'' and ``galactose'' divide the
-samples. Within each condition, two different types of ``ua159'' and ``tw1'' divide
-them again. The pair of ``glucose ua159'' and ``galactose tw1'' is the most
-distinguished.
+\subsection{Correct bias in short reads}
 
-%------------------------------------------------------------
-\subsection{Inference using single-factor data} 
-\label{subsec:ua159only}
-%------------------------------------------------------------
+library(seqbias) 
+library(Rsamtools) 
+ref_fn <- "/v4scratch/sc2265/rnaseq/output/ua159/1/data/NC_004350.fna"
+ref_f <- FaFile( ref_fn ) 
+open.FaFile( ref_f )
+reads_fn <- "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.sorted.bam"
+indexBam(reads_fn)
+ref_seqs <- scanFaIndex( ref_f )
 
-We create a \Rclass{Smutans} object using the \Rclass{CountDataSet}
-\Robject{smutansGenes}. The title option would be useful in distinguishing
-different objects of \Rclass{Smutans} as we have more R objects with the type.  
-A subsequent call to
-\Rfunction{smutans.de2} uses a two-factor data set to test differential
-expression of genes.  The \Rfunction{smutans.de2} estimates effective library
-sizes, and dispersions using functions provided by \Rpackage{DESeq}. Then, it
-tests genes for differential expression.
-<<ua159onlySamples>>=
-ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
-ua159only <- smutans.de2( ua159only, type="ua159", 
-                          condA="glucose", condB="galactose" )
-@
-Because \Rpackage{DESeq} recommended to check
-the dispersion estimates and to make sure that the data quality is as
-expected, we plot the per-gene estimates against the normalized mean
-expressions per gene, and then overlay the fitted curve in red color
-(Fig.\ \ref{figFit}).
+ref_seq <- getSeq(ref_f)
+I.all <- GRanges(seqnames=Rle(c("NC_004350.1"),c(2)),
+                 ranges=IRanges(c(1,1),width=rep(width(ref_seq),2)),
+                 strand=Rle(strand(c("+","-")),c(1,1)))
+seqlengths(I.all) <- c(width(ref_seq))
 
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figFit}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength for the data set of UA159-only sample
-to compare glucose and galactose conditions.}
-\label{figFit}
-\end{figure}
-<<figFit,fig=TRUE>>=
-smutans.plotDispersionEstimates(ua159only)
-@
-\fixme{The following paragraph is adopted from \Rpackage{DESeq} vignette.}
-As we estimated the dispersion from only two samples, we should expect the
-estimates to scatter with quite some sampling variance around their true values.
-Hence, \Rpackage{DESeq} should not use the per-gene estimates directly in the
-test, because using too low dispersion values leads to false positives.  Many of
-the values below the red line are likely to be underestimates of the true
-dispersions, and hence, it is prudent to instead rather use the fitted value.
-On the othe hand, not all of the values above the red line are overestimations,
-and hence, the conservative choice is to keep them instead of replacing them
-with their fitted values.
-Having estimated the dispersion for each gene, we look for differentially
-expressed genes. To contrast two conditions, e.g., to see whether there is
-differential expression between conditions ``glucose'' and ``galactose'',
-\Rfunction{smutans.de2} already performed the 
-\Rfunction{nbinomTest} provided by \Rpackage{DESeq}.  
-Let's list genes with resulting information from \Rpackage{DESeq}.
-<<ua159onlyDE2List>>=
-head(smutans.de2List( ua159only ))
-@
-For each gene, we get its mean expression level (at the base scale) as a joint
-estimate from both conditions, and estimated separately for each condition, the
-fold change from the first to the second condition, i.e., 
-$\frac{baseMeanB}{baseMeanA}$, the logarithm (to basis 2)
-of the fold change, and the $p$ value for the statistical significance of this
-change. The \texttt{padj} column contains the $p$ values, adjusted for multiple
-testing with the Benjamini-Hochberg procedure (see the R function
-\Rfunction{p.adjust}), which controls false discovery rate (FDR). Later, we
-print the gene list out in a CSV file using option file of
-\Rfunction{smutans.de2List}.
-%-----------------------------------------------------------------------------
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figDE}
-\caption{Plot of normalised mean versus $log_2$ fold change
-(this plot is sometimes also called the ``MA-plot'') for the contrast ``glucose''
-versus ``galactose'' using UA159-only sample.}
-\label{figDE}
-\end{figure}
-%-----------------------------------------------------------------------------
-<<figDE,fig=TRUE>>=
-smutans.plotDiffExp(ua159only)
-@
-Figure \ref{figDE} shows a plot of the $\log_2$ fold changes against the base
-means, coloring in red those genes that are significant at 10\% of FDR. 
-Let us use ``tw1'' type to compare ``glucose'' and ``galactose'' conditions.
-We continue to analyze the data set using type ``tw1'' to compare the two
-conditions. We perform the single-factor analysis. 
-<<tw1onlySamples>>=
-tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
-tw1only <- smutans.de2( tw1only, type="tw1", 
-                        condA="glucose", condB="galactose" )
-tabUA159onlyTW1only <- smutans.compareResult( ua159only, tw1only )
-addmargins( tabUA159onlyTW1only )
-@
-The table, \Robject{tabUA159onlyTW1only}, shows numbers of genes found to be 
-differentially expressed in the two analyses: ``UA159 Only'' and ``TW1 Only.''
-Of the total \Sexpr{sum(tabUA159onlyTW1only)} genes, \Sexpr{tabUA159onlyTW1only[2,2]}
-genes were found in both analyses. The analysis using TW1-only sample found
-\Sexpr{tabUA159onlyTW1only[1,2]} genes. The analysis using UA159-only sample found
-\Sexpr{tabUA159onlyTW1only[2,1]} genes.
+I <- random.intervals( ref_seqs, n = 5, m = 100000 )
+seqs <- scanFa( ref_f, I )
+neg_idx <- as.logical( I@strand == '-' ) 
+seqs[neg_idx] <- reverseComplement( seqs[neg_idx] )
+counts <- count.reads( reads_fn, I, binary = T )
+freqs <- kmer.freq( seqs, counts )
+sb <- seqbias.fit( ref_fn, reads_fn, L = 5, R = 20 )
+bias <- seqbias.predict( sb, I )
+bias <- seqbias.predict( sb, I.all )
+bias[[2]][99996]
+seqbias.save(sb,"/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.yml")
+sb <- seqbias.load( ref_fn, "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.yml")
+sb5 <- seqbias.fit( ref_fn, reads_fn, n=5e5, L = 5, R = 20 )
 
-Until now we focused on the effect of condition either using UA159-only or
-TW1-only samples. We now change the gear to see effects of sample types on
-differentially expressed genes. We create two \Rclass{Smutans} objects:
-\Robject{glucoseOnly} and \Robject{galactoseOnly}, and compare them in detecting
-differentially expressed genes:
-<<glucoseOnlySamples>>=
-glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
-glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
-                            condA="ua159", condB="tw1" )
-galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
-galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
-                              condA="ua159", condB="tw1" )
-tabGlucoseOnlyGalactoseOnly <- smutans.compareResult( glucoseOnly, 
-                                                      galactoseOnly )
-addmargins( tabGlucoseOnlyGalactoseOnly )
-@
-Of the total \Sexpr{sum(tabGlucoseOnlyGalactoseOnly)} genes,
-\Sexpr{tabGlucoseOnlyGalactoseOnly[2,2]} genes were found in both analyses at FDR
-rate of 10\%. The analysis using galactose-only sample found
-\Sexpr{tabGlucoseOnlyGalactoseOnly[1,2]} genes. The analysis using glucose-only
-sample found \Sexpr{tabGlucoseOnlyGalactoseOnly[2,1]} genes. 
-Table \ref{tab:de} summarizes the number of differentially expressed genes that
-were found using single-factor analyses at FDR of 10\%.
+N <- width(ref_seq) # genome length
+cnt <- count.reads(reads_fn, I.all, sb = sb5)
+cnt <- count.reads(reads_fn, GRanges("seq1", IRanges(start = 1, width = N)), sb = sb5)
 
-%-----------------------------------------------------------------------------
-\begin{table}[!h]
-\caption{{\bf Numbers of differentially expressed genes in different analyses}}
-\vspace{1ex}
-\noindent \begin{centering}
-\begin{tabular}{ccccl}
-\hline 
-Sample A & Sample B & Number of Genes & Result File \\
-\hline 
-UA159 Glucose & UA159 Galactose & \Sexpr{sum(ua159only@padj < .1,na.rm=T)} & ua159only.csv \\
-TW1 Glucose & TW1 Galactose & \Sexpr{sum(tw1only@padj < .1,na.rm=T)} & tw1only.csv \\
-UA159 Glucose & TW1 Glucose & \Sexpr{sum(glucoseOnly@padj < .1,na.rm=T)} & glucoseOnly.csv \\
-UA159 Galactose & TW1 Galactose & \Sexpr{sum(galactoseOnly@padj < .1,na.rm=T)} & galactoseOnly.csv \\
-% UA159 Glucose & UA159 Galactose & \Sexpr{sum(tabUA159onlyTW1only[2,])} & ua159only.csv \\
-% TW1 Glucose & TW1 Galactose & \Sexpr{sum(tabUA159onlyTW1only[,2])} & tw1only.csv \\
-% UA159 Glucose & TW1 Glucose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[2,])} & glucoseOnly.csv \\
-% UA159 Galactose & TW1 Galactose & \Sexpr{sum(tabGlucoseOnlyGalactoseOnly[,2])} & galactoseOnly.csv \\
-\hline 
-\end{tabular}
-\par\end{centering}
-\label{tab:de}
-\end{table}
-%-----------------------------------------------------------------------------
+readlen <- 100
+cov <- coverage(GRanges("NC_004350.1", IRanges(start = 1:N, width = readlen)), weight = cnt[[1]])
 
-%------------------------------------------------------------
-\subsection{Inference using all of the samples} 
-\label{subsec:allSample}
-%------------------------------------------------------------
+\subsection{Compute the coverage of short reads}
 
-It is inconvenient to use four different analyses to find differentially
-expressed genes under one of two factors. Besides, it would be desirable to use all of
-the data to find differentially expressed genes because using more samples might
-be more powerful in detecting differentially expressed genes.  Let us find genes
-differentially expressed in different conditions of ``glucose'' and
-``galactose'' using both types of ``ua159'' and ``tw1.'' We will compare these
-genes and those found in the single-factor analysis that we did previously,
-i.e., \Robject{ua159only}.
-Let's start with multi-factor the analysis by creating a new \Rclass{Smutans} object, called
-\Robject{allSample}. Although we already used \Robject{allSample} when
-clustering the samples, we recreate it for instructive purposes.
-<<allSamples,results=hide>>=
-allSample <- newSmutans( smutansGenes, 
-                         title="All Samples for Comparing Conditions" )
-allSample <- smutans.de2( allSample )
-@
-The \Rfunction{smutans.de2} uses a two-factor data set to
-perform two-factor analysis. The two factors are type and condition. We are here
-interested in condition effect. We could check 
-the dispersion estimates and to make sure that the data quality is as
-expected by plotting the per-gene estimates against the normalized mean
-expressions per gene, and then overlay the fitted curve.
-<<figFitAllSample,eval=FALSE>>=
-# not run
-smutans.plotDispersionEstimates(allSample)
-@
-%-----------------------------------------------------------------------------
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figCompareUA159onlyAndAllSample}
-\caption{Comparison of $p$ values (unadjusted) from the test using only the
-four paired-end samples and the test using all seven samples. We can see that
-the latter tend to be smaller, as expected from the higher power of the test with all samples.}
-\label{figCompareUA159onlyAndAllSample}
-\end{figure}
-%-----------------------------------------------------------------------------
-<<figCompareUA159onlyAndAllSample,fig=TRUE>>=
-pval.ua159only <- smutans.pval( ua159only )
-pval.allSample <- smutans.pval( allSample )
-bottom = function(x, theta=1e-12) pmax(x, theta)
-plot( bottom(pval.ua159only), bottom(pval.allSample), log="xy", pch=20, cex=.3 )
-abline(a=0, b=1, col="blue")
-@
-<<compareUA159onlyAndAllSample>>=
-tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
-                                                   allSample )
-addmargins( tabUA159onlyAndAllSample )
-@
-Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
-\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
-analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
-genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
-genes.  Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from
-the two-factor analysis versus those from the single-factor analysis. The
-p-values from the two-factor analysis are overall smaller than those from the
-single-factor analysis. We would expect that this was the case because of higher
-power of the two-factor analysis.
+bamFile <- "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.sorted.bam"
+aln <- readGappedAlignments(bamFile)
+aln <- as(aln, "GRanges")
+coverage(aln)
+cov <- coverage(aln, weight = cnt[[1]])
 
-We can show that taking the library type into account was important to have
-good detection power by doing the analysis again using the standard workflow,
-as outlined earlier, and without informing \Rpackage{DESeq} of the library
-types. This anlaysis is not necessary. However, it clearly shows the advantage of the
-two-factor analysis.
-<<allSampleWithoutType>>=
-allSampleWithoutType <- newSmutans( smutansGenes, "All Sample without Type" )
-allSampleWithoutType <- smutans.de2WithoutType( allSampleWithoutType,
-                                                condA="glucose", 
-                                                condB="galactose" )
-tabWithoutWithAllSample <- smutans.compareResult( allSampleWithoutType, 
-                                                  allSample )
-addmargins( tabWithoutWithAllSample )
-@
-Of the total \Sexpr{sum(tabWithoutWithAllSample)} genes,
-\Sexpr{tabWithoutWithAllSample[2,2]} genes were found in both analyses. The
-analysis using all of the sample found \Sexpr{tabWithoutWithAllSample[1,2]}
-genes. The analysis using all of the same sample but without considering type
-found only \Sexpr{tabWithoutWithAllSample[2,1]} genes.
+\subsection{Export coverage to a wiggle file}
 
-As we mentioned above, we can have two analyses using the two-factor analysis.
-Until now we compared conditions.
-We proceed to compare the two types by considering all of the samples.
-<<allSampleType,results=verbatim>>=
-allSampleType <- newSmutans( smutansGenes, title="All samples for type" )
-allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
-tabGlucoseOnlyAllSample <- smutans.compareResult( glucoseOnly, 
-                                                  allSampleType )
-addmargins( tabGlucoseOnlyAllSample )
-tabGalactoseOnlyAllSample <- smutans.compareResult( galactoseOnly, 
-                                                    allSampleType )
-addmargins( tabGalactoseOnlyAllSample )
-@
-Of the total \Sexpr{sum(tabGlucoseOnlyAllSample)} genes,
-\Sexpr{tabGlucoseOnlyAllSample[2,2]} genes were found in both analyses. The
-analysis using all of the sample found \Sexpr{tabGlucoseOnlyAllSample[1,2]}
-genes. The analysis using glucose-only sample
-found \Sexpr{tabGlucoseOnlyAllSample[2,1]} genes.  
-In the comparison using galactose-only sample,
-of the \Sexpr{sum(tabGalactoseOnlyAllSample)} genes,
-\Sexpr{tabGalactoseOnlyAllSample[2,2]} genes were found in both analyses. The
-analysis using all of the sample found \Sexpr{tabGalactoseOnlyAllSample[1,2]}
-genes. The analysis using galactose-only sample
-found \Sexpr{tabGalactoseOnlyAllSample[2,1]} genes.
-
-We print out the result of differentially expressed genes in CSV files. A
-spreadsheet program such as EXCEL should be able to open CSV files. Users who
-are more comfortable with the EXCEL program than R could post-process CSV
-files for sorting and finding differentially expressed genes using different
-q-value cutoff values.
-<<smutanSamples>>=
-smutans.de2List( allSample, "allSample.csv" )
-smutans.de2List( allSampleType, "allSampleType.csv" )
-smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
-smutans.de2List( galactoseOnly, "galactoseOnly.csv" )
-smutans.de2List( ua159only, "ua159only.csv" )
-smutans.de2List( tw1only, "tw1only.csv" )
-head( smutans.de2List( allSample ) )
-@
-The output from \Robject{allSample} is different from that of
-\Robject{ua159only} or \Robject{tw1only}.
-The first column of ``allSample.csv'' is for genes.
-The next three columns show the fitted coefficients, converted to a logarithm
-base 2 scale: ``X.Intercept.'', ``typeua159'', and ``conditionglucose.''
-The log2 fold change due to the condition is shown in the 4th column. As indicated
-by the column name, it is the effect of ``glucose'', i.e., the log ratio of
-``glucose'' versus ``galactose'' or $\log(glucose/galactose)$. 
-Note that the library type also
-had noticeable influence on the expression, and hence would have increased the dispersion
-estimates (and so reduced power) if we had not fitted an effect for it.
-The column \emph{deviance}
-is the deviance of the fit. (Comparing the deviances with a $\chi^2$ likelihood ratio test is
-how \Rfunction{nbinomGLMTest} calculates the $p$ values.) The column, \emph{converged},
-indicates whether the calculation of coefficients and deviance has fully converged.
-(If it is false too often, you can try to change the GLM control parameters, as
-explained in the help page to \Rfunction{fitNbinomGLMs}.)
-The second to the last column, \emph{pval}, is the raw p-values from the test
-\Rfunction{nbinomGLMTest} provided by \Rpackage{DESeq}. The last column,
-\emph{padj}, is the multiple testing adjusted p-values.
-
-%------------------------------------------------------------
-\subsection{Functional category association analyses} 
-\label{subsec:functional}
-%------------------------------------------------------------
-
-Once we find differentially expressed genes, we could cluster them. This could
-be useful to group genes that behave similarly.  These potentially useful
-figures can be plotted from \Rpackage{DESeq}.  We could also find function
-categories that were enriched with differentially expressed genes. For
-clustering differentially expressed genes we need moderate fold changes to make
-them amenable to distance calculation.  Figure \ref{figLfc} shows moderated log
-fold changes versus the original log fold changes.
-
-<<options,results=hide,echo=FALSE>>=
-largeNumber <- 10
-@
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figLfc}
-\caption{Scatterplot of direct (\Robject{lfc}) versus \emph{moderated} log-ratios (\Robject{mod\_lfc}).
-  The moderation criterion used is variance stabilisation.
-  The red points correspond to values that were infinite in \Robject{lfc}
-  and were abitrarily set to $\pm\Sexpr{largeNumber}$ for the purpose of plotting.
-  These values vary in a finite range (as shown in the plot) in \Robject{mod\_lfc}.}
-\label{figLfc}
-\end{figure}
-<<figLfc,fig=TRUE>>=
-select <- 1:40
-smutans.de2TypeLfc( ua159only, condA="glucose", condB="galactose" )
-@
-Now we plot the first \Sexpr{length(select)} genes clustered based on the
-moderated log fold changes.
-
-%-----------------------------------------------------------------------------
-\begin{figure}
-\centering
-\includegraphics[width=.6\textwidth]{smutans-figHeatmap1}
-\caption{Heatmaps showing the expression data of the top \Sexpr{length(select)}
-  differentially expressed genes. Left, the variance stabilization
-  transformed data, right, the original count data.
-  The right plot is dominated by a small number of data points with large values.}
-\label{figHeatmap1}
-\end{figure}
-%-----------------------------------------------------------------------------
-<<figHeatmap1,fig=TRUE>>=
-smutans.de2TypeHeatmap( ua159only )
-@
-
-<<goseqUA159only,eval=FALSE>>=
-data( "smutans.feature.genes" )
-data( "smutans.go.genes" )
-data( "smutans.cat.desc" )
-smutans.de2Goseq ( ua159only, qval=1e-2, 
-                   feature.gene=smutans.feature.genes,
-                   go.genes=smutans.go.genes,
-                   cat.desc=smutans.cat.desc )
-@
-We find some functional categories enriched with differentially expressed genes
-(table \ref{tab:goseq})
-We find functional categories that are enriched with differentially expressed genes.
-To find functional categories enriched with differentially expressed genes we
-first find a list of differentially expressed genes. A set of genes are assigned
-to functional gene categories: \Robject{smutans.go.genes}. Features of the genes in
-UA159 include gene lengths: \Robject{smutans.feature.genes}.
-Functional categories with numbers of genes assigned are in
-\Robject{smutans.cat.desc}.
-%-----------------------------------------------------------------------------
-\begin{table}[!h]
-\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
-genes}}
-\vspace{1ex}
-\noindent \begin{centering}
-\begin{tabular}{cl}
-\hline 
-$q^a$ & Description \\
-\hline 
-%-----------------------------------------------------------------------------
-<<goseqUA159only,echo=FALSE,results=tex>>=
-data( "smutans.feature.genes" )
-data( "smutans.go.genes" )
-data( "smutans.cat.desc" )
-smutans.de2Goseq ( ua159only, qval=1e-2, 
-                   feature.gene=smutans.feature.genes,
-                   go.genes=smutans.go.genes,
-                   cat.desc=smutans.cat.desc )
-@
-%-----------------------------------------------------------------------------
-\hline 
-\end{tabular}
-\par\end{centering}
-\begin{flushleft}
-$^a$False discovery rate estimated by the
-Benjamini-Hochberg method.  All categories having at least ten genes and
-$q\leq 0.05$ are displayed.\\ 
-\end{flushleft}
-\label{tab:goseq}
-\end{table}
-%-----------------------------------------------------------------------------
-
-%--------------------------------------------------
-\section{DE analysis using 835NP, 835P, and UA159 no CSP}
-\label{sec:de835}
-%--------------------------------------------------
-
-We first cluster the \Sexpr{length(colnames(counts(sm835Genes)))} samples.
-\begin{figure}
-\centering
-\includegraphics[width=.9\textwidth]{smutans-figClustNpAndP}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength.}
-\label{figClustNpAndP}
-\end{figure}
-<<figClustNpAndP,fig=TRUE>>=
-data( "sm835Genes" )
-npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
-smutans.de2Clust( npAndP )
-@
-
-We found no functional categories enriched with differentially expressed
-genes in the pairwise comparisons of 835NP, 835P, and UA159 no CSP.
-Let's make a list of differentially expressed genes. 
-Another kind of heatmap plots for
-grouping genes could be interesting. 
-But, basically we have a list of genes with
-some adjusted p-values for detecting differentially expressed genes.
-
-<<look_into_smutansGenes>>=
-head( counts(sm835Genes) )
-pData( sm835Genes )
-@
-
-<<sm835Samples>>=
-npAndP <- newSmutans( sm835Genes, title="835NP vs. 835P" )
-npAndP <- smutans.de2( npAndP, type="ua159", 
-                       condA="835np", condB="835p" )
-npAndUA159 <- newSmutans( sm835Genes, title="835NP vs. UA159" )
-npAndUA159 <- smutans.de2( npAndUA159, type="ua159", 
-                           condA="835np", condB="ua159nocsp" )
-pAndUA159 <- newSmutans( sm835Genes, title="835P vs. UA159" )
-pAndUA159 <- smutans.de2( pAndUA159, type="ua159", 
-                          condA="835p", condB="ua159nocsp" )
-smutans.de2List( npAndP, "npAndP.csv" )
-smutans.de2List( npAndUA159, "npAndUA159.csv" )
-smutans.de2List( pAndUA159, "pAndUA159.csv" )
-@
-
-\begin{figure}
-\centering
-\includegraphics[width=.9\textwidth]{smutans-figHeatmapNpAndP}
-\caption{Empirical (black dots) and fitted (red lines) dispersion values
-plotted against mean expression strength. \textbf{figHeatmapNpAndP}}
-\label{figHeatmapNpAndP}
-\end{figure}
-<<figHeatmapNpAndP,fig=TRUE>>=
-smutans.de2TypeHeatmap( npAndP )
-@
-
-%--------------------------------------------------
-\section{DE analysis using OMZ175 and OMZ/HKRR}
-\label{sec:deomz}
-%--------------------------------------------------
-
-<<look_into_smomzGenes>>=
-data( "smomzGenes" )
-head( counts(smomzGenes) )
-pData( smomzGenes )
-@
-
-<<smomz,results=hide>>=
-omz <- newSmutans( smomzGenes, title="OMZ175 vs. OMZ/HKRR" )
-omz <- smutans.de2( omz, type="ua159", 
-                    condA="OMZ175", condB="OMZHKRR" )
-smutans.de2List( omz, "omz.csv" )
-@
-
-%--------------------------------------------------
-\section{DE analysis using UA159 CSP}
-\label{sec:deua159csp}
-%--------------------------------------------------
-
-<<look_into_smua159cspGenes>>=
-data( "smua159cspGenes" )
-head( counts(smua159cspGenes) )
-pData( smua159cspGenes )
-@
-
-<<smua159csp,results=hide>>=
-ua159csp <- newSmutans( smua159cspGenes, title="UA159 no CSP vs. CSP" )
-ua159csp <- smutans.de2( ua159csp, type="ua159", 
-                         condA="UA159noCSP", condB="UA159CSP" )
-smutans.de2List( ua159csp, "ua159csp.csv" )
-@
-
-%--------------------------------------------------
-\section{DE analysis using SMU86 CSP}
-\label{sec:desmu86csp}
-%--------------------------------------------------
-
-<<look_into_smsmu86cspGenes>>=
-data( "smsmu86cspGenes" )
-head( counts(smsmu86cspGenes) )
-pData( smsmu86cspGenes )
-@
-
-<<smsmu86csp,results=hide>>=
-smu86csp <- newSmutans( smsmu86cspGenes, title="UA159 no CSP vs. CSP" )
-smu86csp <- smutans.de2( smu86csp, type="ua159", 
-                         condA="Smu86noCSP", condB="Smu86CSP" )
-smutans.de2List( smu86csp, "smu86csp.csv" )
-@
-
-%--------------------------------------------------
-\section{DE analysis using predicted transcripts}
-\label{sec:detranscript}
-%--------------------------------------------------
-
-We need ways of supporting the predicted transcripts. We need a way of using
-expression levels and counts of starts and ends of short reads to predict
-transcripts. For now, we just use an ad hoc method for finding transcript start
-and end sites. A likelihood can be constructed by multiplying geometric
-probabilities. Given a range of possible transcription boundaries we can find
-estimates of positions and geometirc parmeter $p$ that maximize the likelihood
-for both ends of a transcript.
-
-\Rpackage{GenomicRanges} might be useful in 
-<<tx,eval=FALSE>>=
-source("http://www.bioconductor.org/biocLite.R")
-biocLite("GenomicRanges")
-@
-
-%--------------------------------------------------
-\section{DE analysis using predicted small RNAs and target genes}
-\label{sec:desmallrna}
-%--------------------------------------------------
-
-Create a file with intergenic regions sequences of \textit{S.\ mutans} UA159.
-Create a file with bacterial genome sequences except \textit{S.\ mutans} NN2025.
-BLAST the intergenic regions against the bacterial genomes. Extract sequence
-alignments from the BLAST output. Refine the alignment using MUSCLE. Convert the
-FASTA-format alignments to MSA-format ones.  Find small RNA candidates using
-RNAz.  Find targets of the small RNA candidates using RNAplex. Targets are
-scored for each small RNA candidate. The scores were used to do functional
-category enrichment tests of targets for each small RNA candidate.
-
-Table \ref{tab:goseq} 
-shows functional categories enriched with target genes for predicted small
-RNAs. Let's use \Rfunction{smutans.mannwhitney} to the package and other data
-objects.
-%-----------------------------------------------------------------------------
-\begin{table}[!h]
-\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
-genes}}
-\vspace{1ex}
-\noindent \begin{centering}
-\begin{tabular}{lcccl}
-\hline 
-small RNA & $q^a$ & count & GO & Description \\
-\hline 
-%-----------------------------------------------------------------------------
-<<goseqUA159only,echo=FALSE,results=tex>>=
-data( "smutans.genes.criteria" )
-data( "smutans.go.genes" )
-data( "smutans.cat.desc" )
-smutans.mannwhitney( qval=0.05,
-                   genes.criteria=smutans.genes.criteria,
-                   go.genes=smutans.go.genes,
-                   cat.desc=smutans.cat.desc )
-@
-%-----------------------------------------------------------------------------
-\hline 
-\end{tabular}
-\par\end{centering}
-\begin{flushleft}
-$^a$False discovery rate estimated by the
-Benjamini-Hochberg method.  All categories having at least ten genes and
-$q\leq 0.05$ are displayed.\\ 
-\end{flushleft}
-\label{tab:goseq}
-\end{table}
-%-----------------------------------------------------------------------------
-
-<<smutansRNA>>=
-data( "smutansGenes2" )
-ua159only <- newSmutans( smutansGenes2, title="UA159 Only" )
-ua159only <- smutans.de2( ua159only, type="ua159", 
-                          condA="glucose", condB="galactose" )
-allSample <- newSmutans( smutansGenes2, 
-                         title="All Samples for Comparing Conditions" )
-allSample <- smutans.de2( allSample )
-tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
-                                                   allSample )
-addmargins( tabUA159onlyAndAllSample )
-@
-Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
-\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
-analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
-genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
-genes. We found a few predicted small RNAs that appeared to be differentially
-expressed.
-<<smutansRNA>>=
-smutans.de2List( allSample )[grep("locus",rownames(smutans.de2List(allSample))),]
-@
-We also found a few predicted small RNAs that appeared to be differentially
-expressed in the comparison of the two types: ``ua159'' and ``tw1.''
-<<smutansRNA>>=
-allSampleType <- newSmutans( smutansGenes2, title="All samples for type" )
-allSampleType <- smutans.de2( allSampleType, compareCondition="no" )
-smutans.de2List( allSampleType )[grep("locus",rownames(smutans.de2List(allSampleType))),]
-@
-
-\section{Prewriting about UA159 versus TW1}
-
-\subsection{Abstract}
-A bacterial transcriptome of the primary etiological agent of human dental caries,
-\textit{Stretptococcus mutans}, is described using RNA-seq data. Putative
-transcripts boundaries are defined. Small RNAs are predicted by computational
-methods, and confirmed by RNA-seq and Southern blot experiments. We refine the
-previously established importance of CcpA regulation as central metabolism and
-virulence gene expression using RNA-seq data.
-CcpA is known to play a direct role in carbon catabolite repression (CCR).
-We describe genes that are differentially expressed in glucose-grown wild-type
-(UA159) and CcpA-deficient (TW1) strains. We found that more
-genes were differentially expressed in the comparison between wild-type and
-CcpA-deficient strains in galactose-grown strains than in
-glucose-grown strains.
-We also found that more genes differentially expressed in wild-type than in
-CcpA-deficient strain when comparing glucose- and galactose-grown cells. These
-two findings were different from those of \cite{Abranches2008}. \fixme{We may
-want to see effects of different p-value cutoff.} We confirmed that CcpA
-modulates the pathogenic potential of \textit{S.\ mutans} through global control
-of gene expression.
-We predicted a few small RNAs based on sequence alignments and RNA secondary
-structures. We confirmed them experimentally. Target genes of predicted small
-RNAs were predicted. The target genes were enriched in some functional
-categories. We listed highly potential small RNAs with their target genes
-functional categories. 
-Transcripts were predicted using RNA-seq short reads mapped on the wild-type
-UA159 genome.
-
-\subsection{Reviews of papers that cited \cite{Abranches2008}}
-
-CCR in bacteria involves regulatory networks that activate or silence genes in
-response to carbon source and availability.
-
-\subsection{Goal: Refinement of \cite{Abranches2008}}
-
-Comparisons of microarray and RNA-seq are not good way to go.
-
-\subsection{Methods and Materials}
-
-\subsubsection{Bacterial strains and growth conditions}
-
-\subsubsection{RNA isolation}
-
-\subsubsection{RNA sequencing}
-
-\subsubsection{Statistical analysis for differential expressions}
-We used an R package, DESeq, \cite{Anders2010} to determine differentially
-expressed genes based on negative binomial model originally developed by \cite{Robinson2007}.
-Steps of analyzing RNA-seq data for determining differentially expressed genes
-are described in \cite{Oshlack2010}. We used a short read aligner called
-\program{BWA} \cite{Li2009a} to map the sequenced short reads on a reference genome,
-\textit{Streptococcus mutans} UA159 (NCBI accession NC\_004350). Because the
-aligner allowed a few gaps for efficient alignment of millions of sequences of
-size being 100 base pairs long, short reads that consist of RNAs and sequencing
-adapter sequences would be not mapped. We used \program{cutadapt} \cite{Martin2011} to
-remove parts of adapter sequences from short reads before mapping on the
-reference genome. Mapped short read alignments were converted into readable
-formats \cite{Li2009b}. We counted short reads aligned to annotated genes in the
-reference genome, forming a table of read counts of the genes. Statistical
-software R \cite{R2011} with \Rpackage{DEseq} \cite{Anders2010}
-was subsequently employed to infer differentially expressed in different
-biological conditions.  We also generated tracks for our recently released
-Streptococcus Genome Browser (http://strep-genome.bscb.cornell.edu) that
-summarize gene expression levels based the short read alignments. This allowed
-us to visualize parts of the reference genome for comparing different RNA
-samples. Further analyses such as functional category enrichment test
-\cite{Young2010} were also employed.
-
-\subsubsection{small RNAs}
-
-RNAz is used to find small RNA candidates. Sequence alignments were a
-critical step for finding small RNAs in the approach. Phylogenies of streptococcus species could
-be considered to choose species related with Streptococcus mutans.
-Alternatively, I find sequences similar to intergenic regions, and refine the
-alignments using a multiple sequence alignment tool such as MUSCLE
-\cite{Edgar2004a}. Steps for
-achieving this goal include: 1. Exract intergenic regions including up- and
-down-stream, 2. BLAST the intergenic regions against non-redundant DB to find
-similar sequences \cite{Altschul1990}, 3. Refine each alignment using MUSCLE \cite{Edgar2004a}, 4. Apply RNAz to the
-alignments for scoring regions for small RNAs
-\cite{Washietl2005,Gruber2010}, and 5. Find transcripts with high
-score of small RNAs. The transcripts must be defined in advance.
-
-RNAplex
-\cite{Tafer2008}
-
-A review paper about small RNA predictions and targets
-\cite{Tafer2008a}
-
-RNAplfold
-\cite{Bernhart2006}
-
-\subsubsection{Transcripts}
-
-``-c 10 -b 25 -force_gp''
-
-\subsubsection{Gene category associations and virulence genes}
-
-We assigned genes to Gene Ontology (GO) categories by comparing the {\em
-  Streptococcus} genes to bacterial proteins from the Uniref90 database
-using BLASTP, and then assigning the same GO classification as the target
-gene of the uniProt GOA database if the match had an $E$-value of
-$<1.0\times10^{-5}.$  
-A gene family was assigned a given
-classification if any of its genes was assigned that classification.  To
-test for associations with replacing gene transfers we used the
-recombination intensity estimated for each gene, and for additive gene
-transfers we used the number of additive transfers inferred for each
-family.  To test for significance, we performed a Mann-Whitney $U$ test of
-the values (recombination intensities or numbers of transfers) associated
-with a given category vs.\ the values for the all other genes/families.  We
-used the \cite{Benjamini1995} method to correct for multiple comparisons.
-
-\subsection{Results}
-
-CcpA affects CCR of \textit{fruA}. 
-\textit{fruA} fructan hydrolase
-When Wen2002 deleted CRE sequencs in the \textit{fruA} promoter region, CCR was
-almost completely alleviated. 
-What are CCR-sensitive genes? 
-What are the mechanism of CcpA on CCR of the \textit{fruA} operon? 
-Expression levels of \textif{fruA} in \textit{S.\ mutans} strains that were
-grown to early exponential phase (OD_600, 0.2) in TV medium supplemented with
-0.5\% glucose or galactose. \fixme{What does this mean?} 
-Deletion of \textit{ccpA} resulted in substantial relief of the CCR of
-\textit{fruA} expression exerted by both glucose and fructose. 
-CcpA could directly interact with the \textit{fruA} promoter region.
-
-The role of CcpA in gene expression during growth on glucose. 
-\textit{S.\ mutans} UA159 and TW1 grown under catabolite-repressing conditions
-(glucose), and catabolite-derepressing conditions (galactose). Under
-glucose-only condition I found xx number of differentially expressed genes. What
-are the functional categories of those differentially expressed genes? What are
-the genes that are most profoundly up- or down-regulated (Smu.1421 to Smu.1424,
-PDH enzyme complex), ... and more? ... This regulatory control was
-short-circuited in the strain growing on glucose and lacking CcpA. I showed that
-some of KEGG pathways were affected by TW1 grown under glucose. We found
-down-regulated genes, which would have something to do with pathogenicity. 
-Upto here, I compared UA159 and TW1 under glucose.
-
-Growth in derepressing conditions induces major changes in the UA159
-transcriptome. Here, I compared glucose and galactose conditions for growing
-UA159. Energy metabolism was enriched with differentially expressed genes. There
-were more up-regulated genes in galactose condition compared with glucose
-condition. Some genes are described in detail.
-
-Now, I compared glucose and galactose for growing TW1.
-
-Now, I compared UA159 and TW1 under galactose.
-
-What genes are differentially expressed?
-
-Name the transcripts.
-I found xx transcripts. Among them xx \% were mono-cistronic operons. xx \% were
-poly-cistronic operons. xx transcripts could not defined because RNA-seq short
-reads were not precise. Transripts were assigned with functional categories, and
-differentially expressed transcripts were found. Transcripts and functional
-categories: how can I assess that transcripts consist of genes with similar
-functional cateogires.
-
-Name the small RNAs.
-RNAz predicted a few small RNAs. Some of them were differentially expressed.
-Some functional categories were enriched with target genes of the predicted
-genes.
-
-\subsection{Discussion}
-We used RNA-seq data to delineate transcript boundaries. Although this map of
-transcripts does not precisely tell us where transcripts start and end, it
-informs us of which open reading frames can be transcribed together. Potentially
-polycistronic operons are defined. We focused on interpretations of operon in
-light of CcpA regulations.
-
-\subsection{Abbreviations}
-
-\begin{enumerate}
-\item CCR Carbon catabolite repression
-\item CcpA ?
-\item IPS intracellular polysaccharide
-\end{enumerate}
+rtracklayer
+export()
 

### 2012-04-24 sh/batch2.sh

@@ -33,6 +33,7 @@ function batch2 {
 #  ucsc-data
     batch2-run-fastqc
     batch2-run-cram
+    batch2-run-cram2fastq
     batch2-run-bwa-align
     batch2-run-samtools-pileup
 
@@ -353,12 +354,8 @@ function copy-data {
   cp \$HOME/$SUBREADALIGN subread-align
 
   # All of the batchjob scripts.
-  cp \$PBS_O_WORKDIR/job-cram .
   cp \$PBS_O_WORKDIR/job-cram2fastq .
-  cp \$PBS_O_WORKDIR/job-fastqc .
   cp \$PBS_O_WORKDIR/job-bwa-align .
-  cp \$PBS_O_WORKDIR/job-de* .
-  cp \$PBS_O_WORKDIR/job-de*.R .
 
   # Create output directories at the compute node.
   mkdir -p $CDATADIR
@@ -367,11 +364,6 @@ function copy-data {
 
   # Copy common data
   cp $RDATADIR/$GENOMEFASTA $CDATADIR
-  cp $RDATADIR/$GENOMEGFF $CDATADIR
-}
-
-function retrieve-data {
-  echo No Copy!
 }
 
 function process-data {
@@ -381,13 +373,14 @@ function process-data {
   g=\$((PBS_ARRAYID-1))
   NUM=\$(printf "%03d" \${FASTQFILES[\$g]})
 
+  cp $RBWADIR/FASTQ\$NUM.cram $CBWADIR
   bash job-cram2fastq \$NUM \\
-    $RDATADIR/FASTQ\$NUM.cram \\
-    $CDATADIR/FASTQ\$NUM.recovered.fq
-  gzip $CDATADIR/FASTQ\$NUM.recovered.fq
+    $CBWADIR/FASTQ\$NUM.cram \\
+    $CBWADIR/FASTQ\$NUM.recovered.fq
+  gzip $CBWADIR/FASTQ\$NUM.recovered.fq
 
   bash job-bwa-align \$NUM \\
-    $CDATADIR/FASTQ\$NUM.recovered.fq.gz \\
+    $CBWADIR/FASTQ\$NUM.recovered.fq.gz \\
     $CBWADIR/FASTQ\$NUM.sorted \\
     $CSUBREADDIR/FASTQ\$NUM.sorted
 
[…]

### 2012-07-17 package/smutans/inst/doc/smutans.Rnw
The

@@
 \cite{Oshlack2010}. Short reads are first mapped to a reference genome using
-approaches such as \program{BWA} \cite{Li2009a} or \program{subread}.
+approaches such as \program{BWA} \cite{Li2009a}.
 Reads fitting into one of the annotated
@@
 %-----------------------------------------------------------------------------
-<<goseqUA159only,echo=FALSE,results=tex>>=
+<<goseqUA159only,eval=FALSE,echo=FALSE,results=tex>>=
 data( "smutans.feature.genes" )
@@
 %-----------------------------------------------------------------------------
-<<goseqUA159only,echo=FALSE,results=tex>>=
+<<goseqUA159only,echo=FALSE,eval=FALSE,results=tex>>=
 data( "smutans.genes.criteria" )

### 2012-07-17 R/xbasegff2txdb.R

@@ -12,17 +12,16 @@ gffFileBasename <- unlist(strsplit(gffFile,"\\."))[1]
 sm.gff <- import.gff3(gffFile)
 sm.gene <- sm.gff[sm.gff$type=="gene",]
 sm.CDS <- sm.gff[sm.gff$type=="CDS",]
-sm.source <- sm.gff[sm.gff$type=="DNA",]
+sm.source <- sm.gff[sm.gff$type=="region",]
 
 transcripts <- 
-  data.frame( tx_id=seq(length(sm.gene$locus_tag)), # integer vector
-              tx_name=sm.gene$locus_tag,            # string vector or factor
-              tx_chrom=sm.gene$space,               # string vector or factor
-              tx_strand=sm.gene$strand,             # strand vector
-              tx_start=start(sm.gene),              # integer vector
-              tx_end=end(sm.gene) )                 # integer vector
+  data.frame( tx_id=seq(length(sm.gene$locus_tag)),
+              tx_name=sm.gene$locus_tag,
+              tx_chrom=sm.gene$space, 
+              tx_strand=sm.gene$strand,
+              tx_start=start(sm.gene),
+              tx_end=end(sm.gene) )
 
-# Each transcript contains an exon. The exon may contain a CDS. 
 # Compare CDS and gene locus tags
 x <- c()
 y <- c()
@@ -48,20 +47,22 @@ rm(i,x,y,z)
 
 # 
 splicings <- 
-  data.frame( tx_id=seq(length(sm.gene$locus_tag)),     # integer A
-              exon_rank=seq(length(sm.gene$locus_tag)), # integer B (unique A-B)
-              exon_name=sm.gene$locus_tag,              # character
-              exon_start=start(sm.gene),                # integer 
-              exon_end=end(sm.gene),                    # integer
-              cds_start=grCDS.start,                    # integer
-              cds_end=grCDS.end )                       # integer
+  data.frame( tx_id=seq(length(sm.gene$locus_tag)),
+              exon_rank=seq(length(sm.gene$locus_tag)),
+              exon_start=start(sm.gene),
+              exon_end=end(sm.gene),
+              exon_name=sm.gene$locus_tag,
+              cds_start=grCDS.start,
+              cds_end=grCDS.end )
 
 chrominfo <-
-  data.frame( chrom=names(sm.gene),    # character
-              length=end(sm.source),   # integer
[…]

### 2012-07-17 pl/bp_genbank2gff3.pl

@@ -1,0 +1,1046 @@
+#!/usr/bin/perl -w
+
+eval 'exec /usr/bin/perl -w -S $0 ${1+"$@"}'
+    if 0; # not running under some shell
+
+#$Id: genbank2gff3.PLS 15088 2008-12-04 02:49:09Z bosborne $;
+
+=pod
+
+=head1 NAME 
+
+genbank2gff3.pl -- Genbank-E<gt>gbrowse-friendly GFF3
+
+=head1 SYNOPSIS
+
+  genbank2gff3.pl [options] filename(s)
+
+  # process a directory containing GenBank flatfiles
+  perl genbank2gff3.pl --dir path_to_files --zip
+
+  # process a single file, ignore explicit exons and introns
+  perl genbank2gff3.pl --filter exon --filter intron file.gbk.gz
+
+  # process a list of files 
+  perl genbank2gff3.pl *gbk.gz
+
+  # process data from URL, with Chado GFF model (-noCDS), and pipe to database loader
+  curl ftp://ftp.ncbi.nih.gov/genomes/Saccharomyces_cerevisiae/CHR_X/NC_001142.gbk \
+  | perl genbank2gff3.pl -noCDS -in stdin -out stdout \
+  | perl gmod_bulk_load_gff3.pl -dbname mychado -organism fromdata
+
+    Options:
+        --dir     -d  path to a list of genbank flatfiles
+        --outdir  -o  location to write GFF files (can be 'stdout' or '-' for pipe)
+        --zip     -z  compress GFF3 output files with gzip
+        --summary -s  print a summary of the features in each contig
+        --filter  -x  genbank feature type(s) to ignore
+        --split   -y  split output to seperate GFF and fasta files for
+                      each genbank record
+        --nolump  -n  seperate file for each reference sequence
+                      (default is to lump all records together into one 
+                       output file for each input file)
+        --ethresh -e  error threshold for unflattener
+                      set this high (>2) to ignore all unflattener errors
+        --[no]CDS -c  Keep CDS-exons, or convert to alternate gene-RNA-protein-exon 
+                      model. --CDS is default. Use --CDS to keep default GFF gene model, 
+                      use --noCDS to convert to g-r-p-e.
+        --format  -f  Input format (SeqIO types): GenBank, Swiss or Uniprot, EMBL work
[…]

### 2012-07-17 pl/extract-msa.pl

@@ -1,26 +1,27 @@
 #!/usr/bin/perl
 ###############################################################################
-# Copyright (C) 2011 Sang Chul Choi
+# Copyright (C) 2011-2012 Sang Chul Choi
 #
-# This file is part of Mauve Analysis.
+# This file is part of RNAseq Analysis.
 # 
-# Mauve Analysis is free software: you can redistribute it and/or modify
+# RNAseq Analysis is free software: you can redistribute it and/or modify
 # it under the terms of the GNU General Public License as published by
 # the Free Software Foundation, either version 3 of the License, or
 # (at your option) any later version.
 #
-# Mauve Analysis is distributed in the hope that it will be useful,
+# RNAseq Analysis is distributed in the hope that it will be useful,
 # but WITHOUT ANY WARRANTY; without even the implied warranty of
 # MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
 # GNU General Public License for more details.
 #
 # You should have received a copy of the GNU General Public License
-# along with Mauve Analysis.  If not, see <http://www.gnu.org/licenses/>.
+# along with RNAseq Analysis.  If not, see <http://www.gnu.org/licenses/>.
 ###############################################################################
 use strict;
 use warnings;
 use Getopt::Long;
 use Pod::Usage;
+use Bio::SeqIO;
 require 'pl/sub-error.pl';
 require 'pl/sub-bed.pl';
 require 'pl/sub-fasta.pl';
@@ -44,6 +45,10 @@ GetOptions( \%params,
             'reference=s',
             'genomeLength=i',
             'in=s',
+            'in1=s',
+            'in2=s',
+            'gene=s',
+            'coverage=f',
             'indir=s',
             'out=s',
             'outdir=s',
@@ -58,6 +63,10 @@ pod2usage(-exitstatus => 0, -verbose => 2) if $man;
 
 my $in;
 my $infile;
+my $in1;
+my $in1file;
+my $in2;
+my $in2file;
 my $out;
 my $outfile;
 my $outdir;
@@ -65,6 +74,12 @@ my $indir;
 my $reference;
 my $muscle;
 my $genomeLength;
+
+my $coverageArg = 0.5;
+if (exists $params{coverage}) {
[…]

### 2012-07-17 pl/extractcds.pl

@@ -1,0 +1,90 @@
+#!/usr/bin/perl
+
+# extractcds
+# C. Letondal, letondal@pasteur.fr
+
+use Bio::SeqIO;
+use Getopt::Std;
+use Getopt::Long;
+
+&init;
+&process;
+
+sub usage {
+    my $program = `basename $0`;
+    chop($program);
+    print STDERR "
+      $program [ -f format ] [ -p ] [ -l ] [ -g ] [ -h ] seqfile
+      Extracts translation(s) from an Embl or Genbank entry
+
+      -f format  : Genbank or Embl format
+      -p         : show product
+      -g         : show gene
+      -l         : show location
+      -h         : this message
+      seqfile    : entry file (or stdin)
+";
+
+}
+
+sub init {
+    getopts('f:plgh');
+    if ($opt_h) {&usage; exit;}
+    if ($opt_f) {$format = $opt_f; } else { $format = "Genbank";}
+    if ($opt_p) {$show_product = 1;}
+    if ($opt_l) {$show_location = 1;}
+    if ($opt_g) {$show_gene = 1;}
+    $seqfile = ($ARGV[0]) ? $ARGV[0] : "-";
+}
+
+sub process {
+    $seqio  = Bio::SeqIO->new (-format => $format , -file => $seqfile);
+    $out = Bio::SeqIO->new( '-format' => 'fasta', '-fh' => \*STDOUT );
+    while (my $seqobj = $seqio->next_seq())
+    {
+    foreach $feat ( $seqobj->all_SeqFeatures() ) {
+        if ($feat->has_tag('translation')) {
+            # if ($feat->has_tag('protein_id')) {
+            if ($feat->has_tag('locus_tag')) {
+                # @protein_id = $feat->each_tag_value('protein_id');
+                @protein_id = $feat->each_tag_value('locus_tag');
+                $protein_id = $protein_id[0];
+            } elsif ((! $show_gene) && $feat->has_tag('gene')) {
+                @gene = $feat->each_tag_value('gene');
+                $protein_id = $gene[0];
+            } else {
+                $protein_id = "";
+            }
+
+            if ($show_gene && $feat->has_tag('gene')) {
+                @gene = $feat->each_tag_value('gene');
+                $gene = $gene[0];
+            } else {
+                $gene = "";
+            }
+
+            if ($show_product && $feat->has_tag('product')) {
+                @product = $feat->each_tag_value('product');
+                $product = $product[0];
+            } else {
+                $product = "";
+            }
+
[…]

### 2012-07-17 pl/geneontology.pl

@@ -110,6 +110,8 @@ if ($cmd eq "gene2go")
 {
   # Parse gff file.
   my %locus2protein;
+  my $geneid;
+  my $locustagid;
   open GFF, $params{gff} or die "cannot open $params{gff} $!";
   while (<GFF>)
   {
@@ -117,16 +119,38 @@ if ($cmd eq "gene2go")
     next if /^#/;
 # NC_004350.1	RefSeq	CDS	194	1549	.	+	0	ID=NC_004350.1:dnaA:unknown_transcript_1;Parent=NC_004350.1:dnaA;locus_tag=SMU.01;note=binds to the dnaA-box as an ATP-bound complex at the origin of replication during the initiation of chromosomal replication%3B can also affect transcription of multiple genes including itself.;transl_table=11;product=chromosomal replication initiation protein;protein_id=NP_720488.1;db_xref=GI:24378533;db_xref=GeneID:1029426;exon_number=1
     # if (/locus\_tag=(\w+);.+;protein\_id=(\w+);/)
-    if (/RefSeq\s+CDS\s+.+locus\_tag=([\w.]+);.+;protein\_id=(\w+)/)
+
+    # The S. mutans version 2 does not work with the following line.
+    # if (/RefSeq\s+CDS\s+.+locus\_tag=([\w.]+);.+;protein\_id=(\w+)/)
+
+    if (/RefSeq\s+gene\s+.+ID=(\w+);.+;locus\_tag=([\w.]+)/)
     {
-      if (exists $locus2protein{$1})
+      $geneid = $1; 
+      $locustagid = $2;
+    }
+    elsif (/RefSeq\s+CDS\s+.+Parent=(\w+);.+;protein\_id=(\w+)/)
+    {
+      if ($geneid eq $1)
       {
-        die "GFF: $1 was already added.";
+        if (exists $locus2protein{$locustagid})
+        {
+          die "GFF: $locustagid was already added.";
+        }
+        $locus2protein{$locustagid} = $2;
       }
-      $locus2protein{$1} = $2;
     }
   }
   close GFF;
+
+  # Test
+=comment
+  foreach my $locusid (keys %locus2protein)
+  {
+    my $protein = $locus2protein{$locusid};
+    print "$locusid\t$protein\n";
+  }
+  exit; 
+=cut
 
   # Parse BLAST file.
   my %protein2uniref;
@@ -139,6 +163,10 @@ if ($cmd eq "gene2go")
[…]

### 2012-07-17 sh/batch3.sh

@@ -62,14 +62,14 @@ function batch3-variable {
   RNUMBERDIR=$ROUTPUTDIR/$SPECIES/$REPETITION
   RDATADIR=$RNUMBERDIR/data
   RBWADIR=$RNUMBERDIR/bwa
-  RANALYSISDIR=$RNUMBERDIR/run-analysis
+  RRUNANALYSIS=$RNUMBERDIR/run-analysis
 
   # Compute node output directories.
   CBASEDIR=output/$SPECIES
   CNUMBERDIR=$CBASEDIR/$REPETITION
   CDATADIR=$CNUMBERDIR/data
   CBWADIR=$CNUMBERDIR/bwa
-  CANALYSISDIR=$CNUMBERDIR/run-analysis
+  CRUNANALYSIS=$CNUMBERDIR/run-analysis
 }
 
 # Create output directories for the species directory
@@ -101,6 +101,7 @@ cat>$BASEDIR/push-data.sh<<EOF
 #!/bin/bash
 ssh -x $CAC_USERHOST mkdir -p $RBWADIR
 ssh -x $CAC_USERHOST mkdir -p $RDATADIR
+ssh -x $CAC_USERHOST mkdir -p $RRUNANALYSIS
 scp $REFGENOMEFASTA $CAC_USERHOST:$RDATADIR
 scp $REFGENOMETXDB $CAC_USERHOST:$RDATADIR
 for g in $FASTQFILES; do

### 2012-07-17 sh/choose-species.sh

@@ -32,9 +32,6 @@ function choose-species {
       mkdir -p $NUMBERDIR
       mkdir -p $DATADIR
       mkdir -p $BWADIR
-      mkdir -p $SUBREADDIR
-      mkdir -p $MAQDIR
-      mkdir -p $BOWTIEDIR
       mkdir -p $TRANSCRIPTDIR
       mkdir -p $RUNANALYSIS
       break

### 2012-07-17 sh/conf.sh

@@ -35,6 +35,9 @@ function conf {
   SUBREADBUILDINDEX=$(grep ^SUBREADBUILDINDEX\: $CONFFILE | cut -d":" -f2)
   SUBREADALIGN=$(grep ^SUBREADALIGN\: $CONFFILE | cut -d":" -f2)
   SEGEMEHL=$(grep ^SEGEMEHL\: $CONFFILE | cut -d":" -f2)
+  MAKEBLASTDB=$(grep ^MAKEBLASTDB\: $CONFFILE | cut -d":" -f2)
+  BLASTN=$(grep ^BLASTN\: $CONFFILE | cut -d":" -f2)
+  BLASTP=$(grep ^BLASTP\: $CONFFILE | cut -d":" -f2)
 
   # Other global variables
   CAC_USERHOST=$CAC_USERNAME@$CAC_LOGIN

### 2012-07-17 sh/deseq.sh

@@ -1,7 +1,7 @@
 ###############################################################################
-# Copyright (C) 2011 Sang Chul Choi
+# Copyright (C) 2011-2012 Sang Chul Choi
 #
-# This file is part of RNASeq Analysis.
+# This file is part of RNAseq Analysis.
 # 
 # Mauve Analysis is free software: you can redistribute it and/or modify
 # it under the terms of the GNU General Public License as published by
@@ -17,87 +17,351 @@
 # along with Mauve Analysis.  If not, see <http://www.gnu.org/licenses/>.
 ###############################################################################
 
+REPETITION=1
 function deseq {
-  PS3="Choose the species for $FUNCNAME: "
   select SPECIES in ${SPECIESS[@]}; do 
-    if [ "$SPECIES" == "" ];  then
-      echo -e "You need to enter something\n"
-      continue
-    else  
-      echo -n "What repetition do you wish to run? (e.g., 1) "
-      read REPETITION
-      global-variable $SPECIES $REPETITION
-      read-species
+  if [ "$SPECIES" == "" ];  then
+    echo -e "You need to enter something\n"
+    continue
+  else  
+    batch3-repetition
+    batch3-variable
+    deseq-speciesfile 
 
-      echo -n "What is the count file name? (e.g., count-1-11) "
-      read DESEQIN
-      echo -n "Enter the first sample? (e.g., UA159) "
-      read FRISTSAMPLE
-      echo -n "Enter the second sample? (e.g., TW1Glucose) "
-      read SECONDSAMPLE
-      # DESEQIN=count
-
-      RTEMP=$BWADIR/$RANDOM.R
-      OUT=$BWADIR/$DESEQIN.out
-      COMMAND="Rscript $RTEMP > $OUT"
-cat>$RTEMP<<EOF
-library(DESeq)
-countsFile <- "$BWADIR/$DESEQIN.txt"
[…]

### 2012-07-17 sh/goseq.sh

@@ -1,0 +1,186 @@
+###############################################################################
+# Copyright (C) 2012 Sang Chul Choi
+#
+# This file is part of RNAseq Analysis.
+# 
+# RNAseq Analysis is free software: you can redistribute it and/or modify
+# it under the terms of the GNU General Public License as published by
+# the Free Software Foundation, either version 3 of the License, or
+# (at your option) any later version.
+#
+# RNAseq Analysis is distributed in the hope that it will be useful,
+# but WITHOUT ANY WARRANTY; without even the implied warranty of
+# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+# GNU General Public License for more details.
+#
+# You should have received a copy of the GNU General Public License
+# along with RNAseq Analysis.  If not, see <http://www.gnu.org/licenses/>.
+###############################################################################
+
+REPETITION=1
+function goseq {
+  select SPECIES in ${SPECIESS[@]}; do 
+  if [ "$SPECIES" == "" ];  then
+    echo -e "You need to enter something\n"
+    continue
+  else  
+    batch3-repetition
+    batch3-variable
+    batch3-speciesfile 
+    goseq-speciesfile 
+    goseq-push-data
+    goseq-get-data
+    # goseq-make-job
+    goseq-run-goseq
+    batch3-copy-scripts
+    goseq-rmessage
+    break
+  fi
+  done
+}
+
+# Set species file variables
+function goseq-speciesfile {
+  GOAUNIPROT=$(grep ^GOAUNIPROT\: $SPECIESFILE | cut -d":" -f2)
+  GOAUNIPROT=$ROOTANALYSISDIR/$GOAUNIPROT
+  GOBO=$(grep ^GOBO\: $SPECIESFILE | cut -d":" -f2)
+  GOBO=$ROOTANALYSISDIR/$GOBO
[…]

### 2012-07-17 sh/run.sh

@@ -56,7 +56,6 @@ source sh/fastxtoolkit-fastx_quality_stats.sh
 
 source sh/fastq-sample.sh
 source sh/degseq.sh
-source sh/deseq.sh
 source sh/edgeR.sh
 source sh/transcript-cufflinks.sh
 source sh/transcript-genecoverage.sh
@@ -74,6 +73,8 @@ source sh/count-cds.sh
 source sh/batch.sh
 source sh/batch2.sh
 source sh/batch3.sh
+source sh/deseq.sh
+source sh/goseq.sh
 
 #####################################################################
 # Read configuration file
@@ -83,7 +84,7 @@ conf
 #####################################################################
 # Read directories
 #####################################################################
-SPECIESS=$(ls species|grep -v ^sim)
+SPECIESS=$(ls $ROOTANALYSISDIR/species|grep -v ^sim)
 
 #####################################################################
 # Menus
@@ -97,6 +98,8 @@ CHOICES=( init-file-system \
           batch3 \
           ---DE---\
           deseq \
+          ---GO---\
+          goseq \
           ---FASTQ---\
           fastq-sample \
           warranty \
@@ -136,6 +139,7 @@ select CHOICE in ${CHOICES[@]}; do
   elif [ "$CHOICE" == "fastq-sample" ]; then $CHOICE; break
   elif [ "$CHOICE" == "degseq" ]; then $CHOICE; break
   elif [ "$CHOICE" == "deseq" ]; then $CHOICE; break
+  elif [ "$CHOICE" == "goseq" ]; then $CHOICE; break
   elif [ "$CHOICE" == "edgeR" ]; then $CHOICE; break
   elif [ "$CHOICE" == "warranty" ]; then $CHOICE; break
   elif [ "$CHOICE" == "copyright" ]; then $CHOICE; break

### 2012-09-12 package-source/sagalactiae.Rnw
Manual html.

--- 2012-04-13 package/manuscript.Rnw
@@
 \usepackage{xr}
-\usepackage{url}
-\usepackage{../../latex/sty/genres}
-\externaldocument{supplementary}
 
@@
 \pagestyle{myheadings}
-\markright{Transcriptome of Streptococcus mutans}
+\markright{Transcriptome of Streptococcus agalactiae}
 
@@
 \title{Defining Transcriptome using RNA Sequencing\\
-of {\em Streptococcus mutans} }
+of {\em Streptococcus agalactiae} }
 
-\author{Sang Chul Choi$^{1}$, Charles D.\ Danko$^{1}$, 
-Melissa J.\ Hubisz$^{1}$, \\
-Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
-\fixme{Add more coauthors, or change the order of authors if needed} \\
-Robert A.\ Burne$^{3}$
-Michael J.\ Stanhope$^{2}$, and Adam Siepel$^{1}$}
+\author{Vince P.\ Richards$^{1}$, Sang Chul Choi$^{2}$,
+and Adam Siepel$^{2}$, and Michael J.\ Stanhope$^{1}$}
 
@@
 \begin{center}
-$^1$Department of Biological Statistics and Computational Biology,\\
-Cornell University, Ithaca, NY 14853, USA
-\\[1ex]
-$^2$Department of Population Medicine and Diagnostic Sciences,\\
+$^1$Department of Population Medicine and Diagnostic Sciences,\\
 College of Veterinary Medicine, Cornell University, Ithaca, NY 14853, USA
 \\[1ex]
-$^3$Department of Oral Biology,\\
-College of Dentistry, University of Florida, Gainesville, Florida 32610 USA
-\\
+$^2$Department of Biological Statistics and Computational Biology,\\
+Cornell University, Ithaca, NY 14853, USA
 \end{center}
@@
 
-\fixme{Add more corresponding authors.}
-
 \begin{tabular}{lp{4.5in}}
@@
 \vspace{1ex}\\
-{\bf Keywords:}&Carbon catabolite repression, RNA-seq, non-coding RNAs
+{\bf Keywords:}&RNA-seq, non-coding RNAs
 \vspace{1ex}\\
-{\bf Running Head:}&Transcriptome of {\em Streptococcus mutans}
+{\bf Running Head:}&Transcriptome of {\em Streptococcus agalactiae}
 \vspace{1ex}\\ 
@@
 
-\setcounter{page}{2}
+\setcounter{page}{1}
 
@@
 
-\fixme{Add more specific findings while summarizing the results. Connect two
-ideas of the differential expression study and transcriptome study.}
-A bacterial transcriptome of the primary etiological agent of human dental
-caries, \textit{Streptococcus mutans}, is described using deep RNA sequencing.
-The technique of RNA-Seq has allowed 
-the determination of transcript boundaries for annotated genes and
-the identification of previously unknown transcribed regions 
-including mRNAs and small noncoding RNAs in the genome.
-Differential expression studies on the role of CcpA in 
-carbon catabolite repression (CCR) suggested that some of the novel mRNAs and
-small RNAs found in our study could be involved in the regulation of CcpA. Our
-study provides more precise landscape of transcriptome changes in response to
-different carbon sources.
-The knowledge that we gained from the current study of \textit{S. mutans}
-transcriptome will lead to re-evaluation of the bacterial genome.
+A bacterial strain NEM316 of {\em Streptococcus agalactiae} has been known to
+originate in humans. We provide a series of evidence for the hypothesis that it
+is originated from cows.
 
@@
 
-% Biological motivations: S. mutans, CCR and CcpA
-About one fifth of children of ages ranging from 2 to 19 were reported to have
-untreated cavities in the United States \citep{health2010}. Oral bacteria digest
-foods containing carbohydrates left on the teeth to convert them to acids, which
-dissolve the surface of the teeth and create cavities.  The primary etiological
-agent of human dental caries, \textit{Streptococcus mutans}, is one of hundreds
-bacterial species that constitute the oral microbial community
-\citep{Jenkinson2005}.  Bacteria can selectively use substrates from different
-carbon sources via a regulation called carbon catabolite repression (CCR)
-\citep{Gorke2008}.  \textit{S.\ mutans} also possesses biological systems for
-CCR.  CCR were known to regulates the expression of virulence factors in many
-pathogenic bacteria, thereby subjecting a number of genes to CCR.  CcpA was
-shown to regulate metabolism and virulence gene expression in 
-\textit{S.\ mutans} through CCR in transcriptom study using 
-the microarray technique based on DNA hybridization \citep{Abranches2008}.
-
-% Technological motivations: RNA-seq, and microarray, bacterial transcriptomes
-It has been a decade since the first genome of the oral bacteria was finished
-\citep{Ajdic2002}.  Studies of the transcriptome in the bacteria were performed
-based on DNA hybridization
-\citep[e.g.,][]{Klein2010,Lemme2011,Redanz2011,Wen2011} whereas those in other
-bacterial species took advantage of high-throughput sequencing technologies:
-e.g.,
-\textit{Mycoplasma pneumoniae} \citep{Guell2009},
-\textit{Methanosarcina mazei} \citep{Jaeger2009},
-\textit{Bacillus subtilis}  \citep{Irnov2010},
-\textit{Helicobacter pylori} \citep{Sharma2010}, and
-\textit{Sulfolobus solfataricus} \citep{Wurtzel2010}, to name a few.
-The advent of the high-throughput RNA deep sequencing technologies allowed
-annotation-free detection of transcripts, sensitive and dynamic ranges of
-measurements in RNA expression levels, and studies of non-coding RNAs.
-
-% Aims of the study
-We used the standard RNA deep sequencing technique to sequence RNAs in
-\textit{S.\ mutans}, focusing on messenger RNAs expression levels in response to
-different nutrients with UA159 wild-type and \textit{ccpA}-deleted TW1
-mutant-type.  We describe transcripts of the mRNAs using the RNA-seq
-data, and present several small non-coding RNAs in \textit{S.\ mutans}. 
-Using the map of transcripts
-and expression levels of the transcripts from different biological conditions
-allowed us to determine which transcripts and non-coding small RNAs were
-differentially expressed. This provided more refined views of dynamic changes in
-the transcriptome landscape of the etiological agent of human dental caries,
-\textit{S.\ mutans}, than previously reported by \citet{Abranches2008}.
-
 \section*{RESULTS}
 
-<<eval=FALSE,echo=FALSE,results=hide>>=
-# not run when smutans R package is installed
-# Use this when smutans R package is being tested.
-library(DESeq)
-library(GenomicRanges)
-library(plotrix)
-source("smutans/R/class_and_slots.R")
-source("smutans/R/core.R")
-source("smutans/R/methods.R")
-load("smutans/data/smutansData.tx.RData")
-load("smutans/data/smutansData.txGenes.RData")
-load("smutans/data/smutansData.txGrCombined.RData")
-load("smutans/data/smutansData.txAnnotation.RData")
-load("smutans/data/smutansData.txRNAz.RData")
-load("smutans/data/smutansData.txPileup.RData")
-load("smutans/data/smutans.feature.genes.RData")
-load("smutans/data/smutans.genes.criteria.RData")
-load("smutans/data/smutans.go.genes.RData")
-load("smutans/data/smutans.cat.desc.RData")
-load("smutans/data/smutans.keg.genes.RData")
-load("smutans/data/smutans.keg.desc.RData")
-load("smutans/data/smutansGenes.RData")
-load("smutans/data/smutansGenes2.RData")
 @
 <<echo=FALSE,results=hide>>=
-# not run without smutans R package installed
-# Remove eval=FALSE when the R package is installed
 library(smutans)
-inDir = system.file("extdata", package="smutans", mustWork=TRUE)
-smutansGenes <- 
-  readSmutans (countsFile=file.path(inDir, "count-ua159-bwa.txt"),
-               indexFile =file.path(inDir, "count-ua159.txt.index"),
-               condition = c("UA159GLU", "UA159GAL", "TW1GLU", "TW1GAL"),
-               firstFactor = c("UA159GLU", "UA159GAL"),
-               firstFactorLabel = c("ua159", "tw1"),
-               secondFactor = c("UA159GLU", "TW1GLU"),
-               secondFactorLabel = c("glucose","galactose"),
-               name="S. mutans UA159, TW1, Glucose, and Galactose")
-data(smutansData.tx)
-data(smutansData.txGenes)
-data(smutansData.txGrCombined)
-data(smutansData.txAnnotation)
-data(smutansData.txRNAz)
-data(smutansData.txPileup)
-data(smutans.feature.genes)
-data(smutans.genes.criteria)
-data(smutans.go.genes)
-data(smutans.cat.desc)
-data(smutans.keg.genes)
-data(smutans.keg.desc)
-data(smutansGenes2)
-@
 
-<<echo=FALSE,result=hide>>=
-notExpressedTx <- smutansData.txGrCombined[elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5]
-grTx <- smutansData.txGrCombined[!(elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5)]
-# We query genes and annotations (genes and RNAs) againt tx's.
-matchGenesTx <- findOverlaps(smutansData.txGenes,grTx)
-matchAnnotationTx <- findOverlaps(smutansData.txAnnotation,grTx)
-# I use gaps with IRanges because gaps with GenomicRanges does not work as expected.
-grNoAnnotation <- gaps(ranges(smutansData.txAnnotation))
-grNoncoding <- gaps(ranges(smutansData.txGenes))
-grRNAz <- ranges(smutansData.txRNAz)
-# 
-matchUnannotatedRNA <- findOverlaps(grRNAz,grNoAnnotation,type="within")
-txRNAz.unannotated <- grRNAz[as.matrix(matchUnannotatedRNA)[,1]]
+for (i in c("NC_004368")) {
+file.1 <- paste("../output/agalactiae/1/bwa/count-",i,".txt", sep="")              
+file.out.1 <- paste(i,"-sag56-A-milk-B-broth-clust.pdf", sep="")                         
+file.out.2 <- paste(i,"-sag56-A-milk-B-broth.csv", sep="")                               
+file.out.3 <- paste(i,"-sag56-A-milk-B-broth.ps", sep="")
 
-# A bug in GenomicRanges: the last segment is the while genome. I removed it
-# from the noncoding ranges.
-# matchNoncoding <- findOverlaps(grNoncoding,grTx)
+agGenes <-
+  readSmutans(countsFile=file.1,
+              indexFile="../output/agalactiae/1/run-analysis/count.txt.index",
+              condition = c("sag56hm", "sag56hb"),
+              firstFactorLabel = c("NEM316"),
+              secondFactorLabel = c("sag56hm", "sag56hb"),
+              name="S. agalactiae NEM316 in milk versus broth")
 
-# Expressed transcripts without annotated genes
-grTxWGenes <- grTx[unique(as.matrix(matchGenesTx)[,2])]
-grTxWOGenes <- grTx[-unique(as.matrix(matchGenesTx)[,2])]
-grTxWOAnnotation <- grTx[-unique(as.matrix(matchAnnotationTx)[,2])]
+ag <- newSmutans( agGenes , title="S. agalactiae NEM316 in milk versus broth")
 
-# Find transcripts with genes with conflicted strands
-grGenesPlus <- smutansData.txGenes[strand(smutansData.txGenes)=='+']
-grGenesMnus <- smutansData.txGenes[strand(smutansData.txGenes)=='-']
-#mtchPlus <- findOverlaps(grGenesPlus,grTxWGenes)
-#mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
-#txWithConflictedGenes <- intersect(as.matrix(mtchPlus)[,2],as.matrix(mtchMnus)[,2])
+pdf(file.out.1)                                                                 
+smutans.de2Clust( ag )                                                          
+dev.off()
 
-# Tx with genes strands
-grTx2 <- grTxWGenes[!strand(grTxWGenes)=='*']
-
-# Find the number of genes in segments.
-matchGenesTx2 <- findOverlaps(smutansData.txGenes,grTx2,type="within")
-number.genes.tx <- runLength(Rle(as.matrix(matchGenesTx2)[,"subjectHits"]))
-
-# Length distribution of UTR regions
-# grTx2 <- grTxWGenes[-txWithConflictedGenes] 
-# Lengths of 5' and 3' UTRs
-mtchPlus <- findOverlaps(grGenesPlus,grTx2,type="within")
-mtchMnus <- findOverlaps(grGenesMnus,grTx2,type="within")
-length5UTR <- c()
-length3UTR <- c()
-for (i in unique(as.matrix(mtchPlus)[,"subjectHits"])) {
-  x <- as.matrix(mtchPlus)[as.matrix(mtchPlus)[,"subjectHits"]==i,"queryHits"]
-  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
-  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
-  length5UTR <- c(length5UTR,y)
-  length3UTR <- c(length3UTR,z)
-}
-for (i in unique(as.matrix(mtchMnus)[,"subjectHits"])) {
-  x <- as.matrix(mtchMnus)[as.matrix(mtchMnus)[,"subjectHits"]==i,"queryHits"]
-  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
-  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
-  length5UTR <- c(length5UTR,z)
-  length3UTR <- c(length3UTR,y)
-}
-
-# Correlation of expression levels of adjacent genes.
-pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
-pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
-elementMetadata(smutansData.txGenes)["score"] <- pileup.y
-mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
-x <- rle(as.matrix(mtch3)[,"subjectHits"])
-comp.exp.two.genes <- c()
-for (i in x$values[x$lengths > 1]) {
-  y <- combn(as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==i],2)
-  for (j in 1:dim(y)[2]) {
-    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
-  }
-}
-# Correlation of expression levels of adjacent genes between transcripts.
-comp.exp.two.genes.between.transcript <- c()
-for (i in 1:length(x$values)) {
-  if (i == length(x$values)) {
-    break
-  }
-  y1 <- as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==x$values[i]]
-  i2 <- i + 1
-  y2 <- as.matrix(mtch3)[,"queryHits"][as.matrix(mtch3)[,"subjectHits"]==x$values[i2]]
-  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
-  for (j in 1:dim(l)[1]) {
-    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
-  }
+ag <- smutans.de2( ag, type="NEM316", condA="sag56hm", condB="sag56hb")
+smutans.de2List( ag, file=file.out.2 )                                          
+smutans.plotDiffExp(ag,file=file.out.3,qval=0.05)                               
 }
@@
 
-\subsection*{Experimental results from Burne group}
-\fixme{Add some if needed.}
-
-\subsection*{Results from Stanhope group when RNA sequencing preparation}
-\fixme{Add some if needed.}
-
-\subsection*{Cluster of RNA samples}
-
-First using the table of read counts of annotated genes we formed a cluster of
-the 13 RNA-seq samples (fig.\ \ref{figClust}). The effect of nutrients appeared
-to be stronger than that of the deletion of \textit{ccpA} genes from the UA159
-genome.
-
-\subsection*{Predicted transcripts}
-
-We obtained a set of segments of RNA-seq expressions for each of the 13 RNA-seq
-samples.  All of the segments for each set of RNA-seq expression segments were
-scored by the average proportions of segments shared by the segments in the rest
-of sets.  We placed more highly ranked segments first on the reference genome
-based on the scores, and purged more lowly ranked segments from the
-transcriptome map. In doing so, the
-final set of segments was formed by placing non-overlapping highly ranked
-segments.  Average site-wise expression levels of the segments 
-are shown at figure \ref{figGrCombinedExpression}. 
-We designated segments as expressed that had site-wise pileup values
-greater than 5 and proportion of sites of zero site-wise pileup values less than
-10\%.  A gene was also designated as an expressed if the gene belonged to an
-expressed segment.  Some of the predicted segments were not expressed or very
-lowly expressed (\Sexpr{length(notExpressedTx)} 
-out of 
-\Sexpr{length(smutansData.txGrCombined)}).
-Of the total \Sexpr{length(smutansData.txGenes)} of genes,
-\Sexpr{length(as.matrix(matchGenesTx)[,1])}
-(\Sexpr{as.integer(length(as.matrix(matchGenesTx)[,1])/length(smutansData.txGenes)*100)}\%) 
-were designated as expressed. 
-Genes that were not designated as expressed could be due to false gene
-annotations or possibility of true negative expression under the experimental
-conditions. 
-Figure \ref{figLength} shows the length distribution of expressed and
-not-expressed genes. 
-Of the \Sexpr{length(grTx)} expressed segments, 
-\Sexpr{length(grTxWOAnnotation)} contained no annotated genes or RNAs.
-Many of the segments might cover gaps between real operons.
-\fixme{Examine these unannotated regions in the UCSC genome tracks.}
-
-Figure \ref{figTranscriptLength} shows the length distribution of expressed
-segments with annotated coding genes and unannotated regions.  Unannotated
-regions could include potential non-coding regions and novel coding genes.
-Segments of unannotated regions appeared to be shorter than those with
-coding genes. Most of the unannotated regions with length being near 360 bp
-could be non-coding regions, and the rest of the unannotated regions with length
-being near 1000 bp could be new coding genes. Some of the unannotated genes
-could be pseudogene regions.  We used strandnesses of known genes to find
-strands of segments although we could not tell strands of the predicted
-transcripts solely based on our RNA-seq data because of our strand-nonspecific
-RNA-seq technique.  Of
-the \Sexpr{length(grTx)} expressed transcripts, 
-\Sexpr{sum(strand(grTx)=='*')}
-included annotated genes of both strands
-(\Sexpr{as.integer(sum(strand(grTx)=='*')/length(grTx)*100)}\%).
-We filtered out the transcripts with genes with conflicted both strands leaving
-\Sexpr{length(grTx2)} expressed transcripts. 
-Figure \ref{figUTRLength} shows length distributions of 5' and 3' UTR of the
-expressed transcripts without genes with conflicted strands.  Many transcripts
-appeared to start and end near the boundaries of annotated genes.  Gene
-expression levels for adjacent genes within expressed transcripts were more
-correlated than those between adjacent transcripts 
-(fig.\ \ref{figExpressionTwoGene}).
-Among the \Sexpr{length(unique(as.matrix(matchGenesTx2)[,"subjectHits"]))} segments
-with genes of all of the same strands, 
-\Sexpr{sum(number.genes.tx==1)} segments contained a single gene while 
-\Sexpr{sum(number.genes.tx > 1)} contained more than one gene.
-
-\subsection*{Predictions of small RNAs}
-
-We predicted
-\Sexpr{length(colnames(smutans.genes.criteria))-1} putative non-coding RNAs, most of
-which were already annotated.  
-Of the \Sexpr{length(colnames(smutans.genes.criteria))-1} predicted non-coding RNAs
-\Sexpr{length(txRNAz.unannotated)} were in  
-unannotated non-coding regions (table \ref{tabUnannotatedRNA}).  Using
-\program{RNAplex} \citep{Tafer2008} and \program{RNAplfold} \citep{Bernhart2006}
-we predicted target genes of the putative non-coding RNAs.  Table \ref{tab:goseq}
-shows functional categories enriched with target genes for predicted small RNAs.
-We found that a few of the putative small RNAs were structured with stable stems
-(figs.\ \ref{figRNALocus1}, \ref{figRNALocus2}, and \ref{figRNALocus3}).  
-\fixme{We might be able to 
-further experimentally verify potentially important small RNAs.}
-
 \subsection*{Differential expression of RNA transcripts}
-
-<<echo=FALSE,result=hide>>=
-glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
-glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
-                            condA="ua159", condB="tw1" )
-save(glucoseOnly, file=file.path("glucoseOnly.RData"))
-#load("glucoseOnly.RData")
-
-smutans.de2List( glucoseOnly, "glucoseOnly.csv" )
-tableGlucoseOnly <- smutans.de2List( glucoseOnly, qval=1e-3 )
-
-# Gene descriptions
-geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
-geneDescription <- geneDescription[,c(4,7)]
-colnames(geneDescription) <- c("Gene ID","Gene Description")
-# Functional classes
-functionalDescription <- read.csv("feature-genome.out-geneonly-oralgen-functional.txt",head=FALSE)
-functionalDescription <- functionalDescription[,c(2,12)]
-colnames(functionalDescription) <- c("Gene ID","Functional Class")
-# Find potential differentially expressed genes, and add gene descriptions, and
-# functional classes
-x <- smutans.de2List( glucoseOnly, qval=1e-3 )
-x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
-colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
-x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
-x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
-tableGlucoseOnly <- x
-@
-
-Two independent factors of the biological processes included the effect of CcpA
-and the treatment of mono-saccharides (glucose versus galactose), resulting in
-four pair-wise comparisons. We compared two conditions of one factor with the
-other factor being fixed at one condition. For example, we compared UA159
-wild-type and TW1 mutant-type strains under the condition of glucose-grown cells
-to determine differentially expressed transcripts or genes potentially due to
-presence and absence of deactivation of \textit{ccpA} genes.  We performed a
-similar comparison with galactose-grown cells. We replicated four
-biological experiments for the condition combination of UA159 wild-type grown
-in galactose, and three replicates for the other three conditions.
-
-\subsection*{Comparison between wild-type and mutant-type in glucose-grown cells}
-
-Table \ref{tabGlucoseOnly} lists differentially expressed genes in the
-comparison of UA159 wild-type and TW1 mutant-type grown with glucose.
-Of \Sexpr{nrow(counts(smutansGenes))} genes, 
-\Sexpr{nrow(tableGlucoseOnly)} genes were found to be differentially expressed
-at FDR of 0.1\%.  We confirmed that the most up-regulated genes in the TW1
-mutant-type encoded the components of the PDH enzyme complex (SMU.1421,
-SMU.1422, SMU.1423, and SMU.1424) as shown previously.  
-These genes appeared to be transcribed in
-a single operon (fig.\ \ref{figUCSC}). 
-Figure \ref{figPieGlucoseOnly} shows the
-proportions of \program{oralgen} databases functional classes by the number of
-differentially expressed genes.  Notably, we found
-\Sexpr{length(grep("Energy metabolism",tableGlucoseOnly[,"Functional Class"]))}
-genes that were involved in energy metabolism where sugars and fermentation
-subclasses were profound. We found 
-\Sexpr{length(grep("PTS",tableGlucoseOnly[,"Functional Class"]))}
-genes in PTS class of signal transduction. Another subclass, 
-``carbohydrates, organic alcohols, and acids,'' in
-``transport and binding proteins'' was following next in the number of
-differentially expressed genes. We also found  
-\Sexpr{length(grep("hypothetical",tableGlucoseOnly[,"Gene Description"]))}
-genes annotated as hypothetical, which might constitute the unknown functional
-class. These results appeared to be consistent with those of \citet{Abranches2008}.
-
-We further tested functional category enrichments using gene ontology terms and KEGG
-pathways. Table \ref{tab:goseqGlucoseOnly} shows that differentially expressed
-genes were associated with the following gene ontology terms:
-phosphotransferase system, 
-glycogen biosynthetic process, and
-sugar:hydrogen symporter activity.
-Among the KEGG pathways, starch and sucrose metabolism and phosphotransferase
-system (PTS) appeared to be highly associated with the differentially expressed
-genes (table \ref{tab:kegseqGlucoseOnly}).
-Figure \ref{figDotplotGlucoseOnly} shows that differentially expressed genes
-tended to increase in expression levels in response to the deletion of
-\textit{ccpA} genes.  Profoundly upregulated genes were putative Clp proteinase
-(SMU.1425), branched-chain alpha-keto acid dehydrogenase subunit E2 (SMU.1421),
-and putative transcriptional regulator (SMU.1599).  Much less proportion of the
-differentially expressed genes were found to be decreased in expression levels
-in response to the lack of CcpA (fig.\ \ref{figDotplotGlucoseOnly}).
-Down-regulated genes include cytoplasmic alpha-amylase (SMU.1590),
-beta-D-fructosyltransferase (SMU.2028), and putative glycerol uptake facilitator
-protein (SMU.396).  Fructosyltransferase converts sucrose to fructose.
-
-Figure \ref{figFoldPValueGlucoseOnly} shows fold changes and $p$-vallues are
-negatively correlated. \fixme{Why did I want to note this observation?}
-
-\subsection*{Comparison between wild-type and mutant-type in galctose-grown cells}
-
-<<echo=FALSE,result=hide>>=
-galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
-galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
-                              condA="ua159", condB="tw1" )
-save(galactoseOnly, file=file.path("galactoseOnly.RData"))
-#load("galactoseOnly.RData")
-x <- smutans.de2List( galactoseOnly, qval=1e-3 )
-x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
-colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
-x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
-x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
-tableGalactoseOnly <- x
-@
-We found that differentially expressed genes in galactose-grown cells were
-smaller than those in glucose-grown cells, which was different from the result
-of \cite{Abranches2008}. 
-Energy metabolism was the most
-profoundly differentially expressed (fig.\ \ref{figPieGalactoseOnly}).
-Subclasses pyruvate family in amino acid biosynthesis, 
-and degradation of proteins, peptides, and glycopeptides in protein fate
-were following the next. Almost a quarter of the genes were unknown.
-Down-regulated genes included cytoplasmic alpha-amylase (SMU.1590), and
-beta-D-fructosyltransferase (SMU.2028)
-(table \ref{tabGalactoseOnly}). We could not find gene ontology terms that were
-significantly associated with differentially expressed genes. It might be
-because many genes were unknown in their functions 
-(fig.\ \ref{figDotplotGalactoseOnly}).
-However, a few KEGG pathways including valine, leucine and isoleucine
-biosynthesis were found to be significant (table \ref{tab:kegseqGalactoseOnly}).
-
-\subsection*{Comparison between glucose- and galactose-grown wild-type cells}
-
-<<echo=FALSE,result=hide>>=
-ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
-ua159only <- smutans.de2( ua159only, type="ua159", 
-                          condA="glucose", condB="galactose" )
-save(ua159only, file=file.path("ua159only.RData"))
-#load("ua159only.RData")
-x <- smutans.de2List( ua159only, qval=1e-10 )
-x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
-colnames(x) <- c("Gene ID","Glucose","Galactose","fold", "p-value")
-x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
-x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
-tableUA159Only <- x
-@
-Among the four pairwise comparisons this showed the largest number of
-differentially expressed genes (table \ref{tabUA159Only}). We used a smaller
-cutoff value for listing them. 
-Figure \ref{figPieUA159Only} shows proportons of functional categories that
-included differentially expressed genes between glucose-grown and
-galactose-grown UA159 wild-type cells. Qualitatively, energy metabolism and PTS
-classes were profound. 
-``Energy metabolism; Fermentation'' had highly differentially expressed genes
-(fig.\ \ref{figDotplotUA159Only}).
-This was similar to the comparison between UA159
-wild-type and TW1 mutant-type grown in glucose. 
-More than a quarter of the
-differentially expressed genes were unknown, unassigned, or hypothetical.
-Consecutive genes (SMU.1488c to SMU.1496) were upregulated in 
-catabolite-derepressing conditions or in galactose-grown cells
-(\fixme{Are these the genes of the lactose (lac) operon?})
-(table \ref{tabUA159Only}).  Another set of genes (SMU.886 to SMU.889) were also
-upregulated (\fixme{Are these called galKTE?}). Whereas the
-four genes appeared to be transcribed together under galactose condition,
-two segments seemed to be transcribed by dividing the four genes into two pairs
-of genes (fig.\ \ref{figUCSC2}).  Genes from SMU.1535 to SMU.1539, which
-appeared to be transcribed together, were upregulated.
-Of the \Sexpr{nrow(tableUA159Only)} differentially expressed genes at
-$1 \times 10^{-10}$ cutoff of multiple testing adjusted p-value, 
-\Sexpr{length(grep("hypothetical",tableUA159Only[,"Gene Description"]))} were
-hypothetical, and most of them were down-regulated
-(fig.\ \ref{figDotplotUA159Only}).
-We also tested functional category enrichments using gene ontology terms 
-(table \ref{tab:goseqUA159Only}) and KEGG pathways 
-(table \ref{tab:kegseqUA159Only})
-where carbohydrate metabolic process and Galactose metabolism were top listed,
-respectively.
-
-\subsection*{Comparison between glucose- and galactose-grown mutant-type cells}
-
-We found sugar metabolism and PTS signal transduction to be profound in
-differential expression in the comparison of glucose-grown and galactose-grown
-TW1 mutant-type cells (fig.\ \ref{figPieTW1Only}). Many of the genes were also
-hypothetical. We found similar genes differentially expressed in TW1 mutant-type
-cells to those in UA159 wild-type 
-(table \ref{tabTW1Only} and fig.\ \ref{figDotplotTW1Only}).
-The comparisons between glucose-grown and galactose-grown UA159, 
-and that from TW1 mutant-type showed very similar functional categories
-although ranks of gene ontology terms were different
-(tables \ref{tab:goseqUA159Only} and \ref{tab:goseqTW1Only}).
-Lactose metabolic process was top listed in TW1, and
-carbohydrate metabolic process was top listed in UA159.
-KEGG pathways in TW1 were also similar to those in UA159 
-(table \ref{tab:kegseqTW1Only}) .
-
-\subsection*{Comparison between glucose- and galactose-grown cells using all samples}
-As previously figure \ref{figClust} shows the nutrient effect was stronger, and
-the effects of nutrient changes appeared to be similar whether UA159
-wild-type or TW1 mutant-type was used, we used all of the 
-\Sexpr{length(colnames(counts(smutansGenes)))} samples to compare glucose-grown
-and galactose grown cells.
-<<echo=FALSE,results=hide>>=
-allSample <- newSmutans( smutansGenes, 
-                         title="All Samples for Comparing Conditions" )
-allSample <- smutans.de2( allSample, 
-                          condA="glucose", condB="galactose" )
-save(allSample, file=file.path("allSample.RData"))
-#load("allSample.RData")
-tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
-                                                   allSample )
-#addmargins( tabUA159onlyAndAllSample )
-@
-Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
-\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found to be differentially
-expressed in both the analysis of using only UA159-grown cells and the analysis
-of using all of the 13 samples. The
-analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
-genes. The analysis using the partial samples found only
-\Sexpr{tabUA159onlyAndAllSample[2,1]} genes.  We would expect that it was 
-because of higher power of the two-factor analysis with more data.
-
-\subsection*{Differentially expressed small RNAs and transcripts}
-
-<<echo=FALSE,results=hide>>=
-ua159only <- newSmutans( smutansGenes2, title="UA159 Only" )
-ua159only <- smutans.de2( ua159only, type="ua159", 
-                          condA="glucose", condB="galactose" )
-allSample <- newSmutans( smutansGenes2, 
-                         title="All Samples for Comparing Conditions" )
-allSample <- smutans.de2( allSample, 
-                          condA="glucose", condB="galactose" )
-tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
-                                                   allSample )
-# smutans.de2List( allSample )[grep("locus",rownames(smutans.de2List(allSample))),]
-@
-We found a few predicted small RNAs that appeared to be differentially
-expressed: locus69, locus101, locus7, and locus1. 
-Two non-coding regions of locus69 and locus101 seemed to be too weak.
-However, locus1 and locus7 seemed to have some noticeable differences between
-glucose and galactose conditions.
-We also found a few predicted small RNAs that appeared to be differentially
-expressed in the comparison of the two types UA159 and TW1: locus68,
-locus8, and locus10.  The locus68 was overlapped with T-box region.
-The two regions of locus8 and locus10 showed very weak and subtle differences
-between UA159 and TW1. 
-<<echo=FALSE,results=hide>>=
-allSampleType <- newSmutans( smutansGenes2, title="All samples for type" )
-allSampleType <- smutans.de2( allSampleType, compareCondition="no", 
-                              condA="ua159", condB="tw1" )
-# smutans.de2List( allSampleType )[grep("locus",rownames(smutans.de2List(allSampleType))),]
-@
-Figures \ref{figRNALocus1}, \ref{figRNALocus2}, and \ref{figRNALocus3} shows MFE
-structures with encoding positional entropy of differentially expressed putative
-structural non-coding RNAs.
-
-\section*{DISCUSSION}
-
-Our current knowledge of the transcriptome is limited to annotated or predicted
-genes.  Whole transcriptome sequencing using high throughput sequencing
-technologies, or deep RNA sequencing (RNA-seq) allowed to reveal the
-transcriptome topography of \textit{S.\ mutans}.  Instead of aiming at
-pin-pointing transcript start or end sites we focused on differential expression
-of genes in response to nutirent availability and deletion of \textit{ccpA} from
-UA159.  Although we did not use strand-specific RNA-seq that could be important
-in gene-dense transcriptomes of species like bacteria or archaea, our
-application of a hidden Markov model for predicting transcripts to RNA-seq data
-in conjunction with known gene annotations allowed to infer segments of
-putatively co-transcribed regions. In doing so, we attempted to ensure a
-high-quality transcriptome assembly.  Although this map of transcripts did not
-precisely tell us where transcripts start and end, it informed us of which open
-reading frames were transcribed together.  We also found discrepancies of
-annotated genes and their transcriptions: a gene contains two transcripts, and
-non-coding regions contain transcripts.  We hope that our study could serve as a
-foundation for a comprehensive study of the \textit{S.\ mutans} transcriptome.
 
@@
 \subsection*{Bacterial strains and growth conditions}
-\fixme{Burne group describes bacterial strains and growth conditions and other
-necessary works. Please, add texts and SCC will copy-edit them.}
 
 \subsection*{RNA isolation}
-\fixme{Burne group describes RNA isolations and other works. Please, add texts
-and SCC will copy-edit them.}
 
 \subsection*{RNA sequencing}
-Deep sequencing of the RNA samples were performed at the Cornell University Life
-Sciences Core Laboratories Center (Ithaca, NY).  RNA extraction was performed
-using a modified Qiagen RNAease protocol (Qiagen, Inc., Chatsworth, CA): cells
-were pelleted, resuspended in RNA protect reagent, frozen, and then resuspended
-in TE, mechanically disrupted in a bead beater in the presence of glass beads
-and acidic phenol. We used the Qiagen kit to obtain total RNAs.  The RNA sample
-was passed through the Ambion MicrobExpress kit twice, which removed ribosomal
-RNAs only using magnetic beads.  This procedure left small RNAs and transfer
-RNAs intact.  Then, we followed the TruSeq Illumina kit to prepare the cDNA
-libraries.
 
@@
 \subsection*{Transcript predictions}
-RNA transcripts were inferred by applying a hidden Markov model to site-wise
-expression levels \citep{Martin2010}.  We used the pileup command of
-\program{BWA} to convert the short reads alignments to pilup values, which were
-taken as the site-wise expression levels along the genome.  Site-wise expression
-levels are a list of site-wise non-negative integers that represents numbers of
-short reads mapped to a particular genomic position.  We used the following
-options for the transcript inference program, \program{ParseRNAseq}: ``-c 10 -b
-25 -force\_gp'' to bin expression levels into 25 parts, and to allow 10 emission
-states for relative expression intensity. We also used the genes as the
-predicted parts of transcripts.    Because we enriched mRNAs in the cDNA
-preparation, we intented to group annotated genes to segments, which could be
-transcripts. We did not aim at finding precise transcription start or end sites.
-Because we did not have strand-specific RNA-seq of the samples due to limited
-resources, we could not tell in which directions transcripts were oriented
-solely based on RNA-seq data. Instead, we used strands of annotated genes to
-determine strands of predicted transcripts.
 
 \subsection*{Prediction of small RNAs and targets}
-Although our RNA-seq data did not enriched non-coding small RNAs in the cDNA
-preparation, we retained small RNAs in the sample by attempting to deplete only
-ribosomal RNAs from the sample. Because non-coding small RNAs and coding
-mRNAs were mixed in the sample, it was difficult to discriminate small
-non-coding RNAs from mRNAs unless non-coding RNAs expressions were masked by
-neighboring background RNA expressions. 
-Therefore, we sought a way of predicting small RNAs
-using \program{RNAz} \citep{Gruber2010}, which  used homologous sequences and
-RNA secondary structures to predict putative non-coding RNAs. Because sequence
-alignments were a critical step for finding small RNAs in the approach, we
-collated sequences similar to intergenic regions, and refined the alignments using
-a multiple sequence alignment tool. Exracting
-intergenic regions including up- and down-stream, we searched a database of
-bacterial genomes using BLAST \citep{Altschul1990} for similar intergenic
-sequences, subsequently, refining the sequence alignments using MUSCLE
-\citep{Edgar2004a}.  Then, we applied \program{RNAz} to the alignments for
-scoring regions for small RNAs \citep{Washietl2005,Gruber2010}.  Targets of
-transcripts with high scores for small RNAs were also predicted using
-\program{RNAplex} \citep{Tafer2008}, and \program{RNAplfold}
-\citep{Bernhart2006}.  Target genes were scored for each small RNA candidate,
-which was used to perform functional category enrichment tests of targets for
-each small RNA candidate.
 
 \subsection*{Experimental validation of small RNAs}
-\fixme{Burne group might have done or be able to do some experimental
-validations of putative non-coding RNAs shown at table \ref{tabUnannotatedRNA}
-\citep[e.g.,][]{Altuvia2007}.}
 
 \subsection*{Statistical analysis for differential expressions}
-We used \Rpackage{DESeq} \citep{Anders2010} to determine differentially
-expressed genes based on negative binomial model originally developed by
-\citet{Robinson2007}.  Steps of analyzing RNA-seq data for determining
-differentially expressed genes were described in \cite{Oshlack2010}.  We counted
-short reads aligned to annotated genes in the reference genome, forming a table
-of read counts of the genes. Statistical software R \citep{R2011} with
-\Rpackage{DEseq} was subsequently employed to infer differentially expressed
-genes in the different biological conditions.
 
 \subsection*{Gene category associations}
-We compiled three sets of gene categories for testing functional associations.
-Firstly, we assigned genes to Gene Ontology (GO) categories by comparing the
-\textit{S.\ mutans} genes to bacterial proteins from the Uniref90 database using
-\program{BLASTP}, and then assigning the same GO classification as the target
-gene of the uniProt GOA database if the match had an $E$-value of
-$<1.0\times10^{-5}.$  A gene family was assigned a given classification if any
-of its genes was assigned that classification.  Secondly, we collated functional
-classes of genes in the Oral Pathogen Sequence Databases available at
-\url{http://oralgen.lanl.gov}. Thirdly, we used a set of metabolite pathways for
-\textit{S.\ mutans} available at Kyoto Encyclopedia of Genes and Genomes (KEGG)
-\citep{Kanehisa2012}.  Using the sets of gene categories for \textit{S.\ mutans}
-we performed a Mann-Whitney $U$ test of the values of target genes of small RNAs
-associated with a given category versus those for the other categories. We
-used a variation of Fisher's exact test accounting for gene lengths to test
-enrichments of differentially expressed genes \citep{Young2010}.  Both of the
-tests were corrected for multiple testing hypotheses \citep{Benjamini1995}.
 
 \section*{DATA ACCESS}
-
-\subsection*{UCSC genome browser tracks and data availability}
-We created tracks for our recently released Streptococcus Genome Browser
-(\url{http://strep-genome.bscb.cornell.edu}) that summarized the results of our
-\textit{S.\ mutans} transcriptome analysis with known genes, gene expression
-levels based on the short read alignments, predicted putative transcript
-segments, and predictd small non-coding RNAs.  They should be accessed by clade
-\textit{Streptococcus}, and genome \textit{S. mutans UA159}.  These tracks can
-be used to inspect loci of interest and to compare the results of different
-RNA-seq data sets.  They can also be queried and intersected with other tracks
-using the UCSC Table Browser.
 
@@
 
-\renewcommand*{\refname}{REFERENCES}
-\bibliographystyle{../../latex/bst/genres}
+\bibliographystyle{mbe}
 \bibliography{siepel-mutans}

### 2012-09-12 package-source/tutorial.Rnw
Manual html.

%\VignetteIndexEntry{Analyzing Streptococcus mutans RNA-seq data}
%\VignettePackage{smutans}

% To compile this document
% library('weaver'); rm(list=ls()); Sweave('smutans.Rnw', driver=weaver()); system('pdflatex smutans')

\documentclass{article}

\usepackage{Sweave}
\usepackage[letterpaper]{geometry}
\usepackage{hyperref,graphicx}
\usepackage{color}
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=4,height=4.5}
\newcommand{\Robject}[1]{object \texttt{#1}}
\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\Rclass}[1]{class \textit{#1}}
\newcommand{\Rfunction}[1]{{function \small\texttt{#1}}}
\newcommand{\program}[1]{{\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{blue}{#1}}}}

\renewcommand{\floatpagefraction}{0.7}

\author{Sang Chul Choi\\[1em]Cornell University,\\ Ithaca, NY\\[1em]
\texttt{schoi@cornell.edu}}

\title{\textsf{\textbf{A Toturial of BioConductor R packages}}}

\begin{document}

\maketitle

\begin{abstract}
BioConductor R packages have been useful in analyzing RNA-seq data sets. I wish
to record for the instructional purposes.
\end{abstract}
\tableofcontents

\section{Genomes}

\subsection{Sample sequences from a genome}

\section{Short reads}

Steps of analyzing RNA-seq data for determining differentially expressed genes
are described in \cite{Oshlack2010}.

\subsection{Create short reads}

\subsection{Read short reads}

\Rpackage{ShortRead}

<<eval=FALSE,echo=TRUE>>=
library(ShortRead)

fastqFileOrg <- "/v4scratch/sc2265/rnaseq/output/ua159/1/data/FASTQ001.fq.gz"
fastqFile <- "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.prinseq.fq.gz"

fastqFile <- "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.recovered.fq.gz"
fq <- readFastq(fastqFile, withIds=TRUE)
fqOrg <- readFastq(fastqFileOrg, withIds=TRUE)

# Number of reads
length(fq)
# The first 3 Reads
head(sread(fq), 3)
# The first 3 Read quality scores
head(quality(fq), 3)
# The first 3 Read IDs
head(id(fq), 3)
# Tally reads by length
table(width(fq))
# Remove reads with N
fqN <- fq
fq <- fqN[grep("N",sread(fqN),invert=TRUE)]
# 
sampler <- FastqSampler(fastqFile, 100000)
fqSample <- yield(sampler)
report(qa(fqSample),dest="/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/report001")

qa2 <- qa(fq, basename(fastqFile)) 
qa3 <- qa(fqSample, basename(fastqFile)) 
report(qa2,dest="/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/report001")

indexBam(reads_fn)
@
X11 error in report

Reweight each read.

\subsection{Sample short reads}

\subsection{Correct bias in short reads}

<<eval=FALSE,echo=TRUE>>=
library(seqbias) 
library(Rsamtools) 
ref_fn <- "/v4scratch/sc2265/rnaseq/output/ua159/1/data/NC_004350.fna"
ref_f <- FaFile( ref_fn ) 
open.FaFile( ref_f )
reads_fn <- "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.sorted.bam"
indexBam(reads_fn)
ref_seqs <- scanFaIndex( ref_f )

ref_seq <- getSeq(ref_f)
I.all <- GRanges(seqnames=Rle(c("NC_004350.1"),c(2)),
                 ranges=IRanges(c(1,1),width=rep(width(ref_seq),2)),
                 strand=Rle(strand(c("+","-")),c(1,1)))
seqlengths(I.all) <- c(width(ref_seq))

I <- random.intervals( ref_seqs, n = 5, m = 100000 )
seqs <- scanFa( ref_f, I )
neg_idx <- as.logical( I@strand == '-' ) 
seqs[neg_idx] <- reverseComplement( seqs[neg_idx] )
counts <- count.reads( reads_fn, I, binary = T )
freqs <- kmer.freq( seqs, counts )
sb <- seqbias.fit( ref_fn, reads_fn, L = 5, R = 20 )
bias <- seqbias.predict( sb, I )
bias <- seqbias.predict( sb, I.all )
bias <- seqbias.predict( sb5, I.all )

bias[[2]][99996]
seqbias.save(sb,"/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.yml")
sb <- seqbias.load( ref_fn, "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.yml")
sb5 <- seqbias.load( ref_fn, "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001-500000.yml")
sb5 <- seqbias.fit( ref_fn, reads_fn, n=5e5, L = 5, R = 20 )

N <- width(ref_seq) # genome length
cnt <- count.reads(reads_fn, I.all, sb = sb5)
cnt <- count.reads(reads_fn, GRanges("seq1", IRanges(start = 1, width = N)), sb = sb5)

readlen <- 100
cov <- coverage(GRanges("NC_004350.1", IRanges(start = 1:N, width = readlen)), weight = cnt[[1]])
@

\subsection{Load biasseq}

<<eval=FALSE,echo=TRUE>>=
library(seqbias) 
library(Rsamtools) 
ref_fn <- "/v4scratch/sc2265/rnaseq/output/ua159/1/data/NC_004350.fna"
ref_f <- FaFile( ref_fn ) 
open.FaFile( ref_f )
reads_fn <- "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.sorted.bam"
indexBam(reads_fn)
ref_seqs <- scanFaIndex( ref_f )

ref_seq <- getSeq(ref_f)
I.all <- GRanges(seqnames=Rle(c("NC_004350.1"),c(2)),
                 ranges=IRanges(c(1,1),width=rep(width(ref_seq),2)),
                 strand=Rle(strand(c("+","-")),c(1,1)))
seqlengths(I.all) <- c(width(ref_seq))

sb5 <- seqbias.load( ref_fn, "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001-500000.yml")
bias <- seqbias.predict( sb5, I.all )
@

\subsection{Compute the coverage of short reads}

<<eval=FALSE,echo=TRUE>>=
bamFile <- "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.sorted.bam"
aln <- readGappedAlignments(bamFile)
aln <- as(aln, "GRanges")
coverage(aln)
cov <- coverage(aln, weight = cnt[[1]])
bias[[2]][99996]

cvg <- rep(0,width(ref_seq))
s1 <- start(aln)
s2 <- end(aln)
s3 <- strand(aln)
for (i in 1:length(aln)) {
  if (i %% 100000 == 0) {
    print(i)
  }
  if (runValue(s3[i]) == "+") {
    cvg[s1[i]:s2[i]] <- cvg[s1[i]:s2[i]] + 1/bias[[1]][s1[i]]
  } else {
    cvg[s1[i]:s2[i]] <- cvg[s1[i]:s2[i]] + 1/bias[[2]][s2[i]]
  }
}
cvgFile <- "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.cvg"
save(cvg,file=cvgFile)
@

\subsection{Export coverage to a wiggle file}

\section{BWA}

\subsection{Count non-unique reads}

bamFile=/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.sorted.bam
samtools view \$bamFile | grep -v XT:A:U | wc -l

The automatic annotation of bacterial genomes
Emily J. Richardson and Mick Watson

rtracklayer
export()

\section{Parallel running}

\subsection{Multicore}

<<eval=FALSE,echo=TRUE>>=
install.packages(c("multicore", "snow", "foreach", "doMC", "doSNOW"))
install.packages(c("Rmpi", "doMPI"))
library(multicore) 
multicore:::detectCores()
@

\section{Apply functions}

\subsection{lapply}

/home/fs01/sc2265/Downloads/r-devel/b/bin/Rscript 1.R 8 1
<<eval=FALSE,echo=TRUE>>=
library(seqbias)
library(Rsamtools) 
library(ShortRead)

args <- commandArgs(trailingOnly = TRUE)
if (length(args) != 2)
{
  cat ("Rscript 1.R 8 1\n")
  quit("no")
}

bamFile <- "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.sorted.bam"
aln <- readGappedAlignments(bamFile)
aln <- as(aln, "GRanges")

n <- as.integer(args[1]) + 1L
a <- as.integer(seq(1L,length(aln),length.out=n))
a[9] <- a[9] + 1
s.1 <- a[as.integer(args[2])]
s.2 <- a[as.integer(args[2]) + 1] - 1
quit("no")

ref_fn <- "/v4scratch/sc2265/rnaseq/output/ua159/1/data/NC_004350.fna"
ref_f <- FaFile( ref_fn ) 
open.FaFile( ref_f )
ref_seqs <- scanFaIndex( ref_f )
ref_seq <- getSeq(ref_f)
I.all <- GRanges(seqnames=Rle(c("NC_004350.1"),c(2)),
                 ranges=IRanges(c(1,1),width=rep(width(ref_seq),2)),
                 strand=Rle(strand(c("+","-")),c(1,1)))
seqlengths(I.all) <- c(width(ref_seq))

sb5 <- seqbias.load( ref_fn, "/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001-500000.yml")
bias <- seqbias.predict( sb5, I.all )

s1 <- start(aln)
s2 <- end(aln)
s3 <- strand(aln)

# foreach (i=1:length(aln)) %do% {
cvg <- rep(0,width(ref_seq))
# for (i in 1:length(aln)) {
for (i in s.1:s.2) {
  if (runValue(s3[i]) == "+") {
    cvg[s1[i]:s2[i]] <- cvg[s1[i]:s2[i]] + 1/bias[[1]][s1[i]]
  } else {
    cvg[s1[i]:s2[i]] <- cvg[s1[i]:s2[i]] + 1/bias[[2]][s2[i]]
  }
}
cvgFile <- paste("/v4scratch/sc2265/rnaseq/output/ua159/1/bwa/FASTQ001.cvg",args[2],sep=".")
save(cvg,file=cvgFile)
close.FaFile( ref_f )
@

track type=wiggle_0 name="RNA-seq" description="RNA-seq 1" visibility=full
autoScale=on color=0,200,100 priority=30
fixedStep chrom=chr1 start=1 step=1
1
...
1

\subsection{doMC}

<<eval=FALSE,echo=TRUE>>=
library(doMC)
registerDoMC()
a <- foreach(i=1:1000000) %dopar% sqrt(i)
getDoParWorkers()
getDoParName()
getDoParVersion()
options(cores=8)
@

<<eval=FALSE,echo=TRUE>>=
cvg <- foreach (i=1:length(aln),.combine="+") %dopar% {
  if (runValue(s3[i]) == "+") {
    cvg[s1[i]:s2[i]] <- cvg[s1[i]:s2[i]] + 1/bias[[1]][s1[i]]
  } else {
    cvg[s1[i]:s2[i]] <- cvg[s1[i]:s2[i]] + 1/bias[[2]][s2[i]]
  }
  cvg
}
@

\section{Session Info}
<<session>>=
sessionInfo()
@

\bibliographystyle{unsrt}
\bibliography{siepel-mutans}

\end{document}

### 2012-09-12 sh/batch3.sh

@@ -19,6 +19,7 @@
 
 REPETITION=1
 function batch3 {
+  PS3="Choose the species for $FUNCNAME: "
   select SPECIES in ${SPECIESS[@]}; do 
   if [ "$SPECIES" == "" ];  then
     echo -e "You need to enter something\n"
@@ -30,6 +31,7 @@ function batch3 {
     batch3-speciesfile 
     batch3-push-data
     batch3-get-data
+    batch3-clean-data
     batch3-make-job
     batch3-run-qcalignde
     create-index
@@ -90,6 +92,9 @@ function batch3-speciesfile {
   REFGENOMETXDB=$(grep ^REFGENOMETXDB\: $SPECIESFILE | cut -d":" -f2)  
   REFGENOMETXDB=$ROOTANALYSISDIR/$REFGENOMETXDB
   GENOMETXDB=$(basename $REFGENOMETXDB)
+  REFGENOMERRNATXDB=$(grep ^REFGENOMERRNATXDB\: $SPECIESFILE | cut -d":" -f2)  
+  REFGENOMERRNATXDB=$ROOTANALYSISDIR/$REFGENOMERRNATXDB
+  GENOMERRNATXDB=$(basename $REFGENOMERRNATXDB)
   QCALIGNDEWALLTIME=$(grep ^QCALIGNDEWALLTIME\: $SPECIESFILE | cut -d":" -f2)
   MINMAPQ=$(grep ^MINMAPQ\: $SPECIESFILE | cut -d":" -f2)
   MINTRIMQ=$(grep ^MINTRIMQ\: $SPECIESFILE | cut -d":" -f2)
@@ -105,6 +110,7 @@ ssh -x $CAC_USERHOST mkdir -p $RDATADIR
 ssh -x $CAC_USERHOST mkdir -p $RRUNANALYSIS
 scp $REFGENOMEFASTA $CAC_USERHOST:$RDATADIR
 scp $REFGENOMETXDB $CAC_USERHOST:$RDATADIR
+scp $REFGENOMERRNATXDB $CAC_USERHOST:$RDATADIR
 for g in $FASTQFILES; do
   FASTQNUM=FASTQ\$(printf "%03d" \$g)
   FASTQFILE=$ROOTANALYSISDIR/\$(grep ^\$FASTQNUM\: $SPECIESFILE | cut -d":" -f2)
[…]

### 2012-09-12 sh/choose-species.sh

@@ -28,11 +28,11 @@ function choose-species {
       read REPETITION
       global-variable $SPECIES $REPETITION
 
-      mkdir -p $BASERUNANALYSIS
+#      mkdir -p $BASERUNANALYSIS
       mkdir -p $NUMBERDIR
       mkdir -p $DATADIR
       mkdir -p $BWADIR
-      mkdir -p $TRANSCRIPTDIR
+#      mkdir -p $TRANSCRIPTDIR
       mkdir -p $RUNANALYSIS
       break
     fi

### 2012-09-12 sh/make-soft-links.sh

@@ -7,6 +7,8 @@ LOCALBASEDIR=$HOME/Documents/Projects/RNASeq-Analysis
 STORAGEBASEDIR=$HOME/data/RNASeq-Analysis
 # exit 
 
+mkdir -p $LOCALBASEDIR
+
 BASEDIR=`pwd`
 echo Creating soft links to directories in the source code base directory...
 for i in doc pl sh src run; do
@@ -26,3 +28,4 @@ echo Edit the conf file available at $LOCALBASEDIR/conf/README
 # mkdir $LOCALBASEDIR/routput
 # mkdir $LOCALBASEDIR/conf
 # cp $BASEDIR/conf/README.local $LOCALBASEDIR/conf/README
+

### 2012-10-01 package-source/mss-lin-supp.Rnw
I've sent a version of the manuscript to Lin.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Load smutans package and data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<eval=FALSE,echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(goseq)
library(GenomicRanges)
library(IRanges)
library(plotrix)
library(xtable)
library(RColorBrewer)
library(lattice)
library(rtracklayer)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txGrCombined.RData")
load("smutans/data/smutansData.txAnnotation.RData")
load("smutans/data/smutansData.txRNAz.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutans.keg.genes.RData")
load("smutans/data/smutans.keg.desc.RData")
load("smutans/data/smutansGenes.RData")
@
<<load,echo=FALSE,results=hide>>=
# Run with smutans R package version 0.1.5 installed with R 2.14.2
# Remove eval=FALSE when the R package is installed
library(smutans)
# library(lattice)
# library(rtracklayer)
data(smutansData.tx)
data(smutansData.txGenes)
data(smutansData.txGrCombined)
data(smutansData.txAnnotation)
data(smutansData.txRNAz)
data(smutansData.txPileup)
data(smutans.feature.genes)
data(smutans.genes.criteria)
data(smutans.go.genes)
data(smutans.cat.desc)
data(smutans.keg.genes)
data(smutans.keg.desc)
data(smutansGenes)
data(smutansGenes2)
@
<<load,echo=FALSE,results=hide>>=
load(file="mss-lin.RData")
@
\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color} 
\usepackage{setspace} 
\usepackage{rotating}
\usepackage{xr}
\usepackage{url}
\usepackage{longtable}
\usepackage{lscape}
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
\usepackage[round,comma,numbers]{natbib}

\def\@cite#1#2{(#1\if@tempswa , #2\fi)}                                         
\def\@biblabel#1{}

\externaldocument{mss-lin}

% Text layout
\oddsidemargin 0in
\evensidemargin 0in
\topmargin -.5in
\textwidth 6.5in
\textheight 9in
\parskip 0.25in

\makeatletter
\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=6,height=6}
\newcommand{\Rpackage}[1]{R package \textit{#1}}
\newcommand{\program}[1]{{\texttt{#1}}}
\newcommand{\fixme}[1]{{\textbf{Fixme:} \textit{\textcolor{red}{#1}}}}
\newcommand{\change}[1]{\textit{\textcolor{blue}{#1}}}
\newcommand{\comment}[1]{}
\renewcommand{\@biblabel}[1]{\quad#1.}
\renewcommand{\thefootnote}{\textit{\alph{footnote}}}
\makeatother

\pagestyle{myheadings}
\markright{RNA-seq analysis in S. mutans}

% Leave date blank
\date{}

\begin{document}

\setcounter{secnumdepth}{-1}

\noindent \begin{LARGE}{\bf CcpA and Catabolite Repression Explored\\
by RNA-Seq in {\em Streptococcus mutans}}\end{LARGE}\\[2ex]
{\bf 
Lin Zeng, Sang Chul Choi, Charles D.\ Danko, \\
Adam Siepel, Michael J.\ Stanhope, and Robert A.\ Burne}%\\[0.5ex]

\listoftables

\listoffigures

\doublespacing

\clearpage{}\setcounter{figure}{0}
\setcounter{table}{0}

\makeatletter 
\renewcommand{\thefigure}{S\@arabic\c@figure} 
\renewcommand{\thetable}{S\@arabic\c@table} 
\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\tablename}{Supplementary Table}
\makeatother

\clearpage{}
\section*{SUPPLEMENTARY TABLES}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table S1. small RNAs neighboring context
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{landscape}
<<tabNeighborSmallRNA,echo=FALSE,results=tex>>=
print( xtable.tabNeighborSmallRNA,
       tabular.environment='longtable',
       table.placement="!h",
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE,
       sanitize.text.function = function(x){x}) 
@
\end{landscape}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table S2. GO UA159/TW1 grown in glucose
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
genes in the comparison UA159 and TW1 grown in glucose.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq2 ( glucoseOnly, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc, 
                   goqval=1e-3)
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$ We used a multiple-testing adjusted $p$-value of $10^{-3}$ to
determine differentially expressed genes.  For each category having at least ten
genes a variation of Fisher's exact test was performed, and another
multiple-testing adjusted $p$-value, or $q$-value was obtained. The listed
categories were chosen at the cutoff value of $10^{-3}$.
\\ 
\end{flushleft}
\label{tabGoseqGlucoseOnly}
\end{table}
%-----------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table S3. KEGG UA159/TW1 grown in glucose
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf KEGG enrichments for differentially expressed
genes in the comparison UA159 and TW1 grown in glucose.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq2 ( glucoseOnly, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.keg.genes[c(4,3)],
                   cat.desc=smutans.keg.desc,
                   goqval=1e-3)
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$ We used a multiple-testing adjusted $p$-value of $10^{-3}$ to
determine differentially expressed genes.  For each category having at least ten
genes a variation of Fisher's exact test was performed, and another
multiple-testing adjusted $p$-value, or $q$-value was obtained. The listed
categories were chosen at the cutoff value of $10^{-3}$.
\end{flushleft}
\label{tabKeggGlucoseOnly}
\end{table}
%-----------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table S4. GO Glucose/Galactose with UA159
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
genes in the comparison of glucose-grown and galactose-grown UA159 wild-type
cells.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq2 ( ua159only, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc, 
                   goqval=1e-3)
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$ We used a multiple-testing adjusted $p$-value of $10^{-3}$ to
determine differentially expressed genes.  For each category having at least ten
genes a variation of Fisher's exact test was performed, and another
multiple-testing adjusted $p$-value, or $q$-value was obtained. The listed
categories were chosen at the cutoff value of $10^{-3}$.
\end{flushleft}
\label{tabGoseqUA159Only}
\end{table}
%-----------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table S5. KEGG Glucose/Galactose with UA159
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf KEGG enrichments for differentially expressed
genes in the comparison of glucose-grown and galactose-grown UA159 wild-type
cells.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq2 ( ua159only, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.keg.genes[c(4,3)],
                   cat.desc=smutans.keg.desc,
                   goqval=1e-3)
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$ We used a multiple-testing adjusted $p$-value of $10^{-3}$ to
determine differentially expressed genes.  For each category having at least ten
genes a variation of Fisher's exact test was performed, and another
multiple-testing adjusted $p$-value, or $q$-value was obtained. The listed
categories were chosen at the cutoff value of $10^{-3}$.
\end{flushleft}
\label{tabKeggUA159Only}
\end{table}
%-----------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table S6. GO Glucose/Galactose with TW1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
genes in the comparison of glucose-grown and galactose-grown TW1 mutant-type
cells.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq2 ( tw1only, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc, 
                   goqval=1e-3)
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$ We used a multiple-testing adjusted $p$-value of $10^{-3}$ to
determine differentially expressed genes.  For each category having at least ten
genes a variation of Fisher's exact test was performed, and another
multiple-testing adjusted $p$-value, or $q$-value was obtained. The listed
categories were chosen at the cutoff value of $10^{-3}$.
\end{flushleft}
\label{tabGoseqTW1Only}
\end{table}
%-----------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table S7. KEGG Glucose/Galactose with TW1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf KEGG enrichments for differentially expressed
genes in the comparison of glucose-grown and galactose-grown TW1 mutant-type
cells.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq2 ( tw1only, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.keg.genes[c(4,3)],
                   cat.desc=smutans.keg.desc,
                   goqval=1e-3)
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$ We used a multiple-testing adjusted $p$-value of $10^{-3}$ to
determine differentially expressed genes.  For each category having at least ten
genes a variation of Fisher's exact test was performed, and another
multiple-testing adjusted $p$-value, or $q$-value was obtained. The listed
categories were chosen at the cutoff value of $10^{-3}$.
\end{flushleft}
\label{tabKeggTW1Only}
\end{table}
%-----------------------------------------------------------------------------

%%-----------------------------------------------------------------------------
%\begin{table}[!h]
%\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
%genes in the comparison UA159 and TW1 grown in galactose.}}
%\vspace{1ex}
%\noindent \begin{centering}
%\begin{tabular}{cl}
%\hline 
%$q^a$ & Description \\
%\hline 
%%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq2 ( galactoseOnly, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc, 
                   goqval=1e-3)
@
%%-----------------------------------------------------------------------------
%\hline 
%\end{tabular}
%\par\end{centering}
%\begin{flushleft}
%$^a$ We used a multiple-testing adjusted $p$-value of $10^{-3}$ to
%determine differentially expressed genes.  For each category having at least ten
%genes a variation of Fisher's exact test was performed, and another
%multiple-testing adjusted $p$-value, or $q$-value was obtained. The listed
%categories were chosen at the cutoff value of $10^{-3}$.
%\end{flushleft}
%\label{tab:goseqGalactoseOnly}
%\end{table}
%%-----------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table S8. KEGG UA159/TW1 grown in galactose
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-----------------------------------------------------------------------------
%\begin{table}[!h]
%\caption{{\bf KEGG enrichments for differentially expressed
%genes in the comparison UA159 and TW1 grown in galactose.}}
%\vspace{1ex}
%\noindent \begin{centering}
%\begin{tabular}{cl}
%\hline 
%$q^a$ & Description \\
%\hline 
%%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq2 ( galactoseOnly, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.keg.genes[c(4,3)],
                   cat.desc=smutans.keg.desc,
                   goqval=1e-3)
@
%%-----------------------------------------------------------------------------
%\hline 
%\end{tabular}
%\par\end{centering}
%\begin{flushleft}
%$^a$False discovery rate estimated by the
%Benjamini-Hochberg method.  All categories having at least ten genes and
%$q\leq 0.05$ are displayed.\\ 
%\end{flushleft}
%\label{tab:kegseqGalactoseOnly}
%\end{table}
%%-----------------------------------------------------------------------------

% END OF TABLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage{}
\section*{SUPPLEMENTARY FIGURES}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure S1. Expressed/Not-expressed Gene length distribution
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<figLength,fig=TRUE,echo=FALSE>>=
l <- list(expressed=width(smutansData.txGenes[matchMatrix(matchGenesTx)[,1]]), 
          notexpressed=width(smutansData.txGenes[-matchMatrix(matchGenesTx)[,1]]))
bin <- c(as.integer(seq.int(0, 3000, length.out=15)), 8500)
multhist(l,breaks=bin,ylab="Frequency",xlab="Length of transcript in base pairs")
@
\begin{figure}[!h]
\centering
\includegraphics[width=.9\textwidth]{mss-lin-supp-figLength}
\caption{{\bf Length distribution for expressed and not-expressed genes.}
Darker gray
bars represent frequency of expressed genes, and lighter gray ones that of
not-expressed genes. The last bin sums from 3000 bp to 8500 bp.
}
\label{figLength}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure S2. Transcript length distribution with/without annotated genes
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<figTranscriptLength,fig=TRUE,echo=FALSE>>=
l <- list(coding=width(grTxWGenes), 
          noncoding=width(grTxWOGenes))
bin <- c(as.integer(seq.int(0, 10000, length.out=15)), 30000)
multhist(l,breaks=bin,ylab="Frequency",xlab="Length of transcript in base pairs")
@
\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{mss-lin-supp-figTranscriptLength}
\caption{{\bf Length distribution for expressed transcripts with annotated genes
and unannotated genes.}
Darker gray bars represent frequency of expressed transcripts with annotated
genes, and lighter gray ones that of without annotated genes that can be
potential non-coding regions. The last bin sums from 10000 bp to 30000 bp.
}
\label{figTranscriptLength}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure S3. UTR length distribution of transcripts
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<figUTRLength,fig=TRUE,echo=FALSE>>=
l <- list(five=length5UTR, three=length3UTR)
bin <- c(as.integer(seq.int(0, 300, length.out=19)), 6500)
multhist(l,breaks=bin,ylab="Frequency",xlab="Length of UTRs in base pairs")
@
\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{mss-lin-supp-figUTRLength}
\caption{{\bf Length distribution for 5' and 3' UTR.}
Darker gray bars represent frequency of 5' UTR lengths, and lighter gray 3' UTR
lengths. The last bin sums from 300 bp to 6500 bp.
}
\label{figUTRLength}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure S4. RNA secondary structure
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/PredSmallRNA-204}
\caption{{\bf MFE structures drawing encoding positional entropy for
differentially expressed regions.}
}
\label{figRNA204}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure S5. Pie chart UA159/TW1 grown in Glucose
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage{}
<<figPieGlucoseOnly,fig=TRUE,echo=FALSE,width=6,height=8>>=
par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
y <- tableGlucoseOnly[,"Functional Class"]
z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
z <- table(as.factor(sort(unlist(z))))

slices <- as.vector(z)
lbls <- names(z)
pct <- round(slices/sum(slices)*100)
c1 <- brewer.pal(n=9,name="Set1")
pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
@
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{mss-lin-supp-figPieGlucoseOnly}
\caption{{\bf Pie chart distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type and TW1 mutant-type grown in glucose.}
Labels for functional categories are colored with nine different colors. The
first label with red color, 
``Biosynthesis of cofactors, prosthetic groups, and carriers; Thiamine'', starts
at the top, and next labels read to the clockwise. The numbers circling the pie
chart is in per cent.
}
\label{figPieGlucoseOnly}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure S6. Pie chart Glucose/Galactose in UA159
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage{}
<<figPieUA159Only,fig=TRUE,echo=FALSE,width=6,height=10>>=
par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
y <- tableUA159Only[,"Functional Class"]
z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
z <- table(as.factor(sort(unlist(z))))

slices <- as.vector(z)
lbls <- names(z)
pct <- round(slices/sum(slices)*100)
c1 <- brewer.pal(n=9,name="Set1")
pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)

legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
@
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{mss-lin-supp-figPieUA159Only}
\caption{{\bf Pie chart distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type grown in glucose and galactose.}
Labels for functional categories are colored with nine different colors. The
first label with red color, 
``Amino acid biosynthesis; Glutamate family'', starts
at the top, and next labels read to the clockwise. The numbers circling the pie
chart is in per cent.
}
\label{figPieUA159Only}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure S7. Pie chart Glucose/Galactose in TW1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage{}
<<figPieTW1Only,fig=TRUE,echo=FALSE,width=6,height=8>>=
par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
trim <- function (x) gsub("^\\s+|\\s+$", "", x)
y <- tableTW1Only[,"Functional Class"]
z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
z <- table(as.factor(sort(unlist(z))))

slices <- as.vector(z)
lbls <- names(z)
pct <- round(slices/sum(slices)*100)
c1 <- brewer.pal(n=9,name="Set1")
pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
@
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{mss-lin-supp-figPieTW1Only}
\caption{{\bf Pie chart distribution of functional classes of differntially expressed genes
in comparing TW1 mutant-type grown in glucose and galactose.}
Labels for functional categories are colored with nine different colors. The
first label with red color, 
``Amino acid biosynthesis; Histidine family'', starts
at the top, and next labels read to the clockwise. The numbers circling the pie
chart is in per cent.
}
\label{figPieTW1Only}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure S8. Dot chart Glucose/Galactose in TW1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<figDotplotTW1Only,fig=TRUE,echo=FALSE,width=12,height=6>>=
#par(mar=c(4.1, 3.1, 0.1, 0.1))
par(mai=c(0, 0, 0, 0))
y <- tableTW1Only[,"Functional Class"]
# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
# z <- table(as.factor(sort(unlist(z))))
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableTW1Only[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotplot.
dotplot(xm[,"Functional Class"]~xm[,2],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{mss-lin-supp-figDotplotTW1Only}
\caption{{\bf Dot chart distribution of functional classes of differntially expressed genes
in comparing TW1 wild-type grown in glucose and galactose.}
}
\label{figDotplotTW1Only}
\end{sidewaysfigure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure S9. Pie chart UA159/TW1 grown in Galactose
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage{}
<<figPieGalactoseOnly,fig=TRUE,echo=FALSE,width=6,height=8>>=
par(mai=c(0,0,0,0),fig=c(0,1,0,1),mfrow=c(2,1))
y <- tableGalactoseOnly[,"Functional Class"]
z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
z <- table(as.factor(sort(unlist(z))))

slices <- as.vector(z)
lbls <- names(z)
pct <- round(slices/sum(slices)*100)
c1 <- brewer.pal(n=9,name="Set1")
pie(z,labels=pct,main="",col=c1,clockwise=TRUE)
plot(0,0,axes=FALSE,frame.plot=FALSE,xlab="",ylab="",cex=0)
legend(-0.8,1,names(z),bty='n',cex=0.7,fill=c1)
@
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{mss-lin-supp-figPieGalactoseOnly}
\caption{{\bf Pie chart distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type and TW1 mutant-type grown in galactose.}
Labels for functional categories are colored with nine different colors. The
first label with red color, 
``Amino acid biosynthesis; Glutamate family'', starts
at the top, and next labels read to the clockwise. The numbers circling the pie
chart is in per cent.
}
\label{figPieGalactoseOnly}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure S10. Dot chart UA159/TW1 grown in Galactose
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<figDotplotGalactoseOnly,fig=TRUE,echo=FALSE,width=12,height=6>>=
par(mai=c(0, 0, 0, 0))
y <- tableGalactoseOnly[,"Functional Class"]
# z <- lapply(y, function(x) trim(unlist(strsplit(as.character(x),"\\|\\|"))))
# z <- table(as.factor(sort(unlist(z))))
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableGalactoseOnly[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotplot.
dotplot(xm[,"Functional Class"]~xm[,2],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{mss-lin-supp-figDotplotGalactoseOnly}
\caption{{\bf Dot chart distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type and TW1 mutant-type grown in galactose.}
}
\label{figDotplotGalactoseOnly}
\end{sidewaysfigure}

\end{document}

%\clearpage{}
%\begin{figure}
%\centering
%\includegraphics[width=\textwidth]{figures/ucsc}
%\caption{{\bf UCSC genome browser tracks of RNA-seq data.}
%Tracks are shown from the top: known genes, pileup value tracks for UA159 grown
%in glucose, those for TW1 in glucose, and the five segments predicted based on
%RNA-seq data and known genes. Known genes with forward strand are colored red,
%and those of reverse strand blue. The most top of the five predicted segments
%are also colored in the same way. 
%}
%\label{figUCSC}
%\end{figure}

### 2012-10-01 package-source/mss-lin.Rnw
I've sent a version of the manuscript to Lin.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Load smutans package and data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<eval=FALSE,echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(goseq)
library(GenomicRanges)
library(IRanges)
library(plotrix)
library(xtable)
library(RColorBrewer)
library(lattice)
library(rtracklayer)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txGrCombined.RData")
load("smutans/data/smutansData.txAnnotation.RData")
load("smutans/data/smutansData.txRNAz.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutans.keg.genes.RData")
load("smutans/data/smutans.keg.desc.RData")
load("smutans/data/smutansGenes.RData")
@
<<load,echo=FALSE,results=hide>>=
# Run with smutans R package version 0.1.5 installed with R 2.14.2
# Remove eval=FALSE when the R package is installed
# R package version 0.2.1 is a new branch from 0.1.5.
# Have a separate branch from 0.1.9; say, 0.3.1 for further development.
library(smutans)
#library(lattice)
#library(rtracklayer)
data(smutansData.tx)
data(smutansData.txGenes)
data(smutansData.txGrCombined)
data(smutansData.txAnnotation)
data(smutansData.txRNAz)
data(smutansData.txPileup)
data(smutans.feature.genes)
data(smutans.genes.criteria)
data(smutans.go.genes)
data(smutans.cat.desc)
data(smutans.keg.genes)
data(smutans.keg.desc)
data(smutansGenes)
data(smutansGenes2)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Functional association and gene description
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<global,echo=FALSE,results=hide>>=
inDir <- system.file("extdata", package="smutans", mustWork=TRUE)
#inDir <- "extdata"

trim <- function (x) gsub("^\\s+|\\s+$", "", x)

# geneDescription <- read.delim("feature-genome.out-geneonly.txt",head=FALSE,sep="\t")
# Note that feature-genome.out-geneonly.txt contains annotations.
# feature-genome.out-geneonly does not.
geneDescription <- read.delim(file.path(inDir,"feature-genome.out-geneonly.txt"),head=FALSE,sep="\t")
geneDescription <- geneDescription[,c(4,7)]
colnames(geneDescription) <- c("Gene ID","Gene Description")

# functionalDescription <- read.csv("feature-genome.out-geneonly-oralgen-functional.txt",head=FALSE)
functionalDescription <- read.csv(file.path(inDir,"feature-genome.out-geneonly-oralgen-functional.txt"),head=FALSE)
functionalDescription <- functionalDescription[,c(2,12)]
colnames(functionalDescription) <- c("Gene ID","Functional Class")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Transcript analyses R scripts
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<transcript,echo=FALSE,result=hide>>=
notExpressedTx <- smutansData.txGrCombined[elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5]
grTx <- smutansData.txGrCombined[!(elementMetadata(smutansData.txGrCombined)$zero > 0.1 & elementMetadata(smutansData.txGrCombined)$pileup < 5)]
# We query genes and annotations (genes and RNAs) againt tx's.
matchGenesTx <- findOverlaps(smutansData.txGenes,grTx)
matchAnnotationTx <- findOverlaps(smutansData.txAnnotation,grTx)
# I use gaps with IRanges because gaps with GenomicRanges does not work as expected.
grNoAnnotation <- gaps(ranges(smutansData.txAnnotation))
grNoncoding <- gaps(ranges(smutansData.txGenes))
grRNAz <- ranges(smutansData.txRNAz)
# 
matchUnannotatedRNA <- findOverlaps(grRNAz,grNoAnnotation,type="within")
txRNAz.unannotated <- grRNAz[matchMatrix(matchUnannotatedRNA)[,1]]

# A bug in GenomicRanges: the last segment is the while genome. I removed it
# from the noncoding ranges.
# matchNoncoding <- findOverlaps(grNoncoding,grTx)

# Expressed transcripts without annotated genes
grTxWGenes <- grTx[unique(matchMatrix(matchGenesTx)[,2])]
grTxWOGenes <- grTx[-unique(matchMatrix(matchGenesTx)[,2])]
grTxWOAnnotation <- grTx[-unique(matchMatrix(matchAnnotationTx)[,2])]

# Find transcripts with genes with conflicted strands
grGenesPlus <- smutansData.txGenes[strand(smutansData.txGenes)=='+']
grGenesMnus <- smutansData.txGenes[strand(smutansData.txGenes)=='-']
#mtchPlus <- findOverlaps(grGenesPlus,grTxWGenes)
#mtchMnus <- findOverlaps(grGenesMnus,grTxWGenes)
#txWithConflictedGenes <- intersect(matchMatrix(mtchPlus)[,2],matchMatrix(mtchMnus)[,2])

# Tx with genes strands
grTx2 <- grTxWGenes[!strand(grTxWGenes)=='*']

# Find the number of genes in segments.
matchGenesTx2 <- findOverlaps(smutansData.txGenes,grTx2,type="within")
number.genes.tx <- runLength(Rle(matchMatrix(matchGenesTx2)[,"subject"]))

# Length distribution of UTR regions
# grTx2 <- grTxWGenes[-txWithConflictedGenes] 
# Lengths of 5' and 3' UTRs
mtchPlus <- findOverlaps(grGenesPlus,grTx2,type="within")
mtchMnus <- findOverlaps(grGenesMnus,grTx2,type="within")
length5UTR <- c()
length3UTR <- c()
for (i in unique(matchMatrix(mtchPlus)[,"subject"])) {
  x <- matchMatrix(mtchPlus)[matchMatrix(mtchPlus)[,"subject"]==i,"query"]
  y <- min(start(grGenesPlus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesPlus[x])) 
  length5UTR <- c(length5UTR,y)
  length3UTR <- c(length3UTR,z)
}
for (i in unique(matchMatrix(mtchMnus)[,"subject"])) {
  x <- matchMatrix(mtchMnus)[matchMatrix(mtchMnus)[,"subject"]==i,"query"]
  y <- min(start(grGenesMnus[x])) - start(grTx2[i])
  z <- end(grTx2[i]) - max(end(grGenesMnus[x])) 
  length5UTR <- c(length5UTR,z)
  length3UTR <- c(length3UTR,y)
}

# Correlation of expression levels of adjacent genes.
pileup.x <- lapply(ranges(smutansData.txGenes), function(x) seqselect(smutansData.txPileup, x))
pileup.y <- mapply(function(x,y) sum(x)/y, x=pileup.x, y=width(smutansData.txGenes))
elementMetadata(smutansData.txGenes)["score"] <- pileup.y
mtch3 <- findOverlaps(smutansData.txGenes,grTx2)
x <- rle(matchMatrix(mtch3)[,"subject"])
comp.exp.two.genes <- c()
for (i in x$values[x$lengths > 1]) {
  y <- combn(matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==i],2)
  for (j in 1:dim(y)[2]) {
    comp.exp.two.genes <- cbind(comp.exp.two.genes,as.matrix(elementMetadata(smutansData.txGenes)["score"][y[,j],]))
  }
}
# Correlation of expression levels of adjacent genes between transcripts.
comp.exp.two.genes.between.transcript <- c()
for (i in 1:length(x$values)) {
  if (i == length(x$values)) {
    break
  }
  y1 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i]]
  i2 <- i + 1
  y2 <- matchMatrix(mtch3)[,"query"][matchMatrix(mtch3)[,"subject"]==x$values[i2]]
  l <- as.matrix(do.call(expand.grid, list(y1,y2)))
  for (j in 1:dim(l)[1]) {
    comp.exp.two.genes.between.transcript <- cbind(comp.exp.two.genes.between.transcript,as.matrix(elementMetadata(smutansData.txGenes)["score"][l[j,],]))
  }
}
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Prediction of small RNAs and targets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<rna,echo=FALSE,results=hide>>=
# Read Infernal's result
#x <- import.bed("/Users/goshng/Documents/Projects/RNASeq-Analysis/test/rfam/ua159-v1/cm.bed")
#irCm <- ranges(x)$chr1
irCm <- ranges(smutansData.txAnnotation[1:274])
# Read RNAz's result
ir1 <- gaps(ranges(smutansData.txAnnotation))
ir2 <- ranges(smutansData.txRNAz)
irRNAz <- ir2
mtchUnannotatedRNA <- findOverlaps(ir2,ir1,type="within")
txRNAz.unannotated <- ir2[matchMatrix(mtchUnannotatedRNA)[,1]]
# Read Pichon's result
# x <- import.bed("/Users/goshng/Documents/Projects/RNASeq-Analysis/test/pichon/ua159-v1/pichon.bed")
x <- import.bed(file.path(inDir,"pichon.bed"))
irPichon <- ranges(x)$chr1
mtchUnannotatedPichon <- findOverlaps(irPichon,ir1,type="within")
txPichon.unannotated <- irPichon[matchMatrix(mtchUnannotatedPichon)[,1]]
txPichon.reduced <- reduce(txPichon.unannotated)
elementReducedPichon <- findOverlaps(txPichon.reduced, txPichon.unannotated, select="first")
# Read sRNAscanner's result
# x <- import.bed("/Users/goshng/Documents/Projects/RNASeq-Analysis/test/sRNAscanner/ua159-v1/srnascanner.bed")
x <- import.bed(file.path(inDir,"srnascanner.bed"))
irScanner <- ranges(x)$chr1
mtchUnannotatedScanner <- findOverlaps(irScanner,ir1,type="within")
txScanner.unannotated <- irScanner[matchMatrix(mtchUnannotatedScanner)[,1]]
txScanner.reduced <- reduce(txScanner.unannotated)
elementReducedScanner <- findOverlaps(txScanner.reduced, txScanner.unannotated, select="first")
# Merge three results of RNAz, Pichon, and Scanner.
irUnion <- reduce(c(irRNAz,irPichon,irScanner))
names(irUnion) <- paste("PredSmallRNA",seq(length(irUnion)),sep="-")
# Read genes in ranges
irGene <- ranges(smutansData.txGenes)
# Find neighboring genes
left.right <- c()
for (k in seq(length(irUnion))) {
  x <- irUnion[k]
  irGeneUnion <- sort(c(irGene,x))
  i <- grep("PredSmallRNA",names(irGeneUnion))
  j.left <- NA
  j.left.distance <- NA
  j.left.description <- NA
  if (i > 1) {
    j.left <- names(irGeneUnion)[i-1]
    j.left.distance <- start(irGeneUnion)[i] - end(irGeneUnion)[i-1]
    j.left.description <- as.character(geneDescription[geneDescription[,1]==j.left,2])
    if (nchar(j.left.description) > 30) {
      j.left.description <- paste(substr(j.left.description,1,25),"...")
    }
  }
  j.right <- NA
  j.right.distance <- NA
  j.right.description <- NA
  if (i < length(names(irGeneUnion))) {
    j.right <- names(irGeneUnion)[i+1]
    j.right.distance <- start(irGeneUnion)[i+1] - end(irGeneUnion)[i]
    j.right.description <- as.character(geneDescription[geneDescription[,1]==j.right,2])
    if (nchar(j.right.description) > 30) {
      j.right.description <- paste(substr(j.right.description,1,25),"...")
    }
  }
  left.right <- rbind(left.right, c(j.left.description,j.left,j.left.distance,names(x),j.right.distance,j.right,j.right.description))
}
rm(irGeneUnion,k,x,i,j.left,j.left.distance,j.right,j.right.distance)
rm(j.left.description,j.right.description)
colnames(left.right) <- c("Description","Left gene","Distance","Predicted sRNA",
                          "Distance", "Right gene","Description")
xtable.tabNeighborSmallRNA <- xtable( left.right,
  display=c("s","s","s","d","s","d","s","s"),
  digits=c(0,0,0,1,0,1,0,0),
  align=c('l','r','r','l','r','l','l','l'),
  label='tabNeighborSmallRNA',
  caption='{\\bf Neighboring genes of predicted small RNAs}' 
)
################################################################################
# Rename predicted RNAs (determined by RNAz) using irUnion or the union set of
# predicted regions.
data(smutans.genes.criteria)
matchRNAz <- findOverlaps(irRNAz,irUnion)
x <- irUnion[matchMatrix(matchRNAz)[,2]]
smutans.genes.criteria <- smutans.genes.criteria[c(TRUE, colnames(smutans.genes.criteria)[-1] %in% names(irRNAz))]
colnames(smutans.genes.criteria)[-1] <- names(x)[match(colnames(smutans.genes.criteria)[-1],names(irRNAz))]

# DE small RNAs
library(smutans)

countsFile <- file.path(inDir,"count.nocds")
indexFile <- file.path(inDir,"count.nocds.txt.index")
################################################################################
# Comparison of UA159GLU and UA159GAL in small RNAs
sampleA <- "UA159GLU"
sampleB <- "UA159GAL"
csvFile = paste("srna-",sampleA,"-",sampleB,".csv",sep="")
clustFile = paste("srna-",sampleA,"-",sampleB,".pdf",sep="")
sampleTitle <- paste(sampleA, sampleB)
#####################
# Read in count data.
#cutoff.count=10
#conds <- scan(file=indexFile,what="character",quiet=TRUE)
#condition = c(sampleA, sampleB)
#subconds <- conds %in% condition
#conds <- conds[subconds]
#firstFactor <- condition
#countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)       
#rownames(countsTable) <- countsTable$gene
#countsTable <- countsTable[,-1]
#countsTable <- countsTable[,subconds]
#countsTable <- countsTable[(rowSums(countsTable) >= cutoff.count),]
#sGenes <- newCountDataSet( countsTable, data.frame(condition=conds,type="ua159") )
# Read in count data.
#####################
sGenes <-                                                                       
  readSmutans (countsFile=countsFile,
               indexFile =indexFile,
               condition = c(sampleA, sampleB),                                 
               firstFactorLabel = c("reference"),                               
               secondFactorLabel = c(sampleA, sampleB))
omz <- newSmutans( sGenes, title=sampleTitle )                                  
omz <- smutans.de2( omz, type="reference",
                    condA=sampleA, condB=sampleB )
# smutans.de2List( omz, csvFile )
x <- smutans.de2List( omz, qval=1e-1 )                                                 
x <- x[,c("id","log2FoldChange","padj")]
colnames(x) <- c("Predicted small RNA ID","$\\log_2$(gal/glc)", "$P$-value")
x <- x[abs(x[,2])>1,]
xtable.tabRNAua159Only <- xtable( x[order(x[,2],decreasing=TRUE),],
  display=c("s","s","f","E"),
  digits=c(0,0,1,0),
  align=c('l','l','r','r'),
  label='tabRNAua159Only',
  caption='{\\bf List of predicted small RNA regions differentially expressed in UA159 cells growing in TV-glucose (glc) and TV-galactose (gal).}'
)

################################################################################
# Comparison of TW1GLU and TW1GAL in small RNAs
sampleA <- "TW1GLU"
sampleB <- "TW1GAL"
csvFile = paste("srna-",sampleA,"-",sampleB,".csv",sep="")
clustFile = paste("srna-",sampleA,"-",sampleB,".pdf",sep="")
sampleTitle <- paste(sampleA, sampleB)
#####################
# Read in count data.
#cutoff.count=10
#conds <- scan(file=indexFile,what="character",quiet=TRUE)
#condition = c(sampleA, sampleB)
#subconds <- conds %in% condition
#conds <- conds[subconds]
#firstFactor <- condition
#countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)       
#rownames(countsTable) <- countsTable$gene
#countsTable <- countsTable[,-1]
#countsTable <- countsTable[,subconds]
#countsTable <- countsTable[(rowSums(countsTable) >= cutoff.count),]
#sGenes <- newCountDataSet( countsTable, data.frame(condition=conds,type="tw1") )
# Read in count data.
#####################
sGenes <-
  readSmutans (countsFile=countsFile,
               indexFile =indexFile,
               condition = c(sampleA, sampleB),
               firstFactorLabel = c("tw1"),
               secondFactorLabel = c(sampleA, sampleB))
omz <- newSmutans( sGenes, title=sampleTitle )                                  
omz <- smutans.de2( omz, type="tw1",
                    condA=sampleA, condB=sampleB )
# smutans.de2List( omz, csvFile )
x <- smutans.de2List( omz, qval=1e-1 )                                                 
x <- x[,c("id","log2FoldChange","padj")]
colnames(x) <- c("Predicted small RNA ID","$\\log_2$(gal/glc)", "$P$-value")
x <- x[abs(x[,2])>1,]
xtable.tabRNAtw1Only <- xtable( x[order(x[,2],decreasing=TRUE),],
  display=c("s","s","f","E"),
  digits=c(0,0,1,0),
  align=c('l','l','r','r'),
  label='tabRNAtw1Only',
  caption='{\\bf List of predicted small RNA regions differentially expressed in TW1 cells growing in TV-glucose (glc) and TV-galactose (gal).}' 
)

################################################################################
# Comparison of UA159GLU and TW1GLU in small RNAs
sampleA <- "UA159GLU"
sampleB <- "TW1GLU"
csvFile = paste("srna-",sampleA,"-",sampleB,".csv",sep="")
clustFile = paste("srna-",sampleA,"-",sampleB,".pdf",sep="")
sampleTitle <- paste(sampleA, sampleB)
#####################
# Read in count data.
#cutoff.count=10
#conds <- scan(file=indexFile,what="character",quiet=TRUE)
#condition = c(sampleA, sampleB)
#subconds <- conds %in% condition
#conds <- conds[subconds]
#firstFactor <- condition
#countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)       
#rownames(countsTable) <- countsTable$gene
#countsTable <- countsTable[,-1]
#countsTable <- countsTable[,subconds]
#countsTable <- countsTable[(rowSums(countsTable) >= cutoff.count),]
#sGenes <- newCountDataSet( countsTable, data.frame(condition=conds,type="glucose") )
# Read in count data.
#####################
sGenes <-
  readSmutans (countsFile=countsFile,
               indexFile =indexFile,
               condition = c(sampleA, sampleB),
               firstFactorLabel = c("glucose"),
               secondFactorLabel = c(sampleA, sampleB))
omz <- newSmutans( sGenes, title=sampleTitle )                                  
omz <- smutans.de2( omz, type="glucose",
                    condA=sampleA, condB=sampleB )
# smutans.de2List( omz, csvFile )
x <- smutans.de2List( omz, qval=1e-1 )                                                 
x <- x[,c("id","log2FoldChange","padj")]
colnames(x) <- c("Predicted small RNA ID","$\\log_2$(TW1/UA159)", "$P$-value")
x <- x[abs(x[,2])>1,]
xtable.tabRNAGlucoseOnly <- xtable( x[order(x[,2],decreasing=TRUE),],
  display=c("s","s","f","E"),
  digits=c(0,0,1,0),
  align=c('l','l','r','r'),
  label='tabRNAGlucoseOnly',
  caption='{\\bf List of predicted small RNA regions differentially expressed in UA159 and TW1 when growing in TV-glucose.}' 
)

################################################################################
# Comparison of UA159GAL and TW1GAL in small RNAs
sampleA <- "UA159GAL"
sampleB <- "TW1GAL"
csvFile = paste("srna-",sampleA,"-",sampleB,".csv",sep="")
clustFile = paste("srna-",sampleA,"-",sampleB,".pdf",sep="")
sampleTitle <- paste(sampleA, sampleB)
#####################
# Read in count data.
#cutoff.count=10
#conds <- scan(file=indexFile,what="character",quiet=TRUE)
#condition = c(sampleA, sampleB)
#subconds <- conds %in% condition
#conds <- conds[subconds]
#firstFactor <- condition
#countsTable <- read.delim (countsFile, header=TRUE, stringsAsFactors=TRUE)       
#rownames(countsTable) <- countsTable$gene
#countsTable <- countsTable[,-1]
#countsTable <- countsTable[,subconds]
#countsTable <- countsTable[(rowSums(countsTable) >= cutoff.count),]
#sGenes <- newCountDataSet( countsTable, data.frame(condition=conds,type="galactose") )
# Read in count data.
#####################
sGenes <-
  readSmutans (countsFile=countsFile,
               indexFile =indexFile,
               condition = c(sampleA, sampleB),
               firstFactorLabel = c("galactose"),
               secondFactorLabel = c(sampleA, sampleB))
omz <- newSmutans( sGenes, title=sampleTitle )                                  
omz <- smutans.de2( omz, type="galactose",
                    condA=sampleA, condB=sampleB )
# smutans.de2List( omz, csvFile )
x <- smutans.de2List( omz, qval=1e-1 )                                                 
x <- x[,c("id","log2FoldChange","padj")]
colnames(x) <- c("Predicted small RNA ID","$\\log_2$(TW1/UA159)", "$P$-value")
x <- x[abs(x[,2])>1,]
xtable.tabRNAGalactoseOnly <- xtable( x[order(x[,2],decreasing=TRUE),],
  display=c("s","s","f","E"),
  digits=c(0,0,1,0),
  align=c('l','l','r','r'),
  label='tabRNAGalactoseOnly',
  caption='{\\bf List of predicted small RNA regions differentially expressed in UA159 and TW1 when growing in TV-galactose.}' 
)
@
<<glucoseOnly,echo=FALSE,results=hide>>=
glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
                            condA="ua159", condB="tw1" )
x <- smutans.de2List( glucoseOnly, qval=1e-3 )
x <- x[,c("id","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","$\\log_2$(TW1/UA159)", "$P$-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x <- x[abs(x[,2])>1,]
xtable.tabGlucoseOnly <- xtable( x[order(x[,2],decreasing=TRUE),],
  display=c("s","s","f","E","s"),
  digits=c(0,0,1,0,0),
  align=c('l','l','r','r','p{3.5in}'),
  label='tabGlucoseOnly',
  caption='{\\bf List of genes differentially expressed in UA159 and TW1 when growing in TV-glucose.}' 
)
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
tableGlucoseOnly <- x
@
<<ua159only,echo=FALSE,results=tex>>=
ua159only <- newSmutans( smutansGenes, title="UA159 Only" )
ua159only <- smutans.de2( ua159only, type="ua159", 
                          condA="glucose", condB="galactose" )
#save(ua159only, file=file.path("ua159only.RData"))
#load("ua159only.RData")
x <- smutans.de2List( ua159only, qval=1e-3 )
x <- x[,c("id","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","$\\log_2$(gal/glc)", "$P$-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x <- x[abs(x[,2])>1,]
xtable.tabUA159Only <- xtable( x[order(x[,2],decreasing=TRUE),],
  display=c("s","s","f","E","s"),
  digits=c(0,0,1,0,0),
  align=c('l','l','r','r','p{3.5in}'),
  label='tabUA159Only',
  caption='{\\bf List of genes differentially expressed in UA159 cells growing in TV-glucose (glc) and TV-galactose (gal).}'
)
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
tableUA159Only <- x
@
<<tw1only,echo=FALSE,results=tex>>=
tw1only <- newSmutans( smutansGenes, title="TW1 Only" )
tw1only <- smutans.de2( tw1only, type="tw1", 
                          condA="glucose", condB="galactose" )
#save(tw1only, file=file.path("tw1only.RData"))
#load("tw1only.RData")
x <- smutans.de2List( tw1only, qval=1e-3 )
x <- x[,c("id","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","$\\log_2$(gal/glc)", "$P$-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x <- x[abs(x[,2])>1,]
xtable.tabTW1Only <- xtable( x[order(x[,2],decreasing=TRUE),],
  display=c("s","s","f","E","s"),
  digits=c(0,0,1,0,0),
  align=c('l','l','r','r','p{3.5in}'),
  label='tabTW1Only',
  caption='{\\bf List of genes differentially expressed in TW1 cells growing in TV-glucose (glc) and TV-galactose (gal).}' 
)
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
tableTW1Only <- x
@
<<galactoseOnly,echo=FALSE,results=tex>>=
galactoseOnly <- newSmutans( smutansGenes, title="Galactose Only" )
galactoseOnly <- smutans.de2( galactoseOnly, condition="galactose", 
                              condA="ua159", condB="tw1" )
#save(galactoseOnly, file=file.path("galactoseOnly.RData"))
#load("galactoseOnly.RData")
x <- smutans.de2List( galactoseOnly, qval=1e-3 )
x <- x[,c("id","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","$\\log_2$(TW1/UA159)", "$P$-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x <- x[abs(x[,2])>1,]
xtable.tabGalactoseOnly <- xtable( x[order(x[,2],decreasing=TRUE),],
  display=c("s","s","f","E","s"),
  digits=c(0,0,1,0,0),
  align=c('l','l','r','r','p{3.5in}'),
  label='tabGalactoseOnly',
  caption='{\\bf List of genes differentially expressed in UA159 and TW1 when growing in TV-galactose}.'
)
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
tableGalactoseOnly <- x
@
<<save,echo=FALSE,results=hide>>=
save.image(file="mss-lin.RData")
@
<<sequence,eval=FALSE,echo=FALSE,results=hide>>=
library(Biostrings)
x <- readFASTA("NC_004350.fna")
y <- DNAString(x[[1]]$seq)
irUnion[35]
i <- 35;toString(subseq(y,start(irUnion)[i],end(irUnion)[i]))
i <- 71;toString(subseq(y,start(irUnion)[i],end(irUnion)[i]))
i <- 116;toString(subseq(y,start(irUnion)[i],end(irUnion)[i]))
i <- 117;toString(subseq(y,start(irUnion)[i],end(irUnion)[i]))
i <- 204;toString(subseq(y,start(irUnion)[i],end(irUnion)[i]))
i <- 205;toString(subseq(y,start(irUnion)[i],end(irUnion)[i]))
rm(x,y)
# PredSmallRNA-35, PredSmallRNA-71, PredSmallRNA-116, PredSmallRNA-117, and 
# PredSmallRNA-204.
# In the other three sets of pairwise comparison, we found only a few
# differentially expressed regions:
# the region, PredSmallRNA-35, (Glucose/Galactose - TW1 background);
# PredSmallRNA-204, and PredSmallRNA-205 (UA159/TW1 - Glucose condition); and
# PredSmallRNA-116, and PredSmallRNA-205 (UA159/TW1 - Galactose condition).
@

\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color} 
\usepackage{setspace} 
\usepackage{rotating}
\usepackage{xr}
\usepackage{url}
\usepackage{longtable}
\usepackage{lscape}
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
\usepackage[round,comma,numbers]{natbib}

\usepackage{xcolor}
\usepackage{titlesec}
\titleformat{\section}[runin]{\bfseries\filcenter}{}{1em}{}
\titleformat{\subsection}[runin]{\bfseries}{}{1em}{}
\setcounter{secnumdepth}{0}
%\titlespacing*{\section}{0pt}{0pt}{0pt}
%\titlespacing*{\subsection}{0pt}{0pt}{0pt}

% \usepackage{../../latex/sty/genres}
\def\@cite#1#2{(#1\if@tempswa , #2\fi)}
\def\@biblabel#1{}

% Supplementary material
\externaldocument{mss-lin-supp}

% Text layout
\oddsidemargin 0in
\evensidemargin 0in
\topmargin -.5in
\textwidth 6.5in
\textheight 9in
\parskip 0.25in

\makeatletter
\SweaveOpts{keep.source=TRUE,eps=FALSE,include=FALSE,width=6,height=6}
\newcommand{\fixme}[1]{\textbf{(COMMENT BY: }\textit{\textcolor{red}{#1}})}
\newcommand{\change}[1]{\textit{\textcolor{blue}{#1}}}
\newcommand{\comment}[1]{}
\makeatother

% \pagestyle{myheadings}
\pagestyle{empty}
\markright{RNA-seq analysis in S. mutans}

\begin{document}

\begin{titlepage}

\title{CcpA and Catabolite Repression Explored by RNA-Seq in\\
{\em Streptococcus mutans}}

\author{Lin Zeng$^{1,4}$, Sang Chul Choi$^{2,3,4}$, Charles D.\ Danko$^{3}$, \\
Adam Siepel$^{3}$, Michael J.\ Stanhope$^{2}$, and Robert A.\ Burne$^{1}$}

% Melissa J.\ Hubisz$^{3}$, \\
% Paulina Dmitrievna Pavinski Bitar$^{2}$, \\

\date{ }
\maketitle
\thispagestyle{empty}

\begin{footnotesize}
\begin{center}
$^1$Department of Oral Biology,\\
College of Dentistry, University of Florida, Gainesville, Florida 32610 USA
\\[1ex]
$^2$Department of Population Medicine and Diagnostic Sciences,\\
College of Veterinary Medicine, Cornell University, Ithaca, NY 14853, USA
\\[1ex]
$^3$Department of Biological Statistics and Computational Biology,\\
Cornell University, Ithaca, NY 14853, USA
\\[1ex]
$^4$LZ and SCC contributed equally to this work.
\\
\end{center}
\end{footnotesize}

\vspace{1in}

\begin{tabular}{lp{4.5in}}
{\bf Submission type:}& Research Article
\vspace{1ex}\\
{\bf Keywords:}&Carbon catabolite repression, CcpA, RNA deep sequencing,
biofilm, dental caries
\vspace{1ex}\\
{\bf Running Head:}&RNA-seq analysis in {\em S.\ mutans}
\vspace{1ex}\\ 
{\bf Corresponding Author:}&
\begin{minipage}[t]{4in}
 Robert Burne\\
 Department of Oral Biology, University of Florida\\
 College of Dentistry, P. O.  Box 100424\\
 Gainesville, FL 32610\\
 Phone: +1-352-273-8850\\
 Fax: +1-352-273-8829\\
 Email: rburne@dental.ufl.edu
\end{minipage}
\end{tabular}

\end{titlepage}

\setcounter{page}{2}

\doublespacing
\raggedright
\parindent 2em
\parskip 0pt

\section*{ABSTRACT}

A bacterial transcriptome of the primary etiological agent of human dental
caries, {\em Streptococcus mutans}, is described here using RNA deep sequencing.
In comparison with previously published data generated using microarray
assays, differential expression profiles of the transcriptome in the context of
a strain lacking the CcpA protein and carbohydrates revealed good agreement with
DNA microarrays, yet expanded considerably the repertoire of DNA sequences that
showed statistically-significant changes in expression as a function of the
presence of CcpA and growth carbohydrate. Some of novel mRNAs and small RNAs
appear by virtue of their location or identity to influence the regulation
and prioritization of carbohydrates.
Our study demonstrated the validity of
RNA-seq as an alternative to DNA microarrays in studying gene regulation in {\em
S.\ mutans} by yielding a more precise landscape of transcriptomic changes in
response to specific mutations and growth conditions.

\clearpage{}

\section*{INTRODUCTION}

About one fifth of children between the ages of 2 and 19 were reported to have
untreated dental caries in the United States (National Center for Health
Statistics, 2010). It is generally accepted that the presence of acid-tolerant
bacteria, a carbohydrate-rich diet and a susceptible host are all required for
development of dental caries. In fact, carbohydrates introduced into the oral
cavity provide the preferred energy sources for the members of the oral
microbiome that are regarded as significant contributors to the caries process:
aciduric streptococci and {\em Actinomyces} subspecies, lactobacilli,
bifidobacteria and {\em Scardovia} species. These organisms are particularly
effective at converting host- and diet-derived carbohydrates into organic acids
that drive demineralization of the tooth.

Cariogenic bacteria, including the primary etiological agent of human dental
caries, {\em Streptococcus mutans}, are usually equipped with multiple pathways
for the internalization and catabolism of carbohydrates \citep{Ajdic2002}. The
development of these repertoires of carbohydrate catabolic pathways likely
reflects adaptation to the complex combination of carbohydrates that are
secreted by the host in glycoproteins and other glycoconjugates, as well as to
the variety of simple and complex carbohydrates that became part of the human
diet a few thousand years ago. Because of the complexities of the repertoire of
carbohydrates to which oral biofilms are exposed and the intermittent feeding
patterns of humans, it is reasonable to conclude that many of the most abundant
members of the oral microbiota have evolved sophisticated pathways to rapidly
and efficiently prioritize the assimilation, catabolism and storage of
carbohydrates. Likewise, when the diet becomes enriched for carbohydrates, the
microorganisms \change{that} must adapt their utilization of substrates as the
physico-chemical and microbial composition of the biofilms acquires the
characteristics associated with enhanced cariogenic potential. This includes
enrichment for the aforementioned aciduric organisms, lower biofilm pH and other
changes in microenvironments, e.g. redox that are well documented to alter
bacterial gene expression and physiology
\citep{Ahn2007,Burne2009,Kolenbrander2000,Marquis1995}.

Carbon catabolite repression (CCR) allows bacteria to utilize carbon sources in
a selective fashion, turning off non-essential catabolic functions while
activating pathways required for utilization of preferred carbohydrates and
other carbon sources \citep{Deutscher2008,Gorke2008}. In a number of low GC
content Gram-positive pathogenic bacteria, CCR has also been shown to affect the
expression of a variety of virulence factors in response to the source and
amount of carbohydrate \citep{Gorke2008}. CCR in these organisms is primarily
exerted through the catabolite control protein A, CcpA. When preferred
carbohydrate sources are present, as sensed through the accumulation of
glycolytic intermediates, a protein kinase is activated (HprK) that can
phosphorylate the general sugar:phosphotransferase protein HPr at serine 46
(HPr-Ser46-P). Hpr-Ser46-P functions as a co-factor for the binding of CcpA to
conserved catabolite response elements (CRE) found near the start sites of
target genes, which can activate or repress gene expression depending on the
gene, other factors and the position of the CRE \citep{Deutscher2008}.

CcpA was shown to regulate carbohydrate metabolism and virulence expression in
{\em S.\ mutans} in a transcriptomic study using the microarray technique based
on DNA hybridization \citep{Abranches2008}. A RegPrecise search
(http://regprecise.lbl.gov) of the {\em S.\ mutans} UA159 genome yielded 99
genes in 48 operons with potential CcpA CREs detected in their promoter regions
\citep{Novichkov2010}. Although only a small portion of these genes have been
confirmed to be subject to CCR regulation in {\em S.\ mutans}, many more have
homologous genes whose regulation in related bacterial species have been
demonstrated to be regulated by CcpA.  A comparison with the results of our
previous Microarray analysis indicated that at a twofold cutoff, near half of
the operons suggested by the RegPrecise website were absent in the genes found
differentially expressed by the study \citep{Abranches2008}. Therefore, we
intend to take advantage of this {\em in silico} work and compare it to genes
identified by both Microarray and RNA-seq analyses as having significantly
differential expression.  This comparison could be used as one of the criteria
for validating the two technologies. Furthermore, another {\em S.\ mutans} study
using microarrays that included all intergenic regions (IGR) of UA159 has
reported a number of IGRs having differential expression in cell subpopulations
sorted according to the expression levels of a {\em comX-GFP} fusion, a reporter
that responds to extracellular bacterial quorum-sensing signal CSP
(competence-stimulating peptide) \citep{Lemme2011}. Some of these IGRs likely
encode as yet unknown, regulatory RNA or transcripts that may help regulate the
development of competence, biofilm formation and stress tolerance in {\em S.\
mutans}. We speculate that some other IGRs in the genome of {\em S.\ mutans}
could play regulatory roles in carbohydrate metabolism.

As the experiment of choice to many who studied bacterial transcriptome in the
last decade or so, microarray analysis has proven to be reliable, rapid and
comparatively economical \citep{Conway2003}. However, in standard arrays only
genes or transcripts that are included in a predetermined set of probes can be
detected in any given assay, therefore the scope of microarray analysis became a
limiting factor when a function was mediated by unannotated transcripts or
genes. Very recently, RNA deep sequencing (RNA-seq) has facilitated
annotation-free \change{probe-free?} detection of transcripts, with greater
sensitivity and dynamic ranges in RNA expression levels than microarrays
\citep{Croucher2010,Filiatrault2011}. In addition, RNA-seq allows for the
analysis of RNAs in non-coding regions, small RNAs (sRNAs) and antisense
transcripts. However as a relatively new tool in modern bacteriology, cautions
and considerations are due while embracing this technology. These include the
inherent bias due to steps that require inverse \change{reverse?} transcription
and ligation, and particularly for bacterial samples, relatively low
signal-to-noise ratios partly due to the presence of a large ribosomal RNA
population. In order to test the applicability of RNA-seq in transcriptome study
in {\em S.\ mutans}, we used standard RNA deep sequencing technique to sequence
RNAs in 13 {\em S.\ mutans} samples, focusing on messenger and small RNA
expression levels in response to different carbohydrates (glucose and galactose)
in both the wild-type strain (UA159) and a {\em ccpA} mutant (TW1)
\citep{Abranches2008}. As we have previously conducted Microarray analysis with
the same set of strains growing in identical conditions, these new data allowed
us to validate the RNA-seq technology while expanding the scope of RNAs that may
play a role in regulation of carbohydrate metabolism, a central factor in {\em
S.\ mutans} virulence.

\section*{MATERIALS AND METHODS}

\subsection*{Bacterial strains and growth conditions}

{\em Streptococcus mutans} strains UA159 and TW1 \citep{Abranches2008} were
maintained on BHI (Difco Laboratories, Detroit, MI) agar plates, and bacterial
cultures used for extraction of RNA were prepared with Tryptone-vitamin
\citep{Burne1999} based medium supplemented with 0.5\% of glucose or galactose
(Sigma, St. Louis, MO). Four repeats, each in the volume of 15 mL, were included
for the culture of strain UA159 growing in TV-galactose, while 3 repeats were
used for each of the other 3 cultures. Bacterial cultures were incubated
statically in the presence of 5\% of CO$_2$ at $37^\circ$C until they reached
mid-exponential phase (OD$_{600}$ $\approx$ 0.5), harvested by centrifugation at
$4^\circ$C for 10 min, treated with bacterial RNAprotect Bacteria Reagent
(Qiagen, Germantown, MD), and immediately stored at $-80^\circ$C.

\subsection*{RNA isolation, enrichment, and sequencing}

Total RNA was extracted from frozen bacterial cells using the RNeasy Mini kit
(Qiagen) according to previously published protocols \citep{Ahn2005}. To remove
16S and 23S rRNAs, 10 $\mu$g of high-quality total RNA was processed using a
MICROB{\em Express}{\texttrademark}  Bacterial mRNA Enrichment Kit (Ambion of Life
Technologies, Grand Island, NY), twice, before precipitating with ethanol and
resuspended in 25 $\mu$L of nuclease-free water. The final quality of enriched
mRNA samples was analyzed using an Agilent Bioanalyzer (Agilent Technologies,
Santa Clara, CA).
\change{
We generated cDNA libraries from the enriched mRNA samples using the TruSeq
Illumina kit (Illumina, San Diego, CA), following instructions from the
supplier. Deep sequencing of the RNA samples were performed at the Cornell
University Life Sciences Core Laboratories Center (Ithaca, NY). 
}
\fixme{RAB:more details please: e.g., adapter sequences? Can the protocols be easily obtained using info provided here?}
\fixme{SCC:The raw RNA-seq data set can tell what adapter sequences we used.
Mike and Paulina should know how precise the protocol is.
I need to ask Mike and Paulina about this part. 
I removed a subsection ``RNA sequencing'' becuase it is redundant.}

\subsection*{Short reads alignments}

Approximately 20 million short-reads were obtained for each sample.  Because the
aligner, BWA \citep{Li2009a}, allowed a few gaps for efficient alignments of
millions of reads of size being around 100 base pairs long, shorter reads that
consisting of partially sequencing adapters would not be mapped.  We, therefore,
removed adapter sequences of each short-read \citep{Martin2011}, and trimmed the
resulting sequences by quality score from the 3'-end \citep{Schmieder2011}
before mapping onto the reference genome of strain UA159 (GenBank accession
AE014133) using the short-read aligner.  Mapped short-read alignments were then
converted into readable formats using SAMTOOLS \citep{Li2009b}.

\subsection*{Transcript predictions}

RNA transcripts were inferred by applying a hidden Markov model to site-wise
expression levels \citep{Martin2010}. A pileup command of BWA was used to
convert the short-read alignments into pileup values, which were taken as the
site-wise expression levels along the genome. Site-wise expression levels are a
list of non-negative integers that represent numbers of short-reads mapped at a
particular genomic position. The transcript inference program, ParseRNAseq,
with the following options was used: ``-c 10 -b 25 -force gp'' to bin expression
levels into 25 parts, and to allow 10 emission states for relative expression
intensity. Genes that were annotated in close approximation were used as the
predicted parts of a transcript, while no information regarding the precise
transcription initiation- or stop-sites were pursued in this study due to
limited resources. For similar reasons, the orientation of each transcript could
not be verified solely based on RNA-seq data; instead we used the information of
annotated genes to determine the strands of predicted transcripts.

\subsection*{Prediction of small RNAs and targets}

Although our RNA-seq protocol did not specifically enrich non-coding small RNAs
in the cDNA preparation, small RNAs were retained in the RNA samples as only
ribosomal RNAs were depleted using specific oligonucleotides. Consequently, it
was difficult to discriminate cDNA originated from non-coding RNAs small from
that of mRNAs, as non-coding RNAs expressions were often masked by the expression
of neighboring background mRNA. Therefore, we sought a way of predicting small
RNAs using a computer program called RNAz \citep{Gruber2010}, which used homologous sequences
and RNA secondary structures to predict putative non-coding small RNAs. Because
sequence alignment was a critical step for finding small RNAs via this approach,
intergenic regions that also included up- and down-stream sequences were
extracted, BLAST-searched against a database of bacterial genomes
\citep{Altschul1990}, and the resultant sequence alignments were further refined
using a program, MUSCLE \citep{Edgar2004a}. Subsequently, RNAz was applied
to the alignments for scoring intergenic regions for putative small RNAs
\citep{Washietl2005}. Target genes for each candidate small RNA were also
predicted using RNAplex \citep{Tafer2008} and RNAplfold \citep{Bernhart2006},
and the resultant target genes were used to perform functional category
enrichment tests based on their scores by these two programs.
\change{
In addition to the above procedure for predicting small RNAs, we also employed a
method of rho-independent terminators (RITs) identification
\citep{Kingsford2007}, to prepare candidate small RNAs, that were subsequently
scored using RNAz, and a transcriptional signal-based method to identify
intergenic sRNA transcriptional units (TUs) \citep{Sridhar2010}. 
}

\subsection*{Statistical analysis for differential expressions}

The R package DESeq \citep{Anders2010} was used to determine differentially
expressed genes on the basis of the negative binomial model \citep{Robinson2007}.
Detailed steps of analyzing RNA-seq data for identifying differentially
expressed genes were well described \citep{Oshlack2010}; short-reads aligned to
a particular annotated gene in the reference genome were counted, generating a
table of read counts of all the protein coding genes. Statistical software R
\citep{R2011} with the R package DESeq was then employed to infer differentially
expressed genes in the four biological conditions from two factors of UA159/TW1
and glucose/galactose. \change{For normalizing expression levels among samples,
total sequencing depths for each sample were estimated as the median of the
ratios of the sample's counts to geometric mean across samples
\citep{Anders2010,Robinson2010}.}

\subsection*{Gene functional category associations}

Three sets of gene categories were compiled for testing functional associations
of differentially expressed genes. First, genes were each assigned to Gene
Ontology (GO) categories by comparing to bacterial proteins from the Uniref90
database using the BLASTP program, and a GO classification was assigned if the
match had an $E$-value of $<1.0\times10^{-5}.$ A gene family was assigned a
given classification if any of its genes was assigned that classification.
Second, we collated functional classes of genes in the Oral Pathogen Sequence
Databases available at \url{http://oralgen.lanl.gov}. Third, genes were mapped
onto a set of metabolite pathways for {\em S.\ mutans} available at Kyoto
Encyclopedia of Genes and Genomes (KEGG) \citep{Kanehisa2012}. Using the sets of
gene categories for {\em S.\ mutans}, we performed a Mann-Whitney {\em U} test of the
values of target genes of small RNAs in association with a given category versus
those for the other categories, and also conducted a variation of Fisher's exact
test accounting for gene lengths to test enrichments of differentially expressed
genes \citep{Young2010}. Both tests were corrected for multiple testing
hypotheses \citep{Benjamini1995}.
\fixme{RAB:do we need more explanation here or is the
reference sufficient?}
\fixme{SCC:These are enough for reproducing the results.}

\subsection*{UCSC genome browser tracks and data availability}

We created tracks for recently released Streptococcus Genome Browser
(\url{http://strep-genome.bscb.cornell.edu}) that summarized the results of our
{\em S.\ mutans} transcriptomic analysis with known genes, gene expression
levels based on the short-read alignments, predicted putative transcript
segments, and predicted non-coding small RNAs. The results can be accessed by
clade ``Streptococcus,'' genome ``S. mutans UA159,'' and \change{assembly ``January
2006.''} These tracks can be used to inspect loci of interest and to compare the
results of different RNA-seq data sets. They can also be queried and intersected
with other tracks using the UCSC Table Browser.

\section*{RESULTS and DISCUSSION}

\subsection*{Clustering of RNA samples}

Based on the results of read counts of all annotated genes, a total of 13
RNA-seq samples were clustered without supervision. As illustrated in Figure
\ref{figClust}, the effect of carbohydrate source appeared to be significantly
stronger than that of the loss of the {\em ccpA} gene. Nevertheless, the
replicates of the same bacterial strain growing in the same carbohydrate
conditions clearly clustered together, indicating that the transcriptomic shifts
were the results of both sugar specificity and regulation by CcpA.

\subsection*{Predicted transcripts}

We obtained a set of transcripts using the RNA-seq data for each of the 13
RNA-seq samples. In order to summarize the 13 sets of predicted transcripts,
all of the transcripts of each set were scored based on their
average site-wise expression levels in relation to transcripts in the rest of
set. More highly-ranked transcripts were placed first on the reference genome
based on their scores, while lower-ranked transcripts were purged from the
transcriptome map. In doing so, the final set of transcripts was formed as
non-overlapping transcripts. The distribution of expression levels
of the predicted transcripts are shown in Figure \ref{figGrCombinedExpression}.
We designated transcripts as expressed whose average site-wise expression levels
were greater than 5 and proportion of sites of zero site-wise expression levels
were less than 10\%. Of 
\Sexpr{length(smutansData.txGrCombined)} predicted transcripts,
\Sexpr{length(notExpressedTx)} 
ones were either not expressed or very poorly expressed.  Genes that belonged to
expressed transcripts were designated as expressed.  Of the total 
\Sexpr{length(smutansData.txGenes)}
of genes, 
\Sexpr{length(as.matrix(matchGenesTx)[,1])}
(\Sexpr{as.integer(length(as.matrix(matchGenesTx)[,1])/length(smutansData.txGenes)*100)}\%) 
were designated as expressed (Figure \ref{figLength}). 
Genes that were not designated as expressed could
be because of incorrect gene annotations, or no expression under the
experimental conditions. Short reads might be not be alignable on duplicate
genes because reads could be aligned to multiple places, and those were not
mapped unequivocally \citep{Li2009a}.
Of the 
\Sexpr{length(grTx)} 
expressed transcripts, 
\Sexpr{length(grTxWOAnnotation)} 
contained no annotated genes or RNAs
\fixme{SCC: I might need to find those expressed transcripts on unannotated regions in a table}.
To assess the tranascript map by measuring how constant expression levels
between genes within polycistronic transcripts, gene expression levels for
adjacent genes within expressed transcripts were more correlated than those
between adjacent transcripts (Figure \ref{figExpressionTwoGene}).  Among the 
\Sexpr{length(unique(matchMatrix(matchGenesTx2)[,"subject"]))} 
final segments, 
\Sexpr{sum(number.genes.tx==1)} 
segments contained a single gene while 
\Sexpr{sum(number.genes.tx > 1)} 
contained more than one gene.

Figure \ref{figTranscriptLength} shows the length distribution of expressed
transcripts with annotated
coding genes and unannotated regions. Unannotated regions could include
potential non-coding regions and novel coding genes. Transcripts of unannotated
regions appeared to be shorter than those with coding genes. Most of the
unannotated regions with length near 360 bp could be non-coding regions, and the
rest of the unannotated regions with length near 1000 bp could be new coding
genes. Some of the unannotated genes may represent pseudogene regions. We used
strandnesses of known genes to find strands of segments although we could not
tell strands of the predicted transcripts solely based on our RNA-seq data
because a strand-specific RNA-Seq technique was not employed here. Of the 
\Sexpr{length(grTx)} 
expressed transcripts, 
\Sexpr{sum(strand(grTx)=='*')} 
included annotated genes of both strands 
(\Sexpr{as.integer(sum(strand(grTx)=='*')/length(grTx)*100)}\%).
We then filtered out the transcripts containing genes with conflicting
strandness, leaving 
\Sexpr{length(grTx2)}
expressed transcripts. Figure \ref{figUTRLength} shows length distributions of
5' and 3' UTR of our finally determined  
\Sexpr{length(grTx2)}
transcripts.

\subsection*{Prediction of small RNAs and targets}

\change{
We merged the three sets of regions that were predicted to contain 
small RNAs from the three different prediction methods. 
While the three sets of predicted regions contained
\Sexpr{length(irRNAz)} (RNAz),
\Sexpr{length(irPichon)} (rho-independent terminators based identification), and
\Sexpr{length(irScanner)} (transcriptional signal-based method),
the number of elements in the union set was \Sexpr{length(irUnion)}.
Of the \Sexpr{length(irUnion)} predicted regions of the union set,
\Sexpr{sum(countOverlaps(irUnion,c(irPichon,irRNAz,irScanner)) > 1)}
regions were constructed from multiple predictions. 
Among the predicted regions of the union set,
\Sexpr{sum(countOverlaps(irUnion,irCm) > 0)}
regions overlapped regions annotated by Infernal (v1.0)
using default options for models in RFAM 9.1 \citep{Gardner2009,Nawrocki2009}
(\Sexpr{length(irCm)} structural RNA regions). 
Because we focused our search of small RNAs on non-genic regions, only 
\Sexpr{sum(countOverlaps(irUnion,irGene) > 0)}
predicted small RNA regions overlapped known genes.  We surveyed neighboring
genes of regions that could contain small RNAs, showing that some of predictions
were clustered together, and others were lonely placed in between known genes
(Table \ref{tabNeighborSmallRNA}).  The functional categories of predicted
target genes of RNAz predicted small RNAs are shown in Table \ref{tabGoseq}.  We
also used the union set of predicted small RNAs containing regions to determine
which regions were differentially expressed.  Because of a smaller set of the
union regions than the gene set of {\em S.\ mutans}, we found only a few regions
that were differentially expressed. For example, when comparing UA159 wild-type
cells grown in glucose and galactose, among 243 predicted regions, 129 regions
were removed from the DE analysis because of their low or zero expression
levels. The following five regions were expressed differentially:
PredSmallRNA-35, PredSmallRNA-71, PredSmallRNA-116, PredSmallRNA-117, and 
PredSmallRNA-204. In the other three sets of pairwise comparison, we found only
a few differentially expressed regions:
the region, PredSmallRNA-35, (Glucose/Galactose - TW1 background);
PredSmallRNA-204, and PredSmallRNA-205 (UA159/TW1 - Glucose condition); and
PredSmallRNA-116, and PredSmallRNA-205 (UA159/TW1 - Galactose condition).
We obtained the MFE structure drawing with encoding base-pair probabilities
for PredSmallRNA-204 regions (Figure \ref{figRNA204}), which shows high
probability of base-pairing among bases in the region.
}

\fixme{RAB:Why these sRNA do not appear to match that in Table S1? If possible,
can we combine these two tables into one?.}
\fixme{SCC:Table \ref{tabNeighborSmallRNA} is a table for regions that were
predicted to contain small RNAs, while table \ref{tabGoseq} is table for targets
of predicted small RNAs.}

\fixme{RAB:Also please make sure these structures are of the ones we want to
present.} 
\fixme{SCC:Now, we have only one RNA secondary structure in the supp.}

\fixme{SCC:I remove the section of differentially expressed small RNAs. We now
have only one section for small RNAs. I deleted most of the text of the previous
analyses of DE study of some small RNA predictions by RNAz.}
% \subsection*{Differentially expressed small RNAs and transcripts}

\subsection*{Differential expression of RNA transcripts}

Two independent factors of the biological processes being studied here included
the effect of CcpA and the treatment growth in glucose versus galactose,
resulting in four pair-wise comparisons. We compared two conditions of one
factor with the other factor being fixed at one condition. The cut-off for
designating a gene as being differentially expressed was a change in transcript
levels of at least 2-fold \change{and $P$-value of $10^{-3}$}.
\fixme{SCC:I have updated tables and values mentioned in the main text using
these criteria of 2-fold change and $10^{-3}$ $P$-value.}

\paragraph{Comparison between wild-type and ccpA mutant in glucose-grown cells:}
Table \ref{tabGlucoseOnly} lists differentially expressed genes in the
comparison of wild-type UA159 and the {\em ccpA} mutant strain TW1 grown with
glucose. 
Of 
\Sexpr{nrow(counts(smutansGenes))}
genes, 
\Sexpr{sum(tableGlucoseOnly[,2]>0)}
genes
were found to have at least twofold increase in expression levels in the mutant
relative to the wild-type strain. At the same time, 
\Sexpr{sum(tableGlucoseOnly[,2]<0)}
genes showed at least
twofold reduction in expression in the loss of CcpA. Among these 
\Sexpr{nrow(tableGlucoseOnly)}
differentially
expressed genes, 
\Sexpr{length(grep("Energy metabolism",tableGlucoseOnly[,"Functional Class"]))}
% \change{Used to be 20}
were involved in energy metabolism including glycolysis,
fermentation and sugar utilization; 
\Sexpr{length(grep("PTS",tableGlucoseOnly[,"Functional Class"]))}
% \change{Used to be 10}
encoded Enzyme II components of the PTS 
(see Tables \ref{tabGoseqGlucoseOnly} and \ref{tabKeggGlucoseOnly} for gene
ontology and KEGG terms that were enriched with differentially expressed genes
in this comparison)
and were required for transporting glucose, mannose, sucrose and cellobiose; and
\Sexpr{length(grep("Two-component",tableGlucoseOnly[,"Functional Class"])) + length(grep("Regulatory",tableGlucoseOnly[,"Functional Class"]))}
were classified as regulatory or two-component system genes
(Figures \ref{figDotplotGlucoseOnly}, and \ref{figPieGlucoseOnly}).
We also found                                                                   
\Sexpr{length(grep("Hypothetical|Unknown|Unassigned",tableGlucoseOnly[,"Functional Class"]))}
% \change{Used to be 8}
genes annotated as unknown functions.  We confirmed that the most up-regulated genes
in TW1 encoded the components of the pyruvate dehydrogenase (PDH) enzyme complex
(SMU.1421c$\sim$1424c) as reported previously \citep{Abranches2008}.
Down-regulated genes included a cytoplasmic alpha-amylase (SMU.1590) that was
located downstream of {\em ccpA} \citep{Simpson1998}, and a fructosyltransferase
gene {\em ftf} (SMU.2028) encoding the enzyme involved in converting sucrose to
a fructan homopolymer \citep{Shiroza1988}. These results appeared to be
generally consistent with those of the microarray analysis
\citep{Abranches2008}.

In agreement with the {\em in silico} prediction provided by the RegPrecise
website, results from the RNA-seq study (but not that of microarray study)
identified the following as differentially expressed: genes encoding enzymes
pyruvate kinase (SMU.1190c; decreased by twofold in TW1) and pyruvate-formate
lyase (SMU.402c; increased by twofold in TW1), a putative thiamine biosynthesis
lipoprotein (SMU.1088), SMU.1125c of a putative ribonucleoside-metabolism
operon, a ribosome associated protein (SMU.500), the sucrose-6-phosphate
hydrolase (SMU.1843) and sucrose-PTS EII (SMU.1841c) \citep{Sato1989}, the major
glucose-PTS EII$^{\mathrm{Man}}$ (SMU.1877$\sim$1879) \citep{Abranches2003}, the
cellobiose-PTS operon (SMU.1596c$\sim$1600c) \citep{Zeng2009} and another
$\beta$-glucoside-PTS EII (SMU.980) \citep{Cote2000}. On the other hand, genes
identified only in the microarray study that matched the RegPrecise predictions
included: glucose-6-phosphate isomerase (SMU.307), a putative
ribulose-monophosphate-PTS EIIC component (SMU.270) and a hypothetical protein
(SMU.799c).

There also appeared to be evidence suggesting improved consistency using RNA-seq
technology. For example, a proven CCR-sensitive operon, the glycogen
biosynthesis {\em glg} cluster that spans SMU.1535c$\sim$1539c
\citep{Abranches2008,Harris1992}, was shown here by RNA-seq to be uniformly
derepressed in strain TW1 growing on glucose. The microarray analysis of the
same strains grown in the same condition, however, indicated that the expression
of only one of the genes, SMU.1537c, was altered in a statistically significant
way. The same was true with an inducible fructose-PTS operon (SMU.870$\sim$872).
Furthermore, RNA-seq analysis, at a twofold cutoff, showed disproportionally
fewer genes 
(\Sexpr{sum(tableGlucoseOnly[,2]<0)} out of \Sexpr{nrow(tableGlucoseOnly)}) 
% (11 out of 62) 
with decreased expression in TW1 as compared to UA159, whereas
microarray analysis showed 
% RESULT NOT FROM THIS RESULT
110 
genes with decreased expression and 
% RESULT NOT FROM THIS RESULT
61 
genes with
increased expression in TW1 at the same cutoff, when both were growing in
TV-glucose. Considering the collective evidence regarding CCR in Gram-positive
bacteria, it was conceivable that CcpA exerted its function predominantly through
negative regulations \citep{Deutscher2008}.

\paragraph{Comparison between glucose- and galactose-grown UA159 cells:}
Figure \ref{figPieUA159Only} shows proportions of functional categories that
included differentially expressed genes between glucose-grown and
galactose-grown UA159 cells. Among the four pairwise comparisons this showed the
largest number 
\Sexpr{nrow(tableUA159Only)} 
% \change{(used to be 139)}
of genes being
differentially expressed (Table \ref{tabUA159Only}). Again, energy metabolism
had the most
(\Sexpr{as.integer(length(grep("Energy metabolism",tableUA159Only[,"Functional Class"]))/nrow(tableUA159Only)*100)}\%)
% \change{used to be (33\%)}
differentially expressed genes and as many as 
% \change{used to be 13}
\Sexpr{length(grep("PTS",tableUA159Only[,"Functional Class"]))}
PTS component genes were
affected, reflecting potential changes in uptake of lactose, glucose, fructose,
mannose, cellobiose, sucrose and trehalose
(see Table \ref{tabGoseqUA159Only} to see that carbohydrate metabolic process is
enriched with genes expressed differentially in the change of energy source
carbohydrate).
The most highly enriched KEGG category in the effect of galactose on UA159 was
galactose metabolism (Table \ref{tabKeggUA159Only}).
Clearly, glucose is a potent
activator of CCR, negatively affecting the expression of various glycolytic
components and PTS operons in {\em S.\ mutans} genome. 
Notably in {\em S.\ mutans}, the
effects of glucose on some of these PTS operons, including EII$^{\mathrm{Lac}}$
\citep{Zeng2010} EII$^{\mathrm{Cel}}$
\citep{Zeng2009} and EII$^{\mathrm{Fru}}$ (SMU.1956c$\sim$1961c)
\citep{Zeng2008} have been shown to be mediated primarily
through the major glucose-PTS EII$^{\mathrm{Man}}$, despite the ability of CcpA
to exert its
regulation indirectly by regulating the expression of EII$^{\mathrm{Man}}$
\citep{Zeng2008}. Therefore,
the transcriptomic alteration shown here is likely a compounded result of these
two tiers of regulation. Notwithstanding, this profound change in gene
expression was also partly due to the fact that galactose and its derivatives
could serve as inducers for activating expression of the Leloir pathway and
tagatose pathway genes that are responsible for assimilating galactose and
lactose \citep{Ajdic1998,Zeng2010}. These include consecutive genes SMU.1488c to SMU.1496c
(tagatose pathway) and genes SMU.886 to SMU.889 (Leloir pathway). Interestingly,
over a quarter 
(\Sexpr{as.integer(length(grep("Hypothetical|Unknown|Unassigned",tableUA159Only[,"Functional Class"]))/nrow(tableUA159Only)*100)}\%)
of all differentially expressed genes were considered
hypothetical or of unknown function,
and most 
(\Sexpr{as.integer(sum(tableUA159Only[grep("Hypothetical|Unknown|Unassigned",tableUA159Only[,"Functional Class"]),2]<0)/length(grep("Hypothetical|Unknown|Unassigned",tableUA159Only[,"Functional Class"]))*100)}\%)
% (\Sexpr{as.integer(sum(tableUA159Only[,2]<0)/nrow(tableUA159Only)*100)}\%)
% \change{used to be 26} genes
of these were
up-regulated in the presence of glucose (Figure \ref{figDotplotUA159Only}). Notably, other genes
suppressed by glucose include the {\em glg} operon (SMU.1535c$\sim$1539c), the
glucan-synthesizing enzymes {\em gtfBC} (SMU.1004$\sim$1005)
\citep{Shiroza1987,Ueda1988}, a glucan-binding protein {\em gbpC} (SMU.1396c)
\citep{Sato1997} and a NAD-dependent aldehyde dehydrogenase {\em gabD}
(SMU.2127).

Our previous microarray analysis of the same set of samples showed similar
patterns of change in gene expression, with significant portion (28 out of 90)
of the affected genes encoding hypothetical proteins \citep{Abranches2008}. In comparison with the
predictions at RegPrecise website, the RNA-seq analysis identified 
12 
\fixme{SCC:Can Lin change this number 12 using the updated Table
\ref{tabUA159Only}?}
presumably
CCR-sensitive operons, 
8 
\fixme{SCC:Can Lin change this number 8 as well using the updated Table
\ref{tabUA159Only}?}
of which were missing from the microarray study
performed under identical conditions: SMU.113$\sim$114 (fructose-1-phosphate kinase
and a fructose-PTS EIIBC) \citep{Wen2001}, SMU.1088 (a putative thiamine biosynthesis
lipoprotein), SMU.1843 ({\em scrB}), SMU.2127 ({\em gabD}), SMU.1596$\sim$1600 (cellobiose
operon), SMU.1877$\sim$1879 (EII$^{\mathrm{Man}}$), SMU.980 ($\beta$-glucoside PTS EII) and the
sucrose-PTS ({\em scrA}). Conversely, 8 operons from the list of RegPrecise were
identified by Microarray study and 4 exclusively: SMU.575c ({\em lrgA})
\citep{Ahn2010}, SMU.396
(a glycerol-uptake facilitator protein, {\em glpF}), SMU.877$\sim$883 (the multiple-sugar
metabolism pathway, {\em msm}) \citep{Russell1992} and SMU.402 (pyruvate-formate lyase).

\paragraph{Comparison between glucose- and galactose-grown cells in the
absence of CcpA:} 
As mentioned above, the effect of sugar appears dominant to that of loss of
CcpA.  Similar to observations made with UA159, when transcriptomes from TW1
grown in glucose or galactose were contrasted, energy metabolism 
(\Sexpr{as.integer(length(grep("Energy metabolism",tableTW1Only[,"Functional Class"]))/nrow(tableTW1Only)*100)}\%)
% \change{used to be (36\%)}
was the largest category of genes differentially expressed
(Figure \ref{figPieTW1Only}).
Similarly to the comparison of UA159 cells grown in glucose and galactose,
carbohydrate metabolic process was highly ranked in the functional categories
enriched with differentially expressed genes (Table \ref{tabGoseqTW1Only}),
and a KEGG category, called galactose metabolism, was found to be enriched with
the differentially expressed genes (Table \ref{tabKeggTW1Only}).
In contrast to
wild-type cells however, only
\Sexpr{length(grep("PTS",tableTW1Only[,"Functional Class"]))}
% \change{used to be 7}
PTS genes were identified that encoded PTS components for lactose, fructose and
mannitol. Notably up-regulated by galactose in TW1 were the tagatose and Leloir
pathways for galactose/lactose metabolism, the {\em glg} operon for glycogen
metabolism, {\em gbpC}, {\em gtfB} and the {\em msm} pathway.  A total of 
\Sexpr{sum(tableTW1Only[,2]<0)} 
% \change{used to be 57}
genes were found down-regulated 
in galactose-grown TW1 cells (Table \ref{tabTW1Only}), with majority 
(\Sexpr{as.integer(sum(tableTW1Only[grep("Energy metabolism",tableTW1Only[,"Functional Class"]),2]<0)/sum(tableTW1Only[,2]<0)*100)}\%)
of
which being designated as hypothetical proteins (Figure \ref{figDotplotTW1Only}).

Different from our previous Microarray study that saw as many as 515 genes
differentially expressed by at least twofold in TW1, with more than 50\% having
lower expression in galactose conditions, RNA-seq showed only 
\Sexpr{nrow(tableTW1Only)} 
\change{used to be 130}
\fixme{SCC: Where do we find this number 130? The number 42 is the number of rows in Table 4.}
in total
affected in TW1 as a function of the growth carbohydrate. As the Microarray
study was carried out using the total RNA and the RNA samples used in RNA-seq
analysis were enriched for mRNA and other non-rRNA populations, this discrepancy
perhaps implies an improved signal-to-noise ratio for the RNA-seq technology,
since the difference between TW1 cells grown in glucose and galactose conditions
is expected to be greatly reduced due to the loss of CcpA.

\paragraph{Comparison between the wild-type and ccpA mutant growing in
TV-galactose:} 
As
reported previously, galactose is less effective than glucose at triggering CCR
in {\em S.\ mutans} \citep{Zeng2008}, so the finding that there were fewer differences in the
transcriptomes of UA159 and TW1 in galactose-grown cells was expected. Although
energy metabolism was still the predominant category 
(\Sexpr{as.integer(length(grep("Energy metabolism",tableGalactoseOnly[,"Functional Class"]))/nrow(tableGalactoseOnly)*100)}\%)
% \change{used to be (27\%)}
among all
differentially expressed genes 
(Figures \ref{figPieGalactoseOnly}, and \ref{figDotplotGalactoseOnly}).
Similar to what was observed in glucose-grown cells, a
few genes including those for PDH components (SMU.1421c$\sim$1424c) were up-regulated
in TW1, and alpha-amylase (SMU.1590c) and $\beta$-D-fructosyltransferase {\em
ftf}
(SMU.2028c) were down-regulated. In contrast, no major PTS genes were
differentially expressed in the mutant during growth in galactose. Because both
strains were grown in galactose, genes involved in galactose and lactose
utilization were likely being expressed similarly and therefore no differences
were noted. Consequently, no gene ontology terms were found significantly
associated with differentially expressed genes. These results are largely
consistent with our previous microarray analysis \citep{Abranches2008}.

\section*{CONCLUDING REMARKS}

Our current knowledge of the {\em S.\ mutans} transcriptome is limited to annotated or
predicted genes. Whole transcriptome sequencing using high throughput sequencing
technologies, or deep RNA sequencing (RNA-seq) allowed us to reveal the
transcriptomic topography of {\em S.\ mutans}. Instead of aiming at pin-pointing
transcript start or end sites we focused on differential expression of genes in
response to nutrient availability and deletion of {\em ccpA} from UA159. Although we
did not use strand-specific RNA-seq that could be important in gene-dense
transcriptomes of species like bacteria or archaea, our application of a hidden
Markov model for predicting transcripts to RNA-seq data in conjunction with
known gene annotations allowed for inference of segments of putatively
co-transcribed regions. In doing so, we attempted to ensure a high-quality
transcriptome assembly. Although this map of transcripts did not precisely tell
us where transcripts start and end, it does provide information as to which open
reading frames could be co-transcribed under the conditions tested. 
In comparison with our previous Microarray study of differential gene expression
using the same set of samples, we observed good agreement between these two
methodologies, as well as identification of genes by each method alone that
appear to match an {\em in silico} analysis published online at RegPrecise website,
and consistent with other independent studies. Therefore, we have identified
additional genes or transcripts that are differentially regulated in {\em S.\ mutans}
in response to carbohydrate change or loss of CcpA. We also found discrepancies
of annotated genes and their transcription, for example a gene containing two
transcripts and non-coding regions producing significant number of transcripts.
We hope that our work could serve as a foundation for a comprehensive study of
the {\em S.\ mutans} transcriptome.

\section*{ACKNOWLEDGMENTS}

\fixme{SCC:I wish to add A.S. and C.D.D. as coauthors. I am not sure whether
this is okay. Mike would know better than I do about who should be in the list
of coauthors.}
This study was funded by the National Institute of Allergy and Infectious
Disease, US National Institutes of Health, under grant number AI073368-01A2 (to
M.J.S. and A.S). Additional support was provided by National Science Foundation
Career Award DBI-0644111 and a David and Lucile Packard Fellowship for Science
and Engineering (to A.S). We
thank Christopher Browngardt for assistance with labeling functional classes of
{\em S.\ mutans} genes.

\renewcommand*{\refname}{REFERENCES}
\bibliographystyle{jbactas}
\bibliography{burne}
\clearpage{}

\section*{TABLES}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table 1. Functional category enrichments of target genes
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology enrichments for target genes of putative small RNAs.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{lcccl}
\hline 
small RNA & $q^a$ & count & GO & Description \\
\hline 
%-----------------------------------------------------------------------------
<<tabGoseq,echo=FALSE,results=tex>>=
smutans.mannwhitney( qval=0.05,
                     genes.criteria=smutans.genes.criteria,
                     go.genes=smutans.go.genes,
                     cat.desc=smutans.cat.desc )
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tabGoseq}
\end{table}
%-----------------------------------------------------------------------------

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table 2. UA159 vs. TW1 with glucose fixed
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage{}
<<tabGlucoseOnly,echo=FALSE,results=tex>>=
print( xtable.tabGlucoseOnly,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE,
       sanitize.text.function = function(x){x}) 
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table 3. Glucose vs. Galactose with UA159 fixed
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage{}
<<tabUA159Only,echo=FALSE,results=tex>>=
print( xtable.tabUA159Only, 
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE,
       sanitize.text.function = function(x){x}) 
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TAble 4. Glucose vs. Galactose with TW1 fixed
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage{}
<<tabTW1Only,echo=FALSE,results=tex>>=
print( xtable.tabTW1Only,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE,
       sanitize.text.function = function(x){x}) 
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% UA159 vs. TW1 with galactose fixed
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage{}
<<tabGalactoseOnly,echo=FALSE,results=tex>>=
print( xtable.tabGalactoseOnly, 
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE, 
       sanitize.text.function = function(x){x}) 
@

\clearpage{}
\section*{FIGURE LEGENDS}
\setlength{\parindent}{0pt}
\setlength{\parskip}{2ex}

{\bf Figure \ref{figClust}.
Cluster of the 13 RNA-seq samples.} 
Heatmap shows the Euclidean distances between pairs of the samples as calculated
from the variance-stabilizing transformation of the count data. One factor is
the kind of growth carbohydrate (glucose or galactose), and another is the
presence or absense of CcpA from the UA159 strain of {\em S.\ mutans}. The {\em
ccpA} deletion mutant type is designated tw1.

{\bf Figure \ref{figGrCombinedExpression}.
Distribution of expression levels of the predicted transcripts.}
The last bin sums from the expression level ranging from 1000 to 50,000. The
expression level is measured as the average number of reads mapped on predicted
transcripts.

{\bf Figure \ref{figExpressionTwoGene}.
Scatter plot of expression levels of pairs of adjacent genes located in the same
transcripts.}
(Top) Two genes in the same tranascripts are plotted using their expression
levels in log10 scale.
(Bottom) Two genes in different tranascripts are plotted using their expression
levels in log10 scale.

{\bf Figure \ref{figDotplotGlucoseOnly}. 
Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type and TW1 mutant-type grown in glucose.} 
Functional class ``Energy metabolism; Sugars'' contains 8 although 7 points are
shown, because two values of the 8 points are not different enough to be
displayed in two points.

{\bf Figure \ref{figDotplotUA159Only}. 
Distribution of functional classes of differntially expressed genes
in comparing UA159 wild-type grown in glucose and galactose.}

\clearpage{}

\section*{FIGURES}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 1. Cluster of 13 RNA-seq samples using genes expression
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<figClust,echo=FALSE,result=hide,fig=TRUE>>=
allSample <- newSmutans( smutansGenes )
smutans.de2Clust( allSample )
@
\begin{figure}[!h]
\centering
\includegraphics[width=.9\textwidth]{mss-lin-figClust}
\caption{}
\label{figClust}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 2. Distribution of transcript expression 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<figGrCombinedExpression,fig=TRUE,echo=FALSE>>=
l <- list(elementMetadata(smutansData.txGrCombined)$pileup)
bin <- c(as.integer(seq.int(0, 1000, length.out=15)), 50000)
multhist(l,breaks=bin,ylab="Frequency of Transcripts",xlab="Expression Level")
@
\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{mss-lin-figGrCombinedExpression}
\caption{}
\label{figGrCombinedExpression}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 3. Correlation of gene expression within and among transcripts
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<figExpressionTwoGene,fig=TRUE,echo=FALSE>>=
plot( log10(comp.exp.two.genes[1,]), log10(comp.exp.two.genes[2,]),
      xlab="Expression of first Gene in log10",
      ylab="Expression of second Gene in log10",
      pch=19,cex=0.5)
abline(b=1,lty=2)
@
<<figExpressionTwoGeneBetweenTranscript,fig=TRUE,echo=FALSE>>=
plot( log10(comp.exp.two.genes.between.transcript[1,]),
      log10(comp.exp.two.genes.between.transcript[2,]),
      xlab="Expression of first Gene in log10",
      ylab="Expression of second Gene in log10",
      pch=19,cex=0.5)
abline(b=1,lty=2)
@
\begin{figure}
\centering
\includegraphics[width=.6\textwidth]{mss-lin-figExpressionTwoGene}
\includegraphics[width=.6\textwidth]{mss-lin-figExpressionTwoGeneBetweenTranscript}
\caption{}
\label{figExpressionTwoGene}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 4. UA159 and TW1 in Glucose
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<figDotplotGlucoseOnly,fig=TRUE,echo=FALSE,width=12,height=6>>=
#par(mar=c(4.1, 3.1, 0.1, 0.1))
par(mai=c(0, 0, 0, 0))
y <- tableGlucoseOnly[,"Functional Class"]
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableGlucoseOnly[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotplot.
dotplot(xm[,"Functional Class"]~xm[,2],
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{mss-lin-figDotplotGlucoseOnly}
\caption{}
\label{figDotplotGlucoseOnly}
\end{sidewaysfigure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 5. Glucose and Galactose in UA159
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<figDotplotUA159Only,fig=TRUE,echo=FALSE,width=12,height=6>>=
par(mai=c(0, 0, 0, 0))
y <- tableUA159Only[,"Functional Class"]
z <- sapply(y, function(x) {x1=strsplit(as.character(x),"\\|\\|"); lapply(x1, trim)})
xm <- tableUA159Only[rep(seq(length(z)),sapply(z,length)),]
xm[,"Functional Class"] <- unlist(z)
# Use xm to plot a dotplot.
dotplot(xm[,"Functional Class"]~xm[,2], 
        main="",xlab="Log2 fold change",col="black",pch=88)
@
\begin{sidewaysfigure}
\centering
\includegraphics[width=0.9\textwidth]{mss-lin-figDotplotUA159Only}
\caption{}
\label{figDotplotUA159Only}
\end{sidewaysfigure}

\end{document}

### 2012-10-01 R/sara-add-annotation.R

@@ -1,0 +1,22 @@
+args <- commandArgs(trailingOnly = TRUE)                                           
+if (length(args) != 2)
+{
+  cat ("Rscript R/find-il-position.R IL1-1_SNP\n")
+  quit("no")
+}
+#args<- c("email/from/sara-palmer/090712/ua159csp-f-chris.csv", "email/to/sara-palmer/071412/smu86csp.csv")
+
+x <- read.csv(args[1],head=T)
+d <- read.csv(args[2],head=T)
+
+# SMU_1056's fake annotation is added for the sake of running the script without
+# error. Sara should be able to let me know the annotation.
+stopifnot(sum(d$id %in% x[,1]) == nrow(d))
+
+# Select 2nd column (OralGen), 5th column (gene name), 6th column function .
+d <- data.frame(d[,-1], oralgen.annotation=paste("=HYPERLINK(\"",x[match(d$id,x[,1]),4],"\",\"",x[match(d$id,x[,1]),2],"\")",sep=""),
+                        genename=x[match(d$id,x[,1]),5],
+                        some.annotation=x[match(d$id,x[,1]),6])
+                   
+write.csv(d,file=paste(sub(".csv","",args[2]),"-annotation.csv",sep=""),row.names=FALSE)
+

### 2012-10-01 R/sara-create-chris-gff.R

@@ -1,0 +1,29 @@
+library(rtracklayer)
+
+#args<- c("output/sara/gff/44.gff", "email/to/sara-palmer/080812/UA159-11VS1-SMU44.csv")
+args <- commandArgs(trailingOnly = TRUE)                                           
+if (length(args) != 2)
+{
+  cat ("Rscript R/sara-create-chris-gff.R output/sara/gff/44.gff email/to/sara-palmer/080812/UA159-11VS1-SMU44.csv\n")
+  quit("no")
+}
+
+gff.range <- import.gff3(args[1])
+gff.range.cds <- gff.range[gff.range$type=="CDS",]
+x <- gff.range.cds
+
+# x <- read.csv(args[1],head=T)
+d <- read.csv(args[2],head=T)
+
+# SMU_1056's fake annotation is added for the sake of running the script without
+# error. Sara should be able to let me know the annotation.
+stopifnot(sum(d$id %in% gff.range.cds$locus_tag) == nrow(d))
+
+# Select 2nd column (OralGen), 5th column (gene name), 6th column function .
+
+x.note <- unlist(lapply(x$Note,function(x) if(length(x)==0) NA else x))
+d <- data.frame(d[,-1],
+                note.annotation=x.note[match(d$id,x$locus_tag)],
+                product.annotation=x$product[match(d$id,x$locus_tag)])
+
+write.csv(d,file=paste(sub(".csv","",args[2]),"-annotation.csv",sep=""),row.names=FALSE)

### 2012-10-01 R/sara-nine-pattern.R

@@ -1,0 +1,60 @@
+args <- commandArgs(trailingOnly = TRUE)                                           
+if (length(args) != 2)
+{
+  cat ("Rscript R/find-il-position.R IL1-1_SNP\n")
+  quit("no")
+}
+#args<- c("email/to/sara-palmer/092412/ua159ph-annotation.csv", "email/to/sara-palmer/092412/smu21ph-annotation.csv")
+
+x <- read.csv(args[1])
+y <- read.csv(args[2])
+
+# Two-Fold Changes
+for (i in c("plus","minus","zero")) {
+  for (j in c("plus","minus","zero")) {
+    m <- x$padj < 0.1 & y$padj < 0.1
+    if (i == "plus" & j == "plus") {
+      m <- m & x$log2FoldChange > 1 & y$log2FoldChange > 1
+    } else if (i == "plus" & j == "minus") {
+      m <- m & x$log2FoldChange > 1 & y$log2FoldChange < -1
+    } else if (i == "plus" & j == "zero") {
+      m <- m & x$log2FoldChange > 1 & abs(y$log2FoldChange) < 1
+    } else if (i == "minus" & j == "plus") {
+      m <- m & x$log2FoldChange < -1 & y$log2FoldChange > 1
+    } else if (i == "minus" & j == "minus") {
+      m <- m & x$log2FoldChange < -1 & y$log2FoldChange < -1
+    } else if (i == "minus" & j == "zero") {
+      m <- m & x$log2FoldChange < -1 & abs(y$log2FoldChange) < 1
+    } else if (i == "zero" & j == "plus") {
+      m <- m & abs(x$log2FoldChange) < 1 & y$log2FoldChange > 1
+    } else if (i == "zero" & j == "minus") {
+      m <- m & abs(x$log2FoldChange) < 1 & y$log2FoldChange < -1
+    } else if (i == "zero" & j == "zero") {
+      m <- m & abs(x$log2FoldChange) < 1 & abs(y$log2FoldChange) < 1
+    }
+    m[is.na(m)] <- FALSE
+    d <- data.frame(id=x[m,"id"],
+                    x.baseMean=x[m,"baseMean"],
+                    x.baseMeanA=x[m,"baseMeanA"],
+                    x.baseMeanB=x[m,"baseMeanB"],
+                    x.foldChange=x[m,"foldChange"],
+                    x.log2FoldChange=x[m,"log2FoldChange"],
+                    x.padj=x[m,"padj"],
+                    y.baseMean=y[m,"baseMean"],
[…]

### 2012-10-01 sh/batch3.sh

@@ -99,6 +99,7 @@ function batch3-speciesfile {
   MINMAPQ=$(grep ^MINMAPQ\: $SPECIESFILE | cut -d":" -f2)
   MINTRIMQ=$(grep ^MINTRIMQ\: $SPECIESFILE | cut -d":" -f2)
   KEEPBAM=$(grep ^KEEPBAM\: $SPECIESFILE | cut -d":" -f2)
+  COUNTMODE=$(grep ^COUNTMODE\: $SPECIESFILE | cut -d":" -f2)
 }
 
 # Send input files to the remote machine.
@@ -110,7 +111,7 @@ ssh -x $CAC_USERHOST mkdir -p $RDATADIR
 ssh -x $CAC_USERHOST mkdir -p $RRUNANALYSIS
 scp $REFGENOMEFASTA $CAC_USERHOST:$RDATADIR
 scp $REFGENOMETXDB $CAC_USERHOST:$RDATADIR
-scp $REFGENOMERRNATXDB $CAC_USERHOST:$RDATADIR
+# scp $REFGENOMERRNATXDB $CAC_USERHOST:$RDATADIR
 for g in $FASTQFILES; do
   FASTQNUM=FASTQ\$(printf "%03d" \$g)
   FASTQFILE=$ROOTANALYSISDIR/\$(grep ^\$FASTQNUM\: $SPECIESFILE | cut -d":" -f2)
@@ -466,7 +467,7 @@ cat>>$BASEDIR/job-de.R<<EOF
 # Remove strands from both feature and BAM alignments.
 strand(feature.txnc) <- '*'
 strand(bv) <- '*'
-olap <- summarizeOverlaps(feature.txnc,bv,mode="IntersectionStrict")
+olap <- summarizeOverlaps(feature.txnc,bv,mode="$COUNTMODE")
 cl <- assays(olap)\$counts[,1]
 cl.cds <- cl[elementMetadata(feature.txnc)\$type=="CDS"]
 cl.nocds <- cl[elementMetadata(feature.txnc)\$type=="NOCDS"]
@@ -474,7 +475,7 @@ cl.tx <- cl[!elementMetadata(feature.txnc)\$type=="NG"]
 cl.ng <- cl[elementMetadata(feature.txnc)\$type=="NG"]
 
 strand(bv.multiple) <- '*'
-olap <- summarizeOverlaps(feature.txnc,bv.multiple,mode="IntersectionStrict")
[…]

### 2012-10-01 sh/copyright.sh

@@ -20,4 +20,9 @@ function copyright {
   less COPYING
 }
 
+function quit {
+  cat <<EOF
+RNA-seq Analysis - Bye!
+EOF
+}
 

### 2012-10-01 sh/deseq.sh

@@ -143,7 +143,7 @@ library(smutans)
 args <- commandArgs(trailingOnly = TRUE)
 if (length(args) != 3)
 {
-  cat ("Rscript job-simulate.R UA159 1SM1 out\n")
+  cat ("Rscript job-deseq.R UA159 1SM1 out\n")
   quit("yes")
 }
 sampleA <- args[1]

### 2012-10-01 sh/run.sh

@@ -76,6 +76,7 @@ source sh/batch3.sh
 source sh/deseq.sh
 source sh/goseq.sh
 source sh/mcl.sh
+source sh/sara-batch.sh
 
 #####################################################################
 # Read configuration file
@@ -91,80 +92,29 @@ SPECIESS=$(ls $ROOTANALYSISDIR/species|grep -v ^sim)
 # Menus
 #####################################################################
 PS3="Select the menu : "
-CHOICES=( # init-file-system \
-          choose-species \
-          ---ANNOTATION---\
+CHOICES=( choose-species \
           convert-gff2txdb \
-          ---BATCH---\
           batch3 \
-          ---DE---\
           deseq \
           goseq \
           mcl \
-          ---FASTQ---\
           fastq-sample \
+          sara-batch \
           warranty \
           copyright \
           quit )
-select CHOICE in ${CHOICES[@]}; do 
-  if [ "$CHOICE" == "" ];  then
-    echo -e "You need to enter something\n"
-    continue
-  elif [ "$CHOICE" == "init-file-system" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "choose-species" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "cp-genome" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "bwa-index-genome" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "bowtie-index-genome" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "bowtie-align" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "bowtie-refflat" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "segemehl-index-genome" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "bwa-align" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "bwa-align-per-fastq" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "bwa-samse" ]; then $CHOICE; break
-  elif [ "$CHOICE" == "bwa-mpileup" ]; then $CHOICE; break
[…]

### 2012-10-01 sh/sara-batch.sh

@@ -1,0 +1,58 @@
+###############################################################################
+# Copyright (C) 2012 Sang Chul Choi
+#
+# This file is part of RNASeq Analysis.
+# 
+# RNASeq Analysis is free software: you can redistribute it and/or modify
+# it under the terms of the GNU General Public License as published by
+# the Free Software Foundation, either version 3 of the License, or
+# (at your option) any later version.
+#
+# RNASeq Analysis is distributed in the hope that it will be useful,
+# but WITHOUT ANY WARRANTY; without even the implied warranty of
+# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+# GNU General Public License for more details.
+#
+# You should have received a copy of the GNU General Public License
+# along with RNASeq Analysis.  If not, see <http://www.gnu.org/licenses/>.
+###############################################################################
+
+function sara-batch {
+#  rm -f email/to/sara-palmer/080812/*-annotation.csv;
+#  for i in email/to/sara-palmer/080812/*.csv; do
+#    Rscript R/sara-add-annotation.R \
+#      email/from/sara-palmer/090712/ua159csp-f-chris.csv $i
+#  done
+
+#  rm -rf output/sara/gff; mkdir output/sara/gff
+#  for i in 44 20 21 56 69 57 86; do
+#    cat data/denovo/SMU${i}_allcontigs.gbf \
+#      | perl pl/bp_genbank2gff3.pl -in stdin -out stdout \
+#      | perl -e "while (<>) { if (/^>/) { last; } print; }" \
+#      > output/sara/gff/$i.gff
+#  done
+  Rscript R/sara-create-chris-gff.R \
+    output/sara/gff/44.gff email/to/sara-palmer/080812/UA159-11VS1-SMU44.csv
+  Rscript R/sara-create-chris-gff.R \
+    output/sara/gff/20.gff email/to/sara-palmer/080812/UA159-15JP3-SMU20.csv
+  Rscript R/sara-create-chris-gff.R \
+    output/sara/gff/21.gff email/to/sara-palmer/080812/UA159-1SM1-SMU21.csv
+  Rscript R/sara-create-chris-gff.R \
+    output/sara/gff/56.gff email/to/sara-palmer/080812/UA159-N29-SMU56.csv
+  Rscript R/sara-create-chris-gff.R \
[…]
